### 2020-09-11 paper_draft/main.tex
initial commit

\documentclass[11pt]{article}
\usepackage{apacite}
\usepackage{syntonly}
%\syntaxonly
%
%Margin - 1 inch on all sides
%
\usepackage[letterpaper]{geometry}
\usepackage{times}
\geometry{top=1.0in, bottom=1.0in, left=1.0in, right=1.0in}

%
%Doublespacing
%
\usepackage{setspace}
%\doublespacing

%
%Rotating tables (e.g. sideways when too long)
%
\usepackage{rotating}

%\usepackage{hyperref}
%
%Fancy-header package to modify header/page numbering (insert last name)
%
\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{} 
\chead{} 
\rhead{\thepage} 
\lfoot{} 
\cfoot{} 
\rfoot{} 
\renewcommand{\headrulewidth}{0pt} 
\renewcommand{\footrulewidth}{0pt} 
%To make sure we actually have header 0.5in away from top edge
%12pt is one-sixth of an inch. Subtract this from 0.5in to get headsep value
\setlength\headsep{0.5in}

%
%Works cited environment
%(to start, use \begin{workscited...}, each entry preceded by \bibent)
% - from Ryan Alcock's MLA style file
%
%\newcommand{\bibent}{\noindent \hangindent 40pt}
%\newenvironment{workscited}{\newpage \begin{center} References \end{center}}{\newpage }

\usepackage{tipa}
\newcommand{\nt}[1]{\textipa{[#1]}} % narrow transcription
\newcommand{\wt}[1]{\textipa{/#1/}} % wide transcription

\usepackage{graphicx} %for images
\graphicspath{{images/}}
\usepackage{caption}

\usepackage{enumitem} %customize list numberings

%
%Begin document
%

\title{Cross-language Phonetic Transfer in Bilingual Speech}
\date{}

\begin{document}

\maketitle

\bibliographystyle{apacite}

\section{Introduction}
\subsection{Literature review}

Research on the phonological systems of bi/multilingual speakers suggests that speakers maintain separate, but phonologically linked sound categories for both their languages. These show cross-language influence in both perception and production. There is a lot of work on-
\begin{enumerate}[label=(\roman*)]

\item Long-term influence due to acquisition of multiple languages-- affects not only L2, but also L1. Bidirectional influence-- shapes phonetic categories in bilinguals, which differ from that of monolingual speakers\cite{caramazza1973acquisition, flege1987production, guion2003vowel}.

\item Temporary phonetic drift due to language exposure \cite{sancier1997gestural,chang2012rapid,tobin2017phonetic}
\end{enumerate}
These studies- look at the two languages of a bilingual speaker as used separately.
\citeA{grosjean1994going} first studied dynamic phonetic influence due to simultaneous activation of both languages of a bilingual speaker. Although they did not find any evidence of cross-language influence, and concluded that the switch between language systems is immediate and complete at the phonetic level, subsequent studies using similar paradigms have reported different results.

\paragraph{}These compare bilingual speakers’ productions during mixed-language usage to their own productions in a single language. The findings vary greatly—studies have shown unidirectional influence of L1 on L2 \cite{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \cite{tsui2019impact,elias2017effects}, bidirectional convergence \cite{bullock2009trying, olson2016role}, divergence \cite{bullock2009trying,vsimavckova2018patterns} and no influence \cite{muldner2019phonetics,schwartz2015language}. Partly attributed to differences in methodology, populations, and language pairs, along with variation in individual switching strategies.

\paragraph{}Overall, the results suggest that cross-language phonetic influence in bilinguals increases during mixed-language processing, and both L1 and L2 sounds can show gradient phonetic shift in production as a result. The extent of shift is mediated by language-specific (sound system), context-specific (eg. language mode), and speaker-specific (proficiency, dominance) constraints. Studies eliciting spontaneous bilingual speech have also noted the influence of social factors \cite{khattab2009phonetic}.

\paragraph{}Most studies of this phenomenon have focused on consonant sounds, using differences in typical VOT of stop consonants to measure phonetic shift. Studies on vowel sounds are relatively fewer and much more recent, but suggest that vowels too show dynamic cross-language transfer. The studies so far report varied results: \citeA{elias2017effects,simonet2014phonetic,simonet2020increased} have reported L2 influence on L1 in code-switching (CS) tasks, whereas \citeA{muldner2019phonetics} found no significant increase in cross-language influence. As far as we are aware, there is no study yet on L1 influence on L2 vowels during mixed language use.

\paragraph{}The majority of studies reviewed so far have used CS (both spontaneous and scripted) to induce co-activation of languages. Other methods used include delayed repetition \cite{simonet2014phonetic}, reading single- or mixed-language word lists \cite{simonet2020increased}, and interpreting across languages \cite{vsimavckova2015immediate,vsimavckova2018patterns}. \citeA{olson2013bilingual} argues— CS (switching languages within a discourse, in connected speech) allows for planning and pragmatic considerations, which independently affect production, and might therefore obscure the underlying phonetic interaction due to mixed-language processing. They propose cued picture-naming as an alternative paradigm to induce switches in language—it lacks discourse context and predictability, so offers a more controlled way to observe phonetic effects due to language-processing manipulations. Following this, other studies \cite{goldrick2014language,tsui2019impact} have used this paradigm to study dynamic phonetic transfer in bilinguals.

\paragraph{}While phonetic transfer is usually measured at the point of the language-switch, previous  research has noted that the observed transfer could potentially have been triggered by two different sources— local phonetic processing (the switch itself), or global language co-activation (language mode \cite{grosjean1998studying} or context of the task). To examine this, several studies have manipulated language mode /context in addition to language-switching. Excepting one study (\citeNP{olson2013bilingual}; language context operationalized as proportion of words from each language during a test block) which found the balanced language context to mitigate transfer, results generally suggest that cross-language transfer due to switching either increases in a mixed language mode \cite{simonet2014phonetic}, or remains unaffected by it \cite{schwartz2015language,vsimavckova2015immediate,olson2016role,simonet2020increased}.

\subsection{Present study}
\paragraph{}Dynamic phonetic transfer has been shown to be mediated by language-specific constraints, and vary across groups. However, existing studies have largely focused on a limited set of phonologically related language pairs. We extend the research to a new language pair and population—Bengali-English bilingual speakers in India. Long history of language contact, rapid CS in this population is common, socially unmarked. Therefore, we expect any phonetic effects observed in a CS paradigm, to be at least partially representative of what happens during daily language use experience.

\paragraph{}The small body of existing work on transfer in vowels so far has reported L2 influence on L1 due to language mixing. Many earlier studies on VOT, however, have shown transfer from L1 to L2. Building on the current line of questioning about the nature of transfer in vowels, we ask whether simultaneous co-activation causes an increased L1 influence on the production of L2 vowels. If this influence is characterized as accentedness \cite{goldrick2014language}, then-- can ask whether language mixing temporarily leads more accented L2 productions.

\paragraph{}The results of \citeA{simonet2014phonetic} and \citeA{elias2017effects} suggest that phonetic interaction during switching affects vowel categories which are common across the two languages, as well as those which are not. For the purpose of this study, two English vowels were chosen:
(i) \wt{\ae}--exists as a category in both Bengali and General Indian English (GIE)\cite{masica1972sound}, and (ii) \wt{2}-- exists as a category in English, but not Bengali. It is assumed that the lack of a mid-central vowel in Bengali allows greater room for variablity due to shift (if any) in \nt{2} compared to \nt{\ae}. This criteria has been used to explain differences in the extent and direction of cross-language transfer in numerous studies measuring VOT. Therefore, we ask whether a parallel difference in the extent of shift is found in vowels.

\paragraph{}Both CS and cued picture naming have been used in different studies to elicit mixed-language productions, and have shown a similar range of variation in results. Following from the ongoing debates about  switching paradigms, we collect production data from the same participant pool using both test paradigms, allowing for a direct comparison of responses.

\paragraph{}Since previous work largely shows no inhibitory effect of a mixed language mode on temporary phonetic effects of language switching, switched and non-switched utterances were not presented in separate test blocks in this study (cf \citeNP{bullock2009trying,muldner2019phonetics,olson2016role}, etc), but rather alternated randomly within each block. Instructions were given in mixed language. Thus, maintained a bilingual language mode throughout the experiment. Also more ecologically valid, because daily language experience of the target population-- not likely to provide stable, monolingual ambient language conditions; constantly encounter mixed-language input (both auditory and visual).

\paragraph{}The research questions can thus be summarized as follows:
\begin{enumerate}[label=(\roman*)]
	\item Does mixed-language processing temporarily increase L1 influence on L2 vowel productions?
	\item If so, does the language-specific distribution of vowels constrain the extent of shift?
	\item Do two commonly used paradims-- code-switching and cued picture-naming, elicit different degrees of cross-langauge transfer?
\end{enumerate}

\section{Methodology} %ID will write this
Design: 10 participants*20 target words*2 utterance types *2 tasks *4 iterations = 3200 items

\subsection{Participants}

\begin{itemize}
	\item 10 Bengali-English bilingual speakers (5 female)
	\item age range-- 19 to 28
	\item selected on the basis of response to a Language Background Questionnaire, to match for language experience and LSRW skills in L1 and L2 
\end{itemize}

\subsection{Stimuli}
\begin{itemize}
	\item Target words: 20 monosyllabic English words containing \nt{2} or \nt{ae} (not lexicalized loanwords)
	\item Words with pre-vocalic voiced plosives (/b/ and /d/) were selected, to minimize coarticulatory effects on the target vowel
	\item Filler words: 10 monosyllabic English words not containing the target vowels
\end{itemize}

\subsection{Procedure}
\begin{enumerate}[]
	\item Cued picture-naming: Participants were presented with slides in the following sequence-
	\begin{itemize}
		\item Language cue- a word in either English or Bengali orthography
		\item a picture displayed for 50ms; named by the participant in either English or Bengali, depending on the language cue 
		\item Target word- English word printed in English orthography; read out by the participant
		\item Distracter math problem
	\end{itemize}
Since all target words were in English, language cue was used to manipulate the preceding word, giving target words in non-switched and switched utterances. Engish and Bengali orthographies are visually distinct, so it was used to cue language.
  \item Code-switching: Each slide contained one target word embedded in either an English (unilingual) or a Bengali (bilingual; target produced as a code switch from Bengali) carrier sentence; read out by the participant. Carrier sentences:
  \begin{itemize}
  	\item Unilingual sentence: That is a yellow [Target Word].
  	\item Mixed-language sentence: \textipa{o-ta \ae k-ta ka:lo} [Target Word].
  \end{itemize}

  The sound preceding the target word was uniform across al sentences (the final sound in ‘yellow’ is assimilated to [o] in Indian English), to minimize the possible effects of coarticulation on the target word. A specifier-noun construction was used to ensure identical word order in both languages, giving uniform prosodic context for the target word. For bilingual utterances, mixed orthography was used. \\
  Four iterations of each task were recorded. Every target word appeared twice during a test block; once in a switched and once in a non-switched context. Order of items and conditions (switched vs non-switched) was randomized within each trial. Participants alternated between the two tasks.

\end{enumerate}
\subsection{Acoustic analysis}
 The first iteration of each task was treated as a trial. An additional 67 tokens (2.79\%) were excluded due to errors in presentation/production, giving 2333 target word items for analysis. Recodings were manually segmented and annotated in PRAAT \cite{boersma2016praat}, F1 and F2 of the target vowels \nt{2, ae} were measured at 5\% and 50\% into the vowels.

\section{Results} %ID will write this
\begin{itemize}
	\item A linear mixed effects model with utterance type (unilingual vs bilingual), vowel, and gender as fixed effects shows a significant effect of utterance type on lobanov normalized F1 at 5\% (p=0.0096). Normalized F1 is higher in the bilingual utterances.

	\item No effect of utterance type on F1 at 50\% of the vowel, or on normalized F2 values.

	\item \nt{2} is lowered to a greater extent than \nt{\ae}.

	\item Significant effect of task

	\item \wt{d}-initial target words show more lowering than \wt{b}-initial words %I had subsetted the data by onset consonant, and saw this. Not sure what to make of it. The b-initial set by itself doesn't show significant lowering.

\end{itemize}

\section{Discussion}
Results are in line with previous studies which find that CS between languages does not lead to a parallel switch in phonology. Instead, gradient, momentary phonetic shifts are observed.
\begin{itemize}
	\item Both the vowels are lowered (shift towards related L1 category) in the bilingual utterances compared to speakers’ own unilingual productions, showing increased cross-language influence during mixed-language processing. 
	\item Lowering at 5\% of the vowel, but not at 50\%, suggests that only the initial part of the vowel is receptive to dynamic shift, not the steady state
	\item No change in F2
	\item More shift in \nt{2} compared to \nt{\ae}-- suggests that the extent of shift is constrained by the distribution of vowels and room for variability. Since we don't have monolingual Bengali data, can't say anything about the extent of shift relative to the initial distance between the categories. 
	\item CS in connected speech leads to greater cross-language phonetic interaction than unplanned switches through a cued picture naming task, suggesting that at least part of the observed effects in CS paradigms can be attributed to discourse awareness and planning.
\end{itemize}

\section{Conclusion}
Mixed language processing temporarily increases cross-language phonetic transfer, causing L2 vowels to converge towards L1. The extent of shift is mediated by language-specific distribution of sound categories, and is greater in connected speech.

\bibliography{references}

\end{document}

### 2020-09-11 RCode/bengali_phon_acco.R
adding R code

@@ -1,0 +1,197 @@
+attach(bengali_phon_acco_raw)
+library(lattice)
+library(plyr)
+library(phonR)
+library(lme4)
+library(ggplot2)
+library(dplyr)
+
+normF1 <- with(bengali_phon_acco_raw, normLobanov(F1_5), group=Subject)
+normF2 <- with(bengali_phon_acco_raw, normLobanov(F2_5), group=Subject)
+normF1mid <- with(bengali_phon_acco_raw, normLobanov(F1_mid), group=Subject)
+normF2mid <- with(bengali_phon_acco_raw, normLobanov(F2_mid), group=Subject)
+
+with(bengali_phon_acco_raw, plotVowels(normF1, normF2, group=Vowel, pch.tokens = Vowel))
+par(mfrow = c(1, 1))
+with(bengali_phon_acco_raw, plotVowels(normF1, normF2, plot.tokens = TRUE, pch.tokens = Vowel, 
+                    plot.means = TRUE, pch.means = Context, 
+                      ellipse.line = TRUE, pretty = TRUE))
+boxplot(normF1~Context*Vowel, data=bengali_phon_acco_raw)
+
+vowel_vt <- bengali_phon_acco_raw %>%
+  filter(Vowel == "ʌ")
+task_p <- subset(bengali_phon_acco_raw, (vowel=="ʌ"))
+p <- ggplot(bengali_phon_acco_raw, aes(x=Context, y=normF1)) + 
+  geom_boxplot() + xlab("Language context: B (Bengali) and E (English) and Tasks: S (Sentence) and P (Picture naming)") + ylab("Lobanov Normalized F1 at 5% of the vowel")
+p + facet_grid(Vowel~Task)
+
+p <- ggplot(bengali_phon_acco_raw, aes(x=Context, y=normF1)) + 
+  geom_boxplot() + xlab("Language context: B (Bengali) and E (English) and Tasks: S (Sentence) and P (Picture naming)") + ylab("Normalized F1")
+p + facet_grid(Vowel~Task)
+
+ggplot(bengali_phon_acco_raw, aes(x=context, y=normF1, fill=vowel))
+geom_boxplot()
+
+phon_acco.model = lmer(normF1mid ~ Vowel + Task + Context + Gender + (1+Context|Word), data=bengali_phon_acco_raw, REML=FALSE)
[…]

### 2020-09-11 paper_draft/main.tex
Added author info

@@
 \documentclass[11pt]{article}
-\usepackage{apacite}
+% \usepackage{apacite}
 \usepackage{syntonly}
@@
 
+\author{Auromita Mitra, Indranil Dutta}
+
 \begin{document}
@@
 
-\bibliographystyle{apacite}
+% \bibliographystyle{apacite}
 
@@
 
-\item Long-term influence due to acquisition of multiple languages-- affects not only L2, but also L1.
+{}\item Long-term influence due to acquisition of multiple languages-- affects not only L2, but also L1.
 Bidirectional influence-- shapes phonetic categories in bilinguals, which differ from that of monolingual speakers\cite{caramazza1973acquisition, flege1987production, guion2003vowel}.
@@
 These studies- look at the two languages of a bilingual speaker as used separately.
-\citeA{grosjean1994going} first studied dynamic phonetic influence due to simultaneous activation of both languages of a bilingual speaker.
+\cite{grosjean1994going} first studied dynamic phonetic influence due to simultaneous activation of both languages of a bilingual speaker.
 Although they did not find any evidence of cross-language influence, and concluded that the switch between language systems is immediate and complete at the phonetic level, subsequent studies using similar paradigms have reported different results.
 
-\paragraph{}These compare bilingual speakers’ productions during mixed-language usage to their own productions in a single language.
+These compare bilingual speakers’ productions during mixed-language usage to their own productions in a single language.
 The findings vary greatly—studies have shown unidirectional influence of L1 on L2 \cite{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \cite{tsui2019impact,elias2017effects}, bidirectional convergence \cite{bullock2009trying, olson2016role}, divergence \cite{bullock2009trying,vsimavckova2018patterns} and no influence \cite{muldner2019phonetics,schwartz2015language}.
@@
 
-\paragraph{}Overall, the results suggest that cross-language phonetic influence in bilinguals increases during mixed-language processing, and both L1 and L2 sounds can show gradient phonetic shift in production as a result.
+Overall, the results suggest that cross-language phonetic influence in bilinguals increases during mixed-language processing, and both L1 and L2 sounds can show gradient phonetic shift in production as a result.
 The extent of shift is mediated by language-specific (sound system), context-specific (eg.
@@
 
-\paragraph{}Most studies of this phenomenon have focused on consonant sounds, using differences in typical VOT of stop consonants to measure phonetic shift.
+Most studies of this phenomenon have focused on consonant sounds, using differences in typical VOT of stop consonants to measure phonetic shift.
 Studies on vowel sounds are relatively fewer and much more recent, but suggest that vowels too show dynamic cross-language transfer.
-The studies so far report varied results: \citeA{elias2017effects,simonet2014phonetic,simonet2020increased} have reported L2 influence on L1 in code-switching (CS) tasks, whereas \citeA{muldner2019phonetics} found no significant increase in cross-language influence.
+The studies so far report varied results: \cite{elias2017effects,simonet2014phonetic,simonet2020increased} have reported L2 influence on L1 in code-switching (CS) tasks, whereas \cite{muldner2019phonetics} found no significant increase in cross-language influence.
 As far as we are aware, there is no study yet on L1 influence on L2 vowels during mixed language use.
 
-\paragraph{}The majority of studies reviewed so far have used CS (both spontaneous and scripted) to induce co-activation of languages.
+The majority of studies reviewed so far have used CS (both spontaneous and scripted) to induce co-activation of languages.
 Other methods used include delayed repetition \cite{simonet2014phonetic}, reading single- or mixed-language word lists \cite{simonet2020increased}, and interpreting across languages \cite{vsimavckova2015immediate,vsimavckova2018patterns}.
-\citeA{olson2013bilingual} argues— CS (switching languages within a discourse, in connected speech) allows for planning and pragmatic considerations, which independently affect production, and might therefore obscure the underlying phonetic interaction due to mixed-language processing.
-They propose cued picture-naming as an alternative paradigm to induce switches in language—it lacks discourse context and predictability, so offers a more controlled way to observe phonetic effects due to language-processing manipulations.
+\cite{olson2013bilingual} argues- CS (switching languages within a discourse, in connected speech) allows for planning and pragmatic considerations, which independently affect production, and might therefore obscure the underlying phonetic interaction due to mixed-language processing.
+They propose cued picture-naming as an alternative paradigm to induce switches in language-it lacks discourse context and predictability, so offers a more controlled way to observe phonetic effects due to language-processing manipulations.
 Following this, other studies \cite{goldrick2014language,tsui2019impact} have used this paradigm to study dynamic phonetic transfer in bilinguals.
 
-\paragraph{}While phonetic transfer is usually measured at the point of the language-switch, previous  research has noted that the observed transfer could potentially have been triggered by two different sources— local phonetic processing (the switch itself), or global language co-activation (language mode \cite{grosjean1998studying} or context of the task).
+While phonetic transfer is usually measured at the point of the language-switch, previous research has noted that the observed transfer could potentially have been triggered by two different sources— local phonetic processing (the switch itself), or global language co-activation (language mode \cite{grosjean1998studying} or context of the task).
 To examine this, several studies have manipulated language mode /context in addition to language-switching.
-Excepting one study (\citeNP{olson2013bilingual}; language context operationalized as proportion of words from each language during a test block) which found the balanced language context to mitigate transfer, results generally suggest that cross-language transfer due to switching either increases in a mixed language mode \cite{simonet2014phonetic}, or remains unaffected by it \cite{schwartz2015language,vsimavckova2015immediate,olson2016role,simonet2020increased}.
+Excepting one study (\cite{olson2013bilingual}; language context operationalized as proportion of words from each language during a test block) which found the balanced language context to mitigate transfer, results generally suggest that cross-language transfer due to switching either increases in a mixed language mode \cite{simonet2014phonetic}, or remains unaffected by it \cite{schwartz2015language,vsimavckova2015immediate,olson2016role,simonet2020increased}.
 
@@
 
-\paragraph{}The results of \citeA{simonet2014phonetic} and \citeA{elias2017effects} suggest that phonetic interaction during switching affects vowel categories which are common across the two languages, as well as those which are not.
+\paragraph{}The results of \cite{simonet2014phonetic} and \cite{elias2017effects} suggest that phonetic interaction during switching affects vowel categories which are common across the two languages, as well as those which are not.
 For the purpose of this study, two English vowels were chosen:
@@
 
+\bibliographystyle{unsrt}
+
 \bibliography{references}

### 2020-10-15 paper_draft/main.tex
auromita's edits- introduction

@@
 \documentclass[11pt]{article}
-% \usepackage{apacite}
+
 \usepackage{syntonly}
@@
 \geometry{top=1.0in, bottom=1.0in, left=1.0in, right=1.0in}
-
+\usepackage{apacite}
 %
@@
 
-\title{Cross-language Phonetic Transfer in Bilingual Speech}
+\title{Mixed language processing increases cross-language phonetic transfer in Bengali-English Bilinguals}
 \date{}
@@
 
-% \bibliographystyle{apacite}
+%\bibliographystyle{apacite}
 
-\section{Introduction}
-\subsection{Literature review}
+\section{Phonetic transfer}
 
 Research on the phonological systems of bi/multilingual speakers suggests that speakers maintain separate, but phonologically linked sound categories for both their languages.
-These show cross-language influence in both perception and production.
-There is a lot of work on-
+These show cross-language influence in both perception and production
+%citation
+.
+In production, this kind of influence (variously termed transfer, drift, accommodation, interference-- used interchangeably here to indicate any interaction between two sets of phonetic norms) has been studied in two conditions:
+(i) While a bilingual speaker is operating in any one of their languages.
+These studies often compare the speech of bilinguals to monolingual norms, and view transfer as changes to long-term memory representations as a result of acquiring an L2 \cite{guion2003vowel,caramazza1973acquisition,flege1987production}; (ii) When both languages of a bilingual speaker are co-activated.
+These compare productions during mixed-language processing to participants' own productions while using a single language.
+The resulting transfer is variously thought to involve short-term memory, online processing costs, language mode, and context-awareness.
+This is the focus of the present study.
+
+\paragraph{}Studies of the phonetic effects of mixed-language production are relatively fewer, and much more recent.
+Existing work has largely focused on a few pairs of phonologically related languages, and used temporal properties of consonants (in particular, VOT) to measure transfer.
+However, great variation in results (cf sec.\ref{asymmetries} ) suggests that data from a wider variety of populations, language pairs,  and phonetic features is required in order to make meaningful generalizations.
+There is no work yet on phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language.
+Nevertheless, widespread multilingualism in this part of the world suggests that phonetic
+%line--rephrase
+behavior in these populations can improve our understanding of cross-language interactions at the phonetic level, as they are more likely to reflect real-world experience with mixed-language processing.
+
+\paragraph{} The present study looks at phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India.
+We measure the spectral properties (F1 and F2) of two English vowels to check if L1 influence on L2 increases during mixed-language use relative to a participant's unilingual baseline production.
+Mixed-language data is elicited in two switching paradigms-- code-switching and cued picture-naming.
+The results show an effect of dynamic transfer on L2 vowels for the first time, and a difference in the extent of transfer between the two paradigms.
+The findings are discussed in light of recent proposals about asymmetries in transfer effects, and debates about the role of connected speech in enhancing cross-language interaction.
+
+\paragraph{}The next section gives background information on Bengali and English in India, focusing on vowel systems.
+The rest of the introduction  discusses causes, duration, and possible sources of asymmetries in transfer reported in previous studies, paradigms used, and the design of the present study in light of these.
+
+\subsection{Bengali and English in India}
+
+\subsubsection{Demography} 
+\paragraph{}Bengali (also, Bangla) is an Indo-Aryan language spoken in Indian and Bangladesh.
+In India, more than 97 million people speak Bengali as a first language (Census of India, 2011), mostly in the state of West Bengal.
+A majority of this population also speaks other additional languages.
+
+\paragraph{}Indian English (IE) refers to the variety of English that has developed in the Indian subcontinent.
+In India, it is spoken as an L2 by 129 million people (Census of India, 2011).
+English is one of the two official languages, used in education, law, media, as a lingua franca mainly for an educated elite in most metropolitan regions, and carries a high prestige value \cite{tollefson2014language}.
+%citations 
+In spite of the different L1s of its speakers, recent literature suggests that IE has a target phonology that is distinct from any of these, as well as from other native varieties of English (see \citeA{sirsa2013effects} for a review).
+Thus, regional variations are seen as the result of L1-influence on a common underlying target.
+Early attempts to describe IE phonologically (General Indian English or GIE, \cite{masica1972sound}) were primarily attempts at standardization for pedagogical purposes.
+Excepting a small minority of specific communities \cite{pandey201517}, there are hardly any L1 speakers of IE.\\
+
+\paragraph{}Considering these facts, the present study focuses on short-term phonetic interaction during mixed-language use, because:
 \begin{enumerate}[label=(\roman*)]
+	\item Since the population is multilingual, we expect long-term representations to be affected by multiple languages. 
+	\item Given that examples of an IE phonology without L1 `influence' are very rare, it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a given `baseline' condition.
+\end{enumerate}
 
-{}\item Long-term influence due to acquisition of multiple languages-- affects not only L2, but also L1.
-Bidirectional influence-- shapes phonetic categories in bilinguals, which differ from that of monolingual speakers\cite{caramazza1973acquisition, flege1987production, guion2003vowel}.
+\subsubsection{Vowel systems} \label{vowel systems}
+The vowel inventory of Western Bangla (the variety spoken by the participants of the present study) consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \cite{garry2001facts}.
+Note that there is no mid-central vowel.
+Of relevance to this study, the category \nt{\ae} in Bengali is lower than the corresponding English category \cite{chatterji1921bengali}.\\
+% citation \cite{bhattacharyarelated}
 
-\item Temporary phonetic drift due to language exposure \cite{sancier1997gestural,chang2012rapid,tobin2017phonetic}
+The vowel system of IE contains the monopthongs \nt{I, i, E, e, \ae, @/2, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \cite{wells1982accents, masica1972sound}.
+Their distribution in the F1XF2 space is shown in (insert table 1.7, Pandey).
+There is a single mid-central vowel corresponding to the categories \nt{2,@,3:}, which are treated as distinct in most native varieties of English.
+Since the English items used in this study are traditionally transcribed with \nt{2}, we use this symbol to indicate the mid-central vowel throughout.
+%In IE, the dipthong \nt{@U} is realized as \nt{o:}.
+
+\subsection{What causes transfer and what does it affect?}
+
+Many studies have distinguished between changes in memory \textsc{representations} due to the acquisition of multiple sound systems (cf SLM \citeNP{flege1995second,flege2007language}; PAM-L2 \citeNP{best2007nonnative}), and interaction during \textsc{processing} (transfer vs interference; \citeNP{grosjean2012attempt}, competence vs performance interference; \citeNP{paradis1993linguistic}).
+Given their transient nature (cf sec.\ref{duration}), dynamic changes in production during mixed-language use are generally thought to involve the latter \cite{elias2017effects,simonet2014phonetic}.
+What triggers this interaction?
+\citeA{olson2016role} argues that while cross-language phonetic effects are largely measured at the point of the switch, it could have two potential sources:
+\begin{enumerate}
+	\item The local point of switch itself
+	\item Global co-activation of two languages -- bilingual language mode \cite{grosjean1998studying} 
 \end{enumerate}
-These studies- look at the two languages of a bilingual speaker as used separately.
-\cite{grosjean1994going} first studied dynamic phonetic influence due to simultaneous activation of both languages of a bilingual speaker.
-Although they did not find any evidence of cross-language influence, and concluded that the switch between language systems is immediate and complete at the phonetic level, subsequent studies using similar paradigms have reported different results.
 
-These compare bilingual speakers’ productions during mixed-language usage to their own productions in a single language.
-The findings vary greatly—studies have shown unidirectional influence of L1 on L2 \cite{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \cite{tsui2019impact,elias2017effects}, bidirectional convergence \cite{bullock2009trying, olson2016role}, divergence \cite{bullock2009trying,vsimavckova2018patterns} and no influence \cite{muldner2019phonetics,schwartz2015language}.
-Partly attributed to differences in methodology, populations, and language pairs, along with variation in individual switching strategies.
+A number of studies have specifically manipulated language mode, both in the presence and absence of switching.
+Overall, results suggest that (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \cite{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use -- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \cite{olson2016role,tsui2019impact} or spontaneous conversation \cite{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
+(iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \cite{olson2016role} found no additive effects, \cite{olson2013bilingual} found the balanced language context to inhibit transfer compared to unbalanced contexts.
+Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in separate monolingual test block, separated by a few hours to days \cite{schwartz2015language, bullock2009trying,antoniou2011inter, elias2017effects,vsimavckova2015immediate,vsimavckova2018patterns}.
 
-Overall, the results suggest that cross-language phonetic influence in bilinguals increases during mixed-language processing, and both L1 and L2 sounds can show gradient phonetic shift in production as a result.
-The extent of shift is mediated by language-specific (sound system), context-specific (eg.
-language mode), and speaker-specific (proficiency, dominance) constraints.
-Studies eliciting spontaneous bilingual speech have also noted the influence of social factors \cite{khattab2009phonetic}.
+\paragraph{}\citeA{grosjean1998studying} suggests that various aspects of the communicative setting, including exposure to (spoken or written) stimuli in multiple languages, and awareness of the interlocutor's being bilingual, could trigger a bilingual mode.
+Since the participants in this study are in an environment which largely contains mixed-language input, multilingual interlocutors, and no stable ambient language, we expect that phonetic transfer during daily language use, if any, takes place in a bilingual mode.
+Thus to preserve ecological validity, we elicited both unilingual and switched utterances in a bilingual language mode (cf section Methodology).
+Any observed differences in this paradigm would result from interaction during online processing.
 
-Most studies of this phenomenon have focused on consonant sounds, using differences in typical VOT of stop consonants to measure phonetic shift.
-Studies on vowel sounds are relatively fewer and much more recent, but suggest that vowels too show dynamic cross-language transfer.
-The studies so far report varied results: \cite{elias2017effects,simonet2014phonetic,simonet2020increased} have reported L2 influence on L1 in code-switching (CS) tasks, whereas \cite{muldner2019phonetics} found no significant increase in cross-language influence.
-As far as we are aware, there is no study yet on L1 influence on L2 vowels during mixed language use.
+In a bilingual mode, both language systems are expected to be (nearly) equally accessible throughout the test block.
+Thus, a consistent difference between switched and non-switched tokens in such a paradigm would be possible only if the effects were highly localized (if not, we should expect a gradual convergence over the course of the experiment).
+This is discussed in the next section.
 
-The majority of studies reviewed so far have used CS (both spontaneous and scripted) to induce co-activation of languages.
-Other methods used include delayed repetition \cite{simonet2014phonetic}, reading single- or mixed-language word lists \cite{simonet2020increased}, and interpreting across languages \cite{vsimavckova2015immediate,vsimavckova2018patterns}.
-\cite{olson2013bilingual} argues- CS (switching languages within a discourse, in connected speech) allows for planning and pragmatic considerations, which independently affect production, and might therefore obscure the underlying phonetic interaction due to mixed-language processing.
-They propose cued picture-naming as an alternative paradigm to induce switches in language-it lacks discourse context and predictability, so offers a more controlled way to observe phonetic effects due to language-processing manipulations.
-Following this, other studies \cite{goldrick2014language,tsui2019impact} have used this paradigm to study dynamic phonetic transfer in bilinguals.
+\subsection{Duration: How long do effects last?} \label{duration}
 
-While phonetic transfer is usually measured at the point of the language-switch, previous research has noted that the observed transfer could potentially have been triggered by two different sources— local phonetic processing (the switch itself), or global language co-activation (language mode \cite{grosjean1998studying} or context of the task).
-To examine this, several studies have manipulated language mode /context in addition to language-switching.
-Excepting one study (\cite{olson2013bilingual}; language context operationalized as proportion of words from each language during a test block) which found the balanced language context to mitigate transfer, results generally suggest that cross-language transfer due to switching either increases in a mixed language mode \cite{simonet2014phonetic}, or remains unaffected by it \cite{schwartz2015language,vsimavckova2015immediate,olson2016role,simonet2020increased}.
+Studies which compare bilinguals' phonological systems with monolinguals see them as relatively stable over time.
+The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods, \cite<e.g.>{guion2003vowel, caramazza1973acquisition}.
+However, longitudinal studies (citations) and between-subject comparisons of bilingual speakers with different durations of L2 exposure suggest that the interaction between the sound systems is dynamic.
+Over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts-- e.g.
+\citeA{bohn1992production}
+%found that an approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language transfer.
+These have been interpreted as changes to long-term phonological representations.
 
-\subsection{Present study}
-\paragraph{}Dynamic phonetic transfer has been shown to be mediated by language-specific constraints, and vary across groups.
-However, existing studies have largely focused on a limited set of phonologically related language pairs.
-We extend the research to a new language pair and population—Bengali-English bilingual speakers in India.
-Long history of language contact, rapid CS in this population is common, socially unmarked.
-Therefore, we expect any phonetic effects observed in a CS paradigm, to be at least partially representative of what happens during daily language use experience.
+Moreover, changes due to transfer are not unidirectional or irreversible.
+\citeA{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language.
+This not only showed that cross-language interaction can be triggered in an order of months, but also that the effects can be reversed within a similar time range.
+A recent study by \citeA{tobin2017phonetic} found comparable effects in an even shorter duration (2-4 weeks).
+A short-term longitudinal study by \citeA{chang2012rapid} showed that over the course of the first five weeks of learning an L2, there was a gradual convergence of L1 towards L2.
+For some sounds and features (though not others, cf sec.\ref{asymmetry between sounds}), transfer was additive over time.
+How long these effects last in the absence of regular L2 input was not tested.
+% if effects are cumulative, then the difference between switched and non-switched tokes should be lesser in later test blocks 
+(previous two paragraphs needed?)
 
-\paragraph{}The small body of existing work on transfer in vowels so far has reported L2 influence on L1 due to language mixing.
-Many earlier studies on VOT, however, have shown transfer from L1 to L2.
-Building on the current line of questioning about the nature of transfer in vowels, we ask whether simultaneous co-activation causes an increased L1 influence on the production of L2 vowels.
-If this influence is characterized as accentedness \cite{goldrick2014language}, then-- can ask whether language mixing temporarily leads more accented L2 productions.
+\paragraph{}Studies looking at mixed language use within an utterance mostly analyze only the switched (target) token.
+Thus, very few direct measurements of the duration of transfer effects are available.
+One study which measured this \cite{bullock2009trying} did not find any residual effects on the matrix language following the switch, suggesting that transfer during code-switching is localized.
 
-\paragraph{}The results of \cite{simonet2014phonetic} and \cite{elias2017effects} suggest that phonetic interaction during switching affects vowel categories which are common across the two languages, as well as those which are not.
-For the purpose of this study, two English vowels were chosen:
-(i) \wt{\ae}--exists as a category in both Bengali and General Indian English (GIE)\cite{masica1972sound}, and (ii) \wt{2}-- exists as a category in English, but not Bengali.
-It is assumed that the lack of a mid-central vowel in Bengali allows greater room for variablity due to shift (if any) in \nt{2} compared to \nt{\ae}.
-This criteria has been used to explain differences in the extent and direction of cross-language transfer in numerous studies measuring VOT.
-Therefore, we ask whether a parallel difference in the extent of shift is found in vowels.
+More indirect evidence for this comes from studies which have elicited switched and non-switched tokens in the same test block \cite{tsui2019impact,olson2013bilingual}, and still found differences between the two, suggesting that changes due to transfer are quickly `reset'-- in an order of seconds.
+However, note that these studies measure VOT, which is a temporal feature.
+We have no apriori reason to assume that these durations generalize to vowel quality.
+However, findings from sub-categorical phonetic shifts triggered by other factors (such as convergence towards an interlocutor) do evidence rapid shifts in vowel quality within similar time-frames \cite{pardo2010expressing,babel2010dialect,babel2012evidence}.
+Thus in the current study, we present switched and non-switched tokens in the same test block to induce a bilingual mode.
+We expect the intervening words between two subsequent targets to undo any residual effects of transfer.
 
-\paragraph{}Both CS and cued picture naming have been used in different studies to elicit mixed-language productions, and have shown a similar range of variation in results.
-Following from the ongoing debates about  switching paradigms, we collect production data from the same participant pool using both test paradigms, allowing for a direct comparison of responses.
+\subsection{Asymmetries between languages in extent and direction of transfer} \label{asymmetries}
+Previous research has largely reported asymmetrical shifts between languages. Two sources of such differences are discussed below:
 
-\paragraph{}Since previous work largely shows no inhibitory effect of a mixed language mode on temporary phonetic effects of language switching, switched and non-switched utterances were not presented in separate test blocks in this study (cf \citeNP{bullock2009trying,muldner2019phonetics,olson2016role}, etc), but rather alternated randomly within each block.
-Instructions were given in mixed language.
-Thus, maintained a bilingual language mode throughout the experiment.
-Also more ecologically valid, because daily language experience of the target population-- not likely to provide stable, monolingual ambient language conditions; constantly encounter mixed-language input (both auditory and visual).
+\subsubsection{L1 vs L2 status} 
+\paragraph{}Flege's Speech Learning Model (SLM) \citeyear{flege1995second,flege2007language} posits that the sound categories of a bilingual speaker exist in a common phonological space, and thus in principle, both L1 and L2 sounds can affect one another.
+Thus, the patterns of cross-language transfer depend on the mapping of L2 phonemes in relation to existing L1 categories.
+In non-switched production, this is evidenced through pervasive L1 influence on contrasts which are perceptually linked to an existing L1 contrast (equivalence classification-- \citeNP{flege1984limits,flege1987production}), and
+the observation that both L1 and L2 sound systems of bilinguals differ from corresponding monolingual speakers \cite{guion2003vowel}.
 
-\paragraph{}The research questions can thus be summarized as follows:
-\begin{enumerate}[label=(\roman*)]
-	\item Does mixed-language processing temporarily increase L1 influence on L2 vowel productions?
-	\item If so, does the language-specific distribution of vowels constrain the extent of shift?
-	\item Do two commonly used paradims-- code-switching and cued picture-naming, elicit different degrees of cross-langauge transfer?
-\end{enumerate}
+\paragraph{}In mixed-language production, the role of language status is less clear-- studies have reported unidirectional influence of L1 on L2 \cite{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \cite{tsui2019impact,elias2017effects, olson2013bilingual}, bidirectional convergence \cite{bullock2009trying, olson2016role}, divergence \cite{bullock2009trying,vsimavckova2018patterns}, and no influence \cite{muldner2019phonetics,schwartz2015language}.
+Since most existing studies measure shifts in VOT, the observed asymmetries between languages have been variously explained either in terms of their L1 vs L2 status, or language-specific differences between long- and short-lag VOT languages (cf sec.
+\ref{sound systems}).
+
+\citeA{olson2013bilingual} first found unidirectional VOT shift of L1 towards L2 in two different groups -- native speakers of Spanish (short-lag) and English (long-lag), matched for proficiency and age of L2 acquisition.
+This suggests that beyond language-specific differences, the L1 vs L2 status of the language does mediate transfer.
+They explain the asymmetry in terms of the Inhibitory Control Model (ICM) \cite{green1998mental} -- to select a phonetic realization from one language, the other must be inhibited.
+Being the `stronger' language, more inhibition is required on the L1 while speaking in an L2, than vice-versa.
+Thus, switching into L1 incurs a greater switch-cost (more cross-language influence) than switching into L2.
+The lesser switch cost results in a lack of visible cross-language transfer effects on L2.
+\citeA{tsui2019impact} found comparable results, but only in the groups which were non-balanced in language dominance.
+
+\paragraph{}Since VOT is independently affected by short- vs long-lag differences, studies of other sound contrasts/features can avoid this conflation.
+However, given the implications of the ICM, we would first have to establish that these L2 categories can indeed be affected by dynamic interference.
+The few existing studies which have measured vowel quality \cite{simonet2014phonetic,muldner2019phonetics,elias2017effects} and phonological processes \cite{simonet2020increased,schwartz2015language} have examined transfer effects on L1.
+Thus, this study builds on the existing work by checking if the L2 vowel quality of proficient bilinguals is affected by mixed-language processing.
+
+\subsubsection{Sound systems of the languages:} \label{sound systems} In contrast to the cognitive factors discussed above, many studies have attributed asymmetries to language-internal factors.
+Specifically, a pattern that emerges from the body of work on VOT is that shift in long-lag VOTs towards the short-lag norms of the other language is much more consistent and systematic than the reverse \cite{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid}.
+In general, short-lag VOTs seem to resist accommodative shift.
+\citeA{bullock2009trying} suggests that because long-lag languages offer a greater range of acceptable VOTs, there is more `room' for movement.
+In contrast, shift in short-lag languages would risk the loss of a phonological contrast (VOT is the primary cue for voicing contrast in most of these languages).
+This is a linguistic/phonological constraint of transfer.
+In a similar vein, \citeA{antoniou2011inter} found that while English stops shifted towards Greek (a short-lag language), the reverse was seen only in a subset of sounds-- word-medial non-nasalized stops.
+They attribute this to the low frequency of these sounds in Greek, which makes them less `stable', and thus more amenable to shift.
+This suggests that the distribution of sounds in a language can affect the kind of transfer patterns observed.
+
+\paragraph{}These results highlight that beyond interaction in cognitive processing, phonetic transfer is ultimately a linguistic phenomenon, and thus subject to language-specific constraints.
+Therefore, to make meaningful generalizations, it is imperative to consider data from a variety of language pairs.
+There are very few studies on phonologically(?) unrelated languages.
+This is the first transfer study on an Indo-Aryan language.
+
+Since Bengali has a 4-way laryngeal contrast, VOT does not just encode a voiced/voiceless distinction, and moreover is not the only cue for the corresponding phonological categories \cite{dmitrieva2020acoustic}.
+Thus, effects of Bengali on English VOT are likely to be affected by additional phonological factors.
+To avoid any potential confounds, we chose to look at vowels as a target for transfer.
+
+\subsection{Asymmetries between sounds:}\label{asymmetry between sounds}
+
+Existing research reports asymmetrical transfer effects not only between languages, but also among different sounds/features in a language, in both long-term and transient interactions.
+Studies which have examined multiple sound categories have found that interactions between individual sounds pairs do not necessarily reflect the overall pattern of global (system-wide) shift \cite{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot be seen as an atomic measure.
+In light of the discussion in section \ref{sound systems}, this is not surprising -- if a general linguistic principle of `room for movement' constrains transfer, then we would expect it to apply to individual sounds too.
+Once again, this emphasizes the need to study a wider range of sound contrasts.
+
+\paragraph{}In the present study, we examine two vowel contrasts in Indian English-- the mid-central vowel \nt{2} and the low front vowel \nt{\ae}.
+As table () shows, the former is in a less crowded part of the vowel space, and thus has more latitude for movement without risking contrast.
+Thus, considering phonological constraints, we expect a greater degree of shift in \nt{2} compared to \nt{\ae}.
+
+However, there is another potential source of asymmetry between the two sounds, which could lead to an alternative pattern:
+
+\nt{2} is not present as a phoneme in Bengali, whereas \nt{\ae} is a common category across both languages (cf sec.\ref{vowel systems}).
+Flege's Speech Learning Model (SLM) \cite{flege1995second,flege2007language} suggests that sound categories which are common across languages influence each other because they share a common acoustic-phonetic space.
+Other categories can in principle be produced with native-like accuracy, showing little effect of transfer.
+Given that our participants are highly proficient bilinguals, and assuming that cross-language correspondences remain the same in representation and processing, this would lead us to expect more transfer in the common category, \nt{\ae}.
+
+This also seems to follow from the fact that there is no obvious competing L1 category (no specific target for transfer) during the production of \nt{2}.
+However, it is unlikely that this should altogether preclude a shift in \nt{2}, since at least one existing study has reported transfer effects on a non-common vowel category in a comparable paradigm \cite{simonet2014phonetic}.
+Here, the Catalan \nt{O} moved towards an acoustically close category \nt{o}, which is common to Catalan and Spanish, suggesting that shift might be towards a set of phonetic norms rather than a particular target.
+
+Thus, we expect the English \nt{\ae} to lower towards the corresponding Bengali category(cf sec.\ref{vowel systems}).
+The vowel \nt{2} is expected to shift towards an acoustically close category in IE -- a probable candidate being the low vowel \nt{a:}, as anecdotal observations of heavily-accented English suggest that the sounds are perceived as close by Bengali speakers.
+The extent of transfer in each vowel could plausibly show either of the patterns outlined above.
+
+\subsection{A note on paradigms: connected speech or not?}
+
+Studies of phonetic transfer during mixed-language use have largely used spontaneous or scripted code-switching (CS, defined as the use of multiple languages within a single utterance, ie in connected speech; \citeNP{myers1993dueling}) to co-activate the two languages.
+Other paradigms include cued picture-naming, delayed repetition, interpreting across languages, reading word-lists etc.
+\citeA{olson2013bilingual} argues that code-switching introduces discursive factors such as discourse planning and pragmatic considerations, which ultimately show patterns of language practice, rather than cognitive behavior.
+This finds support from studies which report transfer effects before the switch point \cite{bullock2009trying}, or other evidence of planning such as hyperarticulation \cite{muldner2019phonetics} and interlocutor-awareness in spontaneous CS \cite{khattab2013phonetic}.
+To avoid such confounds, Olson suggests cued picture-naming (where pictures are named in rapid succession in either language, based on the given language cue) as an alternative paradigm to isolate purely phonetic effects in controlled setups.
+
+\paragraph{}Note that while there is evidence for planning in CS, exactly how this affects phonetic transfer cannot be predicted from the existing literature-- while \cite{bullock2009trying} found tokens immediately preceding the switch point to be phonetically different from the rest, one group showed convergence towards the switch language(explained in terms of proficiency) while the other showed divergence (explained in terms of extra-linguistic factors).
+The participants in Muldner et al.'s \citeyear{muldner2019phonetics} study hyperarticulated switched tokens (suggesting some degree of planning), and showed no significant shift.
+Moreover, recall that Olson's \citeyear{olson2016role} findings suggest that there is a limit on transfer-- multiple factors do not necessarily lead to additive effects.
+Thus, the existence of additional factors in CS does not automatically mean that this should affect the extent of transfer.
+
+\paragraph{}While much more work is necessary to tease out the various factors and their interactions, the focus of the present study is to check whether, in an experimental setting, the postulated discursive elements in a CS paradigm affect the extent of observed transfer, compared to a paradigm which arguably lacks them--cued picture-naming.
+We are less concerned with the exact nature of differences, than with their comparability.
+While studies using each of these paradigms to induce transfer exist, comparison across studies is difficult, since they also differ in other factors.
+This study provides a starting point for such comparison by eliciting productions of the same tokens from the same participants, in both paradigms.
+A difference in the extent of transfer will suggest that the planning effects in connected speech affect transfer independently of the phonetic results of switching, which has implications that future studies should look into.
+% in discussion--cognitive load, also--check- by how much the time reduced for each of the paradigms (1st vs last iteration).
 

### 2021-01-18 writeup_final/writeup_final.tex
new folder writeup_final
Added analysis, results, discussion

--- 2020-10-15 paper_draft/main.tex
@@
+% !TeX spellcheck = en_US
 \documentclass[11pt]{article}
-
+\usepackage{apacite}
 \usepackage{syntonly}
+\usepackage{booktabs}
+\usepackage{dcolumn}
 %\syntaxonly
@@
 \geometry{top=1.0in, bottom=1.0in, left=1.0in, right=1.0in}
-\usepackage{apacite}
+
 %
@@
 %\doublespacing
-
+\usepackage{comment}
 %
@@
 
-\author{Auromita Mitra, Indranil Dutta}
-
 \begin{document}
@@
 
-%\bibliographystyle{apacite}
+	\bibliographystyle{apacite}
+%\begin{markdown}
 
@@
  Research on the phonological systems of bi/multilingual speakers suggests that speakers maintain separate, but phonologically linked sound categories for both their languages.
-These show cross-language influence in both perception and production
+These show cross-language influence in both perception and production.
  %citation
-.
 In production, this kind of influence (variously termed transfer, drift, accommodation, interference-- used interchangeably here to indicate any interaction between two sets of phonetic norms) has been studied in two conditions:
@@
 The resulting transfer is variously thought to involve short-term memory, online processing costs, language mode, and context-awareness.
-This is the focus of the present study.
+The present study is concerned with the latter.
 
-\paragraph{}Studies of the phonetic effects of mixed-language production are relatively fewer, and much more recent.
-Existing work has largely focused on a few pairs of phonologically related languages, and used temporal properties of consonants (in particular, VOT) to measure transfer.
+\paragraph{ }Existing research on the phonetic effects of mixed-language production has largely focused on a few pairs of phonologically related languages, and used temporal properties of consonants (in particular, voice onset time or VOT) to measure transfer.
 However, great variation in results (cf sec.\ref{asymmetries} ) suggests that data from a wider variety of populations, language pairs,  and phonetic features is required in order to make meaningful generalizations.
-There is no work yet on phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language.
+There is no work yet on short-term phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language.
 Nevertheless, widespread multilingualism in this part of the world suggests that phonetic
 %line--rephrase
-behavior in these populations can improve our understanding of cross-language interactions at the phonetic level, as they are more likely to reflect real-world experience with mixed-language processing.
+ behavior in these populations can add to the understanding of such cross-language interactions, as they are more likely to reflect real-world experience with mixed-language processing.
 
-\paragraph{} The present study looks at phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India.
-We measure the spectral properties (F1 and F2) of two English vowels to check if L1 influence on L2 increases during mixed-language use relative to a participant's unilingual baseline production.
-Mixed-language data is elicited in two switching paradigms-- code-switching and cued picture-naming.
-The results show an effect of dynamic transfer on L2 vowels for the first time, and a difference in the extent of transfer between the two paradigms.
-The findings are discussed in light of recent proposals about asymmetries in transfer effects, and debates about the role of connected speech in enhancing cross-language interaction.
+\paragraph{} The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India.
+We measure the spectral properties (F1 and F2) of two English vowels to check if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2.
+Mixed-language data is elicited in two switching paradigms: code-switching and cued picture-naming.
+The results demonstrate a dynamic effect of transfer on L2 vowels, but no difference in the extent of transfer between the two paradigms.
+Findings are discussed in light of recent proposals about asymmetries in transfer effects and the role of connected speech in introducing discursive factors to transfer studies.
 
-\paragraph{}The next section gives background information on Bengali and English in India, focusing on vowel systems.
+\paragraph{}The next section provides background information on Bengali and English in India, focusing on vowel systems.
 The rest of the introduction  discusses causes, duration, and possible sources of asymmetries in transfer reported in previous studies, paradigms used, and the design of the present study in light of these.
@@
 	\subsubsection{Demography} 
-\paragraph{}Bengali (also, Bangla) is an Indo-Aryan language spoken in Indian and Bangladesh.
+	\paragraph{}Bengali (also, Bangla) is an Indo-Aryan language primarily spoken in India and Bangladesh.
 In India, more than 97 million people speak Bengali as a first language (Census of India, 2011), mostly in the state of West Bengal.
@@
 Thus, regional variations are seen as the result of L1-influence on a common underlying target.
-Early attempts to describe IE phonologically (General Indian English or GIE, \cite{masica1972sound}) were primarily attempts at standardization for pedagogical purposes.
-Excepting a small minority of specific communities \cite{pandey201517}, there are hardly any L1 speakers of IE.\\
+	%Early attempts to describe IE phonologically (General Indian English or GIE, \cite{masica1972sound}) were primarily attempts at standardization for pedagogical purposes. 
+	Excepting a small minority of specific communities \cite{pandey201517}, there are very few L1 speakers of IE.\\
 
@@
 	\subsubsection{Vowel systems} \label{vowel systems}
-The vowel inventory of Western Bangla (the variety spoken by the participants of the present study) consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \cite{garry2001facts}.
-Note that there is no mid-central vowel.
-Of relevance to this study, the category \nt{\ae} in Bengali is lower than the corresponding English category \cite{chatterji1921bengali}.\\
-% citation \cite{bhattacharyarelated}
+	The vowel inventory of Western Bangla (the variety spoken by the participants in this study) consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \cite{garry2001facts}.
+Note that there is no mid-central vowel category.
 
 	The vowel system of IE contains the monopthongs \nt{I, i, E, e, \ae, @/2, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \cite{wells1982accents, masica1972sound}.
-Their distribution in the F1XF2 space is shown in (insert table 1.7, Pandey).
-There is a single mid-central vowel corresponding to the categories \nt{2,@,3:}, which are treated as distinct in most native varieties of English.
+%Their distribution in the F1XF2 space is shown in (insert table 1.7, Pandey).
+	There is a single mid-central vowel corresponding to the categories \nt{2,@,3:}, which are treated as distinct in many native varieties of English.
 Since the English items used in this study are traditionally transcribed with \nt{2}, we use this symbol to indicate the mid-central vowel throughout.
@@
 
-Many studies have distinguished between changes in memory \textsc{representations} due to the acquisition of multiple sound systems (cf SLM \citeNP{flege1995second,flege2007language}; PAM-L2 \citeNP{best2007nonnative}), and interaction during \textsc{processing} (transfer vs interference; \citeNP{grosjean2012attempt}, competence vs performance interference; \citeNP{paradis1993linguistic}).
+Existing research has distinguished between changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf SLM \cite{flege1995second,flege2007language}; PAM-L2 \cite{best2007nonnative}), and interaction during \textsc{processing} (transfer vs interference \cite{grosjean2012attempt}; competence vs performance interference \cite{paradis1993linguistic}).
 Given their transient nature (cf sec.\ref{duration}), dynamic changes in production during mixed-language use are generally thought to involve the latter \cite{elias2017effects,simonet2014phonetic}.
 What triggers this interaction?
-\citeA{olson2016role} argues that while cross-language phonetic effects are largely measured at the point of the switch, it could have two potential sources:
+\citeA{olson2016role} argues that while cross-language phonetic effects are largely measured at the point of language switch, it could have two potential sources:
 \begin{enumerate}
@@
 Overall, results suggest that (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \cite{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use -- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \cite{olson2016role,tsui2019impact} or spontaneous conversation \cite{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
-(iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \cite{olson2016role} found no additive effects, \cite{olson2013bilingual} found the balanced language context to inhibit transfer compared to unbalanced contexts.
-Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in separate monolingual test block, separated by a few hours to days \cite{schwartz2015language, bullock2009trying,antoniou2011inter, elias2017effects,vsimavckova2015immediate,vsimavckova2018patterns}.
+(iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \citeA{olson2016role} found no additive effects, \citeA{olson2013bilingual} found the balanced language context to inhibit transfer compared to unbalanced contexts.
+Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days \cite{schwartz2015language, bullock2009trying,antoniou2011inter, elias2017effects,vsimavckova2015immediate,vsimavckova2018patterns}.
 
-\paragraph{}\citeA{grosjean1998studying} suggests that various aspects of the communicative setting, including exposure to (spoken or written) stimuli in multiple languages, and awareness of the interlocutor's being bilingual, could trigger a bilingual mode.
+\paragraph{}\citeA{grosjean1998studying} suggests that various aspects of the communicative setting, including exposure to (spoken or written) stimuli in multiple languages and awareness of the interlocutor's being bilingual, could trigger a bilingual mode.
 Since the participants in this study are in an environment which largely contains mixed-language input, multilingual interlocutors, and no stable ambient language, we expect that phonetic transfer during daily language use, if any, takes place in a bilingual mode.
@@
 
-Studies which compare bilinguals' phonological systems with monolinguals see them as relatively stable over time.
+%Studies which compare bilinguals speakers' phonological systems with monolinguals see them as relatively stable over time.
 The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods, \cite<e.g.>{guion2003vowel, caramazza1973acquisition}.
-However, longitudinal studies (citations) and between-subject comparisons of bilingual speakers with different durations of L2 exposure suggest that the interaction between the sound systems is dynamic.
+However, longitudinal studies %(citations)
+%and between-subject comparisons of bilingual speakers who differ in the duration of L2 exposure suggest that interaction between the sound systems is dynamic.
 Over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts-- e.g.
@@
 
-Moreover, changes due to transfer are not unidirectional or irreversible.
+%Moreover, changes due to transfer are not unidirectional or irreversible.
 \citeA{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language.
@@
 How long these effects last in the absence of regular L2 input was not tested.
-% if effects are cumulative, then the difference between switched and non-switched tokes should be lesser in later test blocks 
-(previous two paragraphs needed?)
+% if effects are cumulative, then the difference between switched and non-switched tokes should be lesser in later test blocks (previous two paragraphs needed?)
 
-\paragraph{}Studies looking at mixed language use within an utterance mostly analyze only the switched (target) token.
+\paragraph{}Studies examining mixed-language use within an utterance mostly analyze just the switched (target) token.
 Thus, very few direct measurements of the duration of transfer effects are available.
-One study which measured this \cite{bullock2009trying} did not find any residual effects on the matrix language following the switch, suggesting that transfer during code-switching is localized.
+One study which measured this \cite{bullock2009trying} did not find any residual effects on the matrix language following a switch, suggesting that transfer during code-switching is localized.
 
-More indirect evidence for this comes from studies which have elicited switched and non-switched tokens in the same test block \cite{tsui2019impact,olson2013bilingual}, and still found differences between the two, suggesting that changes due to transfer are quickly `reset'-- in an order of seconds.
-However, note that these studies measure VOT, which is a temporal feature.
+More indirect evidence for this comes from studies which have elicited switched and non-switched tokens in the same test block \cite{tsui2019impact,olson2013bilingual}, and still reported differences between the two, suggesting that changes due to transfer are quickly `reset'-- in an order of seconds.
+Note that all these studies measure VOT, which is a temporal feature.
 We have no apriori reason to assume that these durations generalize to vowel quality.
-However, findings from sub-categorical phonetic shifts triggered by other factors (such as convergence towards an interlocutor) do evidence rapid shifts in vowel quality within similar time-frames \cite{pardo2010expressing,babel2010dialect,babel2012evidence}.
-Thus in the current study, we present switched and non-switched tokens in the same test block to induce a bilingual mode.
+However, findings from sub-categorical phonetic shifts triggered by other factors (such as convergence towards an interlocutor) do evidence rapid shifts in vowel quality within comparable time-frames \cite{pardo2010expressing,babel2010dialect,babel2012evidence}.
+Thus in the current study, we present switched and non-switched tokens randomly within the same test block to induce a bilingual mode.
 We expect the intervening words between two subsequent targets to undo any residual effects of transfer.
@@
 \subsection{Asymmetries between languages in extent and direction of transfer} \label{asymmetries}
-Previous research has largely reported asymmetrical shifts between languages. Two sources of such differences are discussed below:
+Previous research has largely reported asymmetrical shifts between languages. Two proposed sources of such differences that are relevant to this study are discussed below:
 
@@
 Thus, the patterns of cross-language transfer depend on the mapping of L2 phonemes in relation to existing L1 categories.
-In non-switched production, this is evidenced through pervasive L1 influence on contrasts which are perceptually linked to an existing L1 contrast (equivalence classification-- \citeNP{flege1984limits,flege1987production}), and
-the observation that both L1 and L2 sound systems of bilinguals differ from corresponding monolingual speakers \cite{guion2003vowel}.
+In non-switched production, this is evidenced through a pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (equivalence classification-- \citeNP{flege1984limits,flege1987production}), and
+the observation that both the L1 and L2 sound systems of bilinguals differ from corresponding monolingual speakers \cite{guion2003vowel}.
 
@@
 
-\citeA{olson2013bilingual} first found unidirectional VOT shift of L1 towards L2 in two different groups -- native speakers of Spanish (short-lag) and English (long-lag), matched for proficiency and age of L2 acquisition.
-This suggests that beyond language-specific differences, the L1 vs L2 status of the language does mediate transfer.
+ \paragraph{}\citeA{olson2013bilingual} first reported a unidirectional VOT shift of L1 towards L2 in two different groups -- native speakers of Spanish (short-lag) and English (long-lag), matched for proficiency and age of L2 acquisition.
+This suggested that beyond language-specific differences, the L1 vs L2 status of the language does mediate transfer.
 They explain the asymmetry in terms of the Inhibitory Control Model (ICM) \cite{green1998mental} -- to select a phonetic realization from one language, the other must be inhibited.
-Being the `stronger' language, more inhibition is required on the L1 while speaking in an L2, than vice-versa.
+Being the `stronger' language, more inhibition is required on the L1.
 Thus, switching into L1 incurs a greater switch-cost (more cross-language influence) than switching into L2.
 The lesser switch cost results in a lack of visible cross-language transfer effects on L2.
-\citeA{tsui2019impact} found comparable results, but only in the groups which were non-balanced in language dominance.
+\citeA{tsui2019impact} found comparable results, but only in participants who were not equally dominant in both languages.
+Balanced bilinguals did not demonstrate any transfer effects in VOT.
+They propose that this is because balanced bilinguals have better inhibitory control (low switch costs in both languages) due to greater experience with language switching.
 
-\paragraph{}Since VOT is independently affected by short- vs long-lag differences, studies of other sound contrasts/features can avoid this conflation.
+ \paragraph{}Since VOT is independently affected by short- vs long-lag differences, studies of other sound contrasts/features can avoid this conflation, and give a clearer picture of the effect of language status.
 However, given the implications of the ICM, we would first have to establish that these L2 categories can indeed be affected by dynamic interference.
-The few existing studies which have measured vowel quality \cite{simonet2014phonetic,muldner2019phonetics,elias2017effects} and phonological processes \cite{simonet2020increased,schwartz2015language} have examined transfer effects on L1.
-Thus, this study builds on the existing work by checking if the L2 vowel quality of proficient bilinguals is affected by mixed-language processing.
+The few existing studies which focus on vowel quality \cite{simonet2014phonetic,muldner2019phonetics,elias2017effects} and phonological processes \cite{simonet2020increased,schwartz2015language} have all examined transfer effects on L1.
+As far as we are aware, there has been no work yet on vowels in L2.
+Thus, this study builds on the existing work by checking whether the L2 vowel quality of proficient bilinguals can be affected by mixed-language processing.
 
-\subsubsection{Sound systems of the languages:} \label{sound systems} In contrast to the cognitive factors discussed above, many studies have attributed asymmetries to language-internal factors.
-Specifically, a pattern that emerges from the body of work on VOT is that shift in long-lag VOTs towards the short-lag norms of the other language is much more consistent and systematic than the reverse \cite{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid}.
+ \subsubsection{Differences between sound systems of the languages} \label{sound systems} In addition to the cognitive factors discussed above, many studies have attributed the asymmetries to language-internal factors.
+Specifically, a pattern that emerges from the body of work on VOT is that a shift in long-lag VOT towards the short-lag norms of the other language is much more consistent and systematic than the converse \cite{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid}.
 In general, short-lag VOTs seem to resist accommodative shift.
-\citeA{bullock2009trying} suggests that because long-lag languages offer a greater range of acceptable VOTs, there is more `room' for movement.
-In contrast, shift in short-lag languages would risk the loss of a phonological contrast (VOT is the primary cue for voicing contrast in most of these languages).
-This is a linguistic/phonological constraint of transfer.
+\citeA{bullock2009trying} suggest that because long-lag languages offer a greater range of acceptable VOTs, there is more `room' for movement.
+In contrast, shift in short-lag languages could risk the loss of a phonological contrast (VOT is the primary cue for voicing contrast in most of these languages).
+This is a language-specific phonological constraint on transfer.
   In a similar vein, \citeA{antoniou2011inter} found that while English stops shifted towards Greek (a short-lag language), the reverse was seen only in a subset of sounds-- word-medial non-nasalized stops.
 They attribute this to the low frequency of these sounds in Greek, which makes them less `stable', and thus more amenable to shift.
-This suggests that the distribution of sounds in a language can affect the kind of transfer patterns observed.
+This suggests that the distribution of sounds in a language could affect the kind of transfer patterns observed.
 
-\paragraph{}These results highlight that beyond interaction in cognitive processing, phonetic transfer is ultimately a linguistic phenomenon, and thus subject to language-specific constraints.
+  \paragraph{}These results highlight that beyond interaction during the cognitive processing of sounds, phonetic transfer is ultimately a linguistic phenomenon, and thus subject to language-specific constraints.
 Therefore, to make meaningful generalizations, it is imperative to consider data from a variety of language pairs.
-There are very few studies on phonologically(?) unrelated languages.
-This is the first transfer study on an Indo-Aryan language.
+Existing research on phonetic transfer largely revolves around a small set of phonologically related languages.
+Thus, the present study extends the scope of this research to a new set of languages-- Indian English and Bengali (an Indo-Aryan language).
 
-Since Bengali has a 4-way laryngeal contrast, VOT does not just encode a voiced/voiceless distinction, and moreover is not the only cue for the corresponding phonological categories \cite{dmitrieva2020acoustic}.
-Thus, effects of Bengali on English VOT are likely to be affected by additional phonological factors.
-To avoid any potential confounds, we chose to look at vowels as a target for transfer.
+ \paragraph{}Since Bengali has a 4-way laryngeal contrast, VOT does not encode a binary voiced/voiceless distinction, and moreover is not the only cue for the corresponding phonological categories \cite{dmitrieva2020acoustic}.
+Thus, the effects of Bengali on English VOT are likely to be affected by additional phonological factors.
+Thus, avoiding any potential confounds was another reason we chose to focus on vowels as a target for transfer.
 
@@
 
-Existing research reports asymmetrical transfer effects not only between languages, but also among different sounds/features in a language, in both long-term and transient interactions.
+ Existing research reports asymmetrical transfer effects not only between languages, but also between different sounds/features in a language, in both long-term and transient interactions.
 Studies which have examined multiple sound categories have found that interactions between individual sounds pairs do not necessarily reflect the overall pattern of global (system-wide) shift \cite{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot be seen as an atomic measure.
-In light of the discussion in section \ref{sound systems}, this is not surprising -- if a general linguistic principle of `room for movement' constrains transfer, then we would expect it to apply to individual sounds too.
+In light of the discussion in section \ref{sound systems}, this is not surprising -- if a general linguistic principle of `room for movement' constrains transfer, then we should expect it to apply to individual sounds too.
 Once again, this emphasizes the need to study a wider range of sound contrasts.
 
-\paragraph{}In the present study, we examine two vowel contrasts in Indian English-- the mid-central vowel \nt{2} and the low front vowel \nt{\ae}.
-As table () shows, the former is in a less crowded part of the vowel space, and thus has more latitude for movement without risking contrast.
-Thus, considering phonological constraints, we expect a greater degree of shift in \nt{2} compared to \nt{\ae}.
+ \paragraph{}In the present study, we examine two vowel categories in Indian English-- the mid-central vowel \nt{2} and the low front vowel \nt{\ae}.
+There are two plausible sources of asymmetry between these:
+ \begin{enumerate}
+ 	\item Position in the vowel space of IE: \nt{2} is in a less crowded part of the vowel space, and thus has more latitude for movement without risking contrast, particularly in the height (F1) dimension.
+Thus, considering purely phonological constraints on IE, we would expect that a greater degree of shift is possible in \nt{2} compared to \nt{\ae}.
+Any movement in \nt{\ae} is expected to be limited to front-back (F2) dimension.
+ 	\item Target of transfer: Since we do not elicit unilingual L1 utterances in this study, we cannot directly compare the target vowels to baseline L1 vowels.
+However, the body of existing research on phonetic transfer suggests that shifts are targeted with respect to categories in the other language, rather than random.
+Thus, another source of asymmetry is the fact that \nt{2} is not present as a phoneme in Bengali, whereas \nt{\ae} is a common category across both languages (cf sec.\ref{vowel systems}).
+ \end{enumerate}
 
-However, there is another potential source of asymmetry between the two sounds, which could lead to an alternative pattern:
+%table
 
-\nt{2} is not present as a phoneme in Bengali, whereas \nt{\ae} is a common category across both languages (cf sec.\ref{vowel systems}).
   Flege's Speech Learning Model (SLM) \cite{flege1995second,flege2007language} suggests that sound categories which are common across languages influence each other because they share a common acoustic-phonetic space.
-Other categories can in principle be produced with native-like accuracy, showing little effect of transfer.
+(So if Bengali and English differ in their canonical realizations of \nt{\ae}, then we should expect the English \nt{\ae} to shift towards the corresponding Bengali category in the mixed condition.--- is there any literature on this?)
+
+  %Other categories can in principle be produced with native-like accuracy, showing little effect of transfer.
 Given that our participants are highly proficient bilinguals, and assuming that cross-language correspondences remain the same in representation and processing, this would lead us to expect more transfer in the common category, \nt{\ae}.
 
-This also seems to follow from the fact that there is no obvious competing L1 category (no specific target for transfer) during the production of \nt{2}.
+ There is no obvious competing L1 category during the production of \nt{2}.
 However, it is unlikely that this should altogether preclude a shift in \nt{2}, since at least one existing study has reported transfer effects on a non-common vowel category in a comparable paradigm \cite{simonet2014phonetic}.
@@
 
-Thus, we expect the English \nt{\ae} to lower towards the corresponding Bengali category(cf sec.\ref{vowel systems}).
-The vowel \nt{2} is expected to shift towards an acoustically close category in IE -- a probable candidate being the low vowel \nt{a:}, as anecdotal observations of heavily-accented English suggest that the sounds are perceived as close by Bengali speakers.
-The extent of transfer in each vowel could plausibly show either of the patterns outlined above.
+  The vowel \nt{2} is thus expected to shift towards an acoustically close category in IE -- a probable candidate being the low vowel \nt{a:}, as anecdotal observations of heavily-accented English suggest that the sounds are perceived as close by Bengali speakers.
 
-\subsection{A note on paradigms: connected speech or not?}
+\subsection{A note on paradigms: connected speech or not?} \label{paradigms}
 
@@
 Moreover, recall that Olson's \citeyear{olson2016role} findings suggest that there is a limit on transfer-- multiple factors do not necessarily lead to additive effects.
-Thus, the existence of additional factors in CS does not automatically mean that this should affect the extent of transfer.
+Thus, the existence of additional factors in CS does not automatically mean that this should affect the extent of transfer compared to any other paradigm.
+It is possible that the effects in the studies mentioned above could be attributed to any other aspect of the experimental setup, rather than a result of code-switching in particular.
 
@@
 While studies using each of these paradigms to induce transfer exist, comparison across studies is difficult, since they also differ in other factors.
-This study provides a starting point for such comparison by eliciting productions of the same tokens from the same participants, in both paradigms.
+This study thus provides a starting point for such comparison by eliciting productions of the same tokens from the same participants, in both paradigms.
 A difference in the extent of transfer will suggest that the planning effects in connected speech affect transfer independently of the phonetic results of switching, which has implications that future studies should look into.
-% in discussion--cognitive load, also--check- by how much the time reduced for each of the paradigms (1st vs last iteration).
 
-\section{Methodology} %ID will write this
-Design: 10 participants*20 target words*2 utterance types *2 tasks *4 iterations = 3200 items
+\subsection{Research questions and hypotheses}
+\begin{itemize}
+	\item Are L2 vowel categories of proficient bilingual speakers affected by dynamic phonetic transfer during mixed-language production?
+	\item (Asymmetries-- direction of shift in each vowel) 
+	\item Is there an observable difference in the extent of transfer between the cued picture-naming and the code-switching paradigms? 
+\end{itemize}
+
+\section{Methodology} 
+The experiment had the following factorial design: 10 participants*20 target words*2 conditions *2 tasks *4 iterations = 3200 target items
 
@@
 
-\begin{itemize}
-	\item 10 Bengali-English bilingual speakers (5 female)
-	\item age range-- 19 to 28
-	\item selected on the basis of response to a Language Background Questionnaire, to match for language experience and LSRW skills in L1 and L2 
-\end{itemize}
+10 Bengali-English bilingual speakers (male=5, age range 19 to 28), completed the tasks.
+All the participants speak other languages in addition to Bengali and English.
+At the time of recording, they were living in or around the EFL University campus in Hyderabad, which has a linguistically diverse population.
+Speakers were selected on the basis of responses to a Language Background Questionnaire (sec.\ref{appendix}) to ensure comparable LSRW skills in L1 and L2 across participants, and were compensated for their time.
 
 \subsection{Stimuli}
-\begin{itemize}
-	\item Target words: 20 monosyllabic English words containing \nt{2} or \nt{ae} (not lexicalized loanwords)
-	\item Words with pre-vocalic voiced plosives (/b/ and /d/) were selected, to minimize coarticulatory effects on the target vowel
-	\item Filler words: 10 monosyllabic English words not containing the target vowels
-\end{itemize}
+Twenty monosyllabic English words containing the vowels \nt{2} or \nt{ae} were selected as target words.
+To minimize coarticulatory effects on the vowel, the pre-vocalic consonants were all voiced plosives (\nt{b} or \nt{d}).
+10 monosyllabic English words that do not contain the target vowel served as filler items.
+English words which are lexicalized loanwords in Bengali were avoided.
+The list of words can be found in sec.\ref{appendix}.
 
 \subsection{Procedure}
+Participants were recorded individually in a sound attenuated room. %, using a ... microphone type/description?
+Productions were recorded in Praat \cite{boersma2016praat} at a 44.1K sampling rate.
+Every participant performed four iterations each of two tasks-- cued picture-naming and code-switching.
+Utterances in the two conditions (unilingual and mixed) were randomly interspersed within each iteration.
+In the unilingual condition, the target word was preceded by another English word in the utterance, whereas in the mixed condition, the target was produced as a switch from Bengali.
+Every target word appeared twice during a test block; once in a unilingual and once in a mixed condition.
+Participants alternated between the two tasks, and were allowed to take a short break after each test block.
+
+The tasks are described below:
+
 \begin{enumerate}[]
@@
 		\item Language cue- a word in either English or Bengali orthography
-		\item a picture displayed for 50ms; named by the participant in either English or Bengali, depending on the language cue 
-		\item Target word- English word printed in English orthography; read out by the participant
+		\item A picture displayed for 50ms; named by the participant in either English or Bengali, depending on the language cue 
+		\item Target -- an English word; read out by the participant
 		\item Distracter math problem
 	\end{itemize}
-Since all target words were in English, language cue was used to manipulate the preceding word, giving target words in non-switched and switched utterances.
-Engish and Bengali orthographies are visually distinct, so it was used to cue language.
-  \item Code-switching: Each slide contained one target word embedded in either an English (unilingual) or a Bengali (bilingual; target produced as a code switch from Bengali) carrier sentence; read out by the participant.
-Carrier sentences:
+	Since all the target words were in English, the language cue was used to manipulate the preceding word, giving target words in non-switched (unilingual) and switched (mixed) utterances.
+The orthographies of English and Bengali are visually distinct, and were thus used to cue language.
+
+	\item Code-switching: Each slide contained one target word embedded in either an English (unilingual) or a Bengali (mixed; target produced as a switch from Bengali) carrier sentence which was read out by the participant.
+The carrier sentences are given below:
 	\begin{itemize}
@@
 
-  The sound preceding the target word was uniform across al sentences (the final sound in ‘yellow’ is assimilated to [o] in Indian English), to minimize the possible effects of coarticulation on the target word.
+	The sound preceding the target word was uniform across all sentences (the final sound in `yellow' is produced as [o] in Indian English), to minimize any possible differences due to coarticulatory effects on the target word.
 A specifier-noun construction was used to ensure identical word order in both languages, giving uniform prosodic context for the target word.
-For bilingual utterances, mixed orthography was used.
+Mixed orthography was used for the mixed-language utterances.
 \\
-  Four iterations of each task were recorded.
-Every target word appeared twice during a test block; once in a switched and once in a non-switched context.
-Order of items and conditions (switched vs non-switched) was randomized within each trial.
-Participants alternated between the two tasks.
 
 \end{enumerate}
-\subsection{Acoustic analysis}
- The first iteration of each task was treated as a trial.
+\subsection{Acoustic analysis}\label{analysis}
+The first iteration of each task was treated as a trial, and excluded from analysis.
 An additional 67 tokens (2.79\%) were excluded due to errors in presentation/production, giving 2333 target word items for analysis.
-Recodings were manually segmented and annotated in PRAAT \cite{boersma2016praat}, F1 and F2 of the target vowels \nt{2, ae} were measured at 5\% and 50\% into the vowels.
+Recordings were manually segmented and annotated in Praat \cite{boersma2016praat}, and the first and second formant frequencies (F1 and F2) of the target vowels \nt{2, ae} were measured at five points-- in 10\% increments starting at 5\% into the vowel.
+This gave measures of F1 and F2 at 5\%, 15\%, 25\%, 35\% and 45\% into the vowel, allowing us to observe any dynamic changes in the extent of transfer over the course of producing the vowel.
 
-\section{Results} %ID will write this
-\begin{itemize}
-	\item A linear mixed effects model with utterance type (unilingual vs bilingual), vowel, and gender as fixed effects shows a significant effect of utterance type on lobanov normalized F1 at 5\% (p=0.0096).
-Normalized F1 is higher in the bilingual utterances.
+\section{Results}
 
-	\item No effect of utterance type on F1 at 50\% of the vowel, or on normalized F2 values.
+%Figures ... show the findings of the experiment. 
+All statistical analyses were performed in R \cite{r}.
+The dependent variables were vowel height (F1) and backness (F2).
+We fitted a series of linear mixed effects models using the \emph{lme4} package \cite{lme4}, to check the effect of experimental conditions on lobanov normalized F1 and F2 at each of the five points in the vowel (sec.\ref{analysis}).
+For all mixed models, the alpha criterion was set at $|t| > 2 $.
+The independent variables or fixed factors, along with their levels, are given in table \ref{table variables}.
+Subject and Item were treated as random factors.
 
-	\item \nt{2} is lowered to a greater extent than \nt{\ae}.
+\begin{table}[h]
+	\caption{Fixed factors}
+	\centering
+	\begin{tabular}{l l}
+		\hline 
+		Factor & Levels\\ 
+		\hline
+		Gender & male, female\\
+		Vowel & \textbackslash vt (\nt{2}), \textbackslash ae (\nt{\ae}) \\
+		Task (corresponding to the two paradigms) & p (Picture: cued picture-naming, s (Sentence: code-switching)\\
+		Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)\\ [1ex]
+		\hline	
+	\end{tabular}
+\label{table variables}
+\end{table}
 
-	\item Significant effect of task
+Since we wanted to model height and backness at different points in the vowel using a single model, allowing us to track if/how transfer changes over time, we arbitrarily chose a point in the vowel (F1 at 25\%; henceforth $F1\_25$) to identify an optimal model, and then fitted this to all other points.
 
-	\item \wt{d}-initial target words show more lowering than \wt{b}-initial words %I had subsetted the data by onset consonant, and saw this.
-Not sure what to make of it.
-The b-initial set by itself doesn't show significant lowering.
+Vowel quality and gender are expected to systematically affect F1.
+Thus, we began with a model that had Vowel and Gender as fixed effects, random intercepts for Subject and Item, and random slopes for each within-unit variable that was relevant to our hypotheses-- by-subject slopes for Task, Context, and Vowel; by-item slopes for Task and Context.
+Next, adding Context as a fixed effect significantly improved the model fit ($\chi^2(1) = 5.54, p= 0.01854$), suggesting that the experimental condition affects vowel height.
+However, since we expected the extent of change in vowel height between the conditions to depend on the identity of the vowel (cf sec.\ref{asymmetries}), we added a two-way interaction factor between Vowel and Context.
+As expected, this was found to significantly improve model fit ($\chi^2(1) = 12.61, p= 0.0003837$).
+To check if the experimental paradigm affected F1, we added Task as a fixed effect, and this marginally improved the model ($\chi^2(1) = 3.74, p= 0.05302$).
+Following from the hypothesis in sec.\ref{paradigms}, we expected the extent of shift to differ across the two paradigms.
+However, no significant improvement was achieved by adding an interaction effect between Task and Context.
+Subsequent additions of other interactions, and removing each of the fixed effects in turn, did not lead to any further improvement.
 
-\end{itemize}
+Context and Vowel are both between-subject factors.
+Since our hypothesis concerns their interaction, we modified the random effects structure by adding a by-subject random slope for the interaction term \cite{brauer2018linear}.
+This improved the model fit ($\chi^2(5) = 11.43, p = 0.04342$).
+Thus, we report on the model including Gender, Task, Context, Vowel, and a Context-Vowel interaction as fixed effects, random intercepts for Subject and Item, by-subject random slopes for Task, Context, Vowel and the Context-Vowel interaction, and by-item random slopes for Task and Context.
+
+The following sections discuss the results for vowel height and backness respectively:
+
+\subsection{Vowel Height (F1)}
+
+Table \ref{table_f1_fixed_effects} summarizes the results of the fixed effects from the model for F1 at each point.
+As expected, we observe a significant effect of gender, such that male speakers have a lower mean F1 than female speakers.
+Vowel is not a significant predictor of F1.
+This seems surprising, given that the phonemic category is expected to systematically affect vowel quality.
+However, a boxplot visualization (Figure\ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two categories and the extent of height difference between them.
+Note that these measures are only from the English (unilingual) condition, suggesting that the variability is a feature of the participants' IE vowel categories, independent of the effects of switching.
+The lack of any independent effect of Context on F1 suggests that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels.
+Instead, the significant interaction between Context and Vowel indicates a lowering in \nt{2}, but not in \nt{\ae}, in the mixed-language condition.
+The results also reveal a significant effect of Task, such that the sentence task (code-switching) lowers both vowels.
+This effect is strongest at the beginning of the vowel, and disappears by 45\%.
+This is an unexpected result, since we have no principled reason to expect code-switching to induce an across-board lowering of vowels.
+As noted earlier, there is no significant interaction of Task with any of the other fixed factors.
+We used a series of ANOVAs to compare the above model with a null model that differs in the absence of the main factor which concerns our hypothesis-- Vowel*Context-- at each point in the vowel.
+We find that the effect is not significant at 5\%, and increases over the course of articulating the vowel.
+%ANOVA table add
+
+\begin{figure}
+
+	\includegraphics{vowels_by_subject_english}
+	\caption{Subject-wise mean F1 values for \nt{2} and \nt{\ae} uttered in the unilingual English condition}
+	\label{boxplot_F1}
+\end{figure}
+
+\begin{table}[htbp] \centering 
+	\caption{Model outputs for F1 at each of the five points in the vowel} 
+	\label{table_f1_fixed_effects} 
+	\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
+		\\[-1.8ex]\hline 
+		\hline \\[-1.8ex] 
+		& \multicolumn{5}{c}{\textit{Dependent variable:}} \\ 
+		\cline{2-6} 
+		\\[-1.8ex] & F1\_5 & F1\_15 & F1\_25 & F1\_35 & F1\_45 \\ 
+		\\[-1.8ex] & (Model 1) & (Model 2) & (Model 3) & (Model 4) & (Model 5)\\ 
+		\hline \\[-1.8ex] 
+		(Intercept) & 0.447$^{**}$ & 0.718$^{***}$ & 0.750$^{***}$ & 0.863$^{***}$ & 0.921$^{***}$ \\ 
+		& (0.211) & (0.176) & (0.153) & (0.152) & (0.150) \\ 
+		& & & & & \\ 
+		Gendermale & $-$1.171$^{***}$ & $-$1.526$^{***}$ & $-$1.586$^{***}$ & $-$1.668$^{***}$ & $-$1.681$^{***}$ \\ 
+		& (0.237) & (0.184) & (0.168) & (0.177) & (0.175) \\ 
+		& & & & & \\ 
+		Contexte & $-$0.010 & $-$0.015 & 0.033 & 0.025 & 0.037 \\ 
+		& (0.058) & (0.057) & (0.042) & (0.040) & (0.035) \\ 
+		& & & & & \\ 
+		Vowel\textbackslash vt & 0.022 & $-$0.070 & $-$0.034 & $-$0.131 & $-$0.216 \\ 
+		& (0.170) & (0.175) & (0.170) & (0.174) & (0.168) \\ 
+		& & & & & \\ 
+		Tasks & 0.319$^{***}$ & 0.235$^{***}$ & 0.172$^{**}$ & 0.133$^{*}$ & 0.102 \\ 
+		& (0.091) & (0.083) & (0.081) & (0.077) & (0.073) \\ 
+		& & & & & \\ 
+		Contexte:Vowel\textbackslash vt & $-$0.126$^{**}$ & $-$0.138$^{**}$ & $-$0.185$^{***}$ & $-$0.191$^{***}$ & $-$0.192$^{***}$ \\ 
+		& (0.056) & (0.060) & (0.055) & (0.064) & (0.066) \\ 
+		& & & & & \\ 
+		\hline \\[-1.8ex] 
+		Observations & 2,333 & 2,333 & 2,333 & 2,333 & 2,333 \\ 
+		Log Likelihood & $-$1,771.558 & $-$1,589.316 & $-$1,278.310 & $-$1,313.795 & $-$1,380.083 \\ 
+		Akaike Inf. Crit. & 3,599.116 & 3,234.631 & 2,612.621 & 2,683.589 & 2,816.165 \\ 
+		Bayesian Inf. Crit. & 3,760.253 & 3,395.769 & 2,773.758 & 2,844.727 & 2,977.303 \\ 
+		\hline 
+		\hline \\[-1.8ex] 
+		\textit{Note:}  & \multicolumn{5}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
+	\end{tabular} 
+\end{table}
+
+\subsection{Vowel backness (F2)}
+
+Table \ref{table_F2_fixed_effects} summarizes the results of fixed effects from the model for F2 at each point.
+Once again, we see that male speakers have a lower mean F2 than female speakers, as expected.
+The estimates for Vowel show that excepting the beginning of the vowel, the category \nt{2} is consistently produced as more backed than \nt{\ae}.
+Unlike vowel height, Context has an independent effect on vowel backness, such that both the vowels in this study have lower F2 values in the English (unilingual) condition, i.e.
+they are fronted in the mixed condition.
+A significant interaction between Vowel and Context shows that there is a difference in the extent of fronting between the two vowels -- the shift is greater in the case of \nt{\ae}.
+We see a significant effect of Task on F2, such that vowels in the sentence task (code-switching) were more backed than those in the picture-naming task.
+As with F1, this is an unexpected result.
+A lack of significant interaction with Context suggests that there is no difference in the extent of shift between the two paradigms.
+The results of ANOVAs comparing the full model to a null model which lacks the Vowel*Context interaction factor show that as with vowel height, the change in vowel backness between the two conditions is not significant at 5\%, and increases over the course of the vowel.
+
+\begin{table}[htbp] \centering 
+	\caption{Model outputs for F2 at each of the five points in the vowel} 
+	\label{table_F2_fixed_effects} 
+	\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
+		\\[-1.8ex]\hline 
+		\hline \\[-1.8ex] 
+		& \multicolumn{5}{c}{\textit{Dependent variable:}} \\ 
+		\cline{2-6} 
+		\\[-1.8ex] & F2\_5 & F2\_15 & F2\_25 & F2\_35 & F2\_45 \\ 
+		\\[-1.8ex] & (Model 1) & (Model 2) & (Model 3) & (Model 4) & (Model 5)\\ 
+		\hline \\[-1.8ex] 
+		(Intercept) & 0.607$^{***}$ & 0.838$^{***}$ & 0.778$^{***}$ & 0.827$^{***}$ & 0.826$^{***}$ \\ 
+		& (0.222) & (0.213) & (0.211) & (0.203) & (0.204) \\ 
+		& & & & & \\ 
+		Gendermale & $-$0.839$^{***}$ & $-$0.891$^{***}$ & $-$0.807$^{***}$ & $-$0.750$^{***}$ & $-$0.748$^{***}$ \\ 
+		& (0.124) & (0.106) & (0.107) & (0.116) & (0.109) \\ 
+		& & & & & \\ 
+		Contexte & 0.003 & $-$0.088$^{**}$ & $-$0.065$^{**}$ & $-$0.063 & $-$0.103$^{***}$ \\ 
+		& (0.036) & (0.036) & (0.032) & (0.044) & (0.033) \\ 
+		& & & & & \\ 
+		Vowel\textbackslash vt & $-$0.217 & $-$0.615$^{***}$ & $-$0.651$^{***}$ & $-$0.801$^{***}$ & $-$0.836$^{***}$ \\ 
+		& (0.175) & (0.183) & (0.175) & (0.168) & (0.166) \\ 
+		& & & & & \\
+		Tasks & $-$0.234$^{***}$ & $-$0.194$^{***}$ & $-$0.152$^{***}$ & $-$0.141$^{***}$ & $-$0.092$^{**}$ \\ 
+		& (0.029) & (0.035) & (0.050) & (0.047) & (0.045) \\ 
+		& & & & & \\
+		Contexte:Vowel\textbackslash vt & 0.031 & 0.110$^{***}$ & 0.128$^{***}$ & 0.091$^{**}$ & 0.147$^{***}$ \\ 
+		& (0.050) & (0.041) & (0.040) & (0.046) & (0.043) \\ 
+		& & & & & \\   
+		\hline \\[-1.8ex] 
+		Observations & 2,333 & 2,333 & 2,333 & 2,333 & 2,333 \\ 
+		Log Likelihood & $-$1,283.531 & $-$1,478.494 & $-$1,174.087 & $-$1,222.682 & $-$1,019.033 \\ 
+		Akaike Inf. Crit. & 2,623.062 & 3,012.988 & 2,404.175 & 2,501.363 & 2,094.066 \\ 
+		Bayesian Inf. Crit. & 2,784.199 & 3,174.125 & 2,565.312 & 2,662.501 & 2,255.204 \\ 
+		\hline 
+		\hline \\[-1.8ex] 
+		\textit{Note:}  & \multicolumn{5}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
+	\end{tabular} 
+\end{table}
 
 \section{Discussion}
-Results are in line with previous studies which find that CS between languages does not lead to a parallel switch in phonology. Instead, gradient, momentary phonetic shifts are observed.
-\begin{itemize}
-	\item Both the vowels are lowered (shift towards related L1 category) in the bilingual utterances compared to speakers’ own unilingual productions, showing increased cross-language influence during mixed-language processing.
-	\item Lowering at 5\% of the vowel, but not at 50\%, suggests that only the initial part of the vowel is receptive to dynamic shift, not the steady state
-	\item No change in F2
-	\item More shift in \nt{2} compared to \nt{\ae}-- suggests that the extent of shift is constrained by the distribution of vowels and room for variability.
-Since we don't have monolingual Bengali data, can't say anything about the extent of shift relative to the initial distance between the categories.
-	\item CS in connected speech leads to greater cross-language phonetic interaction than unplanned switches through a cued picture naming task, suggesting that at least part of the observed effects in CS paradigms can be attributed to discourse awareness and planning.
-\end{itemize}
+
+(Extent and direction of shift)
+
+Our third research question concerned the experimental paradigms.
+\citeA{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, "discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
+This calls to question the comparability of results between studies using these different paradigms.
+We tested this by comparing productions of the same items from the same group of participants in a scripted CS paradigm and a cued picture-naming paradigm.
+We found no significant interaction of Task either with Context, or with the Vowel*Context interaction, suggesting that the extent of phonetic transfer was not affected by the paradigm.
+There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in transfer at all.
+This would mean that the hyperarticulation seen, for example, in \citeA{muldner2019phonetics} is better interpreted as resulting from some other aspect of the experimental setup, rather than the act of code-switching itself.
+While planning could additionally affect the words preceding the switch, the effects on the switched token itself do not change.
+(ii) Additionally, our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
+Thus, another possible explanation is that the discursive effects discussed by \citeA{olson2013bilingual} are triggered by more naturalistic paradigms where attention to meaning and interlocutor are important.
+More within-study comparisons with less structured paradigms, and spontaneous code-switching data, can shed light on this.
+The results of the present study suggest that for the purposes of observing dynamic phonetic transfer, elicitation through cued-picture-naming vs (controlled) code-switching does not seem to significantly alter the behavior under study, and thus the choice of paradigm can be based on other considerations.
+Moreover, this suggests that the results from such studies can be compared directly.
+
+Paradigm did affect vowel quality, but this applied uniformly across experimental conditions-- all vowels were more backed and lowered in the code-switching task.
+Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this is an artifact of the particular stimuli used here-- resulting from some corarticulatory effect that was absent in the picture-naming task.
+While this is intriguing, it is beyond the scope of the present study.
 
 \section{Conclusion}
-Mixed language processing temporarily increases cross-language phonetic transfer, causing L2 vowels to converge towards L1.
-The extent of shift is mediated by language-specific distribution of sound categories, and is greater in connected speech.
+Mixed language processing temporarily increases cross-language phonetic transfer, causing L2 vowels to (converge towards L1 norms ?).
+The extent and direction of such shift is mediated by language-specific distribution of sound categories.
+In a controlled experimental setup to study transfer effects, the postulated discursive factors in a code-switching paradigm does not change the patterns of observed transfer.
 
-\bibliographystyle{unsrt}
+\section{Appendix} \label{appendix}
 
+%\end{markdown}
+%\bibliography{F:/DISHA_DATA/Disha_study/LINGUISTICS/Shastri/references_srsf.bib}
+%\bibliography{C:/Users/91767/Desktop/Project/Writeup/references.bib}
 \bibliography{references}

### 2021-01-18 RCode/bengali_phon_acco_new.R
R code for interaction model

@@ -1,0 +1,328 @@
+formant_data_5 <- read.csv("~/Desktop/Disha_Project/formant_data_5.csv")
+attach(formant_data_5)
+# detach(formant_data_5)
+
+library(lattice)
+library(plyr)
+library(phonR)
+library(lme4)
+library(ggplot2)
+library(dplyr)
+library(stargazer)
+
+
+normF1 <- with(formant_data_5,normLobanov(cbind(F1_5,F1_15,F1_25,F1_35,F1_45)),group=Subject)
+normF1 <- as.data.frame(normF1)
+
+normF2 <- with(formant_data_5,normLobanov(cbind(F2_5,F2_15,F2_25,F2_35,F2_45)),group=Subject)
+normF2 <- as.data.frame(normF2)
+
+
+with(formant_data_5, plotVowels(F1_15, F2_15, Vowel, group = Context, plot.tokens = FALSE, 
+              plot.means = TRUE , var.sty.by = Vowel,var.col.by = Context, pch.tokens = Vowel, cex.tokens = 0.8,
+              cex.means = 0.8, alpha.means = 0.3,alpha.tokens = 0.3,ellipse.line = TRUE,  ellipse.fill = TRUE, pretty = TRUE))
+
+# par(mfrow = c(1, 1))
+# with(bengali_phon_acco_raw, plotVowels(normF1, normF2, plot.tokens = TRUE, pch.tokens = Vowel, 
+#                     plot.means = TRUE, pch.means = Context, 
+#                       ellipse.line = TRUE, pretty = TRUE))
+# boxplot(normF1~context*vowel, data=bengali_phon_acco_raw)
+
+# vowel_vt <- bengali_phon_acco_raw %>%
+#   filter(Vowel == "ʌ")
+# task_p <- subset(bengali_phon_acco_raw, (vowel=="ʌ"))
+# p <- ggplot(bengali_phon_acco_raw, aes(x=Context, y=normF1)) + 
+#   geom_boxplot() + xlab("Language context: B (Bengali) and E (English) and Tasks: S (Sentence) and P (Picture naming)") + ylab("Lobanov Normalized F1 at 5% of the vowel")
+# p + facet_grid(Vowel~Task)
+# 
+# p <- ggplot(bengali_phon_acco_raw, aes(x=Context, y=Vowel_Duration)) + 
+#   geom_boxplot() + xlab("Language context: B (Bengali) and E (English) and Tasks: S (Sentence) and P (Picture naming)") + ylab("Vowel duration (in ms)")
[…]

### 2021-03-19 writeup_final/writeup_final.tex
added Bengali data, figures, conclusion

@@
 
-\title{Mixed language processing increases cross-language phonetic transfer in Bengali-English Bilinguals}
+\title{Mixed language processing increases cross-language phonetic transfer in Bengali-English Bilinguals}  %\footnote{This paper is based on a research project at the Speech and Language Processing Lab at EFL University, Hyderabad.
+Guide: Prof.
+Indranil Dutta.
+To be communicated in 2021.}
 \date{}
@@
 %\begin{markdown}
+
+\begin{abstract}
+	Existing research demonstrates that mixed-language productions of bilingual speakers differ from productions in a single language.
+Studies have largely focused on temporal properties of consonants, and reported varied, and often contradictory, results.
+The present study examines the effect of mixed-language processing on two L2 vowel categories in a group of highly proficient Bengali-English bilingual speakers in India.
+Following from a discussion in \cite{olson2013bilingual}, we compare productions elicited in two paradigms-- cued picture-naming and code-switching.
+Results indicate a dynamic increase in cross-language phonetic transfer during mixed-language use leading to a shift towards L1 norms, asymmetrical shifts in the two categories, and no difference in transfer patterns between the two paradigms.
+\end{abstract}
 
@@
 \paragraph{} The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India.
-We measure the spectral properties (F1 and F2) of two English vowels to check if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2.
+We measure spectral properties (F1 and F2) of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2.
 Mixed-language data is elicited in two switching paradigms: code-switching and cued picture-naming.
@@
 Thus to preserve ecological validity, we elicited both unilingual and switched utterances in a bilingual language mode (cf section Methodology).
-Any observed differences in this paradigm would result from interaction during online processing.
+Any observed differences in this paradigm would arguably result from interaction during online processing.
 
@@
 As far as we are aware, there has been no work yet on vowels in L2.
-Thus, this study builds on the existing work by checking whether the L2 vowel quality of proficient bilinguals can be affected by mixed-language processing.
+Thus, this study builds on the existing work by examining whether the L2 vowel quality of proficient bilinguals can be affected by mixed-language processing.
 
@@
 \citeA{bullock2009trying} suggest that because long-lag languages offer a greater range of acceptable VOTs, there is more `room' for movement.
-In contrast, shift in short-lag languages could risk the loss of a phonological contrast (VOT is the primary cue for voicing contrast in most of these languages).
+In contrast, shift in short-lag languages could risk the loss of a phonological contrast (VOT being the primary cue for voicing contrast in these languages).
 This is a language-specific phonological constraint on transfer.
@@
 
- \paragraph{}Since Bengali has a 4-way laryngeal contrast, VOT does not encode a binary voiced/voiceless distinction, and moreover is not the only cue for the corresponding phonological categories \cite{dmitrieva2020acoustic}.
+ \paragraph{}Since Bengali has a four-way laryngeal contrast, VOT does not encode a binary voiced/voiceless distinction, and moreover is not the only cue for the corresponding phonological categories \cite{dmitrieva2020acoustic}.
 Thus, the effects of Bengali on English VOT are likely to be affected by additional phonological factors.
@@
 Thus, considering purely phonological constraints on IE, we would expect that a greater degree of shift is possible in \nt{2} compared to \nt{\ae}.
-Any movement in \nt{\ae} is expected to be limited to front-back (F2) dimension.
+Any movement in \nt{\ae} is expected to be primarily in the front-back (F2) dimension.
  	\item Target of transfer: Since we do not elicit unilingual L1 utterances in this study, we cannot directly compare the target vowels to baseline L1 vowels.
@@
  Flege's Speech Learning Model (SLM) \cite{flege1995second,flege2007language} suggests that sound categories which are common across languages influence each other because they share a common acoustic-phonetic space.
-(So if Bengali and English differ in their canonical realizations of \nt{\ae}, then we should expect the English \nt{\ae} to shift towards the corresponding Bengali category in the mixed condition.--- is there any literature on this?)
+Thus, if Bengali and English differ in their canonical realizations of \nt{\ae}, then we should expect the English \nt{\ae} to shift towards the corresponding Bengali category in the mixed condition.
 
@@
 
+\paragraph{}As noted above, we did not elicit unilingual Bengali productions from the participants in this study.
+To estimate a baseline value for the relevant Bengali vowels, productions in the unilingual English condition were compared to data from five female Bengali speakers from \citeA{dutta2016using}.
+Formant values were extracted from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel.
+Comparisons of Lobanov-normalized F1 and F2 suggests that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, compared to the English categories \nt{\ae} and \nt{2} respectively.
+
 \subsection{A note on paradigms: connected speech or not?} \label{paradigms}
@@
 
-\paragraph{}Note that while there is evidence for planning in CS, exactly how this affects phonetic transfer cannot be predicted from the existing literature-- while \cite{bullock2009trying} found tokens immediately preceding the switch point to be phonetically different from the rest, one group showed convergence towards the switch language(explained in terms of proficiency) while the other showed divergence (explained in terms of extra-linguistic factors).
+\paragraph{}Note that while there is evidence for planning in CS, exactly how this affects phonetic transfer cannot be predicted from the existing literature-- while \cite{bullock2009trying} found tokens immediately preceding the switch point to be phonetically distinct from the rest, one group showed convergence towards the switch language(explained in terms of proficiency) while the other showed divergence (explained in terms of extra-linguistic factors).
 The participants in Muldner et al.'s \citeyear{muldner2019phonetics} study hyperarticulated switched tokens (suggesting some degree of planning), and showed no significant shift.
@@
 Thus, the existence of additional factors in CS does not automatically mean that this should affect the extent of transfer compared to any other paradigm.
-It is possible that the effects in the studies mentioned above could be attributed to any other aspect of the experimental setup, rather than a result of code-switching in particular.
+It is possible that the effects in the studies mentioned above could be attributed to some other aspect of the experimental setup, rather than a result of code-switching in particular.
 
@@
 	\item Are L2 vowel categories of proficient bilingual speakers affected by dynamic phonetic transfer during mixed-language production?
-	\item (Asymmetries-- direction of shift in each vowel) 
+	\item If so, we expect an asymmetry between the vowels in the extent and direction of shift.
+Based on comparisons with Bengali vowels from \citeA{dutta2016using} we expect \nt{2} to be lowered and fronted in the mixed condition.
+\nt{\ae} is expected to show lowering, but to a lesser extent than \nt{2}, as it is in a part of the IE vowel space that has greater vowel density.
 	\item Is there an observable difference in the extent of transfer between the cued picture-naming and the code-switching paradigms? 
@@
 English words which are lexicalized loanwords in Bengali were avoided.
-The list of words can be found in sec.\ref{appendix}.
+%The list of words can be found in sec.\ref{appendix}.
 
@@
 
-\section{Results}
+\section{Results} \label{results}
 
@@
 The dependent variables were vowel height (F1) and backness (F2).
-We fitted a series of linear mixed effects models using the \emph{lme4} package \cite{lme4}, to check the effect of experimental conditions on lobanov normalized F1 and F2 at each of the five points in the vowel (sec.\ref{analysis}).
+We fitted a series of linear mixed effects models using the \emph{lme4} package \cite{lme4}, to examine the effect of experimental conditions on lobanov normalized F1 and F2 at each of the five points in the vowel (sec.\ref{analysis}).
 For all mixed models, the alpha criterion was set at $|t| > 2 $.
@@
 
-Context and Vowel are both between-subject factors.
+Context and Vowel are both within-subject factors.
 Since our hypothesis concerns their interaction, we modified the random effects structure by adding a by-subject random slope for the interaction term \cite{brauer2018linear}.
@@
 
-\subsection{Vowel Height (F1)}
+\subsection{Vowel Height -- F1}
 
@@
 
-	\includegraphics{vowels_by_subject_english}
+	\includegraphics{vowel_by_subject_ggplot}
 	\caption{Subject-wise mean F1 values for \nt{2} and \nt{\ae} uttered in the unilingual English condition}
@@
 
-\subsection{Vowel backness (F2)}
+\subsection{Vowel backness -- F2}
 
@@
 
-\section{Discussion}
+\section{Discussion and Conclusion}
 
-(Extent and direction of shift)
+With regard to our first research question, the results confirm that the L2 vowels of proficient bilinguals can be affected by mixed-language processing.
+Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.
+This is significant in the context of an ICM-based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, because it demonstrates that at least for vowels, the postulated lower switch cost on an L2 does not entirely mask transfer effects.
+Thus, following from the discussion in sec.
+\ref{asymmetries}, it might be fruitful for future studies to focus on vowels as a site for transfer.
+This would both expand the scope of the existing research (cf sec.
+\ref{asymmetry between sounds}) and avoid confounding factors arising from independent differences between long- and short-lag VOT languages.
+
+We hypothesized an asymmetrical shift between the two vowels, owing to differences in (i) their positions in the Indian English vowel space, and (ii) the existence of a corresponding category in the L1 Bengali inventory.
+The significant interaction between Vowel and Context in both F1 and F2 confirm that the experimental condition affected both height and backness differently for each vowel.
+As expected, we find a greater degree of shift in \nt{2}, possibly owing to its position in a less crowded part of the IE vowel space.
+Figure \ref{vowels_e_b} shows F1xF2 plots for the vowels in the unilingual and mixed (dotted ellipses) conditions, in comparison to the related Bengali categories (colored ellipses) from \citeA{dutta2016using}.
+The category \nt{2} is lowered and fronted in the mixed language condition as expected.
+The direction of shift in \nt{\ae} is surprising-- it is fronted in the mixed-language condition.
+While phonological constraints on IE suggested that a shift in the F1 dimension could potentially risk the loss of contrast and is thus expected to be minimal, Figure \ref{vowels_e_b} shows that \nt{\ae} in Bengali is not fronted with respect to the English category.
+Thus, fronting in the mixed condition cannot be explained in relation the former, at least from the present data.
+Data from a larger number of speakers, both for IE and Bengali, is needed to explore this further.
+
+\begin{figure}
+
+	\includegraphics{vowels_e_b_final}
+	\caption{Vowel categories in unilinual English (solid lines), mixed condition (dotted lines), and baseline Bengali (colored ellipses)}
+	\label{vowels_e_b}
+\end{figure}
 
 Our third research question concerned the experimental paradigms.
-\citeA{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, "discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
+\citeA{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
 This calls to question the comparability of results between studies using these different paradigms.
@@
 
-\section{Conclusion}
-Mixed language processing temporarily increases cross-language phonetic transfer, causing L2 vowels to (converge towards L1 norms ?).
+\paragraph{}
+Mixed language processing temporarily increases cross-language phonetic transfer, causing an L2 vowel to converge towards a related L1 category.
 The extent and direction of such shift is mediated by language-specific distribution of sound categories.
-In a controlled experimental setup to study transfer effects, the postulated discursive factors in a code-switching paradigm does not change the patterns of observed transfer.
+In a controlled experimental setup to study transfer effects, the postulated discursive factors in a code-switching paradigm does not change the patterns of observed transfer relative to a cued picture-naming paradigm.
 
-\section{Appendix} \label{appendix}
+%\section{Conclusion}
+
+%\section{Appendix} \label{appendix}
 

### 2021-05-03 draft_glossa/glossa-template.tex
glossa_draft

--- 2021-03-19 writeup_final/writeup_final.tex
@@
-% !TeX spellcheck = en_US
-\documentclass[11pt]{article}
-\usepackage{apacite}
-\usepackage{syntonly}
-\usepackage{booktabs}
-\usepackage{dcolumn}
-%\syntaxonly
-%
-%Margin - 1 inch on all sides
-%
-\usepackage[letterpaper]{geometry}
-\usepackage{times}
-\geometry{top=1.0in, bottom=1.0in, left=1.0in, right=1.0in}
+% !TeX document-id = {6d544fe2-4255-4108-8022-d809ff1d2c53}
+% TeX program = xelatexmk %doesn't work, no xelatexmk option under 'build'. Changed default engine to xelatex (options-->configure-->build-->default build-->xelatex)
+% see http://info.semprag.org/basics for a full description of this template
+\documentclass[charis,linguex]{glossa}
+% possible options: 
+% [times] for Times font (default if no option is chosen)
+% [cm] for Computer Modern font 
+% [lucida] for Lucida font (not freely available)
+% [brill] open type font, freely downloadable for non-commercial use from http://www.brill.com/about/brill-fonts; requires xetex
+% [charis] for CharisSIL font, freely downloadable from http://software.sil.org/charis/
+% for the Brill an CharisSIL fonts, you have to use the XeLatex typesetting engine (not pdfLatex)
+% for headings, tables, captions, etc., Fira Sans is used: https://www.fontsquirrel.com/fonts/fira-sans
+% [biblatex] for using biblatex (the default is natbib, do not load the natbib package in this file, it is loaded automatically via the document class glossa.cls)
+% [linguex] loads the linguex example package
+% !!
+a note on the use of linguex: in glossed examples, the third line of the example (the translation) needs to be prefixed with \glt.
+This is to allow a first line with the name of the language and the source of the example.
+See example (2) in the text for an illustration.
+% !! a note on the use of bibtex: for PhD dissertations to typeset correctly in the references list, the Address field needs to contain the city (for US cities in the format "Santa Cruz, CA")
 
-%
-%Doublespacing
-%
-\usepackage{setspace}
-%\doublespacing
-\usepackage{comment}
-%
-%Rotating tables (e.g. sideways when too long)
-%
-\usepackage{rotating}
+%\addbibresource{sample.bib} 
+% the above line is for use with biblatex
+% replace this by the name of your bib-file (extension .bib is required)
+% comment out if you use natbib/bibtex
 
-%\usepackage{hyperref}
-%
-%Fancy-header package to modify header/page numbering (insert last name)
-%
-\usepackage{fancyhdr}
-\pagestyle{fancy}
-\lhead{} 
-\chead{} 
-\rhead{\thepage} 
-\lfoot{} 
-\cfoot{} 
-\rfoot{} 
-\renewcommand{\headrulewidth}{0pt} 
-\renewcommand{\footrulewidth}{0pt} 
-%To make sure we actually have header 0.5in away from top edge
-%12pt is one-sixth of an inch. Subtract this from 0.5in to get headsep value
-\setlength\headsep{0.5in}
+%for natbib: replace \cite with \citep (parenthetical); \citea with \cite (in-text)
 
-%
-%Works cited environment
-%(to start, use \begin{workscited...}, each entry preceded by \bibent)
-% - from Ryan Alcock's MLA style file
-%
-%\newcommand{\bibent}{\noindent \hangindent 40pt}
-%\newenvironment{workscited}{\newpage \begin{center} References \end{center}}{\newpage }
+\let\B\relax %to resolve a conflict in the definition of these commands between xyling and xunicode (the latter called by fontspec, called by charis)
+\let\T\relax
+\usepackage{xyling} %for trees; the use of xyling with the CharisSIL font produces poor results in the branches. This problem does not arise with the packages qtree or forest.
+\usepackage[linguistics]{forest} %for nice trees!
 
@@
 
-\usepackage{graphicx} %for images
-\graphicspath{{images/}}
-\usepackage{caption}
+\usepackage{tabularx}
 
-\usepackage{enumitem} %customize list numberings
+%\graphicspath{{images/}}
 
-%
-%Begin document
-%
+% \pdf* commands provide metadata for the PDF output. ASCII characters only!
+\pdfauthor{Full author name}
+\pdftitle{Full title}
+\pdfkeywords{Full keyword list, separated, by, commas}
 
-\title{Mixed language processing increases cross-language phonetic transfer in Bengali-English Bilinguals}  %\footnote{This paper is based on a research project at the Speech and Language Processing Lab at EFL University, Hyderabad.
-Guide: Prof.
-Indranil Dutta.
-To be communicated in 2021.}
-\date{}
+\title[Mixed language processing increases cross-language phonetic transfer in Bengali-English Bilinguals}
+% Optional short title inside square brackets, for the running headers.
+
+\author[]% short form of the author names for the running header. If no short author is given, no authors print in the headers.
+{%as many authors as you like, each separated by \AND.
+  \spauthor{.\\ 
+  \institute{.\\
+  \small{%105, Bd. Raspail, 75005 Paris\\
+	}
+  }
+  \AND
+  \spauthor{.\\
+  \institute{.\\
+  \small{%Warmoesberg 26, 1000 Brussel\\
+  }
+  }%
+}
+
+%%%Notes
+% bengali sentence--gloss
 
@@
 
+\sffamily
 \maketitle
 
-	\bibliographystyle{apacite}
-%\begin{markdown}
+Word count: 4269
 
@@
 The present study examines the effect of mixed-language processing on two L2 vowel categories in a group of highly proficient Bengali-English bilingual speakers in India.
-Following from a discussion in \cite{olson2013bilingual}, we compare productions elicited in two paradigms-- cued picture-naming and code-switching.
+Following from a discussion in \citep{olson2013bilingual}, we compare productions elicited in two paradigms-- cued picture-naming and code-switching.
 Results indicate a dynamic increase in cross-language phonetic transfer during mixed-language use leading to a shift towards L1 norms, asymmetrical shifts in the two categories, and no difference in transfer patterns between the two paradigms.
 \end{abstract}
+
+\begin{keywords}
+
+\end{keywords}
+
+\rmfamily
 
@@
 (i) While a bilingual speaker is operating in any one of their languages.
-These studies often compare the speech of bilinguals to monolingual norms, and view transfer as changes to long-term memory representations as a result of acquiring an L2 \cite{guion2003vowel,caramazza1973acquisition,flege1987production}; (ii) When both languages of a bilingual speaker are co-activated.
+These studies often compare the speech of bilinguals to monolingual norms, and view transfer as changes to long-term memory representations as a result of acquiring an L2 \citep{guion2003vowel,caramazza1973acquisition,flege1987production}; (ii) When both languages of a bilingual speaker are co-activated.
 These compare productions during mixed-language processing to participants' own productions while using a single language.
@@
 
-\paragraph{ }Existing research on the phonetic effects of mixed-language production has largely focused on a few pairs of phonologically related languages, and used temporal properties of consonants (in particular, voice onset time or VOT) to measure transfer.
-However, great variation in results (cf sec.\ref{asymmetries} ) suggests that data from a wider variety of populations, language pairs,  and phonetic features is required in order to make meaningful generalizations.
+Existing research on the phonetic effects of mixed-language production has largely focused on a few pairs of phonologically related languages, and used temporal properties of consonants (in particular, voice onset time or VOT) to measure transfer.
+However, great variation in results (cf sec.\ref{asymmetries} ) suggests that data from a wider variety of
+%elaborate on 'great variation'-- differences between language pairs, plus sociolinguistic situation (effects of language experience etc)-- clearer.
+'Types of features'--separate line?- Brief mention - discursive factors in VOT.
+populations, language pairs,  and phonetic features is required in order to make meaningful generalizations.
 There is no work yet on short-term phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language.
@@
 
-\paragraph{} The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India.
+The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India.
 We measure spectral properties (F1 and F2) of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2.
@@
 
-\paragraph{}The next section provides background information on Bengali and English in India, focusing on vowel systems.
+The next section provides background information on Bengali and English in India, focusing on vowel systems.
 The rest of the introduction  discusses causes, duration, and possible sources of asymmetries in transfer reported in previous studies, paradigms used, and the design of the present study in light of these.
@@
 \subsubsection{Demography}
-	\paragraph{}Bengali (also, Bangla) is an Indo-Aryan language primarily spoken in India and Bangladesh.
+
+Bengali (also, Bangla) is an Indo-Aryan language primarily spoken in India and Bangladesh.
 In India, more than 97 million people speak Bengali as a first language (Census of India, 2011), mostly in the state of West Bengal.
@@
 
-	\paragraph{}Indian English (IE) refers to the variety of English that has developed in the Indian subcontinent.
+Indian English (IE) refers to the variety of English that has developed in the Indian subcontinent.
 In India, it is spoken as an L2 by 129 million people (Census of India, 2011).
-English is one of the two official languages, used in education, law, media, as a lingua franca mainly for an educated elite in most metropolitan regions, and carries a high prestige value \cite{tollefson2014language}.
+English is one of the two official languages, used in education, law, media, as a lingua franca mainly for an educated elite in most metropolitan regions, and carries a high prestige value \citep{tollefson2014language}.
 %citations 
-	In spite of the different L1s of its speakers, recent literature suggests that IE has a target phonology that is distinct from any of these, as well as from other native varieties of English (see \citeA{sirsa2013effects} for a review).
+In spite of the different L1s of its speakers, recent literature suggests that IE has a target phonology that is distinct from any of these, as well as from other native varieties of English (see \cite{sirsa2013effects} for a review).
 Thus, regional variations are seen as the result of L1-influence on a common underlying target.
-	%Early attempts to describe IE phonologically (General Indian English or GIE, \cite{masica1972sound}) were primarily attempts at standardization for pedagogical purposes. 
-	Excepting a small minority of specific communities \cite{pandey201517}, there are very few L1 speakers of IE.\\
+%Early attempts to describe IE phonologically (General Indian English or GIE, \citep{masica1972sound}) were primarily attempts at standardization for pedagogical purposes. 
+Excepting a small minority of specific communities \citep{pandey201517}, there are very few L1 speakers of IE.\\
 
-	\paragraph{}Considering these facts, the present study focuses on short-term phonetic interaction during mixed-language use, because:
+Considering these facts, the present study focuses on short-term phonetic interaction during mixed-language use, because:
 \begin{enumerate}[label=(\roman*)]
@@
 \subsubsection{Vowel systems} \label{vowel systems}
-	The vowel inventory of Western Bangla (the variety spoken by the participants in this study) consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \cite{garry2001facts}.
+The vowel inventory of Western Bangla (the variety spoken by the participants in this study) consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \citep{garry2001facts}.
 Note that there is no mid-central vowel category.
 
-	The vowel system of IE contains the monopthongs \nt{I, i, E, e, \ae, @/2, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \cite{wells1982accents, masica1972sound}.
+The vowel system of IE contains the monopthongs \nt{I, i, E, e, \ae, @/2, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \citep{wells1982accents, masica1972sound}.
 %Their distribution in the F1XF2 space is shown in (insert table 1.7, Pandey).
@@
 
-Existing research has distinguished between changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf SLM \cite{flege1995second,flege2007language}; PAM-L2 \cite{best2007nonnative}), and interaction during \textsc{processing} (transfer vs interference \cite{grosjean2012attempt}; competence vs performance interference \cite{paradis1993linguistic}).
-Given their transient nature (cf sec.\ref{duration}), dynamic changes in production during mixed-language use are generally thought to involve the latter \cite{elias2017effects,simonet2014phonetic}.
+Existing research has distinguished between changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf SLM \citep{flege1995second,flege2007language}; PAM-L2 \citep{best2007nonnative}), and interaction during \textsc{processing} (transfer vs interference \citep{grosjean2012attempt}; competence vs performance interference \citep{paradis1993linguistic}).
+Given their transient nature (cf sec.\ref{duration}), dynamic changes in production during mixed-language use are generally thought to involve the latter \citep{elias2017effects,simonet2014phonetic}.
 What triggers this interaction?
-\citeA{olson2016role} argues that while cross-language phonetic effects are largely measured at the point of language switch, it could have two potential sources:
+\cite{olson2016role} argues that while cross-language phonetic effects are largely measured at the point of language switch, it could have two potential sources:
 \begin{enumerate}
 	\item The local point of switch itself
-	\item Global co-activation of two languages -- bilingual language mode \cite{grosjean1998studying} 
+	\item Global co-activation of two languages -- bilingual language mode \citep{grosjean1998studying} 
 \end{enumerate}
@@
 A number of studies have specifically manipulated language mode, both in the presence and absence of switching.
-Overall, results suggest that (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \cite{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use -- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \cite{olson2016role,tsui2019impact} or spontaneous conversation \cite{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
-(iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \citeA{olson2016role} found no additive effects, \citeA{olson2013bilingual} found the balanced language context to inhibit transfer compared to unbalanced contexts.
-Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days \cite{schwartz2015language, bullock2009trying,antoniou2011inter, elias2017effects,vsimavckova2015immediate,vsimavckova2018patterns}.
+Overall, results suggest that (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use -- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \citep{olson2016role,tsui2019impact} or spontaneous conversation \citep{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
+(iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \cite{olson2016role} found no additive effects, \cite{olson2013bilingual} found the balanced language context to inhibit transfer compared to unbalanced contexts.
+Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days \cite{Pschwartz2015language, bullock2009trying,antoniou2011inter, elias2017effects,vsimavckova2015immediate,vsimavckova2018patterns}.
 
-\paragraph{}\citeA{grosjean1998studying} suggests that various aspects of the communicative setting, including exposure to (spoken or written) stimuli in multiple languages and awareness of the interlocutor's being bilingual, could trigger a bilingual mode.
+\cite{grosjean1998studying} suggests that various aspects of the communicative setting, including exposure to (spoken or written) stimuli in multiple languages and awareness of the interlocutor's being bilingual, could trigger a bilingual mode.
 Since the participants in this study are in an environment which largely contains mixed-language input, multilingual interlocutors, and no stable ambient language, we expect that phonetic transfer during daily language use, if any, takes place in a bilingual mode.
@@
 Over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts-- e.g.
-\citeA{bohn1992production}
+\cite{bohn1992production}
 %found that an approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language transfer.
@@
 %Moreover, changes due to transfer are not unidirectional or irreversible.
-\citeA{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language.
+\cite{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language.
 This not only showed that cross-language interaction can be triggered in an order of months, but also that the effects can be reversed within a similar time range.
-A recent study by \citeA{tobin2017phonetic} found comparable effects in an even shorter duration (2-4 weeks).
-A short-term longitudinal study by \citeA{chang2012rapid} showed that over the course of the first five weeks of learning an L2, there was a gradual convergence of L1 towards L2.
+A recent study by \cite{tobin2017phonetic} found comparable effects in an even shorter duration (2-4 weeks).
+A short-term longitudinal study by \cite{chang2012rapid} showed that over the course of the first five weeks of learning an L2, there was a gradual convergence of L1 towards L2.
 For some sounds and features (though not others, cf sec.\ref{asymmetry between sounds}), transfer was additive over time.
@@
 
-\paragraph{}Studies examining mixed-language use within an utterance mostly analyze just the switched (target) token.
+Studies examining mixed-language use within an utterance mostly analyze only the switched (target) token.
 Thus, very few direct measurements of the duration of transfer effects are available.
-One study which measured this \cite{bullock2009trying} did not find any residual effects on the matrix language following a switch, suggesting that transfer during code-switching is localized.
+One study which measured this \citep{bullock2009trying} did not find any residual effects on the matrix language following a switch, suggesting that transfer during code-switching is localized.
 
-More indirect evidence for this comes from studies which have elicited switched and non-switched tokens in the same test block \cite{tsui2019impact,olson2013bilingual}, and still reported differences between the two, suggesting that changes due to transfer are quickly `reset'-- in an order of seconds.
+More indirect evidence for this comes from studies which have elicited switched and non-switched tokens in the same test block \citep{tsui2019impact,olson2013bilingual}, and still reported differences between the two, suggesting that changes due to transfer are quickly `reset'-- in an order of seconds.
 Note that all these studies measure VOT, which is a temporal feature.
 We have no apriori reason to assume that these durations generalize to vowel quality.
-However, findings from sub-categorical phonetic shifts triggered by other factors (such as convergence towards an interlocutor) do evidence rapid shifts in vowel quality within comparable time-frames \cite{pardo2010expressing,babel2010dialect,babel2012evidence}.
+However, findings from sub-categorical phonetic shifts triggered by other factors (such as convergence towards an interlocutor) do evidence rapid shifts in vowel quality within comparable time-frames \citep{pardo2010expressing,babel2010dialect,babel2012evidence}.
 Thus in the current study, we present switched and non-switched tokens randomly within the same test block to induce a bilingual mode.
@@
 \subsubsection{L1 vs L2 status} 
-\paragraph{}Flege's Speech Learning Model (SLM) \citeyear{flege1995second,flege2007language} posits that the sound categories of a bilingual speaker exist in a common phonological space, and thus in principle, both L1 and L2 sounds can affect one another.
+Flege's Speech Learning Model (SLM) \citeyear{flege1995second,flege2007language} posits that the sound categories of a bilingual speaker exist in a common phonological space, and thus in principle, both L1 and L2 sounds can affect one another.
 Thus, the patterns of cross-language transfer depend on the mapping of L2 phonemes in relation to existing L1 categories.
-In non-switched production, this is evidenced through a pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (equivalence classification-- \citeNP{flege1984limits,flege1987production}), and
-the observation that both the L1 and L2 sound systems of bilinguals differ from corresponding monolingual speakers \cite{guion2003vowel}.
+In non-switched production, this is evidenced through a pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (equivalence classification-- \cite{flege1984limits,flege1987production}), and
+the observation that both the L1 and L2 sound systems of bilinguals differ from corresponding monolingual speakers \citep{guion2003vowel}.
 
-\paragraph{}In mixed-language production, the role of language status is less clear-- studies have reported unidirectional influence of L1 on L2 \cite{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \cite{tsui2019impact,elias2017effects, olson2013bilingual}, bidirectional convergence \cite{bullock2009trying, olson2016role}, divergence \cite{bullock2009trying,vsimavckova2018patterns}, and no influence \cite{muldner2019phonetics,schwartz2015language}.
+In mixed-language production, the role of language status is less clear-- studies have reported unidirectional influence of L1 on L2 \citep{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \citep{tsui2019impact,elias2017effects, olson2013bilingual}, bidirectional convergence \citep{bullock2009trying, olson2016role}, divergence \citep{bullock2009trying,vsimavckova2018patterns}, and no influence \citep{muldner2019phonetics,schwartz2015language}.
 Since most existing studies measure shifts in VOT, the observed asymmetries between languages have been variously explained either in terms of their L1 vs L2 status, or language-specific differences between long- and short-lag VOT languages (cf sec.
@@
 
- \paragraph{}\citeA{olson2013bilingual} first reported a unidirectional VOT shift of L1 towards L2 in two different groups -- native speakers of Spanish (short-lag) and English (long-lag), matched for proficiency and age of L2 acquisition.
+\cite{olson2013bilingual} first reported a unidirectional VOT shift of L1 towards L2 in two different groups -- native speakers of Spanish (short-lag) and English (long-lag), matched for proficiency and age of L2 acquisition.
 This suggested that beyond language-specific differences, the L1 vs L2 status of the language does mediate transfer.
-They explain the asymmetry in terms of the Inhibitory Control Model (ICM) \cite{green1998mental} -- to select a phonetic realization from one language, the other must be inhibited.
+They explain the asymmetry in terms of the Inhibitory Control Model (ICM) \citep{green1998mental} -- to select a phonetic realization from one language, the other must be inhibited.
 Being the `stronger' language, more inhibition is required on the L1.
@@
 The lesser switch cost results in a lack of visible cross-language transfer effects on L2.
-\citeA{tsui2019impact} found comparable results, but only in participants who were not equally dominant in both languages.
+\cite{tsui2019impact} found comparable results, but only in participants who were not equally dominant in both languages.
 Balanced bilinguals did not demonstrate any transfer effects in VOT.
@@
 
- \paragraph{}Since VOT is independently affected by short- vs long-lag differences, studies of other sound contrasts/features can avoid this conflation, and give a clearer picture of the effect of language status.
+Since VOT is independently affected by short- vs long-lag differences, studies of other sound contrasts/features can avoid this conflation, and give a clearer picture of the effect of language status.
 However, given the implications of the ICM, we would first have to establish that these L2 categories can indeed be affected by dynamic interference.
-The few existing studies which focus on vowel quality \cite{simonet2014phonetic,muldner2019phonetics,elias2017effects} and phonological processes \cite{simonet2020increased,schwartz2015language} have all examined transfer effects on L1.
+The few existing studies which focus on vowel quality \citep{simonet2014phonetic,muldner2019phonetics,elias2017effects} and phonological processes \citep{simonet2020increased,schwartz2015language} have all examined transfer effects on L1.
 As far as we are aware, there has been no work yet on vowels in L2.
@@
 In general, short-lag VOTs seem to resist accommodative shift.
-\citeA{bullock2009trying} suggest that because long-lag languages offer a greater range of acceptable VOTs, there is more `room' for movement.
+\cite{bullock2009trying} suggest that because long-lag languages offer a greater range of acceptable VOTs, there is more `room' for movement.
 In contrast, shift in short-lag languages could risk the loss of a phonological contrast (VOT being the primary cue for voicing contrast in these languages).
 This is a language-specific phonological constraint on transfer.
-  In a similar vein, \citeA{antoniou2011inter} found that while English stops shifted towards Greek (a short-lag language), the reverse was seen only in a subset of sounds-- word-medial non-nasalized stops.
+In a similar vein, \cite{antoniou2011inter} found that while English stops shifted towards Greek (a short-lag language), the reverse was seen only in a subset of sounds-- word-medial non-nasalized stops.
 They attribute this to the low frequency of these sounds in Greek, which makes them less `stable', and thus more amenable to shift.
@@
 
-  \paragraph{}These results highlight that beyond interaction during the cognitive processing of sounds, phonetic transfer is ultimately a linguistic phenomenon, and thus subject to language-specific constraints.
+These results highlight that beyond interaction during the cognitive processing of sounds, phonetic transfer is ultimately a linguistic phenomenon, and thus subject to language-specific phonological constraints.
 Therefore, to make meaningful generalizations, it is imperative to consider data from a variety of language pairs.
@@
 
- \paragraph{}Since Bengali has a four-way laryngeal contrast, VOT does not encode a binary voiced/voiceless distinction, and moreover is not the only cue for the corresponding phonological categories \cite{dmitrieva2020acoustic}.
+Since Bengali has a four-way laryngeal contrast, VOT does not encode a binary voiced/voiceless distinction, and moreover is not the only cue for the corresponding phonological categories \citep{dmitrieva2020acoustic}.
 Thus, the effects of Bengali on English VOT are likely to be affected by additional phonological factors.
@@
 
- \subsection{Asymmetries between sounds:}\label{asymmetry between sounds}
+\subsection{Asymmetries between sounds}\label{asymmetry between sounds}
 
 Existing research reports asymmetrical transfer effects not only between languages, but also between different sounds/features in a language, in both long-term and transient interactions.
-Studies which have examined multiple sound categories have found that interactions between individual sounds pairs do not necessarily reflect the overall pattern of global (system-wide) shift \cite{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot be seen as an atomic measure.
+Studies which have examined multiple sound categories have found that interactions between individual sounds pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot be seen as an atomic measure.
 In light of the discussion in section \ref{sound systems}, this is not surprising -- if a general linguistic principle of `room for movement' constrains transfer, then we should expect it to apply to individual sounds too.
@@
 
- \paragraph{}In the present study, we examine two vowel categories in Indian English-- the mid-central vowel \nt{2} and the low front vowel \nt{\ae}.
-There are two plausible sources of asymmetry between these:
+In the present study, we examine two vowel categories in Indian English-- the mid-central vowel \nt{2} and the low front vowel \nt{\ae}. There are two plausible sources of asymmetry between these:
 \begin{enumerate}
@@
 
- Flege's Speech Learning Model (SLM) \cite{flege1995second,flege2007language} suggests that sound categories which are common across languages influence each other because they share a common acoustic-phonetic space.
+Flege's Speech Learning Model (SLM) \citep{flege1995second,flege2007language} suggests that sound categories which are common across languages influence each other because they share a common acoustic-phonetic space.
 Thus, if Bengali and English differ in their canonical realizations of \nt{\ae}, then we should expect the English \nt{\ae} to shift towards the corresponding Bengali category in the mixed condition.
@@
 There is no obvious competing L1 category during the production of \nt{2}.
-However, it is unlikely that this should altogether preclude a shift in \nt{2}, since at least one existing study has reported transfer effects on a non-common vowel category in a comparable paradigm \cite{simonet2014phonetic}.
+However, it is unlikely that this should altogether preclude a shift in \nt{2}, since at least one existing study has reported transfer effects on a non-common vowel category in a comparable paradigm \citep{simonet2014phonetic}.
 Here, the Catalan \nt{O} moved towards an acoustically close category \nt{o}, which is common to Catalan and Spanish, suggesting that shift might be towards a set of phonetic norms rather than a particular target.
@@
 
-\paragraph{}As noted above, we did not elicit unilingual Bengali productions from the participants in this study.
-To estimate a baseline value for the relevant Bengali vowels, productions in the unilingual English condition were compared to data from five female Bengali speakers from \citeA{dutta2016using}.
+As noted above, we did not elicit unilingual Bengali productions from the participants in this study.
+To estimate a baseline value for the relevant Bengali vowels, productions in the unilingual English condition were compared to data from five female Bengali speakers from \cite{dutta2016using}.
 Formant values were extracted from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel.
@@
 
-Studies of phonetic transfer during mixed-language use have largely used spontaneous or scripted code-switching (CS, defined as the use of multiple languages within a single utterance, ie in connected speech; \citeNP{myers1993dueling}) to co-activate the two languages.
+Studies of phonetic transfer during mixed-language use have largely used spontaneous or scripted code-switching (CS, defined as the use of multiple languages within a single utterance, ie in connected speech; \citealp{myers1993dueling}) to co-activate the two languages.
 Other paradigms include cued picture-naming, delayed repetition, interpreting across languages, reading word-lists etc.
-\citeA{olson2013bilingual} argues that code-switching introduces discursive factors such as discourse planning and pragmatic considerations, which ultimately show patterns of language practice, rather than cognitive behavior.
-This finds support from studies which report transfer effects before the switch point \cite{bullock2009trying}, or other evidence of planning such as hyperarticulation \cite{muldner2019phonetics} and interlocutor-awareness in spontaneous CS \cite{khattab2013phonetic}.
+\cite{olson2013bilingual} argues that code-switching introduces discursive factors such as discourse planning and pragmatic considerations, which ultimately show patterns of language practice, rather than cognitive behavior.
+This finds support from studies which report transfer effects before the switch point \citep{bullock2009trying}, or other evidence of planning such as hyperarticulation \citep{muldner2019phonetics} and interlocutor-awareness in spontaneous CS \citep{khattab2013phonetic}.
 To avoid such confounds, Olson suggests cued picture-naming (where pictures are named in rapid succession in either language, based on the given language cue) as an alternative paradigm to isolate purely phonetic effects in controlled setups.
 
-\paragraph{}Note that while there is evidence for planning in CS, exactly how this affects phonetic transfer cannot be predicted from the existing literature-- while \cite{bullock2009trying} found tokens immediately preceding the switch point to be phonetically distinct from the rest, one group showed convergence towards the switch language(explained in terms of proficiency) while the other showed divergence (explained in terms of extra-linguistic factors).
+Note that while there is evidence for planning in CS, exactly how this affects phonetic transfer cannot be predicted from the existing literature-- while \citep{bullock2009trying} found tokens immediately preceding the switch point to be phonetically distinct from the rest, one group showed convergence towards the switch language(explained in terms of proficiency) while the other showed divergence (explained in terms of extra-linguistic factors).
 The participants in Muldner et al.'s \citeyear{muldner2019phonetics} study hyperarticulated switched tokens (suggesting some degree of planning), and showed no significant shift.
@@
 
-\paragraph{}While much more work is necessary to tease out the various factors and their interactions, the focus of the present study is to check whether, in an experimental setting, the postulated discursive elements in a CS paradigm affect the extent of observed transfer, compared to a paradigm which arguably lacks them--cued picture-naming.
+While much more work is necessary to tease out the various factors and their interactions, the focus of the present study is to check whether, in an experimental setting, the postulated discursive elements in a CS paradigm affect the extent of observed transfer, compared to a paradigm which arguably lacks them--cued picture-naming.
 We are less concerned with the exact nature of differences, than with their comparability.
@@
 	\item If so, we expect an asymmetry between the vowels in the extent and direction of shift.
-Based on comparisons with Bengali vowels from \citeA{dutta2016using} we expect \nt{2} to be lowered and fronted in the mixed condition.
+Based on comparisons with Bengali vowels from \cite{dutta2016using} we expect \nt{2} to be lowered and fronted in the mixed condition.
 \nt{\ae} is expected to show lowering, but to a lesser extent than \nt{2}, as it is in a part of the IE vowel space that has greater vowel density.
@@
 Participants were recorded individually in a sound attenuated room. %, using a ... microphone type/description?
-Productions were recorded in Praat \cite{boersma2016praat} at a 44.1K sampling rate.
+Productions were recorded in Praat \citep{boersma2016praat} at a 44.1K sampling rate.
 Every participant performed four iterations each of two tasks-- cued picture-naming and code-switching.
@@
 An additional 67 tokens (2.79\%) were excluded due to errors in presentation/production, giving 2333 target word items for analysis.
-Recordings were manually segmented and annotated in Praat \cite{boersma2016praat}, and the first and second formant frequencies (F1 and F2) of the target vowels \nt{2, ae} were measured at five points-- in 10\% increments starting at 5\% into the vowel.
+Recordings were manually segmented and annotated in Praat \citep{boersma2016praat}, and the first and second formant frequencies (F1 and F2) of the target vowels \nt{2, ae} were measured at five points-- in 10\% increments starting at 5\% into the vowel.
 This gave measures of F1 and F2 at 5\%, 15\%, 25\%, 35\% and 45\% into the vowel, allowing us to observe any dynamic changes in the extent of transfer over the course of producing the vowel.
@@
 %Figures ... show the findings of the experiment. 
-All statistical analyses were performed in R \cite{r}.
+All statistical analyses were performed in R \citep{r}.
 The dependent variables were vowel height (F1) and backness (F2).
-We fitted a series of linear mixed effects models using the \emph{lme4} package \cite{lme4}, to examine the effect of experimental conditions on lobanov normalized F1 and F2 at each of the five points in the vowel (sec.\ref{analysis}).
+We fitted a series of linear mixed effects models using the \emph{lme4} package \citep{lme4}, to examine the effect of experimental conditions on lobanov normalized F1 and F2 at each of the five points in the vowel (sec.\ref{analysis}).
 For all mixed models, the alpha criterion was set at $|t| > 2 $.
@@
 	\caption{Fixed factors}
+	\sffamily
 	\centering
@@
 		Gender & male, female\\
-		Vowel & \textbackslash vt (\nt{2}), \textbackslash ae (\nt{\ae}) \\
+		Vowel & \nt{2}, \nt{\ae} \\
 		Task (corresponding to the two paradigms) & p (Picture: cued picture-naming, s (Sentence: code-switching)\\
@@
 Context and Vowel are both within-subject factors.
-Since our hypothesis concerns their interaction, we modified the random effects structure by adding a by-subject random slope for the interaction term \cite{brauer2018linear}.
+Since our hypothesis concerns their interaction, we modified the random effects structure by adding a by-subject random slope for the interaction term \citep{brauer2018linear}.
 This improved the model fit ($\chi^2(5) = 11.43, p = 0.04342$).
@@
 Instead, the significant interaction between Context and Vowel indicates a lowering in \nt{2}, but not in \nt{\ae}, in the mixed-language condition.
-The results also reveal a significant effect of Task, such that the sentence task (code-switching) lowers both vowels.
+The results also reveal a significant effect of Task, such that the sentence task (code-switching) leads to higher F1 values in both vowels, relative to the cued picture-naming task.
 This effect is strongest at the beginning of the vowel, and disappears by 45\%.
@@
 %ANOVA table add
-
-\begin{figure}
-
-	\includegraphics{vowel_by_subject_ggplot}
-	\caption{Subject-wise mean F1 values for \nt{2} and \nt{\ae} uttered in the unilingual English condition}
-	\label{boxplot_F1}
-\end{figure}
 
@@
 
+\begin{figure}
+
+	\includegraphics[scale=0.8]{vowel_by_subject_ggplot}
+	\caption{Subject-wise mean F1; unilingual English}
+	\label{boxplot_F1}
+\end{figure}
+
 \section{Discussion and Conclusion}
@@
 Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.
-This is significant in the context of an ICM-based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, because it demonstrates that at least for vowels, the postulated lower switch cost on an L2 does not entirely mask transfer effects.
+This is significant in the context of an ICM-based analysis of between-language asymmetries in phonetic transfer as proposed by \citep{olson2013bilingual}, because it demonstrates that at least for vowels, the postulated lower switch cost on an L2 does not entirely mask transfer effects.
 Thus, following from the discussion in sec.
@@
 As expected, we find a greater degree of shift in \nt{2}, possibly owing to its position in a less crowded part of the IE vowel space.
-Figure \ref{vowels_e_b} shows F1xF2 plots for the vowels in the unilingual and mixed (dotted ellipses) conditions, in comparison to the related Bengali categories (colored ellipses) from \citeA{dutta2016using}.
+Figure \ref{vowels_e_b} shows F1xF2 plots for the vowels in the unilingual and mixed (dotted ellipses) conditions, in comparison to the related Bengali categories (colored ellipses) from \cite{dutta2016using}.
 The category \nt{2} is lowered and fronted in the mixed language condition as expected.
@@
 
+% only ae -- shift significant?
+% difference between vowels-- transfer--makes reference to phonological categories--
+% knitr -- put the R code for figures in here so changes update here
+%in intro--say-- paper argues that these asymetries (seen in previous lit) point to the (urgent:)) need for diversifying the languages and sound features studied in order to make generalizations
 \begin{figure}
 
-	\includegraphics{vowels_e_b_final}
+	\includegraphics[scale=0.8]{vowels_e_b_final}
 	\caption{Vowel categories in unilinual English (solid lines), mixed condition (dotted lines), and baseline Bengali (colored ellipses)}
@@
 Our third research question concerned the experimental paradigms.
-\citeA{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
+\cite{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
 This calls to question the comparability of results between studies using these different paradigms.
@@
 We found no significant interaction of Task either with Context, or with the Vowel*Context interaction, suggesting that the extent of phonetic transfer was not affected by the paradigm.
-There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in transfer at all.
-This would mean that the hyperarticulation seen, for example, in \citeA{muldner2019phonetics} is better interpreted as resulting from some other aspect of the experimental setup, rather than the act of code-switching itself.
+There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in observable transfer at all.
+This would mean that the hyperarticulation seen, for example, in \cite{muldner2019phonetics} is better interpreted as resulting from some other aspect of the experimental setup, rather than the act of code-switching itself.
 While planning could additionally affect the words preceding the switch, the effects on the switched token itself do not change.
-(ii) Additionally, our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
-Thus, another possible explanation is that the discursive effects discussed by \citeA{olson2013bilingual} are triggered by more naturalistic paradigms where attention to meaning and interlocutor are important.
+(ii) Our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
+Thus, another possible explanation is that the discursive effects discussed by \cite{olson2013bilingual} are triggered by more naturalistic paradigms where attention to meaning and interlocutor are important.
 More within-study comparisons with less structured paradigms, and spontaneous code-switching data, can shed light on this.
 The results of the present study suggest that for the purposes of observing dynamic phonetic transfer, elicitation through cued-picture-naming vs (controlled) code-switching does not seem to significantly alter the behavior under study, and thus the choice of paradigm can be based on other considerations.
-Moreover, this suggests that the results from such studies can be compared directly.
+Moreover, this suggests that (all else being equal) the results from such studies can be compared directly.
 
 Paradigm did affect vowel quality, but this applied uniformly across experimental conditions-- all vowels were more backed and lowered in the code-switching task.
-Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this is an artifact of the particular stimuli used here-- resulting from some corarticulatory effect that was absent in the picture-naming task.
+Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this lowering is an artifact of the particular stimuli used here-- resulting from some corarticulatory effect that was absent in the picture-naming task.
 While this is intriguing, it is beyond the scope of the present study.
 
-\paragraph{}
 Mixed language processing temporarily increases cross-language phonetic transfer, causing an L2 vowel to converge towards a related L1 category.
@@
 
-%\section{Conclusion}
+%\nocite{*} %this is to get all the entries of the sample bibliography; delete this line for an actual Glossa submission
 
-%\section{Appendix} \label{appendix}
-
-%\end{markdown}
-%\bibliography{F:/DISHA_DATA/Disha_study/LINGUISTICS/Shastri/references_srsf.bib}
-%\bibliography{C:/Users/91767/Desktop/Project/Writeup/references.bib}
-\bibliography{references}
+%\printbibliography %for use with biblatex; comment out if you use natbib
+\bibliography{references,sample} %for use with natbib; comment out if you use biblatex, and change 'sample' by the name of your bib-file
 
 \end{document}
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% TeX-engine: luatex
+%%% End:

### 2021-05-03 writeup_final/writeup_final.tex
glossa_draft

@@
 \paragraph{ }Existing research on the phonetic effects of mixed-language production has largely focused on a few pairs of phonologically related languages, and used temporal properties of consonants (in particular, voice onset time or VOT) to measure transfer.
-However, great variation in results (cf sec.\ref{asymmetries} ) suggests that data from a wider variety of populations, language pairs,  and phonetic features is required in order to make meaningful generalizations.
+However, great variation in results (cf sec.\ref{asymmetries} ) suggests that data from a wider variety of
+%elaborate on 'great variation'-- differences between language pairs, plus sociolinguistic situation (effects of language experience etc)-- clearer.
+'Types of features'--separate line?- Brief mention - discursive factors in VOT.
+ populations, language pairs,  and phonetic features is required in order to make meaningful generalizations.
 There is no work yet on short-term phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language.
@@
 
+%
 \paragraph{} The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India.
@@
 
-\paragraph{}Studies examining mixed-language use within an utterance mostly analyze just the switched (target) token.
+\paragraph{}Studies examining mixed-language use within an utterance mostly analyze only the switched (target) token.
 Thus, very few direct measurements of the duration of transfer effects are available.
@@
 
-  \paragraph{}These results highlight that beyond interaction during the cognitive processing of sounds, phonetic transfer is ultimately a linguistic phenomenon, and thus subject to language-specific constraints.
+  \paragraph{}These results highlight that beyond interaction during the cognitive processing of sounds, phonetic transfer is ultimately a linguistic phenomenon, and thus subject to language-specific phonological constraints.
 Therefore, to make meaningful generalizations, it is imperative to consider data from a variety of language pairs.
@@
 
- \subsection{Asymmetries between sounds:}\label{asymmetry between sounds}
+ \subsection{Asymmetries between sounds}\label{asymmetry between sounds}
 
@@
 
+% only ae -- shift significant?
+% difference between vowels-- transfer--makes reference to phonological categories--
+% knitr -- put the R code for figures in here so changes update here
+%in intro--say-- paper argues that these asymetries (seen in previous lit) point to the (urgent:)) need for diversifying the languages and sound features studied in order to make generalizations
 \begin{figure}
@@
 We found no significant interaction of Task either with Context, or with the Vowel*Context interaction, suggesting that the extent of phonetic transfer was not affected by the paradigm.
-There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in transfer at all.
+There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in observable transfer at all.
 This would mean that the hyperarticulation seen, for example, in \citeA{muldner2019phonetics} is better interpreted as resulting from some other aspect of the experimental setup, rather than the act of code-switching itself.
 While planning could additionally affect the words preceding the switch, the effects on the switched token itself do not change.
-(ii) Additionally, our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
+(ii) Our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
 Thus, another possible explanation is that the discursive effects discussed by \citeA{olson2013bilingual} are triggered by more naturalistic paradigms where attention to meaning and interlocutor are important.
@@
 The results of the present study suggest that for the purposes of observing dynamic phonetic transfer, elicitation through cued-picture-naming vs (controlled) code-switching does not seem to significantly alter the behavior under study, and thus the choice of paradigm can be based on other considerations.
-Moreover, this suggests that the results from such studies can be compared directly.
+Moreover, this suggests that (all else being equal) the results from such studies can be compared directly.
 
 Paradigm did affect vowel quality, but this applied uniformly across experimental conditions-- all vowels were more backed and lowered in the code-switching task.
-Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this is an artifact of the particular stimuli used here-- resulting from some corarticulatory effect that was absent in the picture-naming task.
+Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this lowering is an artifact of the particular stimuli used here-- resulting from some corarticulatory effect that was absent in the picture-naming task.
 While this is intriguing, it is beyond the scope of the present study.

### 2021-05-04 draft_glossa/glossa-template.tex
fixed tables, word count

@@
 
-\title[Mixed language processing increases cross-language phonetic transfer in Bengali-English Bilinguals}
+\title{Mixed language processing increases cross-language phonetic transfer in Bengali-English Bilinguals}
 % Optional short title inside square brackets, for the running headers.
@@
   \institute{.\\
-  \small{%105, Bd. Raspail, 75005 Paris\\
+  \small{.\\
 	}
@@
   \institute{.\\
-  \small{%Warmoesberg 26, 1000 Brussel\\
+  \small{.\\
   }
@@
 
-Word count: 4269
+Word count: 6944
 
@@
 
-%Studies which compare bilinguals speakers' phonological systems with monolinguals see them as relatively stable over time.
+Studies which compare bilinguals speakers' phonological systems with monolinguals see them as relatively stable over time.
 The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods, \cite<e.g.>{guion2003vowel, caramazza1973acquisition}.
 However, longitudinal studies %(citations)
-%and between-subject comparisons of bilingual speakers who differ in the duration of L2 exposure suggest that interaction between the sound systems is dynamic.
+and between-subject comparisons of bilingual speakers who differ in the duration of L2 exposure suggest that interaction between the sound systems is dynamic.
 Over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts-- e.g.
 \cite{bohn1992production}
-%found that an approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language transfer.
+found that an approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language transfer.
 These have been interpreted as changes to long-term phonological representations.
 
-%Moreover, changes due to transfer are not unidirectional or irreversible.
+Moreover, changes due to transfer are not unidirectional or irreversible.
 \cite{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language.
@@
 At the time of recording, they were living in or around the EFL University campus in Hyderabad, which has a linguistically diverse population.
-Speakers were selected on the basis of responses to a Language Background Questionnaire (sec.\ref{appendix}) to ensure comparable LSRW skills in L1 and L2 across participants, and were compensated for their time.
+Speakers were selected on the basis of responses to a Language Background Questionnaire to ensure comparable LSRW skills in L1 and L2 across participants, and were compensated for their time.
 
@@
 
-\begin{table}[h]
+\begin{table}
 	\caption{Fixed factors}
@@
 	\centering
-	\begin{tabular}{l l}
+	\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
 		\hline 
@@
 
+\begin{figure}
+
+	\includegraphics[scale=0.8]{vowel_by_subject_ggplot}
+	\caption{Subject-wise mean F1; unilingual English}
+	\label{boxplot_F1}
+\end{figure}
+
 \begin{table}[htbp] \centering 
@@
 		& & & & & \\ 
-		Vowel\textbackslash vt & 0.022 & $-$0.070 & $-$0.034 & $-$0.131 & $-$0.216 \\ 
+		Vowel\nt{2} & 0.022 & $-$0.070 & $-$0.034 & $-$0.131 & $-$0.216 \\ 
 		& (0.170) & (0.175) & (0.170) & (0.174) & (0.168) \\ 
@@
 		& & & & & \\ 
-		Contexte:Vowel\textbackslash vt & $-$0.126$^{**}$ & $-$0.138$^{**}$ & $-$0.185$^{***}$ & $-$0.191$^{***}$ & $-$0.192$^{***}$ \\ 
+		Contexte:Vowel\nt{2} & $-$0.126$^{**}$ & $-$0.138$^{**}$ & $-$0.185$^{***}$ & $-$0.191$^{***}$ & $-$0.192$^{***}$ \\ 
 		& (0.056) & (0.060) & (0.055) & (0.064) & (0.066) \\ 
@@
 		& & & & & \\ 
-		Vowel\textbackslash vt & $-$0.217 & $-$0.615$^{***}$ & $-$0.651$^{***}$ & $-$0.801$^{***}$ & $-$0.836$^{***}$ \\ 
+		Vowel\nt{2} & $-$0.217 & $-$0.615$^{***}$ & $-$0.651$^{***}$ & $-$0.801$^{***}$ & $-$0.836$^{***}$ \\ 
 		& (0.175) & (0.183) & (0.175) & (0.168) & (0.166) \\ 
@@
 		& & & & & \\
-		Contexte:Vowel\textbackslash vt & 0.031 & 0.110$^{***}$ & 0.128$^{***}$ & 0.091$^{**}$ & 0.147$^{***}$ \\ 
+		Contexte:Vowel\nt{2} & 0.031 & 0.110$^{***}$ & 0.128$^{***}$ & 0.091$^{**}$ & 0.147$^{***}$ \\ 
 		& (0.050) & (0.041) & (0.040) & (0.046) & (0.043) \\ 
@@
 \end{table}
-
-\begin{figure}
-
-	\includegraphics[scale=0.8]{vowel_by_subject_ggplot}
-	\caption{Subject-wise mean F1; unilingual English}
-	\label{boxplot_F1}
-\end{figure}
 
@@
 
-%\nocite{*} %this is to get all the entries of the sample bibliography; delete this line for an actual Glossa submission
-
 %\printbibliography %for use with biblatex; comment out if you use natbib
-\bibliography{references,sample} %for use with natbib; comment out if you use biblatex, and change 'sample' by the name of your bib-file
+\bibliography{references} %for use with natbib; comment out if you use biblatex, and change 'sample' by the name of your bib-file
 

### 2021-05-24 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
author info and affil

--- 2021-05-04 draft_glossa/glossa-template.tex
@@
 % \pdf* commands provide metadata for the PDF output. ASCII characters only!
-\pdfauthor{Full author name}
-\pdftitle{Full title}
-\pdfkeywords{Full keyword list, separated, by, commas}
+\pdfauthor{Auromita Mitra, Indranil Dutta}
+\pdftitle{Mixed language processing increases cross-language phonetic transfer in Bengali-English Bilinguals}
+\pdfkeywords{phonetic transfer, vowel formants, Bengali, English}
 
@@
 
-\author[]% short form of the author names for the running header. If no short author is given, no authors print in the headers.
+\author[Auromita \& Indranil]% short form of the author names for the running header. If no short author is given, no authors print in the headers.
 {%as many authors as you like, each separated by \AND.
-  \spauthor{.\\ 
-  \institute{.\\
-  \small{.\\
-	}
+  \spauthor{Auromita Mitra\\ 
+  \institute{The EFL University}\\
+  \small{%105, Bd. Raspail, 75005 Paris\\
+  auromita.mitra@gmail.com}
   }
   \AND
-  \spauthor{.\\
-  \institute{.\\
-  \small{.\\
-  }
+  \spauthor{Indranil Dutta \\
+  \institute{Jadavpur University}\\
+  \small{%Warmoesberg 26, 1000 Brussel\\
+  indranildutta.lnl@jadavpuruniversity.in}
   }%
 }
-
-%%%Notes
-% bengali sentence--gloss
-
 \begin{document}
@@
 Studies which compare bilinguals speakers' phonological systems with monolinguals see them as relatively stable over time.
-The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods, \cite<e.g.>{guion2003vowel, caramazza1973acquisition}.
+The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods, \cite{guion2003vowel, caramazza1973acquisition}.
 However, longitudinal studies %(citations)
@@
 Existing research on phonetic transfer largely revolves around a small set of phonologically related languages.
-Thus, the present study extends the scope of this research to a new set of languages-- Indian English and Bengali (an Indo-Aryan language).
+Thus, the present study extends the scope of this research to a new set of languages---Indian English and Bengali (an Indo-Aryan language).
 
-Since Bengali has a four-way laryngeal contrast, VOT does not encode a binary voiced/voiceless distinction, and moreover is not the only cue for the corresponding phonological categories \citep{dmitrieva2020acoustic}.
+Since Bengali has a four-way laryngeal contrast, VOT does not encode a binary voiced---voiceless distinction, and moreover is not the only cue for the corresponding phonological categories \citep{dmitrieva2020acoustic}.
 Thus, the effects of Bengali on English VOT are likely to be affected by additional phonological factors.

### 2021-05-24 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
figure legends, glosses

@@
 
-\usepackage{tabularx}
-
 %\graphicspath{{images/}}
@@
 }
+%\usepackage{gb4e}
 \begin{document}
@@
 		\item Unilingual sentence: That is a yellow [Target Word].
-		\item Mixed-language sentence: \textipa{o-ta \ae k-ta ka:lo} [Target Word].
+		\item Mixed-language sentence: \textipa{o-\:ta \ae k-\:ta kalo} [Target Word]. 
 	\end{itemize}
+
+Gloss and translation for Bengali sentence:
+
+	o-ta \quad Ek-ta \quad kalo [target word]\\
+	that-CL one-CL black [target word]\\
+	that is a black [target word]\\
 

### 2021-05-24 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
emdashes added

--- 2021-05-24 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
@@
 
-Paradigm did affect vowel quality, but this applied uniformly across experimental conditions-- all vowels were more backed and lowered in the code-switching task.
-Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this lowering is an artifact of the particular stimuli used here-- resulting from some corarticulatory effect that was absent in the picture-naming task.
+Paradigm did affect vowel quality, but this applied uniformly across experimental conditions---all vowels were more backed and lowered in the code-switching task.
+Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this lowering is an artifact of the particular stimuli used here---resulting from some corarticulatory effect that was absent in the picture-naming task.
 While this is intriguing, it is beyond the scope of the present study.

### 2021-05-24 RCode/bengali_phon_acco_new.R
IPA, legends in figures

@@ -1,4 +1,4 @@
-formant_data_5 <- read.csv("~/Desktop/Disha_Project/formant_data_5.csv")
+formant_data_5 <- read.csv("~/Desktop/Disha_Project/formant_data_5.csv", check.names = TRUE)
 attach(formant_data_5)
 detach(formant_data_5)
 
@@ -33,7 +33,7 @@ detach(bengali)
 bengaliAE <- subset (bengali, V1=="A"|V1=="E")
 attach(bengaliAE)
 detach(bengaliAE)
-bengaliAE_bd
+
 
 
 #####
@@ -97,26 +97,29 @@ detach(taskp)
 
 # subject-wise F1 in English context
 attach(contexte)
-ggplot(contexte, aes(x = Subject, y = enormF1$F1_25, fill = Vowel)) + ylab("Normalized F1") +geom_boxplot()
+remapping <- c(`\\ae` = "æ", `\\vt` = "ʌ")
+contexte$unicodevowel <- remapping[as.character(contexte$Vowel)]
+ggplot(contexte, aes(x = Subject, y = enormF1$F1_25, fill = unicodevowel)) + ylab("Normalized F1") +scale_fill_discrete(name="Vowel")+ geom_boxplot()
 detach(contexte)
 
 
 # vowels: unilingual English, mixed, and baseline Bengali
-xlim <- c(3,-3)
-ylim = c(3,-3)
+xlim <- c(2.5,-2.5)
+ylim = c(3.5,-2.5)
 
-remapping <- c(A = "a:", E = "\\ae")
+remapping <- c(A = "a:", E = "æ")
 bengaliAE$unicodevowel <- remapping[as.character(bengaliAE$V1)]
 with(bengaliAE,plotVowels(F1_V1_T55,F2_V1_T55, unicodevowel, plot.tokens = F, plot.means = TRUE,
-                        pch.tokens = V1, pch.means = unicodevowel, cex.tokens = 0.8, var.col.by = V1,
-                        cex.means = 0.8, alpha.means = 0.8,alpha.tokens = 0.1,ellipse.line = F,  
-                        ellipse.fill = T, pretty = TRUE, xlim = xlim, ylim = ylim, axes=FALSE))
+                        pch.tokens = V1, pch.means = unicodevowel, cex.tokens = 0.8, var.col.by = unicodevowel,
+                        cex.means = 1, alpha.means = 0.8,alpha.tokens = 0.1,ellipse.line = F,  
[…]

### 2021-05-25 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
r code addtions

@@
 
-\author[Auromita \& Indranil]% short form of the author names for the running header. If no short author is given, no authors print in the headers.
-{%as many authors as you like, each separated by \AND.
-  \spauthor{Auromita Mitra\\ 
-  \institute{The EFL University}\\
-  \small{%105, Bd. Raspail, 75005 Paris\\
-  auromita.mitra@gmail.com}
-  }
-  \AND
-  \spauthor{Indranil Dutta \\
-  \institute{Jadavpur University}\\
-  \small{%Warmoesberg 26, 1000 Brussel\\
-  indranildutta.lnl@jadavpuruniversity.in}
-  }%
-}
+% \author[Auromita \& Indranil]% short form of the author names for the running header. If no short author is given, no authors print in the headers.
+% {%as many authors as you like, each separated by \AND.
+%   \spauthor{Auromita Mitra\\ 
+%   \institute{The EFL University}\\
+%   \small{%105, Bd. Raspail, 75005 Paris\\
+%   auromita.mitra@gmail.com}
+%   }
+%   \AND
+%   \spauthor{Indranil Dutta \\
+%   \institute{Jadavpur University}\\
+%   \small{%Warmoesberg 26, 1000 Brussel\\
+%   indranildutta.lnl@jadavpuruniversity.in}
+%   }%
+% }
 %\usepackage{gb4e}

### 2021-05-25 RCode/bengali_phon_acco_new.R

@@ -1,4 +1,4 @@
-formant_data_5 <- read.csv("~/Desktop/Disha_Project/formant_data_5.csv")
+formant_data_5 <- read.csv("formant_data_5.csv")
 attach(formant_data_5)
 detach(formant_data_5)
 
@@ -27,7 +27,7 @@ with(formant_data_5, plotVowels(normF1$F1_15, normF2$F2_15, Vowel, group = Conte
 
 
 ## bengali baseline data
-bengali <- read.csv("~/Desktop/Disha_Project/bengali_shruti/anti.csv", stringsAsFactors = TRUE, header = TRUE)
+bengali <- read.csv("anti.csv", stringsAsFactors = TRUE, header = TRUE)
 attach(bengali)
 detach(bengali)
 bengaliAE <- subset (bengali, V1=="A"|V1=="E")

### 2021-05-29 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex

@@
 In India, it is spoken as an L2 by 129 million people (Census of India, 2011).
-English is one of the two official languages, used in education, law, media, as a lingua franca mainly for an educated elite in most metropolitan regions, and carries a high prestige value \citep{tollefson2014language}.
+English is one of the two official languages, used in education, law, media, as a \emph{lingua franca} mainly for an educated elite in most metropolitan regions, and carries a high prestige value \citep{tollefson2014language}.
 %citations 
@@
 %Early attempts to describe IE phonologically (General Indian English or GIE, \citep{masica1972sound}) were primarily attempts at standardization for pedagogical purposes. 
-Excepting a small minority of specific communities \citep{pandey201517}, there are very few L1 speakers of IE.\\
+Except for a small minority of specific communities \citep{pandey201517}, there are very few L1 speakers of IE.\\
 

### 2021-05-31 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
introduction edits

@@
 
-Research on the phonological systems of bi/multilingual speakers suggests that speakers maintain separate, but phonologically linked sound categories for both their languages.
-These show cross-language influence in both perception and production.
+Bi/multilingual speakers can distinguish between the phonetic norms of their languages \cite{caramazza1973acquisition,macleod2010impact,bosch2003simultaneous} and therefore maintain separate sound categories for each language.
+However, these categories are not autonomous, and show cross-language influence in both perception and production \cite{flege1995second,fowler2008cross,flege2002assessing}.
 %citation
@@
 Existing research on the phonetic effects of mixed-language production has largely focused on a few pairs of phonologically related languages, and used temporal properties of consonants (in particular, voice onset time or VOT) to measure transfer.
-However, great variation in results (cf sec.\ref{asymmetries} ) suggests that data from a wider variety of
+However, great variation in results (elaborated in sec.\ref{asymmetries} ) suggests that data from a wider variety of
 %elaborate on 'great variation'-- differences between language pairs, plus sociolinguistic situation (effects of language experience etc)-- clearer.
@@
 %line--rephrase
-behavior in these populations can add to the understanding of such cross-language interactions, as they are more likely to reflect real-world experience with mixed-language processing.
+behavior in these populations can be particularly valuable to understanding the nature of such cross-language interactions, as they are more likely to reflect real-world experience with mixed-language processing.
 
@@
 The results demonstrate a dynamic effect of transfer on L2 vowels, but no difference in the extent of transfer between the two paradigms.
-Findings are discussed in light of recent proposals about asymmetries in transfer effects and the role of connected speech in introducing discursive factors to transfer studies.
+Findings are discussed in light of recent proposals about asymmetries in transfer effects and the role of connected speech in introducing discursive factors to transfer studies.\\
 
-The next section provides background information on Bengali and English in India, focusing on vowel systems.
-The rest of the introduction  discusses causes, duration, and possible sources of asymmetries in transfer reported in previous studies, paradigms used, and the design of the present study in light of these.
+The rest of this section is structured as follows: \ref{bengali_english_in_india} provides background information on Bengali and Indian English focusing on vowel systems, and suggests on the basis of socio-linguistic facts that for the present population, short-term phonetic interaction during mixed-language use is an ecologically valid paradigm that is likely to be a part of everyday language-use experience.
+\ref{causes} discusses two potential sources of transfer during language processing-- global co-activation of multiple languages vs the local act of switching between languages.
+Given the highly multilingual setting that characterizes this population, we note that the former is expected to be a constant feature and therefore any observed phonetic interaction is better understood as resulting from the latter.
+The design of the present study is motivated on the basis of this discussion.
+A logical consequence of this model is that such phonetic interaction must be highly localized.
+\ref{duration} elaborates on this by reviewing literature on the duration and `reversiblity' of transfer effects.
+A recurring pattern that emerges from existing research on short-term phonetic transfer is that the phonetic influence does not affect both languages of the bilingual speaker equally, although the nature of these reported differences varies greatly across studies.
+In \ref{asymmetries}, we review some of these findings, focusing on proposed sources of asymmetries, in order to highlight the factors that seem to mediate such phonetic interaction.
+We argue that these results point to an urgent need for expanding the range of language pairs and sound categories examined in order to make meaningful generalizations about the nature of short-term phonetic transfer.
+Following from this, \ref{asymmetry between sounds} discusses reported asymmetries between different sound categories of a single language, and develops the hypotheses of the present study on the basis of these.
+In \ref{paradigms}, we motivate a secondary aim of this study: to verify the postulated differences between two paradigms for eliciting short-term phonetic interaction in an experimental setup, namely code-switching and cued picture-naming.
+\ref{questions_and_hypotheses} summarizes the research questions and hypotheses.
 
-\subsection{Bengali and English in India}
+\subsection{Bengali and English in India} \label{bengali_english_in_india}
 
@@
 Thus, regional variations are seen as the result of L1-influence on a common underlying target.
-%Early attempts to describe IE phonologically (General Indian English or GIE, \citep{masica1972sound}) were primarily attempts at standardization for pedagogical purposes. 
-Except for a small minority of specific communities \citep{pandey201517}, there are very few L1 speakers of IE.\\
+Excepting a small minority of specific communities \citep{pandey201517}, there are very few L1 speakers of IE.\\
 
@@
 
-\subsection{What causes transfer and what does it affect?}
+\subsection{What causes transfer and what does it affect?}\label{causes}
 
-Existing research has distinguished between changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf SLM \citep{flege1995second,flege2007language}; PAM-L2 \citep{best2007nonnative}), and interaction during \textsc{processing} (transfer vs interference \citep{grosjean2012attempt}; competence vs performance interference \citep{paradis1993linguistic}).
+Existing research has distinguished changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during processing (transfer vs interference \citep{grosjean2012attempt}; competence vs performance interference \citep{paradis1993linguistic}).
 Given their transient nature (cf sec.\ref{duration}), dynamic changes in production during mixed-language use are generally thought to involve the latter \citep{elias2017effects,simonet2014phonetic}.
@@
 Since the participants in this study are in an environment which largely contains mixed-language input, multilingual interlocutors, and no stable ambient language, we expect that phonetic transfer during daily language use, if any, takes place in a bilingual mode.
-Thus to preserve ecological validity, we elicited both unilingual and switched utterances in a bilingual language mode (cf section Methodology).
+Thus to preserve ecological validity, we elicited both unilingual and switched utterances in a bilingual language mode.
 Any observed differences in this paradigm would arguably result from interaction during online processing.
@@
 
-\subsection{Research questions and hypotheses}
+\subsection{Research questions and hypotheses}\label{questions_and_hypotheses}
 \begin{itemize}
@@
 
-\section{Methodology} 
+\section{Methodology} \label{methodology}
 The experiment had the following factorial design: 10 participants*20 target words*2 conditions *2 tasks *4 iterations = 3200 target items

### 2021-06-01 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
minor changes

--- 2021-05-29 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
@@
 
-Considering these facts, the present study focuses on short-term phonetic interaction during mixed-language use, because:
+Considering these facts, the present study focuses on short-term phonetic interaction during mixed-language use for the following reasons:
 \begin{enumerate}[label=(\roman*)]
@@
 \cite{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
-This calls to question the comparability of results between studies using these different paradigms.
+This calls to question the comparability of results across studies using these different paradigms.
 We tested this by comparing productions of the same items from the same group of participants in a scripted CS paradigm and a cued picture-naming paradigm.

### 2021-06-08 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
edits misc

--- 2021-05-31 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
@@
 Bi/multilingual speakers can distinguish between the phonetic norms of their languages \cite{caramazza1973acquisition,macleod2010impact,bosch2003simultaneous} and therefore maintain separate sound categories for each language.
-However, these categories are not autonomous, and show cross-language influence in both perception and production \cite{flege1995second,fowler2008cross,flege2002assessing}.
-%citation
-In production, this kind of influence (variously termed transfer, drift, accommodation, interference-- used interchangeably here to indicate any interaction between two sets of phonetic norms) has been studied in two conditions:
-(i) While a bilingual speaker is operating in any one of their languages.
-These studies often compare the speech of bilinguals to monolingual norms, and view transfer as changes to long-term memory representations as a result of acquiring an L2 \citep{guion2003vowel,caramazza1973acquisition,flege1987production}; (ii) When both languages of a bilingual speaker are co-activated.
-These compare productions during mixed-language processing to participants' own productions while using a single language.
-The resulting transfer is variously thought to involve short-term memory, online processing costs, language mode, and context-awareness.
+However, these categories are not autonomous-- they show cross-language influence in both perception and production \cite{flege1995second,fowler2008cross,flege2002assessing}.
+What causes, affects, and results from such influence?
+Answers to these questions provide important insights into how language `systems' are represented and processed in the mind.
+%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
+%show cross-language-- rephrase
+
+%keep just the types here, and put the next 2 lines in the causes&effects section?
+Cross-language influence at the level of sounds can be studied broadly in two kinds of conditions:
+(i) While a bilingual speaker is operating in any one of their languages, by comparing bilingual speech to monolingual norms \citep[e.g.][]{guion2003vowel,caramazza1973acquisition,flege1987production}; %These studies often compare the speech of bilinguals to monolingual norms, and view the influence as changes to long-term memory representations as a result of acquiring an L2 \citep{guion2003vowel,caramazza1973acquisition,flege1987production};
+(ii) When both languages of a bilingual speaker are co-activated, by comparing bilingual speakers' mixed-language speech to their own norms while using a single language \citep[e.g.][]{grosjean1994going, bullock2009trying,elias2017effects, simonet2014phonetic}.
+%These compare productions during mixed-language use to participants' own productions while using a single language.
+The resulting influence is variously thought to involve short-term memory, online processing costs, language mode, and context-awareness.
 The present study is concerned with the latter.
+In production, researchers have variously referred to this kind of influence as transfer, drift, accommodation, and interference-- these terms are used interchangeably here to indicate any interaction between two sets of phonetic norms.
 
-Existing research on the phonetic effects of mixed-language production has largely focused on a few pairs of phonologically related languages, and used temporal properties of consonants (in particular, voice onset time or VOT) to measure transfer.
-However, great variation in results (elaborated in sec.\ref{asymmetries} ) suggests that data from a wider variety of
+Existing research on the phonetic effects of mixed-language production has largely focused on a few pairs of phonologically related languages.
+A majority of these studies use temporal properties of consonants (in particular, voice onset time or VOT) to measure transfer.
+However, the reported results vary greatly across studies, and appear to be contingent upon both language-specific features and language experience of the participants.
+This suggests that data from a wider variety of
 %elaborate on 'great variation'-- differences between language pairs, plus sociolinguistic situation (effects of language experience etc)-- clearer.
 'Types of features'--separate line?- Brief mention - discursive factors in VOT.
-populations, language pairs,  and phonetic features is required in order to make meaningful generalizations.
+populations, language pairs,  and phonetic features needs to be considered in order to make meaningful generalizations.
 There is no work yet on short-term phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language.
@@
 %line--rephrase
-behavior in these populations can be particularly valuable to understanding the nature of such cross-language interactions, as they are more likely to reflect real-world experience with mixed-language processing.
+behavior in these populations can be particularly valuable towards understanding the nature of such cross-language interactions, as they are likely to reflect real-world experience with mixed-language processing.\\
 
@@
 Mixed-language data is elicited in two switching paradigms: code-switching and cued picture-naming.
-The results demonstrate a dynamic effect of transfer on L2 vowels, but no difference in the extent of transfer between the two paradigms.
-Findings are discussed in light of recent proposals about asymmetries in transfer effects and the role of connected speech in introducing discursive factors to transfer studies.\\
+We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction.
+The results demonstrate a dynamic effect of transfer on L2 vowels during mixed-language production, but no difference in the extent of transfer between the two paradigms.
+Findings are discussed in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.\\
 
-The rest of this section is structured as follows: \ref{bengali_english_in_india} provides background information on Bengali and Indian English focusing on vowel systems, and suggests on the basis of socio-linguistic facts that for the present population, short-term phonetic interaction during mixed-language use is an ecologically valid paradigm that is likely to be a part of everyday language-use experience.
+The rest of this section is structured as follows: \ref{bengali_english_in_india} provides background information on Bengali and Indian English focusing on vowel systems.
 \ref{causes} discusses two potential sources of transfer during language processing-- global co-activation of multiple languages vs the local act of switching between languages.
@@
 In India, it is spoken as an L2 by 129 million people (Census of India, 2011).
-English is one of the two official languages, used in education, law, media, as a \emph{lingua franca} mainly for an educated elite in most metropolitan regions, and carries a high prestige value \citep{tollefson2014language}.
+English is one of the two official languages, used in education, law, media, as a \emph{lingua franca} mainly for an educated elite in most metropolitan regions, and carries a high prestige value \citep{pandey201517, kachru1981english, tollefson2014language, kachru1983indianization}.
 %citations 
-In spite of the different L1s of its speakers, recent literature suggests that IE has a target phonology that is distinct from any of these, as well as from other native varieties of English (see \cite{sirsa2013effects} for a review).
-Thus, regional variations are seen as the result of L1-influence on a common underlying target.
-Excepting a small minority of specific communities \citep{pandey201517}, there are very few L1 speakers of IE.\\
+Recent research comparing the segmental and suprasegmental properties of IE spoken in different parts of the country report many commonalities (see \cite{sirsa2013effects} for a review).
+This suggests that in spite of the varied L1s of its speakers, IE has a target phonology that is distinct from any of these, as well as from other native varieties of English.
+Thus, it is fruitful to think of regional variations as resulting from L1-influence on a common underlying target.
+Excepting a small minority of specific communities \citep{pandey201517, wells1982accents, coelho1997anglo}, there are very few L1 speakers of IE.\\
 
-Considering these facts, the present study focuses on short-term phonetic interaction during mixed-language use, because:
+%%%%%%%%%%%%%%%%%%%%%%%do something%%%%%%%%%%%%%%%
+Considering the language usage patterns suggested by this demography, the present study focuses on short-term phonetic interaction during mixed-language use, because:
 \begin{enumerate}[label=(\roman*)]
@@
 \end{enumerate}
+
+The socio-linguistic facts about English in India also suggest that for the present population, mixed-language processing of English and an L1 is an ecologically valid paradigm that is likely to be a part of everyday language experience.
+Thus, data from this population is valuable to the understanding of short-term phonetic interaction as it occurs in .
 
@@
 %Their distribution in the F1XF2 space is shown in (insert table 1.7, Pandey).
-There is a single mid-central vowel corresponding to the categories \nt{2,@,3:}, which are treated as distinct in many native varieties of English.
+A single mid-central vowel corresponds to the categories \nt{2,@,3:}, which are treated as distinct in many native varieties of English \citep{nihalani1979indian,wells1982accents,hickey2005legacies,bansal1969intelligibility}.
 Since the English items used in this study are traditionally transcribed with \nt{2}, we use this symbol to indicate the mid-central vowel throughout.
-%In IE, the dipthong \nt{@U} is realized as \nt{o:}.
 
@@
 
-Existing research has distinguished changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during processing (transfer vs interference \citep{grosjean2012attempt}; competence vs performance interference \citep{paradis1993linguistic}).
-Given their transient nature (cf sec.\ref{duration}), dynamic changes in production during mixed-language use are generally thought to involve the latter \citep{elias2017effects,simonet2014phonetic}.
+Existing research has distinguished changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf.
+Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during accessing, processing, or articulation of these categories (cf.
+transfer vs interference \citep{grosjean2012attempt}; competence vs performance interference \citep{paradis1993linguistic}).
+Given their transient nature, dynamic changes in production during mixed-language use are generally attributed to the latter, e.g.
+online processing costs \citep{olson2013bilingual,tsui2019impact,vsimavckova2015immediate}, language mode \cite{simonet2014phonetic}, context-awareness \citep{khattab2013phonetic}.
 What triggers this interaction?
@@
 \cite{grosjean1998studying} suggests that various aspects of the communicative setting, including exposure to (spoken or written) stimuli in multiple languages and awareness of the interlocutor's being bilingual, could trigger a bilingual mode.
-Since the participants in this study are in an environment which largely contains mixed-language input, multilingual interlocutors, and no stable ambient language, we expect that phonetic transfer during daily language use, if any, takes place in a bilingual mode.
-Thus to preserve ecological validity, we elicited both unilingual and switched utterances in a bilingual language mode.
+The participants in the present study are in an environment which largely contains mixed-language input, multilingual interlocutors, and no stable ambient language.
+Therefore we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode.
+To preserve ecological validity, we elicited both unilingual and switched utterances in a bilingual language mode.
 Any observed differences in this paradigm would arguably result from interaction during online processing.

### 2021-06-08 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
Commented line No.106

--- 2021-06-01 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
@@
 
-Indian English (IE) refers to the variety of English that has developed in the Indian subcontinent.
+% Indian English (IE) refers to the variety of English that has developed in the Indian subcontinent.
 In India, it is spoken as an L2 by 129 million people (Census of India, 2011).
@@
 %citations 
-In spite of the different L1s of its speakers, recent literature suggests that IE has a target phonology that is distinct from any of these, as well as from other native varieties of English (see \cite{sirsa2013effects} for a review).
+% In spite of the different L1s of its speakers, recent literature suggests that IE has a target phonology that is distinct from any of these, as well as from other native varieties of English (see \cite{sirsa2013effects} for a review).
 Thus, regional variations are seen as the result of L1-influence on a common underlying target.

### 2021-06-08 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex

--- 2021-06-08 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
@@
 Considering the language usage patterns suggested by this demography, the present study focuses on short-term phonetic interaction during mixed-language use, because:
+
+Considering these facts, the present study focuses on short-term phonetic interaction during mixed-language use for the following reasons:
+
 \begin{enumerate}[label=(\roman*)]

### 2021-06-08 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
abstract

@@
 %\usepackage{gb4e}
+
 \begin{document}
@@
 
-Word count: 6944
+Word count: 7627
 
 \begin{abstract}
-Existing research demonstrates that mixed-language productions of bilingual speakers differ from productions in a single language.
-Studies have largely focused on temporal properties of consonants, and reported varied, and often contradictory, results.
-The present study examines the effect of mixed-language processing on two L2 vowel categories in a group of highly proficient Bengali-English bilingual speakers in India.
-Following from a discussion in \citep{olson2013bilingual}, we compare productions elicited in two paradigms-- cued picture-naming and code-switching.
-Results indicate a dynamic increase in cross-language phonetic transfer during mixed-language use leading to a shift towards L1 norms, asymmetrical shifts in the two categories, and no difference in transfer patterns between the two paradigms.
+This study investigates the phonetic outcome of mixed-language production in a group of highly proficient bilingual speakers of Bengali and English, to better understand the nature of cross-language phonetic interaction.
+Existing research has largely focused on a small set of phonologically related languages and used voice onset time (VOT) to measure transfer.
+Results show that transfer is mediated by both language-specific features and linguistic experience of participants, pointing to an urgent need for diversifying the range of languages and populations studied.
+We compare productions from a yet-unstudied population in two conditions—unilingual L2 (English) and mixed (English and Bengali).
+The spectral properties of two L2 vowel categories— \nt{\ae}(common to L1 and L2) and \nt{2} (absent in L1) are examined as sites for transfer.
+To understand how this unfolds dynamically, we measure F1 and F2 at five points over the course of articulating the vowel.
+We find a dynamic change in both vowels during mixed production, with \nt{2} shifting towards a related L1 category.
+We interpret this as a temporary increase in cross-language phonetic influence during mixed-language use, leading to a shift towards L1 norms.
+asymmetrical shifts in the two categories, and no difference in transfer patterns between the two paradigms.
 \end{abstract}
@@
 
-Bi/multilingual speakers can distinguish between the phonetic norms of their languages \cite{caramazza1973acquisition,macleod2010impact,bosch2003simultaneous} and therefore maintain separate sound categories for each language.
-However, these categories are not autonomous-- they show cross-language influence in both perception and production \cite{flege1995second,fowler2008cross,flege2002assessing}.
-What causes, affects, and results from such influence?
-Answers to these questions provide important insights into how language `systems' are represented and processed in the mind.
-%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
+Bi/multilingual speakers can distinguish between the phonetic norms of their languages and maintain separate sound categories for each language \citep{caramazza1973acquisition,macleod2010impact,bosch2003simultaneous}.
+However, these categories are not autonomous-- they influence each other across languages in both perception and production \citep[e.g.][]{flege1995second,fowler2008cross,flege2002assessing}, and the nature of such interaction provides crucial insights into how language `systems' are cognitively represented and processed.
+This study reports on the phonetic outcomes of mixed-language production in proficient bilingual speakers of Bengali and English.\\%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
 %show cross-language-- rephrase
@@
 (i) While a bilingual speaker is operating in any one of their languages, by comparing bilingual speech to monolingual norms \citep[e.g.][]{guion2003vowel,caramazza1973acquisition,flege1987production}; %These studies often compare the speech of bilinguals to monolingual norms, and view the influence as changes to long-term memory representations as a result of acquiring an L2 \citep{guion2003vowel,caramazza1973acquisition,flege1987production};
-(ii) When both languages of a bilingual speaker are co-activated, by comparing bilingual speakers' mixed-language speech to their own norms while using a single language \citep[e.g.][]{grosjean1994going, bullock2009trying,elias2017effects, simonet2014phonetic}.
+(ii) When both languages of a bilingual speaker are co-activated, by comparing speakers' mixed-language speech to their own norms while using a single language \citep[e.g.][]{grosjean1994going, bullock2009trying,elias2017effects, simonet2014phonetic}.
 %These compare productions during mixed-language use to participants' own productions while using a single language.
@@
 The present study is concerned with the latter.
-In production, researchers have variously referred to this kind of influence as transfer, drift, accommodation, and interference-- these terms are used interchangeably here to indicate any interaction between two sets of phonetic norms.
+In production, researchers have variously referred to this kind of influence as transfer, drift, accommodation, and interference-- these terms are used interchangeably here to indicate any interaction between two sets of phonetic norms.\\
 
@@
 
-Considering these facts, the present study focuses on short-term phonetic interaction during mixed-language use for the following reasons:
-
 \begin{enumerate}[label=(\roman*)]
@@
 
-\subsection{Duration: How long do effects last?} \label{duration}
+\subsection{Duration of effects} \label{duration}
 

### 2021-06-08 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
edits

@@
 This study investigates the phonetic outcome of mixed-language production in a group of highly proficient bilingual speakers of Bengali and English, to better understand the nature of cross-language phonetic interaction.
-Existing research has largely focused on a small set of phonologically related languages and used voice onset time (VOT) to measure transfer.
-Results show that transfer is mediated by both language-specific features and linguistic experience of participants, pointing to an urgent need for diversifying the range of languages and populations studied.
-We compare productions from a yet-unstudied population in two conditions—unilingual L2 (English) and mixed (English and Bengali).
+Existing research has largely focused on a small set of phonologically related languages and used voice onset time (VOT) to measure phonetic transfer.
+Results show that this interaction is mediated by both language-specific features and linguistic experience of participants, necessitating diversification in the range of languages and populations studied.
+We compare productions from a yet-unstudied population in two conditions-- unilingual L2 (English) and mixed (Bengali and English).
 The spectral properties of two L2 vowel categories— \nt{\ae}(common to L1 and L2) and \nt{2} (absent in L1) are examined as sites for transfer.
 To understand how this unfolds dynamically, we measure F1 and F2 at five points over the course of articulating the vowel.
-We find a dynamic change in both vowels during mixed production, with \nt{2} shifting towards a related L1 category.
+We find dynamic and  asymmetrical changes in both vowels during mixed production, with \nt{2} shifting towards a related L1 category.
 We interpret this as a temporary increase in cross-language phonetic influence during mixed-language use, leading to a shift towards L1 norms.
-asymmetrical shifts in the two categories, and no difference in transfer patterns between the two paradigms.
+Mixed language is elicited in two switching paradigms: cued picture-naming and code-switching, to assess if differences between the paradigms independently influence the outcome of phonetic interaction.
+Results reveal identical transfer patterns across the two paradigms.
+Findings are discussed in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.
 \end{abstract}
@@
 However, these categories are not autonomous-- they influence each other across languages in both perception and production \citep[e.g.][]{flege1995second,fowler2008cross,flege2002assessing}, and the nature of such interaction provides crucial insights into how language `systems' are cognitively represented and processed.
-This study reports on the phonetic outcomes of mixed-language production in proficient bilingual speakers of Bengali and English.\\%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
+Based on the productions of proficient bilingual speakers of Bengali and English, this study reports a temporary increase in cross-language phonetic interaction during mixed-language use.\\%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
 %show cross-language-- rephrase
@@
 The present study is concerned with the latter.
-In production, researchers have variously referred to this kind of influence as transfer, drift, accommodation, and interference-- these terms are used interchangeably here to indicate any interaction between two sets of phonetic norms.\\
+In production, this kind of influence has been variously termed \textit{transfer, drift, accommodation,} and \textit{interference}-- we use these interchangeably here to indicate any interaction between two sets of phonetic norms.\\
 
-Existing research on the phonetic effects of mixed-language production has largely focused on a few pairs of phonologically related languages.
+Existing research on the phonetic effects of mixed-language production has largely focused on a limited set of phonologically related languages.
 A majority of these studies use temporal properties of consonants (in particular, voice onset time or VOT) to measure transfer.
@@
 'Types of features'--separate line?- Brief mention - discursive factors in VOT.
-populations, language pairs,  and phonetic features needs to be considered in order to make meaningful generalizations.
-There is no work yet on short-term phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language.
-Nevertheless, widespread multilingualism in this part of the world suggests that phonetic
+populations, language pairs,  and phonetic features needs to be considered in order to make meaningful generalizations. Widespread multilingualism in the Indian subcontinent suggests that phonetic
 %line--rephrase
-behavior in these populations can be particularly valuable towards understanding the nature of such cross-language interactions, as they are likely to reflect real-world experience with mixed-language processing.\\
+behavior in these populations can be particularly valuable towards understanding the nature of such cross-language interactions, as they are likely to reflect real-world experience with mixed-language processing.
+However, there is no work yet on short-term phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language.
+\\
 
 The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India.
-We measure spectral properties (F1 and F2) of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2.
-Mixed-language data is elicited in two switching paradigms: code-switching and cued picture-naming.
+We measure spectral properties: F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2.
+Mixed-language data is elicited in two switching paradigms: cued picture-naming and code-switching.
 We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction.
@@
 \ref{duration} elaborates on this by reviewing literature on the duration and `reversiblity' of transfer effects.
-A recurring pattern that emerges from existing research on short-term phonetic transfer is that the phonetic influence does not affect both languages of the bilingual speaker equally, although the nature of these reported differences varies greatly across studies.
-In \ref{asymmetries}, we review some of these findings, focusing on proposed sources of asymmetries, in order to highlight the factors that seem to mediate such phonetic interaction.
+A recurring pattern that emerges from existing research on short-term phonetic transfer is that both languages of the bilingual speaker are not equally affected, although the nature of these reported differences varies greatly across studies.
+In \ref{asymmetries}, we review some of these findings, focusing on proposed sources of asymmetries, in order to highlight the factors that mediate such phonetic interaction (or fail to do so).
 We argue that these results point to an urgent need for expanding the range of language pairs and sound categories examined in order to make meaningful generalizations about the nature of short-term phonetic transfer.
 Following from this, \ref{asymmetry between sounds} discusses reported asymmetries between different sound categories of a single language, and develops the hypotheses of the present study on the basis of these.
-In \ref{paradigms}, we motivate a secondary aim of this study: to verify the postulated differences between two paradigms for eliciting short-term phonetic interaction in an experimental setup, namely code-switching and cued picture-naming.
+In \ref{paradigms}, we motivate a secondary aim of this study: to verify the postulated differences between two paradigms for eliciting short-term phonetic interaction in an experimental setup, namely cued picture-naming and code-switching.
 \ref{questions_and_hypotheses} summarizes the research questions and hypotheses.
@@
 
-	o-ta \quad Ek-ta \quad kalo [target word]\\
+	o-\textipa{\:t}a \quad \textipa{\ae}k-\textipa{\:t}a \quad kalo [target word]\\
 	that-CL one-CL black [target word]\\

### 2021-06-11 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
final manuscript

@@
 %for natbib: replace \cite with \citep (parenthetical); \citea with \cite (in-text)
-
+\usepackage{hyperref}
 \let\B\relax %to resolve a conflict in the definition of these commands between xyling and xunicode (the latter called by fontspec, called by charis)
@@
 % \pdf* commands provide metadata for the PDF output. ASCII characters only!
-\pdfauthor{Auromita Mitra, Indranil Dutta}
-\pdftitle{Mixed language processing increases cross-language phonetic transfer in Bengali-English Bilinguals}
+%\pdfauthor{Auromita Mitra, Indranil Dutta}
+%\pdftitle{Mixed language processing increases cross-language phonetic transfer in Bengali-English Bilinguals}
 \pdfkeywords{phonetic transfer, vowel formants, Bengali, English}
 
-\title{Mixed language processing increases cross-language phonetic transfer in Bengali-English Bilinguals}
+\title{Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals}
 % Optional short title inside square brackets, for the running headers.
@@
 
-Word count: 7627
+Word count: 9263
 
@@
 Existing research has largely focused on a small set of phonologically related languages and used voice onset time (VOT) to measure phonetic transfer.
-Results show that this interaction is mediated by both language-specific features and linguistic experience of participants, necessitating diversification in the range of languages and populations studied.
+Results indicate that such interaction is mediated by both language-specific features and linguistic experience, highlighting the need for diversification in the range of languages and populations studied.
 We compare productions from a yet-unstudied population in two conditions-- unilingual L2 (English) and mixed (Bengali and English).
-The spectral properties of two L2 vowel categories— \nt{\ae}(common to L1 and L2) and \nt{2} (absent in L1) are examined as sites for transfer.
+The spectral properties of two L2 vowel categories— \nt{\ae} (common to L1 and L2) and \nt{2} (absent in L1) are examined as sites for transfer.
 To understand how this unfolds dynamically, we measure F1 and F2 at five points over the course of articulating the vowel.
-We find dynamic and  asymmetrical changes in both vowels during mixed production, with \nt{2} shifting towards a related L1 category.
+We find dynamic and asymmetrical changes in both vowels during mixed production, with \nt{2} shifting towards a related L1 category \nt{a:}.
 We interpret this as a temporary increase in cross-language phonetic influence during mixed-language use, leading to a shift towards L1 norms.
 Mixed language is elicited in two switching paradigms: cued picture-naming and code-switching, to assess if differences between the paradigms independently influence the outcome of phonetic interaction.
-Results reveal identical transfer patterns across the two paradigms.
-Findings are discussed in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.
+Results reveal no difference in transfer patterns across the two paradigms.
+We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the postulated role of connected speech in introducing discursive factors to transfer studies.
 \end{abstract}
@@
 \begin{keywords}
-
+ phonetic transfer; vowel quality; bilingual speech production; L1-L2 interaction; Indian English; Bengali 
 \end{keywords}
@@
 Bi/multilingual speakers can distinguish between the phonetic norms of their languages and maintain separate sound categories for each language \citep{caramazza1973acquisition,macleod2010impact,bosch2003simultaneous}.
-However, these categories are not autonomous-- they influence each other across languages in both perception and production \citep[e.g.][]{flege1995second,fowler2008cross,flege2002assessing}, and the nature of such interaction provides crucial insights into how language `systems' are cognitively represented and processed.
-Based on the productions of proficient bilingual speakers of Bengali and English, this study reports a temporary increase in cross-language phonetic interaction during mixed-language use.\\%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
+However, these categories are not autonomous-- they influence each other across languages in both perception and production \citep[e.g.][]{flege1995second,fowler2008cross,flege2002assessing}, and the nature of such interaction provides crucial insights into how language ``systems" are cognitively represented and processed.
+Based on the productions of proficient bilingual speakers of Bengali and English, this study reports that cross-language phonetic influence temporarily increases during mixed-language use.\\%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
 %show cross-language-- rephrase
@@
 %keep just the types here, and put the next 2 lines in the causes&effects section?
-Cross-language influence at the level of sounds can be studied broadly in two kinds of conditions:
+Cross-language influence at the level of sounds can be studied in broadly two kinds of conditions:
 (i) While a bilingual speaker is operating in any one of their languages, by comparing bilingual speech to monolingual norms \citep[e.g.][]{guion2003vowel,caramazza1973acquisition,flege1987production}; %These studies often compare the speech of bilinguals to monolingual norms, and view the influence as changes to long-term memory representations as a result of acquiring an L2 \citep{guion2003vowel,caramazza1973acquisition,flege1987production};
@@
 However, the reported results vary greatly across studies, and appear to be contingent upon both language-specific features and language experience of the participants.
-This suggests that data from a wider variety of
+This emphasizes the importance of considering data from a wider variety of
 %elaborate on 'great variation'-- differences between language pairs, plus sociolinguistic situation (effects of language experience etc)-- clearer.
 'Types of features'--separate line?- Brief mention - discursive factors in VOT.
-populations, language pairs,  and phonetic features needs to be considered in order to make meaningful generalizations. Widespread multilingualism in the Indian subcontinent suggests that phonetic
+populations, language pairs,  and phonetic features in order to make meaningful generalizations. Widespread multilingualism in the Indian subcontinent suggests that phonetic
 %line--rephrase
@@
 The results demonstrate a dynamic effect of transfer on L2 vowels during mixed-language production, but no difference in the extent of transfer between the two paradigms.
-Findings are discussed in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.\\
+We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.\\
 
@@
 A logical consequence of this model is that such phonetic interaction must be highly localized.
-\ref{duration} elaborates on this by reviewing literature on the duration and `reversiblity' of transfer effects.
+\ref{duration} elaborates on this by reviewing literature on the duration and ``reversiblity" of transfer effects.
 A recurring pattern that emerges from existing research on short-term phonetic transfer is that both languages of the bilingual speaker are not equally affected, although the nature of these reported differences varies greatly across studies.
@@
 %%%%%%%%%%%%%%%%%%%%%%%do something%%%%%%%%%%%%%%%
-Considering the language usage patterns suggested by this demography, the present study focuses on short-term phonetic interaction during mixed-language use, because:
+Considering the language usage patterns suggested by this demography, the present study focuses specifically on short-term phonetic interaction during mixed-language use, because:
 
 \begin{enumerate}[label=(\roman*)]
-	\item Since the population is multilingual, we expect long-term representations to be affected by multiple languages. 
-	\item Given that examples of an IE phonology without L1 `influence' are very rare, it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a given `baseline' condition.
+	\item In multilingual populations, long-term representations of sound categories are expected to be affected by multiple languages. 
+	\item Given that examples of an IE phonology without L1 ``influence" are very rare, it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a given baseline condition.
 \end{enumerate}
@@
 The socio-linguistic facts about English in India also suggest that for the present population, mixed-language processing of English and an L1 is an ecologically valid paradigm that is likely to be a part of everyday language experience.
-Thus, data from this population is valuable to the understanding of short-term phonetic interaction as it occurs in .
+Thus, data from this population is valuable to the understanding of short-term phonetic interaction.
 
@@
 
-Existing research has distinguished changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf.
+Existing research has distinguished changes in category \textit{representations} due to the acquisition of multiple sound systems (cf.
 Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during accessing, processing, or articulation of these categories (cf.
@@
 Given their transient nature, dynamic changes in production during mixed-language use are generally attributed to the latter, e.g.
-online processing costs \citep{olson2013bilingual,tsui2019impact,vsimavckova2015immediate}, language mode \cite{simonet2014phonetic}, context-awareness \citep{khattab2013phonetic}.
+online processing costs \citep{olson2013bilingual,tsui2019impact,vsimavckova2015immediate}, language mode \citep{simonet2014phonetic}, context-awareness \citep{khattab2013phonetic}.
 What triggers this interaction?
@@
 A number of studies have specifically manipulated language mode, both in the presence and absence of switching.
-Overall, results suggest that (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use -- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \citep{olson2016role,tsui2019impact} or spontaneous conversation \citep{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
+Overall, results suggest that: (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use -- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \citep{olson2016role,tsui2019impact} or spontaneous conversation \citep{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
 (iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \cite{olson2016role} found no additive effects, \cite{olson2013bilingual} found the balanced language context to inhibit transfer compared to unbalanced contexts.
-Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days \cite{Pschwartz2015language, bullock2009trying,antoniou2011inter, elias2017effects,vsimavckova2015immediate,vsimavckova2018patterns}.
+Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days \citep{schwartz2015language, bullock2009trying,antoniou2011inter, elias2017effects,vsimavckova2015immediate,vsimavckova2018patterns}.\\
 
@@
 Any observed differences in this paradigm would arguably result from interaction during online processing.
+\\
 
-In a bilingual mode, both language systems are expected to be (nearly) equally accessible throughout the test block.
-Thus, a consistent difference between switched and non-switched tokens in such a paradigm would be possible only if the effects were highly localized (if not, we should expect a gradual convergence over the course of the experiment).
+In a bilingual mode, both language systems are expected to be nearly equally accessible throughout the test block.
+Thus, a consistent difference between switched and non-switched tokens in such a paradigm would be possible only if the effects were highly localized-- if not, we should expect a gradual convergence over the course of the experiment.
 This is discussed in the next section.
 
-\subsection{Duration of effects} \label{duration}
+\subsection{Duration of transfer effects} \label{duration}
 
-Studies which compare bilinguals speakers' phonological systems with monolinguals see them as relatively stable over time.
-The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods, \cite{guion2003vowel, caramazza1973acquisition}.
+Research comparing bilinguals speakers' phonological systems with monolinguals treats them as relatively stable over time.
+The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods \citep[e.g.][]{guion2003vowel, caramazza1973acquisition}.
 However, longitudinal studies %(citations)
-and between-subject comparisons of bilingual speakers who differ in the duration of L2 exposure suggest that interaction between the sound systems is dynamic.
-Over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts-- e.g.
-\cite{bohn1992production}
-found that an approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language transfer.
+and between-subject comparisons of bilingual speakers with different durations of L2 exposure suggest that interaction between the sound systems is dynamic: over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts.
+For example, \cite{bohn1992production}
+observed that approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language transfer.
 These have been interpreted as changes to long-term phonological representations.
@@
 \cite{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language.
-This not only showed that cross-language interaction can be triggered in an order of months, but also that the effects can be reversed within a similar time range.
-A recent study by \cite{tobin2017phonetic} found comparable effects in an even shorter duration (2-4 weeks).
-A short-term longitudinal study by \cite{chang2012rapid} showed that over the course of the first five weeks of learning an L2, there was a gradual convergence of L1 towards L2.
-For some sounds and features (though not others, cf sec.\ref{asymmetry between sounds}), transfer was additive over time.
+This not only evidenced that cross-language interaction can be triggered in an order of months, but also that the effects can be reversed within a similar time range.
+A more recent study by \cite{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks).
+Based on a short-term longitudinal study, \cite{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2.
+For some sounds and features (though not others, cf.
+sec.\ref{asymmetry between sounds}), transfer was additive over time.
 How long these effects last in the absence of regular L2 input was not tested.
+\\
 % if effects are cumulative, then the difference between switched and non-switched tokes should be lesser in later test blocks (previous two paragraphs needed?)
 
-Studies examining mixed-language use within an utterance mostly analyze only the switched (target) token.
-Thus, very few direct measurements of the duration of transfer effects are available.
-One study which measured this \citep{bullock2009trying} did not find any residual effects on the matrix language following a switch, suggesting that transfer during code-switching is localized.
-
-More indirect evidence for this comes from studies which have elicited switched and non-switched tokens in the same test block \citep{tsui2019impact,olson2013bilingual}, and still reported differences between the two, suggesting that changes due to transfer are quickly `reset'-- in an order of seconds.
-Note that all these studies measure VOT, which is a temporal feature.
+The majority of studies on the phonetics of mixed-language use only analyze the switched (target) token.
+Thus, there are few direct measurements of the duration of short-term transfer effects.
+One study which measured this in a code-switching paradigm \citep{bullock2009trying} did not find any residual effects on the matrix language following a switch, suggesting that transfer during code-switching is localized.
+Indirect evidence in support of this comes from experiments which have elicited switched and non-switched tokens in the same test block and still reported differences between the two \citep[e.g.][]{tsui2019impact,olson2013bilingual}, suggesting that changes due to transfer are quickly ``reset"-- in an order of seconds.\\
+Note that all the studies discussed above measure VOT, which is a temporal feature.
 We have no apriori reason to assume that these durations generalize to vowel quality.
-However, findings from sub-categorical phonetic shifts triggered by other factors (such as convergence towards an interlocutor) do evidence rapid shifts in vowel quality within comparable time-frames \citep{pardo2010expressing,babel2010dialect,babel2012evidence}.
+However, findings from sub-categorical phonetic shifts triggered by other factors (such as convergence towards an interlocutor) do evidence rapid shifts in vowel quality within comparable time-frames \citep[e.g.][]{pardo2010expressing,babel2010dialect,babel2012evidence}.
 Thus in the current study, we present switched and non-switched tokens randomly within the same test block to induce a bilingual mode.
@@
 \subsection{Asymmetries between languages in extent and direction of transfer} \label{asymmetries}
-Previous research has largely reported asymmetrical shifts between languages. Two proposed sources of such differences that are relevant to this study are discussed below:
+A recurring pattern that emerges from existing research on short-term phonetic transfer is that the extent and patterns of phonetic shift are not equivalent across the two languages of the bilingual speaker.
+Studies vary greatly in the nature of differences that they report.
+The potential causes for such asymmetry are of interest because they point towards the factors that mediate cross-language phonetic interaction.
+In this section, we discuss two proposed sources of such asymmetry that are relevant to the present study:
 
 \subsubsection{L1 vs L2 status} 
-Flege's Speech Learning Model (SLM) \citeyear{flege1995second,flege2007language} posits that the sound categories of a bilingual speaker exist in a common phonological space, and thus in principle, both L1 and L2 sounds can affect one another.
-Thus, the patterns of cross-language transfer depend on the mapping of L2 phonemes in relation to existing L1 categories.
-In non-switched production, this is evidenced through a pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (equivalence classification-- \cite{flege1984limits,flege1987production}), and
-the observation that both the L1 and L2 sound systems of bilinguals differ from corresponding monolingual speakers \citep{guion2003vowel}.
+Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} posits that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another.
+Thus, the patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories.
+In non-switched production, this is evidenced through a pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf.
+\textit{equivalence classification}-- \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][]{guion2003vowel}.
+\\
 
 In mixed-language production, the role of language status is less clear-- studies have reported unidirectional influence of L1 on L2 \citep{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \citep{tsui2019impact,elias2017effects, olson2013bilingual}, bidirectional convergence \citep{bullock2009trying, olson2016role}, divergence \citep{bullock2009trying,vsimavckova2018patterns}, and no influence \citep{muldner2019phonetics,schwartz2015language}.
-Since most existing studies measure shifts in VOT, the observed asymmetries between languages have been variously explained either in terms of their L1 vs L2 status, or language-specific differences between long- and short-lag VOT languages (cf sec.
+Since most existing studies measure shifts in VOT, the observed asymmetries between languages have been variously explained either in terms of their L1 vs L2 status, or language-specific differences between long- and short-lag VOT languages.
+%(cf sec.
 \ref{sound systems}).
@@
 \cite{olson2013bilingual} first reported a unidirectional VOT shift of L1 towards L2 in two different groups -- native speakers of Spanish (short-lag) and English (long-lag), matched for proficiency and age of L2 acquisition.
-This suggested that beyond language-specific differences, the L1 vs L2 status of the language does mediate transfer.
-They explain the asymmetry in terms of the Inhibitory Control Model (ICM) \citep{green1998mental} -- to select a phonetic realization from one language, the other must be inhibited.
-Being the `stronger' language, more inhibition is required on the L1.
-Thus, switching into L1 incurs a greater switch-cost (more cross-language influence) than switching into L2.
-The lesser switch cost results in a lack of visible cross-language transfer effects on L2.
-\cite{tsui2019impact} found comparable results, but only in participants who were not equally dominant in both languages.
+This established that beyond language-specific differences, the L1 vs L2 status of the language does mediate transfer.
+They interpret this asymmetry in terms of the Inhibitory Control Model (ICM) \citep{green1998mental}: to select a phonetic realization from one language, the other must be inhibited.
+Since L1 is the ``stronger" language, the amount of inhibition required on L1 while using an L2 is greater than the inhibition required on an L2 while using the L1.
+This greater initial inhibition on L1 means that switching into L1 incurs a greater ``switch cost" (more cross-language influence) than switching into L2.
+Lower switch cost results in a lack of visible cross-language transfer effects on L2.
+\cite{tsui2019impact} report comparable results, but only in participants who were not equally dominant in both languages.
 Balanced bilinguals did not demonstrate any transfer effects in VOT.
-They propose that this is because balanced bilinguals have better inhibitory control (low switch costs in both languages) due to greater experience with language switching.
+They propose that this is because balanced bilinguals have better inhibitory control (low switch cost in both languages) due to greater experience with language switching.\\
 
-Since VOT is independently affected by short- vs long-lag differences, studies of other sound contrasts/features can avoid this conflation, and give a clearer picture of the effect of language status.
-However, given the implications of the ICM, we would first have to establish that these L2 categories can indeed be affected by dynamic interference.
-The few existing studies which focus on vowel quality \citep{simonet2014phonetic,muldner2019phonetics,elias2017effects} and phonological processes \citep{simonet2020increased,schwartz2015language} have all examined transfer effects on L1.
-As far as we are aware, there has been no work yet on vowels in L2.
+Since differences between long-lag and short-lag languages independently affect the phonetic realization of VOT, focusing on other phonetic features as sites for transfer can avoid this conflation, and clarify the precise effect of language status.
+However, given the implications of the ICM, it is necessary to first establish that these L2 categories can indeed be affected by dynamic interference, particularly in proficient bilinguals.
+The few existing studies on vowel quality \citep{simonet2014phonetic,muldner2019phonetics,elias2017effects} or phonological processes \citep{simonet2020increased,schwartz2015language} have examined transfer effects on L1 categories.
+As far as we are aware, there is no work yet on vowels in L2.
 Thus, this study builds on the existing work by examining whether the L2 vowel quality of proficient bilinguals can be affected by mixed-language processing.
 
-\subsubsection{Differences between sound systems of the languages} \label{sound systems} In addition to the cognitive factors discussed above, many studies have attributed the asymmetries to language-internal factors.
-Specifically, a pattern that emerges from the body of work on VOT is that a shift in long-lag VOT towards the short-lag norms of the other language is much more consistent and systematic than the converse \cite{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid}.
+\subsubsection{Differences between sound systems of the languages} \label{sound systems} In addition to the cognitive factors discussed above, many researchers have attributed the observed asymmetries to language-internal factors.
+Specifically, a pattern that emerges from the body of work on VOT is that a shift in long-lag VOT towards the short-lag norms of the other language is much more consistent and systematic than the converse \citep{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid}.
 In general, short-lag VOTs seem to resist accommodative shift.
-\cite{bullock2009trying} suggest that because long-lag languages offer a greater range of acceptable VOTs, there is more `room' for movement.
+\cite{bullock2009trying} suggest that because long-lag languages offer a greater range of acceptable VOTs, there is more ``room" for movement.
 In contrast, shift in short-lag languages could risk the loss of a phonological contrast (VOT being the primary cue for voicing contrast in these languages).
 This is a language-specific phonological constraint on transfer.
-In a similar vein, \cite{antoniou2011inter} found that while English stops shifted towards Greek (a short-lag language), the reverse was seen only in a subset of sounds-- word-medial non-nasalized stops.
-They attribute this to the low frequency of these sounds in Greek, which makes them less `stable', and thus more amenable to shift.
-This suggests that the distribution of sounds in a language could affect the kind of transfer patterns observed.
+In a similar vein, \cite{antoniou2011inter} observe that while the VOT of English stops shifted towards Greek (a short-lag language), the reverse was seen only in a specific subset of sounds-- word-medial non-nasalized stops.
+They attribute this to the low frequency of these sounds in Greek, making them less ``stable", and thus more amenable to cross-language influence.
+This suggests that the distribution of sounds in a language could affect the kind of transfer patterns observed.\\
 
@@
 Therefore, to make meaningful generalizations, it is imperative to consider data from a variety of language pairs.
-Existing research on phonetic transfer largely revolves around a small set of phonologically related languages.
-Thus, the present study extends the scope of this research to a new set of languages---Indian English and Bengali (an Indo-Aryan language).
+Existing research on phonetic transfer largely revolves around a limited set of phonologically related languages.
+Thus, the present study extends the scope of this research to a new pair of languages--- Indian English and Bengali.
 
-Since Bengali has a four-way laryngeal contrast, VOT does not encode a binary voiced---voiceless distinction, and moreover is not the only cue for the corresponding phonological categories \citep{dmitrieva2020acoustic}.
-Thus, the effects of Bengali on English VOT are likely to be affected by additional phonological factors.
-Thus, avoiding any potential confounds was another reason we chose to focus on vowels as a target for transfer.
+Bengali has a four-way laryngeal contrast.
+Thus, VOT does not encode a binary voiced-voiceless distinction, and moreover is not the only relevant cue for the corresponding phonological categories \citep{dmitrieva2020acoustic}.
+This means that any effect of L1 Bengali on L2 English VOT is likely to be mediated by additional phonological factors.
+Thus, avoiding any potential confounds was another reason we chose to focus on vowel quality as a target for transfer.
 
@@
 
-Existing research reports asymmetrical transfer effects not only between languages, but also between different sounds/features in a language, in both long-term and transient interactions.
-Studies which have examined multiple sound categories have found that interactions between individual sounds pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot be seen as an atomic measure.
-In light of the discussion in section \ref{sound systems}, this is not surprising -- if a general linguistic principle of `room for movement' constrains transfer, then we should expect it to apply to individual sounds too.
-Once again, this emphasizes the need to study a wider range of sound contrasts.
+Asymmetries in the extent and patterns of transfer have not only been observed between languages, but also between different sounds/features of a language, in both long-term and transient interactions.
+Studies which have examined multiple sound categories have found that interactions between individual sounds pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep[e.g.][]{chang2012rapid,elias2017effects}, suggesting that \textit{extent of transfer} cannot treated as an atomic measure.
+In light of the discussion about VOT in section \ref{sound systems}, this is not surprising --- if a general linguistic principle of ``room for movement" constrains transfer, then we should expect it to apply to individual sounds too.
+Once again, this emphasizes the importance of examining a wider range of sound contrasts.\\
 
-In the present study, we examine two vowel categories in Indian English-- the mid-central vowel \nt{2} and the low front vowel \nt{\ae}. There are two plausible sources of asymmetry between these:
+In the present study, we focus on two vowel categories in Indian English-- the mid-central vowel \nt{2} and the low front vowel \nt{\ae}. There are two plausible sources of asymmetry between these:
 \begin{enumerate}
-	\item Position in the vowel space of IE: \nt{2} is in a less crowded part of the vowel space, and thus has more latitude for movement without risking contrast, particularly in the height (F1) dimension.
-Thus, considering purely phonological constraints on IE, we would expect that a greater degree of shift is possible in \nt{2} compared to \nt{\ae}.
-Any movement in \nt{\ae} is expected to be primarily in the front-back (F2) dimension.
-	\item Target of transfer: Since we do not elicit unilingual L1 utterances in this study, we cannot directly compare the target vowels to baseline L1 vowels.
-However, the body of existing research on phonetic transfer suggests that shifts are targeted with respect to categories in the other language, rather than random.
-Thus, another source of asymmetry is the fact that \nt{2} is not present as a phoneme in Bengali, whereas \nt{\ae} is a common category across both languages (cf sec.\ref{vowel systems}).
-\end{enumerate}
+	\item Position in the IE vowel space: compared to \nt{\ae}, \nt{2} exists in a part of the vowel space that has a lower vowel density.
+This affords a greater latitude for movement without risking the loss of a contrast, particularly in the vowel height (F1) dimension.
+Thus, considering purely phonological constraints on IE, we expect that a greater degree of shift is possible in \nt{2} compared to \nt{\ae}.
+Any movement in \nt{\ae} is expected to be primarily in the backness (F2) dimension.
+	\item Target of transfer: findings from existing research on cross-language phonetic interaction suggest that changes in production during mixed-language use are not random, but rather targeted with respect to categories in the other language.
+Thus, another source of asymmetry is the fact that the category \nt{2} is absent in Bengali, whereas \nt{\ae} is a common category across both languages (cf.
+sec.\ref{vowel systems}).
+\end{enumerate}\\
 
@@
 
-Flege's Speech Learning Model (SLM) \citep{flege1995second,flege2007language} suggests that sound categories which are common across languages influence each other because they share a common acoustic-phonetic space.
+Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} posits that sound categories which are common across languages influence each other because they share a common acoustic-phonetic space.
 Thus, if Bengali and English differ in their canonical realizations of \nt{\ae}, then we should expect the English \nt{\ae} to shift towards the corresponding Bengali category in the mixed condition.
@@
 However, it is unlikely that this should altogether preclude a shift in \nt{2}, since at least one existing study has reported transfer effects on a non-common vowel category in a comparable paradigm \citep{simonet2014phonetic}.
-Here, the Catalan \nt{O} moved towards an acoustically close category \nt{o}, which is common to Catalan and Spanish, suggesting that shift might be towards a set of phonetic norms rather than a particular target.
+Here, participants' Catalan \nt{O} shifted towards an acoustically close category \nt{o}, which is common to Catalan and Spanish, suggesting that movement might be towards a set of phonetic norms rather than a specific corresponding category.
 
-The vowel \nt{2} is thus expected to shift towards an acoustically close category in IE -- a probable candidate being the low vowel \nt{a:}, as anecdotal observations of heavily-accented English suggest that the sounds are perceived as close by Bengali speakers.
+We expect the vowel \nt{2} to shift towards an acoustically/perceptually close category in IE -- a probable candidate being the low vowel \nt{a:}, as anecdotal observations of Bengali-accented English suggest that the sounds are perceived as close by Bengali speakers.
+\\
 
-As noted above, we did not elicit unilingual Bengali productions from the participants in this study.
-To estimate a baseline value for the relevant Bengali vowels, productions in the unilingual English condition were compared to data from five female Bengali speakers from \cite{dutta2016using}.
+In the present study, we do not elicit unilingual Bengali productions.
+To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we used Bengali production data from five female speakers from \cite{dutta2021}.
 Formant values were extracted from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel.
-Comparisons of Lobanov-normalized F1 and F2 suggests that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, compared to the English categories \nt{\ae} and \nt{2} respectively.
+Next, we compared these to our participants' productions in the unilingual English condition.
+Comparisons of Lobanov-normalized F1 and F2 suggest that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.
 
@@
 
-Studies of phonetic transfer during mixed-language use have largely used spontaneous or scripted code-switching (CS, defined as the use of multiple languages within a single utterance, ie in connected speech; \citealp{myers1993dueling}) to co-activate the two languages.
-Other paradigms include cued picture-naming, delayed repetition, interpreting across languages, reading word-lists etc.
-\cite{olson2013bilingual} argues that code-switching introduces discursive factors such as discourse planning and pragmatic considerations, which ultimately show patterns of language practice, rather than cognitive behavior.
-This finds support from studies which report transfer effects before the switch point \citep{bullock2009trying}, or other evidence of planning such as hyperarticulation \citep{muldner2019phonetics} and interlocutor-awareness in spontaneous CS \citep{khattab2013phonetic}.
-To avoid such confounds, Olson suggests cued picture-naming (where pictures are named in rapid succession in either language, based on the given language cue) as an alternative paradigm to isolate purely phonetic effects in controlled setups.
+This section motivates a secondary aim of this study: to verify if the extent and patterns of phonetic interaction in bilingual participants are independently mediated by the choice of experimental paradigm.
+Studies of phonetic transfer during mixed-language use have largely employed spontaneous or scripted code-switching (CS) to elicit mixed productions.
+CS is defined as the use of multiple languages within a single utterance, i.e.
+in connected speech \citep{myers1993dueling}.
+Other paradigms which have been used include cued picture-naming, delayed repetition, interpreting across languages, reading word-lists etc.
+\cite{olson2013bilingual} argues that code-switching introduces discursive factors such as discourse planning and pragmatic considerations, which ultimately reflect patterns of language practice, rather than cognitive behavior.
+This finds support from studies which report transfer effects \textit{before} the switch point \citep{bullock2009trying}, or other evidence for planning such as hyperarticulation \citep{muldner2019phonetics} and interlocutor-awareness in spontaneous CS \citep{khattab2013phonetic}.
+To avoid such confounds, \cite{olson2013bilingual} endorses cued picture-naming (where pictures are named in rapid succession in either language, based on the given language cue) as an alternative paradigm to isolate purely phonetic effects in controlled setups.
+This observation has important implications for how the existing literature on short-term phonetic interaction is interpreted: if the discursive properties of a CS paradigm independently affect the phonetic outcome of mixed-language use, this calls to question the comparability of results across studies using these different paradigms.
+Identifying paradigm-specific differences is thus particularly relevant for this line of research, as generalizations often rely crucially on the assumption of such comparability.
+\\
 
-Note that while there is evidence for planning in CS, exactly how this affects phonetic transfer cannot be predicted from the existing literature-- while \citep{bullock2009trying} found tokens immediately preceding the switch point to be phonetically distinct from the rest, one group showed convergence towards the switch language(explained in terms of proficiency) while the other showed divergence (explained in terms of extra-linguistic factors).
-The participants in Muldner et al.'s \citeyear{muldner2019phonetics} study hyperarticulated switched tokens (suggesting some degree of planning), and showed no significant shift.
-Moreover, recall that Olson's \citeyear{olson2016role} findings suggest that there is a limit on transfer-- multiple factors do not necessarily lead to additive effects.
-Thus, the existence of additional factors in CS does not automatically mean that this should affect the extent of transfer compared to any other paradigm.
-It is possible that the effects in the studies mentioned above could be attributed to some other aspect of the experimental setup, rather than a result of code-switching in particular.
+Note that while there is evidence for planning in CS, precisely how this affects phonetic interaction cannot be predicted from the existing literature--- while \cite{bullock2009trying} found tokens immediately preceding the switch point to be phonetically distinct from the rest, one participant group converged towards the switch language (explained in terms of proficiency) while the other showed divergence (explained in terms of extra-linguistic factors).
+The participants in Muldner et al.'s \citeyearpar{muldner2019phonetics} study hyperarticulated switched tokens (indicating some degree of planning), and did not evidence a significant shift during mixed-language use.
+Moreover, recall that the findings of \cite{olson2016role} suggest that there is a limit on the extent of transfer-- multiple factors do not necessarily result in additive effects.
+Thus, the presence of additional factors in CS does not necessitate that this should alter the extent of observable transfer compared to any other paradigm.
+It is possible that the outcomes in the studies discussed above are better attributed to some other aspect of the experimental setup, rather than to the act of code-switching in particular.\\
 
-While much more work is necessary to tease out the various factors and their interactions, the focus of the present study is to check whether, in an experimental setting, the postulated discursive elements in a CS paradigm affect the extent of observed transfer, compared to a paradigm which arguably lacks them--cued picture-naming.
-We are less concerned with the exact nature of differences, than with their comparability.
-While studies using each of these paradigms to induce transfer exist, comparison across studies is difficult, since they also differ in other factors.
-This study thus provides a starting point for such comparison by eliciting productions of the same tokens from the same participants, in both paradigms.
-A difference in the extent of transfer will suggest that the planning effects in connected speech affect transfer independently of the phonetic results of switching, which has implications that future studies should look into.
+The precise contribution of various (linguistic and extra-linguistic) factors and their interactions to the phonetics mixed-language use merits detailed investigation.
+In the present study, our focus is limited to verifying whether, in an experimental setting, the postulated discursive elements of a CS paradigm alter the extent of observable transfer compared to a paradigm that arguably lacks these--- cued picture-naming.
+We are less concerned with the exact nature of differences than with their comparability.
+While researchers have successfully employed each of these paradigms to induce phonetic interaction, comparison across studies is difficult since they also differ in other factors.
+We provide a starting point for such comparison by eliciting productions of the same tokens from the same set of participants, in both paradigms.
+A difference in results will suggest that the planning involved in connected speech affects transfer above and beyond the phonetic outcome of switching itself, which has implications that future work should consider.
 
@@
 \begin{itemize}
-	\item Are L2 vowel categories of proficient bilingual speakers affected by dynamic phonetic transfer during mixed-language production?
-	\item If so, we expect an asymmetry between the vowels in the extent and direction of shift.
-Based on comparisons with Bengali vowels from \cite{dutta2016using} we expect \nt{2} to be lowered and fronted in the mixed condition.
+	\item \textbf{L2 vowels}: Are L2 vowel categories of proficient bilingual speakers affected by dynamic phonetic transfer during mixed-language production, as evidenced by differences in the spectral properties of L2 vowels in switched and non-switched tokens?
+	\item \textbf{Asymmetry}: If so, we expect an asymmetry between the vowels in the extent and direction of shift.
+Based on comparisons with Bengali vowels from \cite{dutta2021} we expect \nt{2} to be lowered and fronted in the mixed condition.
 \nt{\ae} is expected to show lowering, but to a lesser extent than \nt{2}, as it is in a part of the IE vowel space that has greater vowel density.
-	\item Is there an observable difference in the extent of transfer between the cued picture-naming and the code-switching paradigms? 
+	\item \textbf{Paradigm}: Is there an observable difference between the extent of transfer in a cued picture-naming vs a code-switching paradigm? 
 \end{itemize}
@@
 \section{Methodology} \label{methodology}
-The experiment had the following factorial design: 10 participants*20 target words*2 conditions *2 tasks *4 iterations = 3200 target items
+The experiment had the following factorial design: 10 participants* 20 target words* 2 conditions* 2 tasks* 4 iterations = 3200 target tokens.
 
@@
 10 monosyllabic English words that do not contain the target vowel served as filler items.
-English words which are lexicalized loanwords in Bengali were avoided.
+We avoided English words that are lexicalized loanwords in Bengali.
 %The list of words can be found in sec.\ref{appendix}.
@@
 Utterances in the two conditions (unilingual and mixed) were randomly interspersed within each iteration.
-In the unilingual condition, the target word was preceded by another English word in the utterance, whereas in the mixed condition, the target was produced as a switch from Bengali.
+In the unilingual condition, the target word was preceded by another English word in the utterance.
+In the mixed condition, the target was preceded by a Bengali word and thus produced as a switch from Bengali.
 Every target word appeared twice during a test block; once in a unilingual and once in a mixed condition.
-Participants alternated between the two tasks, and were allowed to take a short break after each test block.
+Participants alternated between the two tasks, and were allowed to take a short break after each test block.\\
 
@@
 	\end{itemize}
-	Since all the target words were in English, the language cue was used to manipulate the preceding word, giving target words in non-switched (unilingual) and switched (mixed) utterances.
-The orthographies of English and Bengali are visually distinct, and were thus used to cue language.
+	Since all the target words were in English, the language cue was used to manipulate the \textit{preceding} word, giving target words in non-switched (unilingual) and switched (mixed) utterances.
+The orthographies of English and Bengali are visually distinct, and were therefore used to cue language.
 
@@
 	\begin{itemize}
-		\item Unilingual sentence: That is a yellow [Target Word].
-		\item Mixed-language sentence: \textipa{o-\:ta \ae k-\:ta kalo} [Target Word]. 
+		\item Unilingual sentence: That is a yellow [\textit{target word}].
+		\item Mixed-language sentence: \textipa{o-\:ta \ae k-\:ta kalo} [\textit{target word}]. 
 	\end{itemize}
@@
 
-	o-\textipa{\:t}a \quad \textipa{\ae}k-\textipa{\:t}a \quad kalo [target word]\\
-	that-CL one-CL black [target word]\\
-	that is a black [target word]\\
+	o-\textipa{\:t}a \quad \textipa{\ae}k-\textipa{\:t}a \quad kalo [\textit{target word}]\\
+	that-\textsc{clf}  one-\textsc{clf}  black [\textit{target word}]\\
+	`that is a black [\textit{target word}]'\\
 
-	The sound preceding the target word was uniform across all sentences (the final sound in `yellow' is produced as [o] in Indian English), to minimize any possible differences due to coarticulatory effects on the target word.
-A specifier-noun construction was used to ensure identical word order in both languages, giving uniform prosodic context for the target word.
-Mixed orthography was used for the mixed-language utterances.
+	The sound preceding the target word was uniform across all sentences (the final sound in \textit{yellow} is produced as [o] in Indian English), to minimize any possible differences due to coarticulatory effects on the target vowel.
+A specifier-noun construction was used to ensure identical word order in both languages, giving a uniform prosodic context for the target word across condition.
+In presenting the mixed-language utterances, we used mixed orthography.
 \\
@@
 \subsection{Acoustic analysis}\label{analysis}
-The first iteration of each task was treated as a trial, and excluded from analysis.
-An additional 67 tokens (2.79\%) were excluded due to errors in presentation/production, giving 2333 target word items for analysis.
+The first iteration of each task was treated as a trial, and excluded from further analyses.
+An additional 67 tokens (2.79\%) were excluded due to errors in presentation/production, giving 2333 target word tokens for analysis.
 Recordings were manually segmented and annotated in Praat \citep{boersma2016praat}, and the first and second formant frequencies (F1 and F2) of the target vowels \nt{2, ae} were measured at five points-- in 10\% increments starting at 5\% into the vowel.
@@
 
-\begin{table}
+\begin{table}[h]
 	\caption{Fixed factors}
@@
 
-Since we wanted to model height and backness at different points in the vowel using a single model, allowing us to track if/how transfer changes over time, we arbitrarily chose a point in the vowel (F1 at 25\%; henceforth $F1\_25$) to identify an optimal model, and then fitted this to all other points.
+Our target in this analysis was to model vowel quality at different points in the vowel using the same mixed-effects model, allowing us to track if/how transfer changes over time.
+To do this, we arbitrarily chose a point in the vowel: F1 at 25\% (henceforth F1$\_$25) to identify an optimal model, and then fitted this to all other points.
 
@@
 Next, adding Context as a fixed effect significantly improved the model fit ($\chi^2(1) = 5.54, p= 0.01854$), suggesting that the experimental condition affects vowel height.
-However, since we expected the extent of change in vowel height between the conditions to depend on the identity of the vowel (cf sec.\ref{asymmetries}), we added a two-way interaction factor between Vowel and Context.
+Following from our hypothesis \textbf{Asymmetry}, we expected the extent of change in vowel height between the conditions to be mediated by the identity of the vowel.
+Thus, we added a two-way interaction factor between Context and Vowel (Context*Vowel).
 As expected, this was found to significantly improve model fit ($\chi^2(1) = 12.61, p= 0.0003837$).
-To check if the experimental paradigm affected F1, we added Task as a fixed effect, and this marginally improved the model ($\chi^2(1) = 3.74, p= 0.05302$).
-Following from the hypothesis in sec.\ref{paradigms}, we expected the extent of shift to differ across the two paradigms.
+To assess whether the experimental paradigm affected F1, we added Task as a fixed effect.
+This marginally improved the model ($\chi^2(1) = 3.74, p= 0.05302$).
+Towards addressing our research question \textbf{Paradigm}, we wanted to assess whether the extent of shift differs across the two paradigms.
 However, no significant improvement was achieved by adding an interaction effect between Task and Context.
@@
 This improved the model fit ($\chi^2(5) = 11.43, p = 0.04342$).
-Thus, we report on the model including Gender, Task, Context, Vowel, and a Context-Vowel interaction as fixed effects, random intercepts for Subject and Item, by-subject random slopes for Task, Context, Vowel and the Context-Vowel interaction, and by-item random slopes for Task and Context.
+Thus, we report on a model that includes Gender, Task, Context, Vowel, Context*Vowel as fixed effects, random intercepts for Subject and Item, by-subject random slopes for Task, Context, Vowel and Context*Vowel, and by-item random slopes for Task and Context.\\
 
@@
 Vowel is not a significant predictor of F1.
-This seems surprising, given that the phonemic category is expected to systematically affect vowel quality.
-However, a boxplot visualization (Figure\ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two categories and the extent of height difference between them.
-Note that these measures are only from the English (unilingual) condition, suggesting that the variability is a feature of the participants' IE vowel categories, independent of the effects of switching.
+This is surprising, given that the phonological category is expected to systematically affect vowel quality.
+However, a boxplot visualization (Figure\ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two categories \nt{2, \ae} and the extent of height difference between them.
+Note that these measures are only from the English (unilingual) condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching.
 The lack of any independent effect of Context on F1 suggests that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels.
@@
 As noted earlier, there is no significant interaction of Task with any of the other fixed factors.
-We used a series of ANOVAs to compare the above model with a null model that differs in the absence of the main factor which concerns our hypothesis-- Vowel*Context-- at each point in the vowel.
-We find that the effect is not significant at 5\%, and increases over the course of articulating the vowel.
+We used a series of ANOVAs to compare the model elaborated above with a null model that differs in the absence of the main factor which concerns our hypothesis, Vowel*Context, at each point in the vowel.
+We find that the difference is not significant at 5\%, and increases over the course of articulating the vowel.
 %ANOVA table add
 
-\begin{figure}
+\begin{figure}[h] \centering
 
@@
 		& & & & & \\ 
-		Gendermale & $-$1.171$^{***}$ & $-$1.526$^{***}$ & $-$1.586$^{***}$ & $-$1.668$^{***}$ & $-$1.681$^{***}$ \\ 
+		Gender(male) & $-$1.171$^{***}$ & $-$1.526$^{***}$ & $-$1.586$^{***}$ & $-$1.668$^{***}$ & $-$1.681$^{***}$ \\ 
 		& (0.237) & (0.184) & (0.168) & (0.177) & (0.175) \\ 
 		& & & & & \\ 
-		Contexte & $-$0.010 & $-$0.015 & 0.033 & 0.025 & 0.037 \\ 
+		Context(e) & $-$0.010 & $-$0.015 & 0.033 & 0.025 & 0.037 \\ 
 		& (0.058) & (0.057) & (0.042) & (0.040) & (0.035) \\ 
 		& & & & & \\ 
-		Vowel\nt{2} & 0.022 & $-$0.070 & $-$0.034 & $-$0.131 & $-$0.216 \\ 
+		Vowel(\textipa{2}) & 0.022 & $-$0.070 & $-$0.034 & $-$0.131 & $-$0.216 \\ 
 		& (0.170) & (0.175) & (0.170) & (0.174) & (0.168) \\ 
 		& & & & & \\ 
-		Tasks & 0.319$^{***}$ & 0.235$^{***}$ & 0.172$^{**}$ & 0.133$^{*}$ & 0.102 \\ 
+		Task(s) & 0.319$^{***}$ & 0.235$^{***}$ & 0.172$^{**}$ & 0.133$^{*}$ & 0.102 \\ 
 		& (0.091) & (0.083) & (0.081) & (0.077) & (0.073) \\ 
 		& & & & & \\ 
-		Contexte:Vowel\nt{2} & $-$0.126$^{**}$ & $-$0.138$^{**}$ & $-$0.185$^{***}$ & $-$0.191$^{***}$ & $-$0.192$^{***}$ \\ 
+		Context(e):Vowel(\textipa{2}) & $-$0.126$^{**}$ & $-$0.138$^{**}$ & $-$0.185$^{***}$ & $-$0.191$^{***}$ & $-$0.192$^{***}$ \\ 
 		& (0.056) & (0.060) & (0.055) & (0.064) & (0.066) \\ 
@@
 The estimates for Vowel show that excepting the beginning of the vowel, the category \nt{2} is consistently produced as more backed than \nt{\ae}.
-Unlike vowel height, Context has an independent effect on vowel backness, such that both the vowels in this study have lower F2 values in the English (unilingual) condition, i.e.
+Unlike vowel height, Context has an independent effect on vowel backness, such that both the vowels in this study have higher F2 values in the Bengali (mixed) condition, i.e.
 they are fronted in the mixed condition.
-A significant interaction between Vowel and Context shows that there is a difference in the extent of fronting between the two vowels -- the shift is greater in the case of \nt{\ae}.
-We see a significant effect of Task on F2, such that vowels in the sentence task (code-switching) were more backed than those in the picture-naming task.
-As with F1, this is an unexpected result.
+A significant interaction between Vowel and Context reveals an asymmetry in the extent of fronting between the two vowels -- the shift is greater in the case of \nt{\ae}.
+We see a significant effect of Task on F2, such that vowels in the Sentence task (code-switching) were more backed than those in the picture-naming task.
+As with F1, this is an unexpected result, as we have no principled reason to expect code-switching to induce an across-board backing of vowels.
 A lack of significant interaction with Context suggests that there is no difference in the extent of shift between the two paradigms.
-The results of ANOVAs comparing the full model to a null model which lacks the Vowel*Context interaction factor show that as with vowel height, the change in vowel backness between the two conditions is not significant at 5\%, and increases over the course of the vowel.
+The results of ANOVAs comparing the full model to a null model that lacks the Vowel*Context interaction factor reveal that as with vowel height, the change in vowel backness between the two conditions is not significant at 5\%, and increases over the course of the vowel.
 
@@
 		& & & & & \\ 
-		Gendermale & $-$0.839$^{***}$ & $-$0.891$^{***}$ & $-$0.807$^{***}$ & $-$0.750$^{***}$ & $-$0.748$^{***}$ \\ 
+		Gender(male) & $-$0.839$^{***}$ & $-$0.891$^{***}$ & $-$0.807$^{***}$ & $-$0.750$^{***}$ & $-$0.748$^{***}$ \\ 
 		& (0.124) & (0.106) & (0.107) & (0.116) & (0.109) \\ 
 		& & & & & \\ 
-		Contexte & 0.003 & $-$0.088$^{**}$ & $-$0.065$^{**}$ & $-$0.063 & $-$0.103$^{***}$ \\ 
+		Context(e) & 0.003 & $-$0.088$^{**}$ & $-$0.065$^{**}$ & $-$0.063 & $-$0.103$^{***}$ \\ 
 		& (0.036) & (0.036) & (0.032) & (0.044) & (0.033) \\ 
 		& & & & & \\ 
-		Vowel\nt{2} & $-$0.217 & $-$0.615$^{***}$ & $-$0.651$^{***}$ & $-$0.801$^{***}$ & $-$0.836$^{***}$ \\ 
+		Vowel(\textipa{2}) & $-$0.217 & $-$0.615$^{***}$ & $-$0.651$^{***}$ & $-$0.801$^{***}$ & $-$0.836$^{***}$ \\ 
 		& (0.175) & (0.183) & (0.175) & (0.168) & (0.166) \\ 
 		& & & & & \\
-		Tasks & $-$0.234$^{***}$ & $-$0.194$^{***}$ & $-$0.152$^{***}$ & $-$0.141$^{***}$ & $-$0.092$^{**}$ \\ 
+		Task(s) & $-$0.234$^{***}$ & $-$0.194$^{***}$ & $-$0.152$^{***}$ & $-$0.141$^{***}$ & $-$0.092$^{**}$ \\ 
 		& (0.029) & (0.035) & (0.050) & (0.047) & (0.045) \\ 
 		& & & & & \\
-		Contexte:Vowel\nt{2} & 0.031 & 0.110$^{***}$ & 0.128$^{***}$ & 0.091$^{**}$ & 0.147$^{***}$ \\ 
+		Context(e):Vowel(\textipa{2}) & 0.031 & 0.110$^{***}$ & 0.128$^{***}$ & 0.091$^{**}$ & 0.147$^{***}$ \\ 
 		& (0.050) & (0.041) & (0.040) & (0.046) & (0.043) \\ 
@@
 
-\section{Discussion and Conclusion}
+\section{Discussion}
 
-With regard to our first research question, the results confirm that the L2 vowels of proficient bilinguals can be affected by mixed-language processing.
+With regard to our first research question \textbf{L2 vowels}, the results confirm that the L2 vowels of proficient bilinguals can be affected by mixed-language processing.
 Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.
-This is significant in the context of an ICM-based analysis of between-language asymmetries in phonetic transfer as proposed by \citep{olson2013bilingual}, because it demonstrates that at least for vowels, the postulated lower switch cost on an L2 does not entirely mask transfer effects.
+In the context of an ICM--based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, this is a significant finding: it demonstrates that the postulated lower switch cost on an L2 does not entirely mask transfer effects on L2 vowels.
+Moreover, our findings differ from \cite{tsui2019impact} who report on the basis of VOT measurements that balanced bilinguals do not evidence an effect of language-switching.
+The authors attribute this to greater experience with language switching, leading to lower switch-costs in both languages.
+We find that even proficient bilinguals with extensive experience of language-switching demonstrate cross-language transfer effects in vowel production.
 Thus, following from the discussion in sec.
-\ref{asymmetries}, it might be fruitful for future studies to focus on vowels as a site for transfer.
-This would both expand the scope of the existing research (cf sec.
-\ref{asymmetry between sounds}) and avoid confounding factors arising from independent differences between long- and short-lag VOT languages.
+\ref{asymmetries}, it might be fruitful for future studies to focus on vowels as a potential site for transfer.
+This would both expand the scope of the existing research (cf.
+sec.
+\ref{asymmetry between sounds}) and avoid confounding factors arising from independent differences between long- and short-lag VOT languages.\\
 
-We hypothesized an asymmetrical shift between the two vowels, owing to differences in (i) their positions in the Indian English vowel space, and (ii) the existence of a corresponding category in the L1 Bengali inventory.
+We hypothesized an asymmetrical shift between the two vowels (\textbf{Asymmetry}), owing to differences in (i) their positions in the Indian English vowel space, and (ii) the presence/absence of a corresponding category in the L1 Bengali inventory.
 The significant interaction between Vowel and Context in both F1 and F2 confirm that the experimental condition affected both height and backness differently for each vowel.
-As expected, we find a greater degree of shift in \nt{2}, possibly owing to its position in a less crowded part of the IE vowel space.
-Figure \ref{vowels_e_b} shows F1xF2 plots for the vowels in the unilingual and mixed (dotted ellipses) conditions, in comparison to the related Bengali categories (colored ellipses) from \cite{dutta2016using}.
+As expected, we find a greater degree of shift in \nt{2}, possibly owing to its position in a less dense part of the IE vowel space.
+Figure \ref{vowels_e_b} shows $F1xF2$ plots for vowel productions in the unilingual and mixed (dotted ellipses) conditions, in comparison to the related Bengali categories (colored ellipses) from \cite{dutta2021}.
 The category \nt{2} is lowered and fronted in the mixed language condition as expected.
@@
 Thus, fronting in the mixed condition cannot be explained in relation the former, at least from the present data.
-Data from a larger number of speakers, both for IE and Bengali, is needed to explore this further.
+Recall that the Bengali vowel productions used here are from a different set of speakers than our participants.
+Assuming some amount of individual differences in the precise phonetic realization of vowel categories in both languages, this could be a potential source of explanation for this result.
+Data from a larger number of speakers for both languages, as well as within-speaker comparisons of English and Bengali vowels, are required to clarify this.
+\\
 
-% only ae -- shift significant?
-% difference between vowels-- transfer--makes reference to phonological categories--
-% knitr -- put the R code for figures in here so changes update here
-%in intro--say-- paper argues that these asymetries (seen in previous lit) point to the (urgent:)) need for diversifying the languages and sound features studied in order to make generalizations
-\begin{figure}
-
+\begin{figure}[h] \centering
 	\includegraphics[scale=0.8]{vowels_e_b_final}
@@
 
-Our third research question concerned the experimental paradigms.
+Our third research question \textbf{Paradigm} concerned the effect of experimental paradigm on the phonetic outcome of mixed-language use.
 \cite{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
-This calls to question the comparability of results across studies using these different paradigms.
-We tested this by comparing productions of the same items from the same group of participants in a scripted CS paradigm and a cued picture-naming paradigm.
+If such discursive properties independently affect the phonetic outcome of language mixing in an experimental setup, this calls to question the comparability of results across studies using these different paradigms.
+We verified this by comparing productions of the same items from the same group of participants in a scripted CS paradigm and a cued picture-naming paradigm.
 We found no significant interaction of Task either with Context, or with the Vowel*Context interaction, suggesting that the extent of phonetic transfer was not affected by the paradigm.
 There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in observable transfer at all.
-This would mean that the hyperarticulation seen, for example, in \cite{muldner2019phonetics} is better interpreted as resulting from some other aspect of the experimental setup, rather than the act of code-switching itself.
+This would mean that the pre-switch effects observed, for example, in \cite{muldner2019phonetics} and \cite{bullock2009trying} are better interpreted as resulting from some other aspect of the experimental setup, than the act of code-switching itself.
 While planning could additionally affect the words preceding the switch, the effects on the switched token itself do not change.
@@
 Thus, another possible explanation is that the discursive effects discussed by \cite{olson2013bilingual} are triggered by more naturalistic paradigms where attention to meaning and interlocutor are important.
-More within-study comparisons with less structured paradigms, and spontaneous code-switching data, can shed light on this.
-The results of the present study suggest that for the purposes of observing dynamic phonetic transfer, elicitation through cued-picture-naming vs (controlled) code-switching does not seem to significantly alter the behavior under study, and thus the choice of paradigm can be based on other considerations.
+More within-study comparisons with less structured paradigms, and spontaneous code-switching data, can clarify this.
+The results of the present study indicate that for the purposes of observing dynamic phonetic transfer in an experimental setup, elicitation through cued-picture-naming vs (controlled) code-switching does not significantly alter the behavior under study.
+Thus, the choice of paradigm can be based on other considerations.
 Moreover, this suggests that (all else being equal) the results from such studies can be compared directly.
 
-Paradigm did affect vowel quality, but this applied uniformly across experimental conditions---all vowels were more backed and lowered in the code-switching task.
+The paradigm did affect vowel quality, but this applied uniformly across experimental conditions: all vowels were more backed and lowered in the code-switching task.
 Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this lowering is an artifact of the particular stimuli used here---resulting from some corarticulatory effect that was absent in the picture-naming task.
 While this is intriguing, it is beyond the scope of the present study.
+\\
 
-Mixed language processing temporarily increases cross-language phonetic transfer, causing an L2 vowel to converge towards a related L1 category.
-The extent and direction of such shift is mediated by language-specific distribution of sound categories.
-In a controlled experimental setup to study transfer effects, the postulated discursive factors in a code-switching paradigm does not change the patterns of observed transfer relative to a cued picture-naming paradigm.
+\section{Conclusion}
+This study examined the phonetic outcome of mixed-language production in a group of highly proficient bilingual speakers of Bengali and Indian English.
+Specifically, we examined the spectral quality of two L2 vowel categories \nt{2} and \nt{\ae} produced in two experimental conditions: unilingual (English) and mixed (switching from Bengali to English).
+Comparisons of F1 and F2 reveal that mixed language processing temporarily increases phonetic interaction between the languages, causing \nt{2} to converge towards a related L1 category \nt{a:}.
+The extent and direction of such shift is asymmetrical across the two categories, which parallels findings in prior studies.
+This asymmetry appears to be mediated by the language-specific distribution of vowel categories.
+Mixed productions were elicited in two switching paradigms: code-switching and cued picture-naming, to assess whether the postulated discursive elements in code-switching independently influences the outcome of phonetic transfer.
+Results suggest that in an experimental setup, elicitation through cued-picture-naming vs (controlled) code-switching does not significantly alter the behavior under study.
+\\
+
+This study has a number of limitations: (i) the `baseline' values for Bengali vowel categories were estimated from a different set of speakers than the participants of this study, potentially obscuring the role of individual differences in the realization of vowel categories; (ii) we used a frame sentence to elicit code-switched productions such that not all the sentences were pragmatically well-formed.
+This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study.
+Future work should address this by considering data from more naturalistic code-switching paradigms.
+
+\section*{Abbreviations}
+\textsc{clf} = classifier
+
+\section*{Data availability}
+The data files and analysis code are available at:  \url{https://osf.io/dsb2x/?view_only=5cb821070bc74a038a293239157c3b11}
+
+\section*{Competing interests}
+The authors have no competing interests to declare.
 

### 2021-06-22 draft_BLC/BLC_AM-ID_manuscript.tex
BLC draft

--- 2021-06-11 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
@@
 % see http://info.semprag.org/basics for a full description of this template
-\documentclass[charis,linguex]{glossa}
+%\documentclass[times,linguex]{glossa}
+\documentclass[12 pt]{article}
+\usepackage[letterpaper]{geometry}
+%\usepackage{times}
+\geometry{top=1.5in, bottom=1.0in, left=1.0in, right=1.0in}
 % possible options: 
@@
 % !! a note on the use of bibtex: for PhD dissertations to typeset correctly in the references list, the Address field needs to contain the city (for US cities in the format "Santa Cruz, CA")
-
 %\addbibresource{sample.bib} 
@@
 %for natbib: replace \cite with \citep (parenthetical); \citea with \cite (in-text)
-\usepackage{hyperref}
-\let\B\relax %to resolve a conflict in the definition of these commands between xyling and xunicode (the latter called by fontspec, called by charis)
-\let\T\relax
-\usepackage{xyling} %for trees; the use of xyling with the CharisSIL font produces poor results in the branches. This problem does not arise with the packages qtree or forest.
-\usepackage[linguistics]{forest} %for nice trees!
+\usepackage[hidelinks]{hyperref}
+%\renewcommand{\sectionautorefname}{\S}
+\usepackage{xcolor}
+\hypersetup{
+	colorlinks,
+	linkcolor={blue!50!black},
+	citecolor={blue!50!black},
+	urlcolor={blue!80!black}
+}
+
+\usepackage{graphicx} %for images
+\graphicspath{{images/}}
+\usepackage{caption}
+
+\usepackage{booktabs}
+\usepackage{enumitem} %customize list numberings
+
+\usepackage[natbibapa]{apacite}
+\usepackage{natbib}
 
@@
 \newcommand{\wt}[1]{\textipa{/#1/}} % wide transcription
+%\usepackage{natbib}
 
-%\graphicspath{{images/}}
+\usepackage{fancyhdr}
+\pagestyle{fancy}
+\fancyhf{}
+\renewcommand{\headrulewidth}{0pt}
+\fancyhead[R]{\thepage}
 
-% \pdf* commands provide metadata for the PDF output. ASCII characters only!
-%\pdfauthor{Auromita Mitra, Indranil Dutta}
-%\pdftitle{Mixed language processing increases cross-language phonetic transfer in Bengali-English Bilinguals}
-\pdfkeywords{phonetic transfer, vowel formants, Bengali, English}
+\usepackage{setspace}
+\doublespacing
+
+\usepackage{comment}
+
+\newlength\mystoreparindent
+\newenvironment{myparindent}[1]{%
+	\setlength{\mystoreparindent}{\the\parindent}
+	\setlength{\parindent}{#1}
+}{%
+	\setlength{\parindent}{\mystoreparindent}
+}
 
 \title{Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals}
-% Optional short title inside square brackets, for the running headers.
 
@@
 \begin{document}
+\bibliographystyle{apacite}
+%\sffamily
+%\maketitle
 
-\sffamily
-\maketitle
+%Word count: 9263
 
-Word count: 9263
+\begin{singlespace}
 
-\begin{abstract}
-This study investigates the phonetic outcome of mixed-language production in a group of highly proficient bilingual speakers of Bengali and English, to better understand the nature of cross-language phonetic interaction.
-Existing research has largely focused on a small set of phonologically related languages and used voice onset time (VOT) to measure phonetic transfer.
-Results indicate that such interaction is mediated by both language-specific features and linguistic experience, highlighting the need for diversification in the range of languages and populations studied.
-We compare productions from a yet-unstudied population in two conditions-- unilingual L2 (English) and mixed (Bengali and English).
-The spectral properties of two L2 vowel categories— \nt{\ae} (common to L1 and L2) and \nt{2} (absent in L1) are examined as sites for transfer.
-To understand how this unfolds dynamically, we measure F1 and F2 at five points over the course of articulating the vowel.
-We find dynamic and asymmetrical changes in both vowels during mixed production, with \nt{2} shifting towards a related L1 category \nt{a:}.
-We interpret this as a temporary increase in cross-language phonetic influence during mixed-language use, leading to a shift towards L1 norms.
-Mixed language is elicited in two switching paradigms: cued picture-naming and code-switching, to assess if differences between the paradigms independently influence the outcome of phonetic interaction.
-Results reveal no difference in transfer patterns across the two paradigms.
-We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the postulated role of connected speech in introducing discursive factors to transfer studies.
-\end{abstract}
+\begin{myparindent}{0pt}
+Short title: Phonetic transfer\\
 
-\begin{keywords}
- phonetic transfer; vowel quality; bilingual speech production; L1-L2 interaction; Indian English; Bengali 
-\end{keywords}
+Full title: Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals\\
 
-\rmfamily
+Authors:\\
+
+Auromita Mitra
+
+EFL University
+
+Hyderabad\\
+
+Indranil Dutta
+
+Jadavpur University
+
+Kolkata\\
+
+Acknowledgements:\\
+
+Address for correspondence:\\
+Auromita Mitra,
+
+EFL University,
+
+Hyderabad- 500007,
+
+auromita.mitra@gmail.com
+
+\end{myparindent}
+
+\end{singlespace}
+
+\newpage
+\begin{myparindent}{0pt}
+Abstract:\\
+This study investigates the phonetic outcome of mixed-language speech in Bengali and Indian English, towards understanding cross-language interaction in highly proficient bilinguals.
+We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in unilingual vs mixed productions.
+Results reveal dynamic and asymmetrical changes in both vowels, \nt{2} shifting towards a related L1 category \nt{a:}.
+We interpret this as a temporary increase in cross-language phonetic interaction during mixed-language use, leading to a shift towards L1 norms, evidencing transfer effects on L2 vowels for the first time.
+We elicit mixed productions through two tasks: cued picture-naming and code-switching, to assess if experimental paradigm independently influences the behavior under study.
+Results reveal no difference across paradigms.
+We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction between sound systems and the postulated role of connected speech in introducing discursive factors to transfer studies.\\
+
+Keywords: phonetic transfer; vowel quality; bilingual speech production; L1-L2 interaction; Indian English; Bengali
+
+\end{myparindent}
+
+\newpage
 
@@
 
-\subsubsection{Demography}
+\subsubsection*{Demography}
 
@@
 
-\subsubsection{Vowel systems} \label{vowel systems}
+\subsubsection*{Vowel systems} \label{vowel systems}
 The vowel inventory of Western Bangla (the variety spoken by the participants in this study) consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \citep{garry2001facts}.
@@
 
-\subsubsection{L1 vs L2 status} 
+\subsubsection*{L1 vs L2 status} 
 Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} posits that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another.
@@
 
-\subsubsection{Differences between sound systems of the languages} \label{sound systems} In addition to the cognitive factors discussed above, many researchers have attributed the observed asymmetries to language-internal factors.
+\subsubsection*{Differences between sound systems of the languages} \label{sound systems} In addition to the cognitive factors discussed above, many researchers have attributed the observed asymmetries to language-internal factors.
 Specifically, a pattern that emerges from the body of work on VOT is that a shift in long-lag VOT towards the short-lag norms of the other language is much more consistent and systematic than the converse \citep{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid}.
@@
 sec.\ref{vowel systems}).
-\end{enumerate}\\
+\end{enumerate}
 
@@
 	\caption{Fixed factors}
-	\sffamily
 	\centering
-	\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
-		\hline 
-		Factor & Levels\\ 
-		\hline
-		Gender & male, female\\
+	\resizebox{\textwidth}{!}{%
+		\begin{tabular}{@{}ll@{}}
+			\toprule
+			Factor                                                 & Levels                                                            \\ \midrule
+			Gender                                                 & male, female                                                      \\
 			Vowel                                                  & \nt{2}, \nt{\ae} \\
-		Task (corresponding to the two paradigms) & p (Picture: cued picture-naming, s (Sentence: code-switching)\\
-		Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)\\ [1ex]
-		\hline	
-	\end{tabular}
+			Task (corresponding to the two paradigms)              & p (Picture: cued picture-naming, s (Sentence: code-switching)     \\
+			Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)                       \\ \bottomrule
+		\end{tabular}%
+	}
 \label{table variables}
@@
 Thus, we began with a model that had Vowel and Gender as fixed effects, random intercepts for Subject and Item, and random slopes for each within-unit variable that was relevant to our hypotheses-- by-subject slopes for Task, Context, and Vowel; by-item slopes for Task and Context.
-Next, adding Context as a fixed effect significantly improved the model fit ($\chi^2(1) = 5.54, p= 0.01854$), suggesting that the experimental condition affects vowel height.
+Next, adding Context as a fixed effect significantly improved the model fit ($\chi^2(1) = 5.54, p= .01854$), suggesting that the experimental condition affects vowel height.
 Following from our hypothesis \textbf{Asymmetry}, we expected the extent of change in vowel height between the conditions to be mediated by the identity of the vowel.
 Thus, we added a two-way interaction factor between Context and Vowel (Context*Vowel).
-As expected, this was found to significantly improve model fit ($\chi^2(1) = 12.61, p= 0.0003837$).
+As expected, this was found to significantly improve model fit ($\chi^2(1) = 12.61, p= .0003837$).
 To assess whether the experimental paradigm affected F1, we added Task as a fixed effect.
-This marginally improved the model ($\chi^2(1) = 3.74, p= 0.05302$).
+This marginally improved the model ($\chi^2(1) = 3.74, p= .05302$).
 Towards addressing our research question \textbf{Paradigm}, we wanted to assess whether the extent of shift differs across the two paradigms.
@@
 Since our hypothesis concerns their interaction, we modified the random effects structure by adding a by-subject random slope for the interaction term \citep{brauer2018linear}.
-This improved the model fit ($\chi^2(5) = 11.43, p = 0.04342$).
+This improved the model fit ($\chi^2(5) = 11.43, p = .04342$).
 Thus, we report on a model that includes Gender, Task, Context, Vowel, Context*Vowel as fixed effects, random intercepts for Subject and Item, by-subject random slopes for Task, Context, Vowel and Context*Vowel, and by-item random slopes for Task and Context.\\
@@
 As expected, we find a greater degree of shift in \nt{2}, possibly owing to its position in a less dense part of the IE vowel space.
-Figure \ref{vowels_e_b} shows $F1xF2$ plots for vowel productions in the unilingual and mixed (dotted ellipses) conditions, in comparison to the related Bengali categories (colored ellipses) from \cite{dutta2021}.
+Figure \ref{vowels_e_b} shows $F1XF2$ plots for vowel productions in the unilingual and mixed (dotted ellipses) conditions, in comparison to the related Bengali categories (colored ellipses) from \cite{dutta2021}.
 The category \nt{2} is lowered and fronted in the mixed language condition as expected.
@@
 This asymmetry appears to be mediated by the language-specific distribution of vowel categories.
-Mixed productions were elicited in two switching paradigms: code-switching and cued picture-naming, to assess whether the postulated discursive elements in code-switching independently influences the outcome of phonetic transfer.
+Mixed productions were elicited in two switching paradigms: code-switching and cued picture-naming, to assess whether the postulated discursive elements in code-switching independently influence the outcome of phonetic transfer.
 Results suggest that in an experimental setup, elicitation through cued-picture-naming vs (controlled) code-switching does not significantly alter the behavior under study.
@@
 \section*{Data availability}
-The data files and analysis code are available at:  \url{https://osf.io/dsb2x/?view_only=5cb821070bc74a038a293239157c3b11}
+The data files and analysis code are openly available at:  \url{https://osf.io/dsb2x/?view_only=5cb821070bc74a038a293239157c3b11}
 

### 2021-07-04 draft_BLC/BLC_AM-ID_manuscript.tex
BLC formatting

@@
 \geometry{top=1.5in, bottom=1.0in, left=1.0in, right=1.0in}
-% possible options: 
-% [times] for Times font (default if no option is chosen)
-% [cm] for Computer Modern font 
-% [lucida] for Lucida font (not freely available)
-% [brill] open type font, freely downloadable for non-commercial use from http://www.brill.com/about/brill-fonts; requires xetex
-% [charis] for CharisSIL font, freely downloadable from http://software.sil.org/charis/
-% for the Brill an CharisSIL fonts, you have to use the XeLatex typesetting engine (not pdfLatex)
-% for headings, tables, captions, etc., Fira Sans is used: https://www.fontsquirrel.com/fonts/fira-sans
-% [biblatex] for using biblatex (the default is natbib, do not load the natbib package in this file, it is loaded automatically via the document class glossa.cls)
-% [linguex] loads the linguex example package
-% !!
-a note on the use of linguex: in glossed examples, the third line of the example (the translation) needs to be prefixed with \glt.
-This is to allow a first line with the name of the language and the source of the example.
-See example (2) in the text for an illustration.
-% !! a note on the use of bibtex: for PhD dissertations to typeset correctly in the references list, the Address field needs to contain the city (for US cities in the format "Santa Cruz, CA")
-%\addbibresource{sample.bib} 
-% the above line is for use with biblatex
-% replace this by the name of your bib-file (extension .bib is required)
-% comment out if you use natbib/bibtex
 
-%for natbib: replace \cite with \citep (parenthetical); \citea with \cite (in-text)
 \usepackage[hidelinks]{hyperref}
@@
 \usepackage{caption}
+\captionsetup[table]{skip=0pt,singlelinecheck=off}
+\captionsetup[figure]{skip=0pt,singlelinecheck=off}
 
@@
 \newcommand{\wt}[1]{\textipa{/#1/}} % wide transcription
-%\usepackage{natbib}
 
@@
 
+School of languages and Linguistics
+
 Jadavpur University
@@
 
-Acknowledgements:\\
+Acknowledgments:\\
+We are grateful to Maumita Bhowmik and Anannya Mondal for their help with data collection and processing, and to all the participants of the study.\\
 
@@
 We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in unilingual vs mixed productions.
-Results reveal dynamic and asymmetrical changes in both vowels, \nt{2} shifting towards a related L1 category \nt{a:}.
+Results reveal dynamic and asymmetrical changes in both vowels;  \nt{2} shifting towards a related L1 category \nt{a:}.
 We interpret this as a temporary increase in cross-language phonetic interaction during mixed-language use, leading to a shift towards L1 norms, evidencing transfer effects on L2 vowels for the first time.
@@
 
-\section{Phonetic transfer}
+\section{Introduction: Phonetic transfer}
 
 Bi/multilingual speakers can distinguish between the phonetic norms of their languages and maintain separate sound categories for each language \citep{caramazza1973acquisition,macleod2010impact,bosch2003simultaneous}.
-However, these categories are not autonomous-- they influence each other across languages in both perception and production \citep[e.g.][]{flege1995second,fowler2008cross,flege2002assessing}, and the nature of such interaction provides crucial insights into how language ``systems" are cognitively represented and processed.
-Based on the productions of proficient bilingual speakers of Bengali and English, this study reports that cross-language phonetic influence temporarily increases during mixed-language use.\\%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
+However, these categories are not autonomous-- they influence each other across languages in both perception and production \citep[e.g.][]{flege1995second,fowler2008cross,flege2002assessing}, and the nature of such interaction provides crucial insights into how language `systems' are cognitively represented and processed.
+Based on the productions of proficient bilingual speakers of Bengali and English, this study reports that cross-language phonetic influence temporarily increases during mixed-language use.%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
 %show cross-language-- rephrase
@@
 The present study is concerned with the latter.
-In production, this kind of influence has been variously termed \textit{transfer, drift, accommodation,} and \textit{interference}-- we use these interchangeably here to indicate any interaction between two sets of phonetic norms.\\
+In production, this kind of influence has been variously termed \textsc{transfer, drift, accommodation,} and \textsc{interference}-- we use these interchangeably here to indicate any interaction between two sets of phonetic norms.
 
@@
 However, there is no work yet on short-term phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language.
-\\
 
@@
 The results demonstrate a dynamic effect of transfer on L2 vowels during mixed-language production, but no difference in the extent of transfer between the two paradigms.
-We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.\\
+We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.
 
@@
 A logical consequence of this model is that such phonetic interaction must be highly localized.
-\ref{duration} elaborates on this by reviewing literature on the duration and ``reversiblity" of transfer effects.
+\ref{duration} elaborates on this by reviewing literature on the duration and `reversiblity' of transfer effects.
 A recurring pattern that emerges from existing research on short-term phonetic transfer is that both languages of the bilingual speaker are not equally affected, although the nature of these reported differences varies greatly across studies.
@@
 In India, it is spoken as an L2 by 129 million people (Census of India, 2011).
-English is one of the two official languages, used in education, law, media, as a \emph{lingua franca} mainly for an educated elite in most metropolitan regions, and carries a high prestige value \citep{pandey201517, kachru1981english, tollefson2014language, kachru1983indianization}.
+English is one of the two official languages, used in education, law, media, as a lingua franca mainly for an educated elite in most metropolitan regions, and carries a high prestige value \citep{pandey201517, kachru1981english, tollefson2014language, kachru1983indianization}.
 %citations 
@@
 Thus, it is fruitful to think of regional variations as resulting from L1-influence on a common underlying target.
-Excepting a small minority of specific communities \citep{pandey201517, wells1982accents, coelho1997anglo}, there are very few L1 speakers of IE.\\
+Excepting a small minority of specific communities \citep{pandey201517, wells1982accents, coelho1997anglo}, there are very few L1 speakers of IE.
 
@@
 	\item In multilingual populations, long-term representations of sound categories are expected to be affected by multiple languages. 
-	\item Given that examples of an IE phonology without L1 ``influence" are very rare, it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a given baseline condition.
+	\item Given that examples of an IE phonology without L1 `influence' are very rare, it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a given baseline condition.
 \end{enumerate}
@@
 
-Existing research has distinguished changes in category \textit{representations} due to the acquisition of multiple sound systems (cf.
+Existing research has distinguished changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf.
 Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during accessing, processing, or articulation of these categories (cf.
@@
 	\item The local point of switch itself
-	\item Global co-activation of two languages -- bilingual language mode \citep{grosjean1998studying} 
+	\item Global co-activation of two languages -- \textsc{bilingual language mode} \citep{grosjean1998studying} 
 \end{enumerate}
@@
 A number of studies have specifically manipulated language mode, both in the presence and absence of switching.
-Overall, results suggest that: (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use -- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \citep{olson2016role,tsui2019impact} or spontaneous conversation \citep{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
-(iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \cite{olson2016role} found no additive effects, \cite{olson2013bilingual} found the balanced language context to inhibit transfer compared to unbalanced contexts.
-Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days \citep{schwartz2015language, bullock2009trying,antoniou2011inter, elias2017effects,vsimavckova2015immediate,vsimavckova2018patterns}.\\
+Overall, results suggest that: (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use--- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \citep{olson2016role,tsui2019impact} or spontaneous conversation \citep{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
+(iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \cite{olson2016role} found no additive effects of language mode, \cite{olson2013bilingual} found a balanced language context to inhibit transfer compared to unbalanced contexts.
+Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days \citep{schwartz2015language, bullock2009trying,antoniou2011inter, elias2017effects,vsimavckova2015immediate,vsimavckova2018patterns}.
 
-\cite{grosjean1998studying} suggests that various aspects of the communicative setting, including exposure to (spoken or written) stimuli in multiple languages and awareness of the interlocutor's being bilingual, could trigger a bilingual mode.
+\cite{grosjean1998studying} suggests that various aspects of the communicative setting, including exposure to (spoken or written) stimuli in multiple languages and awareness of the interlocutor's being bilingual, could trigger a bilingual language mode.
 The participants in the present study are in an environment which largely contains mixed-language input, multilingual interlocutors, and no stable ambient language.
@@
 To preserve ecological validity, we elicited both unilingual and switched utterances in a bilingual language mode.
-Any observed differences in this paradigm would arguably result from interaction during online processing.
-\\
+Any observed differences in this paradigm would arguably result from interaction during online processing of sounds while switching between languages.
 
@@
 
-Moreover, changes due to transfer are not unidirectional or irreversible.
+Moreover, changes due to transfer are not necessarily unidirectional or irreversible.
 \cite{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language.
@@
 How long these effects last in the absence of regular L2 input was not tested.
-\\
 % if effects are cumulative, then the difference between switched and non-switched tokes should be lesser in later test blocks (previous two paragraphs needed?)
 
-The majority of studies on the phonetics of mixed-language use only analyze the switched (target) token.
+The research discussed above concerns situations where the bilingual speaker is operating in any one of their languages.
+When it comes to the phonetics of mixed-language use, the majority of existing studies only analyze the switched (target) token.
 Thus, there are few direct measurements of the duration of short-term transfer effects.
 One study which measured this in a code-switching paradigm \citep{bullock2009trying} did not find any residual effects on the matrix language following a switch, suggesting that transfer during code-switching is localized.
-Indirect evidence in support of this comes from experiments which have elicited switched and non-switched tokens in the same test block and still reported differences between the two \citep[e.g.][]{tsui2019impact,olson2013bilingual}, suggesting that changes due to transfer are quickly ``reset"-- in an order of seconds.\\
+Indirect evidence in support of this comes from experiments which have elicited switched and non-switched tokens in the same test block and still reported differences between the two \citep[e.g.][]{tsui2019impact,olson2013bilingual}, suggesting that changes due to transfer are quickly `reset'--- in an order of seconds.\\
 Note that all the studies discussed above measure VOT, which is a temporal feature.
@@
 In non-switched production, this is evidenced through a pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf.
-\textit{equivalence classification}-- \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][]{guion2003vowel}.
-\\
+\textsc{equivalence classification}-- \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][]{guion2003vowel}.
 
@@
 This established that beyond language-specific differences, the L1 vs L2 status of the language does mediate transfer.
-They interpret this asymmetry in terms of the Inhibitory Control Model (ICM) \citep{green1998mental}: to select a phonetic realization from one language, the other must be inhibited.
-Since L1 is the ``stronger" language, the amount of inhibition required on L1 while using an L2 is greater than the inhibition required on an L2 while using the L1.
-This greater initial inhibition on L1 means that switching into L1 incurs a greater ``switch cost" (more cross-language influence) than switching into L2.
+They interpret this asymmetry in terms of the Inhibitory Control Model (ICM) \citep{green1998mental}: to select a phonetic realization from one language, the other must be \textsc{inhibited}.
+Since L1 is the `stronger' language, the amount of inhibition required on L1 while using an L2 is greater than the inhibition required on an L2 while using the L1.
+This greater initial inhibition on L1 means that switching into L1 incurs a greater \textsc{switch cost} (more cross-language influence) than switching into L2.
 Lower switch cost results in a lack of visible cross-language transfer effects on L2.
@@
 Balanced bilinguals did not demonstrate any transfer effects in VOT.
-They propose that this is because balanced bilinguals have better inhibitory control (low switch cost in both languages) due to greater experience with language switching.\\
+They propose that this is because balanced bilinguals have better inhibitory control (low switch cost in both languages) due to greater experience with language switching.
 
 Since differences between long-lag and short-lag languages independently affect the phonetic realization of VOT, focusing on other phonetic features as sites for transfer can avoid this conflation, and clarify the precise effect of language status.
-However, given the implications of the ICM, it is necessary to first establish that these L2 categories can indeed be affected by dynamic interference, particularly in proficient bilinguals.
+However, given the implications of the ICM, it is necessary to first establish that these L2 categories can indeed be affected by dynamic interference, particularly in bilinguals with extensive language-switching experience.
 The few existing studies on vowel quality \citep{simonet2014phonetic,muldner2019phonetics,elias2017effects} or phonological processes \citep{simonet2020increased,schwartz2015language} have examined transfer effects on L1 categories.
@@
 In general, short-lag VOTs seem to resist accommodative shift.
-\cite{bullock2009trying} suggest that because long-lag languages offer a greater range of acceptable VOTs, there is more ``room" for movement.
+\cite{bullock2009trying} suggest that because long-lag languages offer a greater range of acceptable VOTs, there is more `room' for movement.
 In contrast, shift in short-lag languages could risk the loss of a phonological contrast (VOT being the primary cue for voicing contrast in these languages).
@@
 They attribute this to the low frequency of these sounds in Greek, making them less ``stable", and thus more amenable to cross-language influence.
-This suggests that the distribution of sounds in a language could affect the kind of transfer patterns observed.\\
+This suggests that the distribution of sounds in a language could affect the kind of transfer patterns observed.
 
@@
 Asymmetries in the extent and patterns of transfer have not only been observed between languages, but also between different sounds/features of a language, in both long-term and transient interactions.
-Studies which have examined multiple sound categories have found that interactions between individual sounds pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep[e.g.][]{chang2012rapid,elias2017effects}, suggesting that \textit{extent of transfer} cannot treated as an atomic measure.
-In light of the discussion about VOT in section \ref{sound systems}, this is not surprising --- if a general linguistic principle of ``room for movement" constrains transfer, then we should expect it to apply to individual sounds too.
-Once again, this emphasizes the importance of examining a wider range of sound contrasts.\\
+Studies that have examined multiple sound categories have found that interactions between individual sounds pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep[e.g.][]{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot treated as an atomic measure.
+In light of the discussion about VOT in section \ref{sound systems}, this is not surprising --- if a general linguistic principle of `room for movement' constrains transfer, then we should expect it to apply to individual sounds too.
+Once again, this emphasizes the importance of examining a wider range of sound contrasts.
 
@@
 We expect the vowel \nt{2} to shift towards an acoustically/perceptually close category in IE -- a probable candidate being the low vowel \nt{a:}, as anecdotal observations of Bengali-accented English suggest that the sounds are perceived as close by Bengali speakers.
-\\
 
@@
 \cite{olson2013bilingual} argues that code-switching introduces discursive factors such as discourse planning and pragmatic considerations, which ultimately reflect patterns of language practice, rather than cognitive behavior.
-This finds support from studies which report transfer effects \textit{before} the switch point \citep{bullock2009trying}, or other evidence for planning such as hyperarticulation \citep{muldner2019phonetics} and interlocutor-awareness in spontaneous CS \citep{khattab2013phonetic}.
+This finds support from studies which report transfer effects \textsc{before} the switch point \citep{bullock2009trying}, or other evidence for planning such as hyperarticulation \citep{muldner2019phonetics} and interlocutor-awareness in spontaneous CS \citep{khattab2013phonetic}.
 To avoid such confounds, \cite{olson2013bilingual} endorses cued picture-naming (where pictures are named in rapid succession in either language, based on the given language cue) as an alternative paradigm to isolate purely phonetic effects in controlled setups.
 This observation has important implications for how the existing literature on short-term phonetic interaction is interpreted: if the discursive properties of a CS paradigm independently affect the phonetic outcome of mixed-language use, this calls to question the comparability of results across studies using these different paradigms.
-Identifying paradigm-specific differences is thus particularly relevant for this line of research, as generalizations often rely crucially on the assumption of such comparability.
-\\
+Because the reported outcomes of short-term phonetic interaction vary so greatly across studies (cf.
+section \ref{asymmetries}), generalizations about the mechanisms that underlie these patterns often rely crucially on the assumption of such comparability.
+Therefore, identifying paradigm-specific differences is particularly relevant for this line of research.
 
@@
 The participants in Muldner et al.'s \citeyearpar{muldner2019phonetics} study hyperarticulated switched tokens (indicating some degree of planning), and did not evidence a significant shift during mixed-language use.
-Moreover, recall that the findings of \cite{olson2016role} suggest that there is a limit on the extent of transfer-- multiple factors do not necessarily result in additive effects.
+Moreover, recall that the findings of \cite{olson2016role} suggest that there is a limit on the extent of transfer--- multiple factors do not necessarily result in additive effects.
 Thus, the presence of additional factors in CS does not necessitate that this should alter the extent of observable transfer compared to any other paradigm.
-It is possible that the outcomes in the studies discussed above are better attributed to some other aspect of the experimental setup, rather than to the act of code-switching in particular.\\
+It is possible that the outcomes in the studies discussed above are better attributed to some other aspect of the experimental setup, rather than to the act of code-switching in particular.
 
-The precise contribution of various (linguistic and extra-linguistic) factors and their interactions to the phonetics mixed-language use merits detailed investigation.
-In the present study, our focus is limited to verifying whether, in an experimental setting, the postulated discursive elements of a CS paradigm alter the extent of observable transfer compared to a paradigm that arguably lacks these--- cued picture-naming.
+The precise contribution of various (linguistic and extra-linguistic) factors and their interactions to the phonetics mixed-language use merits detailed investigation in its own right.
+In the present study, however, our focus is limited to verifying whether, in an experimental setting, the postulated discursive elements of a CS paradigm alter the extent of observable transfer compared to a paradigm that arguably lacks these--- cued picture-naming.
 We are less concerned with the exact nature of differences than with their comparability.
@@
 Every target word appeared twice during a test block; once in a unilingual and once in a mixed condition.
-Participants alternated between the two tasks, and were allowed to take a short break after each test block.\\
-
+Participants alternated between the two tasks, and were allowed to take a short break after each test block.
 The tasks are described below:
@@
 \begin{enumerate}[]
-	\item Cued picture-naming: Participants were presented with slides in the following sequence-
+	\item Cued picture-naming: Participants were presented with slides in the following sequence:
 	\begin{itemize}
@@
 
-	o-\textipa{\:t}a \quad \textipa{\ae}k-\textipa{\:t}a \quad kalo [\textit{target word}]\\
+o-\textipa{\:t}a \qquad \textipa{\ae}k-\textipa{\:t}a \quad kalo [\textit{target word}]\\
 that-\textsc{clf}  one-\textsc{clf}  black [\textit{target word}]\\
-	`that is a black [\textit{target word}]'\\
+``that is a black [\textit{target word}]"\\
 
@@
 
-%Figures ... show the findings of the experiment. 
 All statistical analyses were performed in R \citep{r}.
@@
 The independent variables or fixed factors, along with their levels, are given in table \ref{table variables}.
-Subject and Item were treated as random factors.
+Subject and Item were treated as random factors.\\
 
-\begin{table}[h]
-	\caption{Fixed factors}
-	\centering
-	\resizebox{\textwidth}{!}{%
-		\begin{tabular}{@{}ll@{}}
-			\toprule
-			Factor                                                 & Levels                                                            \\ \midrule
-			Gender                                                 & male, female                                                      \\
-			Vowel                                                  & \nt{2}, \nt{\ae} \\
-			Task (corresponding to the two paradigms)              & p (Picture: cued picture-naming, s (Sentence: code-switching)     \\
-			Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)                       \\ \bottomrule
-		\end{tabular}%
-	}
-\label{table variables}
-\end{table}
+< Insert Table \ref{table variables} about here > \\
 
@@
 This improved the model fit ($\chi^2(5) = 11.43, p = .04342$).
-Thus, we report on a model that includes Gender, Task, Context, Vowel, Context*Vowel as fixed effects, random intercepts for Subject and Item, by-subject random slopes for Task, Context, Vowel and Context*Vowel, and by-item random slopes for Task and Context.\\
+Thus, we report on a model that includes Gender, Task, Context, Vowel, Context*Vowel as fixed effects, random intercepts for Subject and Item, by-subject random slopes for Task, Context, Vowel and Context*Vowel, and by-item random slopes for Task and Context.
 
-The following sections discuss the results for vowel height and backness respectively:
+Figure \ref{vowels_e_b} summarizes the findings of the experiment ($F1XF2$ plots for vowel productions in the unilingual and mixed conditions, in comparison to the related Bengali categories from \cite{dutta2021}).
+The following sections discuss the results for vowel height and backness respectively.
+
+< Insert Figure \ref{vowels_e_b} about here >
 
@@
 This is surprising, given that the phonological category is expected to systematically affect vowel quality.
-However, a boxplot visualization (Figure\ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two categories \nt{2, \ae} and the extent of height difference between them.
+However, a boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two categories \nt{2, \ae} and the extent of height difference between them.
 Note that these measures are only from the English (unilingual) condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching.
@@
 We find that the difference is not significant at 5\%, and increases over the course of articulating the vowel.
+\\
 %ANOVA table add
 
-\begin{figure}[h] \centering
+< Insert Table \ref{table_f1_fixed_effects} about here >\\
 
-	\includegraphics[scale=0.8]{vowel_by_subject_ggplot}
-	\caption{Subject-wise mean F1; unilingual English}
-	\label{boxplot_F1}
-\end{figure}
+< Insert Figure \ref{boxplot_F1} about here >\\
 
-\begin{table}[htbp] \centering 
+\subsection{Vowel backness -- F2}
+
+Table \ref{table_f2_fixed_effects} summarizes the results of fixed effects from the model for F2 at each point.
+Once again, we see that male speakers have a lower mean F2 than female speakers, as expected.
+The estimates for Vowel show that excepting the beginning of the vowel, the category \nt{2} is consistently produced as more backed than \nt{\ae}.
+Unlike vowel height, Context has an independent effect on vowel backness, such that both the vowels in this study have higher F2 values in the Bengali (mixed) condition, i.e.
+they are fronted in the mixed condition.
+A significant interaction between Vowel and Context reveals an asymmetry in the extent of fronting between the two vowels -- the shift is greater in the case of \nt{\ae}.
+We see a significant effect of Task on F2, such that vowels in the Sentence task (code-switching) were more backed than those in the picture-naming task.
+As with F1, this is an unexpected result, as we have no principled reason to expect code-switching to induce an across-board backing of vowels.
+A lack of significant interaction with Context suggests that there is no difference in the extent of shift between the two paradigms.
+The results of ANOVAs comparing the full model to a null model that lacks the Vowel*Context interaction factor reveal that as with vowel height, the change in vowel backness between the two conditions is not significant at 5\%, and increases over the course of the vowel.\\
+
+< Insert Table \ref{table_f2_fixed_effects} about here >
+
+\section{Discussion}
+
+With regard to our first research question \textbf{L2 vowels}, the results confirm that the L2 vowels of proficient bilinguals can be affected by mixed-language processing.
+Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.
+In the context of an ICM--based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, this is a significant finding: it demonstrates that the postulated lower switch cost on an L2 does not entirely mask transfer effects on L2 vowels.
+Moreover, our findings differ from \cite{tsui2019impact} who report on the basis of VOT measurements that balanced bilinguals do not evidence an effect of language-switching.
+\citeauthor{tsui2019impact} attribute the absence of transfer effects to greater experience with language switching in these participants, leading to lower switch-costs in both languages.
+We find that even proficient bilinguals with extensive experience of language switching demonstrate cross-language transfer effects in vowel production.
+Thus, following from the discussion in sec.
+\ref{asymmetries}, it might be fruitful for future studies to focus on vowels as a potential site for transfer.
+This would both expand the empirical scope of the existing research (cf.
+sec.
+\ref{asymmetry between sounds}) and avoid confounding factors arising from independent differences between long- and short-lag VOT languages.
+
+We hypothesized an asymmetrical shift between the two vowels (\textbf{Asymmetry}), owing to differences in (i) their positions in the Indian English vowel space, and (ii) the presence/absence of a corresponding category in the L1 Bengali inventory.
+The significant interaction between Vowel and Context in both F1 and F2 confirm that the experimental condition affected both height and backness differently for each vowel.
+As expected, we find a greater degree of shift in \nt{2}, possibly owing to its position in a less dense part of the IE vowel space.
+The category \nt{2} is lowered and fronted in the mixed language condition as expected (Figure \ref{vowels_e_b}).
+The direction of shift in \nt{\ae} is surprising-- it is fronted in the mixed-language condition.
+While phonological constraints on IE suggested that a shift in the F1 dimension could potentially risk the loss of contrast and is thus expected to be minimal, Figure \ref{vowels_e_b} shows that \nt{\ae} in Bengali is not fronted with respect to the English category.
+Thus, fronting in the mixed condition cannot be explained in relation the former, at least from the present data.
+Recall that the Bengali vowel productions used here are from a different set of speakers than our participants.
+Assuming some amount of individual differences in the precise phonetic realization of vowel categories in both languages, this could be a potential source of explanation for this result.
+Data from a larger number of speakers for both languages, as well as within-speaker comparisons of English and Bengali vowels, are required to clarify this.
+
+Our third research question \textbf{Paradigm} concerned the effect of experimental paradigm on the phonetic outcome of mixed-language use.
+\cite{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
+If such discursive properties independently affect the phonetic outcome of language mixing in an experimental setup, this calls to question the comparability of results across studies using these different paradigms.
+We verified this by comparing productions of the same items from the same group of participants in a scripted CS paradigm and a cued picture-naming paradigm.
+We found no significant interaction of Task either with Context, or with the Vowel*Context interaction, suggesting that the extent of phonetic transfer was not affected by the paradigm.
+There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in observable transfer at all.
+This would mean that the pre-switch effects observed, for example, in \cite{muldner2019phonetics} and \cite{bullock2009trying} are better interpreted as resulting from some other aspect of the experimental setup, than the act of code-switching itself.
+While planning could additionally affect the words preceding the switch, the effects on the switched token itself do not change.
+(ii) Our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
+Thus, another possible explanation is that the discursive effects discussed by \cite{olson2013bilingual} are triggered by more naturalistic paradigms where attention to meaning and interlocutor are important.
+More within-study comparisons with less structured paradigms, and spontaneous code-switching data, can clarify this.
+The results of the present study indicate that for the purposes of observing dynamic phonetic transfer in an experimental setup, elicitation through cued-picture-naming vs (controlled) code-switching does not significantly alter the behavior under study.
+Thus, the choice of paradigm can be based on other considerations.
+Moreover, this suggests that (all else being equal) the results from such studies can be compared directly.
+
+The paradigm did affect vowel quality, but this applied uniformly across experimental conditions: all vowels were more backed and lowered in the code-switching task.
+Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this lowering is an artifact of the particular stimuli used here---resulting from some corarticulatory effect that was absent in the picture-naming task.
+While this is intriguing, it is beyond the scope of the present study.
+\\
+
+\section{Conclusion}
+This study examined the phonetic outcome of mixed-language production in a group of highly proficient bilingual speakers of Bengali and Indian English.
+Specifically, we examined the spectral quality of two L2 vowel categories \nt{2} and \nt{\ae} produced in two experimental conditions: unilingual (English) and mixed (switching from Bengali to English).
+Comparisons of F1 and F2 reveal that mixed language processing temporarily increases phonetic interaction between the languages, causing \nt{2} to converge towards a related L1 category \nt{a:}.
+The extent and direction of such shift is asymmetrical across the two vowel categories, which parallels findings in prior studies.
+This asymmetry appears to be mediated by the language-specific distribution of vowel categories.
+Mixed productions were elicited in two switching paradigms: code-switching and cued picture-naming, to assess whether the postulated discursive elements in code-switching independently influence the outcome of phonetic transfer.
+Results suggest that in an experimental setup, elicitation through cued-picture-naming vs (controlled) code-switching does not significantly alter the behavior under study.
+
+This study has a number of limitations: (i) the `baseline' values for Bengali vowel categories were estimated from a different set of speakers than the participants of this study, potentially obscuring the role of individual differences in the realization of vowel categories; (ii) we used a frame sentence to elicit code-switched productions such that not all the sentences were pragmatically well-formed.
+This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study.
+Future work should address this by considering data from more naturalistic code-switching paradigms.
+
+\section*{Abbreviations}
+\textsc{clf} = classifier
+
+\section*{Data availability}
+The data files and analysis code are openly available at:  \url{https://osf.io/dsb2x/?view_only=5cb821070bc74a038a293239157c3b11}
+
+\section*{Competing interests}
+The authors declare none.
+
+%\printbibliography %for use with biblatex; comment out if you use natbib
+\bibliography{references} %for use with natbib; comment out if you use biblatex, and change 'sample' by the name of your bib-file
+
+\newpage
+\section*{Tables}
+\begin{table}[h]
+	\caption{Fixed factors}
+	%\centering
+	\resizebox{\textwidth}{!}{%
+		\begin{tabular}{@{}ll@{}}
+			\toprule
+			Factor                                                 & Levels                                                            \\ \midrule
+			Gender                                                 & male, female                                                      \\
+			Vowel                                                  & \nt{2}, \nt{\ae} \\
+			Task (corresponding to the two paradigms)              & p (Picture: cued picture-naming, s (Sentence: code-switching)     \\
+			Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)                       \\ \bottomrule
+		\end{tabular}%
+	}
+	\label{table variables}
+\end{table}
+
+\newpage
+\begin{table}[htbp] %\centering 
 	\caption{Model outputs for F1 at each of the five points in the vowel} 
@@
 
-\subsection{Vowel backness -- F2}
+\newpage
 
-Table \ref{table_F2_fixed_effects} summarizes the results of fixed effects from the model for F2 at each point.
-Once again, we see that male speakers have a lower mean F2 than female speakers, as expected.
-The estimates for Vowel show that excepting the beginning of the vowel, the category \nt{2} is consistently produced as more backed than \nt{\ae}.
-Unlike vowel height, Context has an independent effect on vowel backness, such that both the vowels in this study have higher F2 values in the Bengali (mixed) condition, i.e.
-they are fronted in the mixed condition.
-A significant interaction between Vowel and Context reveals an asymmetry in the extent of fronting between the two vowels -- the shift is greater in the case of \nt{\ae}.
-We see a significant effect of Task on F2, such that vowels in the Sentence task (code-switching) were more backed than those in the picture-naming task.
-As with F1, this is an unexpected result, as we have no principled reason to expect code-switching to induce an across-board backing of vowels.
-A lack of significant interaction with Context suggests that there is no difference in the extent of shift between the two paradigms.
-The results of ANOVAs comparing the full model to a null model that lacks the Vowel*Context interaction factor reveal that as with vowel height, the change in vowel backness between the two conditions is not significant at 5\%, and increases over the course of the vowel.
-
-\begin{table}[htbp] \centering 
+\begin{table}[htbp] %\centering 
 	\caption{Model outputs for F2 at each of the five points in the vowel} 
-	\label{table_F2_fixed_effects} 
+	\label{table_f2_fixed_effects} 
 	\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
@@
 
-\section{Discussion}
+\newpage
 
-With regard to our first research question \textbf{L2 vowels}, the results confirm that the L2 vowels of proficient bilinguals can be affected by mixed-language processing.
-Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.
-In the context of an ICM--based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, this is a significant finding: it demonstrates that the postulated lower switch cost on an L2 does not entirely mask transfer effects on L2 vowels.
-Moreover, our findings differ from \cite{tsui2019impact} who report on the basis of VOT measurements that balanced bilinguals do not evidence an effect of language-switching.
-The authors attribute this to greater experience with language switching, leading to lower switch-costs in both languages.
-We find that even proficient bilinguals with extensive experience of language-switching demonstrate cross-language transfer effects in vowel production.
-Thus, following from the discussion in sec.
-\ref{asymmetries}, it might be fruitful for future studies to focus on vowels as a potential site for transfer.
-This would both expand the scope of the existing research (cf.
-sec.
-\ref{asymmetry between sounds}) and avoid confounding factors arising from independent differences between long- and short-lag VOT languages.\\
+\section*{Figures}
 
-We hypothesized an asymmetrical shift between the two vowels (\textbf{Asymmetry}), owing to differences in (i) their positions in the Indian English vowel space, and (ii) the presence/absence of a corresponding category in the L1 Bengali inventory.
-The significant interaction between Vowel and Context in both F1 and F2 confirm that the experimental condition affected both height and backness differently for each vowel.
-As expected, we find a greater degree of shift in \nt{2}, possibly owing to its position in a less dense part of the IE vowel space.
-Figure \ref{vowels_e_b} shows $F1XF2$ plots for vowel productions in the unilingual and mixed (dotted ellipses) conditions, in comparison to the related Bengali categories (colored ellipses) from \cite{dutta2021}.
-The category \nt{2} is lowered and fronted in the mixed language condition as expected.
-The direction of shift in \nt{\ae} is surprising-- it is fronted in the mixed-language condition.
-While phonological constraints on IE suggested that a shift in the F1 dimension could potentially risk the loss of contrast and is thus expected to be minimal, Figure \ref{vowels_e_b} shows that \nt{\ae} in Bengali is not fronted with respect to the English category.
-Thus, fronting in the mixed condition cannot be explained in relation the former, at least from the present data.
-Recall that the Bengali vowel productions used here are from a different set of speakers than our participants.
-Assuming some amount of individual differences in the precise phonetic realization of vowel categories in both languages, this could be a potential source of explanation for this result.
-Data from a larger number of speakers for both languages, as well as within-speaker comparisons of English and Bengali vowels, are required to clarify this.
-\\
-
-\begin{figure}[h] \centering
-	\includegraphics[scale=0.8]{vowels_e_b_final}
-	\caption{Vowel categories in unilinual English (solid lines), mixed condition (dotted lines), and baseline Bengali (colored ellipses)}
+\begin{figure}[h] %\centering
+	\includegraphics[scale=1]{vowels_e_b_final}
+	\caption{Vowel quality in unilinual English (solid lines), mixed condition (dotted lines), and baseline Bengali (colored ellipses) productions}
 	\label{vowels_e_b}
@@
 
-Our third research question \textbf{Paradigm} concerned the effect of experimental paradigm on the phonetic outcome of mixed-language use.
-\cite{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
-If such discursive properties independently affect the phonetic outcome of language mixing in an experimental setup, this calls to question the comparability of results across studies using these different paradigms.
-We verified this by comparing productions of the same items from the same group of participants in a scripted CS paradigm and a cued picture-naming paradigm.
-We found no significant interaction of Task either with Context, or with the Vowel*Context interaction, suggesting that the extent of phonetic transfer was not affected by the paradigm.
-There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in observable transfer at all.
-This would mean that the pre-switch effects observed, for example, in \cite{muldner2019phonetics} and \cite{bullock2009trying} are better interpreted as resulting from some other aspect of the experimental setup, than the act of code-switching itself.
-While planning could additionally affect the words preceding the switch, the effects on the switched token itself do not change.
-(ii) Our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
-Thus, another possible explanation is that the discursive effects discussed by \cite{olson2013bilingual} are triggered by more naturalistic paradigms where attention to meaning and interlocutor are important.
-More within-study comparisons with less structured paradigms, and spontaneous code-switching data, can clarify this.
-The results of the present study indicate that for the purposes of observing dynamic phonetic transfer in an experimental setup, elicitation through cued-picture-naming vs (controlled) code-switching does not significantly alter the behavior under study.
-Thus, the choice of paradigm can be based on other considerations.
-Moreover, this suggests that (all else being equal) the results from such studies can be compared directly.
+\newpage
 
-The paradigm did affect vowel quality, but this applied uniformly across experimental conditions: all vowels were more backed and lowered in the code-switching task.
-Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this lowering is an artifact of the particular stimuli used here---resulting from some corarticulatory effect that was absent in the picture-naming task.
-While this is intriguing, it is beyond the scope of the present study.
-\\
+\begin{figure}[h] %\centering
 
-\section{Conclusion}
-This study examined the phonetic outcome of mixed-language production in a group of highly proficient bilingual speakers of Bengali and Indian English.
-Specifically, we examined the spectral quality of two L2 vowel categories \nt{2} and \nt{\ae} produced in two experimental conditions: unilingual (English) and mixed (switching from Bengali to English).
-Comparisons of F1 and F2 reveal that mixed language processing temporarily increases phonetic interaction between the languages, causing \nt{2} to converge towards a related L1 category \nt{a:}.
-The extent and direction of such shift is asymmetrical across the two categories, which parallels findings in prior studies.
-This asymmetry appears to be mediated by the language-specific distribution of vowel categories.
-Mixed productions were elicited in two switching paradigms: code-switching and cued picture-naming, to assess whether the postulated discursive elements in code-switching independently influence the outcome of phonetic transfer.
-Results suggest that in an experimental setup, elicitation through cued-picture-naming vs (controlled) code-switching does not significantly alter the behavior under study.
-\\
-
-This study has a number of limitations: (i) the `baseline' values for Bengali vowel categories were estimated from a different set of speakers than the participants of this study, potentially obscuring the role of individual differences in the realization of vowel categories; (ii) we used a frame sentence to elicit code-switched productions such that not all the sentences were pragmatically well-formed.
-This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study.
-Future work should address this by considering data from more naturalistic code-switching paradigms.
-
-\section*{Abbreviations}
-\textsc{clf} = classifier
-
-\section*{Data availability}
-The data files and analysis code are openly available at:  \url{https://osf.io/dsb2x/?view_only=5cb821070bc74a038a293239157c3b11}
-
-\section*{Competing interests}
-The authors have no competing interests to declare.
-
-%\printbibliography %for use with biblatex; comment out if you use natbib
-\bibliography{references} %for use with natbib; comment out if you use biblatex, and change 'sample' by the name of your bib-file
+	\includegraphics[scale=1]{vowel_by_subject_ggplot}
+	\caption{Subject-wise mean F1; unilingual English}
+	\label{boxplot_F1}
+\end{figure}
 

### 2021-10-19 draft_BLC/BLC_AM-ID_manuscript.tex
BLC reviewer comments

@@
 
-EFL University
+EFL University,
 
-Hyderabad\\
+Hyderabad,
+
+India\\
 
@@
 
-School of languages and Linguistics
+Jadavpur University,
 
-Jadavpur University
+Kolkata,
 
-Kolkata\\
+India\\
 
@@
 This study investigates the phonetic outcome of mixed-language speech in Bengali and Indian English, towards understanding cross-language interaction in highly proficient bilinguals.
-We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in unilingual vs mixed productions.
+We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in unilingual vs.
+mixed productions.
 Results reveal dynamic and asymmetrical changes in both vowels;  \nt{2} shifting towards a related L1 category \nt{a:}.
@@
 Results reveal no difference across paradigms.
-We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction between sound systems and the postulated role of connected speech in introducing discursive factors to transfer studies.\\
+We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction, postulated discursive factors in code-switching, and the issue of comparability between paradigms in transfer studies.\\
 
@@
 The rest of this section is structured as follows: \ref{bengali_english_in_india} provides background information on Bengali and Indian English focusing on vowel systems.
-\ref{causes} discusses two potential sources of transfer during language processing-- global co-activation of multiple languages vs the local act of switching between languages.
+\ref{causes} discusses two potential sources of transfer during language processing-- global co-activation of multiple languages vs.
+the local act of switching between languages.
 Given the highly multilingual setting that characterizes this population, we note that the former is expected to be a constant feature and therefore any observed phonetic interaction is better understood as resulting from the latter.
@@
 A logical consequence of this model is that such phonetic interaction must be highly localized.
-\ref{duration} elaborates on this by reviewing literature on the duration and `reversiblity' of transfer effects.
+\ref{duration} elaborates on this by reviewing literature on the duration and `reversibility' of transfer effects.
 A recurring pattern that emerges from existing research on short-term phonetic transfer is that both languages of the bilingual speaker are not equally affected, although the nature of these reported differences varies greatly across studies.
@@
 Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during accessing, processing, or articulation of these categories (cf.
-transfer vs interference \citep{grosjean2012attempt}; competence vs performance interference \citep{paradis1993linguistic}).
+transfer vs.
+interference \citep{grosjean2012attempt}; competence vs.
+performance interference \citep{paradis1993linguistic}).
 Given their transient nature, dynamic changes in production during mixed-language use are generally attributed to the latter, e.g.
@@
 
-\subsubsection*{L1 vs L2 status} 
+\subsubsection*{L1 vs. L2 status} 
 Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} posits that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another.
@@
 We provide a starting point for such comparison by eliciting productions of the same tokens from the same set of participants, in both paradigms.
-A difference in results will suggest that the planning involved in connected speech affects transfer above and beyond the phonetic outcome of switching itself, which has implications that future work should consider.
+A difference in results will suggest that the planning involved in connected speech and code-switching affects transfer above and beyond the phonetic outcome of switching itself, which has implications that future work should consider.
 
@@
 \nt{\ae} is expected to show lowering, but to a lesser extent than \nt{2}, as it is in a part of the IE vowel space that has greater vowel density.
-	\item \textbf{Paradigm}: Is there an observable difference between the extent of transfer in a cued picture-naming vs a code-switching paradigm? 
+	\item \textbf{Paradigm}: Is there an observable difference between the extent of transfer in a cued picture-naming vs. a code-switching paradigm? 
 \end{itemize}
@@
 At the time of recording, they were living in or around the EFL University campus in Hyderabad, which has a linguistically diverse population.
-Speakers were selected on the basis of responses to a Language Background Questionnaire to ensure comparable LSRW skills in L1 and L2 across participants, and were compensated for their time.
+Speakers were selected on the basis of responses to a Language Background Questionnaire to ensure comparable LSRW (listening, speaking, reading, writing) skills in L1 and L2 across participants, and were compensated for their time.
 
 \subsection{Stimuli}
-Twenty monosyllabic English words containing the vowels \nt{2} or \nt{ae} were selected as target words.
+Twenty monosyllabic English words containing the vowels \nt{2} or \nt{\ae} were selected as target words.
 To minimize coarticulatory effects on the vowel, the pre-vocalic consonants were all voiced plosives (\nt{b} or \nt{d}).
@@
 	\end{itemize}
-	Since all the target words were in English, the language cue was used to manipulate the \textit{preceding} word, giving target words in non-switched (unilingual) and switched (mixed) utterances.
+	Since all the target words were in English, the language cue was used to manipulate the \textsc{preceding} word, giving target words in non-switched (unilingual) and switched (mixed) utterances.
 The orthographies of English and Bengali are visually distinct, and were therefore used to cue language.
@@
 An additional 67 tokens (2.79\%) were excluded due to errors in presentation/production, giving 2333 target word tokens for analysis.
-Recordings were manually segmented and annotated in Praat \citep{boersma2016praat}, and the first and second formant frequencies (F1 and F2) of the target vowels \nt{2, ae} were measured at five points-- in 10\% increments starting at 5\% into the vowel.
+Recordings were manually segmented and annotated in Praat \citep{boersma2016praat}, and the first and second formant frequencies (F1 and F2) of the target vowels \nt{2, \ae} were measured at five points-- in 10\% increments starting at 5\% into the vowel.
 This gave measures of F1 and F2 at 5\%, 15\%, 25\%, 35\% and 45\% into the vowel, allowing us to observe any dynamic changes in the extent of transfer over the course of producing the vowel.
@@
 Vowel quality and gender are expected to systematically affect F1.
-Thus, we began with a model that had Vowel and Gender as fixed effects, random intercepts for Subject and Item, and random slopes for each within-unit variable that was relevant to our hypotheses-- by-subject slopes for Task, Context, and Vowel; by-item slopes for Task and Context.
+Thus, we began with a model that had Vowel and Gender as fixed effects, random intercepts for Subject and Item, and random slopes for each within-unit variable that was relevant to our hypotheses: by-subject slopes for Task, Context, and Vowel; by-item slopes for Task and Context.
 Next, adding Context as a fixed effect significantly improved the model fit ($\chi^2(1) = 5.54, p= .01854$), suggesting that the experimental condition affects vowel height.
@@
 \ref{asymmetries}, it might be fruitful for future studies to focus on vowels as a potential site for transfer.
-This would both expand the empirical scope of the existing research (cf.
+This would both expand the empirical scope of existing research (cf.
 sec.
 \ref{asymmetry between sounds}) and avoid confounding factors arising from independent differences between long- and short-lag VOT languages.
+More generally, these findings point to a need for expanding the range of features in which transfer effects are studied, in order to make robust generalizations about the mechanisms underlying observed patterns.
 
@@
 More within-study comparisons with less structured paradigms, and spontaneous code-switching data, can clarify this.
-The results of the present study indicate that for the purposes of observing dynamic phonetic transfer in an experimental setup, elicitation through cued-picture-naming vs (controlled) code-switching does not significantly alter the behavior under study.
+The results of the present study indicate that for the purposes of observing dynamic phonetic transfer in an experimental setup, elicitation through cued-picture-naming vs.
+(controlled) code-switching does not significantly alter the behavior under study.
 Thus, the choice of paradigm can be based on other considerations.
@@
 Comparisons of F1 and F2 reveal that mixed language processing temporarily increases phonetic interaction between the languages, causing \nt{2} to converge towards a related L1 category \nt{a:}.
-The extent and direction of such shift is asymmetrical across the two vowel categories, which parallels findings in prior studies.
+The extent and direction of shift is asymmetrical across the two vowel categories, which parallels findings in prior studies.
 This asymmetry appears to be mediated by the language-specific distribution of vowel categories.
-Mixed productions were elicited in two switching paradigms: code-switching and cued picture-naming, to assess whether the postulated discursive elements in code-switching independently influence the outcome of phonetic transfer.
-Results suggest that in an experimental setup, elicitation through cued-picture-naming vs (controlled) code-switching does not significantly alter the behavior under study.
+These findings evidence an effect of phonetic transfer on L2 vowels for the first time, and raise questions for proposed inhibition-based accounts of cross-language phonetic interaction.
+Mixed productions were elicited in two switching paradigms: code-switching and cued picture-naming, to verify whether the postulated discursive elements in code-switching independently influence the outcome of phonetic transfer.
+Results suggest that in an experimental setup, elicitation through cued-picture-naming vs.
+(controlled) code-switching does not significantly alter the behavior under study.
 
-This study has a number of limitations: (i) the `baseline' values for Bengali vowel categories were estimated from a different set of speakers than the participants of this study, potentially obscuring the role of individual differences in the realization of vowel categories; (ii) we used a frame sentence to elicit code-switched productions such that not all the sentences were pragmatically well-formed.
+This study has a number of limitations: (i) the `baseline' values for Bengali vowel categories were estimated from a different set of speakers than the participants of this study, potentially obscuring the role of individual differences in the realization of vowel categories; within-speaker comparisons of L1 and L2 vowels is likely to offer a more fine-grained view of the direction of shift in relation to L1 categories.
+(ii) we used a frame sentence to elicit code-switched productions such that not all the sentences were pragmatically well-formed.
 This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study.
@@
 \section*{Data availability}
-The data files and analysis code are openly available at:  \url{https://osf.io/dsb2x/?view_only=5cb821070bc74a038a293239157c3b11}
+The materials, data files, and analysis code for this study are openly available in OSF at:  \url{https://doi.org/10.17605/OSF.IO/DSB2X}.
 
@@
 	\includegraphics[scale=1]{vowels_e_b_final}
-	\caption{Vowel quality in unilinual English (solid lines), mixed condition (dotted lines), and baseline Bengali (colored ellipses) productions}
+	\caption{Vowel quality in unilinual English, mixed, and baseline Bengali productions}
 	\label{vowels_e_b}
@@
 	\includegraphics[scale=1]{vowel_by_subject_ggplot}
-	\caption{Subject-wise mean F1; unilingual English}
+	\caption{Subject-wise mean F1 in unilingual English productions}
 	\label{boxplot_F1}

### 2022-02-01 RCode/transfer_analysis.Rmd
revisions: data in long format, Rmd file

---
title: "Phonetic Tansfer analysis code"
author: "Auromita"
date: "1/23/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
```

Load packages and data, create objects for lobanov normalized F1 and F2, create dataframe.

(note: file paths relative to the .Rmd file, not the global R environment)

```{r, include=FALSE}

library(tidyverse)
library(phonR)
library(lme4)
library(lmerTest)
library(stargazer)
library(pwr)

df <- read.csv("formant_data_5.csv", check.names = TRUE) %>%
  rename_all(tolower) %>%
  mutate_if(is.character,as.factor)

str(df)

attach(df)

# pool measures from all 5 points in the vowel, then from that-- calculate the average F1 for each speaker, and normalize all data points from that speaker wrt that.

normF1 <- with(df,normLobanov(cbind(f1_5,f1_15,f1_25,f1_35,f1_45)),group=subject)
normF2 <- with(df,normLobanov(cbind(f2_5,f2_15,f2_25,f2_35,f2_45)),group=subject)

df_lobanov <- data.frame(subject,gender,task,word,context,vowel,normF1,normF2)

detach(df)

rm(df,normF1,normF2)

# data frame with time as variable
df_time <- df_lobanov %>%
  rename(f1_1=f1_5, f1_2=f1_15, f1_3=f1_25, f1_4=f1_35, f1_5=f1_45, f2_1=f2_5, f2_2=f2_15, f2_3=f2_25, f2_4=f2_35, f2_5=f2_45) %>%
  # dplyr pivot() functions: convert data between long(indexed) and wide(cartesian) formats
  pivot_longer(starts_with("f"),
               names_to = c(".value", "time"),
               names_sep = "_",
               names_transform = list(time = as.factor)
               )
# Since there are multiple observations per row (F1 and F2) that need to go in separate columns in the output, the ".value" lets you specify that part of the variable name will be used to split the data in the output. The rest of the name becomes the variable "time"

```

LME models:

```{r}

## main model

phon_acco.model = lmer(f1 ~ 
                         gender + context*vowel + time*vowel + task +
                         (1|word) + (1+context*vowel|subject), 
                         data= df_time, REML=FALSE)
summary(phon_acco.model)
summary(rePCA(phon_acco.model))

null.model = lmer(f1 ~ 
                    gender + context*vowel + task + 
                    (1|word) + (1+context*vowel|subject), 
                    data=df_time, REML=FALSE)
summary(null.model)

anova(phon_acco.model,null.model)

coef(phon_acco.model)

```

### 2022-02-01 RCode/bengali_phon_acco_new.R

@@ -1,32 +1,35 @@
-<<<<<<< HEAD
-formant_data_5 <- read.csv("formant_data_5.csv")
-=======
-formant_data_5 <- read.csv("formant_data_5.csv", check.names = TRUE)
->>>>>>> 093ce24fb2b2845200c064d597ff33aaf389134e
-attach(formant_data_5)
-# detach(formant_data_5)
+# analysis code for phonetic_transfer
+# file paths relative to R project environment (root = phonetic_transfer)
 
-library(lattice)
-library(plyr)
+library(tidyverse)
 library(phonR)
 library(lme4)
-library(ggplot2)
-library(dplyr)
-library(tidyr)
+library(lmerTest)
 library(stargazer)
 
 
+formant_data_5 <- read.csv("RCode/formant_data_5.csv", check.names = TRUE)%>%
+  mutate_if(is.character,as.factor)
+str(formant_data_5)
+
+attach(formant_data_5)
+
+
+
+# pool measures from all 5 points in the vowel, then from that-- calculate the average F1 for each speaker, and normalize all data points from that speaker wrt that.  
 normF1 <- with(formant_data_5,normLobanov(cbind(F1_5,F1_15,F1_25,F1_35,F1_45)),group=Subject)
-normF1 <- as.data.frame(normF1)
 
 normF2 <- with(formant_data_5,normLobanov(cbind(F2_5,F2_15,F2_25,F2_35,F2_45)),group=Subject)
-normF2 <- as.data.frame(normF2)
 
 
-with(formant_data_5, plotVowels(normF1$F1_15, normF2$F2_15, Vowel, group = Context, plot.tokens = FALSE, 
-              plot.means = TRUE , var.sty.by = Vowel,var.col.by = Context, pch.means = Vowel, cex.tokens = 0.8,
-              cex.means = 0.8, alpha.means = 0.3,alpha.tokens = 0.3,ellipse.line = TRUE,  ellipse.fill = TRUE,
-              fill.opacity = 0.1, pretty = TRUE))
+df_lobanov <- data.frame(Subject,Gender,Task,Word,Context,Vowel,normF1,normF2)
+
+
+detach(formant_data_5)
+# with(formant_data_5, plotVowels(normF1$F1_15, normF2$F2_15, Vowel, group = Context, plot.tokens = FALSE, 
[…]

### 2022-02-20 RCode/transfer_analysis.Rmd
analysis log-mean normalization, rmd file

@@
 
-(note: file paths relative to the .Rmd file, not the global R environment)
+(note: file paths relative to project root folder (phonetic_transfer))
 
@@
 library(pwr)
+library(smir)
 
-df <- read.csv("formant_data_5.csv", check.names = TRUE) %>%
+df <- read.csv("RCode/formant_data_5.csv", check.names = TRUE) %>%
   rename_all(tolower) %>%
@@
 
-rm(df,normF1,normF2)
+rm(normF1,normF2)
 
 # data frame with time as variable
-df_time <- df_lobanov %>%
+df_lobanov_time <- df_lobanov %>%
   rename(f1_1=f1_5, f1_2=f1_15, f1_3=f1_25, f1_4=f1_35, f1_5=f1_45, f2_1=f2_5, f2_2=f2_15, f2_3=f2_25, f2_4=f2_35, f2_5=f2_45) %>%
@@
 
+rm(df_lobanov)
+```
+
+Log-mean normalization:
+
+```{r}
+
+# data in long format for log-mean normalization 
+df_long <- df %>%
+  select(!c(vowel_duration, f1_mid)) %>%
+  rename(f1_1=f1_5, f1_2=f1_15, f1_3=f1_25, f1_4=f1_35, f1_5=f1_45, f2_1=f2_5, f2_2=f2_15, f2_3=f2_25, f2_4=f2_35, f2_5=f2_45, iteration=filename) %>%
+  pivot_longer(starts_with("f"), # pivot all the columns whose names start with "f"
+               values_to = "ff_raw",
+               # drop the f part, store the rest as values for new variables "formant" and "time"
+               names_to = c(NA, "formant", "time"), 
+               # how to divide up the variable name (f + formant + _ + time)
+               names_pattern = "(.)(.)_(.)") %>% 
+  # reorder columns
+  relocate("subject", "gender", "task", "word", "context", "vowel", "time", "formant", "ff_raw") %>%
+  mutate_if(is.character, as.factor) %>%
+  # new column with log-transformed FFs
+  mutate(ff_log = log(ff_raw)) %>%
+  # composite variable vowel-formant (estimate for the dialectal reference pattern-- the target ff for that vowel in the dialect)
+  mutate (vowel.formant = factor(interaction(vowel, formant)))
+
+# fitting a regression model to predict log-transformed formant-frequencies (ff-log) for a given vowel, formant and speaker as the sum of a speaker displacement term (subject) and normalized vowel-formant effects (vowel.formant)  (Barreda&Neary 2018)
+
+### only vowel midpoints for figures
+df_long <- df_long %>%
+  filter(time==5)
+
+attach(df_long)
+
+# forcing the intercept to 0 so that coefficients are calculated for all levels of "subject"
+M = lm (ff_log ~ 0 + subject + vowel.formant, contrasts = list(vowel.formant = contr.sum))
+summary(M)
+
+# coefficients of the speaker effect (estimate of speaker parameter)
+speaker_coeff = dummy.coef(M)$subject
+# coefficients of the vowel.formant effect (estimate of dialect reference/target formant for each vowel)
+vowel_coeff = dummy.coef(M)$vowel.formant
+
+detach(df_long)
+
+# from each log-transformed ff of a particular speaker, want to subtract the speaker coefficient
+
+temp <- as.data.frame(speaker_coeff) %>%
+  rownames_to_column(var = "subject")
+
+df_long <- df_long %>%
+  full_join(temp, by = "subject")
+
+df_long <- df_long %>%
+  mutate(ff_log_normalized = ff_log - speaker_coeff)
+
+df_long <- df_long %>%
+  mutate(ff_exp_normalized = exp(ff_log_normalized))
+
+rm(temp,speaker_coeff,vowel_coeff)
+
+# putting F1 and F2 in separate columns for plotting (pivot_wider doesn't work because there are no unique identifiers for each vowel, since there were multiple reps in each condition).
+So crude workaround:
+
+df_clean <- df_long %>%
+  select(subject, gender, task, iteration, word, context, vowel, time, formant, ff_log_normalized, ff_exp_normalized, ff_raw)
+
+f1 <- df_clean %>%
+  filter(formant == 1)
+
+f2 <- df_clean %>%
+  filter(formant == 2)
+
+df_normalized <- bind_cols(f1,f2$ff_log_normalized,f2$ff_exp_normalized, f2$ff_raw)%>%
+  mutate(formant=NULL) %>%
+  rename(f1_log = ff_log_normalized, f1_exp = ff_exp_normalized, f1_raw = ff_raw, f2_log =...13, f2_exp = ...14, f2_raw = ...15)
+
+rm(df_clean, f1, f2)
+
+```
+
+Baseline Bengali data from the Shruti corpus:
+
+```{r, include=FALSE}
+
+# list of files in directory (paths relative to project root)
+file_list <- list.files(path="shruti/merged_speaker_data", full.names = T)
+vowel_list <- "oh|u|i|ee|A|E|aa"
+
+# create dataframe by reading data from each file in folder
+df_ben <- file_list %>%
+  set_names(.) %>%
+  # apply read_table() to each file, read only specified columns
+  map_df(read_table, .id = "FileName", col_types = cols_only(rowLabel ='f', Time ='d', F1 ='d', F2 ='d')) %>%
+  # rename columns
+  rename(speaker=FileName, vowel=rowLabel, time_absolute=Time, f1=F1, f2=F2) %>%
+  # only keep rows that have vowels
+  filter(grepl(vowel_list, vowel)) %>%
+  # clean column name (remove file path)
+  mutate(speaker = str_replace(speaker, "shruti/merged_speaker_data/", "")) %>%
+  mutate(speaker = str_replace(speaker, "_merged.txt", "")) %>%
+  mutate(speaker = as.factor(speaker)) %>%
+  droplevels()
+
+# columns for gender and time point (1-20)
+
+male = c("abd", "chandan", "kunal", "plb", "sayan", "sudha", "swarnendu", "amitava", "deb", "mainak", "rajashree", "sm", "suman", "xxx", "bd", "jit", "moy", "samaresh", "smt", "suparna", "chan", "jyotirmoy", "padma", "sandi", "styabrata", "suparnakdas")
+
+female = c("manika", "msm", "punam", "rita",  "ritwika",  "shyamoshree",  "suranjana",  "tml")
+
+timepoints <- c(1:20)
+
+# add column for gender using mutate() and ifelse()
+# add column for time (iterate over 1:20; length.out: desired length of final vector (here- total no. of rows))
+df_ben <- df_ben %>%
+  mutate(gender = ifelse(speaker %in% male, "male", "female"),
+         time = rep(timepoints, length.out = 2458340), .after = time_absolute,
+         time = as.factor(time)
+         )
+
+# Lobanov normalization (normalizing over all vowels here. For the English data-- only have 2 vowels.)
+
+# attach(bengali_df)
+# lobanov <- with(bengali_df,normLobanov(cbind(f1,f2), group = speaker))
+# bengali_lobanov <- data.frame(speaker, gender, vowel, time, lobanov)
+# rm(lobanov)
+# detach(bengali_df)
+
+rm(female, file_list, male, timepoints, vowel_list)
+
+```
+
+Log-mean normalization for Bengali data:
+
+```{r}
+
+# subset of dataframe with only vowels of interest (A and E) at vowel midpoints (timepoints 8-12),
+# removing outlier formant values (<300 Hz): ~ 11% of rows. In full dataset (all vowels): ~14%
+
+df_ben_AE <- df_ben %>%
+  filter(grepl("A|E", vowel), time %in% (8:12), f1>500) %>%
+  droplevels() %>%
+  mutate(time_absolute = NULL) %>%
+  relocate(gender, .after = speaker)
+
+# data in long format for log-mean normalization
+
+df_ben_long <- df_ben_AE %>%
+  pivot_longer(starts_with("f"), # pivot all the columns whose names start with "f"
+               values_to = "ff_raw",
+               # drop the f part, store the rest as value for new variable "formant" 
+               names_to = c(NA, "formant"), 
+               # how to divide up the variable name (f + formant)
+               names_pattern = "(.)(.)") %>% 
+  mutate_if(is.character, as.factor) %>%
+  # new column with log-transformed FFs
+  mutate(ff_log = log(ff_raw)) %>%
+  # composite variable vowel-formant (estimate for the dialectal reference pattern-- the target ff for that vowel in the dialect)
+  mutate (vowel.formant = factor(interaction(vowel, formant)))
+
+# fitting a regression model to predict log-transformed formant-frequencies (ff-log) for a given vowel, formant and speaker as the sum of a speaker displacement term (subject) and normalized vowel-formant effects (vowel.formant)  (Barreda&Neary 2018)
+
+attach(df_ben_long)
+
+# forcing the intercept to 0 so that coefficients are calculated for all levels of "subject"
+M_ben = lm (ff_log ~ 0 + speaker + vowel.formant, contrasts = list(vowel.formant = contr.sum))
+summary(M)
+
+# coefficients of the speaker effect (estimate of speaker parameter)
+speaker_coeff = dummy.coef(M_ben)$speaker
+# coefficients of the vowel.formant effect (estimate of dialect reference/target formant for each vowel)
+vowel_coeff = dummy.coef(M_ben)$vowel.formant
+
+detach(df_ben_long)
+
+# from each log-transformed ff of a particular speaker, want to subtract the speaker coefficient
+
+temp <- as.data.frame(speaker_coeff) %>%
+  rownames_to_column(var = "speaker")
+
+df_ben_long <- df_ben_long %>%
+  full_join(temp, by = "speaker")
+
+df_ben_long <- df_ben_long %>%
+  mutate(ff_log_normalized = ff_log - speaker_coeff)
+
+df_ben_long <- df_ben_long %>%
+  mutate(ff_exp_normalized = exp(ff_log_normalized))
+
+rm(temp,speaker_coeff,vowel_coeff)
+
+# putting F1 and F2 in separate columns for plotting (pivot_wider doesn't work because there are no unique identifiers for each vowel, since there were multiple reps in each condition).
+So crude workaround:
+
+df_clean <- df_ben_long %>%
+  select(speaker, gender, vowel, time, formant, ff_log_normalized, ff_exp_normalized)
+
+f1 <- df_clean %>%
+  filter(formant == 1)
+
+f2 <- df_clean %>%
+  filter(formant == 2)
+
+df_ben_normalized <- bind_cols(f1, f2$ff_log_normalized, f2$ff_exp_normalized) %>%
+  mutate(formant=NULL) %>%
+  rename(f1_log=ff_log_normalized, f1_exp=ff_exp_normalized, f2_log=...8, f2_exp=...9)
+
+rm(df_clean, f1, f2)
+
+variance <- var(df_ben$f1)
+variancef2 <- var(df_ben$f2)
+
+```
+
+Plotting vowel data:
+
+```{r}
+
+# IPA vowel symbols
+remapping_english <- c(`\\ae` = "æ", `\\vt` = "ʌ")
+remapping_bengali <- c(A = "a:", E = "æ")
+
+# add column with unicode vowels
+df_ben_normalized$unicodevowel <- remapping_bengali[as.character(df_ben_normalized$vowel)]
+
+df_normalized$unicodevowel <- remapping_english[as.character(df_normalized$vowel)]
+
+# plot: unilingual English, mixed, and baseline Bengali
+xlim <- c(1,-0.2)
+ylim = c(0,-1)
+
+# Bengali vowels
+with(df_ben_normalized,plotVowels(f1_log, f2_log, vowel,
+                        plot.tokens = F, plot.means = TRUE, pch.tokens = unicodevowel, pch.means = unicodevowel, 
+                        var.col.by = vowel, var.sty.by = vowel, 
+                        cex.tokens = 0.8, cex.means = 1, alpha.means = 0.8,alpha.tokens = 0.3,
+                        ellipse.line = F, ellipse.fill = T, fill.opacity = 0.5,
+                        poly.line = F, poly.order = c("A","E"),
+                        xlim = xlim, ylim = ylim, 
+                        legend.kwd="bottomleft", legend.args = list(legend=c("baseline æ","baseline a:")), 
+                        axes=FALSE, pretty = TRUE))
+par(new=TRUE)
+
+# English vowels
+with(df_normalized, plotVowels(f1_log, f2_log, vowel, group = context, 
+                               plot.tokens = F, plot.means = TRUE , pch.tokens = unicodevowel, pch.means = unicodevowel,
+                               var.col.by = NA, var.sty.by = context, 
+                               cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
+                               ellipse.line = T,  ellipse.fill = T,
+                               poly.line = F, poly.order = c("\\ae","\\vt"),
+                               xlim = xlim, ylim = ylim, 
+                               legend.kwd = "bottomright", legend.args = list(legend=c("unilingual","mixed")),
+                               fill.opacity = 0.1, pretty = TRUE))
+
 ```
@@
 phon_acco.model = lmer(f1 ~ 
-                         gender + context*vowel + time*vowel + task +
+                         gender + context*vowel + context*task +
                          (1|word) + (1+context*vowel|subject), 
@@
 summary(rePCA(phon_acco.model))
+
+fixef(phon_acco.model)
+powerSim(phon_acco.model)
+
+trial <- lmer(f1 ~
+                context*vowel + gender + task +
+                (1+context*vowel|subject) + (1|word),
+                data = df_time, REML = FALSE)
+
+summary(trial)
+
+fixef(trial)["contexte:tasks"] <- 0.1
+powerSim(trial, test = fixed("tasks"), nsim = 20)
+pc <- powerCurve (trial, along = "subject", test = fixed("task"), nsim = 20)
+plot(pc)
 

### 2022-02-26 RCode/transfer_analysis.Rmd
analysis models

@@
 library(lmerTest)
-library(stargazer)
 library(pwr)
-library(smir)
+library(simr)
 
@@
 # data frame with time as variable
+
+# lobanov-normalized FFs
 df_lobanov_time <- df_lobanov %>%
@@
 
-rm(df_lobanov)
+# raw FFs
+df_time <- df %>%
+  select(!c("filename", "f1_mid", "vowel_duration")) %>%
+  rename(f1_1=f1_5, f1_2=f1_15, f1_3=f1_25, f1_4=f1_35, f1_5=f1_45, f2_1=f2_5, f2_2=f2_15, f2_3=f2_25, f2_4=f2_35, f2_5=f2_45) %>%
+  # dplyr pivot() functions: convert data between long(indexed) and wide(cartesian) formats
+  pivot_longer(starts_with("f"),
+               names_to = c(".value", "time"),
+               names_sep = "_",
+               names_transform = list(time = as.factor)
+               )
+
 ```
@@
 
-```{r}
+```{r, eval=FALSE, include=FALSE}
 
@@
 # Lobanov normalization (normalizing over all vowels here. For the English data-- only have 2 vowels.)
-
 # attach(bengali_df)
@@
 df_ben_AE <- df_ben %>%
-  filter(grepl("A|E", vowel), time %in% (8:12), f1>500) %>%
+  filter(grepl("A|E", vowel), time %in% (8:12), f1>600) %>%
   droplevels() %>%
@@
 df_clean <- df_ben_long %>%
-  select(speaker, gender, vowel, time, formant, ff_log_normalized, ff_exp_normalized)
+  select(speaker, gender, vowel, time, formant, ff_log_normalized, ff_exp_normalized, ff_raw)
 
@@
 
-df_ben_normalized <- bind_cols(f1, f2$ff_log_normalized, f2$ff_exp_normalized) %>%
+df_ben_normalized <- bind_cols(f1, f2$ff_log_normalized, f2$ff_exp_normalized, f2$ff_raw) %>%
   mutate(formant=NULL) %>%
-  rename(f1_log=ff_log_normalized, f1_exp=ff_exp_normalized, f2_log=...8, f2_exp=...9)
+  rename(f1_log=ff_log_normalized, f1_exp=ff_exp_normalized, f1_raw=ff_raw, f2_log=...9, f2_exp=...10, f2_raw=...11)
 
 rm(df_clean, f1, f2)
-
-variance <- var(df_ben$f1)
-variancef2 <- var(df_ben$f2)
 
@@
 # plot: unilingual English, mixed, and baseline Bengali
-xlim <- c(1,-0.2)
-ylim = c(0,-1)
+xlim <- c(2500,600)
+ylim = c(1100,400)
+
+temp <- df_ben %>%
+  filter(time %in% (8:12), f1>300)
+
+with(temp,plotVowels(f1, f2, vowel,
+                        plot.tokens = F, plot.means = TRUE, pch.tokens = vowel, pch.means = vowel, 
+                        var.col.by = vowel, var.sty.by = vowel, 
+                        cex.tokens = 0.8, cex.means = 1, alpha.means = 0.8,alpha.tokens = 0.3,
+                        ellipse.line = F, ellipse.fill = T, fill.opacity = 0.3,
+                        poly.line = F, poly.order = c("A","E"),
+                        xlim = xlim, ylim = ylim, 
+                        legend.kwd="bottomleft", legend.args = list(legend=c("baseline æ","baseline a:")), 
+                        axes=T, pretty = TRUE))
 
 # Bengali vowels
-with(df_ben_normalized,plotVowels(f1_log, f2_log, vowel,
+with(df_ben_normalized,plotVowels(f1_raw, f2_raw, vowel,
                         plot.tokens = F, plot.means = TRUE, pch.tokens = unicodevowel, pch.means = unicodevowel, 
@@
                         cex.tokens = 0.8, cex.means = 1, alpha.means = 0.8,alpha.tokens = 0.3,
-                        ellipse.line = F, ellipse.fill = T, fill.opacity = 0.5,
+                        ellipse.line = F, ellipse.fill = T, fill.opacity = 0.2,
                         poly.line = F, poly.order = c("A","E"),
@@
 # English vowels
-with(df_normalized, plotVowels(f1_log, f2_log, vowel, group = context, 
+with(df_normalized, plotVowels(f1_raw, f2_raw, vowel, group = context, 
                                plot.tokens = F, plot.means = TRUE , pch.tokens = unicodevowel, pch.means = unicodevowel,
@@
 
-LME models:
+## LME models:
+
+```{r}
+# plot for English vowels:
+
+# Lobanov-normalized FFs
+xlim = c(2.5, -2)
+ylim = c(2.3, -2)
+
+with(df_lobanov_time, plotVowels(f1, f2, vowel, group = task,
+                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = vowel,
+                               var.col.by = task, var.sty.by = NULL, 
+                               cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
+                               ellipse.line = F,  ellipse.fill = T,
+                               poly.line = F, poly.order = c("\\ae","\\vt"),
+                               xlim = xlim, ylim = ylim, 
+                               legend.kwd = "bottomright", legend.args = gender,
+                               fill.opacity = 0.1, pretty = TRUE))
+
+# raw FFs
+with(df_time, plotVowels(f1, f2, vowel, group = task, 
+                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = vowel,
+                               var.col.by = task, var.sty.by = context, 
+                               cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
+                               ellipse.line = T,  ellipse.fill = T,
+                               poly.line = F, poly.order = c("\\ae","\\vt"),
+                               xlim = c(2500, 700), ylim = c(1400, 100), 
+                               legend.kwd = "bottomright", legend.args = gender,
+                               fill.opacity = 0.1, pretty = TRUE))
+
+```
+
+Gender difference is not removed after lobanov normalization. Normalizing by gender:
 
@@
 
+df_lobanov_gender_time <- df_time %>%
+  mutate(f1 = with(df_lobanov_time, normLobanov(f1), group = gender),
+         f2 = with(df_lobanov_time, normLobanov(f2), group = gender))
+
+```
+
 ## main model
 
+writeup-- (using data from all 5 points.
+This should be okay because the pre-vocalic consonants are all the same, so any effect of the consonant should be uniform.
+A difference between contexts in the early part of the vowel suggests differences in target FFs)
+
+Gender effects: try:
+- normalize by gender
+- use log-mean normalized values
+
+Hypotheses:
+
+- FFs are different across vowels
+- FFS are different across contexts
+- the two vowel categories are differently affected (precise direction/extent of shift in FFs depends on category, position in vowel space, presence/absence of contrast in L1)
+So main effect of interest: context*vowel
+
+- if transfer patterns differ across tasks, then expect: context\*task interaction, OR context\*vowel\*task interaction
+
+- FFs should change across time, so effect of time on FF. But what time is doing to the FF (direction of change)-- depends on the vowel category. So expect: time\*vowel interaction
+
+- if transfer patterns change across time, then expect: context\*time interaction, OR context\*vowel\*time interaction
+
+Random effects structure:
+
+- subject: intercept, context, vowel (target FF different across speakers), context\*vowel interaction (transfer pattern different across speakers)
+
+-item: intercept (FF for vowel different in each item), context (the effect of context is different for each word), NOT vowel (between-item variable), NOT gender (even though within-item, no reason to believe that effect of gender on FF is different for each item)
+
+```{r, results='hide'}
+temp <- df_lobanov_time %>%
+  filter(time == "5")
+
+# F1
+
+# full model
+phon_acco.model = lmer(f1 ~ 
+                         gender + time*vowel + vowel*context*task +
+                         (1|word) + (1+context+vowel|subject), 
+                         data= df_lobanov_time, REML=FALSE)
+summary(phon_acco.model)
+summary(rePCA(phon_acco.model))
+
+# null model
+null.model = lmer(f1~
+                   gender + time*vowel + vowel*context + task*vowel +
+                   (1|word) + (1+context+vowel|subject), 
+                   data= df_lobanov_time, REML=FALSE)
+
+summary(null.model)
+
+anova(phon_acco.model, null.model)
+```
+
+Random effects structure: (1+context|word) + (1+context+vowel|subject). But by-word random slope for context is explaining barely any variance, and null model is not converging. So removing that.
+
+- Full model1: gender + time\*vowel + vowel\*context
+- null model1: gender + time*vowel
+full model is better
+- null model2: gender + time*vowel + context
+full model still better
+
+Shows: F1 differs across contexts (mediated by vowel category). So transfer is happening.
+
+Now want to see: effect of task?
+If there is a difference in transfer patterns across tasks, then expect task to interact with the vowel\*context variable 
+- full model: gender + time\*vowel + vowel\*context\*task 
+-task by itself affects F1 (unexpected)
+- marginal effect of context\*task (effect of context on F1 is moderated by task) -- expected
+- significant effect of vowel\*context\*task (transfer pattern is moderated by task) -- expected
+
+- null model: gender + time\*vowel + vowel\*context
+full model better
+
+- null model: gender + time\*vowel + vowel\*context + task
+(because adding task is significant on its own, and adding it as an interaction term increases model parameters, want to see if having it as an interaction significantly improves fit)
+full model better -- task does interact with context and vowel
+
+- null model: gender + time\*vowel + vowel\*context + task\*context 
+to see if three-way interaction is needed. Is task only interacting with context?
+nope, context\*task not significant here
+full model better
+
+gender + time\*vowel + vowel\*context + task\*vowel 
+to see if three-way interaction is needed. Is task only interacting with vowel?
+context\*vowel is significant
+but anova shows that the full model is still better
+
+So: need the task\*context\*vowel interaction
+
+Want to see: is the effect of task local (decreases with time)? Since the effect of juat task on F1 is unexpected, want to see if it might be an artifact of the paradigm/stimuli
+
+```{r}
+# F2
 phon_acco.model = lmer(f1 ~ 
@@
 summary(rePCA(phon_acco.model))
+
+```
+
+Power analysis:
+
+```{r}
 

### 2022-04-16 RCode/transfer_analysis.Rmd
manuscript revisions, added literature

@@
 knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
+
+options(scipen=999) # turn off scientific e notation, show full numbers. To turn on: options(scipen=0)
 ```
@@
 library(phonR)
+library(stargazer)
+library(modelsummary)
 library(lme4)
@@
 library(simr)
+library(EMAtools) # to calculate Cohen's d values from lmer objects
 
+# replacing vowel symbols with unicode:
+remapping_english <- c(`\\ae` = "æ", `\\vt` = "ʌ")
+
+# read in data
 df <- read.csv("RCode/formant_data_5.csv", check.names = TRUE) %>%
   rename_all(tolower) %>%
+  mutate(vowel = remapping_english[as.character(vowel)]) %>%
   mutate_if(is.character,as.factor)
@@
 
-```{r, include=FALSE}
+```{r, eval=FALSE}
 
@@
 
-```{r}
+```{r, eval=FALSE}
 
@@
 
-with(df_lobanov_time, plotVowels(f1, f2, vowel, group = task,
-                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = vowel,
-                               var.col.by = task, var.sty.by = NULL, 
+with(df_lobanov_gender_time, plotVowels(f1, f2, vowel, group = interaction(task,context),
+                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = interaction(task,context),
+                               var.col.by = interaction(task,context), var.sty.by = gender, 
                                cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
-                               ellipse.line = F,  ellipse.fill = T,
+                               ellipse.line = F,  ellipse.fill = F,
                                poly.line = F, poly.order = c("\\ae","\\vt"),
-                               xlim = xlim, ylim = ylim, 
+                               xlim = c(1,-1), ylim = c(0.5,-1), 
                                legend.kwd = "bottomright", legend.args = gender,
@@
 
+```
+
+```{r}
+
 # raw FFs
-with(df_time, plotVowels(f1, f2, vowel, group = task, 
+with(df_time, plotVowels(f1, f2, vowel, group = interaction(gender, task), 
                                plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = vowel,
-                               var.col.by = task, var.sty.by = context, 
+                               var.col.by = interaction(gender, task), var.sty.by = task, 
                                cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
-                               ellipse.line = T,  ellipse.fill = T,
+                               ellipse.line = F,  ellipse.fill = F,
                                poly.line = F, poly.order = c("\\ae","\\vt"),
@@
 
-df_lobanov_gender_time <- df_time %>%
-  mutate(f1 = with(df_lobanov_time, normLobanov(f1), group = gender),
-         f2 = with(df_lobanov_time, normLobanov(f2), group = gender))
+gender_normalized <- with(df_lobanov_time, normLobanov(cbind(f1,f2), group=gender))
 
+df_lobanov_gender_time <- df_lobanov_time %>%
+  select(!c(f1,f2)) %>%
+  mutate(f1 = gender_normalized$f1,
+         f2 = gender_normalized$f2)
+
+rm(gender_normalized)  
 ```
@@
 Gender effects: try:
-- normalize by gender
+- normalize by gender (done)
 - use log-mean normalized values
@@
 ```{r, results='hide'}
-temp <- df_lobanov_time %>%
-  filter(time == "5")
+
+temp <- df_lobanov_gender_time %>%
+  mutate(time = as.numeric(time)) %>%
+  mutate(time = scale(time))
 
@@
 # full model
-phon_acco.model = lmer(f1 ~ 
-                         gender + time*vowel + vowel*context*task +
-                         (1|word) + (1+context+vowel|subject), 
-                         data= df_lobanov_time, REML=FALSE)
+phon_acco.model_f1 = lmer(f1 ~ 
+                         time*vowel + vowel*context*task + time*task +
+                         (1+context|word) + (1+context+vowel|subject), 
+                         data= temp, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
 summary(phon_acco.model)
@@
 
+# Cohen's d values (from EMAtools package):
+d_F1 <- lme.dscore(phon_acco.model, data=temp, type='lme4')
+
+##--> find way to export fixed effects table (estimate, SE, p values) to LaTex
+
+stargazer(phon_acco.model)
+
 # null model
 null.model = lmer(f1~
-                   gender + time*vowel + vowel*context + task*vowel +
-                   (1|word) + (1+context+vowel|subject), 
-                   data= df_lobanov_time, REML=FALSE)
+                   time*vowel + vowel*context*task + time*task +
+                   (1+context|word) + (1+context+vowel|subject), 
+                   data= temp, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
 
 summary(null.model)
+summary(rePCA(null.model))
 
 anova(phon_acco.model, null.model)
+
+# some weird interaction with lmerTest package. Detach that first
+stargazer(phon_acco.model)
+modelsummary(phon_acco.model)
+
 ```
 
-Random effects structure: (1+context|word) + (1+context+vowel|subject). But by-word random slope for context is explaining barely any variance, and null model is not converging. So removing that.
+Random effects structure: (1+context|word) + (1+context+vowel|subject).
+But by-word random slope for context is explaining barely any variance, and null model is not converging.
+So removing that: (1|word) + (1+context+vowel|subject)
 
-- Full model1: gender + time\*vowel + vowel\*context
-- null model1: gender + time*vowel
+- Full model: time\*vowel + vowel\*context
+- null model: time*vowel
 full model is better
-- null model2: gender + time*vowel + context
+- null model: time*vowel + context
 full model still better
@@
 
-gender + time\*vowel + vowel\*context + task\*vowel 
+- null model: gender + time\*vowel + vowel\*context + task\*vowel 
 to see if three-way interaction is needed. Is task only interacting with vowel?
-context\*vowel is significant
+task\*vowel is significant
 but anova shows that the full model is still better
@@
 
-Want to see: is the effect of task local (decreases with time)? Since the effect of juat task on F1 is unexpected, want to see if it might be an artifact of the paradigm/stimuli
+vowel\\vt:contexte:tasks      -1.185e-01  3.551e-02  1.162e+04  -3.338 0.000848 ***
+vowelvt:contexte has a negative slope (for vt, in the English context, F1 is lower--vowel is higher, i.e. vowel is lower in b context. This is the main transfer effect.)
+Now, the interaction of this with tasks has a negative slope (this entire thing is increased (becomes more in magnitude) in the s task compared to the p task)
+i.e., more transfer in the s task?
+
+Want to see: is the effect of task local (decreases with time)? Since the effect of just task on F1 is unexpected, want to see if it might be an artifact of the paradigm/stimuli
+tasks:time     -7.506e-02  1.751e-02  1.162e+04  -4.286 1.83e-05 ***
+(tasks has a positive slope (F1 increases/vowel lowers in s task);
+time has a positive slope (F1 increases/vowel lowers with time);
+interaction term has a negative slope: the effect of tasks (lowering of vowel) reduces with increase in time: effect of task is more in the beginning: local effect)
+
+Is transfer pattern changing with time? 
+vowel\*context\*time: not significant
+context\*time: not significant
+
+full model: gender + vowel\*context\*task\*time
+null model: gender + time\*vowel + vowel\*context\*task + task\*time -- final model
+null model better (time doesn't need to interact with context, or with vowel\*context)-- transfer pattern is not changing over time-- the targets are different
+No effect of context on formant dynamics. There is a linear scaling of the trajectory--targets are different.
+
+Do transfer patterns differ across gender?
+full model: time\*vowel + vowel\*context\*task\*gender + task\*time
+no effect of vowel\*context\*gender-- no difference in transfer patterns
+but: task\*gender is significant: men show more lowering in the sentence task than females --unexpected
 
@@
 # F2
-phon_acco.model = lmer(f1 ~ 
-                         gender + context*vowel + context*task +
-                         (1|word) + (1+context*vowel|subject), 
-                         data= df_time, REML=FALSE)
+
+# full model
+phon_acco.model_f2 = lmer(f2 ~ 
+                         vowel*time + vowel*context*task + task*time +
+                         (1+context|word) + (1+context+vowel|subject), 
+                         data= temp, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
 summary(phon_acco.model)
@@
 
+# Cohen's d values (from EMAtools package):
+d_F2 <- lme.dscore(phon_acco.model, data=temp, type='lme4')
+
+##--> find way to export fixed effects table (estimate, SE, p values) to LaTex
+
+# null model
+null.model = lmer(f2~
+                   time*vowel + vowel*context*task*time +
+                   (1+context|word) + (1+context+vowel|subject), 
+                    data= temp, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(null.model)
+
+anova(phon_acco.model, null.model)
+
+stargazer(phon_acco.model)
 ```
+
+- Full model: time\*vowel + vowel\*context
+effect of vowel: vt is more backed than ae
+effect of time: F2 changes with time, but direction is moderated by vowel: ae is fronted, vt less so
+marginal effect of context: both vowels are fronted in b context
+main effect of vowel*context: ae is fronted in the b context, vt is very very slightly backed
+- null model: time*vowel
+full model is better
+- null model: time*vowel + context
+full model still better
+
+Shows: F2 differs across contexts (mediated by vowel category). So transfer is happening in front-back dimension.
+
+Now want to see: effect of task?
+If there is a difference in transfer patterns across tasks, then expect task to interact with the vowel\*context variable 
+- full model: time\*vowel + vowel\*context\*task 
+fails to converge.
+fitting 0-correlation model (remove correlations from by-subject random effects): still doesn't converge.
+VCV matrix shows by-subject slopes for vowel have correlation of 1.
+So removing that.
+Converges now.
+independent effect of task: vowels are backed in the s task
+
+-task by itself affects F2 (unexpected)
+- marginal effect of context\*task (effect of context on F2 is moderated by task) -- more fronting (in b context) during s task
+- effect of vowel\*context\*task: not significant
+
+- null model: gender + time\*vowel + vowel\*context
+full model better
+---
+- null model: gender + time\*vowel + vowel\*context + task
+(because adding task is significant on its own, and adding it as an interaction term increases model parameters, want to see if having it as an interaction significantly improves fit)
+full model better -- task does interact with context and vowel
+
+- null model: gender + time\*vowel + vowel\*context + task\*context 
+to see if three-way interaction is needed. Is task only interacting with context?
+nope, context\*task not significant here
+full model better
+
+- null model: gender + time\*vowel + vowel\*context + task\*vowel 
+to see if three-way interaction is needed. Is task only interacting with vowel?
+task\*vowel is significant
+but anova shows that the full model is still better
+
+So: need the task\*context\*vowel interaction
+
+vowel\\vt:contexte:tasks      -1.185e-01  3.551e-02  1.162e+04  -3.338 0.000848 ***
+vowelvt:contexte has a negative slope (for vt, in the English context, F1 is lower--vowel is higher, i.e. vowel is lower in b context. This is the main transfer effect.)
+Now, the interaction of this with tasks has a negative slope (this entire thing is increased (becomes more in magnitude) in the s task compared to the p task)
+i.e., more transfer in the s task?
+
+Want to see: is the effect of task local (decreases with time)? Since the effect of just task on F1 is unexpected, want to see if it might be an artifact of the paradigm/stimuli
+tasks:time     -7.506e-02  1.751e-02  1.162e+04  -4.286 1.83e-05 ***
+(tasks has a positive slope (F1 increases/vowel lowers in s task);
+time has a positive slope (F1 increases/vowel lowers with time);
+interaction term has a negative slope: the effect of tasks (lowering of vowel) reduces with increase in time: effect of task is more in the beginning: local effect)
+
+Is transfer pattern changing with time? 
+vowel\*context\*time: not significant
+context\*time: not significant
+
+full model: gender + vowel\*context\*task\*time
+null model: gender + time\*vowel + vowel\*context\*task + task\*time -- final model
+null model better (time doesn't need to interact with context, or with vowel\*context)-- transfer pattern is not changing over time-- the targets are different
+No effect of context on formant dynamics. There is a linear scaling of the trajectory--targets are different.
+
+Do transfer patterns differ across gender?
+full model: time\*vowel + vowel\*context\*task\*gender + task\*time
+no effect of vowel\*context\*gender-- no difference in transfer patterns
+but: task\*gender is significant: men show more lowering in the sentence task than females --unexpected
 
@@
 
-fixef(phon_acco.model)
-powerSim(phon_acco.model)
+# F1:
 
-trial <- lmer(f1 ~
-                context*vowel + gender + task +
-                (1+context*vowel|subject) + (1|word),
-                data = df_time, REML = FALSE)
+fixef(phon_acco.model_f1)
 
-summary(trial)
+fixef(phon_acco.model_f1)["vowelʌ:contexte:tasks"] <- - 0.1
 
-fixef(trial)["contexte:tasks"] <- 0.1
-powerSim(trial, test = fixed("tasks"), nsim = 20)
-pc <- powerCurve (trial, along = "subject", test = fixed("task"), nsim = 20)
+# power for interaction term: compare current model to one without the interaction term
+powerSim(phon_acco.model_f1, nsim = 20, test = fcompare(f1 ~ time*vowel + vowel*context + time*task))
+# shows 90% power
+
+pc <- powerCurve (phon_acco.model_f1, along = "subject", test = fixed("vowelʌ:contexte:tasks"), nsim = 20)
 plot(pc)
 
-null.model = lmer(f1 ~ 
-                    gender + context*vowel + task + 
-                    (1|word) + (1+context*vowel|subject), 
-                    data=df_time, REML=FALSE)
-summary(null.model)
+```
 
-anova(phon_acco.model,null.model)
+Reporting power analyses:
 
-coef(phon_acco.model)
+Now that you have conducted your power analysis you will need to report it in APA style.
+Thissentence (or two) is usually placed in the Results section of a research report near the start of theresults when the assumptions and adequacy of the sample size are being reported.The sentence that you write should include the following:
 
-```
+- The type of statistical test that you will be using to analyse the data (e.g., an independentsample t-test)
+- The type of power analysis that you did–either a priori or post hoc (generally researchersreport the a priori power analysis)
+- The name of the software program that you used to conduct the power analysis and acitation to the authors and date of publication (see the G\*Power website listed above forthe journal article to cite)
+- The level of power aspired to for a prior analyses, or the actual power achieved for post hocanalysesThe estimated or actual effect size used in the calculationThe alpha level used in the calculation
+- For an a priori power analysis, report the sample size needed to achieve the minimumacceptable power level.For post hoc analysis, report achieved sample size.
+
+All this detail needs to be reported concisely and in one or two sentences at most.I’ve written an example reportfor question 1 for a power of .80 below:
+
+An a priori power analysis was conducted using G*Power3 (Faul, Erdfelder, Lang, & Buchner,2007) to test the difference between two independent group means using a two-tailed test,a medium effect size (d= .50), and an alpha of .05.Result showed that a total sample of 128 participants with two equal sized groups ofn= 64 was required to achieve a power of .80.
+
+-------
+
+A statistical power analysis was performed for sample size estimation, based on data from pilot
+study/published study X (N=…), comparing …. to ….. The effect size (ES) in this study was ….,
+considered to be extremely large/large/medium/small using Cohen's (1988) criteria. With an
+alpha = .05 and power = 0.80, the projected sample size needed with this effect size (GPower
+3.1 or other software) is approximately N = ….. for this simplest between/within group
+comparison. Thus, our proposed sample size of ..N+.. will be more than adequate for the main
+objective of this study and should also allow for expected attrition and our additional objectives
+of controlling for possible mediating/moderating factors/subgroup analysis, etc.

### 2022-04-16 draft_BLC/BLC_AM-ID_manuscript.tex
manuscript revisions, added literature

@@
 \usepackage[letterpaper]{geometry}
-%\usepackage{times}
+\usepackage{times}
 \geometry{top=1.5in, bottom=1.0in, left=1.0in, right=1.0in}
@@
 }
+
+\usepackage{easyReview}
 
@@
 The present study is concerned with the latter.
-In production, this kind of influence has been variously termed \textsc{transfer, drift, accommodation,} and \textsc{interference}-- we use these interchangeably here to indicate any interaction between two sets of phonetic norms.
+In production, this kind of influence has been variously termed \textsc{transfer, drift, accommodation,} and \textsc{interference}-- we use the term \textsc{transfer} here, to indicate any interaction between two sets of phonetic norms.
 
@@
 We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction.
-The results demonstrate a dynamic effect of transfer on L2 vowels during mixed-language production, but no difference in the extent of transfer between the two paradigms.
+The results demonstrate a dynamic effect of transfer on L2 vowels during mixed-language production, \alert{but no difference in the extent of transfer between the two paradigms --and slightly different patterns of shift among the paradigms?}.
 We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.
@@
 \ref{questions_and_hypotheses} summarizes the research questions and hypotheses.
+
+Existing studies have studied phonetic transfer along different acoustic features.
+While discussing the literature in subsequent sections, we will often indicate the acoustic feature in parentheses following a citation, for clarity.
 
@@
 
-The vowel system of IE contains the monopthongs \nt{I, i, E, e, \ae, @/2, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \citep{wells1982accents, masica1972sound}.
+The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \citep{wells1982accents, masica1972sound}.
 %Their distribution in the F1XF2 space is shown in (insert table 1.7, Pandey).
@@
 Given their transient nature, dynamic changes in production during mixed-language use are generally attributed to the latter, e.g.
-online processing costs \citep{olson2013bilingual,tsui2019impact,vsimavckova2015immediate}, language mode \citep{simonet2014phonetic}, context-awareness \citep{khattab2013phonetic}.
+online processing costs \citep[][VOT]{olson2013bilingual,tsui2019impact,vsimavckova2015immediate}, language mode \citep[][vowel quality]{simonet2014phonetic}, context-awareness \citep[][phonological variables]{khattab2013phonetic}.
 What triggers this interaction?
@@
 A number of studies have specifically manipulated language mode, both in the presence and absence of switching.
-Overall, results suggest that: (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use--- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \citep{olson2016role,tsui2019impact} or spontaneous conversation \citep{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
-(iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \cite{olson2016role} found no additive effects of language mode, \cite{olson2013bilingual} found a balanced language context to inhibit transfer compared to unbalanced contexts.
-Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days \citep{schwartz2015language, bullock2009trying,antoniou2011inter, elias2017effects,vsimavckova2015immediate,vsimavckova2018patterns}.
+Overall, results suggest that: (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep[][vowel quality]{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use--- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \citep[][VOT]{olson2016role,tsui2019impact} or spontaneous conversation \citep[][VOT]{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
+(iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \cite[][VOT]{olson2016role} found no additive effects of language mode, \cite[][VOT]{olson2013bilingual} found a balanced language context to inhibit transfer compared to unbalanced contexts.
+Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days (\cite{elias2017effects}; vowel quality), \citep[][VOT]{schwartz2015language, bullock2009trying,antoniou2011inter, vsimavckova2015immediate,vsimavckova2018patterns}.
 
@@
 
-Research comparing bilinguals speakers' phonological systems with monolinguals treats them as relatively stable over time.
+Research comparing bilingual speakers' phonological systems with monolinguals treats them as relatively stable over time.
 The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods \citep[e.g.][]{guion2003vowel, caramazza1973acquisition}.
@@
 and between-subject comparisons of bilingual speakers with different durations of L2 exposure suggest that interaction between the sound systems is dynamic: over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts.
-For example, \cite{bohn1992production}
+For example, \cite[][vowel quality]{bohn1992production}
 observed that approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language transfer.
@@
 Moreover, changes due to transfer are not necessarily unidirectional or irreversible.
-\cite{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language.
+\cite[][VOT]{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language.
 This not only evidenced that cross-language interaction can be triggered in an order of months, but also that the effects can be reversed within a similar time range.
-A more recent study by \cite{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks).
-Based on a short-term longitudinal study, \cite{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2.
+A more recent study by \cite[][VOT]{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks).
+Based on a short-term longitudinal study, \cite[][VOT]{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2.
 For some sounds and features (though not others, cf.
@@
 
-The research discussed above concerns situations where the bilingual speaker is operating in any one of their languages.
+The research discussed above concerns situations where a speaker is operating in any one of their languages.
 When it comes to the phonetics of mixed-language use, the majority of existing studies only analyze the switched (target) token.
 Thus, there are few direct measurements of the duration of short-term transfer effects.
-One study which measured this in a code-switching paradigm \citep{bullock2009trying} did not find any residual effects on the matrix language following a switch, suggesting that transfer during code-switching is localized.
-Indirect evidence in support of this comes from experiments which have elicited switched and non-switched tokens in the same test block and still reported differences between the two \citep[e.g.][]{tsui2019impact,olson2013bilingual}, suggesting that changes due to transfer are quickly `reset'--- in an order of seconds.\\
+One study which measured this in a code-switching paradigm \citep[][VOT]{bullock2009trying} did not find any residual effects on the matrix language following a switch, suggesting that transfer during code-switching is localized.
+Indirect evidence in support of this comes from experiments which have elicited switched and non-switched tokens in the same test block and still reported differences between the two \citep[e.g.][VOT]{tsui2019impact,olson2013bilingual}, suggesting that changes due to transfer are quickly `reset'--- in an order of seconds.\\
 Note that all the studies discussed above measure VOT, which is a temporal feature.
@@
 In non-switched production, this is evidenced through a pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf.
-\textsc{equivalence classification}-- \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][]{guion2003vowel}.
+\textsc{equivalence classification}-- \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.
 
-In mixed-language production, the role of language status is less clear-- studies have reported unidirectional influence of L1 on L2 \citep{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \citep{tsui2019impact,elias2017effects, olson2013bilingual}, bidirectional convergence \citep{bullock2009trying, olson2016role}, divergence \citep{bullock2009trying,vsimavckova2018patterns}, and no influence \citep{muldner2019phonetics,schwartz2015language}.
+In mixed-language production, the role of language status is less clear-- studies have reported unidirectional influence of L1 on L2 \citep[][VOT]{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \citep[][VOT]{tsui2019impact, olson2013bilingual}, (\cite{elias2017effects}, vowel quality) bidirectional convergence \citep[][VOT]{bullock2009trying, olson2016role}, divergence \citep[][VOT]{bullock2009trying,vsimavckova2018patterns}, and no influence \citep[][vowel quality]{muldner2019phonetics}, \citep[][phonological process]{schwartz2015language}.
 Since most existing studies measure shifts in VOT, the observed asymmetries between languages have been variously explained either in terms of their L1 vs L2 status, or language-specific differences between long- and short-lag VOT languages.
@@
 
-Bengali has a four-way laryngeal contrast.
-Thus, VOT does not encode a binary voiced-voiceless distinction, and moreover is not the only relevant cue for the corresponding phonological categories \citep{dmitrieva2020acoustic}.
-This means that any effect of L1 Bengali on L2 English VOT is likely to be mediated by additional phonological factors.
-Thus, avoiding any potential confounds was another reason we chose to focus on vowel quality as a target for transfer.
+Since Bengali has a four-way laryngeal contrast, VOT does not encode a binary voiced-voiceless distinction.
+Moreover, VOT is not the only relevant cue for the corresponding phonological categories in the language \citep{dmitrieva2020acoustic}.
+This means that any effect of L1 Bengali on L2 English VOT is likely to be affected by additional phonological factors.
+Thus, avoiding any potential confounds was another reason we chose to focus on vowel quality as a target for transfer in the present study.
 
@@
 Asymmetries in the extent and patterns of transfer have not only been observed between languages, but also between different sounds/features of a language, in both long-term and transient interactions.
-Studies that have examined multiple sound categories have found that interactions between individual sounds pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep[e.g.][]{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot treated as an atomic measure.
+Studies that have examined multiple sound categories have found that interactions between individual sound pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep[e.g.][vowel quality]{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot treated as an atomic measure.
 In light of the discussion about VOT in section \ref{sound systems}, this is not surprising --- if a general linguistic principle of `room for movement' constrains transfer, then we should expect it to apply to individual sounds too.
@@
 The first iteration of each task was treated as a trial, and excluded from further analyses.
-An additional 67 tokens (2.79\%) were excluded due to errors in presentation/production, giving 2333 target word tokens for analysis.
+An additional 67 tokens (2.79\%) were excluded due to errors in presentation/production, giving 2333 target word tokens (\nt{2}: 1137, \nt{\ae}: 1196) for analysis.
 Recordings were manually segmented and annotated in Praat \citep{boersma2016praat}, and the first and second formant frequencies (F1 and F2) of the target vowels \nt{2, \ae} were measured at five points-- in 10\% increments starting at 5\% into the vowel.
 This gave measures of F1 and F2 at 5\%, 15\%, 25\%, 35\% and 45\% into the vowel, allowing us to observe any dynamic changes in the extent of transfer over the course of producing the vowel.
+Individual vocal tract differences affect formant frequencies across speakers.
+To avoid this confound, we used Lobanov normalization (using the PhonR package \citep{phonR} in R \citep{r}, representing each speaker's formant frequencies as the scaled distance from their own mean (\alert{citation for lobanov normalization?}).
+Plotting the vowels (figure) shows a difference between male and female speakers, such that male speakers have lower formant frequencies.
+Normalizing by speaker does not get rid of this difference.
+Since gender-related differences in formant frequencies are not of interest in this study, we normalized individual values by gender (representing each speaker's values as a scaled distance from the group mean).
+Statistical analyses were performed on these normalized values.
 
@@
 All statistical analyses were performed in R \citep{r}.
-The dependent variables were vowel height (F1) and backness (F2).
-We fitted a series of linear mixed effects models using the \emph{lme4} package \citep{lme4}, to examine the effect of experimental conditions on lobanov normalized F1 and F2 at each of the five points in the vowel (sec.\ref{analysis}).
+We used linear mixed effects models to test the effect of experimental conditions on vowel height (F1) and backness (F2) respectively.
+Since formants were extracted from five points over the course of articulating the vowel, time was coded numerically (1-5), and treated as an independent variable.
+This was scaled and centered prior to analysis and treated as a continuous predictor.
+Models were fit using the \emph{lme4} package \citep{lme4}, and \emph{lmerTest} \citep{lmerTest} was used to estimate p values for individual variables.
 For all mixed models, the alpha criterion was set at $|t| > 2 $.
+The dependent variables were Lobanov normalized F1 and F2, and we fitted separate models for these.
 The independent variables or fixed factors, along with their levels, are given in table \ref{table variables}.
@@
 
-Our target in this analysis was to model vowel quality at different points in the vowel using the same mixed-effects model, allowing us to track if/how transfer changes over time.
-To do this, we arbitrarily chose a point in the vowel: F1 at 25\% (henceforth F1$\_$25) to identify an optimal model, and then fitted this to all other points.
+\subsection{Models}
 
-Vowel quality and gender are expected to systematically affect F1.
-Thus, we began with a model that had Vowel and Gender as fixed effects, random intercepts for Subject and Item, and random slopes for each within-unit variable that was relevant to our hypotheses: by-subject slopes for Task, Context, and Vowel; by-item slopes for Task and Context.
-Next, adding Context as a fixed effect significantly improved the model fit ($\chi^2(1) = 5.54, p= .01854$), suggesting that the experimental condition affects vowel height.
-Following from our hypothesis \textbf{Asymmetry}, we expected the extent of change in vowel height between the conditions to be mediated by the identity of the vowel.
-Thus, we added a two-way interaction factor between Context and Vowel (Context*Vowel).
-As expected, this was found to significantly improve model fit ($\chi^2(1) = 12.61, p= .0003837$).
-To assess whether the experimental paradigm affected F1, we added Task as a fixed effect.
-This marginally improved the model ($\chi^2(1) = 3.74, p= .05302$).
-Towards addressing our research question \textbf{Paradigm}, we wanted to assess whether the extent of shift differs across the two paradigms.
-However, no significant improvement was achieved by adding an interaction effect between Task and Context.
-Subsequent additions of other interactions, and removing each of the fixed effects in turn, did not lead to any further improvement.
+\subsubsection*{Fixed effects}
 
-Context and Vowel are both within-subject factors.
-Since our hypothesis concerns their interaction, we modified the random effects structure by adding a by-subject random slope for the interaction term \citep{brauer2018linear}.
-This improved the model fit ($\chi^2(5) = 11.43, p = .04342$).
-Thus, we report on a model that includes Gender, Task, Context, Vowel, Context*Vowel as fixed effects, random intercepts for Subject and Item, by-subject random slopes for Task, Context, Vowel and Context*Vowel, and by-item random slopes for Task and Context.
+Our study involved two IE vowels. Vowel category is expected to systematically affect formant frequencies (FFs).
 
-Figure \ref{vowels_e_b} summarizes the findings of the experiment ($F1XF2$ plots for vowel productions in the unilingual and mixed conditions, in comparison to the related Bengali categories from \cite{dutta2021}).
-The following sections discuss the results for vowel height and backness respectively.
+The experimental treatment of interest is Context, i.e.
+whether the target item was presented in a unilingual (English) or mixed (Bengali) context.
+Our main research question (\bf{L2 vowels}) asks whether L2 vowel quality (formant frequencies) of proficient bilinguals is affected by the Context of utterance.
+We hypothesize that the direction of shift in the mixed condition is towards the corresponding L1 category.
+Thus, we expect the effect of Context on formant frequencies, i.e.
+the direction of shift, to be affected by the vowel category (c.f.
+the second hypothesis \bf{Asymmetry}).
+Thus, our main effect of interest is the interaction term Context*Vowel.
+This corresponds to the transfer pattern.
 
-< Insert Figure \ref{vowels_e_b} about here >
+Our third hypothesis (\bf{Paradigm}) was that the language-mixing paradigm, or Task, will affect the extent and direction of transfer.
+If this is the case, we expect Task to interact with the transfer term, giving a three way Task*Context*Vowel interaction.
+
+Since we measured FFs at five points in the vowel, Time is expected to affect the FF.
+However, formant dynamics are expected to depend on the vowel category, and so we expect the interaction of Vowel*Time to predict formant frequency.
+If transfer patterns change across time, then we should expect it to interact with the transfer term, giving a Context*Vowel*Time interaction.
+
+\subsubsection*{Random effects structure}
+
+Subject and Item were treated as random factors.
+Given our hypotheses, we included random intercepts for Subject and Item, by-subject random slopes for Vowel (formant targets for vowel categories differ across individuals) and Context (individuals differ in their response to the experimental condition), and a by-item random slope for Context (the effect of experimental conditions differ across words).
+\alert{(convergence problems with by-word random slope for context-- changing the optimizer to bobyqa solved the problem)}
+
+We tested the significance of our hypothesized explanatory variables using t-values and p-values for individual variables in the models, and tested overall model fit by comparing the full model to a corresponding null model that lacks the variable of interest using ANOVAs.
+The following subsections summarize and discuss the model outputs for F1 and F2 respectively:
+
+\subsection{Power analysis}
+We were able to recruit 10 participants.
+To confirm whether the experiment design was powerful enough to detect the effects we are investigating, we carried out post-hoc power analyses using the simr package \citep{simr} in R.
+The target of the analysis was to calculate the power for detecting an effect for our main variables of interest: vowel*context (to test the hypotheses L2-vowels and Asymmetry), and vowel*context*task (to test the hypothesis Paradigm).
+
+Power is a function of sample size and effect size.
+Thus, a key issue in power analyses is to estimate the expected effect size for the treatment/factor of interest \citep{brysbaert2018power, kumle2021estimating}.
+This may be expressed as a standardized effect size or estimate ($\beta$), Cohen's d, or partial $\eta ^2$.
+Effect sizes in simr are expressed as unstandardized effect sizes (model estimates).
+In the absence (to the best of our knowledge) of literature on typical effect sizes for spectral differences in cross-language vowel production, we made a conservative estimate of effect sizes as |$\beta$| > 0.1.
+This amounts to the assumption that the smallest effect size that is meaningful to us (smallest effect size of interest; SESOI \citep{kumle2021estimating}) is 0.1, and we want to confirm whether the study has enough power to detect such an effect, if it were to exist.
+Note that this is an arbitrary assumption; only more discussion among researchers in this field will lead to a consensus on what is considered a typical cross-language effect.
+In the  Calculations of d values from our models using the package EMAtools \citep{ematools} show that they range from 0.01 to 1.06 (no need).
+
+While differences between paradigms have been discussed in previous literature and theoretically motivated, there is no work yet which has measured differences in a controlled experimental setup, and thus no reported effect sizes to act as a guideline.
+Since we are assuming $\beta$ = 0.1 to be the smallest difference across conditions that we are interested in, we will assume that a difference of similar magnitude across paradigms will be of theoretical interest to us.
+Thus, the power analysis fixes effect sizes of both variables at $\beta$ = 0.1, and report the power.
+
+--> in cover letter: power analyses for complex mixed models are not trivial and is an active pursuit.
+While we have done our best to follow available literature, we would love to have more feedback on this.
+\subsubsection*{transfer term: vowel*context}
+To calculate the power for the interaction term using simr, we set the estimate for the term to 0.05, and compare the full model against a model that lacks the interaction term.
+On the basis of 100 simulations, the power for
+
+\subsubsection*{paradigm differences: vowel*context*task}
 
@@
 
-Table \ref{table_f1_fixed_effects} summarizes the results of the fixed effects from the model for F1 at each point.
+\subsubsection*{Model output summary}
+
+First, to test hypothesis 1 (\bf{L2 transfer}), we fitted a model with vowel*time, and vowel*context as fixed effects.
+The transfer term (vowel*context) is significant in this model (standardized effect size ($\beta$)= -2.330e-01, standardized error (SE)=  5.244e-02, t= -4.44, p=0.0002).
+Comparing this to a null model that lacks the transfer term shows that the full model is significantly better ($\chi^2(2) = 14.42, p=  0.0007$).
+To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction.
+The full model was better ($\chi^2(1) = 12.51, p= 0.0004$).
+
+To test the hypothesis \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects.
+This three-way interaction term is significant ($\beta$= -1.955e-01, SE=  5.453e-02, t= -3.59, p= 0.0003).
+We also find a significant effect of task on FFs ($\beta$=3.300e-01, SE=2.668e-02, t=12.371, p=2e-16).
+Comparing this model to null models that lack the fixed effect Task ($\chi^2(4) = 572.61, p=  2.2e-16$) shows that the full model is better.
+Because Task has an independent effect on F1, and adding interactions increases model parameters, we fitted a second null model with Task as a separate fixed effect without the interaction term to test whether the interaction term is needed.
+However, the full model was still a better fit for the data ($\chi^2(3) = 23.82, p=  2.722e-05$).
+Next, we tested two other null models to verify that the full model with the three-way interaction term is indeed the optimal model for the data, by replacing the three-way interaction term with: (i) vowel*context + task*context (to test if task only interacts with context, rather than the transfer term vowel*context).
+The full model is better ($\chi^2(2) = 23.28, p= 8.78e-06$); (ii) vowel*context + task*vowel (to test if task only interacts with vowel, rather than the transfer term vowel*context).
+The full model is better ($\chi^2(2) = 13.36, p= 0.001$).
+This shows that the three-way interaction is the best fit for the data.
+\alert{later, during interpreting part?
+--> Want to see: is the effect of task local (decreases with time)?
+Since the effect of just task on F1 is unexpected, want to see if it might be an artifact of the paradigm/stimuli} --
+Because an independent effect of task on FFs is unexpected (c.f.
+), we wanted to see if the effect might be local, i.e.
+decrease over time.
+We added a task*time interaction term to the full model, which is significant ($\beta$= -1.216e-01, SE=  1.347e-02, t= -9.028, p= 2e-16 ) and improves model fit ($\chi^2(1) = 81.22, p= 2.2e-16$).
+
+Next, to examine the dynamics of transfer, we added a three-way interaction between the transfer term (vowel*context) and time.
+This term has a marginally significant effect ($\beta$= -4.604e-02, SE=  2.692e-02, t= -1.71, p= 0.08), but does not significantly improve model fit compared to a null model lacking the vowel*context*time term ($\chi^2(2) = 3.41, p= 0.18$).
+
+Table summary:
+
+\subsubsection*{Interpretation}
+
+---older stuff:
+Table \ref{table_f1_fixed_effects} summarizes the results of the fixed effects from the model for F1.
 As expected, we observe a significant effect of gender, such that male speakers have a lower mean F1 than female speakers.
@@
 
+---end older stuff
+
 \subsection{Vowel backness -- F2}
+
+\subsubsection*{title}
+
+First, to test hypothesis 1 (\bf{L2 transfer}), we fitted a model with vowel*time, and vowel*context as fixed effects.
+The transfer term (vowel*context) is significant in this model (standardized effect size ($\beta$)= 1.234e-01, standardized error (SE)=  2.055e-02, t= 6.00, p=5.46e-06).
+Comparing this to a null model that lacks the transfer term shows that the full model is significantly better ($\chi^2(2) = 16.20, p= 0.0003$).
+To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction.
+The full model was better ($\chi^2(1) = 16.02, p= 6.245e-05$).
+
+To test the hypothesis \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects.
+While task has an independent effect on F2 ($\beta$= -8.758e-02, SE= 1.756e-02, t= -4.98, p= 6.18e-07), the three-way interaction term is not significant ($\beta$= 3.349e-02, SE= 3.565e-02, t= 0.94, p= 0.34).
+However, there are significant two-way interactions between task*context ($\beta$= -1.099e-01, SE= 2.483e-02, t= -4.427, p= 9.66e-06) and task*vowel ($\beta$= -8.983e-02, SE= 2.535e-02, t= -3.544, p= 0.000396).
+Comparing this to a null model that lacks the fixed effect Task confirms that the full model is still a better fit for the data ($\chi^2(4) = 572.61, p=  2.2e-16$), indicating that an optimal model should contain the term Task.
+To confirm whether Task interacts with any of the other terms, we fitted a model with Task as a separate fixed effect without the interaction term.
+Comparing this to the full (interaction) model shows that the latter is still a better fit ($\chi^2(3) = 44.8, p= 1.02e-09$).
+Next, we tested two other null models to verify that the three-way interaction term is indeed optimal for the data, by replacing the three-way interaction term with: (i) vowel*context + task*context (to test if task only interacts with context, rather than the transfer term vowel*context).
+The full model is better ($\chi^2(2) = 17.28, p= 0.0001$); (ii) vowel*context + task*vowel (to test if task only interacts with vowel, rather than the transfer term vowel*context).
+The full model is better ($\chi^2(2) = 28.45, p= 6.633e-07$).
+This shows that the three-way interaction is the best fit for the data.
+
+\alert{later, during interpreting part?
+--> Want to see: is the effect of task local (decreases with time)?
+Since the effect of just task on F1 is unexpected, want to see if it might be an artifact of the paradigm/stimuli} --
+Because an independent effect of task on FFs is unexpected (c.f.
+), we wanted to see if the effect might be local, i.e.
+decrease over time.
+We added a task*time interaction term to the full model, which is significant ($\beta$= 6.237e-02, SE=  8.876e-03, t= 7.027, p= 2.23e-12 ) and improves model fit ($\chi^2(1) = 49.271, p= 2.23e-12$).
+\alert{interpretation: direction of effects: tasks has negative estimate (vowels are backed in the s task).
+task*time has a positive estimate-- means that the effect of task is mitigated/reduced with increase in time}
+
+Next, to examine the dynamics of transfer, we added a three-way interaction between the transfer term (vowel*context) and time.
+This term is not significant ($\beta$= -6.274e-03, SE= 1.774e-02, t= -0.35, p= 0.72), and the addition does not significantly improve model fit compared to a null model lacking the vowel*context*time term ($\chi^2(2) = 2.2443, p= 0.32$).
+
+Thus, the optimal model is: vowel*time + vowel*context*task + task*time + (1+context|word) + (1+context+vowel|subject)
+
+Table .. summarizes the fixed effects for the optimal model for F2. Model coefficient estimates ($\beta$), standard errors, and p-values. p-vales smaller than 0.05 are in bold. (do the same for F1)
+
+[put at end of subsection: after summary of model results, say-- want to see how the direction of shift compares to corresponding Bengali categories] Figure \ref{vowels_e_b} summarizes the findings of the experiment ($F1XF2$ plots for vowel productions in the unilingual and mixed conditions, in comparison to the related Bengali categories from \cite{dutta2021}).
+The following sections discuss the results for vowel height and backness respectively.
+
+< Insert Figure \ref{vowels_e_b} about here >
+
+---older stuff:
 
@@
 			Task (corresponding to the two paradigms)              & p (Picture: cued picture-naming, s (Sentence: code-switching)     \\
-			Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)                       \\ \bottomrule
+			Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)                       \\
+			Time                                                   & continuous variable, -1.6 - 1.6 \\
+			 \bottomrule
 		\end{tabular}%
@@
 
+\newpage
+
+\begin{table}[!htbp] \centering 
+	\caption{} 
+	\label{} 
+	\begin{tabular}{@{\extracolsep{5pt}}lc} 
+		\\[-1.8ex]\hline 
+		\hline \\[-1.8ex] 
+		& \multicolumn{1}{c}{\textit{Dependent variable:}} \\ 
+		\cline{2-2} 
+		\\[-1.8ex] & f1 \\ 
+		\hline \\[-1.8ex] 
+		time & 0.139$^{***}$ \\ 
+		& (0.012) \\ 
+		& \\ 
+		vowelʌ & $-$0.147 \\ 
+		& (0.223) \\ 
+		& \\ 
+		contexte & $-$0.013 \\ 
+		& (0.068) \\ 
+		& \\ 
+		tasks & 0.330$^{***}$ \\ 
+		& (0.027) \\ 
+		& \\ 
+		time:vowelʌ & $-$0.171$^{***}$ \\ 
+		& (0.013) \\ 
+		& \\ 
+		vowelʌ:contexte & $-$0.137$^{**}$ \\ 
+		& (0.061) \\ 
+		& \\ 
+		vowelʌ:tasks & 0.009 \\ 
+		& (0.038) \\ 
+		& \\ 
+		contexte:tasks & 0.073$^{*}$ \\ 
+		& (0.038) \\ 
+		& \\ 
+		time:tasks & $-$0.122$^{***}$ \\ 
+		& (0.013) \\ 
+		& \\ 
+		vowelʌ:contexte:tasks & $-$0.196$^{***}$ \\ 
+		& (0.054) \\ 
+		& \\ 
+		Constant & $-$0.053 \\ 
+		& (0.178) \\ 
+		& \\ 
+		\hline \\[-1.8ex] 
+		Observations & 11,665 \\ 
+		Log Likelihood & $-$12,954.910 \\ 
+		Akaike Inf. Crit. & 25,951.830 \\ 
+		Bayesian Inf. Crit. & 26,106.480 \\ 
+		\hline 
+		\hline \\[-1.8ex] 
+		\textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
+	\end{tabular} 
+\end{table}
+
+\newpage
+
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+\begin{table}[]
+	\centering
+	\caption{optimal model: F1}
+	\label{table_model_f1}
+	\begin{tabular}{@{}llllll@{}}
+		\toprule
+		& Estimate   & Std. Error & t value & Pr(>|t|)    &  \\ \midrule
+		(Intercept)           & -5.347e-02 & 1.784e-01  & -0.300  & 0.76863     &  \\
+		time                  & 1.391e-01  & 1.156e-02  & 12.040  & < 2e-16 *** &  \\
+		vowelʌ                & -1.475e-01 & 2.230e-01  & -0.661  & 0.51870     &  \\
+		contexte              & -1.342e-02 & 6.836e-02  & -0.196  & 0.84627     &  \\
+		tasks                 & 3.300e-01  & 2.659e-02  & 12.414  & < 2e-16 *** &  \\
+		time:vowelʌ           & -1.709e-01 & 1.347e-02  & -12.688 & < 2e-16 *** &  \\
+		vowelʌ:contexte       & -1.371e-01 & 6.088e-02  & -2.251  & 0.03130 *   &  \\
+		vowelʌ:tasks          & 9.002e-03  & 3.850e-02  & 0.234   & 0.81512     &  \\
+		contexte:tasks        & 7.334e-02  & 3.760e-02  & 1.951   & 0.05112 .   &  \\
+		time:tasks            & -1.216e-01 & 1.347e-02  & -9.028  & < 2e-16 *** &  \\
+		vowelʌ:contexte:tasks &            &            &         &             &  \\ \bottomrule
+	\end{tabular}
+\end{table}
+
 \end{document}

### 2022-04-16 RCode/bengali_phon_acco_new.R

@@ -34,7 +34,7 @@ detach(formant_data_5)
 
 
 ## bengali baseline data
-bengali <- read.csv("baseline_bengali.csv", stringsAsFactors = TRUE, header = TRUE)
+bengali <- read.csv("RCode/baseline_bengali.csv", stringsAsFactors = TRUE, header = TRUE)
 attach(bengali)
 detach(bengali)
 bengaliAE <- subset (bengali, V1=="A"|V1=="E")

### 2022-04-21 draft_BLC/BLC_AM-ID_manuscript.tex
revision minor edits

@@
 English is one of the two official languages, used in education, law, media, as a lingua franca mainly for an educated elite in most metropolitan regions, and carries a high prestige value \citep{pandey201517, kachru1981english, tollefson2014language, kachru1983indianization}.
-%citations 
+
+\alert{note that there are large regional variations in Indian English, and the literature on IEphonology is thus not "standardized" or uncontested.
+For the present study, we focus on a variety of IE spoken by an educated elite, used in news channels, taught in classrooms (citations)}
+
 Recent research comparing the segmental and suprasegmental properties of IE spoken in different parts of the country report many commonalities (see \cite{sirsa2013effects} for a review).
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%do something%%%%%%%%%%%%%%%
 Considering the language usage patterns suggested by this demography, the present study focuses specifically on short-term phonetic interaction during mixed-language use, because:
@@
 
+\alert{discussion: this confirms that the local act of switching does cause transfer, beyond language mode.
+Most existing studies looking at local effect of switching measure VOT.
+We find that the sae happens in vowels.
+Moreover, the result of transfer is interesting-- in VOT, it can only increase or decrease, which makes it impossible to know if it is the particular consonant that becomes more like its L1 counterpart, or if there is some more system-wide shift.
+Because vowel quality measures have directionality as well as quantity, we can see finer patterns.
+We find that it is not merely a shift towards a Bengali counterpart, but rather a shift towards Bengali \textit{norms}-- a category that doesn't exist in Bengali moves towards a related Bengali category.
+What does this say about what happens during online processing and speech planning?
+--- look at discussion section of the mode vs switching papers: what do they say about this?
+Can this finding be accounted for with that?}
+
+\alert{It makes sense that a bilingual language mode makes the representations from boh languages available.
+But how are these processed during the act of using language?
+If these representtions are combined in creative ways, then are we moving between languages during a single discourse?
+Can we think of intermediate representations, not of individual sounds, but rather of sound systems?
+}
+
 \subsection{Duration of transfer effects} \label{duration}
@@
 For example, \cite[][vowel quality]{bohn1992production}
-observed that approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language transfer.
+observed that approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language influence.
 These have been interpreted as changes to long-term phonological representations.
@@
 We expect the intervening words between two subsequent targets to undo any residual effects of transfer.
+
+\alert{discussion: found that like VOT, vowel quality shifts due to switching are also highly localized, and reset in an order of seconds.
+This behavior is similar to other context-driven shifts in the literature}
 
@@
 The few existing studies on vowel quality \citep{simonet2014phonetic,muldner2019phonetics,elias2017effects} or phonological processes \citep{simonet2020increased,schwartz2015language} have examined transfer effects on L1 categories.
-As far as we are aware, there is no work yet on vowels in L2.
+As far as we are aware, there is no work yet on short-term transfer in L2 vowels.
 Thus, this study builds on the existing work by examining whether the L2 vowel quality of proficient bilinguals can be affected by mixed-language processing.
+
+\alert{discussion: more about ICM, assumptions about processing}
 
@@
 Studies that have examined multiple sound categories have found that interactions between individual sound pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep[e.g.][vowel quality]{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot treated as an atomic measure.
-In light of the discussion about VOT in section \ref{sound systems}, this is not surprising --- if a general linguistic principle of `room for movement' constrains transfer, then we should expect it to apply to individual sounds too.
+In light of the discussion about VOT in section \ref{sound systems}, this is not surprising --- if a general linguistic principle of `room for movement' and contrast constrains transfer, then we should expect it to apply to individual sounds too.
 Once again, this emphasizes the importance of examining a wider range of sound contrasts.
@@
 
+\alert{in discussion- say: fact that nt shows more shift-- shows that this influence is not just at the level of individual sounds (that way would expect more shif in the common category because it's "confusing".
+But this-- shows that the L1 influence is at a system-level.
+More inclined towards using a Bengali-like phonological system.
+Has implications for cognitive models of language and what happens during short-term interaction)}
+
 In the present study, we do not elicit unilingual Bengali productions.
-To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we used Bengali production data from five female speakers from \cite{dutta2021}.
+To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we processed and used productions from a freely available Bengali corpus SHRUTI \citep{shruticorpus}.
 Formant values were extracted from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel.
 Next, we compared these to our participants' productions in the unilingual English condition.
-Comparisons of Lobanov-normalized F1 and F2 suggest that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.
+\alert{fig.
+..
+plots} F1-F2 plots show that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.
 
@@
 Note that this is an arbitrary assumption; only more discussion among researchers in this field will lead to a consensus on what is considered a typical cross-language effect.
+
+"Determining the SESOI for (G)LMMs is difficult in a simulation-based approach where effect sizes are indicated through the model’s unstandardized beta coefficients.
+relating effect sizes to beta coefficients in complex models is far from trivial and the authors therefore refrain from making specific recommendations." \citep{kumle2021estimating}
+
+Next, following \cite{kumle2021estimating}, we ran a power analysis assuming smaller effect sizes by reducing the beta-coefficients by 15\%.
+(Use the model summary, make a vector of all beta-values, including intercept, add as argument of SESOI= )
+
 In the  Calculations of d values from our models using the package EMAtools \citep{ematools} show that they range from 0.01 to 1.06 (no need).

### 2022-05-09 RCode/transfer_analysis.Rmd
draft

@@
 library(phonR)
+library(lme4)
 library(stargazer)
 library(modelsummary)
-library(lme4)
 library(lmerTest)
@@
 library(EMAtools) # to calculate Cohen's d values from lmer objects
+
+# package mixedpower (power analysis)
+if (!require("devtools")) {
+    install.packages("devtools", dependencies = TRUE)}
+
+   devtools::install_github("DejanDraschkow/mixedpower") 
+library(mixedpower)
 
@@
                          data= temp, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
-summary(phon_acco.model)
+
+summary(phon_acco.model_f1)
 summary(rePCA(phon_acco.model))
@@
 
-stargazer(phon_acco.model)
+stargazer(phon_acco.model_f1)
 
@@
                          data= temp, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
-summary(phon_acco.model)
+summary(phon_acco.model_f2)
 summary(rePCA(phon_acco.model))
@@
 
-fixef(phon_acco.model_f1)["vowelʌ:contexte:tasks"] <- - 0.1
+# using the mixedpower package
 
-# power for interaction term: compare current model to one without the interaction term
-powerSim(phon_acco.model_f1, nsim = 20, test = fcompare(f1 ~ time*vowel + vowel*context + time*task))
-# shows 90% power
+# dummy numeric variable for subject:
+remapping_subjects <- c("f1"="01", "f2"="02", "f3"="03", "f4"="04", "f5"="05", "m1"="06", "m2"="07", "m3"="08", "m4"="09", "m5"="10")
 
-pc <- powerCurve (phon_acco.model_f1, along = "subject", test = fixed("vowelʌ:contexte:tasks"), nsim = 20)
-plot(pc)
+temp <- temp %>%
+  mutate(subject_num = remapping_subjects[as.character(subject)]) %>%
+  mutate(subject_num = as.numeric(subject_num))
+
+power_trial <- mixedpower(model = phon_acco.model_f1, data = temp,
+                          fixed_effects = c("vowel","context","time","task"),
+                          simvar = "subject_num", steps = c(5,10,15,20),
+                          critical_value = 2)
+
+power_plot <- multiplotPower(power_trial, filename = "powerplot_f1_databased.png")
+
+# next, ran a power analysis assuming smaller effect sizes by reducing the beta-coefficients by 15\%.
+(Use the model summary, make a vector of all beta-values, including intercept, add as argument of SESOI= )
+
+beta.f1 <-fixef(phon_acco.model_f1) # unname() #creates a named vector. To remove names, function: unname(beta.f1)
+
+beta.f1
+
+# setting beta-values to 85% of original
+beta.f1<- beta.f1 %>%
+  map(~ . * 0.95) %>% # map each element of the vector to the result of applying the function f to it (creates a list)
+  unlist()  # convert back to a named vector
+
+beta.f1
+
+power_f1_minus5 <- mixedpower(model = phon_acco.model_f1, data = temp,
+                          fixed_effects = c("vowel","context","time","task"),
+                          simvar = "subject_num", steps = c(5,8,10,12,15), n_sim = 500,
+                          SESOI = beta.f1, databased = F,
+                          critical_value = 2)
+
+powerplot_f1_minus5 <- multiplotPower(power_f1_minus5, filename = "powerplot_f1_minus5.png")
+
+```
+
+done: F1: -5, -15, +15
+F2: -15, +15
+
+```{r}
+
+# F2:
+
+# using the mixedpower package
+
+# dummy numeric variable for subject:
+remapping_subjects <- c("f1"="01", "f2"="02", "f3"="03", "f4"="04", "f5"="05", "m1"="06", "m2"="07", "m3"="08", "m4"="09", "m5"="10")
+
+temp <- temp %>%
+  mutate(subject_num = remapping_subjects[as.character(subject)]) %>%
+  mutate(subject_num = as.numeric(subject_num))
+
+power_f2_databased <- mixedpower(model = phon_acco.model_f2, data = temp,
+                          fixed_effects = c("vowel","context","time","task"),
+                          simvar = "subject_num", steps = c(5,10,15,20),
+                          critical_value = 2)
+
+powerplot_f2_databased <- multiplotPower(power_f2_databased, filename = "powerplot_f2_databased.png")
+
+# next, ran a power analysis assuming smaller effect sizes by reducing the beta-coefficients by 15\%.
+(Use the model summary, make a vector of all beta-values, including intercept, add as argument of SESOI= )
+
+beta.f2 <-fixef(phon_acco.model_f2) #creates a named vector. To remove names, function: unname(beta.f1)
+
+beta.f2
+
+# setting beta-values to 85% of original
+beta.f2 <- beta.f2 %>%
+  map(~ . * 1.15) %>% # map each element of the vector to the result of applying the function f to it (creates a list)
+  unlist()  # convert back to a named vector
+
+beta.f2
+
+power_f2_plus15 <- mixedpower(model = phon_acco.model_f2, data = temp,
+                          fixed_effects = c("vowel","context","time","task"),
+                          simvar = "subject_num", steps = c(5,7,10,12,15), n_sim = 500,
+                          SESOI = beta.f2, databased = F,
+                          critical_value = 2)
+
+powerplot_f2_plus15 <- multiplotPower(power_f2_plus15, filename = "powerplot_f2_plus15.png")
 
@@
 Now that you have conducted your power analysis you will need to report it in APA style.
-Thissentence (or two) is usually placed in the Results section of a research report near the start of theresults when the assumptions and adequacy of the sample size are being reported.The sentence that you write should include the following:
+This sentence (or two) is usually placed in the Results section of a research report near the start of the results when the assumptions and adequacy of the sample size are being reported.The sentence that you write should include the following:
 
@@
 - The type of power analysis that you did–either a priori or post hoc (generally researchersreport the a priori power analysis)
-- The name of the software program that you used to conduct the power analysis and acitation to the authors and date of publication (see the G\*Power website listed above forthe journal article to cite)
+- The name of the software program that you used to conduct the power analysis and a citation to the authors and date of publication (see the G\*Power website listed above forthe journal article to cite)
 - The level of power aspired to for a prior analyses, or the actual power achieved for post hocanalysesThe estimated or actual effect size used in the calculationThe alpha level used in the calculation
@@
 of controlling for possible mediating/moderating factors/subgroup analysis, etc.
+
+------
+There is literature saying that power analysis is not very relevant if an effect has already been observed (cite), and should not change the interpretation of already observed significant effects.
+Here, did a retrospective power analysis to guide how null results should be interpreted.
+say:power for vowel\*context is lower than 80%.
+Means that: for a effect of this size, if it is present, there is a 40% chance that we may not observe it in the data.
+This means that if we get a null result, we should be careful about interpreting this as an absence of effect, and future studies should instead repeat this with larger sample sizes.
+For vowel\*context\*task, however, power is more than 80%.
+Thus, we can reasonably interpret the absence of statistical significance as an absence of the effect.
+
+```{r}
+# scratch pad: additional code for manuscript revision
+
+temp <- df_lobanov %>%
+  filter(vowel == "æ")
+
+# total: 2333; ʌ: 1137, æ: 1196
+
+```

### 2022-05-09 draft_BLC/BLC_AM-ID_manuscript.tex
draft

@@
 
-\alert{discussion: this confirms that the local act of switching does cause transfer, beyond language mode.
-Most existing studies looking at local effect of switching measure VOT.
-We find that the sae happens in vowels.
-Moreover, the result of transfer is interesting-- in VOT, it can only increase or decrease, which makes it impossible to know if it is the particular consonant that becomes more like its L1 counterpart, or if there is some more system-wide shift.
-Because vowel quality measures have directionality as well as quantity, we can see finer patterns.
-We find that it is not merely a shift towards a Bengali counterpart, but rather a shift towards Bengali \textit{norms}-- a category that doesn't exist in Bengali moves towards a related Bengali category.
-What does this say about what happens during online processing and speech planning?
---- look at discussion section of the mode vs switching papers: what do they say about this?
-Can this finding be accounted for with that?}
-
-\alert{It makes sense that a bilingual language mode makes the representations from boh languages available.
-But how are these processed during the act of using language?
-If these representtions are combined in creative ways, then are we moving between languages during a single discourse?
-Can we think of intermediate representations, not of individual sounds, but rather of sound systems?
-}
-
 \subsection{Duration of transfer effects} \label{duration}
@@
 
-\alert{in discussion- say: fact that nt shows more shift-- shows that this influence is not just at the level of individual sounds (that way would expect more shif in the common category because it's "confusing".
-But this-- shows that the L1 influence is at a system-level.
-More inclined towards using a Bengali-like phonological system.
-Has implications for cognitive models of language and what happens during short-term interaction)}
-
-In the present study, we do not elicit unilingual Bengali productions.
-To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we processed and used productions from a freely available Bengali corpus SHRUTI \citep{shruticorpus}.
-Formant values were extracted from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel.
-Next, we compared these to our participants' productions in the unilingual English condition.
+In the present study, our research question concerns whether L2 vowels are produced differently in unilingual vs mixed conditions.
+We do not elicit unilingual Bengali productions.
+To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we used production data from the freely available Bengali corpus, SHRUTI \citep{shruticorpus}.
+Formant values were extracted from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, to get formant estimates from the steady state, which should most closely approximate the production target.
+Next, we compared these to our participants' productions in the unilingual English condition to understand the relative positions of the categories.
 \alert{fig.
 ..
-plots} F1-F2 plots show that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.
+plots-- from current plot- remove the dotted line.
+Filter by context, then remove group=context from figure code} F1-F2 plots show that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.
 
@@
 
-\subsection{Participants}
+\subsection{Participants} \label{participants}
 
@@
 To avoid this confound, we used Lobanov normalization (using the PhonR package \citep{phonR} in R \citep{r}, representing each speaker's formant frequencies as the scaled distance from their own mean (\alert{citation for lobanov normalization?}).
-Plotting the vowels (figure) shows a difference between male and female speakers, such that male speakers have lower formant frequencies.
-Normalizing by speaker does not get rid of this difference.
-Since gender-related differences in formant frequencies are not of interest in this study, we normalized individual values by gender (representing each speaker's values as a scaled distance from the group mean).
+Plotting the vowels \alert{(figure)} shows a difference between male and female speakers, such that male speakers have overall lower formant frequencies.
+Speaker normalization using the Lobanov method does not get rid of this difference.
+Since gender-related differences in formant frequencies are not of interest in this study, we normalized individual values by gender (representing each value as a scaled distance from the group mean).
 Statistical analyses were performed on these normalized values.
@@
 This was scaled and centered prior to analysis and treated as a continuous predictor.
-Models were fit using the \emph{lme4} package \citep{lme4}, and \emph{lmerTest} \citep{lmerTest} was used to estimate p values for individual variables.
+We fitted models using the \emph{lme4} package \citep{lme4}, and used \emph{lmerTest} \citep{lmerTest} to obtain estimates of p values for individual variables.
 For all mixed models, the alpha criterion was set at $|t| > 2 $.
-The dependent variables were Lobanov normalized F1 and F2, and we fitted separate models for these.
-The independent variables or fixed factors, along with their levels, are given in table \ref{table variables}.
+The dependent variables were normalized F1 and F2, and we fitted separate models for these.
+The independent variables or fixed factors, along with their levels, are given in table \ref{table variables} \alert{add variable type with each: factor, numerical, ...}.
 Subject and Item were treated as random factors.\\
@@
 the direction of shift, to be affected by the vowel category (c.f.
-the second hypothesis \bf{Asymmetry}).
+section \ref{asymmetry between sounds}).
 Thus, our main effect of interest is the interaction term Context*Vowel.
@@
 
-Our third hypothesis (\bf{Paradigm}) was that the language-mixing paradigm, or Task, will affect the extent and direction of transfer.
+Our third research question (\bf{Paradigm}) asks whether the language-mixing paradigm, or Task, affects the extent and direction of transfer.
 If this is the case, we expect Task to interact with the transfer term, giving a three way Task*Context*Vowel interaction.
@@
 Since we measured FFs at five points in the vowel, Time is expected to affect the FF.
-However, formant dynamics are expected to depend on the vowel category, and so we expect the interaction of Vowel*Time to predict formant frequency.
+However, the precise effect of formant dynamics is expected to depend on the vowel category.
+Thus, we expect the interaction of Vowel*Time to predict formant frequency.
 If transfer patterns change across time, then we should expect it to interact with the transfer term, giving a Context*Vowel*Time interaction.
@@
 Given our hypotheses, we included random intercepts for Subject and Item, by-subject random slopes for Vowel (formant targets for vowel categories differ across individuals) and Context (individuals differ in their response to the experimental condition), and a by-item random slope for Context (the effect of experimental conditions differ across words).
-\alert{(convergence problems with by-word random slope for context-- changing the optimizer to bobyqa solved the problem)}
 
 We tested the significance of our hypothesized explanatory variables using t-values and p-values for individual variables in the models, and tested overall model fit by comparing the full model to a corresponding null model that lacks the variable of interest using ANOVAs.
-The following subsections summarize and discuss the model outputs for F1 and F2 respectively:
-
-\subsection{Power analysis}
-We were able to recruit 10 participants.
-To confirm whether the experiment design was powerful enough to detect the effects we are investigating, we carried out post-hoc power analyses using the simr package \citep{simr} in R.
-The target of the analysis was to calculate the power for detecting an effect for our main variables of interest: vowel*context (to test the hypotheses L2-vowels and Asymmetry), and vowel*context*task (to test the hypothesis Paradigm).
-
-Power is a function of sample size and effect size.
-Thus, a key issue in power analyses is to estimate the expected effect size for the treatment/factor of interest \citep{brysbaert2018power, kumle2021estimating}.
-This may be expressed as a standardized effect size or estimate ($\beta$), Cohen's d, or partial $\eta ^2$.
-Effect sizes in simr are expressed as unstandardized effect sizes (model estimates).
-In the absence (to the best of our knowledge) of literature on typical effect sizes for spectral differences in cross-language vowel production, we made a conservative estimate of effect sizes as |$\beta$| > 0.1.
-This amounts to the assumption that the smallest effect size that is meaningful to us (smallest effect size of interest; SESOI \citep{kumle2021estimating}) is 0.1, and we want to confirm whether the study has enough power to detect such an effect, if it were to exist.
-Note that this is an arbitrary assumption; only more discussion among researchers in this field will lead to a consensus on what is considered a typical cross-language effect.
-
-"Determining the SESOI for (G)LMMs is difficult in a simulation-based approach where effect sizes are indicated through the model’s unstandardized beta coefficients.
-relating effect sizes to beta coefficients in complex models is far from trivial and the authors therefore refrain from making specific recommendations." \citep{kumle2021estimating}
-
-Next, following \cite{kumle2021estimating}, we ran a power analysis assuming smaller effect sizes by reducing the beta-coefficients by 15\%.
-(Use the model summary, make a vector of all beta-values, including intercept, add as argument of SESOI= )
-
-In the  Calculations of d values from our models using the package EMAtools \citep{ematools} show that they range from 0.01 to 1.06 (no need).
-
-While differences between paradigms have been discussed in previous literature and theoretically motivated, there is no work yet which has measured differences in a controlled experimental setup, and thus no reported effect sizes to act as a guideline.
-Since we are assuming $\beta$ = 0.1 to be the smallest difference across conditions that we are interested in, we will assume that a difference of similar magnitude across paradigms will be of theoretical interest to us.
-Thus, the power analysis fixes effect sizes of both variables at $\beta$ = 0.1, and report the power.
-
---> in cover letter: power analyses for complex mixed models are not trivial and is an active pursuit.
-While we have done our best to follow available literature, we would love to have more feedback on this.
-\subsubsection*{transfer term: vowel*context}
-To calculate the power for the interaction term using simr, we set the estimate for the term to 0.05, and compare the full model against a model that lacks the interaction term.
-On the basis of 100 simulations, the power for
-
-\subsubsection*{paradigm differences: vowel*context*task}
+The following subsections summarize and discuss the model outputs for F1 and F2 respectively.
 
@@
 First, to test hypothesis 1 (\bf{L2 transfer}), we fitted a model with vowel*time, and vowel*context as fixed effects.
-The transfer term (vowel*context) is significant in this model (standardized effect size ($\beta$)= -2.330e-01, standardized error (SE)=  5.244e-02, t= -4.44, p=0.0002).
+The transfer term (vowel*context) is significant in this model (model coefficient/ unstandardized effect size ($\beta$)= -2.330e-01, standard error (SE)=  5.244e-02, t= -4.44, p=0.0002).
 Comparing this to a null model that lacks the transfer term shows that the full model is significantly better ($\chi^2(2) = 14.42, p=  0.0007$).
@@
 This shows that the three-way interaction is the best fit for the data.
-\alert{later, during interpreting part?
---> Want to see: is the effect of task local (decreases with time)?
-Since the effect of just task on F1 is unexpected, want to see if it might be an artifact of the paradigm/stimuli} --
 Because an independent effect of task on FFs is unexpected (c.f.
@@
 
-Table summary:
+Thus, the optimal model was: f1 ~ time*vowel + vowel*context*task + time*task + (1 + context|word) + (1 + context + vowel|subject).
+Table \ref{table_f1} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model.
+\alert{p-vales smaller than 0.05 are in bold.}
 
 \subsubsection*{Interpretation}
+We extracted formant estimates from five uniformly spaced points in the vowel, from the time at 5\% into the vowel (at the start of articulation) to 45\% (at the mid-part, or steady state, which is expected to correspond to the articulatory target for the category).
+As expected, we see that F1 values are significantly predicted by Time in the model, reflecting th dynamic nature of articulation which causes acoustic variation over the course of time.
+The positive slope of the estimate for Time indicates that F1 increases, i.e.
+the two categories in the study show lowering over the course of articulation.
+Moreover, the significant Time*Vowel interaction shows that the F1 trajectory is different for the two vowels.
+This is expected, as the categories are expected to present different articulatory targets.
 
----older stuff:
-Table \ref{table_f1_fixed_effects} summarizes the results of the fixed effects from the model for F1.
-As expected, we observe a significant effect of gender, such that male speakers have a lower mean F1 than female speakers.
  Vowel is not a significant predictor of F1.
-This is surprising, given that the phonological category is expected to systematically affect vowel quality.
+This appears surprising, given that the phonological category is expected to systematically affect vowel quality.
 However, a boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two categories \nt{2, \ae} and the extent of height difference between them.
 Note that these measures are only from the English (unilingual) condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching.
-The lack of any independent effect of Context on F1 suggests that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels.
-Instead, the significant interaction between Context and Vowel indicates a lowering in \nt{2}, but not in \nt{\ae}, in the mixed-language condition.
-The results also reveal a significant effect of Task, such that the sentence task (code-switching) leads to higher F1 values in both vowels, relative to the cued picture-naming task.
-This effect is strongest at the beginning of the vowel, and disappears by 45\%.
-This is an unexpected result, since we have no principled reason to expect code-switching to induce an across-board lowering of vowels.
-As noted earlier, there is no significant interaction of Task with any of the other fixed factors.
-We used a series of ANOVAs to compare the model elaborated above with a null model that differs in the absence of the main factor which concerns our hypothesis, Vowel*Context, at each point in the vowel.
-We find that the difference is not significant at 5\%, and increases over the course of articulating the vowel.
-\\
-%ANOVA table add
+Given that estimates for fixed effects are pooled across participants, this explains the lack of reliable differences between \nt{2} and \nt{\ae}.
+The negative slope of the estimate for Context shows that F1 is lower, i.e.
+vowels are higher, in the unilingual (e) condition compared to the mixed (b) condition.
+Put more intuitively, vowels are lowered in the mixed condition.
+However, the lack significance indicates that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels.
+Instead, the significant interaction between Vowel and Context signals an asymmetrical shift in the categories \nt{2} and \nt{\ae} in the mixed-language condition.
+Specifically, the Vowel*Context interaction has a negative slope, showing that the lowering is greater in \nt{2}.
 
-< Insert Table \ref{table_f1_fixed_effects} about here >\\
+The results also reveal a significant effect of Task, such that the sentence task (code-switching) leads to higher F1 values in both vowels relative to the cued picture-naming task as: vowels are lowered in the code-switching task, independent of experimental condition.
+This is surprising, as we have no aproiri reason to expect a uniform difference in vowel quality across tasks.
+A significant interaction with Time with a positive slope of the estimate shows that this effect is local: it is strongest at the beginning of the vowel, and reduces over the course of the vowel.
+This is consistent with our assumption that the unexpected Task effect is an artifact of the paradigm, possibly resulting from coarticulatory effects with the preceding sound in the carrier phrase.
+Of greater interest to us is the significant Vowel*Context*Task interaction, which indicates a difference in transfer patterns across paradigms.
+Specifically, the negative slope of the estimate suggests that transfer in the F1 (height) dimension are greater in the code-switching (s) task compared to the picture-naming (p) task.\\
 
@@
 
----end older stuff
-
 \subsection{Vowel backness -- F2}
 
-\subsubsection*{title}
+\subsubsection*{Model output summary}
 
@@
 
-\alert{later, during interpreting part?
---> Want to see: is the effect of task local (decreases with time)?
-Since the effect of just task on F1 is unexpected, want to see if it might be an artifact of the paradigm/stimuli} --
-Because an independent effect of task on FFs is unexpected (c.f.
-), we wanted to see if the effect might be local, i.e.
+Because an independent effect of task on FFs is unexpected \alert{(c.f.
+next section)}, we wanted to see if the effect might be local, i.e.
 decrease over time.
 We added a task*time interaction term to the full model, which is significant ($\beta$= 6.237e-02, SE=  8.876e-03, t= 7.027, p= 2.23e-12 ) and improves model fit ($\chi^2(1) = 49.271, p= 2.23e-12$).
-\alert{interpretation: direction of effects: tasks has negative estimate (vowels are backed in the s task).
-task*time has a positive estimate-- means that the effect of task is mitigated/reduced with increase in time}
 
@@
 
-Thus, the optimal model is: vowel*time + vowel*context*task + task*time + (1+context|word) + (1+context+vowel|subject)
+Thus, the optimal model is: vowel*time + vowel*context*task + task*time + (1+context|word) + (1+context+vowel|subject).
+Table \ref{table_f2} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model.
+\alert{p-vales smaller than 0.05 are in bold.}
 
-Table .. summarizes the fixed effects for the optimal model for F2. Model coefficient estimates ($\beta$), standard errors, and p-values. p-vales smaller than 0.05 are in bold. (do the same for F1)
-
-[put at end of subsection: after summary of model results, say-- want to see how the direction of shift compares to corresponding Bengali categories] Figure \ref{vowels_e_b} summarizes the findings of the experiment ($F1XF2$ plots for vowel productions in the unilingual and mixed conditions, in comparison to the related Bengali categories from \cite{dutta2021}).
+Statistical analyses reveal that the vowel \nt{2} is lowered and fronted, while \nt{\ae} is fronted, in the mixed condition.
+Previous literature on phonetic transfer suggests that such shifts are not random, but rather directional and targeted.
+We want to understand how the direction of shift compares to corresponding Bengali categories.
+Figure \ref{vowels_e_b} summarizes the findings of the experiment ($F1XF2$ plots for vowel productions in the unilingual and mixed conditions, in comparison to the related Bengali categories from \cite{dutta2021}).
+This confirms our hypotheses (c.f.
+section \ref{asymmetry between sounds}) that \nt{\ae} moves towards the corresponding \nt{\ae} category in Bengali, whereas the shift in \nt{2} is towards the related Bengali category \nt{a:}.
 The following sections discuss the results for vowel height and backness respectively.
@@
 
----older stuff:
+\subsection*{Interpretation}
 
-Table \ref{table_f2_fixed_effects} summarizes the results of fixed effects from the model for F2 at each point.
-Once again, we see that male speakers have a lower mean F2 than female speakers, as expected.
+As with F1, we find that Time significantly influences F2 values in the model, reflecting the dynamic nature of articulation which causes acoustic variation over the course of time.
+Moreover, the Time*Vowel interaction shows that this trajectory differs across the two vowels: while the positive slope for Time indicates that F2 increases, i.e.
+vowels are fronted with time, the negative slope of the Time*Vowel interaction shows that this effect is smaller for \nt{2}.
+This is expected, as figure \alert{ref} confirms that the category \nt{\ae} is overall more fronted than \nt{2}.
+
+Unlike in the F1 dimension, Vowel is a significant predictor of F2, with \nt{2} having significantly lower F2, i.e. produced as more backed, than \nt{\ae}.
+
+As with vowel height, there is no significant independent effect of Context
+
 The estimates for Vowel show that excepting the beginning of the vowel, the category \nt{2} is consistently produced as more backed than \nt{\ae}.
@@
 
+\subsection{Power analysis}
+We were able to recruit 10 participants (c.f.
+section \ref{participants}).
+To confirm whether the experiment design was powerful enough to detect the effects we are investigating, we carried out post-hoc power analyses using the mixedpower package \citep{kumle2021estimating} in R.
+The target of the analysis was to calculate the power for detecting an effect for our main variables of interest: vowel*context (to test the hypotheses L2-vowels and Asymmetry), and vowel*context*task (to test the hypothesis Paradigm).
+
+Power is a function of sample size and effect size.
+Thus, a key issue in power analyses is to estimate the expected effect size for the treatment/factor of interest \citep{brysbaert2018power, kumle2021estimating}.
+This may be expressed as a standardized effect size or estimate ($\beta$), Cohen's d, or partial $\eta ^2$.
+
+Using observed effect sizes from the data for power analyses is a bad practice and has been shown to be meaningless \citep{hoenig2001abuse, gelman2019observed_effect_size}, as such "observed power" depends on the p-value.
+Therefore, we need a reasonable estimate of effect size.
+One way to do this is to use effect sizes reported in other existing literature (however, see \cite{brysbaert2018power} who recommends against using effect sizes from single published studies, as these tend to be inflated), or meta analyses.
+An alternative approach is to define a smallest effect size that is meaningful to us (smallest effect size of interest; SESOI \citep{kumle2021estimating}), and test whether the study has enough power to detect such an effect, if it were to exist.
+This needs to be theoretically motivated.
+To the best of our knowledge, there is no existing literature on typical effect sizes for spectral differences in cross-language vowel production, and not many studies that report model estimates.
+Only more discussion among researchers in this field will lead to a consensus on what is considered a typical or minimally meaningful cross-language effect.
+
+Similarly, while differences between language-switching paradigms have been informally discussed in previous literature, we are not aware of any study that has measured these differences in a controlled experimental setup, and thus no reported effect sizes to act as a guideline.
+
+Cohen (cite) arbitrarily classifies effect sizes (cohen's d values) as small, medium, and large.
+Although there ar arguments against the meaningfulness of these categories \alert{citations}, these are often used as ballpark values for power analyses in psychological studies (for a discussion, see \cite{brysbaert2018power}).
+However, this approach does not generalize well to mixed-effects analyses, particularly for complex models with multiple random effects and interactions, as is typical in psycholinguistic and bilingualism research.
+"Determining the SESOI for (G)LMMs is difficult in a simulation-based approach where effect sizes are indicated through the model’s unstandardized beta coefficients...relating effect sizes to beta coefficients in complex models is far from trivial and the authors therefore refrain from making specific recommendations.
+\citep{kumle2021estimating}".
+
+The results of a power analysis are only as meaningful as the assumptions that go into it.
+To avoid making arbitrary assumptions, we followed the approach in \cite{brysbaert2018power} (and discussed in \cite{kumle2021estimating}) of taking effect sized of the study of interest and directly modifying all the beta coefficients by a fixed percentage.
+Note that we are not assuming that doing so gets us closer to a "true effect size".
+Instead, we run simulations with a range of effect sized and report the power of our study to plausibly detect an effect of that size.
+The results are shown in the plots in figure ...
+These show power over a range of sample sizes, given the assumed effect size.
+
+--> in cover letter: power analyses for complex mixed models are not trivial and is an active pursuit.
+While we have done our best to follow available literature, we would love to have more feedback on this.
+
 \section{Discussion}
 
-With regard to our first research question \textbf{L2 vowels}, the results confirm that the L2 vowels of proficient bilinguals can be affected by mixed-language processing.
+Taken as a whole, these results bear on our original research questions in the following ways: regarding our first question, whether L2 vowels of proficient bilinguals can be affected by short-term transfer from L1 (\textbf{L2 vowels}), our results confirm that this indeed happens during mixed-language use.
+Both the common category \nt{\ae} and the L2-exclusive category \nt{2} show spectral shifts during mixed-language productions compared to unilingual productions.
+As hypothesized (\textbf{Asymmetry}), the results reveal asymmetries in the extent and direction of shifts between the two vowels.
+Specifically, \nt{2} is lowered and fronted, while \nt{\ae} is fronted, with the extent of shift being greater in \nt{2}.
+Comparision with canonical Bengali vowel productions shows that the shift in \nt{\ae} is towards the corresponding Bengali category, while that in \nt{2} is towards the Bengali category \nt{a:}.
+Regarding our third research question, whether paradigm used to elicit language switching independently affects patterns of transfer (\textbf{Paradigm}), analyzing the role of Task in the results confirms this.
+We find a greater degree of transfer during code-switching compared to the cued picture-naming task.
+However, the difference is small, and needs to be replicated (?).
+These results and their implications are discussed in the following subsections.
+
+\subsection{Phonetic transfer in L2 vowels}
+
+One of the main aims of this study is to understand short-term phonetic transfer in L2, towards understanding language processing in proficient bilinguals.
+Two interrelated questions, discussed in section \ref{introduction} (....) bear on this: (i) what causes transfer in mixed-language speech; and (ii) what causes observed asymmetries between transfer in L1 vs L2.
+Together, these serve as guidelines for how we posit that the language systems of th bilingual exist in relation to one another, and consequently, what a cognitive model should look like-- any model should be able to account for these observations.
+
+-- what causes transfer: language mode vs switching
+
+\alert{discussion: We presented data in the same session, such that the language mode (participants' overall language ``setting" remained the same).
+The results show differences between between the switched and non-switched tokens in spite of this.
+this confirms that the local act of switching does cause transfer, even in a bilingual language mode, paralleling results from \cite{olson2016role, tsui2019impact, piccinini2015voice}.
+All of these examined the local effect of switching on a temporal feature of speech, namely, VOT.
+The results of the present study indicate that the same happens in vowels.
+Moreover, the result of transfer is interesting-- in VOT, it can only increase or decrease, which makes it impossible to know if it is the particular consonant that becomes more like its L1 counterpart, or if there is some more system-wide shift.
+--* check discussion section of these three studies and see if they have talked about this --* Because vowel quality measures show directionality as well as magnitude, we can see finer patterns.
+We find that it is not merely a shift towards a Bengali counterpart, but rather a shift towards Bengali \textit{norms}-- a category that doesn't exist in Bengali moves towards a related Bengali category.
+This result parallels \cite{simonet2014phonetic}, who found the Catalan ...
+of bilingual speakers shifts towards a related L1 Spanish category ..
+in a bilingual language mode (although, this study did not examine switched speech).
+(check discussion section once) What does this say about what happens during online processing and speech planning?
+--- look at discussion section of the mode vs switching papers (those three + ones that found no local effect of switching): what do they say about this?
+Can this finding be accounted for with that?}
+
+\alert{It makes sense that a bilingual language mode makes the representations from both languages available.
+But what do these representations look like, and how are these processed during the act of using language?
+Examining speech during switching can shed light on this.
+We argue that what we see here is an example of \textit{creativity}-- beyond a simple L2-L1 mapping, there is a method to the movement such that it makes reference to entities beyond the level of the individual sound-- i.e., the system-level.
+If representations are combined in creative ways, then are we truly ``moving between" languages, in a real sense, during online processing within a single discourse?
+Can we think of intermediate representations, not of individual sounds, but rather of sound systems?
+While more detailed empirical data will eventually enable us to map out patterns of transfer during switched speech, this interpretation is in line with the kind of model proposed by \cite{khattab2013phonetic} based on observation from the use of phonological variables in bilingual children (check-- this the one where they talked about moving between systems-- spectrum of choices?).
+This visualizes/conceptualizes the bilingual language system as a \textit{continuum of sound systems}, from which the bilingual speaker is constantly choosing, guided by a variety of different pressures.
+We propose that vowels are a good feature to examine: in addition to continuous measures (as opposed to phonological variables), they also give us the 2-dimensional data (as opposed to VOT) that is suited to further testing and refining such a model.}
+
+--- differences between L1 vs L2 transfer:
+
+Studies of short-term phonetic transfer consistently find differences between transfer patterns in L1 vs L2.
+The bulk of this literature is based on the feature VOT.
+While vowels have been show to induce L2-to-L1 influence, the present study demonstrates L1-to-L2 transfer.
+
+Previous studies with VOT have often found a lack of parallel transfer from L1 to L2.
+One postulated model that is consistent with this finding is the ICM, first proposed by \cite{green1986control} at the level of the lexicon, but extended to sound systems by \cite{olson2013bilingual,tsui2019impact}.
+This proposes that ...\alert{read Olson2013 discussion, more about assumptions of ICM and consequences about models of processing}  As discussed in section \ref{causes}, vowels provide a fertile testing ground for this theory in sound systems.
+
 Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.
@@
 
-We hypothesized an asymmetrical shift between the two vowels (\textbf{Asymmetry}), owing to differences in (i) their positions in the Indian English vowel space, and (ii) the presence/absence of a corresponding category in the L1 Bengali inventory.
-The significant interaction between Vowel and Context in both F1 and F2 confirm that the experimental condition affected both height and backness differently for each vowel.
-As expected, we find a greater degree of shift in \nt{2}, possibly owing to its position in a less dense part of the IE vowel space.
+\alert{this is an example where studying different features allows us to compare the predictions of different underlying process models.
+While most studies usually focus on a single feature due to ease of creating materials, there have been studies which have measured multiple features (e.g.
+..., or no need?).
+Measuring multiple features within a single design, with the same group of participants, would give a powerful tool to compare different models at little added cost to researchers.
+This would sidestep the unavoidable confounds that come with comparing results across studies, or even replication}
+
+\subsection{Asymmetries in Phonetic Transfer}
+The literature on short-term phonetic transfer has frequently noted differences in transfer patterns between different sound categories in a languages.
+This has been attributed to both general linguistic principles such as contrast preservation, and language-specific factors such as frequency and distribution of categories.
+Examining these asymmetries more closely, and understanding their sources, helps us figure out what factors affect/moderate transfer and thus, by extension, what level of representation and processing this interaction happens at, and the nature of short-term intra-speaker variation.
+The present study examined transfer in two L2 categories, \nt{2} and \nt{\ae}. We hypothesized an asymmetrical shift between the two vowels (\textbf{Asymmetry}), owing to differences in:
+ \begin{itemize}
+ 	\item their position in the Indian English vowel space: \nt{\ae} is in a more crowded part of the vowel space
+ 	\item the presence/absence of a corresponding category in the L1 Bengali inventory
+ \end{itemize}
+
+The significant interaction between Vowel and Context in both F1 and F2 in the present study confirm that the experimental condition affected both height and backness differently for each vowel.
+Examining the model coeffieicnets ($\beta$) values shows that the transfer effect is greater for ...
+compared to ...
+(\alert{check interaction effect and unpack}) a greater degree of shift in \nt{2}, possibly owing to its position in a less dense part of the IE vowel space.
 The category \nt{2} is lowered and fronted in the mixed language condition as expected (Figure \ref{vowels_e_b}).
@@
 Thus, fronting in the mixed condition cannot be explained in relation the former, at least from the present data.
-Recall that the Bengali vowel productions used here are from a different set of speakers than our participants.
-Assuming some amount of individual differences in the precise phonetic realization of vowel categories in both languages, this could be a potential source of explanation for this result.
-Data from a larger number of speakers for both languages, as well as within-speaker comparisons of English and Bengali vowels, are required to clarify this.
+\alert{remove-- Recall that the Bengali vowel productions used here are from a different set of speakers than our participants.
+Assuming some amount of individual differences in the precise phonetic realization of vowel categories in both languages, this could be a potential source of explanation for this result.} Data from a larger number of speakers for both languages, as well as within-speaker comparisons of English and Bengali vowels, are required to clarify this.
 
+\alert{The fact that nt shows more shift-- shows that this influence is not just at the level of individual sounds (that way would expect more shift in the common category because it's "confusing".
+--check what \cite{simonet2014phonetic} says about this, about why shift happens.
+But this-- shows that the L1 influence is at a system-level.
+More inclined towards using a Bengali-like phonological system.
+Has implications for cognitive models of language and what happens during short-term interaction)}
+
+Switching paradigm
 Our third research question \textbf{Paradigm} concerned the effect of experimental paradigm on the phonetic outcome of mixed-language use.
@@
 \begin{table}[h]
-	\caption{Fixed factors}
+	\caption{Fixed effects}
 	%\centering
@@
 			\toprule
-			Factor                                                 & Levels                                                            \\ \midrule
+			Variable name                                                 & Levels                                                            \\ \midrule
 			Gender                                                 & male, female                                                      \\
@@
 			Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)                       \\
-			Time                                                   & continuous variable, -1.6 - 1.6 \\
+			Time                                                   & continuous variable, range -1.6 to 1.6, scaled and centered\\
 			 \bottomrule
@@
 \newpage
-\begin{table}[htbp] %\centering 
-	\caption{Model outputs for F1 at each of the five points in the vowel} 
-	\label{table_f1_fixed_effects} 
-	\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
-		\\[-1.8ex]\hline 
-		\hline \\[-1.8ex] 
-		& \multicolumn{5}{c}{\textit{Dependent variable:}} \\ 
-		\cline{2-6} 
-		\\[-1.8ex] & F1\_5 & F1\_15 & F1\_25 & F1\_35 & F1\_45 \\ 
-		\\[-1.8ex] & (Model 1) & (Model 2) & (Model 3) & (Model 4) & (Model 5)\\ 
-		\hline \\[-1.8ex] 
-		(Intercept) & 0.447$^{**}$ & 0.718$^{***}$ & 0.750$^{***}$ & 0.863$^{***}$ & 0.921$^{***}$ \\ 
-		& (0.211) & (0.176) & (0.153) & (0.152) & (0.150) \\ 
-		& & & & & \\ 
-		Gender(male) & $-$1.171$^{***}$ & $-$1.526$^{***}$ & $-$1.586$^{***}$ & $-$1.668$^{***}$ & $-$1.681$^{***}$ \\ 
-		& (0.237) & (0.184) & (0.168) & (0.177) & (0.175) \\ 
-		& & & & & \\ 
-		Context(e) & $-$0.010 & $-$0.015 & 0.033 & 0.025 & 0.037 \\ 
-		& (0.058) & (0.057) & (0.042) & (0.040) & (0.035) \\ 
-		& & & & & \\ 
-		Vowel(\textipa{2}) & 0.022 & $-$0.070 & $-$0.034 & $-$0.131 & $-$0.216 \\ 
-		& (0.170) & (0.175) & (0.170) & (0.174) & (0.168) \\ 
-		& & & & & \\ 
-		Task(s) & 0.319$^{***}$ & 0.235$^{***}$ & 0.172$^{**}$ & 0.133$^{*}$ & 0.102 \\ 
-		& (0.091) & (0.083) & (0.081) & (0.077) & (0.073) \\ 
-		& & & & & \\ 
-		Context(e):Vowel(\textipa{2}) & $-$0.126$^{**}$ & $-$0.138$^{**}$ & $-$0.185$^{***}$ & $-$0.191$^{***}$ & $-$0.192$^{***}$ \\ 
-		& (0.056) & (0.060) & (0.055) & (0.064) & (0.066) \\ 
-		& & & & & \\ 
-		\hline \\[-1.8ex] 
-		Observations & 2,333 & 2,333 & 2,333 & 2,333 & 2,333 \\ 
-		Log Likelihood & $-$1,771.558 & $-$1,589.316 & $-$1,278.310 & $-$1,313.795 & $-$1,380.083 \\ 
-		Akaike Inf. Crit. & 3,599.116 & 3,234.631 & 2,612.621 & 2,683.589 & 2,816.165 \\ 
-		Bayesian Inf. Crit. & 3,760.253 & 3,395.769 & 2,773.758 & 2,844.727 & 2,977.303 \\ 
-		\hline 
-		\hline \\[-1.8ex] 
-		\textit{Note:}  & \multicolumn{5}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
+
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+\begin{table}[h]
+	\centering
+	\caption{Optimal model: F1}
+	\label{table_f1}
+	\begin{tabular}{@{}lllll@{}}
+		\toprule
+		& Estimate  & Std. Error & t value & p value          \\ \midrule
+		(Intercept)           & -0.053474 & 0.178435   & -0.300  & 0.769            \\
+		time                  & 0.139140  & 0.011556   & 12.040  & \textbf{< 0.001 ***} \\
+		vowel\textipa{2}                & -0.147480 & 0.223022   & -0.661  & 0.519            \\
+		contexte              & -0.013419 & 0.068357   & -0.196  & 0.846            \\
+		tasks                 & 0.330034  & 0.026586   & 12.414  & \textbf{< 0.001 ***} \\
+		time:vowel\textipa{2}           & -0.170892 & 0.013469   & -12.688 & \textbf{< 0.001 ***} \\
+		vowel\textipa{2}:contexte       & -0.137062 & 0.060884   & -2.251  & \textbf{0.031 *}   \\
+		vowel\textipa{2}:tasks          & 0.009002  & 0.038498   & 0.234   & 0.815            \\
+		contexte:tasks        & 0.073341  & 0.037597   & 1.951   & 0.051 .          \\
+		time:tasks            & -0.121604 & 0.013469   & -9.028  & \textbf{< 0.001 ***} \\
+		vowel\textipa{2}:contexte:tasks & -0.195613 & 0.054338   & -3.600  & \textbf{< 0.001 ***} \\ \bottomrule
 	\end{tabular}
@@
 
-\begin{table}[htbp] %\centering 
-	\caption{Model outputs for F2 at each of the five points in the vowel} 
-	\label{table_f2_fixed_effects} 
-	\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
-		\\[-1.8ex]\hline 
-		\hline \\[-1.8ex] 
-		& \multicolumn{5}{c}{\textit{Dependent variable:}} \\ 
-		\cline{2-6} 
-		\\[-1.8ex] & F2\_5 & F2\_15 & F2\_25 & F2\_35 & F2\_45 \\ 
-		\\[-1.8ex] & (Model 1) & (Model 2) & (Model 3) & (Model 4) & (Model 5)\\ 
-		\hline \\[-1.8ex] 
-		(Intercept) & 0.607$^{***}$ & 0.838$^{***}$ & 0.778$^{***}$ & 0.827$^{***}$ & 0.826$^{***}$ \\ 
-		& (0.222) & (0.213) & (0.211) & (0.203) & (0.204) \\ 
-		& & & & & \\ 
-		Gender(male) & $-$0.839$^{***}$ & $-$0.891$^{***}$ & $-$0.807$^{***}$ & $-$0.750$^{***}$ & $-$0.748$^{***}$ \\ 
-		& (0.124) & (0.106) & (0.107) & (0.116) & (0.109) \\ 
-		& & & & & \\ 
-		Context(e) & 0.003 & $-$0.088$^{**}$ & $-$0.065$^{**}$ & $-$0.063 & $-$0.103$^{***}$ \\ 
-		& (0.036) & (0.036) & (0.032) & (0.044) & (0.033) \\ 
-		& & & & & \\ 
-		Vowel(\textipa{2}) & $-$0.217 & $-$0.615$^{***}$ & $-$0.651$^{***}$ & $-$0.801$^{***}$ & $-$0.836$^{***}$ \\ 
-		& (0.175) & (0.183) & (0.175) & (0.168) & (0.166) \\ 
-		& & & & & \\
-		Task(s) & $-$0.234$^{***}$ & $-$0.194$^{***}$ & $-$0.152$^{***}$ & $-$0.141$^{***}$ & $-$0.092$^{**}$ \\ 
-		& (0.029) & (0.035) & (0.050) & (0.047) & (0.045) \\ 
-		& & & & & \\
-		Context(e):Vowel(\textipa{2}) & 0.031 & 0.110$^{***}$ & 0.128$^{***}$ & 0.091$^{**}$ & 0.147$^{***}$ \\ 
-		& (0.050) & (0.041) & (0.040) & (0.046) & (0.043) \\ 
-		& & & & & \\   
-		\hline \\[-1.8ex] 
-		Observations & 2,333 & 2,333 & 2,333 & 2,333 & 2,333 \\ 
-		Log Likelihood & $-$1,283.531 & $-$1,478.494 & $-$1,174.087 & $-$1,222.682 & $-$1,019.033 \\ 
-		Akaike Inf. Crit. & 2,623.062 & 3,012.988 & 2,404.175 & 2,501.363 & 2,094.066 \\ 
-		Bayesian Inf. Crit. & 2,784.199 & 3,174.125 & 2,565.312 & 2,662.501 & 2,255.204 \\ 
-		\hline 
-		\hline \\[-1.8ex] 
-		\textit{Note:}  & \multicolumn{5}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+\begin{table}[h]
+	\centering
+	\caption{Optimal model: F2}
+	\label{table_f2}
+	\begin{tabular}{@{}lllll@{}}
+		\toprule
+		& Estimate & Std. Error & t value & p value          \\ \midrule
+		(Intercept)           & 0.23202  & 0.22100    & 1.05    & 0.304            \\
+		vowel\textipa{2}                & -0.38596 & 0.14858    & -2.60   & \textbf{0.015 *}   \\
+		time                  & 0.04287  & 0.00761    & 5.63    & \textbf{< 0.001 ***} \\
+		contexte              & -0.01720 & 0.02823    & -0.61   & 0.548            \\
+		tasks                 & -0.08758 & 0.01752    & -5.00   & \textbf{< 0.001 ***} \\
+		vowel\textipa{2}:time           & -0.16289 & 0.00888    & -18.35  & \textbf{< 0.001 ***} \\
+		vowel\textipa{2}:contexte       & 0.11043  & 0.02751    & 4.01    & \textbf{< 0.001 ***} \\
+		vowel\textipa{2}:tasks          & -0.08983 & 0.02530    & -3.55   & \textbf{< 0.001 ***} \\
+		contexte:tasks        & -0.10990 & 0.02477    & -4.44   & \textbf{< 0.001 ***} \\
+		time:tasks            & 0.06237  & 0.00888    & 7.03    & \textbf{< 0.001 ***} \\
+		vowel\textipa{2}:contexte:tasks & 0.03349  & 0.03557    & 0.94    & 0.34652          \\ \bottomrule
 	\end{tabular}
@@
 
+f1: contexte had -ve slope, vt also has -ve slope, so vt is moving up in e ie down in b condition.
+vowel*context has -ve slope, which means that vt is moving more than ae.
+Now, vowel*context*task has -ve slope, which means in the s task, compared to p, there is  more downward movement.
+So more transfer in s task.
+
+f2: vt has -ve slope, contexte has -ve slope, which means e condition e things are more back, ie vowels are fronted in b condition.
+But vowel*context has a +ve slope-- vt is less backed in the e condition that ae, ie it is less fronted in the b condition, ie ae is fronted more, or shows more transfer effect, than vt.
+vowel*context*task also has +ve slope, which means thatthe -ve effect of contexte are further reduced when the task is s, ie fronting in mixed condition is less in s task
+
 \newpage
-
 \section*{Figures}
@@
 
-\newpage
-
-\begin{table}[!htbp] \centering 
-	\caption{} 
-	\label{} 
-	\begin{tabular}{@{\extracolsep{5pt}}lc} 
-		\\[-1.8ex]\hline 
-		\hline \\[-1.8ex] 
-		& \multicolumn{1}{c}{\textit{Dependent variable:}} \\ 
-		\cline{2-2} 
-		\\[-1.8ex] & f1 \\ 
-		\hline \\[-1.8ex] 
-		time & 0.139$^{***}$ \\ 
-		& (0.012) \\ 
-		& \\ 
-		vowelʌ & $-$0.147 \\ 
-		& (0.223) \\ 
-		& \\ 
-		contexte & $-$0.013 \\ 
-		& (0.068) \\ 
-		& \\ 
-		tasks & 0.330$^{***}$ \\ 
-		& (0.027) \\ 
-		& \\ 
-		time:vowelʌ & $-$0.171$^{***}$ \\ 
-		& (0.013) \\ 
-		& \\ 
-		vowelʌ:contexte & $-$0.137$^{**}$ \\ 
-		& (0.061) \\ 
-		& \\ 
-		vowelʌ:tasks & 0.009 \\ 
-		& (0.038) \\ 
-		& \\ 
-		contexte:tasks & 0.073$^{*}$ \\ 
-		& (0.038) \\ 
-		& \\ 
-		time:tasks & $-$0.122$^{***}$ \\ 
-		& (0.013) \\ 
-		& \\ 
-		vowelʌ:contexte:tasks & $-$0.196$^{***}$ \\ 
-		& (0.054) \\ 
-		& \\ 
-		Constant & $-$0.053 \\ 
-		& (0.178) \\ 
-		& \\ 
-		\hline \\[-1.8ex] 
-		Observations & 11,665 \\ 
-		Log Likelihood & $-$12,954.910 \\ 
-		Akaike Inf. Crit. & 25,951.830 \\ 
-		Bayesian Inf. Crit. & 26,106.480 \\ 
-		\hline 
-		\hline \\[-1.8ex] 
-		\textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
-	\end{tabular} 
-\end{table}
-
-\newpage
-
-% Please add the following required packages to your document preamble:
-% \usepackage{booktabs}
-\begin{table}[]
-	\centering
-	\caption{optimal model: F1}
-	\label{table_model_f1}
-	\begin{tabular}{@{}llllll@{}}
-		\toprule
-		& Estimate   & Std. Error & t value & Pr(>|t|)    &  \\ \midrule
-		(Intercept)           & -5.347e-02 & 1.784e-01  & -0.300  & 0.76863     &  \\
-		time                  & 1.391e-01  & 1.156e-02  & 12.040  & < 2e-16 *** &  \\
-		vowelʌ                & -1.475e-01 & 2.230e-01  & -0.661  & 0.51870     &  \\
-		contexte              & -1.342e-02 & 6.836e-02  & -0.196  & 0.84627     &  \\
-		tasks                 & 3.300e-01  & 2.659e-02  & 12.414  & < 2e-16 *** &  \\
-		time:vowelʌ           & -1.709e-01 & 1.347e-02  & -12.688 & < 2e-16 *** &  \\
-		vowelʌ:contexte       & -1.371e-01 & 6.088e-02  & -2.251  & 0.03130 *   &  \\
-		vowelʌ:tasks          & 9.002e-03  & 3.850e-02  & 0.234   & 0.81512     &  \\
-		contexte:tasks        & 7.334e-02  & 3.760e-02  & 1.951   & 0.05112 .   &  \\
-		time:tasks            & -1.216e-01 & 1.347e-02  & -9.028  & < 2e-16 *** &  \\
-		vowelʌ:contexte:tasks &            &            &         &             &  \\ \bottomrule
-	\end{tabular}
-\end{table}
-
 \end{document}

### 2022-05-19 RCode/transfer_analysis.Rmd
manuscript revision: major edits, figures

@@
 library(lme4)
-library(stargazer)
-library(modelsummary)
+#library(stargazer)
+#library(modelsummary)
 library(lmerTest)
-library(pwr)
-library(simr)
-library(EMAtools) # to calculate Cohen's d values from lmer objects
+#library(pwr)
+#library(simr)
+#library(EMAtools) # to calculate Cohen's d values from lmer objects
 
@@
 
-# replacing vowel symbols with unicode:
+# for replacing vowel symbols with unicode:
 remapping_english <- c(`\\ae` = "æ", `\\vt` = "ʌ")
+remapping_bengali <- c(A = "a:", E = "æ")
 
@@
 
-# data frame with time as variable
+# data frame with time as variable (long format)
 
@@
 
-# raw FFs
+# raw FFs with time as variable
 df_time <- df %>%
@@
 
-```{r, eval=FALSE}
+```{r, include=FALSE}
 
@@
 
+# subset of dataframe with only vowels of interest (A and E) at vowel midpoints (timepoints 8-12),
+# removing outlier formant values (<300 Hz): ~ 11% of rows. In full dataset (all vowels): ~14%
+
+df_ben_AE <- df_ben %>%
+  filter(grepl("A|E", vowel), time %in% (8:12), f1>600) %>%
+  droplevels() %>%
+  mutate(time_absolute = NULL) %>%
+  relocate(gender, .after = speaker) %>%
+  mutate (vowel = remapping_bengali[as.character(vowel)])
+
 ```
@@
 
-```{r, eval=FALSE}
+```{r, eval=FALSE, include=FALSE}
 
@@
 
-```{r, eval=FALSE}
+```{r}
 
-# IPA vowel symbols
-remapping_english <- c(`\\ae` = "æ", `\\vt` = "ʌ")
-remapping_bengali <- c(A = "a:", E = "æ")
-
-# add column with unicode vowels
-df_ben_normalized$unicodevowel <- remapping_bengali[as.character(df_ben_normalized$vowel)]
-
-df_normalized$unicodevowel <- remapping_english[as.character(df_normalized$vowel)]
+# add column with unicode vowels (if needed)
+#df_normalized$unicodevowel <- remapping_english[as.character(df_normalized$vowel)]
 
@@
 xlim <- c(2500,600)
-ylim = c(1100,400)
+ylim <- c(1100,500)
 
-temp <- df_ben %>%
-  filter(time %in% (8:12), f1>300)
+## plot for baseline Bengali vs unilingual English
+df_eng_unilingual <- df_time %>%
+  filter(time %in% (3:5), context == "e")
 
-with(temp,plotVowels(f1, f2, vowel,
+with(df_ben_AE,plotVowels(f1, f2, vowel,
+                        plot.tokens = F, plot.means = TRUE, pch.tokens = vowel, pch.means = vowel, 
+                        var.col.by = NULL, var.sty.by = NULL, 
+                        cex.tokens = 0.8, cex.means = 1, alpha.means = 0.8,alpha.tokens = 0.3,
+                        ellipse.line = F, ellipse.fill = T, fill.opacity = 0.2, 
+                        poly.line = F, poly.order = c("a:","æ"),
+                        xlim = xlim, ylim = ylim, 
+                        #legend.kwd="bottomleft", legend.args = list(legend=c("baseline æ","baseline a:")), 
+                        pretty = TRUE))
+par(new=TRUE)
+
+# English vowels
+with(df_eng_unilingual, plotVowels(f1, f2, vowel, 
+                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = vowel,
+                               var.col.by = NULL, var.sty.by = NULL, 
+                               cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
+                               ellipse.line = T, ellipse.fill = F, fill.opacity = 0.1,
+                               poly.line = F, poly.order = c("æ","ʌ"),
+                               xlim = xlim, ylim = ylim, 
+                               #legend.kwd = "bottomright", 
+                               axes=FALSE, pretty = TRUE))
+
+rm(df_eng_unilingual)
+
+```
+
+```{r}
+## final result plot:
+
+df_eng_plots <- df_time %>%
+  filter(time %in% (3:5))
+
+# Bengali vowels
+with(df_ben_AE,plotVowels(f1, f2, vowel,
                         plot.tokens = F, plot.means = TRUE, pch.tokens = vowel, pch.means = vowel, 
@@
                         cex.tokens = 0.8, cex.means = 1, alpha.means = 0.8,alpha.tokens = 0.3,
-                        ellipse.line = F, ellipse.fill = T, fill.opacity = 0.3,
-                        poly.line = F, poly.order = c("A","E"),
+                        ellipse.line = F, ellipse.fill = T, fill.opacity = 0.2,
+                        poly.line = F, poly.order = c("a:","æ"),
                         xlim = xlim, ylim = ylim, 
                         legend.kwd="bottomleft", legend.args = list(legend=c("baseline æ","baseline a:")), 
-                        axes=T, pretty = TRUE))
-
-# Bengali vowels
-with(df_ben_normalized,plotVowels(f1_raw, f2_raw, vowel,
-                        plot.tokens = F, plot.means = TRUE, pch.tokens = unicodevowel, pch.means = unicodevowel, 
-                        var.col.by = vowel, var.sty.by = vowel, 
-                        cex.tokens = 0.8, cex.means = 1, alpha.means = 0.8,alpha.tokens = 0.3,
-                        ellipse.line = F, ellipse.fill = T, fill.opacity = 0.2,
-                        poly.line = F, poly.order = c("A","E"),
-                        xlim = xlim, ylim = ylim, 
-                        legend.kwd="bottomleft", legend.args = list(legend=c("baseline æ","baseline a:")), 
-                        axes=FALSE, pretty = TRUE))
+                        pretty = TRUE))
 par(new=TRUE)
@@
 # English vowels
-with(df_normalized, plotVowels(f1_raw, f2_raw, vowel, group = context, 
-                               plot.tokens = F, plot.means = TRUE , pch.tokens = unicodevowel, pch.means = unicodevowel,
-                               var.col.by = NA, var.sty.by = context, 
+with(df_eng_plots, plotVowels(f1, f2, vowel, group = context,
+                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = vowel,
+                               var.col.by = NULL, var.sty.by = context, 
                                cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
-                               ellipse.line = T,  ellipse.fill = T,
-                               poly.line = F, poly.order = c("\\ae","\\vt"),
+                               ellipse.line = T, ellipse.fill = T, fill.opacity = 0.1,
+                               poly.line = F, poly.order = c("æ","ʌ"),
                                xlim = xlim, ylim = ylim, 
                                legend.kwd = "bottomright", legend.args = list(legend=c("unilingual","mixed")),
-                               fill.opacity = 0.1, pretty = TRUE))
+                               axes=FALSE, pretty = TRUE))
+
+rm(df_eng_plots)
 
@@
 
-```{r}
+```{r, eval=FALSE, include=FALSE}
 # plot for English vowels:
@@
 
-```{r, results='hide'}
+```{r, eval=FALSE}
 
@@
 summary(phon_acco.model_f1)
-summary(rePCA(phon_acco.model))
+summary(rePCA(phon_acco.model_f1))
 
@@
 
-stargazer(phon_acco.model_f1)
+#stargazer(phon_acco.model_f1)
 
@@
 
-anova(phon_acco.model, null.model)
+anova(phon_acco.model_f1, null.model)
 
 # some weird interaction with lmerTest package. Detach that first
-stargazer(phon_acco.model)
-modelsummary(phon_acco.model)
+#stargazer(phon_acco.model)
+#modelsummary(phon_acco.model)
 
@@
 
-```{r}
+```{r,eval=FALSE}
 # F2
@@
 summary(phon_acco.model_f2)
-summary(rePCA(phon_acco.model))
+summary(rePCA(phon_acco.model_f2))
 
@@
 
-anova(phon_acco.model, null.model)
+anova(phon_acco.model_f2, null.model)
 
-stargazer(phon_acco.model)
 ```
@@
 
-```{r}
+```{r, include=FALSE, eval=FALSE}
 
@@
 beta.f1<- beta.f1 %>%
-  map(~ . * 0.95) %>% # map each element of the vector to the result of applying the function f to it (creates a list)
+  map(~ . * 1.05) %>% # map each element of the vector to the result of applying the function f to it (creates a list)
   unlist()  # convert back to a named vector
@@
 
-power_f1_minus5 <- mixedpower(model = phon_acco.model_f1, data = temp,
+power_f1_plus5 <- mixedpower(model = phon_acco.model_f1, data = temp,
                           fixed_effects = c("vowel","context","time","task"),
@@
 
-powerplot_f1_minus5 <- multiplotPower(power_f1_minus5, filename = "powerplot_f1_minus5.png")
+powerplot_f1_plus5 <- multiplotPower(power_f1_plus5, filename = "powerplot_f1_plus5.png")
 
@@
 beta.f2 <- beta.f2 %>%
-  map(~ . * 1.15) %>% # map each element of the vector to the result of applying the function f to it (creates a list)
+  map(~ . * 1.05) %>% # map each element of the vector to the result of applying the function f to it (creates a list)
   unlist()  # convert back to a named vector
@@
 
-power_f2_plus15 <- mixedpower(model = phon_acco.model_f2, data = temp,
+power_f2_plus5 <- mixedpower(model = phon_acco.model_f2, data = temp,
                           fixed_effects = c("vowel","context","time","task"),
-                          simvar = "subject_num", steps = c(5,7,10,12,15), n_sim = 500,
+                          simvar = "subject_num", steps = c(5,8,10,12,15), n_sim = 500,
                           SESOI = beta.f2, databased = F,
@@
 
-powerplot_f2_plus15 <- multiplotPower(power_f2_plus15, filename = "powerplot_f2_plus15.png")
+ powerplot_f2_plus5 <- multiplotPower(power_f2_plus5, filename = "powerplot_f2_plus5.png")
 
@@
 
-```{r}
+```{r, eval=FALSE, include=FALSE}
 # scratch pad: additional code for manuscript revision

### 2022-05-19 draft_BLC/BLC_AM-ID_manuscript.tex
manuscript revision: major edits, figures

@@
 \captionsetup[figure]{skip=0pt,singlelinecheck=off}
+\usepackage{subcaption}
 
@@
 We elicit mixed productions through two tasks: cued picture-naming and code-switching, to assess if experimental paradigm independently influences the behavior under study.
-Results reveal no difference across paradigms.
+Results reveal parallel patterns, but a slight difference in magnitude of transfer across paradigms.
 We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction, postulated discursive factors in code-switching, and the issue of comparability between paradigms in transfer studies.\\
@@
 
-\section{Introduction: Phonetic transfer}
+\section{Introduction: Phonetic transfer}\label{introduction}
 
@@
 We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction.
-The results demonstrate a dynamic effect of transfer on L2 vowels during mixed-language production, \alert{but no difference in the extent of transfer between the two paradigms --and slightly different patterns of shift among the paradigms?}.
+The results demonstrate a time-invariant effect of transfer on L2 vowels during mixed-language production, and parallel pattern but different magnitude of transfer between the two paradigms.
 We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.
@@
 
-\alert{note that there are large regional variations in Indian English, and the literature on IEphonology is thus not "standardized" or uncontested.
-For the present study, we focus on a variety of IE spoken by an educated elite, used in news channels, taught in classrooms (citations)}
-
-Recent research comparing the segmental and suprasegmental properties of IE spoken in different parts of the country report many commonalities (see \cite{sirsa2013effects} for a review).
-This suggests that in spite of the varied L1s of its speakers, IE has a target phonology that is distinct from any of these, as well as from other native varieties of English.
+Note that ``Indian English" is not a monolithic entity; there are large regional variations in the English speech of Indian speakers, and the literature on IE phonology is thus neither ``standardized" nor uncontested.
+For the present study, we focus on a variety of IE spoken by an educated elite, used in news channels, often taught as a target in classrooms, and generally considered to be free of any particular regional affiliation \citep[c.f.][who reports this variety to be ``a \textit{de-facto} norm", and thus calls it General(ized) Indian English]{masica1972sound}.
+This sound system is described in section \ref{vowel systems}.
+However, recent research comparing the segmental and suprasegmental properties of IE spoken in different parts of the country also report many commonalities (see \cite{sirsa2013effects} for a review).
+It has thus been suggested that in spite of the varied L1s of its speakers, IE has a target phonology that is distinct from any of these, as well as from other native varieties of English.
 Thus, it is fruitful to think of regional variations as resulting from L1-influence on a common underlying target.
 Excepting a small minority of specific communities \citep{pandey201517, wells1982accents, coelho1997anglo}, there are very few L1 speakers of IE.
+Thus, English is largely learned and used as an L2.
 
@@
 The vowel inventory of Western Bangla (the variety spoken by the participants in this study) consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \citep{garry2001facts}.
+The vowel chart in figure \ref{vowels_bengali} shows their distribution in the F1XF2 space.
 Note that there is no mid-central vowel category.
 
-The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \citep{wells1982accents, masica1972sound}.
-%Their distribution in the F1XF2 space is shown in (insert table 1.7, Pandey).
+The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2/3:, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \citep{wells1982accents, masica1972sound}.
+The vowel chart in figure \ref{vowels_ie} shows their distribution in the F1XF2 space.
 A single mid-central vowel corresponds to the categories \nt{2,@,3:}, which are treated as distinct in many native varieties of English \citep{nihalani1979indian,wells1982accents,hickey2005legacies,bansal1969intelligibility}.
@@
 
-\alert{discussion: more about ICM, assumptions about processing}
-
 \subsubsection*{Differences between sound systems of the languages} \label{sound systems} In addition to the cognitive factors discussed above, many researchers have attributed the observed asymmetries to language-internal factors.
@@
 In the present study, our research question concerns whether L2 vowels are produced differently in unilingual vs mixed conditions.
-We do not elicit unilingual Bengali productions.
-To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we used production data from the freely available Bengali corpus, SHRUTI \citep{shruticorpus}.
-Formant values were extracted from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, to get formant estimates from the steady state, which should most closely approximate the production target.
-Next, we compared these to our participants' productions in the unilingual English condition to understand the relative positions of the categories.
-\alert{fig.
-..
-plots-- from current plot- remove the dotted line.
-Filter by context, then remove group=context from figure code} F1-F2 plots show that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.
+Previous literature on phonetic transfer suggests that such shifts are not random, but rather directional and targeted.
+Although our primary research question merely concerns whether productions in unilingual vs mixed conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories.
+To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we processed and used existing speech data from a freely available Bengali corpus, SHRUTI \citep{shruticorpus}.
+The procedure is described below:
+
+\subsubsection{Baseline Bengali data from the SHRUTI corpus}
+
+The SHRUTI corpus is the largest free corpus of Western Bengali speech that we are aware of.
+It comprises of 21.64 hours of read speech, containing 7383 unique sentences (22012 unique words) from Bengali newspapers in different domains, covering the most frequently used words of the language.
+The speakers are 34 adults (age range 20-40, 75\% male, 25\% female) who have grown up in the Indian state of West Bengal, and speak standard colloquial Bengali.
+All of the speakers have university-level education or higher.
+Given these parallels in linguistic, regional, and socio-economic background, we believe that Bengali speech from this corpus is representative of the average target Bengali vowel categories that are of interest in this study.
+To estimate average vowel positions, we extracted formant values from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, to get estimates from the steady state, which should most closely approximate the production target.
+Next, we compared these to our participants' productions in the unilingual English condition to understand the relative positions of these categories.
+This is visualized in figure \ref{vowels_unilingual_ie_bengali}.
+
+Note that we refrained from using corpus data for average Indian English vowel positions, opting instead to use the participant data from this study, because IE has high regional variability (c.f.
+discussion in section \ref{bengali_english_in_india}).
+To the best of our knowledge, there is no freely available corpus of IE speech from Bengali speakers.
+The F1XF2 plot indicates that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.
 
@@
 \subsection{Participants} \label{participants}
-
-10 Bengali-English bilingual speakers (male=5, age range 19 to 28), completed the tasks.
-All the participants speak other languages in addition to Bengali and English.
+Participants were recruited through an open call at the EFL University campus in Hyderabad, India, circulated through social media.
+Interested volunteers filled a brief language background questionnaire.
+These responses were used to screen participants, in the manner described below.
+As noted in \ref{bengali_english_in_india}, India has a super multilingual an linguistically diverse population, and EFLU being a central university represents an extremely diverse population, as students are from different parts of the country, having both different L1s, and different varieties of English.
+Our aim in screening participants was to minimize between-participant random variability, while maintaining ecological validity considering the actual day-to-day settings in which such behaviors likely take place.
+We chose two metrics to do this: (i) where the participant grew up; (ii) number of years of formal education in Bengali.
+The first ensures that the variety of Bengali that participants are exposed to in their surroundings is similar.
+This would mean that the target representations for Bengali vowels would be expected to be similar.
+This was especially important because we were not eliciting unilingual L1 data in this study, and were using a corpus to estimate the position of Benali vowels in the formant space.
+Based on previous literature (c.f.
+section \ref{asymmetry between sounds}), we expect that L2 vowels shift towards related L1 categories during mixed language utterances.
+Thus, we wanted to make sure that differences between participants in the pattern of shift were actually due to individual differences, rather than different acoustic targets.
+Another reason to control for place of growing up was that it is a fairly reliable predictor of what languages the person was educated in, owing to how the education system works (elaborated below).
+
+The second metric is Medium of Instruction (MoI) in school.
+Most schools in India follow the Three Language Policy, wherein formal education is provided in English, Hindi, and the regional language (the idea and implementation of this policy, and its implications for regional minorities has been a matter of much public debate as well as scholarly work; for discussions see \alert{citation}).
+Sustained formal education in a language is a reliable predictor of LSRW (listening, speaking, reading, writing) skills.
+Since this study provides stimuli in the form of text, we wanted to ensure that all participants were sufficiently comfortable with reading and speaking both Bengali and English, to avoid confounds and uncontrolled variability due to reading/task effort.
+
+With these considerations in mind, we only recruited those participants who had grown up in the state of West Bengal and lived there for the majority of their lives, had parents who spoke Bengali, and had received at least 5 years of formal education in Bengali (all respondents had received 5+ years of education in English and were studying university degrees that are taught in English), and self-reported proficiency in speaking, understanding, and reading Bengali.
+
+A total of 28 bilingual speakers of Bengali and English (18 female) responded to the recruitment call.
+Using the inclusion criteria reported above, 10 volunteers (5 female, 5 male, age range 19 to 28) were invited to participate in a speech production study, and were compensated for their time.
+All participants reported normal hearing, and normal or corrected to normal vision.
+After the completion of the study, we followed up with a more detailed language background survey at a later date to learn more about the language profiles of the participants.
+This survey was a modified and consolidated version of three popular bilingualism profiling tools: the Bilingual Language Profile (BLP) \citep{blp}, the Language Experience and Proficiency Questionnaire (LEAP-Q) \citep{leap-q} which incorporates language history and self-assessed proficiency, and the Bilingual Switching Questionnaire (BSWQ) \citep{language_switching_questionnaire}.
+Together, these incorporate questions about language history, usage, attitudes, and switching experience.
+
+In addition to Bengali and English, all participants reported knowing other languages, the most common being Hindi (n=10).
+Since this situation is representative of the present population and our study concerns only Bengali-English interaction, we only focused on these languages in the survey.
+The average age of acquisition was 2.6 for English (range 0 i.e.
+since birth to 6) and 0.25 (range 0 to 2) for Bengali.
+Thus, all participants can be described as simultaneous bilinguals, having acquired both English and Bengali at a young age.
+They report being comfortable using both languages at a young age, with the average being 7.5 for English and 0.3 for Bengali (on a likert scale rating where 0=for as long as I can remember).
+As mentioned earlier, all participants had been in schools where Bengali and English were both used as Medium of Instruction (MoI).
+On average, participants had had more than 16 years (range 7 to 20+) of formal education in English and 11.8 years (range 8 to 15) in Bengali at the time of the study.
+As described in the inclusion criteria, all participants reported having spent over 15 years in a Bengali-speaking region.
 At the time of recording, they were living in or around the EFL University campus in Hyderabad, which has a linguistically diverse population.
-Speakers were selected on the basis of responses to a Language Background Questionnaire to ensure comparable LSRW (listening, speaking, reading, writing) skills in L1 and L2 across participants, and were compensated for their time.
+In terms of usage, participants show a range of language usage patterns.
+Almost 100\% of the participants report exclusively using Bengali with little to no English while interacting with family, and English at school/the workplace.
+However, language use with friends varies greatly, likely reflecting the language composition of the friend circle.
+The questionnaire asked participants to rate their proficiency in each language.
+All participants rated themselves as highly proficient in speaking and understanding both Bengali and English.
+While self-ratings for reading and writing English were at ceiling, ratings for Bengali literacy (reading and writing) showed some variation, with an average of 4.87 (on a 0-6 likert scale, range 3 to 6) for reading and 4.25 (range 3 to 6) for writing.
+This is a common situation for many educated sections of society, since institutions of higher education beyond the school level largely use English (for discussions about the language politics of higher education in India, see \alert{ \cite{ELE readings find}}).
+The language attitude section of the questionnaire reveals interesting information about the linguistic landscape of the population, and how people experience language in such a setting.
+When asked how much ``like themselves" participants feel while using each language, response were at ceiling for both languages.
+Similarly, most participants identified equally strongly with both an ``English-speaking" and a ``Bengali-speaking" culture, possibly because for most people, such linguistic-cultural identities are not clearly delineated and not exclusive of one another.
+These responses shed light on how multiple languages fit seamlessly into different facets of daily life to create a ``continuous" linguistic experience \alert{(c.f.
+plurality-- from Personal statement)}.
+In such a scenario, it is not surprising that language-switching is common, expected, as well as used is creative ways with communicative purpose.
+Responses to the section on language switching demonstrate this.
+On average, participants do not report being often unable to recall words in the target language while speaking Bengali (average rating 2.5 on a 1-6 likert scale where 1=never and 6=always) or English (average rating 2.3), indicating that switching is not strictly necessary for fulfilling lexical gaps.
+Indeed, participants do not report switching between languages without awareness (1.8) or control (1.25).
+In spite of this, participants report a proclivity to code-switch (rating 3.5).
+Importantly, a large number of participants report volition while switching languages (4.12), and many report it to be situationally motivated (3.12).
+These responses indicate that participants are highly aware of switching habits and move between languages for communicative purposes.
+This fact, combined with the linguistic situation described here and in section \ref{bengali_english_in_india}, are important to bear in mind while interpreting the results of a transfer study in the present population.
 
@@
 10 monosyllabic English words that do not contain the target vowel served as filler items.
-We avoided English words that are lexicalized loanwords in Bengali.
+We avoided English words that are already lexicalized loanwords in Bengali.
 %The list of words can be found in sec.\ref{appendix}.
 
-\subsection{Procedure}
+\subsection{Procedure} \label{procedure}
 Participants were recorded individually in a sound attenuated room. %, using a ... microphone type/description?
 Productions were recorded in Praat \citep{boersma2016praat} at a 44.1K sampling rate.
-Every participant performed four iterations each of two tasks-- cued picture-naming and code-switching.
+Each participant completed four iterations each of two tasks-- cued picture-naming and code-switching.
 Utterances in the two conditions (unilingual and mixed) were randomly interspersed within each iteration.
@@
 
-The sound preceding the target word was uniform across all sentences (the final sound in \textit{yellow} is produced as [o] in Indian English), to minimize any possible differences due to coarticulatory effects on the target vowel.
-A specifier-noun construction was used to ensure identical word order in both languages, giving a uniform prosodic context for the target word across condition.
-In presenting the mixed-language utterances, we used mixed orthography.
+Carrier sentences were chosen such that the sound preceding the target word was uniform across all sentences (the final sound in \textit{yellow} is produced as [o] in Indian English), to minimize any possible differences due to coarticulatory effects on the target vowel.
+To ensure identical word order in both languages, we chose a specifier-noun construction, as this has a parallel structure in Bengali and English.
+This gave a uniform prosodic context for the target word across conditions.
+While presenting the mixed-language utterances, we used mixed orthography.
 \\
@@
 The dependent variables were normalized F1 and F2, and we fitted separate models for these.
-The independent variables or fixed factors, along with their levels, are given in table \ref{table variables} \alert{add variable type with each: factor, numerical, ...}.
+The independent variables or fixed factors, along with their levels, are given in Table \ref{table variables}.
 Subject and Item were treated as random factors.\\
@@
 
-Our study involved two IE vowels. Vowel category is expected to systematically affect formant frequencies (FFs).
-
+Our study involved two IE vowels.
+Vowel category is expected to systematically affect formant frequencies (FFs).
+Thus, Vowel is a fixed effect in the models.
 The experimental treatment of interest is Context, i.e.
@@
 Our main research question (\bf{L2 vowels}) asks whether L2 vowel quality (formant frequencies) of proficient bilinguals is affected by the Context of utterance.
-We hypothesize that the direction of shift in the mixed condition is towards the corresponding L1 category.
-Thus, we expect the effect of Context on formant frequencies, i.e.
-the direction of shift, to be affected by the vowel category (c.f.
-section \ref{asymmetry between sounds}).
+We hypothesize that the direction of such shift in the mixed condition, if any, is towards a corresponding L1 category.
+Therefore, we expect the effect of Context on formant frequencies, i.e.
+the direction of shift, to be moderated by the vowel category (c.f.
+the discussion in Section \ref{asymmetry between sounds}, leading to our second research question \textbf{Asymmetry}).
 Thus, our main effect of interest is the interaction term Context*Vowel.
 This corresponds to the transfer pattern.
-
 Our third research question (\bf{Paradigm}) asks whether the language-mixing paradigm, or Task, affects the extent and direction of transfer.
 If this is the case, we expect Task to interact with the transfer term, giving a three way Task*Context*Vowel interaction.
-
 Since we measured FFs at five points in the vowel, Time is expected to affect the FF.
@@
 
-First, to test hypothesis 1 (\bf{L2 transfer}), we fitted a model with vowel*time, and vowel*context as fixed effects.
+First, to test RQ 1 (\bf{L2 vowels}), we fitted a model with vowel*time, and vowel*context as fixed effects.
 The transfer term (vowel*context) is significant in this model (model coefficient/ unstandardized effect size ($\beta$)= -2.330e-01, standard error (SE)=  5.244e-02, t= -4.44, p=0.0002).
@@
 We also find a significant effect of task on FFs ($\beta$=3.300e-01, SE=2.668e-02, t=12.371, p=2e-16).
-Comparing this model to null models that lack the fixed effect Task ($\chi^2(4) = 572.61, p=  2.2e-16$) shows that the full model is better.
+Comparing this model to a null model that lacks the fixed effect Task ($\chi^2(4) = 572.61, p=  2.2e-16$) shows that the full model is better.
 Because Task has an independent effect on F1, and adding interactions increases model parameters, we fitted a second null model with Task as a separate fixed effect without the interaction term to test whether the interaction term is needed.
@@
 Because an independent effect of task on FFs is unexpected (c.f.
-), we wanted to see if the effect might be local, i.e.
+model interpretation below), we wanted to examine if the effect might be local, i.e.
 decrease over time.
@@
 
-Thus, the optimal model was: f1 ~ time*vowel + vowel*context*task + time*task + (1 + context|word) + (1 + context + vowel|subject).
+Thus, the optimal model was: f1 $\sim$ time*vowel + vowel*context*task + time*task + (1 + context|word) + (1 + context + vowel|subject).
 Table \ref{table_f1} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model.
@@
 \subsubsection*{Interpretation}
-We extracted formant estimates from five uniformly spaced points in the vowel, from the time at 5\% into the vowel (at the start of articulation) to 45\% (at the mid-part, or steady state, which is expected to correspond to the articulatory target for the category).
-As expected, we see that F1 values are significantly predicted by Time in the model, reflecting th dynamic nature of articulation which causes acoustic variation over the course of time.
+We extracted formant estimates from five uniformly spaced time points in the vowel: from the time at 5\% into the vowel (at the start of articulation) to 45\% (at the mid-part, or steady state, which is expected to correspond most closely to the articulatory target for the category).
+As expected, we see that F1 values are significantly predicted by Time in the model, reflecting the dynamic nature of articulation which causes acoustic variation over the course of time.
 The positive slope of the estimate for Time indicates that F1 increases, i.e.
-the two categories in the study show lowering over the course of articulation.
+the two categories in the study show lowering, over the course of articulation.
 Moreover, the significant Time*Vowel interaction shows that the F1 trajectory is different for the two vowels.
@@
 Vowel is not a significant predictor of F1.
-This appears surprising, given that the phonological category is expected to systematically affect vowel quality.
-However, a boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two categories \nt{2, \ae} and the extent of height difference between them.
+This is initially surprising, given that the phonological category is expected to systematically affect vowel quality.
+However, a boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two IE categories \nt{2, \ae} and the extent of height difference between them.
 Note that these measures are only from the English (unilingual) condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching.
 Given that estimates for fixed effects are pooled across participants, this explains the lack of reliable differences between \nt{2} and \nt{\ae}.
+
 The negative slope of the estimate for Context shows that F1 is lower, i.e.
@@
 However, the lack significance indicates that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels.
-Instead, the significant interaction between Vowel and Context signals an asymmetrical shift in the categories \nt{2} and \nt{\ae} in the mixed-language condition.
+Instead, the significant interaction between Vowel and Context signals that the extent and direction of shift is moderated by vowel category.
+There is an asymmetrical shift in the categories \nt{2} and \nt{\ae} in the mixed-language condition.
 Specifically, the Vowel*Context interaction has a negative slope, showing that the lowering is greater in \nt{2}.
+The lack of a significant Vowel*Context*Time interaction shows that transfer patterns do not change across time.
+This indicates a uniform scaling of formant trajectories in the mixed condition, suggesting that the underlying acoustic targets for these vowels are different during unilingual vs mixed utterances.
 
@@
 A significant interaction with Time with a positive slope of the estimate shows that this effect is local: it is strongest at the beginning of the vowel, and reduces over the course of the vowel.
-This is consistent with our assumption that the unexpected Task effect is an artifact of the paradigm, possibly resulting from coarticulatory effects with the preceding sound in the carrier phrase.
+This is consistent with the interpretation that the unexpected Task effect is an artifact of the paradigm, possibly resulting from coarticulatory effects owing to the specific stimuli used in this study.
 Of greater interest to us is the significant Vowel*Context*Task interaction, which indicates a difference in transfer patterns across paradigms.
@@
 
-Because an independent effect of task on FFs is unexpected \alert{(c.f.
-next section)}, we wanted to see if the effect might be local, i.e.
+Because an independent effect of task on FFs is unexpected (c.f.
+model interpretation below), we wanted to see if the effect might be local, i.e.
 decrease over time.
@@
 
-Thus, the optimal model is: vowel*time + vowel*context*task + task*time + (1+context|word) + (1+context+vowel|subject).
+Thus, the optimal model is: f2 $\sim$ vowel*time + vowel*context*task + task*time + (1+context|word) + (1+context+vowel|subject).
 Table \ref{table_f2} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model.
@@
 
-Statistical analyses reveal that the vowel \nt{2} is lowered and fronted, while \nt{\ae} is fronted, in the mixed condition.
-Previous literature on phonetic transfer suggests that such shifts are not random, but rather directional and targeted.
-We want to understand how the direction of shift compares to corresponding Bengali categories.
-Figure \ref{vowels_e_b} summarizes the findings of the experiment ($F1XF2$ plots for vowel productions in the unilingual and mixed conditions, in comparison to the related Bengali categories from \cite{dutta2021}).
-This confirms our hypotheses (c.f.
-section \ref{asymmetry between sounds}) that \nt{\ae} moves towards the corresponding \nt{\ae} category in Bengali, whereas the shift in \nt{2} is towards the related Bengali category \nt{a:}.
-The following sections discuss the results for vowel height and backness respectively.
-
-< Insert Figure \ref{vowels_e_b} about here >
-
 \subsection*{Interpretation}
 
-As with F1, we find that Time significantly influences F2 values in the model, reflecting the dynamic nature of articulation which causes acoustic variation over the course of time.
-Moreover, the Time*Vowel interaction shows that this trajectory differs across the two vowels: while the positive slope for Time indicates that F2 increases, i.e.
+As with F1, we find that Time significantly influences F2 values in the model, reflecting the dynamic nature of speech which causes acoustic variation over the course of articulation.
+Moreover, the Time*Vowel interaction shows that this trajectory differs across the two vowels: while the positive slope for Time indicates that F2 increases i.e.
 vowels are fronted with time, the negative slope of the Time*Vowel interaction shows that this effect is smaller for \nt{2}.
@@
 
-Unlike in the F1 dimension, Vowel is a significant predictor of F2, with \nt{2} having significantly lower F2, i.e. produced as more backed, than \nt{\ae}.
-
-As with vowel height, there is no significant independent effect of Context
-
-The estimates for Vowel show that excepting the beginning of the vowel, the category \nt{2} is consistently produced as more backed than \nt{\ae}.
-Unlike vowel height, Context has an independent effect on vowel backness, such that both the vowels in this study have higher F2 values in the Bengali (mixed) condition, i.e.
-they are fronted in the mixed condition.
-A significant interaction between Vowel and Context reveals an asymmetry in the extent of fronting between the two vowels -- the shift is greater in the case of \nt{\ae}.
-We see a significant effect of Task on F2, such that vowels in the Sentence task (code-switching) were more backed than those in the picture-naming task.
-As with F1, this is an unexpected result, as we have no principled reason to expect code-switching to induce an across-board backing of vowels.
-A lack of significant interaction with Context suggests that there is no difference in the extent of shift between the two paradigms.
-The results of ANOVAs comparing the full model to a null model that lacks the Vowel*Context interaction factor reveal that as with vowel height, the change in vowel backness between the two conditions is not significant at 5\%, and increases over the course of the vowel.\\
-
-< Insert Table \ref{table_f2_fixed_effects} about here >
+Unlike in the F1 dimension, Vowel is a significant predictor of F2, with \nt{2} having significantly lower F2, i.e.
+being produced as more backed, than \nt{\ae}.
+As with vowel height, there is no significant independent effect of Context, showing that mixed language use does not lead to uniform fronting or backing across vowels.
+This is not surprising, as we expect movement to be towards the corresponding L1 category, and thus differ across vowels.
+The negative slope of the estimate indicates that on average, vowels have lower F2 i.e.
+are backed, in the unilingual (e) condition.
+In other words, vowels are fronted in the mixed condition.
+This effect is greater in \nt{\ae}, as indicated by the negative slope of the Vowel*Context interaction.
+In the F2 dimension, too, the lack of a significant Vowel*Context*Time interaction shows that transfer patterns do not change across time, suggesting that the underlying acoustic targets are different in the two conditions.
+
+We see a significant effect of Task on F2, such that vowels in the code-switching (s) task have lower F2, i.e.
+are more backed than those in the picture-naming (p) task.
+As with F1, this is an unexpected result; we have no principled reason to expect code-switching to induce an across-board backing of vowels.
+Vowels are backed in the code-switching task, independent of experimental condition.
+A significant interaction with Time, with a positive slope for the estimate, shows that this effect is local: it is strongest at the beginning of the vowel, and reduces over the course of the vowel.
+This aligns with the possibility that the unexpected Task effect is an artifact of the paradigm and the particular stimuli used.
+The Vowel*Context*Task interaction is of interest, as it relates to our research question of how the test paradigm affects transfer.
+Unlike vowel height, we do not find patterns of shift in vowel backness (F2) to differ across paradigms, as indicated by the lack of a significant effect for the fixed effect.
+
+< Insert Table \ref{table_f2} about here >
+
+.
+
+Statistical analyses reveal that the vowel \nt{2} is lowered and fronted, while \nt{\ae} is fronted, in the mixed condition.
+Based on previous literature, we expect that such shifts in vowel categories result from increased cross-language influence during mixed-language use, i.e.
+shifts are targeted with respect to L1 categories.
+As discussed in section \ref{asymmetry between sounds}, although our primary research question is about whether productions in unilingual vs mixed conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories.
+Figure \ref{vowels_unilingual_ie_bengali} shows the unilingual IE productions of the participants in relation to average positions of related vowel categories in Bengali.
+Figure \ref{vowels_e_b} overlays productions in the mixed condition over this, showing $F1XF2$ plots for IE vowel productions in the unilingual and mixed conditions, in comparison to the related Bengali categories from the SHRUTI corpus \cite{shruticorpus}.
+Given that the Bengali corpus data represent average vowel positions, and are not from the same set of speakers as the participants of this study, we refrained from including these in statistical analyses.
+However, figure \ref{vowels_e_b} provides a heuristic confirmation for our informal hypotheses in section \ref{asymmetry between sounds}) that the IE \nt{\ae} moves towards the corresponding \nt{\ae} category in Bengali, whereas the shift in \nt{2} is towards the related Bengali category \nt{a:}.
+The following sections discuss the implications of these results.
+
+< Insert Figure \ref{vowels_e_b} about here >
 
@@
 section \ref{participants}).
-To confirm whether the experiment design was powerful enough to detect the effects we are investigating, we carried out post-hoc power analyses using the mixedpower package \citep{kumle2021estimating} in R.
+To confirm whether the experiment design was powerful enough to detect the effects we are investigating, we carried out post-hoc power analyses using the \textit{mixedpower} package \citep{kumle2021estimating} in R.
 The target of the analysis was to calculate the power for detecting an effect for our main variables of interest: vowel*context (to test the hypotheses L2-vowels and Asymmetry), and vowel*context*task (to test the hypothesis Paradigm).
@@
 Thus, a key issue in power analyses is to estimate the expected effect size for the treatment/factor of interest \citep{brysbaert2018power, kumle2021estimating}.
-This may be expressed as a standardized effect size or estimate ($\beta$), Cohen's d, or partial $\eta ^2$.
-
-Using observed effect sizes from the data for power analyses is a bad practice and has been shown to be meaningless \citep{hoenig2001abuse, gelman2019observed_effect_size}, as such "observed power" depends on the p-value.
+This may be expressed by a variety of measures, including Cohen's d, $\eta ^2$, and model estimates ($\beta$).
+
+Using observed effect sizes from the data for power analyses has been shown to be meaningless \citep{hoenig2001abuse, gelman2019observed_effect_size}, since such ``observed power" is a function of the p-value.
 Therefore, we need a reasonable estimate of effect size.
@@
 An alternative approach is to define a smallest effect size that is meaningful to us (smallest effect size of interest; SESOI \citep{kumle2021estimating}), and test whether the study has enough power to detect such an effect, if it were to exist.
-This needs to be theoretically motivated.
-To the best of our knowledge, there is no existing literature on typical effect sizes for spectral differences in cross-language vowel production, and not many studies that report model estimates.
-Only more discussion among researchers in this field will lead to a consensus on what is considered a typical or minimally meaningful cross-language effect.
-
-Similarly, while differences between language-switching paradigms have been informally discussed in previous literature, we are not aware of any study that has measured these differences in a controlled experimental setup, and thus no reported effect sizes to act as a guideline.
-
-Cohen (cite) arbitrarily classifies effect sizes (cohen's d values) as small, medium, and large.
-Although there ar arguments against the meaningfulness of these categories \alert{citations}, these are often used as ballpark values for power analyses in psychological studies (for a discussion, see \cite{brysbaert2018power}).
+This value needs to be theoretically motivated.
+To the best of our knowledge, there is no existing literature on typical effect sizes for spectral differences in cross-language vowel production, and not many studies that report effect sizes.
+Only more discussion in this field will lead to a consensus on what is considered a typical or minimally meaningful cross-language effect.
+Similarly, while differences between language-switching paradigms have been informally discussed in previous literature, we are not aware of any study that has measured these differences in a controlled experimental setup, and thus no reported effect sizes for between-paradigm differences in phonetic transfer to act as a guideline.
+
+Cohen \citep{cohen1988statistical} classifies effect sizes (Cohen's d values) as small ($\leq$0.2), medium (0.5), and large ($\geq$ 0.8).
+Although the usefulness of this classification has been debated in the literature, these are often used as ballpark values for power analyses in psychological studies (for a discussion, see \cite{brysbaert2018power}).
 However, this approach does not generalize well to mixed-effects analyses, particularly for complex models with multiple random effects and interactions, as is typical in psycholinguistic and bilingualism research.
-"Determining the SESOI for (G)LMMs is difficult in a simulation-based approach where effect sizes are indicated through the model’s unstandardized beta coefficients...relating effect sizes to beta coefficients in complex models is far from trivial and the authors therefore refrain from making specific recommendations.
+``Determining the SESOI for (G)LMMs is difficult in a simulation-based approach where effect sizes are indicated through the model’s unstandardized beta coefficients...relating effect sizes to beta coefficients in complex models is far from trivial and the authors therefore refrain from making specific recommendations.
 \citep{kumle2021estimating}".
@@
 The results of a power analysis are only as meaningful as the assumptions that go into it.
-To avoid making arbitrary assumptions, we followed the approach in \cite{brysbaert2018power} (and discussed in \cite{kumle2021estimating}) of taking effect sized of the study of interest and directly modifying all the beta coefficients by a fixed percentage.
+To avoid making arbitrary assumptions, we followed the approach in \cite{brysbaert2018power} (and discussed in \cite{kumle2021estimating}) of extracting the effect sizes from the study of interest and directly modifying all the beta coefficients by a fixed percentage.
 Note that we are not assuming that doing so gets us closer to a "true effect size".
-Instead, we run simulations with a range of effect sized and report the power of our study to plausibly detect an effect of that size.
-The results are shown in the plots in figure ...
-These show power over a range of sample sizes, given the assumed effect size.
-
---> in cover letter: power analyses for complex mixed models are not trivial and is an active pursuit.
-While we have done our best to follow available literature, we would love to have more feedback on this.
+Instead, we run simulations with a range of effect sizes and report the power of our study to plausibly detect an effect of that size.
+The results are shown in the plots in figure \ref{} Each panel presents power analysis at a different effect size, and is based on 500 simulations.
+The plots show power (y-axis) over a range of sample sizes (x-axis), given the assumed effect size.
 
@@
 
-Taken as a whole, these results bear on our original research questions in the following ways: regarding our first question, whether L2 vowels of proficient bilinguals can be affected by short-term transfer from L1 (\textbf{L2 vowels}), our results confirm that this indeed happens during mixed-language use.
+Taken as a whole, these results bear on our original research questions in the following ways: regarding our first question of whether L2 vowels of proficient bilinguals can be affected by short-term transfer from L1 (\textbf{L2 vowels}), our results indicate that this indeed happens during mixed-language use.
 Both the common category \nt{\ae} and the L2-exclusive category \nt{2} show spectral shifts during mixed-language productions compared to unilingual productions.
-As hypothesized (\textbf{Asymmetry}), the results reveal asymmetries in the extent and direction of shifts between the two vowels.
+As hypothesized (c.f.
+our second research question \textbf{Asymmetry}), the results reveal asymmetries in the extent and direction of shifts between the two vowels.
 Specifically, \nt{2} is lowered and fronted, while \nt{\ae} is fronted, with the extent of shift being greater in \nt{2}.
-Comparision with canonical Bengali vowel productions shows that the shift in \nt{\ae} is towards the corresponding Bengali category, while that in \nt{2} is towards the Bengali category \nt{a:}.
-Regarding our third research question, whether paradigm used to elicit language switching independently affects patterns of transfer (\textbf{Paradigm}), analyzing the role of Task in the results confirms this.
+Comparison with canonical Bengali vowel productions shows that the shift in \nt{\ae} is towards the corresponding Bengali category \nt{\ae}, while that in \nt{2} is towards the Bengali category \nt{a:}.
+Regarding our third research question of whether the paradigm used to elicit language switching independently affects patterns of transfer (\textbf{Paradigm}), analyzing the role of Task in the results confirms this.
 We find a greater degree of transfer during code-switching compared to the cued picture-naming task.
-However, the difference is small, and needs to be replicated (?).
+However, this difference is limited to the F1 dimension, and needs to be replicated.
 These results and their implications are discussed in the following subsections.
@@
 
-One of the main aims of this study is to understand short-term phonetic transfer in L2, towards understanding language processing in proficient bilinguals.
-Two interrelated questions, discussed in section \ref{introduction} (....) bear on this: (i) what causes transfer in mixed-language speech; and (ii) what causes observed asymmetries between transfer in L1 vs L2.
-Together, these serve as guidelines for how we posit that the language systems of th bilingual exist in relation to one another, and consequently, what a cognitive model should look like-- any model should be able to account for these observations.
-
--- what causes transfer: language mode vs switching
-
-\alert{discussion: We presented data in the same session, such that the language mode (participants' overall language ``setting" remained the same).
-The results show differences between between the switched and non-switched tokens in spite of this.
-this confirms that the local act of switching does cause transfer, even in a bilingual language mode, paralleling results from \cite{olson2016role, tsui2019impact, piccinini2015voice}.
-All of these examined the local effect of switching on a temporal feature of speech, namely, VOT.
+The main aim of this study was to examine short-term phonetic transfer in L2, towards understanding language processing in proficient bilinguals.
+Two interrelated questions, discussed in section \ref{introduction}, bear on this: (i) what causes transfer in mixed-language speech; and (ii) what causes observed asymmetries between transfer in L1 vs L2.
+Together, these serve as the basis for our theories about the language systems of multilingual individuals exist in relation to one another, and consequently, what cognitive models of language should look like.
+
+\paragraph{\textbf{What causes transfer: language mode vs switching}}
+An often discussed issue in transfer research is whether cross-language interaction at the phonetic level during language mixing is triggered by the co-activation of languages that makes representations from both languages available, or by the local act of switching during online processing and articulation of those sounds.
+In this study, we presented unilingual and mixed stimuli in the same session, such that the language mode (participants' global language ``setting") remained the same throughout.
+In spite of this, results demonstrate consistent differences between switched and non-switched tokens.
+This confirms that the local act of switching does trigger transfer, even in a bilingual language mode, paralleling results from \cite{olson2016role, tsui2019impact, piccinini2015voice}.
+These studies all examined the effect of switching on a temporal feature of speech, namely voice onset time or VOT.
 The results of the present study indicate that the same happens in vowels.
-Moreover, the result of transfer is interesting-- in VOT, it can only increase or decrease, which makes it impossible to know if it is the particular consonant that becomes more like its L1 counterpart, or if there is some more system-wide shift.
---* check discussion section of these three studies and see if they have talked about this --* Because vowel quality measures show directionality as well as magnitude, we can see finer patterns.
-We find that it is not merely a shift towards a Bengali counterpart, but rather a shift towards Bengali \textit{norms}-- a category that doesn't exist in Bengali moves towards a related Bengali category.
-This result parallels \cite{simonet2014phonetic}, who found the Catalan ...
-of bilingual speakers shifts towards a related L1 Spanish category ..
-in a bilingual language mode (although, this study did not examine switched speech).
-(check discussion section once) What does this say about what happens during online processing and speech planning?
---- look at discussion section of the mode vs switching papers (those three + ones that found no local effect of switching): what do they say about this?
-Can this finding be accounted for with that?}
-
-\alert{It makes sense that a bilingual language mode makes the representations from both languages available.
-But what do these representations look like, and how are these processed during the act of using language?
+Moreover, the exact effect of transfer is interesting-- VOT is a one-dimensional measure.
+Thus, observing an increase or decrease in VOT in a mixed language condition makes it difficult to infer whether a consonant has moved towards a specific L1/L2 counterpart, or reflects a more general system-wide shift in articulatory norms.
+Since vowel quality measures show directionality as well as magnitude, we can observe finer details in the patterns of shift.
+In the present study, we find that productions of English vowels do not merely shift towards a Bengali counterpart, but rather towards a set of Bengali-like \textit{norms}: a vowel category that doesn't exist in Bengali nonetheless moves towards a related (perceptually close) Bengali category.
+This result parallels \cite{simonet2014phonetic}, who found the Catalan \nt{O} of bilingual speakers shifts towards a related Spanish category, \nt{o}, in a bilingual language mode (note that they did not examine code-switched speech, and focused on vowel height as a target of transfer).
+The study showed that co-activating multiple languages within a discourse causes a transient increase in cross-language phonetic interaction, leading to changes in both L1 and L2 (moderated by language dominance).
+This is interpreted as a follows: a bilingual mode temporarily co-activates both language systems, making both L1 and L2 representations available during processing.
+These representations interact while producing speech in such a language mode, leading to cross-language phonetic transfer.
+But what does this interaction look like, and how are these abstract representations processed during the act of producing speech?
 Examining speech during switching can shed light on this.
-We argue that what we see here is an example of \textit{creativity}-- beyond a simple L2-L1 mapping, there is a method to the movement such that it makes reference to entities beyond the level of the individual sound-- i.e., the system-level.
-If representations are combined in creative ways, then are we truly ``moving between" languages, in a real sense, during online processing within a single discourse?
-Can we think of intermediate representations, not of individual sounds, but rather of sound systems?
-While more detailed empirical data will eventually enable us to map out patterns of transfer during switched speech, this interpretation is in line with the kind of model proposed by \cite{khattab2013phonetic} based on observation from the use of phonological variables in bilingual children (check-- this the one where they talked about moving between systems-- spectrum of choices?).
-This visualizes/conceptualizes the bilingual language system as a \textit{continuum of sound systems}, from which the bilingual speaker is constantly choosing, guided by a variety of different pressures.
-We propose that vowels are a good feature to examine: in addition to continuous measures (as opposed to phonological variables), they also give us the 2-dimensional data (as opposed to VOT) that is suited to further testing and refining such a model.}
-
---- differences between L1 vs L2 transfer:
-
+
+The time course of such shifts provides an insight into this.
+Vowels in the mixed condition do not start out the same, eventually deviating in acoustics from their unilingual counterparts.
+Instead, we see a difference between the conditions right from the start of articulation.
+This suggests that such interaction involves the speech planning stage.
+The pattern of transfer observed in this study, particularly for \nt{2}, also demonstrates that beyond a simple L2-L1 mapping, transfer during online processing there makes reference to entities beyond the level of the individual sound.
+Specifically, such transfer demonstrates a system-level interaction: a sound that does not exist in the L1 repertoire nonetheless acquires a different target, and the target is related to the L1 system in a meaningful way.
+This cannot be reasonably explained as articulatory inertia carried over from uttering L1 sounds.
+It is also difficult to account for in a purely exemplar framework, as the discourse does not contain any instances of ``L1 \nt{2}" that could affect the L2 category.
+If representations are combined in creative ways during online speech processing within an utterance, then can we think of intermediate representations not just for individual sounds, but rather for sound systems?
+While more detailed empirical data will eventually enable us to clarify such questions, this approach aligns with sociophonetic understandings of bilingual phonetic variation, for example as discussed by \cite{khattab2009phonetic,khattab2013phonetic} based on the use of phonological variables by Arabic-English bilingual children.
+This visualizes the bilingual phonetic system as ``a rich store of phonological representations", comprising of both language-internal and cross-language variation, from which the bilingual speaker is constantly choosing, guided by a variety of pressures.
+We propose that vowels are a good feature to examine the plausibility of such systems as cognitive models for language processing: in addition to continuous measures (unlike phonological variables), they also provide two-dimensional data (unlike VOT) that is suited to further test and refine predictions from such a model.
+
+\paragraph{Differences between L1 vs L2 transfer:}
 Studies of short-term phonetic transfer consistently find differences between transfer patterns in L1 vs L2.
 The bulk of this literature is based on the feature VOT.
-While vowels have been show to induce L2-to-L1 influence, the present study demonstrates L1-to-L2 transfer.
-
-Previous studies with VOT have often found a lack of parallel transfer from L1 to L2.
-One postulated model that is consistent with this finding is the ICM, first proposed by \cite{green1986control} at the level of the lexicon, but extended to sound systems by \cite{olson2013bilingual,tsui2019impact}.
-This proposes that ...\alert{read Olson2013 discussion, more about assumptions of ICM and consequences about models of processing}  As discussed in section \ref{causes}, vowels provide a fertile testing ground for this theory in sound systems.
-
+While vowels have been shown to induce L2-to-L1 influence, the present study demonstrates L1-to-L2 transfer in vowels.
+
+Previous studies of language-switching have often reported phonetic transfer from L2 to L1, but a lack of parallel transfer from L1 to L2.
+One postulated model that is consistent with this finding is the Inhibitory Control Model (ICM), first proposed by \cite{green1986control} at the lexical level, but extended to sound systems by \cite{olson2013bilingual,tsui2019impact}.
+This proposes that the lack of transfer effects on L2 reflects the lower inhibition required on L2 representations in the first place, which makes switching into L2 easier, compared to the converse situation, when the high initial inhibition on L1 induces phonetic transfer while switching back into L1.
+As discussed in section \ref{causes}, vowels provide a fertile testing ground for this theory in sound systems, as they avoid the confound of independent language-specific differences between long- and short-lag VOT languages.
 Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.
-In the context of an ICM--based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, this is a significant finding: it demonstrates that the postulated lower switch cost on an L2 does not entirely mask transfer effects on L2 vowels.
-Moreover, our findings differ from \cite{tsui2019impact} who report on the basis of VOT measurements that balanced bilinguals do not evidence an effect of language-switching.
+In the context of an ICM-based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, this is a significant finding: it demonstrates that the postulated lower switch cost on an L2 does not entirely mask transfer effects on L2 vowels.
+There are a couple of possible explanations for this.
+To begin with, the present study focuses solely on L1 (Bengali) to L2 (English) transfer, and not the converse.
+It is theoretically possible that these participant are L2 English dominant.
+That would mean that switching from L1 into L2 induces a greater switch cost, leading to transfer effects on switched L2 productions.
+This would lead us to expect that if the same participants are asked to switch from English to Bengali, we would not see any evidence of transfer.
+While more empirical data could clarify this, the language background of the present participant pool makes this scenario unlikely.
+All participants report regularly using both English and Bengali in a variety of settings, and feeling equally comfortable using bot languages.
+Listening and speaking skills are expected to be most relevant to phonetic transfer in daily language use scenarios, and participants' self-ratings for these skills are at ceiling for both languages.
+One point to note is that participants rated their reading proficiency in Bengali to be lower than that in English.
+Given that the stimuli in this task were presented visually, it is a potential source of explanation.
+However, anecdotal observation suggests that phonetic transfer in L2 productions is common in daily language use too, even in the absence of reading.
+This raises questions about an ICM-based analysis of cross-language asymmetries in phonetic transfer that future research should investigate.
+
+Moreover, our findings differ from \cite{tsui2019impact}, who report on the basis of VOT measurements that balanced bilinguals do not evidence an effect of language-switching.
 \citeauthor{tsui2019impact} attribute the absence of transfer effects to greater experience with language switching in these participants, leading to lower switch-costs in both languages.
-We find that even proficient bilinguals with extensive experience of language switching demonstrate cross-language transfer effects in vowel production.
+As noted in \ref{participants}, the participants in this study are highly proficient bilinguals with extensive experience of language switching.
+Moreover, they live in an environment where they are constantly exposed to mixed-language stimuli, and are thus expected to be habituated to drawing on multiple language systems during online language processing.
+They nonetheless demonstrate cross-language transfer effects in L2 vowel production.
+One way to think about these differences is that vowels are generally more prone phonetic transfer than temporal features such as VOT.
 Thus, following from the discussion in sec.
@@
 More generally, these findings point to a need for expanding the range of features in which transfer effects are studied, in order to make robust generalizations about the mechanisms underlying observed patterns.
-
-\alert{this is an example where studying different features allows us to compare the predictions of different underlying process models.
-While most studies usually focus on a single feature due to ease of creating materials, there have been studies which have measured multiple features (e.g.
-..., or no need?).
-Measuring multiple features within a single design, with the same group of participants, would give a powerful tool to compare different models at little added cost to researchers.
-This would sidestep the unavoidable confounds that come with comparing results across studies, or even replication}
+This is an example where studying different features of speech (e.g.
+VOT vs vowel quality) allows us to compare the predictions of different underlying process models.
+Transfer studies usually focus on a single feature due to ease of creating materials.
+Measuring multiple features within a single design, with the same group of participants, promises to be a powerful tool to compare different models at little added cost to researchers.
+This would sidestep the unavoidable confounds that come with comparing results across studies, or even replication.
+
+Beyond methodological considerations, however, we argue that these differences ultimately raise a broader question about a purely inhibition-based understanding of transfer.
+In light of a compelling body of literature on the sociophonetic and communicative implications of phonetic variation along the L1-L2 continuum \citep[e.g.][]{khattab2013phonetic,agnihotri1979processes,stuart2011hybridity}, is it wise to think of transfer as resulting from an inability to suppress and separate language systems?
+Given that young children are able to use minute phonetic variation in sophisticated ways, might it not be more intuitive to think of transfer as a communication-enabling feature of language, even at a purely cognitive level?
+Research on the cognitive mechanisms underlying cross-language interaction and variation in bilingual speech more generally is still an emerging and exciting line of inquiry.
+We believe that a communication-centered approach to bilingual variation is essential to understand how bilingual language processing functions in the real world.
+To this end, native multilingual populations serve as an important source of data on language behavior that has been largely ignored, but should be harnessed.
 
@@
  	\item their position in the Indian English vowel space: \nt{\ae} is in a more crowded part of the vowel space
- 	\item the presence/absence of a corresponding category in the L1 Bengali inventory
+ 	\item the presence/absence of a corresponding category in the L1 Bengali inventory: \nt{\ae} is a common phonological category in Bengali and English, while \nt{2} does not exist as a category in Bengali
  \end{itemize}
 
-The significant interaction between Vowel and Context in both F1 and F2 in the present study confirm that the experimental condition affected both height and backness differently for each vowel.
-Examining the model coeffieicnets ($\beta$) values shows that the transfer effect is greater for ...
-compared to ...
-(\alert{check interaction effect and unpack}) a greater degree of shift in \nt{2}, possibly owing to its position in a less dense part of the IE vowel space.
-The category \nt{2} is lowered and fronted in the mixed language condition as expected (Figure \ref{vowels_e_b}).
-The direction of shift in \nt{\ae} is surprising-- it is fronted in the mixed-language condition.
-While phonological constraints on IE suggested that a shift in the F1 dimension could potentially risk the loss of contrast and is thus expected to be minimal, Figure \ref{vowels_e_b} shows that \nt{\ae} in Bengali is not fronted with respect to the English category.
-Thus, fronting in the mixed condition cannot be explained in relation the former, at least from the present data.
-\alert{remove-- Recall that the Bengali vowel productions used here are from a different set of speakers than our participants.
-Assuming some amount of individual differences in the precise phonetic realization of vowel categories in both languages, this could be a potential source of explanation for this result.} Data from a larger number of speakers for both languages, as well as within-speaker comparisons of English and Bengali vowels, are required to clarify this.
-
-\alert{The fact that nt shows more shift-- shows that this influence is not just at the level of individual sounds (that way would expect more shift in the common category because it's "confusing".
---check what \cite{simonet2014phonetic} says about this, about why shift happens.
-But this-- shows that the L1 influence is at a system-level.
-More inclined towards using a Bengali-like phonological system.
-Has implications for cognitive models of language and what happens during short-term interaction)}
-
-Switching paradigm
+The results of the study indicate that the effect of the experimental condition differed across the two vowels, showing differences in both the magnitude and direction of shift.
+Overall, the magnitude of shift is greater in \nt{2}.
+Recall (figure \ref{vowels_ie}) that the category \nt{\ae} is in a denser part of the IE vowel space than \nt{2}.
+Specifically: \nt{\ae} is already at the lowermost part of the vowel space.
+Moreover, large movements upwards in the F1 (vowel height) dimension would risk a conflation of the category with the neighboring \nt{E}.
+In contrast, \nt{2} (indicated as \nt{@} in the chart) is in the central part of the vowel space and is not flanked by any immediate neighboring categories.
+Thus, the difference in magnitude of transfer aligns with our hypothesis \textbf{Asymmetry} that the extent of shift in individual phonological categories is constrained by the linguistic principle of contrast preservation.
+An alternative way to interpret this result is that the initial distance between IE \nt{2} and its target L1 category \nt{a:} was greater than that between the IE and Bengali \nt{\ae} categories, leading to a greater visible shift in the former.
+However, our results also parallel previous studies with VOT that have consistently observed greater shifts in the long-lag VOT language (which has a greater acceptable range of VOT, i.e.
+more room for movement) compared to the short-lag VOT language, and attribute these differences to the language-specific distribution of categories \citep{antoniou2011inter, bullock2009trying}.
+This suggests that a general principle of ``room for movement" does play a role in moderating transfer.
+More broadly, these results collectively suggest that phonetic transfer (i) makes reference to system-level representations that go beyond individual sounds; and (ii) respects linguistic principles.
+This has implications both for our understanding of phonetic transfer as a phenomenon (e.g.
+on which mental representations and at what level of processing and is this occurring?) and language as a system (e.g.
+how have transfer phenomena shaped phonological systems in multilingual settings where languages have co-existed for long periods of time?
+What are the linguistic implications of increasingly cosmopolitan urban spaces and identities?).
+
+Our results also indicate asymmetry in the direction of shift between the two categories: the category \nt{2} is lowered and fronted in the mixed language condition, which \nt{\ae} is fronted.
+On the basis of previous literature indicating that both long-term and short-term shifts in phonological categories of multilingual speakers is targeted towards specific categories in the other language, we hypothesized (Section \ref{asymmetry between sounds}) that participants' IE \nt{\ae} would shift towards the Bengali \nt{\ae}, and \nt{2} towards the Bengali \nt{a:}.
+As figure \ref{vowels_e_b} indicates, comparing transfer patterns with Bengali categories from \cite{shruticorpus} suggests that this is borne out.
+The results of \nt{2} align with \citeauthor[][]{simonet2014phonetic}'s (2014) finding that a category which is not common to L1 and L2 nonetheless shifts towards an acoustically/perceptually close category.
+Once again, we see that transfer phenomena make reference to system-level representations.
+
+\subsection{Switching paradigm}
+
 Our third research question \textbf{Paradigm} concerned the effect of experimental paradigm on the phonetic outcome of mixed-language use.
@@
 We verified this by comparing productions of the same items from the same group of participants in a scripted CS paradigm and a cued picture-naming paradigm.
-We found no significant interaction of Task either with Context, or with the Vowel*Context interaction, suggesting that the extent of phonetic transfer was not affected by the paradigm.
-There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in observable transfer at all.
-This would mean that the pre-switch effects observed, for example, in \cite{muldner2019phonetics} and \cite{bullock2009trying} are better interpreted as resulting from some other aspect of the experimental setup, than the act of code-switching itself.
-While planning could additionally affect the words preceding the switch, the effects on the switched token itself do not change.
-(ii) Our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
-Thus, another possible explanation is that the discursive effects discussed by \cite{olson2013bilingual} are triggered by more naturalistic paradigms where attention to meaning and interlocutor are important.
+We found that the direction of transfer did not change across the two paradigms, with \nt{2} getting lowered and fronted, and \nt{\ae} getting fronted, in the mixed condition in both cases.
+However, there was a difference in the magnitude of transfer.
+Specifically, in the F1 dimension, the extent of phonetic transfer is greater in the code-switching task compared to the cued picture-naming task.
+This asymmetry is not seen in the F2 dimension.
+One way to think about this is that during code-switching, there is a greater degree of language co-activation because switches are being made within an utterance, where planning is required: Since participants see the entire utterance and then read it aloud, there is room for utterance planning.
+Compared to this, in the cued picture-naming paradigm participants see the language cue immediately prior the target word, inducing extremely ``local" switches and thus a smaller degree of language co-activation.
+While both of these are local switches and take place in a bilingual language mode where both languages are already co-activated, it is possible that this minute difference in time-frame induces a difference of degrees in the extent of cross-language phonetic interaction.
+This aligns with the findings of \cite{bullock2009trying}, who found evidence of transfer in pre-switch tokens, suggesting planning effects.
+In our study specifically, another factor that could have contributed to this is orthography.
+In both tasks, Bengali and English words were presented in their respective orthographies.
+In the picture-naming task, this was used to cue language-- as described in  section \ref{procedure}, participants saw one slide at a time.
+In the CS task, mixed utterances were presented in mixed orthography.
+It is possible that seeing mixed orthography, particularly for participants who are used to processing mixed-language visual input in their surroundings, triggered increased language co-activation, leading to greater cross-language interaction.
+
+We do not see any difference in the pattern of transfer, or the relative behaviors of the two categories.
+Our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
+Thus, one possible explanation for the overall lack of asymmetry is that the discursive effects discussed by \cite{olson2013bilingual} are triggered by more naturalistic paradigms where attention to meaning and interlocutor are important.
 More within-study comparisons with less structured paradigms, and spontaneous code-switching data, can clarify this.
@@
 (controlled) code-switching does not significantly alter the behavior under study.
-Thus, the choice of paradigm can be based on other considerations.
-Moreover, this suggests that (all else being equal) the results from such studies can be compared directly.
-
-The paradigm did affect vowel quality, but this applied uniformly across experimental conditions: all vowels were more backed and lowered in the code-switching task.
-Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this lowering is an artifact of the particular stimuli used here---resulting from some corarticulatory effect that was absent in the picture-naming task.
-While this is intriguing, it is beyond the scope of the present study.
-\\
+However, elicitation through code-switching offers greater planning, potentially increasing the magnitude of observable effects.
+
+%There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in observable transfer at all.
+This would mean that the pre-switch effects observed, for example, in \cite{muldner2019phonetics} and \cite{bullock2009trying} are better interpreted as resulting from some other aspect of the experimental setup, than the act of code-switching itself.
+While planning could additionally affect the words preceding the switch, the effects on the switched token itself do not change.
+(ii)
+
+An unexpected result in the present study is that the paradigm directly affected vowel quality, across experimental conditions: all vowels were more backed and lowered in the code-switching task.
+There is no principled reason to expect this.
+A closer look at the dynamics of this effect indicate that it is greatest at the beginning of articulation, and decreases steadily with time.
+Given that it is localized at the beginning of the target vowel, it seems likely that this lowering is an artifact of the particular stimuli used here-- resulting from some corarticulatory effect with the preceding word in the carrier phrase that was absent in the picture-naming task.
+While intriguing, a more detailed analysis of this effect is beyond the scope of the present study.
+
+%This study has a number of limitations: (i) the `baseline' values for Bengali vowel categories were estimated from a different set of speakers than the participants of this study, potentially obscuring the role of individual differences in the realization of vowel categories; within-speaker comparisons of L1 and L2 vowels is likely to offer a more fine-grained view of the direction of shift in relation to L1 categories.
+(ii) we used a frame sentence to elicit code-switched productions such that not all the sentences were pragmatically well-formed.
+This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study.
+Future work should address this by considering data from more naturalistic code-switching paradigms.
 
 \section{Conclusion}
-This study examined the phonetic outcome of mixed-language production in a group of highly proficient bilingual speakers of Bengali and Indian English.
+This study investigates how the phonetic systems of highly proficient bilingual speakers interact during online speech processing, by examining the phonetic outcome of mixed-language production in speakers of Bengali and Indian English.
 Specifically, we examined the spectral quality of two L2 vowel categories \nt{2} and \nt{\ae} produced in two experimental conditions: unilingual (English) and mixed (switching from Bengali to English).
-Comparisons of F1 and F2 reveal that mixed language processing temporarily increases phonetic interaction between the languages, causing \nt{2} to converge towards a related L1 category \nt{a:}.
+Comparisons of F1 and F2 across conditions reveal that mixed language processing temporarily increases phonetic interaction between the languages, causing the English \nt{\ae} of participants to converge towards the Bengali category \nt{\ae}, and \nt{2} to converge towards a related L1 category \nt{a:}.
 The extent and direction of shift is asymmetrical across the two vowel categories, which parallels findings in prior studies.
@@
 (controlled) code-switching does not significantly alter the behavior under study.
-
-This study has a number of limitations: (i) the `baseline' values for Bengali vowel categories were estimated from a different set of speakers than the participants of this study, potentially obscuring the role of individual differences in the realization of vowel categories; within-speaker comparisons of L1 and L2 vowels is likely to offer a more fine-grained view of the direction of shift in relation to L1 categories.
-(ii) we used a frame sentence to elicit code-switched productions such that not all the sentences were pragmatically well-formed.
-This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study.
-Future work should address this by considering data from more naturalistic code-switching paradigms.
 
@@
 			Variable name                                                 & Levels                                                            \\ \midrule
-			Gender                                                 & male, female                                                      \\
 			Vowel                                                  & \nt{2}, \nt{\ae} \\
-			Task (corresponding to the two paradigms)              & p (Picture: cued picture-naming, s (Sentence: code-switching)     \\
+			Task (corresponding to the two paradigms)              & p (Picture: cued picture-naming), s (Sentence: code-switching)     \\
 			Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)                       \\
@@
 
-f1: contexte had -ve slope, vt also has -ve slope, so vt is moving up in e ie down in b condition.
-vowel*context has -ve slope, which means that vt is moving more than ae.
-Now, vowel*context*task has -ve slope, which means in the s task, compared to p, there is  more downward movement.
-So more transfer in s task.
-
-f2: vt has -ve slope, contexte has -ve slope, which means e condition e things are more back, ie vowels are fronted in b condition.
-But vowel*context has a +ve slope-- vt is less backed in the e condition that ae, ie it is less fronted in the b condition, ie ae is fronted more, or shows more transfer effect, than vt.
-vowel*context*task also has +ve slope, which means thatthe -ve effect of contexte are further reduced when the task is s, ie fronting in mixed condition is less in s task
-
 \newpage
@@
 \begin{figure}[h] %\centering
+	\includegraphics[scale=1.2]{vowels_ie}
+	\caption{Vowel chart for monophthongs in General(ized) Indian English, from \cite{masica1972sound}}
+	\label{vowels_ie}
+\end{figure}
+
+\newpage
+
+\begin{figure}[h] %\centering
+	\includegraphics[scale=0.7]{vowels_bengali}
+	\caption{Vowel chart for oral monophthongs in Bengali, from \cite{ghosh2016bengali_vowels}}
+	\label{vowels_bengali}
+\end{figure}
+
+\newpage
+
+\begin{figure}[h] %\centering
+	\includegraphics[scale=1]{vowels_unilingual_ie_bengali}
+	\caption{Participants' unilingual English vowels (solid ellipse) compared to Bengali categories from \cite{shruticorpus}}
+	\label{vowels_unilingual_ie_bengali}
+\end{figure}
+
+\newpage
+
+\begin{figure}[h] %\centering
 	\includegraphics[scale=1]{vowels_e_b_final}
-	\caption{Vowel quality in unilinual English, mixed, and baseline Bengali productions}
+	\caption{Vowel quality in unilingual English, mixed, and baseline Bengali productions}
 	\label{vowels_e_b}
@@
 	\label{boxplot_F1}
+\end{figure}
+
+\newpage
+
+\begin{figure} \label{figure_power_analysis_f1}
+	\centering
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f1_minus15} 
+		\caption{Effect size = observed effect size - 15\%} \label{f1_minus15}
+	\end{subfigure}
+
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f1_minus5} 
+		\caption{Effect size = observed effect size - 5\%} \label{f1_minus5}
+	\end{subfigure}
+
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f1_plus5} 
+		\caption{Effect size = observed effect size + 5\%} \label{f1_plus5}
+	\end{subfigure}
+
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f1_plus15} 
+		\caption{Effect size = observed effect size + 15\%} \label{f1_plus15}
+	\end{subfigure}
+
+	\caption{Power analysis for fixed effects of interest at varying sample sizes: F1}
+
+\end{figure}
+
+\newpage
+
+\begin{figure} \label{figure_power_analysis_f2}
+	\centering
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f2_minus15} 
+		\caption{Effect size = observed effect size - 15\%} \label{f2_minus15}
+	\end{subfigure}
+
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f2_minus5} 
+		\caption{Effect size = observed effect size - 5\%} \label{f2_minus5}
+	\end{subfigure}
+
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f2_plus5} 
+		\caption{Effect size = observed effect size + 5\%} \label{f2_plus5}
+	\end{subfigure}
+
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f2_plus15} 
+		\caption{Effect size = observed effect size + 15\%} \label{f2_plus15}
+	\end{subfigure}
+
+	\caption{Power analysis for fixed effects of interest at varying sample sizes: F2}
+
 \end{figure}

### 2022-06-01 draft_BLC/blc_reviewer_response.txt
BLC revision final

Manuscript BLC-21-0133
Response to Reviewers

Dear Editors,

Thank you for the opportunity to submit a revised draft of our manuscript “Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals” for publication in Bilingualism: Language and Cognition. We appreciate the time and effort from your end and that of the reviewers in providing feedback on our manuscript, and in particular the thoughtful comments for improvements to the work. We have incorporated the suggestions made by the reviewers. Please see below, for a point-by-point response to the reviewers’ comments and concerns. All page numbers refer to the revised manuscript file uploaded in the portal.

Sincerely,
Auromita

------------------------------------------------

Reviewers' comments to authors:

Reviewer: 1

I will focus my comments here on a few things I believe should be improved. Let me say first that I am quite impressed by the quality of the writing and the insight demonstrated. This is a very fine project and it has many positive things going on for it. There are a number of issues the authors would do well to address, though.

1. The sample size is just too small. The study recruited only 10 participants. This is not much less than comparable studies, I know. The standards in our field, however, are quickly changing. This study sets out to compare the sizes of the effects of two experimental treatments or conditions. Effect sizes are likely to be very small in the population, if they exist at all. This is designed as a confirmatory study, rather than an exploratory study. It is one thing to explore an issue for the first or second time, and another to confirm a specific effect others believe they have found while exploring the possibility of its existence. A confirmatory experiment needs a power analyses, or at the very least the sample size must be explicitly justified. I suspect the sample is too small to detect any differences between the two treatments (picture naming vs. code-switching), and the differences between the unilingual and the bilingual trials is apparently found to be very small. The likelihood of detecting a significant difference increases with a more powerful experimental design; and if no differences are detected one may propose that none exist in the population. And, for significant differences, a more accurate estimate can be found in more powerful designs--those with larger samples. I'd like to encourage the authors to consider the possibility to return to the field and to continue to collect data, if at all feasible. Otherwise, at the very least, acknowledge the limitations openly and justify your sample size. Changes are this won't be enough, but only the editorial team can decide whether they continue to accept low-power studies or not.

Authors' response (AR): We agree that the sample size is smaller than ideal. We were constrained by resources and participant availability given a stringent inclusion criteria (discussed in response point 2).This was in order to minimize random variability in the sample, given that the population we are working with is highly variable in terms of linguistic experience. Unfortunately, it is not possible to collect more data. Our university has still not fully reopened after the pandemic, and we therefore don't have access to a lab at present. We plan to expand this study by collecting data online in the near future, but that is beyond the scope of the present study. 
We have performed a power analysis and included it in our manuscript (section 3.4, page 35). This shows that the study does have enough power to detect the effects of interest (phonetic transfer: vowel*context interaction, and paradigm differences: vowel*context*task interaction in F1). We acknowledge that the power for detecting a paradigm effect in F2 is low, given the effect sizes in the present data. We have acknowledged the sample size limitation within the text (page 36). However, we would like to note a couple of points:

i) There is a body of literature that debates the usefulness of post-hoc power analyses in interpreting the findings of a completed study where statistical analyses have already been performed:

Lakens, D. (2021, January 4). Sample Size Justification. https://doi.org/10.1525/collabra.33267
Dziak, J. J., Dierker, L. C., & Abar, B. (2020). The interpretation of statistical power after the data have been gathered. Current Psychology, 39(3), 870-877.
Lenth, R. V. (2007). Post hoc power: tables and commentary. Iowa City: Department of Statistics and Actuarial Science, University of Iowa, 1-13.
Gelman, A. (2019). Don’t calculate post-hoc power using observed estimate of effect size. Annals of surgery, 269(1), e9-e10.

This is especially important in situations where norms around expected effect sizes are yet to be established in the field, and thus cannot be readily motivated on theoretical grounds, making it necessary to rely on observed effect sizes from the data.

ii) We still do not know enough about power calculations in complex mixed effect models and effect sizes for cross-language transfer phenomena to make designing and interpreting such analyses straightforward. We have discussed this in some detail in the manuscript (page 36). Although we have done our best to conduct a power analysis with minimum bias, we are open to any methodological suggestions the reviewers might have.

iii) In light of these considerations, we want to draw your attention to the scarcity of literature on language behaviors of native multilingual populations. This makes it doubly difficult to find effect size estimates from comparable populations, and more generally to rely on existing norms/expectations while making methodological decisions. In large part, this is due to resource inequality, leading to a system that propagates more research on already well-studied languages and populations. The norms for experimental research on bilingualism have been established in a Western context. In a setting where linguistic variability at the population level is much greater than canonically studied groups, establishing comparable levels of experimental control requires significantly greater resources. For example, even after a recruitment call that already controlled for language background and education level (recruiting Bengali-English bilinguals in a university campus), we were able to include only 10 out of 28 volunteers (35.7%)  in the final study given our inclusion criteria. Note that these criteria (discussed further later) establish only a minimal level of between-participant similarity in exposure, dominance, habits, and proficiency, such as is extremely common in most contemporary transfer studies. This example demonstrates the degree of linguistic variability in the present population. In such a scenario, is it scientifically justified to apply blanket uniform standards for participant numbers across populations?  
We completely agree that power considerations are important, and will strive for more well-powered designs in future studies. But we also wish to point out that the lack of diversity in existing bilingual data is an equally large impediment to progress in our field. We believe that the norms that propagate this urgently need to change. Given the methodological uncertainties discussed above, the interpretation of this power analysis is tentative at best (c.f. discussion on page 35). On the other hand, existing results from transfer studies clearly demonstrate the need for data from a wider range of populations and language pairs (c.f. discussion in section 1, Introduction). While we respect whatever decision the editorial team makes in this regard, we write this to draw your attention to an alternative perspective on this issue.

2. We need to know a lot more about the participants. Participant information takes a single paragraph. Readers are not necessarily familiar with this particular bilingual population. What's more, there are no indicators of bilingualism or English or Bengali proficiency. The authors would do well to collect information from the participants utilizing one of the available bilingualism surveys, such as the BLP  or the LEAP-Q. Also, linguistic proficiency should be assessed somehow; perhaps a vocabulary size measure or a grammar knowledge measure. Or at lest a self-assessed proficiency questionnaire. This is of course optional, particularly in native bilingual populations. In this case, though, the reader needs to learn a lot more about the society and the bilingual conditions of both the population and the specific sample: language use patterns, ages and contexts of acquisition, literacy, origin, ages, education levels and languages of education, etc. Finally, consider the possibility of reporting the inclusion and exclusion criteria openly and explicitly.

AR: Thank you for pointing this out. As per reviewer suggestions, we have discussed the linguistic situation of the population in greater detail (section 2.1, page 20) . We had collected some language background information from participants prior to the study, which was used to screen participants. We have now collected additional information about literacy, acquisition, usage habits and attitudes (in the form of a collated survey of the BLP + LEAP-Q + Bilingual Switching Questionnaire (BSWQ)) from the participants of the study (we were able to reach 9 of the 10 original participants). We have discussed these responses in the Participants section. This language background survey includes self-ratings for proficiency, which have been shown in existing literature to correlate well with actual performance. Moreover, educational history is a reliable predictor of proficiency for this population, as detailed in the manuscript. We have reported our inclusion criteria, and discussed the motivation behind it.

3. The Discussion section is brief and shallow. I am impressed by the quality of the writing in the Introduction, but the Discussion and Conclusions are no match for it. The Discussion needs to be much longer, much deeper, it needs to engage with theory and contextual findings explicitly and in depth, it needs to interpret the specific findings and explain their implications. I suggest that the "limitations" paragraph be moved out of the last position in the article. It could perhaps be placed in the Discussion section, but not in the Conclusions section. The Conclusions section needs to focus on the positive findings: what have we learned here? What can we say after having conducted this study? Etc. Finally, regarding the Discussion section, there should be a separate subsection entitled "Summary of Findings" (of course, whether there is such a separate section or not is a personal matter, but my point is that there need to be a few paragraphs in which the authors summarize the findings (perhaps by providing standardized effect sizes) in plain English while refraining from interpreting or explaining the data. There needs to be a subsection in which the data are summarized so that the reader can remember the findings before engaging in a discussion of their implications.

AR: We agree with this feedback. Accordingly, we have now rewritten the Discussion section to be longer and more detailed, with an overall summary of findings at the beginning (page 37), and separate subsections for each of our original hypotheses. We have added a separate subsection titled “Model output summary” in the Results section where we report the fitted model, model output, and effect sizes without interpretation. This is followed by a “Model interpretation” subsection which explains the effects in plain language without theoretical interpretation – i.e.  what the numbers mean in terms of acoustics (e.g. the negative lope of the vowel*context interaction indicates that \nt is lowered to a greater extent than \ae in the mixed condition). In the Discussion section, we have interpreted the finding, discussed it in light of our original research questions and hypotheses, and discussed theoretical implications.

Except for my suggestion to keep collecting data to obtain a larger sample (as well as collecting more information from the participants, in the form of some bilingualism survey), I believe my other suggestions are relatively minor. I must say that I was quite impressed while reading the manuscript until (1) I saw the Participants section (small sample size and sparse information about the linguistic histories and experiences of the participants), and (2) I read the Discussion (brief, shallow, avoiding a deep discussion of the implications of the study and lacking a summary of findings devoid of interpretation before moving on to their interpretation). I sincerely hope the authors are willing to consider some of my comments. I believe this is, overall, a fine study.

AR: Thank you for these suggestions and feedback! We hope we have been able to address the major concerns.

------------------------------------------------------------------------------------------
Reviewer: 2

This study set out “to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant’s multilingual baseline production of L2.” The major claim (page 24) is that “Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.”
I am concerned about whether the results actually support this claim. As the ms. makes very clear, the data on baseline production of these vowels in IE was taken from a different study of five female speakers.

AR: As outlined in the Methodology section (page 19), our study had a within-participant design: we elicited IE speech from a group of bilingual speakers in two conditions-- unilingual (IE), and mixed (switching from Bengali to IE). Our statistical analyses compared productions across these two conditions. This means that participants' unilingual IE productions were compared to their own mixed productions, and were found to differ in vowel quality. Thus, the main claim that “Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.” is certainly supported by the results.
Based on previous literature, we expect that such shifts in vowel categories are not random, but rather targeted with respect to L1 categories. Although our primary research question is about whether productions in unilingual vs mixed conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories. Towards this, we used existing data to estimate average formant values of the relevant Bengali vowel categories. Initially, this data was taken from Dutta (2021). However, in light of concerns about the sample size and gender makeup of that data, we have now used a larger dataset consisting of both male and female speakers from the SHRUTI corpus[i] (c.f. response to comment 1 of reviewer 3). The distribution of vowels was found to be the same in both datasets. Page 16 of the revised manuscript contains a detailed discussion of how this data was extracted and used, including speaker information from the corpus. Note that we have clearly mentioned that this baseline Bengali data represents average formant values, and is only used as a heuristic for discussion. We have not included this data in our statistical analyses. 
[i] Das, B., Mandal, S., & Mitra, P. (2011, October). Bengali speech corpus for continuous auutomatic speech recognition system. In 2011 International conference on speech database and assessments (Oriental COCOSDA) (pp. 51-55). IEEE.

The participants in this study were ten male speakers. While we are told that the speakers in this study “were selected on the basis of responses to a Language Background Questionnaire to ensure comparable LSRW...skills in L1 and L2 across participants), we are not given any information on the criteria used in this selection, and no information beyond gender is provided for the baseline group.

AR: We have taken this feedback into account, and reported the inclusion criteria for the study, and the motivations for using those specific metrics for participant selection (section 2.1, page 20; c.f. response to comment 2 of reviewer 1). More information about the baseline Bengali data, including speaker information, was added as a separate subsection on page 16 (c.f. response to comment 1 of reviewer 3).
(We would like to note that the participants in our study were 5 male and 5 female speakers, rather than 10 male speakers, as detailed in section 2.1 (page 20).)

The ms. seems to assume that the 5 female speakers in the baseline study can be taken as representative of all L1 IE speakers, but we have no evidence for that.

AR: We reiterate that the baseline data was for L1 Bengali vowels. However, in light of concerns about the small sample size, we have now used a larger dataset of Bengali speech extracted from the SHRUTI corpus, which contains both male and female speakers, and is expected to be representative of vowel targets of educated Western Bengali speakers from India (c.f. response to comment 1 of reviewer 3).

A second claim (p. 25) is that “As expected, we find a greater degree of shift in [ʌ], possibly owing to its position in a less dense part of the IE vowel space.” This is a very interesting claim but it would be much more convincing if we were provided explicit information on the IE vowel space, in the form of a chart of all the vowels in this language. We also need to know whether acoustic differences in addition to F1 and F2 (such as duration) play a role in distinguishing vowels.

AR: Thanks for this suggestion! We have accordingly added a vowel chart for IE (page 60), and discussed the two categories of interest, [ʌ] and [æ], in greater detail in section 4.2, page 44. As for other acoustic features, while the IE vowel system uses phonological length contrast, the two categories studied here are not part of a length-contrasted minimal pair (they do not have a long or short counterpart). While [ʌ] and [æ] differ in their canonical durations, with [æ] being a longer vowel[i], we are not aware of any perception data from L1 Bengali IE speakers that could illuminate the role of duration as a perceptual cue to these categories.  
[i]Maxwell, O., & Fletcher, J. (2009). Acoustic and durational properties of Indian English vowels. World Englishes, 28(1), 52-69.

Finally, at a number of points the ms. makes reference to the lack of studies of cross-language influence on vowels (as opposed to VOT). I’m not sure this is a fair characterization. While studies of the effect of language switching on vowels may be lacking, there are certainly a number of studies investigating phonetic drift in vowels, for example:
Grace E. Oh, Susan Guion-Anderson, Katsura Aoyama, James E. Flege, Reiko Akahane-Yamada, Tsuneo Yamada. 2011. A one-year longitudinal study of English and Japanese vowel production by Japanese adults and children in an English-speaking setting. Journal of Phonetics 39: 156-167.
Natalia Kartushina and Clara D. Martin. 2019. Third-language learning affects bilinguals’ production in both their native languages: A longitudinal study of dynamic changes in L1, L2 and L3 vowel production. Journal of Phonetics 77: 100920.
Benjamin Lang and Lisa Davidson. 2019. Effects of Exposure and Vowel Space Distribution on Phonetic Drift: Evidence from American English Learners of French. Language and Speech 62:30-60.

AR: Thank you for pointing us to these references. Our discussions about the lack of studies focusing on vowels are very specifically about short-term, transient phonetic interaction, such as is triggered by language mixing or immediate language context. Such studies demonstrate local cross-language influence during speech processing, as opposed to changes to the mental representations of sound categories over longer periods. We distinguish between long- and short-term transfer early in the text (Introduction, page  3), and focus most of our literature review on the latter, unless explicitly specified otherwise. In light of the reviewer's concern, we have rephrased the text on page 10 ("all the studies of mixed-language production discussed above measure VOT, which is a temporal feature.") to indicate this more clearly.

Specific comments
page 6: I found the sentence in (ii) confusing.
AR: Original sentence: "Given that examples of an IE phonology without L1 'influence' are very rare, it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a given baseline condition."
As discussed in the preceding paragraph (page 6.), for the majority of this population, English is learned and used as an L2 to communicate with other L2 speakers. There are very limited L1 IE speakers. Thus, what is collectively understood as "Indian English" is not a monolithic entity, as it is always colored by a speaker's other languages. Examining cross-language interaction in this scenario amounts to observing an increase in the amount of such L1 "accentedness" in IE speech.

page 6: ‘monopthongs’ should be ‘monophthongs’
page 9: First sentence: “bilinguals speakers’...”
page 14: “individual sounds pairs”
AR: Thanks for pointing these out, we have edited these lines in the manuscript.

page 9: “The research discussed above concerns situations where the bilingual speaker is
operating in any one of their languages.” It seems odd to refer to studies investigating the first
five weeks of learning a language as involving bilinguals.
AR: We have rephrased this line on page 10: "The research discussed above concerns situations where a speaker is operating in any one of their languages"

page 13: It wasn’t clear to me how the discussion of the Bengali laryngeal contrast was relevant
here.
AR: We discuss the four-way laryngeal contrast to motivate our use of vowels as the feature of interest in this study, rather than VOT. The large body of literature on phonetic transfer in VOT is largely based around languages where VOT is the primary cue for voicing contrast (voiced vs voiceless). Thus, exploiting canonical VOT differences between languages allows us to ascertain cross-language influence. In Bengali, the voicing contrast is four-way (voiced vs voiceless, aspirated/breathy vs unaspirated), and VOT is not the primary cue for this contrast. If we were to examine changes to L2 IE VOT as a result of L1 Bengali influence, it would be difficult to interpret the observations in a straightforward manner, owing to the additional phonological differences between the languages.

------------------------------------------------------------------------------------------------------------------
Reviewer: 3

Comments to the Author
The paper reports two tasks – cued-picture naming and code switching – investigating cross-language phonetic transfer in bilingual (Bengali-English) speakers. The study focused on the F1 and F2 of two vowels of the L2 (Indian English) language. Both vowels were subject to changes, with a shift toward the L1 norms. The same changes occurred independently of the task. The study thus reports first evidence of transfer effects on L2 vowels.

I've found the study interesting, and the findings having a good potential for significantly contributing to models of bilingual language control. There are some relatively minor issues (mostly methodological) that I think deserve attention and need to be addressed.

1. As baseline data, the authors use Dutta's (2021) dataset, which is composed of production data from 5 female speakers. In the present study, however, participants are 5 females and 5 males. Because of the effects of gender on vowel realization, do the authors think that this difference between the two groups (Dutta vs the present one) might have contributed to the effects reported in the present study?

AR: Thank you for pointing this out. The statistical analyses in this study is within-participant, comparing participants' IE production in a unilingual condition to their own mixed productions. The baseline Bengali data was used as a heuristic tool to understand how the direction of shift relates to Bengali vowel categories. We had initially used data from Dutta (2021), which was a subset of speech data from the SHRUTI corpus[i], extracted for another study. However, we agree with the reviewer's assessment that the small dataset an lack of gender balance makes it not ideal representative of Bengali vowel production. We have now discarded the Dutta (2021) data, and instead used the entirety of the SHRUTI corpus to get average vowel positions. This corpus contains speech from 34 speakers of Bengali (includes both male and female speakers) and is the largest available corpus of Western Bengali speech that we are aware of. We have elaborated on details of the corpus and speakers on page 16. While the relative vowel positions we observe have not changed as a result of using this larger dataset, we believe that this dataset is more 
representative of average Bengali vowels, and address the reviewer's concern that the observed results might be an artifact of the particular speech data used earlier.

[i] Das, B., Mandal, S., & Mitra, P. (2011, October). Bengali speech corpus for continuous auutomatic speech recognition system. In 2011 International conference on speech database and assessments (Oriental COCOSDA) (pp. 51-55). IEEE.

2. Participants.
a) How was the sample size calculated?
AR: This was based on resource constraints and participant availability based on our inclusion criteria (c.f. response to comment 2 of reviewer 1). We have elaborated on this process and reported our inclusion criteria in page 20 of the revised manuscript.

b) Participants should be described better, especially for what concerns their L2 experience. Currently, there is no information about their proficiency level, as well as their degree of exposure and use of their two languages.
AR: Thanks for this suggestion. We have added the suggested content to the manuscript on page 19; c.f. response to comment 2 of reviewer 1.

3. Materials, how many of the 20 words contained each of the two target vowels? Please specify this information in the text.
AR: We have added this information on page  23:
"Twenty monosyllabic English words: ten containing the vowel ʌ and ten containing æ, were selected as target words."

4. Procedures.

a) If I correctly understand, each word was repeated 4 times in each task (i.e., the same word was produced for a total of 8 times by each participant). If this is the case, can the authors exclude any repetition effect? And why did they opt for a so massive repetition, instead of, e.g., increasing the number of items?

AR: Our choice of items was constrained by phonological considerations, to ensure that coarticulatory effects that are known to affect vowel acoustics, such as identity of the pre- and post-vocalic consonant, were minimal, and controlled across items. Thus, we only used words where the pre-vocalic consonant was a voiced plosive /b/ or /d/, and had a balanced set of post-vocalic consonants across the set of items. Taking into consideration lexical frequency (we only included common words to avoid hesitation while reading) and loanword status (we excluded lexicalized loanwords, e.g. "cup") allowed for 10 target items. We opted for a larger number of repetitions because (i) we wanted multiple responses per participant per item to control for random variability and potential production errors, and (ii) to ensure enough data points for robust statistical analyses. Given that items were presented in random order, and participants were presented with intermediate distracter tasks during the cued picture-naming which relies on language priming, we do not think that repetition effects would come into play in this design.

b) Cued-picture naming: why was the picture presented for a so short time (50 ms)? What words were used as language cue and as target?

AR: In this task the picture, named in either Bengali or English, serves to prime the participant for that particular language. The following target word then either complies with the prime (unilingual condition) or deviates from it (inducing switching; mixed condition). Since participants are already in a bilingual language mode, the effects of such priming, if any, are expected to be highly localized (c.f. discussion in section 1.2 and 1.3, page 9). We presented the picture for a short time to avoid a pause between the prime and the target that would potentially undo the effect of priming, or introduce unwarranted planning effects. Another consideration was for productions in both tasks to mimic the time course and prosody of naturalistic speech as much as possible. Thus, we wanted to avoid long pauses or unnatural intonation that might arise from producing the target as an isolated word.

We have specified the words used as langauge cue, and provided a link to the list of target words, in the text (page 25).

5. Analyses.
a) To analyze the changes at different times during the vowel production, I would suggest to run a single model in which the time of measurement (5%, 15%, 25%, 35%, 45%) is entered as a numeric predictor (from 1 to 5). This will reduce the number of statistical models to run and, more importantly, will allow to directly study the time course of the changes.
AR: This is an excellent suggestion, thank you! We have re-run the analysis as suggested.

b) p. 22 ls 21-23. The authors should specify which interactions they are referring to and report the results of the statistical tests.
AR: While we have now re-written this section to reflect the revised analysis, we have reported the effect sizes, standard errors, and p-values associated with all interaction terms of interest, and reported chi-square, degrees of freedom, and p-values for all model comparisons throughout the manuscript.

c) first analysis (p. 22, ls 24 and following). Given its relevance for the hypotheses, why not including the interaction term for the random slope since the beginning of the analysis?
AR: We have re-written this section. We have now motivated and specified the random effects structure at the beginning of the analysis (page 28), and used that throughout, changing only the fixed effects to test our hypotheses.

d) Tables reporting statistical results, please, specify in the legend what the numbers are (I assume the beta values and the standard errors (in parenthesis), but this should be spelt out).
AR: Values are specified in the revised results tables.

e) I think the readers may appreciate few lines commenting the baseline measure (i.e., Dutta's (2021) data). Moreover, the authors should clarify whether and how these baseline data were compared to the results of their study.
AR: We agree that this information could have been clearer in the original manuscript. We have now added a separate subsection in the introduction (page 16) discussing how we used the baseline Bengali measures, and more detailed speaker information for that data (c.f. response to comment 3 of reviewer 1).

6. I have found the discussion a bit unbalanced, as approximately half of it is dedicated to a secondary question (i.e., the impact of the paradigm on the transfer effects). I suggest the authors to re-balance the space, expanding the discussion of their main research question – they might further develop the theoretical implications of their findings, also offering some explanation of the results they consider unexpected.
AR: Thanks for this feedback; we have re-written the discussion section to include the suggested content (c.f. response to comment 3 of reviewer 1).

Minor

- p. 3 ls 39-43, To improve readiness, I would suggest the authors to consistently use only one term throughout the manuscript.
AH: We have modified the text to use the term "transfer" throughout, and specified this in the Introduction on page 3.

- in some places, the authors use capital letters to highlight some concepts, but I don't think they are really needed. Simple lower-case letters may be fine as well.
AR: We have edited these throughout the manuscript.

- Introduction, when discussing previous studies, the authors might clarify what acoustic measure each study has investigated (VOT, ...)
AR: We have added this information to citations throughout the introduction, and specified this on page 5.

- Acoustic analyses, third line, the authors report a total of 2333 target to be analyzed. How many of them were there for each target vowel?
AR: There were 1137 tokens of [ʌ] and 1196 tokens of [æ]; we have added this information into the text on page 24.

- p. 22, first line, please change “mediated” with “affected” or “moderated” or a similar term – mediated refers to mediation analysis, which is not the analysis run by the authors.
AR: We have made this edit.

### 2022-06-01 BLC_manuscript_1/BLC_AM-ID_manuscript.tex
BLC revision final

--- 2021-06-11 draft_glossa-psych/glossa-psych_AM-ID_manuscript.tex
@@
 % see http://info.semprag.org/basics for a full description of this template
-\documentclass[charis,linguex]{glossa}
-% possible options: 
-% [times] for Times font (default if no option is chosen)
-% [cm] for Computer Modern font 
-% [lucida] for Lucida font (not freely available)
-% [brill] open type font, freely downloadable for non-commercial use from http://www.brill.com/about/brill-fonts; requires xetex
-% [charis] for CharisSIL font, freely downloadable from http://software.sil.org/charis/
-% for the Brill an CharisSIL fonts, you have to use the XeLatex typesetting engine (not pdfLatex)
-% for headings, tables, captions, etc., Fira Sans is used: https://www.fontsquirrel.com/fonts/fira-sans
-% [biblatex] for using biblatex (the default is natbib, do not load the natbib package in this file, it is loaded automatically via the document class glossa.cls)
-% [linguex] loads the linguex example package
-% !!
-a note on the use of linguex: in glossed examples, the third line of the example (the translation) needs to be prefixed with \glt.
-This is to allow a first line with the name of the language and the source of the example.
-See example (2) in the text for an illustration.
-% !! a note on the use of bibtex: for PhD dissertations to typeset correctly in the references list, the Address field needs to contain the city (for US cities in the format "Santa Cruz, CA")
+%\documentclass[times,linguex]{glossa}
+\documentclass[12 pt]{article}
+\usepackage[letterpaper]{geometry}
+\usepackage{times}
+\geometry{top=1.5in, bottom=1.0in, left=1.0in, right=1.0in}
 
-%\addbibresource{sample.bib} 
-% the above line is for use with biblatex
-% replace this by the name of your bib-file (extension .bib is required)
-% comment out if you use natbib/bibtex
+\usepackage[hidelinks]{hyperref}
+%\renewcommand{\sectionautorefname}{\S}
+\usepackage{xcolor}
+\hypersetup{
+	colorlinks,
+	linkcolor={blue!50!black},
+	citecolor={blue!50!black},
+	urlcolor={blue!80!black}
+}
 
-%for natbib: replace \cite with \citep (parenthetical); \citea with \cite (in-text)
-\usepackage{hyperref}
-\let\B\relax %to resolve a conflict in the definition of these commands between xyling and xunicode (the latter called by fontspec, called by charis)
-\let\T\relax
-\usepackage{xyling} %for trees; the use of xyling with the CharisSIL font produces poor results in the branches. This problem does not arise with the packages qtree or forest.
-\usepackage[linguistics]{forest} %for nice trees!
+\usepackage{graphicx} %for images
+\graphicspath{{images/}}
+\usepackage{caption}
+\captionsetup[table]{skip=0pt,singlelinecheck=off}
+\captionsetup[figure]{skip=0pt,singlelinecheck=off}
+
+\usepackage{booktabs}
+\usepackage{enumitem} %customize list numberings
+
+\usepackage[natbibapa]{apacite}
+\usepackage{natbib}
 
@@
 
-%\graphicspath{{images/}}
+\usepackage{fancyhdr}
+\pagestyle{fancy}
+\fancyhf{}
+\renewcommand{\headrulewidth}{0pt}
+\fancyhead[R]{\thepage}
 
-% \pdf* commands provide metadata for the PDF output. ASCII characters only!
-%\pdfauthor{Auromita Mitra, Indranil Dutta}
-%\pdftitle{Mixed language processing increases cross-language phonetic transfer in Bengali-English Bilinguals}
-\pdfkeywords{phonetic transfer, vowel formants, Bengali, English}
+\usepackage{setspace}
+\doublespacing
+
+\usepackage{comment}
+
+\newlength\mystoreparindent
+\newenvironment{myparindent}[1]{%
+	\setlength{\mystoreparindent}{\the\parindent}
+	\setlength{\parindent}{#1}
+}{%
+	\setlength{\parindent}{\mystoreparindent}
+}
 
 \title{Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals}
-% Optional short title inside square brackets, for the running headers.
 
@@
 \begin{document}
+\bibliographystyle{apacite}
+%\sffamily
+%\maketitle
 
-\sffamily
-\maketitle
+%Word count: 9263
 
-Word count: 9263
+\begin{singlespace}
 
-\begin{abstract}
-This study investigates the phonetic outcome of mixed-language production in a group of highly proficient bilingual speakers of Bengali and English, to better understand the nature of cross-language phonetic interaction.
-Existing research has largely focused on a small set of phonologically related languages and used voice onset time (VOT) to measure phonetic transfer.
-Results indicate that such interaction is mediated by both language-specific features and linguistic experience, highlighting the need for diversification in the range of languages and populations studied.
-We compare productions from a yet-unstudied population in two conditions-- unilingual L2 (English) and mixed (Bengali and English).
-The spectral properties of two L2 vowel categories— \nt{\ae} (common to L1 and L2) and \nt{2} (absent in L1) are examined as sites for transfer.
-To understand how this unfolds dynamically, we measure F1 and F2 at five points over the course of articulating the vowel.
-We find dynamic and asymmetrical changes in both vowels during mixed production, with \nt{2} shifting towards a related L1 category \nt{a:}.
-We interpret this as a temporary increase in cross-language phonetic influence during mixed-language use, leading to a shift towards L1 norms.
-Mixed language is elicited in two switching paradigms: cued picture-naming and code-switching, to assess if differences between the paradigms independently influence the outcome of phonetic interaction.
-Results reveal no difference in transfer patterns across the two paradigms.
-We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the postulated role of connected speech in introducing discursive factors to transfer studies.
-\end{abstract}
+\begin{myparindent}{0pt}
+Short title: Phonetic transfer\\
 
-\begin{keywords}
- phonetic transfer; vowel quality; bilingual speech production; L1-L2 interaction; Indian English; Bengali 
-\end{keywords}
+Full title: Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals\\
 
-\rmfamily
+Authors:\\
 
-\section{Phonetic transfer}
+Auromita Mitra
+
+EFL University,
+
+Hyderabad,
+
+India\\
+
+Indranil Dutta
+
+Jadavpur University,
+
+Kolkata,
+
+India\\
+
+Acknowledgments:\\
+We are grateful to Maumita Bhowmik and Anannya Mondal for their help with data collection and processing, and to all the participants of the study.\\
+
+Address for correspondence:\\
+Auromita Mitra,
+
+EFL University,
+
+Hyderabad- 500007,
+
+auromita.mitra@gmail.com
+
+\end{myparindent}
+
+\end{singlespace}
+
+\newpage
+\begin{myparindent}{0pt}
+Abstract:\\
+This study investigates the phonetic outcome of mixed-language speech in Bengali and Indian English, towards understanding cross-language interaction in highly proficient bilinguals.
+We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in unilingual vs.
+mixed productions.
+Results reveal dynamic and asymmetrical changes in both vowels;  \nt{2} shifting towards a related L1 category \nt{a:}.
+We interpret this as a temporary increase in cross-language phonetic interaction during mixed-language use, leading to a shift towards L1 norms, evidencing transfer effects on L2 vowels for the first time.
+We elicit mixed productions through two tasks: cued picture-naming and code-switching, to assess if experimental paradigm independently influences the behavior under study.
+Results reveal no difference across paradigms.
+We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction, postulated discursive factors in code-switching, and the issue of comparability between paradigms in transfer studies.\\
+
+Keywords: phonetic transfer; vowel quality; bilingual speech production; L1-L2 interaction; Indian English; Bengali
+
+\end{myparindent}
+
+\newpage
+
+\section{Introduction: Phonetic transfer}
 
 Bi/multilingual speakers can distinguish between the phonetic norms of their languages and maintain separate sound categories for each language \citep{caramazza1973acquisition,macleod2010impact,bosch2003simultaneous}.
-However, these categories are not autonomous-- they influence each other across languages in both perception and production \citep[e.g.][]{flege1995second,fowler2008cross,flege2002assessing}, and the nature of such interaction provides crucial insights into how language ``systems" are cognitively represented and processed.
-Based on the productions of proficient bilingual speakers of Bengali and English, this study reports that cross-language phonetic influence temporarily increases during mixed-language use.\\%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
+However, these categories are not autonomous-- they influence each other across languages in both perception and production \citep[e.g.][]{flege1995second,fowler2008cross,flege2002assessing}, and the nature of such interaction provides crucial insights into how language `systems' are cognitively represented and processed.
+Based on the productions of proficient bilingual speakers of Bengali and English, this study reports that cross-language phonetic influence temporarily increases during mixed-language use.%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
 %show cross-language-- rephrase
@@
 The results demonstrate a dynamic effect of transfer on L2 vowels during mixed-language production, but no difference in the extent of transfer between the two paradigms.
-We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.\\
+We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.
 
 The rest of this section is structured as follows: \ref{bengali_english_in_india} provides background information on Bengali and Indian English focusing on vowel systems.
-\ref{causes} discusses two potential sources of transfer during language processing-- global co-activation of multiple languages vs the local act of switching between languages.
+\ref{causes} discusses two potential sources of transfer during language processing-- global co-activation of multiple languages vs.
+the local act of switching between languages.
 Given the highly multilingual setting that characterizes this population, we note that the former is expected to be a constant feature and therefore any observed phonetic interaction is better understood as resulting from the latter.
@@
 A logical consequence of this model is that such phonetic interaction must be highly localized.
-\ref{duration} elaborates on this by reviewing literature on the duration and ``reversiblity" of transfer effects.
+\ref{duration} elaborates on this by reviewing literature on the duration and `reversibility' of transfer effects.
 A recurring pattern that emerges from existing research on short-term phonetic transfer is that both languages of the bilingual speaker are not equally affected, although the nature of these reported differences varies greatly across studies.
@@
 
-Existing research has distinguished changes in category \textit{representations} due to the acquisition of multiple sound systems (cf.
+Existing research has distinguished changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf.
 Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during accessing, processing, or articulation of these categories (cf.
-transfer vs interference \citep{grosjean2012attempt}; competence vs performance interference \citep{paradis1993linguistic}).
+transfer vs.
+interference \citep{grosjean2012attempt}; competence vs.
+performance interference \citep{paradis1993linguistic}).
 Given their transient nature, dynamic changes in production during mixed-language use are generally attributed to the latter, e.g.
@@
 
-\subsubsection{L1 vs L2 status} 
+\subsubsection*{L1 vs. L2 status} 
 Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} posits that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another.
@@
 
-\subsubsection{Differences between sound systems of the languages} \label{sound systems} In addition to the cognitive factors discussed above, many researchers have attributed the observed asymmetries to language-internal factors.
+\subsubsection*{Differences between sound systems of the languages} \label{sound systems} In addition to the cognitive factors discussed above, many researchers have attributed the observed asymmetries to language-internal factors.
 Specifically, a pattern that emerges from the body of work on VOT is that a shift in long-lag VOT towards the short-lag norms of the other language is much more consistent and systematic than the converse \citep{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid}.
 In general, short-lag VOTs seem to resist accommodative shift.
-\cite{bullock2009trying} suggest that because long-lag languages offer a greater range of acceptable VOTs, there is more ``room" for movement.
+\cite{bullock2009trying} suggest that because long-lag languages offer a greater range of acceptable VOTs, there is more `room' for movement.
 In contrast, shift in short-lag languages could risk the loss of a phonological contrast (VOT being the primary cue for voicing contrast in these languages).
@@
 The independent variables or fixed factors, along with their levels, are given in table \ref{table variables}.
-Subject and Item were treated as random factors.
+Subject and Item were treated as random factors.\\
 
-\begin{table}[h]
-	\caption{Fixed factors}
-	\sffamily
-	\centering
-	\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
-		\hline 
-		Factor & Levels\\ 
-		\hline
-		Gender & male, female\\
-		Vowel & \nt{2}, \nt{\ae} \\
-		Task (corresponding to the two paradigms) & p (Picture: cued picture-naming, s (Sentence: code-switching)\\
-		Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)\\ [1ex]
-		\hline	
-	\end{tabular}
-	\label{table variables}
-\end{table}
+< Insert Table \ref{table variables} about here > \\
 
@@
 Vowel quality and gender are expected to systematically affect F1.
-Thus, we began with a model that had Vowel and Gender as fixed effects, random intercepts for Subject and Item, and random slopes for each within-unit variable that was relevant to our hypotheses-- by-subject slopes for Task, Context, and Vowel; by-item slopes for Task and Context.
-Next, adding Context as a fixed effect significantly improved the model fit ($\chi^2(1) = 5.54, p= 0.01854$), suggesting that the experimental condition affects vowel height.
+Thus, we began with a model that had Vowel and Gender as fixed effects, random intercepts for Subject and Item, and random slopes for each within-unit variable that was relevant to our hypotheses: by-subject slopes for Task, Context, and Vowel; by-item slopes for Task and Context.
+Next, adding Context as a fixed effect significantly improved the model fit ($\chi^2(1) = 5.54, p= .01854$), suggesting that the experimental condition affects vowel height.
 Following from our hypothesis \textbf{Asymmetry}, we expected the extent of change in vowel height between the conditions to be mediated by the identity of the vowel.
 Thus, we added a two-way interaction factor between Context and Vowel (Context*Vowel).
-As expected, this was found to significantly improve model fit ($\chi^2(1) = 12.61, p= 0.0003837$).
+As expected, this was found to significantly improve model fit ($\chi^2(1) = 12.61, p= .0003837$).
 To assess whether the experimental paradigm affected F1, we added Task as a fixed effect.
-This marginally improved the model ($\chi^2(1) = 3.74, p= 0.05302$).
+This marginally improved the model ($\chi^2(1) = 3.74, p= .05302$).
 Towards addressing our research question \textbf{Paradigm}, we wanted to assess whether the extent of shift differs across the two paradigms.
@@
 Since our hypothesis concerns their interaction, we modified the random effects structure by adding a by-subject random slope for the interaction term \citep{brauer2018linear}.
-This improved the model fit ($\chi^2(5) = 11.43, p = 0.04342$).
-Thus, we report on a model that includes Gender, Task, Context, Vowel, Context*Vowel as fixed effects, random intercepts for Subject and Item, by-subject random slopes for Task, Context, Vowel and Context*Vowel, and by-item random slopes for Task and Context.\\
+This improved the model fit ($\chi^2(5) = 11.43, p = .04342$).
+Thus, we report on a model that includes Gender, Task, Context, Vowel, Context*Vowel as fixed effects, random intercepts for Subject and Item, by-subject random slopes for Task, Context, Vowel and Context*Vowel, and by-item random slopes for Task and Context.
 
-The following sections discuss the results for vowel height and backness respectively:
+Figure \ref{vowels_e_b} summarizes the findings of the experiment ($F1XF2$ plots for vowel productions in the unilingual and mixed conditions, in comparison to the related Bengali categories from \cite{dutta2021}).
+The following sections discuss the results for vowel height and backness respectively.
+
+< Insert Figure \ref{vowels_e_b} about here >
 
@@
 We find that the difference is not significant at 5\%, and increases over the course of articulating the vowel.
+\\
 %ANOVA table add
 
-\begin{figure}[h] \centering
+< Insert Table \ref{table_f1_fixed_effects} about here >\\
 
-	\includegraphics[scale=0.8]{vowel_by_subject_ggplot}
-	\caption{Subject-wise mean F1; unilingual English}
-	\label{boxplot_F1}
-\end{figure}
+< Insert Figure \ref{boxplot_F1} about here >\\
 
-\begin{table}[htbp] \centering 
+\subsection{Vowel backness -- F2}
+
+Table \ref{table_f2_fixed_effects} summarizes the results of fixed effects from the model for F2 at each point.
+Once again, we see that male speakers have a lower mean F2 than female speakers, as expected.
+The estimates for Vowel show that excepting the beginning of the vowel, the category \nt{2} is consistently produced as more backed than \nt{\ae}.
+Unlike vowel height, Context has an independent effect on vowel backness, such that both the vowels in this study have higher F2 values in the Bengali (mixed) condition, i.e.
+they are fronted in the mixed condition.
+A significant interaction between Vowel and Context reveals an asymmetry in the extent of fronting between the two vowels -- the shift is greater in the case of \nt{\ae}.
+We see a significant effect of Task on F2, such that vowels in the Sentence task (code-switching) were more backed than those in the picture-naming task.
+As with F1, this is an unexpected result, as we have no principled reason to expect code-switching to induce an across-board backing of vowels.
+A lack of significant interaction with Context suggests that there is no difference in the extent of shift between the two paradigms.
+The results of ANOVAs comparing the full model to a null model that lacks the Vowel*Context interaction factor reveal that as with vowel height, the change in vowel backness between the two conditions is not significant at 5\%, and increases over the course of the vowel.\\
+
+< Insert Table \ref{table_f2_fixed_effects} about here >
+
+\section{Discussion}
+
+With regard to our first research question \textbf{L2 vowels}, the results confirm that the L2 vowels of proficient bilinguals can be affected by mixed-language processing.
+Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.
+In the context of an ICM--based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, this is a significant finding: it demonstrates that the postulated lower switch cost on an L2 does not entirely mask transfer effects on L2 vowels.
+Moreover, our findings differ from \cite{tsui2019impact} who report on the basis of VOT measurements that balanced bilinguals do not evidence an effect of language-switching.
+\citeauthor{tsui2019impact} attribute the absence of transfer effects to greater experience with language switching in these participants, leading to lower switch-costs in both languages.
+We find that even proficient bilinguals with extensive experience of language switching demonstrate cross-language transfer effects in vowel production.
+Thus, following from the discussion in sec.
+\ref{asymmetries}, it might be fruitful for future studies to focus on vowels as a potential site for transfer.
+This would both expand the empirical scope of existing research (cf.
+sec.
+\ref{asymmetry between sounds}) and avoid confounding factors arising from independent differences between long- and short-lag VOT languages.
+More generally, these findings point to a need for expanding the range of features in which transfer effects are studied, in order to make robust generalizations about the mechanisms underlying observed patterns.
+
+We hypothesized an asymmetrical shift between the two vowels (\textbf{Asymmetry}), owing to differences in (i) their positions in the Indian English vowel space, and (ii) the presence/absence of a corresponding category in the L1 Bengali inventory.
+The significant interaction between Vowel and Context in both F1 and F2 confirm that the experimental condition affected both height and backness differently for each vowel.
+As expected, we find a greater degree of shift in \nt{2}, possibly owing to its position in a less dense part of the IE vowel space.
+The category \nt{2} is lowered and fronted in the mixed language condition as expected (Figure \ref{vowels_e_b}).
+The direction of shift in \nt{\ae} is surprising-- it is fronted in the mixed-language condition.
+While phonological constraints on IE suggested that a shift in the F1 dimension could potentially risk the loss of contrast and is thus expected to be minimal, Figure \ref{vowels_e_b} shows that \nt{\ae} in Bengali is not fronted with respect to the English category.
+Thus, fronting in the mixed condition cannot be explained in relation the former, at least from the present data.
+Recall that the Bengali vowel productions used here are from a different set of speakers than our participants.
+Assuming some amount of individual differences in the precise phonetic realization of vowel categories in both languages, this could be a potential source of explanation for this result.
+Data from a larger number of speakers for both languages, as well as within-speaker comparisons of English and Bengali vowels, are required to clarify this.
+
+Our third research question \textbf{Paradigm} concerned the effect of experimental paradigm on the phonetic outcome of mixed-language use.
+\cite{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
+If such discursive properties independently affect the phonetic outcome of language mixing in an experimental setup, this calls to question the comparability of results across studies using these different paradigms.
+We verified this by comparing productions of the same items from the same group of participants in a scripted CS paradigm and a cued picture-naming paradigm.
+We found no significant interaction of Task either with Context, or with the Vowel*Context interaction, suggesting that the extent of phonetic transfer was not affected by the paradigm.
+There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in observable transfer at all.
+This would mean that the pre-switch effects observed, for example, in \cite{muldner2019phonetics} and \cite{bullock2009trying} are better interpreted as resulting from some other aspect of the experimental setup, than the act of code-switching itself.
+While planning could additionally affect the words preceding the switch, the effects on the switched token itself do not change.
+(ii) Our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
+Thus, another possible explanation is that the discursive effects discussed by \cite{olson2013bilingual} are triggered by more naturalistic paradigms where attention to meaning and interlocutor are important.
+More within-study comparisons with less structured paradigms, and spontaneous code-switching data, can clarify this.
+The results of the present study indicate that for the purposes of observing dynamic phonetic transfer in an experimental setup, elicitation through cued-picture-naming vs.
+(controlled) code-switching does not significantly alter the behavior under study.
+Thus, the choice of paradigm can be based on other considerations.
+Moreover, this suggests that (all else being equal) the results from such studies can be compared directly.
+
+The paradigm did affect vowel quality, but this applied uniformly across experimental conditions: all vowels were more backed and lowered in the code-switching task.
+Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this lowering is an artifact of the particular stimuli used here---resulting from some corarticulatory effect that was absent in the picture-naming task.
+While this is intriguing, it is beyond the scope of the present study.
+\\
+
+\section{Conclusion}
+This study examined the phonetic outcome of mixed-language production in a group of highly proficient bilingual speakers of Bengali and Indian English.
+Specifically, we examined the spectral quality of two L2 vowel categories \nt{2} and \nt{\ae} produced in two experimental conditions: unilingual (English) and mixed (switching from Bengali to English).
+Comparisons of F1 and F2 reveal that mixed language processing temporarily increases phonetic interaction between the languages, causing \nt{2} to converge towards a related L1 category \nt{a:}.
+The extent and direction of shift is asymmetrical across the two vowel categories, which parallels findings in prior studies.
+This asymmetry appears to be mediated by the language-specific distribution of vowel categories.
+These findings evidence an effect of phonetic transfer on L2 vowels for the first time, and raise questions for proposed inhibition-based accounts of cross-language phonetic interaction.
+Mixed productions were elicited in two switching paradigms: code-switching and cued picture-naming, to verify whether the postulated discursive elements in code-switching independently influence the outcome of phonetic transfer.
+Results suggest that in an experimental setup, elicitation through cued-picture-naming vs.
+(controlled) code-switching does not significantly alter the behavior under study.
+
+This study has a number of limitations: (i) the `baseline' values for Bengali vowel categories were estimated from a different set of speakers than the participants of this study, potentially obscuring the role of individual differences in the realization of vowel categories; within-speaker comparisons of L1 and L2 vowels is likely to offer a more fine-grained view of the direction of shift in relation to L1 categories.
+(ii) we used a frame sentence to elicit code-switched productions such that not all the sentences were pragmatically well-formed.
+This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study.
+Future work should address this by considering data from more naturalistic code-switching paradigms.
+
+\section*{Abbreviations}
+\textsc{clf} = classifier
+
+\section*{Data availability}
+The materials, data files, and analysis code for this study are openly available in OSF at:  \url{https://doi.org/10.17605/OSF.IO/DSB2X}.
+
+\section*{Competing interests}
+The authors declare none.
+
+%\printbibliography %for use with biblatex; comment out if you use natbib
+\bibliography{references} %for use with natbib; comment out if you use biblatex, and change 'sample' by the name of your bib-file
+
+\newpage
+\section*{Tables}
+\begin{table}[h]
+	\caption{Fixed factors}
+	%\centering
+	\resizebox{\textwidth}{!}{%
+		\begin{tabular}{@{}ll@{}}
+			\toprule
+			Factor                                                 & Levels                                                            \\ \midrule
+			Gender                                                 & male, female                                                      \\
+			Vowel                                                  & \nt{2}, \nt{\ae} \\
+			Task (corresponding to the two paradigms)              & p (Picture: cued picture-naming, s (Sentence: code-switching)     \\
+			Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)                       \\ \bottomrule
+		\end{tabular}%
+	}
+	\label{table variables}
+\end{table}
+
+\newpage
+\begin{table}[htbp] %\centering 
 	\caption{Model outputs for F1 at each of the five points in the vowel} 
@@
 
-\section{Discussion}
+\newpage
 
-With regard to our first research question \textbf{L2 vowels}, the results confirm that the L2 vowels of proficient bilinguals can be affected by mixed-language processing.
-Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.
-In the context of an ICM--based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, this is a significant finding: it demonstrates that the postulated lower switch cost on an L2 does not entirely mask transfer effects on L2 vowels.
-Moreover, our findings differ from \cite{tsui2019impact} who report on the basis of VOT measurements that balanced bilinguals do not evidence an effect of language-switching.
-The authors attribute this to greater experience with language switching, leading to lower switch-costs in both languages.
-We find that even proficient bilinguals with extensive experience of language-switching demonstrate cross-language transfer effects in vowel production.
-Thus, following from the discussion in sec.
-\ref{asymmetries}, it might be fruitful for future studies to focus on vowels as a potential site for transfer.
-This would both expand the scope of the existing research (cf.
-sec.
-\ref{asymmetry between sounds}) and avoid confounding factors arising from independent differences between long- and short-lag VOT languages.\\
+\section*{Figures}
 
-We hypothesized an asymmetrical shift between the two vowels (\textbf{Asymmetry}), owing to differences in (i) their positions in the Indian English vowel space, and (ii) the presence/absence of a corresponding category in the L1 Bengali inventory.
-The significant interaction between Vowel and Context in both F1 and F2 confirm that the experimental condition affected both height and backness differently for each vowel.
-As expected, we find a greater degree of shift in \nt{2}, possibly owing to its position in a less dense part of the IE vowel space.
-Figure \ref{vowels_e_b} shows $F1xF2$ plots for vowel productions in the unilingual and mixed (dotted ellipses) conditions, in comparison to the related Bengali categories (colored ellipses) from \cite{dutta2021}.
-The category \nt{2} is lowered and fronted in the mixed language condition as expected.
-The direction of shift in \nt{\ae} is surprising-- it is fronted in the mixed-language condition.
-While phonological constraints on IE suggested that a shift in the F1 dimension could potentially risk the loss of contrast and is thus expected to be minimal, Figure \ref{vowels_e_b} shows that \nt{\ae} in Bengali is not fronted with respect to the English category.
-Thus, fronting in the mixed condition cannot be explained in relation the former, at least from the present data.
-Recall that the Bengali vowel productions used here are from a different set of speakers than our participants.
-Assuming some amount of individual differences in the precise phonetic realization of vowel categories in both languages, this could be a potential source of explanation for this result.
-Data from a larger number of speakers for both languages, as well as within-speaker comparisons of English and Bengali vowels, are required to clarify this.
-\\
-
-\begin{figure}[h] \centering
-	\includegraphics[scale=0.8]{vowels_e_b_final}
-	\caption{Vowel categories in unilinual English (solid lines), mixed condition (dotted lines), and baseline Bengali (colored ellipses)}
+\begin{figure}[h] %\centering
+	\includegraphics[scale=1]{vowels_e_b_final}
+	\caption{Vowel quality in unilinual English, mixed, and baseline Bengali productions}
 	\label{vowels_e_b}
@@
 
-Our third research question \textbf{Paradigm} concerned the effect of experimental paradigm on the phonetic outcome of mixed-language use.
-\cite{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
-If such discursive properties independently affect the phonetic outcome of language mixing in an experimental setup, this calls to question the comparability of results across studies using these different paradigms.
-We verified this by comparing productions of the same items from the same group of participants in a scripted CS paradigm and a cued picture-naming paradigm.
-We found no significant interaction of Task either with Context, or with the Vowel*Context interaction, suggesting that the extent of phonetic transfer was not affected by the paradigm.
-There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in observable transfer at all.
-This would mean that the pre-switch effects observed, for example, in \cite{muldner2019phonetics} and \cite{bullock2009trying} are better interpreted as resulting from some other aspect of the experimental setup, than the act of code-switching itself.
-While planning could additionally affect the words preceding the switch, the effects on the switched token itself do not change.
-(ii) Our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
-Thus, another possible explanation is that the discursive effects discussed by \cite{olson2013bilingual} are triggered by more naturalistic paradigms where attention to meaning and interlocutor are important.
-More within-study comparisons with less structured paradigms, and spontaneous code-switching data, can clarify this.
-The results of the present study indicate that for the purposes of observing dynamic phonetic transfer in an experimental setup, elicitation through cued-picture-naming vs (controlled) code-switching does not significantly alter the behavior under study.
-Thus, the choice of paradigm can be based on other considerations.
-Moreover, this suggests that (all else being equal) the results from such studies can be compared directly.
+\newpage
 
-The paradigm did affect vowel quality, but this applied uniformly across experimental conditions: all vowels were more backed and lowered in the code-switching task.
-Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this lowering is an artifact of the particular stimuli used here---resulting from some corarticulatory effect that was absent in the picture-naming task.
-While this is intriguing, it is beyond the scope of the present study.
-\\
+\begin{figure}[h] %\centering
 
-\section{Conclusion}
-This study examined the phonetic outcome of mixed-language production in a group of highly proficient bilingual speakers of Bengali and Indian English.
-Specifically, we examined the spectral quality of two L2 vowel categories \nt{2} and \nt{\ae} produced in two experimental conditions: unilingual (English) and mixed (switching from Bengali to English).
-Comparisons of F1 and F2 reveal that mixed language processing temporarily increases phonetic interaction between the languages, causing \nt{2} to converge towards a related L1 category \nt{a:}.
-The extent and direction of such shift is asymmetrical across the two categories, which parallels findings in prior studies.
-This asymmetry appears to be mediated by the language-specific distribution of vowel categories.
-Mixed productions were elicited in two switching paradigms: code-switching and cued picture-naming, to assess whether the postulated discursive elements in code-switching independently influences the outcome of phonetic transfer.
-Results suggest that in an experimental setup, elicitation through cued-picture-naming vs (controlled) code-switching does not significantly alter the behavior under study.
-\\
-
-This study has a number of limitations: (i) the `baseline' values for Bengali vowel categories were estimated from a different set of speakers than the participants of this study, potentially obscuring the role of individual differences in the realization of vowel categories; (ii) we used a frame sentence to elicit code-switched productions such that not all the sentences were pragmatically well-formed.
-This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study.
-Future work should address this by considering data from more naturalistic code-switching paradigms.
-
-\section*{Abbreviations}
-\textsc{clf} = classifier
-
-\section*{Data availability}
-The data files and analysis code are available at:  \url{https://osf.io/dsb2x/?view_only=5cb821070bc74a038a293239157c3b11}
-
-\section*{Competing interests}
-The authors have no competing interests to declare.
-
-%\printbibliography %for use with biblatex; comment out if you use natbib
-\bibliography{references} %for use with natbib; comment out if you use biblatex, and change 'sample' by the name of your bib-file
+	\includegraphics[scale=1]{vowel_by_subject_ggplot}
+	\caption{Subject-wise mean F1 in unilingual English productions}
+	\label{boxplot_F1}
+\end{figure}
 

### 2022-06-01 BLC_manuscript_2/BLC_AM-ID_manuscript.tex
BLC revision final

--- 2022-05-19 draft_BLC/BLC_AM-ID_manuscript.tex
@@
 mixed productions.
-Results reveal dynamic and asymmetrical changes in both vowels;  \nt{2} shifting towards a related L1 category \nt{a:}.
+Results reveal asymmetrical shifts in both vowels, towards related L1 categories.
 We interpret this as a temporary increase in cross-language phonetic interaction during mixed-language use, leading to a shift towards L1 norms, evidencing transfer effects on L2 vowels for the first time.
 We elicit mixed productions through two tasks: cued picture-naming and code-switching, to assess if experimental paradigm independently influences the behavior under study.
-Results reveal parallel patterns, but a slight difference in magnitude of transfer across paradigms.
-We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction, postulated discursive factors in code-switching, and the issue of comparability between paradigms in transfer studies.\\
+Results reveal parallel patterns, but small magnitude differences, across paradigms.
+We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction, postulated discursive factors in code-switching, and the issue of comparability between paradigms in transfer studies.
+\\
 
@@
 The present study is concerned with the latter.
-In production, this kind of influence has been variously termed \textsc{transfer, drift, accommodation,} and \textsc{interference}-- we use the term \textsc{transfer} here, to indicate any interaction between two sets of phonetic norms.
+In production, this kind of influence has been variously termed \textsc{transfer, drift, accommodation,} and \textsc{interference}-- we use the term \textsc{``transfer"} here, to indicate any interaction between two sets of phonetic norms.
 
@@
 
+<Insert figure \ref{vowels_bengali} about here>
+
+<Insert figure \ref{vowels_ie} about here>
+
 \subsection{What causes transfer and what does it affect?}\label{causes}
@@
 Indirect evidence in support of this comes from experiments which have elicited switched and non-switched tokens in the same test block and still reported differences between the two \citep[e.g.][VOT]{tsui2019impact,olson2013bilingual}, suggesting that changes due to transfer are quickly `reset'--- in an order of seconds.\\
-Note that all the studies discussed above measure VOT, which is a temporal feature.
+Note that all the studies of mixed-language production discussed above measure VOT, which is a temporal feature.
 We have no apriori reason to assume that these durations generalize to vowel quality.
@@
 We expect the intervening words between two subsequent targets to undo any residual effects of transfer.
-
-\alert{discussion: found that like VOT, vowel quality shifts due to switching are also highly localized, and reset in an order of seconds.
-This behavior is similar to other context-driven shifts in the literature}
 
@@
 
-\subsubsection{Baseline Bengali data from the SHRUTI corpus}
+\subsubsection*{Baseline Bengali data from the SHRUTI corpus}
 
@@
 discussion in section \ref{bengali_english_in_india}).
-To the best of our knowledge, there is no freely available corpus of IE speech from Bengali speakers.
+To the best of our knowledge, there is no freely available corpus of IE speech from Bengali speakers in India.
+
 The F1XF2 plot indicates that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.
@@
 \subsection{Participants} \label{participants}
-Participants were recruited through an open call at the EFL University campus in Hyderabad, India, circulated through social media.
+Participants were recruited through an open call  circulated through social media at the EFL University campus in Hyderabad, India.
 Interested volunteers filled a brief language background questionnaire.
 These responses were used to screen participants, in the manner described below.
-As noted in \ref{bengali_english_in_india}, India has a super multilingual an linguistically diverse population, and EFLU being a central university represents an extremely diverse population, as students are from different parts of the country, having both different L1s, and different varieties of English.
-Our aim in screening participants was to minimize between-participant random variability, while maintaining ecological validity considering the actual day-to-day settings in which such behaviors likely take place.
-We chose two metrics to do this: (i) where the participant grew up; (ii) number of years of formal education in Bengali.
-The first ensures that the variety of Bengali that participants are exposed to in their surroundings is similar.
+As noted in section \ref{bengali_english_in_india}, India has an extremely multilingual an linguistically diverse population.
+EFLU being a central university houses students from different parts of the country, who speak both different L1s, and different varieties of English.
+Our aim in screening participants was to minimize between-participant random variability in the study, while maintaining ecological validity considering the actual day-to-day settings in which transfer behaviors likely take place.
+We chose two metrics to do this: (i) where the participant grew up; and (ii) number of years of formal education in Bengali.
+The first ensures that the variety of Bengali that individuals are exposed to in their surroundings is similar across participants.
 This would mean that the target representations for Bengali vowels would be expected to be similar.
-This was especially important because we were not eliciting unilingual L1 data in this study, and were using a corpus to estimate the position of Benali vowels in the formant space.
+This was especially important because we were not eliciting unilingual L1 Bengali data in this study, and were using an existing data source to estimate the position of Bengali vowels in the formant space.
 Based on previous literature (c.f.
 section \ref{asymmetry between sounds}), we expect that L2 vowels shift towards related L1 categories during mixed language utterances.
-Thus, we wanted to make sure that differences between participants in the pattern of shift were actually due to individual differences, rather than different acoustic targets.
-Another reason to control for place of growing up was that it is a fairly reliable predictor of what languages the person was educated in, owing to how the education system works (elaborated below).
+Thus, we wanted to ensure that any between-participant asymmetries in the pattern of shift reflected individual differences in transfer, rather than different acoustic targets.
+Another reason to control for location was that it is a fairly reliable predictor of what languages the person was educated in, owing to how the education system functions in India (we elaborate on this below).
 
 The second metric is Medium of Instruction (MoI) in school.
-Most schools in India follow the Three Language Policy, wherein formal education is provided in English, Hindi, and the regional language (the idea and implementation of this policy, and its implications for regional minorities has been a matter of much public debate as well as scholarly work; for discussions see \alert{citation}).
+Most schools in India follow the Three Language Policy, wherein formal education is provided in English, Hindi, and the regional language.
+(The idea and implementation of this policy, and its implications for linguistic minorities, has been a matter of much public debate as well as scholarly work; for discussions see, e.g.
+\cite{tollefson2014language, jhingran2009hundreds, khubchandani1997language, mohanty2009multilingual, ramanathan2005rethinking}).
 Sustained formal education in a language is a reliable predictor of LSRW (listening, speaking, reading, writing) skills.
@@
 
-With these considerations in mind, we only recruited those participants who had grown up in the state of West Bengal and lived there for the majority of their lives, had parents who spoke Bengali, and had received at least 5 years of formal education in Bengali (all respondents had received 5+ years of education in English and were studying university degrees that are taught in English), and self-reported proficiency in speaking, understanding, and reading Bengali.
+With these considerations in mind, we only recruited participants who had grown up in the state of West Bengal and lived there for the majority of their lives, had parents who spoke Bengali, and had received at least 5 years of formal education in Bengali (all respondents had received 5+ years of education in English and were pursuing university degrees that are taught in English), and self-reported proficiency in speaking, understanding, and reading Bengali.
 
@@
 Using the inclusion criteria reported above, 10 volunteers (5 female, 5 male, age range 19 to 28) were invited to participate in a speech production study, and were compensated for their time.
-All participants reported normal hearing, and normal or corrected to normal vision.
-After the completion of the study, we followed up with a more detailed language background survey at a later date to learn more about the language profiles of the participants.
+All participants reported normal hearing, and normal or corrected-to-normal vision.
+After completion of the study, we followed up with a detailed language background survey at a later date to learn more about the language profiles of the participants.
 This survey was a modified and consolidated version of three popular bilingualism profiling tools: the Bilingual Language Profile (BLP) \citep{blp}, the Language Experience and Proficiency Questionnaire (LEAP-Q) \citep{leap-q} which incorporates language history and self-assessed proficiency, and the Bilingual Switching Questionnaire (BSWQ) \citep{language_switching_questionnaire}.
 Together, these incorporate questions about language history, usage, attitudes, and switching experience.
+The complete questionnaire, along with responses, are available in the ``Data" component of the OSF repository for this project (c.f.
+section \ref{data_availability}).
 
@@
 Thus, all participants can be described as simultaneous bilinguals, having acquired both English and Bengali at a young age.
-They report being comfortable using both languages at a young age, with the average being 7.5 for English and 0.3 for Bengali (on a likert scale rating where 0=for as long as I can remember).
-As mentioned earlier, all participants had been in schools where Bengali and English were both used as Medium of Instruction (MoI).
+They report being comfortable using both languages at a young age, with the average being 7.5 for English and 0.3 for Bengali (on a likert scale rating where 0=for as long as I can remember, 5=since I was 5 years old, and so on).
+As mentioned earlier, all participants had attended schools where Bengali and English were both used as Medium of Instruction (MoI).
 On average, participants had had more than 16 years (range 7 to 20+) of formal education in English and 11.8 years (range 8 to 15) in Bengali at the time of the study.
@@
 While self-ratings for reading and writing English were at ceiling, ratings for Bengali literacy (reading and writing) showed some variation, with an average of 4.87 (on a 0-6 likert scale, range 3 to 6) for reading and 4.25 (range 3 to 6) for writing.
-This is a common situation for many educated sections of society, since institutions of higher education beyond the school level largely use English (for discussions about the language politics of higher education in India, see \alert{ \cite{ELE readings find}}).
-The language attitude section of the questionnaire reveals interesting information about the linguistic landscape of the population, and how people experience language in such a setting.
+This is a common situation for many educated sections of society, since institutions of higher education beyond the school level largely use English (for discussions about the language politics of higher education in India, see, for example, \cite{mohanty2009multilingual}).
+The language attitude section of the questionnaire reveals an interesting insights about the linguistic landscape of the population, and how people experience language in such a setting.
 When asked how much ``like themselves" participants feel while using each language, response were at ceiling for both languages.
 Similarly, most participants identified equally strongly with both an ``English-speaking" and a ``Bengali-speaking" culture, possibly because for most people, such linguistic-cultural identities are not clearly delineated and not exclusive of one another.
-These responses shed light on how multiple languages fit seamlessly into different facets of daily life to create a ``continuous" linguistic experience \alert{(c.f.
-plurality-- from Personal statement)}.
+These responses shed light on how multiple languages fit seamlessly into different facets of daily life to create a ``continuous" linguistic experience (c.f.
+"plurality" as defined by \cite{khubchandani2021plural}).
 In such a scenario, it is not surprising that language-switching is common, expected, as well as used is creative ways with communicative purpose.
@@
 In spite of this, participants report a proclivity to code-switch (rating 3.5).
-Importantly, a large number of participants report volition while switching languages (4.12), and many report it to be situationally motivated (3.12).
-These responses indicate that participants are highly aware of switching habits and move between languages for communicative purposes.
+Importantly, a large number of participants report volition while switching languages (4.12), and many report this to be situationally motivated (3.12).
+These responses indicate that participants are highly aware of their switching habits and move between languages for communicative purposes.
 This fact, combined with the linguistic situation described here and in section \ref{bengali_english_in_india}, are important to bear in mind while interpreting the results of a transfer study in the present population.
@@
 \subsection{Stimuli}
-Twenty monosyllabic English words containing the vowels \nt{2} or \nt{\ae} were selected as target words.
+Twenty monosyllabic English words: ten containing the vowel \nt{2} and ten containing \nt{\ae}, were selected as target words.
 To minimize coarticulatory effects on the vowel, the pre-vocalic consonants were all voiced plosives (\nt{b} or \nt{d}).
@@
 We avoided English words that are already lexicalized loanwords in Bengali.
-%The list of words can be found in sec.\ref{appendix}.
+The complete list of target words can be found in the ``Materials" component of the OSF repository for this project (c.f.
+section \ref{data_availability}).
 
@@
 		\item Language cue- a word in either English or Bengali orthography
-		\item A picture displayed for 50ms; named by the participant in either English or Bengali, depending on the language cue 
+		\item A picture displayed for 50 ms; named by the participant in either English or Bengali, depending on the language cue.
+The picture of a fish was used in all trials, prompting the participant to say either \nt{fIS} or \nt{ma: \textteshlig \super{h}}, depending on the language cue
 		\item Target -- an English word; read out by the participant
@@
 	\end{itemize}
-	Since all the target words were in English, the language cue was used to manipulate the \textsc{preceding} word, giving target words in non-switched (unilingual) and switched (mixed) utterances.
+	Since all the target words were in English, the language cue was used to manipulate the preceding word, giving target words in non-switched (unilingual) and switched (mixed) utterances.
 The orthographies of English and Bengali are visually distinct, and were therefore used to cue language.
@@
 Individual vocal tract differences affect formant frequencies across speakers.
-To avoid this confound, we used Lobanov normalization (using the PhonR package \citep{phonR} in R \citep{r}, representing each speaker's formant frequencies as the scaled distance from their own mean (\alert{citation for lobanov normalization?}).
-Plotting the vowels \alert{(figure)} shows a difference between male and female speakers, such that male speakers have overall lower formant frequencies.
+To avoid this confound, we used Lobanov normalization (using the PhonR package \citep{phonR} in R \citep{r}, representing each speaker's formant frequencies as the scaled distance from their own mean).
+Plotting these vowels shows a difference between male and female speakers, such that male speakers have overall lower formant frequencies (figure \ref{gender_before}).
 Speaker normalization using the Lobanov method does not get rid of this difference.
-Since gender-related differences in formant frequencies are not of interest in this study, we normalized individual values by gender (representing each value as a scaled distance from the group mean).
+Since gender-related differences in formant frequencies are not of interest in this study, we normalized individual values by gender (representing each value as a scaled distance from the group mean; figure \ref{gender_after}).
 Statistical analyses were performed on these normalized values.
+
+< Insert figure \ref{gender_before},\ref{gender_after} about here >
 
@@
 
-First, to test RQ 1 (\bf{L2 vowels}), we fitted a model with vowel*time, and vowel*context as fixed effects.
+First, to test our first research question (\bf{L2 vowels}), we fitted a model with vowel*time, and vowel*context as fixed effects.
 The transfer term (vowel*context) is significant in this model (model coefficient/ unstandardized effect size ($\beta$)= -2.330e-01, standard error (SE)=  5.244e-02, t= -4.44, p=0.0002).
@@
 Table \ref{table_f1} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model.
-\alert{p-vales smaller than 0.05 are in bold.}
-
-\subsubsection*{Interpretation}
+p-vales smaller than 0.05 are starred.
+
+\subsubsection*{Model interpretation}
 We extracted formant estimates from five uniformly spaced time points in the vowel: from the time at 5\% into the vowel (at the start of articulation) to 45\% (at the mid-part, or steady state, which is expected to correspond most closely to the articulatory target for the category).
@@
 Table \ref{table_f2} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model.
-\alert{p-vales smaller than 0.05 are in bold.}
-
-\subsection*{Interpretation}
+p-values smaller than 0.05 are in starred.
+
+\subsubsection*{Model interpretation}
 
@@
 vowels are fronted with time, the negative slope of the Time*Vowel interaction shows that this effect is smaller for \nt{2}.
-This is expected, as figure \alert{ref} confirms that the category \nt{\ae} is overall more fronted than \nt{2}.
+This is expected, as figure \ref{vowels_unilingual_ie_bengali} confirms that the category \nt{\ae} is overall more fronted than \nt{2}.
 
@@
 \subsection{Power analysis}
-We were able to recruit 10 participants (c.f.
+Owing to participant availability given our inclusion criteria, we were able to recruit 10 participants for this study (c.f.
 section \ref{participants}).
+We acknowledge that this sample size is on the lower side for studies with comparable designs.
 To confirm whether the experiment design was powerful enough to detect the effects we are investigating, we carried out post-hoc power analyses using the \textit{mixedpower} package \citep{kumle2021estimating} in R.
+The package simulates responses based on a given dataset and runs the specified analyses on these new data, estimating power by calculating how frequently significant effect are obtained in the simulated data.
 The target of the analysis was to calculate the power for detecting an effect for our main variables of interest: vowel*context (to test the hypotheses L2-vowels and Asymmetry), and vowel*context*task (to test the hypothesis Paradigm).
@@
 
-Cohen \citep{cohen1988statistical} classifies effect sizes (Cohen's d values) as small ($\leq$0.2), medium (0.5), and large ($\geq$ 0.8).
+Cohen \citeyearpar{cohen1988statistical} classifies effect sizes (Cohen's d values) as small ($\leq$0.2), medium (0.5), and large ($\geq$ 0.8).
 Although the usefulness of this classification has been debated in the literature, these are often used as ballpark values for power analyses in psychological studies (for a discussion, see \cite{brysbaert2018power}).
@@
 Instead, we run simulations with a range of effect sizes and report the power of our study to plausibly detect an effect of that size.
-The results are shown in the plots in figure \ref{} Each panel presents power analysis at a different effect size, and is based on 500 simulations.
+The results are shown in the plots in figures \ref{figure_power_analysis_f1} and \ref{figure_power_analysis_f2}.
+Each panel presents power analysis at a different effect size, and is based on 500 simulations.
 The plots show power (y-axis) over a range of sample sizes (x-axis), given the assumed effect size.
@@
 
-%This study has a number of limitations: (i) the `baseline' values for Bengali vowel categories were estimated from a different set of speakers than the participants of this study, potentially obscuring the role of individual differences in the realization of vowel categories; within-speaker comparisons of L1 and L2 vowels is likely to offer a more fine-grained view of the direction of shift in relation to L1 categories.
-(ii) we used a frame sentence to elicit code-switched productions such that not all the sentences were pragmatically well-formed.
-This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study.
-Future work should address this by considering data from more naturalistic code-switching paradigms.
+Although the findings of this study are interesting, we acknowledge that it has a number of limitations, including (i) a small sample size; (ii) a frame sentence to elicit code-switched productions such that not all the sentences were pragmatically well-formed.
+This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study; (iii) focus on a single acoustic feature; (iv) the absence of unilingual Bengali data from the same set of speakers as the participants of the study, which would have allowed us to compare each individual's transfer patterns to their own Bengali productions, rather than to average Bengali vowel formants.
+Future work should address these.
+Our results also raise a number of open questions for future research, particularly concerning the paradigm differences and between-sound asymmetries in phonetic transfer.
 
 \section{Conclusion}
-This study investigates how the phonetic systems of highly proficient bilingual speakers interact during online speech processing, by examining the phonetic outcome of mixed-language production in speakers of Bengali and Indian English.
+This study investigated how the phonetic systems of highly proficient bilingual speakers interact during online speech processing, by examining the phonetic outcome of mixed-language production in speakers of Bengali and Indian English.
 Specifically, we examined the spectral quality of two L2 vowel categories \nt{2} and \nt{\ae} produced in two experimental conditions: unilingual (English) and mixed (switching from Bengali to English).
@@
 (controlled) code-switching does not significantly alter the behavior under study.
+However, the magnitude of observed transfer is greater in the code-switching paradigm.
+These findings add to empirical data, highlight the need for diversifying the set of language pairs and populations in transfer studies, and encourage more within-study comparisons of paradigms and features, towards establishing better methodological norms.
+Ultimately, these results contribute to ongoing discussion about the implications of short-term phonetic interaction as a window to understanding language as a mental and social system.
 
@@
 
-\section*{Data availability}
+\section*{Data availability} \label{data_availability}
 The materials, data files, and analysis code for this study are openly available in OSF at:  \url{https://doi.org/10.17605/OSF.IO/DSB2X}.
@@
 		& Estimate  & Std. Error & t value & p value          \\ \midrule
-		(Intercept)           & -0.053474 & 0.178435   & -0.300  & 0.769            \\
-		time                  & 0.139140  & 0.011556   & 12.040  & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}                & -0.147480 & 0.223022   & -0.661  & 0.519            \\
-		contexte              & -0.013419 & 0.068357   & -0.196  & 0.846            \\
-		tasks                 & 0.330034  & 0.026586   & 12.414  & \textbf{< 0.001 ***} \\
-		time:vowel\textipa{2}           & -0.170892 & 0.013469   & -12.688 & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte       & -0.137062 & 0.060884   & -2.251  & \textbf{0.031 *}   \\
-		vowel\textipa{2}:tasks          & 0.009002  & 0.038498   & 0.234   & 0.815            \\
-		contexte:tasks        & 0.073341  & 0.037597   & 1.951   & 0.051 .          \\
-		time:tasks            & -0.121604 & 0.013469   & -9.028  & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte:tasks & -0.195613 & 0.054338   & -3.600  & \textbf{< 0.001 ***} \\ \bottomrule
+		(Intercept)           & -0.053 & 0.179   & -0.300  & 0.769            \\
+		time                  & 0.139  & 0.012   & 12.040  & \textbf{< 0.001 ***} \\
+		vowel\textipa{2}                & -0.147 & 0.223   & -0.661  & 0.519            \\
+		contexte              & -0.013 & 0.068   & -0.196  & 0.846            \\
+		tasks                 & 0.330  & 0.027   & 12.414  & \textbf{< 0.001 ***} \\
+		time:vowel\textipa{2}           & -0.171 & 0.014   & -12.688 & \textbf{< 0.001 ***} \\
+		vowel\textipa{2}:contexte       & -0.137 & 0.060   & -2.251  & \textbf{0.031 *}   \\
+		vowel\textipa{2}:tasks          & 0.009  & 0.038   & 0.234   & 0.815            \\
+		contexte:tasks        & 0.073  & 0.038   & 1.951   & 0.051 .          \\
+		time:tasks            & -0.122 & 0.013   & -9.028  & \textbf{< 0.001 ***} \\
+		vowel\textipa{2}:contexte:tasks & -0.196 & 0.054   & -3.600  & \textbf{< 0.001 ***} \\ \bottomrule
 	\end{tabular}
@@
 		& Estimate & Std. Error & t value & p value          \\ \midrule
-		(Intercept)           & 0.23202  & 0.22100    & 1.05    & 0.304            \\
-		vowel\textipa{2}                & -0.38596 & 0.14858    & -2.60   & \textbf{0.015 *}   \\
-		time                  & 0.04287  & 0.00761    & 5.63    & \textbf{< 0.001 ***} \\
-		contexte              & -0.01720 & 0.02823    & -0.61   & 0.548            \\
-		tasks                 & -0.08758 & 0.01752    & -5.00   & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:time           & -0.16289 & 0.00888    & -18.35  & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte       & 0.11043  & 0.02751    & 4.01    & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:tasks          & -0.08983 & 0.02530    & -3.55   & \textbf{< 0.001 ***} \\
-		contexte:tasks        & -0.10990 & 0.02477    & -4.44   & \textbf{< 0.001 ***} \\
-		time:tasks            & 0.06237  & 0.00888    & 7.03    & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte:tasks & 0.03349  & 0.03557    & 0.94    & 0.34652          \\ \bottomrule
+		(Intercept)           & 0.232  & 0.221    & 1.05    & 0.304            \\
+		vowel\textipa{2}                & -0.386 & 0.148    & -2.60   & \textbf{0.015 *}   \\
+		time                  & 0.043  & 0.008    & 5.63    & \textbf{< 0.001 ***} \\
+		contexte              & -0.017 & 0.028    & -0.61   & 0.548            \\
+		tasks                 & -0.087 & 0.017    & -5.00   & \textbf{< 0.001 ***} \\
+		vowel\textipa{2}:time           & -0.163 & 0.009    & -18.35  & \textbf{< 0.001 ***} \\
+		vowel\textipa{2}:contexte       & 0.110  & 0.028    & 4.01    & \textbf{< 0.001 ***} \\
+		vowel\textipa{2}:tasks          & -0.090 & 0.025    & -3.55   & \textbf{< 0.001 ***} \\
+		contexte:tasks        & -0.110 & 0.025    & -4.44   & \textbf{< 0.001 ***} \\
+		time:tasks            & 0.062  & 0.009    & 7.03    & \textbf{< 0.001 ***} \\
+		vowel\textipa{2}:contexte:tasks & 0.033  & 0.036    & 0.94    & 0.34652          \\ \bottomrule
 	\end{tabular}
@@
 \section*{Figures}
+
+\begin{figure}[h] %\centering
+	\includegraphics[scale=0.7]{vowels_bengali}
+	\caption{Vowel chart for oral monophthongs in Bengali, from \cite{ghosh2016bengali_vowels}}
+	\label{vowels_bengali}
+\end{figure}
+
+\newpage
 
@@
 \begin{figure}[h] %\centering
-	\includegraphics[scale=0.7]{vowels_bengali}
-	\caption{Vowel chart for oral monophthongs in Bengali, from \cite{ghosh2016bengali_vowels}}
-	\label{vowels_bengali}
-\end{figure}
-
-\newpage
-
-\begin{figure}[h] %\centering
 	\includegraphics[scale=1]{vowels_unilingual_ie_bengali}
-	\caption{Participants' unilingual English vowels (solid ellipse) compared to Bengali categories from \cite{shruticorpus}}
+	\caption{Participants' unilingual English vowels (solid ellipse) compared to Bengali categories from the SHRUTI corpus \citep{shruticorpus}}
 	\label{vowels_unilingual_ie_bengali}
@@
 	\label{vowels_e_b}
+\end{figure}
+
+\newpage
+
+\begin{figure} \label{figure_gender_normalization}
+	\centering
+	\begin{subfigure}[t]{0.5\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{gender_before} 
+		\caption{before gender normalization} \label{gender_before}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[t]{0.5\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{gender_after} 
+		\caption{after gender normalization} \label{gender_after}
+	\end{subfigure}
+\caption{Gender differences in Lobanov-normalized vowels}
 \end{figure}

### 2022-06-01 RCode/transfer_analysis.Rmd
BLC revision final

@@
 
-## LME models:
+## Gender differences in IE vowels:
+```{r}
 
-```{r, eval=FALSE, include=FALSE}
-# plot for English vowels:
+# Lobanov-normalized FFs-- before gender normalization
+xlim = c(3, -3)
+ylim = c(4, -4)
 
-# Lobanov-normalized FFs
-xlim = c(2.5, -2)
-ylim = c(2.3, -2)
-
-with(df_lobanov_gender_time, plotVowels(f1, f2, vowel, group = interaction(task,context),
-                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = interaction(task,context),
-                               var.col.by = interaction(task,context), var.sty.by = gender, 
+with(df_lobanov_time, plotVowels(f1, f2, vowel, group = gender,
+                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = vowel,
+                               var.col.by = vowel, var.sty.by = gender, 
                                cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
-                               ellipse.line = F,  ellipse.fill = F,
+                               ellipse.line = T,  ellipse.fill = T,
                                poly.line = F, poly.order = c("\\ae","\\vt"),
-                               xlim = c(1,-1), ylim = c(0.5,-1), 
+                               xlim = xlim, ylim = ylim, 
                                legend.kwd = "bottomright", legend.args = gender,
-                               fill.opacity = 0.1, pretty = TRUE))
-
-```
-
-```{r, eval=FALSE, include=FALSE}
-
-# raw FFs
-with(df_time, plotVowels(f1, f2, vowel, group = interaction(gender, task), 
-                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = vowel,
-                               var.col.by = interaction(gender, task), var.sty.by = task, 
-                               cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
-                               ellipse.line = F,  ellipse.fill = F,
-                               poly.line = F, poly.order = c("\\ae","\\vt"),
-                               xlim = c(2500, 700), ylim = c(1400, 100), 
-                               legend.kwd = "bottomright", legend.args = gender,
-                               fill.opacity = 0.1, pretty = TRUE))
+                               fill.opacity = 0.05, pretty = TRUE))
 
@@
 
-## main model
+Plots after gender normalization:
+
+```{r}
+
+# raw FFs
+with(df_lobanov_gender_time, plotVowels(f1, f2, vowel, group = gender,
+                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = vowel,
+                               var.col.by = vowel, var.sty.by = gender, 
+                               cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
+                               ellipse.line = T,  ellipse.fill = T,
+                               poly.line = F, poly.order = c("\\ae","\\vt"),
+                               xlim = xlim, ylim = ylim, 
+                               legend.kwd = "bottomright", legend.args = gender,
+                               fill.opacity = 0.05, pretty = TRUE))
+
+```
+
+## LME models:
 

### 2022-07-23 draft_BLC/blc_revision1_reviewer_response.txt
starting new revision

Manuscript BLC-21-0133.R1
Response to Reviewers

Dear Editors,

Thank you for the opportunity to submit a revised draft of our manuscript “Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals” for publication in Bilingualism: Language and Cognition. We appreciate the time and effort from your end and that of the reviewers in providing feedback on our manuscript. We have incorporated the suggestions made by the reviewer. Please see below, for a point-by-point response to the reviewers’ comments and concerns. All page numbers refer to the revised manuscript file uploaded in the portal.

Sincerely,
Auromita

---------------------------------------------------

Reviewer: 2

Comments to the Author
I have now had a chance to read the latest version of the manuscript as well as the authors’ response letter. I am satisfied with the changes (or at least most of the changes) made by the authors in response to the reviews and believe the manuscript is closer to being ready for publication than it was. I am satisfied with the fact that the Discussion section has been expanded and strengthened (with theoretical implications being discussed), a lot more information concerning the participants is provided, and an attempt to sample-size justification is made. My only negative substantial comment has to do with sample size and power analyses. The rest are superficial, minor, and I outline them below. For now, let me say that I appreciate the authors’ effort to reach out to the participants to ask them to answer a language-background questionnaire. This information was needed and is now provided. The Discussion now reviews the findings considering past findings (the contextual literature) and outlines some ideas for further research. This is much appreciated. As I said, I believe the manuscript is almost ready for publication.

Let me talk now about sample-size justification. I appreciate the authors’ attempt at justifying their sample size by running some power analyses. Power analyses are difficult when one does not know what effect size to expect, and it is meaningless to run them a posteriori (using observed effect size as expected effect size). The authors are correct, in my opinion, when they observe that running post-hoc power analyses is meaningless, and yet I appreciate the fact that they were willing to run them anyway to satisfy my petition. Now that this is out in the open, I suggest the authors delete the power analysis section from the manuscript: it unnecessarily exposes them to future criticism. Power analyses in general, even when done a priori (during planning), are meaningless because most bilingualism research does not have established prior findings to base some of its assumptions upon, like minimal effect size of interest. (They are useful as an exploration, an estimation of minimal sample size, but, as a field, we cannot take them seriously (yet).) In the absence of this, which is true of this study (and the authors observe), a few strategies are possible: 1) assume a minimal-effect-size-of-interest of d = 0.4 or so (see Brysbaert’s papers below), focusing on normalized effect size and not effect size in absolute values (such as Hertz), and focusing on a single pairwise comparisons of interest rather than the whole combination of fixed effects and interactions one will end up running; 2) forget about power and focus on precision (get the largest sample one can get and leave it at that) (see Cumming’s book, below); 3) forget about wanting to run a confirmatory, dichotomous data analysis (such as mixed-effects regression or analyses of variance) and focus on effect-size estimation and exploration (what most of us do is exploration anyway, and yet we use confirmatory data analyses; see Cumming’s book; however, the field is dependent on inferential statistics and I understand that, as authors, we feel we have to run them to get published); and, 4) acknowledge that our studies are underpowered and thus focus on a combination of providing effect sizes (for future meta-analyses) and justifying one’s sample size as best as possible, being honest. I suggest the authors do 4 (explain and justify why they have a small sample size) and delete their power analyses from the manuscript. There is no way, given what the authors observe and what Brysbaert shows in his papers (below), that a sample size of 10 allows anybody to have sufficient power for almost anything of interest in bilingualism research. Given the complexity of the study’s design, chances are one would need over 100 participants to have sufficient power. This is not going to be possible for all the things the authors explain. So, here is what I recommend the authors do:

1.      Delete the power analyses section from the manuscript.
2.      Focus on explaining and justifying their sample size. This is already done (in the latest version of the manuscript, not the first one). I am personally satisfied with the authors’ justification: they are using strict inclusion criteria in a population that, although large, is extremely variable, and they saw the need to exclude lots of participants. I understand that one cannot just be running one’s studies for years to obtain a sample size of 100. In this case, this would mean excluding another 180 a posteriori. In this case, the sample size has been determined based on resource constraints (Lakens, see below). So, I fully understand now why the authors’ sample size is small and that it would be extremely difficult to have a larger sample size (given the strict inclusion criteria and human limitations: we have to get published and our projects are small and have few resources). Still, this does not mean that a sample size of 10 allows us to make any strong claims. So, I would leave it at that provided the authors state clearly that they are aware their sample size is very small and that only future research with larger sample sizes may confirm or reject their findings, that their findings are preliminary, and that prior research has also used small samples sizes (give some examples). I suggest the authors’ delete their power analysis section because it just shows the study is likely tremendously underpowered and that, in order to not reveal it, the authors ran obscure calculations to try to justify a small sample size.

Brysbaert, M. (2019). How many participants do we have to Include in properly powered experiments? A tutorial of power analysis with reference tables [Review of How many participants do we have to Include in properly powered experiments? A tutorial of power analysis with reference tables]. Journal of Cognition, 2, 16.
Brysbaert, M. (2021). Power considerations in bilingualism research: Time to step up our game. Bilingualism: Language and Cognition, 24(5), 813–818.
Cumming, G. (2013). Understanding the New Statistics: Effect Sizes, Confidence Intervals, and Meta-Analysis. Routledge.
Lakens, D. (2022). Sample size justification. Collabra. Psychology, 8(1). https://doi.org/10.1525/collabra.33267

Authors' response (AR):

The second comment of substance I have has to do with the procedure. I do not fully understand the procedure of the cued-picture naming task. Can you please explain it in greater detail and perhaps provide some sample trials? You could give the examples in the text and then also a schema in a Figure or something like that. To be honest, I don’t know whether the participants are reading out loud (and the cues are orthographic) or they are naming pictures (and then I don’t understand how language is cued). Please, just explain, give examples, and consider adding a schema in a figure. (“The picture of a fish was used in all {sic} trials…” ????)

AR:

The third comment of substance, which I hadn’t noticed before, has to do with acoustic analyses. On page 26, it is said that formant values were Lobanov-normalized. I don’t think this is possible in this study. I believe the authors did something akin to Lobanov-normalization but quite distinct from it. In Lobanov normalization, values are z-scored with reference to values of all the vowels in a speaker’s vowel system. In other words, for one to be able to do a Lobanov-normalization for a given vowel category (the tokens from a category) one MUST have tokens, collected from the same speaker, of all the other vowels in the speaker’s vowel system. In this case, the given vowel token is expressed with reference to its location in a speaker’s vowel space. This DOES allow to normalize across genders. This is a great normalization procedure, but one needs to have collected samples of all the vowels in the system, or at least of the corner vowels, which outline the limits of the system. I don’t think this is what the authors did here (because I don’t think they collected samples of all the vowels of English or at least the corner vowels). Their normalization method is akin to the Lobanov method in that they did z-scoring with all the vowels they had. However, if they didn’t have samples of all the vowels, this is not a true normalization method, and clearly not Lobanov’s normalization procedure. Please, explain better and then explain what this normalization method accomplished. (Since mixed-effects regression was to be used, and all factors are within-speaker, normalization was likely unnecessary.)

AR:

A quick note about terminology. I am a bit uncomfortable with the authors’ use of the terms “unilingual” and “mixed.” In past research, “unilingual” tends to refer to entire recording sessions in which a single language was used, and “bilingual” refers to recording sessions in which two languages were used. I don’t object to the use of “mixed” here, but only to the use of “unilingual.” It seems that all recording sessions in this study were bilingual or mixed. The difference between the two condition has to do with whether there was a code- or language-switch, that is, whether the target English word was produced after another English word or after a Bengali word. Following the terminology used by others (see Olson’s articles, for instance), I believe you should use “switch” vs “nonswitch” contexts or conditions rather than “unilingual” vs “mixed”. If not, at least explain your use of terminology at the outset and why you use it differently from others.

AR:

Finally, I have a comment about punctuation. The authors, in many locations, use the n-dash (or minus sign) instead of the colon. One example: Page 24, line 51. “Language cue- a word…” I have seen this use of the n-dash throughout the manuscript. I don’t think this follows convention. I believe a colon should be used in most places where an n-dash is used. Please, consider, decide, and revise throughout if needed.

AR:

The following are minor comments:

1.      Page 3, lines 51-53. “However, the reported…” Please, include the full list of references (citations) you have in mind to make this statement.

2.      Page 4, lines 20-21: “properties: F1 and F2, of two…” Revise punctuation. Should likely be: “properties, F1 and F2, of two…”

3.      Page 5, lines 47-50: “other additional languages.” Delete either ‘other’ or ‘additional’. I believe this is redundant.

4.      Page 6, lines 54-55: “suggest that for the present” > “suggest that, for the present population, mixed-language processing. Add comma.

5.      Throughout: F1XF2. Perhaps F1 * F2 or F1 x F2. And perhaps then refer to this as “vowel chart”, following convention.

6.      Page 7. Lexical sets: KIT, FLEECE, etc. It turns out you never use the lexical set terminology again in the manuscript. I don’t see the point of introducing them here. Perhaps just delete.

7.      Page 8, lines 43-49. This is the first place where I noticed this, but I have seen this mistake throughout. I am sure this is done by the automatic reference system used in Latex. Notice that, here, the citations should have the name of the author outside of the parentheses and the years should be inside the parenthesis. The mistake is that both author name and year are inside the parentheses. This is a mistake because, without reading the author names (and, if they are inside the parentheses, they are supposedly not part of the sentence), the sentence does not make sense. Please, do an automatic search and revise all instances of citations. I believe there are dozens of places where this mistake is made.

8.      Page 13, lines 10-17. I don’t think this statement is accurate. Perhaps revise after checking the articles cited here. For instance, Simonet (2014) and Simonet/Amengual (2020) do not examine “only transfer effects on L1 vowels.” In these studies, there is a group of Catalan L1 Spanish L2 speakers and a group of Spanish L1 Catalan L2 speakers. Only Catalan vowels are compared across lingualism conditions (unilingual, bilingual settings). The same transfer or interference effects are found for both groups. This means, I believe, that both L1 and L2 vowels have been found to be modulated by language mode or lingualism.

9.      Page 14, line 43. “cannot be treated as an atomic measure.” I don’t know what this means. Could you explain or revise wording?

10.     Page 19, section 1.7. Not sure this list merits its own section. I believe this should be incorporated in the previous section. I also believe the question might need to be revised. You do not only check if vowel categories are modified in one setting relative to the other setting. You also check the direction and size (or extent) of the modification. Should you have a main research question and some follow-up, secondary questions focused on directionality and extent-of-change?

11.     Page 22, lines 10-13. “all participants … (n = 10)”. I don’t think including n = 10 is necessary here. You have said before you had 10 participants and this sentence already says “all.”

12.     Page 23, lines 28-29. “used is creative ways” > “used in creative ways”

13.     Page 25, line 20. “orthographies … are visually distinct” > “orthographies … are distinct” (orthographies are necessarily visual).

14.     Page 37, lines 46-51. Please, revise sentence top include figure numbers, although you will hopefully delete the whole section.

15.     Page 42, line 26-27. “it is a potential”. What does “it” refer to? Please, spell out, e.g.: “, differences reading proficiency are a potential…”

16.     Page 43, full last paragraph in page. This paragraph is quite intriguing. It seems it contains the gist of a great idea or observation. However, the idea is so vague that the paragraph becomes empty. Do the authors have something to say? Please, say it clearly (explaining it in some detail), even if speculative (you have a right to speculation after having explained the data). Are they not ready to explain their idea in some detail? Please, delete hints to ideas of they are not to be explained.

17.     Page 44, line 11: “in a languages” > “in a language” or “in languages”

18.     Page 46, line 9. “Our third research question Paradigm concerned…” I am a bit confused. I understood “Paradigm” to be a hypothesis, not a research question. Please, check section 1.7 and decide if Asymmetry and Paradigm and research questions or hypotheses. If they are questions, what are the hypotheses in this study? Also, when referring to Paradigm and Asymmetry in the Discussion section, please notice that, in most cases, surrounding commas are needed: “Our third research question, Paradigm, concerned…”
Congratulations on a fine study and thank you for the opportunity to learn about this study before it was published. I hope some of my comments are found to be useful.

### 2022-07-23 draft_BLC/BLC_AM-ID_manuscript.tex
starting new revision

@@
 \begin{myparindent}{0pt}
-Short title: Phonetic transfer\\
+Short title: Phonetic - transfer\\
 

### 2022-07-25 draft_BLC/blc_revision1_reviewer_response.txt
power analysis, picture task figures

@@
 
-AR:
+AR: In this task the picture, named in either Bengali or English, serves to prime the participant for that particular language.
+The following target word then either complies with the prime (unilingual condition) or deviates from it (inducing switching; mixed condition).
 
@@
 Also, when referring to Paradigm and Asymmetry in the Discussion section, please notice that, in most cases, surrounding commas are needed: “Our third research question, Paradigm, concerned…”
+
 Congratulations on a fine study and thank you for the opportunity to learn about this study before it was published. I hope some of my comments are found to be useful.

### 2022-07-25 draft_BLC/BLC_AM-ID_manuscript.tex
power analysis, picture task figures

@@
 \usepackage{subcaption}
+
+\usepackage{standalone} %to include separate .tex files that contain mages
 
@@
 \begin{myparindent}{0pt}
-Short title: Phonetic - transfer\\
+Short title: Phonetic transfer\\
 
@@
 \subsection{Participants} \label{participants}
-Participants were recruited through an open call  circulated through social media at the EFL University campus in Hyderabad, India.
+Participants were recruited through an open call circulated on social media at the EFL University campus in Hyderabad.
 Interested volunteers filled a brief language background questionnaire.
@@
 As noted in section \ref{bengali_english_in_india}, India has an extremely multilingual an linguistically diverse population.
-EFLU being a central university houses students from different parts of the country, who speak both different L1s, and different varieties of English.
+The university campus comprises of students from different parts of the country and world, with varying L1s and different varieties of English.
 Our aim in screening participants was to minimize between-participant random variability in the study, while maintaining ecological validity considering the actual day-to-day settings in which transfer behaviors likely take place.
-We chose two metrics to do this: (i) where the participant grew up; and (ii) number of years of formal education in Bengali.
+We chose two metrics to do this: (i) where the participant grew up; and (ii) amount of formal education in Bengali.
 The first ensures that the variety of Bengali that individuals are exposed to in their surroundings is similar across participants.
@@
 Thus, we wanted to ensure that any between-participant asymmetries in the pattern of shift reflected individual differences in transfer, rather than different acoustic targets.
-Another reason to control for location was that it is a fairly reliable predictor of what languages the person was educated in, owing to how the education system functions in India (we elaborate on this below).
+Another reason to control for location was that it is a fairly reliable predictor for language(s) of education, owing to how the education system functions in the country (we elaborate on this below).
 
 The second metric is Medium of Instruction (MoI) in school.
-Most schools in India follow the Three Language Policy, wherein formal education is provided in English, Hindi, and the regional language.
+Most schools in India follow the Three Language Policy, wherein formal education is provided in English, Hindi, and a regional language.
 (The idea and implementation of this policy, and its implications for linguistic minorities, has been a matter of much public debate as well as scholarly work; for discussions see, e.g.
@@
 Sustained formal education in a language is a reliable predictor of LSRW (listening, speaking, reading, writing) skills.
-Since this study provides stimuli in the form of text, we wanted to ensure that all participants were sufficiently comfortable with reading and speaking both Bengali and English, to avoid confounds and uncontrolled variability due to reading/task effort.
+Since this study uses written stimuli, we wanted to ensure that all participants were sufficiently comfortable with reading and speaking both Bengali and English, to avoid confounds and uncontrolled variability due to reading/task effort.
 
@@
 Together, these incorporate questions about language history, usage, attitudes, and switching experience.
-The complete questionnaire, along with responses, are available in the ``Data" component of the OSF repository for this project (c.f.
+The complete questionnaire, along with responses, is available in the ``Data" component of the OSF repository for this project (c.f.
 section \ref{data_availability}).
 
-In addition to Bengali and English, all participants reported knowing other languages, the most common being Hindi (n=10).
+In addition to Bengali and English, all participants reported knowing other languages, the most common being Hindi.
 Since this situation is representative of the present population and our study concerns only Bengali-English interaction, we only focused on these languages in the survey.
@@
 This is a common situation for many educated sections of society, since institutions of higher education beyond the school level largely use English (for discussions about the language politics of higher education in India, see, for example, \cite{mohanty2009multilingual}).
-The language attitude section of the questionnaire reveals an interesting insights about the linguistic landscape of the population, and how people experience language in such a setting.
+The language attitude section of the questionnaire reveals interesting insights about the linguistic landscape of the population, and how people experience language in such a setting.
 When asked how much ``like themselves" participants feel while using each language, response were at ceiling for both languages.
@@
 These responses shed light on how multiple languages fit seamlessly into different facets of daily life to create a ``continuous" linguistic experience (c.f.
-"plurality" as defined by \cite{khubchandani2021plural}).
+``plurality" as defined by \cite{khubchandani2021plural}).
 In such a scenario, it is not surprising that language-switching is common, expected, as well as used is creative ways with communicative purpose.
 Responses to the section on language switching demonstrate this.
-On average, participants do not report being often unable to recall words in the target language while speaking Bengali (average rating 2.5 on a 1-6 likert scale where 1=never and 6=always) or English (average rating 2.3), indicating that switching is not strictly necessary for fulfilling lexical gaps.
+On average, participants do not report being often unable to recall words in the target language while speaking Bengali (average rating 2.5 on a 1-6 likert scale where 1=never and 6=always) or English (average rating 2.3), indicating that switching is not strictly necessary for filling lexical gaps.
 Indeed, participants do not report switching between languages without awareness (1.8) or control (1.25).
@@
 This fact, combined with the linguistic situation described here and in section \ref{bengali_english_in_india}, are important to bear in mind while interpreting the results of a transfer study in the present population.
+
+Given our inclusion criteria, we were able to recruit 10 participants for this study.
+Although existing studies in the field have used comparable participant numbers (e.g ), we acknowledge that this is a small sample size.
+As detailed above, even after a recruitment call that already controlled for language background and education level (recruiting Bengali-English bilinguals in a university campus), we were able to include only 10 out of 28 volunteers (35.7\%) in the final study.
+Note that the criteria used here establish only a minimal level of between-participant similarity in exposure, dominance, habits, and proficiency, such as is extremely common in most comparable studies.
+This demonstrates the degree of linguistic variability in the present population.
+Norms for experimental research on bilingualism have largely been established in a Western context.
+In settings where linguistic variability at the population level is much greater than these canonically studied groups, establishing comparable levels of experimental control requires significantly greater resources.
+Given that small sample sizes are likely to have low statistical power, it is fruitful to treat these findings as preliminary.
+Replication with larger sample sizes will help to confirm the observed effects, and provide much-needed data diversity (c.f.
+discussion in section \ref{introduction}).
+A more detailed discussion of power analyses in bilingualism research is included as supplementary material on OSF (c.f.
+section \ref{data_availability}).
+We hope that this can facilitate more conversation around establishing norms for power and sample size in our field.
 
@@
 < Insert Figure \ref{vowels_e_b} about here >
-
-\subsection{Power analysis}
-Owing to participant availability given our inclusion criteria, we were able to recruit 10 participants for this study (c.f.
-section \ref{participants}).
-We acknowledge that this sample size is on the lower side for studies with comparable designs.
-To confirm whether the experiment design was powerful enough to detect the effects we are investigating, we carried out post-hoc power analyses using the \textit{mixedpower} package \citep{kumle2021estimating} in R.
-The package simulates responses based on a given dataset and runs the specified analyses on these new data, estimating power by calculating how frequently significant effect are obtained in the simulated data.
-The target of the analysis was to calculate the power for detecting an effect for our main variables of interest: vowel*context (to test the hypotheses L2-vowels and Asymmetry), and vowel*context*task (to test the hypothesis Paradigm).
-
-Power is a function of sample size and effect size.
-Thus, a key issue in power analyses is to estimate the expected effect size for the treatment/factor of interest \citep{brysbaert2018power, kumle2021estimating}.
-This may be expressed by a variety of measures, including Cohen's d, $\eta ^2$, and model estimates ($\beta$).
-
-Using observed effect sizes from the data for power analyses has been shown to be meaningless \citep{hoenig2001abuse, gelman2019observed_effect_size}, since such ``observed power" is a function of the p-value.
-Therefore, we need a reasonable estimate of effect size.
-One way to do this is to use effect sizes reported in other existing literature (however, see \cite{brysbaert2018power} who recommends against using effect sizes from single published studies, as these tend to be inflated), or meta analyses.
-An alternative approach is to define a smallest effect size that is meaningful to us (smallest effect size of interest; SESOI \citep{kumle2021estimating}), and test whether the study has enough power to detect such an effect, if it were to exist.
-This value needs to be theoretically motivated.
-To the best of our knowledge, there is no existing literature on typical effect sizes for spectral differences in cross-language vowel production, and not many studies that report effect sizes.
-Only more discussion in this field will lead to a consensus on what is considered a typical or minimally meaningful cross-language effect.
-Similarly, while differences between language-switching paradigms have been informally discussed in previous literature, we are not aware of any study that has measured these differences in a controlled experimental setup, and thus no reported effect sizes for between-paradigm differences in phonetic transfer to act as a guideline.
-
-Cohen \citeyearpar{cohen1988statistical} classifies effect sizes (Cohen's d values) as small ($\leq$0.2), medium (0.5), and large ($\geq$ 0.8).
-Although the usefulness of this classification has been debated in the literature, these are often used as ballpark values for power analyses in psychological studies (for a discussion, see \cite{brysbaert2018power}).
-However, this approach does not generalize well to mixed-effects analyses, particularly for complex models with multiple random effects and interactions, as is typical in psycholinguistic and bilingualism research.
-``Determining the SESOI for (G)LMMs is difficult in a simulation-based approach where effect sizes are indicated through the model’s unstandardized beta coefficients...relating effect sizes to beta coefficients in complex models is far from trivial and the authors therefore refrain from making specific recommendations.
-\citep{kumle2021estimating}".
-
-The results of a power analysis are only as meaningful as the assumptions that go into it.
-To avoid making arbitrary assumptions, we followed the approach in \cite{brysbaert2018power} (and discussed in \cite{kumle2021estimating}) of extracting the effect sizes from the study of interest and directly modifying all the beta coefficients by a fixed percentage.
-Note that we are not assuming that doing so gets us closer to a "true effect size".
-Instead, we run simulations with a range of effect sizes and report the power of our study to plausibly detect an effect of that size.
-The results are shown in the plots in figures \ref{figure_power_analysis_f1} and \ref{figure_power_analysis_f2}.
-Each panel presents power analysis at a different effect size, and is based on 500 simulations.
-The plots show power (y-axis) over a range of sample sizes (x-axis), given the assumed effect size.
 
@@
 
-\begin{figure} \label{figure_power_analysis_f1}
+\begin{figure}[!htb] \label{figure_picture_task}
 	\centering
-	\begin{subfigure}[t]{0.8\textwidth}
+	\begin{subfigure}[t]{0.49\textwidth}
 		\centering
-		\includegraphics[width=\textwidth]{powerplot_f1_minus15} 
-		\caption{Effect size = observed effect size - 15\%} \label{f1_minus15}
+		\includestandalone[mode=image]{picture_task_eng}
+		\caption{English (non-switch) trial} \label{picture_task_english}
 	\end{subfigure}
-
-	\begin{subfigure}[t]{0.8\textwidth}
+	\begin{subfigure}[t]{0.49\textwidth}
 		\centering
-		\includegraphics[width=\textwidth]{powerplot_f1_minus5} 
-		\caption{Effect size = observed effect size - 5\%} \label{f1_minus5}
+		\includestandalone[mode=image]{picture_task_bengali} 
+		\caption{Bengali (switch) trial} \label{picture_trial_bengali}
 	\end{subfigure}
-
-	\begin{subfigure}[t]{0.8\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{powerplot_f1_plus5} 
-		\caption{Effect size = observed effect size + 5\%} \label{f1_plus5}
-	\end{subfigure}
-
-	\begin{subfigure}[t]{0.8\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{powerplot_f1_plus15} 
-		\caption{Effect size = observed effect size + 15\%} \label{f1_plus15}
-	\end{subfigure}
-
-	\caption{Power analysis for fixed effects of interest at varying sample sizes: F1}
-
-\end{figure}
-
-\newpage
-
-\begin{figure} \label{figure_power_analysis_f2}
-	\centering
-	\begin{subfigure}[t]{0.8\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{powerplot_f2_minus15} 
-		\caption{Effect size = observed effect size - 15\%} \label{f2_minus15}
-	\end{subfigure}
-
-	\begin{subfigure}[t]{0.8\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{powerplot_f2_minus5} 
-		\caption{Effect size = observed effect size - 5\%} \label{f2_minus5}
-	\end{subfigure}
-
-	\begin{subfigure}[t]{0.8\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{powerplot_f2_plus5} 
-		\caption{Effect size = observed effect size + 5\%} \label{f2_plus5}
-	\end{subfigure}
-
-	\begin{subfigure}[t]{0.8\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{powerplot_f2_plus15} 
-		\caption{Effect size = observed effect size + 15\%} \label{f2_plus15}
-	\end{subfigure}
-
-	\caption{Power analysis for fixed effects of interest at varying sample sizes: F2}
-
+\caption{Picture-naming task sample sequence: language cue (not read out), picture (named as per lang cue), target (read out), distractor}
 \end{figure}

### 2022-07-25 draft_BLC/power_analysis.tex
power analysis, picture task figures

--- 2022-05-09 draft_BLC/BLC_AM-ID_manuscript.tex
@@
-% !TeX document-id = {6d544fe2-4255-4108-8022-d809ff1d2c53}
-% TeX program = xelatexmk %doesn't work, no xelatexmk option under 'build'. Changed default engine to xelatex (options-->configure-->build-->default build-->xelatex)
-% see http://info.semprag.org/basics for a full description of this template
-%\documentclass[times,linguex]{glossa}
 \documentclass[12 pt]{article}
@@
 
-\title{Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals}
+%-----------------------------
 
-% \author[Auromita \& Indranil]% short form of the author names for the running header. If no short author is given, no authors print in the headers.
-% {%as many authors as you like, each separated by \AND.
-%   \spauthor{Auromita Mitra\\ 
-%   \institute{The EFL University}\\
-%   \small{%105, Bd. Raspail, 75005 Paris\\
-%   auromita.mitra@gmail.com}
-%   }
-%   \AND
-%   \spauthor{Indranil Dutta \\
-%   \institute{Jadavpur University}\\
-%   \small{%Warmoesberg 26, 1000 Brussel\\
-%   indranildutta.lnl@jadavpuruniversity.in}
-%   }%
-% }
-%\usepackage{gb4e}
+%-----------------------
+
+\title{Supplementary material: power analysis}
+\date{}
+%\maketitle
 
@@
 	\bibliographystyle{apacite}
-%\sffamily
-%\maketitle
 
-%Word count: 9263
+\section*{Supplementary material: power analysis}
 
-\begin{singlespace}
+Owing to participant availability given our inclusion criteria, we were able to recruit 10 participants for this study (c.f.
+section Participants in main article).
+As part of the peer review process, we carried out post-hoc power analyses using the \textit{mixedpower} package \citep{kumle2021estimating} in R.
+Although there is a body of literature that debates the usefulness of post-hoc power analyses in interpreting the findings of a completed study where statistical analyses have already been performed \citep{lakens_2021,dziak2020interpretation,lenth2007post,gelman2019observed_effect_size}, we have included this section as a discussion around considerations for power calculation in bilingualism research.
 
-\begin{myparindent}{0pt}
-Short title: Phonetic transfer\\
-
-Full title: Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals\\
-
-Authors:\\
-
-Auromita Mitra
-
-EFL University,
-
-Hyderabad,
-
-India\\
-
-Indranil Dutta
-
-Jadavpur University,
-
-Kolkata,
-
-India\\
-
-Acknowledgments:\\
-We are grateful to Maumita Bhowmik and Anannya Mondal for their help with data collection and processing, and to all the participants of the study.\\
-
-Address for correspondence:\\
-Auromita Mitra,
-
-EFL University,
-
-Hyderabad- 500007,
-
-auromita.mitra@gmail.com
-
-\end{myparindent}
-
-\end{singlespace}
-
-\newpage
-\begin{myparindent}{0pt}
-Abstract:\\
-This study investigates the phonetic outcome of mixed-language speech in Bengali and Indian English, towards understanding cross-language interaction in highly proficient bilinguals.
-We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in unilingual vs.
-mixed productions.
-Results reveal dynamic and asymmetrical changes in both vowels;  \nt{2} shifting towards a related L1 category \nt{a:}.
-We interpret this as a temporary increase in cross-language phonetic interaction during mixed-language use, leading to a shift towards L1 norms, evidencing transfer effects on L2 vowels for the first time.
-We elicit mixed productions through two tasks: cued picture-naming and code-switching, to assess if experimental paradigm independently influences the behavior under study.
-Results reveal no difference across paradigms.
-We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction, postulated discursive factors in code-switching, and the issue of comparability between paradigms in transfer studies.\\
-
-Keywords: phonetic transfer; vowel quality; bilingual speech production; L1-L2 interaction; Indian English; Bengali
-
-\end{myparindent}
-
-\newpage
-
-\section{Introduction: Phonetic transfer}
-
-Bi/multilingual speakers can distinguish between the phonetic norms of their languages and maintain separate sound categories for each language \citep{caramazza1973acquisition,macleod2010impact,bosch2003simultaneous}.
-However, these categories are not autonomous-- they influence each other across languages in both perception and production \citep[e.g.][]{flege1995second,fowler2008cross,flege2002assessing}, and the nature of such interaction provides crucial insights into how language `systems' are cognitively represented and processed.
-Based on the productions of proficient bilingual speakers of Bengali and English, this study reports that cross-language phonetic influence temporarily increases during mixed-language use.%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
-%show cross-language-- rephrase
-
-%keep just the types here, and put the next 2 lines in the causes&effects section?
-Cross-language influence at the level of sounds can be studied in broadly two kinds of conditions:
-(i) While a bilingual speaker is operating in any one of their languages, by comparing bilingual speech to monolingual norms \citep[e.g.][]{guion2003vowel,caramazza1973acquisition,flege1987production}; %These studies often compare the speech of bilinguals to monolingual norms, and view the influence as changes to long-term memory representations as a result of acquiring an L2 \citep{guion2003vowel,caramazza1973acquisition,flege1987production};
-(ii) When both languages of a bilingual speaker are co-activated, by comparing speakers' mixed-language speech to their own norms while using a single language \citep[e.g.][]{grosjean1994going, bullock2009trying,elias2017effects, simonet2014phonetic}.
-%These compare productions during mixed-language use to participants' own productions while using a single language.
-The resulting influence is variously thought to involve short-term memory, online processing costs, language mode, and context-awareness.
-The present study is concerned with the latter.
-In production, this kind of influence has been variously termed \textsc{transfer, drift, accommodation,} and \textsc{interference}-- we use the term \textsc{transfer} here, to indicate any interaction between two sets of phonetic norms.
-
-Existing research on the phonetic effects of mixed-language production has largely focused on a limited set of phonologically related languages.
-A majority of these studies use temporal properties of consonants (in particular, voice onset time or VOT) to measure transfer.
-However, the reported results vary greatly across studies, and appear to be contingent upon both language-specific features and language experience of the participants.
-This emphasizes the importance of considering data from a wider variety of
-%elaborate on 'great variation'-- differences between language pairs, plus sociolinguistic situation (effects of language experience etc)-- clearer.
-'Types of features'--separate line?- Brief mention - discursive factors in VOT.
-populations, language pairs,  and phonetic features in order to make meaningful generalizations. Widespread multilingualism in the Indian subcontinent suggests that phonetic
-%line--rephrase
-behavior in these populations can be particularly valuable towards understanding the nature of such cross-language interactions, as they are likely to reflect real-world experience with mixed-language processing.
-However, there is no work yet on short-term phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language.
-
-The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India.
-We measure spectral properties: F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2.
-Mixed-language data is elicited in two switching paradigms: cued picture-naming and code-switching.
-We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction.
-The results demonstrate a dynamic effect of transfer on L2 vowels during mixed-language production, \alert{but no difference in the extent of transfer between the two paradigms --and slightly different patterns of shift among the paradigms?}.
-We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.
-
-The rest of this section is structured as follows: \ref{bengali_english_in_india} provides background information on Bengali and Indian English focusing on vowel systems.
-\ref{causes} discusses two potential sources of transfer during language processing-- global co-activation of multiple languages vs.
-the local act of switching between languages.
-Given the highly multilingual setting that characterizes this population, we note that the former is expected to be a constant feature and therefore any observed phonetic interaction is better understood as resulting from the latter.
-The design of the present study is motivated on the basis of this discussion.
-A logical consequence of this model is that such phonetic interaction must be highly localized.
-\ref{duration} elaborates on this by reviewing literature on the duration and `reversibility' of transfer effects.
-A recurring pattern that emerges from existing research on short-term phonetic transfer is that both languages of the bilingual speaker are not equally affected, although the nature of these reported differences varies greatly across studies.
-In \ref{asymmetries}, we review some of these findings, focusing on proposed sources of asymmetries, in order to highlight the factors that mediate such phonetic interaction (or fail to do so).
-We argue that these results point to an urgent need for expanding the range of language pairs and sound categories examined in order to make meaningful generalizations about the nature of short-term phonetic transfer.
-Following from this, \ref{asymmetry between sounds} discusses reported asymmetries between different sound categories of a single language, and develops the hypotheses of the present study on the basis of these.
-In \ref{paradigms}, we motivate a secondary aim of this study: to verify the postulated differences between two paradigms for eliciting short-term phonetic interaction in an experimental setup, namely cued picture-naming and code-switching.
-\ref{questions_and_hypotheses} summarizes the research questions and hypotheses.
-
-Existing studies have studied phonetic transfer along different acoustic features.
-While discussing the literature in subsequent sections, we will often indicate the acoustic feature in parentheses following a citation, for clarity.
-
-\subsection{Bengali and English in India} \label{bengali_english_in_india}
-
-\subsubsection*{Demography}
-
-Bengali (also, Bangla) is an Indo-Aryan language primarily spoken in India and Bangladesh.
-In India, more than 97 million people speak Bengali as a first language (Census of India, 2011), mostly in the state of West Bengal.
-A majority of this population also speaks other additional languages.
-
-Indian English (IE) refers to the variety of English that has developed in the Indian subcontinent.
-In India, it is spoken as an L2 by 129 million people (Census of India, 2011).
-English is one of the two official languages, used in education, law, media, as a lingua franca mainly for an educated elite in most metropolitan regions, and carries a high prestige value \citep{pandey201517, kachru1981english, tollefson2014language, kachru1983indianization}.
-
-\alert{note that there are large regional variations in Indian English, and the literature on IEphonology is thus not "standardized" or uncontested.
-For the present study, we focus on a variety of IE spoken by an educated elite, used in news channels, taught in classrooms (citations)}
-
-Recent research comparing the segmental and suprasegmental properties of IE spoken in different parts of the country report many commonalities (see \cite{sirsa2013effects} for a review).
-This suggests that in spite of the varied L1s of its speakers, IE has a target phonology that is distinct from any of these, as well as from other native varieties of English.
-Thus, it is fruitful to think of regional variations as resulting from L1-influence on a common underlying target.
-Excepting a small minority of specific communities \citep{pandey201517, wells1982accents, coelho1997anglo}, there are very few L1 speakers of IE.
-
-Considering the language usage patterns suggested by this demography, the present study focuses specifically on short-term phonetic interaction during mixed-language use, because:
-
-\begin{enumerate}[label=(\roman*)]
-	\item In multilingual populations, long-term representations of sound categories are expected to be affected by multiple languages. 
-	\item Given that examples of an IE phonology without L1 `influence' are very rare, it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a given baseline condition.
-\end{enumerate}
-
-The socio-linguistic facts about English in India also suggest that for the present population, mixed-language processing of English and an L1 is an ecologically valid paradigm that is likely to be a part of everyday language experience.
-Thus, data from this population is valuable to the understanding of short-term phonetic interaction.
-
-\subsubsection*{Vowel systems} \label{vowel systems}
-The vowel inventory of Western Bangla (the variety spoken by the participants in this study) consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \citep{garry2001facts}.
-Note that there is no mid-central vowel category.
-
-The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \citep{wells1982accents, masica1972sound}.
-%Their distribution in the F1XF2 space is shown in (insert table 1.7, Pandey).
-A single mid-central vowel corresponds to the categories \nt{2,@,3:}, which are treated as distinct in many native varieties of English \citep{nihalani1979indian,wells1982accents,hickey2005legacies,bansal1969intelligibility}.
-Since the English items used in this study are traditionally transcribed with \nt{2}, we use this symbol to indicate the mid-central vowel throughout.
-
-\subsection{What causes transfer and what does it affect?}\label{causes}
-
-Existing research has distinguished changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf.
-Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during accessing, processing, or articulation of these categories (cf.
-transfer vs.
-interference \citep{grosjean2012attempt}; competence vs.
-performance interference \citep{paradis1993linguistic}).
-Given their transient nature, dynamic changes in production during mixed-language use are generally attributed to the latter, e.g.
-online processing costs \citep[][VOT]{olson2013bilingual,tsui2019impact,vsimavckova2015immediate}, language mode \citep[][vowel quality]{simonet2014phonetic}, context-awareness \citep[][phonological variables]{khattab2013phonetic}.
-What triggers this interaction?
-\cite{olson2016role} argues that while cross-language phonetic effects are largely measured at the point of language switch, it could have two potential sources:
-\begin{enumerate}
-	\item The local point of switch itself
-	\item Global co-activation of two languages -- \textsc{bilingual language mode} \citep{grosjean1998studying} 
-\end{enumerate}
-
-A number of studies have specifically manipulated language mode, both in the presence and absence of switching.
-Overall, results suggest that: (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep[][vowel quality]{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use--- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \citep[][VOT]{olson2016role,tsui2019impact} or spontaneous conversation \citep[][VOT]{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
-(iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \cite[][VOT]{olson2016role} found no additive effects of language mode, \cite[][VOT]{olson2013bilingual} found a balanced language context to inhibit transfer compared to unbalanced contexts.
-Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days (\cite{elias2017effects}; vowel quality), \citep[][VOT]{schwartz2015language, bullock2009trying,antoniou2011inter, vsimavckova2015immediate,vsimavckova2018patterns}.
-
-\cite{grosjean1998studying} suggests that various aspects of the communicative setting, including exposure to (spoken or written) stimuli in multiple languages and awareness of the interlocutor's being bilingual, could trigger a bilingual language mode.
-The participants in the present study are in an environment which largely contains mixed-language input, multilingual interlocutors, and no stable ambient language.
-Therefore we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode.
-To preserve ecological validity, we elicited both unilingual and switched utterances in a bilingual language mode.
-Any observed differences in this paradigm would arguably result from interaction during online processing of sounds while switching between languages.
-
-In a bilingual mode, both language systems are expected to be nearly equally accessible throughout the test block.
-Thus, a consistent difference between switched and non-switched tokens in such a paradigm would be possible only if the effects were highly localized-- if not, we should expect a gradual convergence over the course of the experiment.
-This is discussed in the next section.
-
-\subsection{Duration of transfer effects} \label{duration}
-
-Research comparing bilingual speakers' phonological systems with monolinguals treats them as relatively stable over time.
-The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods \citep[e.g.][]{guion2003vowel, caramazza1973acquisition}.
-However, longitudinal studies %(citations)
-and between-subject comparisons of bilingual speakers with different durations of L2 exposure suggest that interaction between the sound systems is dynamic: over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts.
-For example, \cite[][vowel quality]{bohn1992production}
-observed that approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language influence.
-These have been interpreted as changes to long-term phonological representations.
-
-Moreover, changes due to transfer are not necessarily unidirectional or irreversible.
-\cite[][VOT]{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language.
-This not only evidenced that cross-language interaction can be triggered in an order of months, but also that the effects can be reversed within a similar time range.
-A more recent study by \cite[][VOT]{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks).
-Based on a short-term longitudinal study, \cite[][VOT]{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2.
-For some sounds and features (though not others, cf.
-sec.\ref{asymmetry between sounds}), transfer was additive over time.
-How long these effects last in the absence of regular L2 input was not tested.
-% if effects are cumulative, then the difference between switched and non-switched tokes should be lesser in later test blocks (previous two paragraphs needed?)
-
-The research discussed above concerns situations where a speaker is operating in any one of their languages.
-When it comes to the phonetics of mixed-language use, the majority of existing studies only analyze the switched (target) token.
-Thus, there are few direct measurements of the duration of short-term transfer effects.
-One study which measured this in a code-switching paradigm \citep[][VOT]{bullock2009trying} did not find any residual effects on the matrix language following a switch, suggesting that transfer during code-switching is localized.
-Indirect evidence in support of this comes from experiments which have elicited switched and non-switched tokens in the same test block and still reported differences between the two \citep[e.g.][VOT]{tsui2019impact,olson2013bilingual}, suggesting that changes due to transfer are quickly `reset'--- in an order of seconds.\\
-Note that all the studies discussed above measure VOT, which is a temporal feature.
-We have no apriori reason to assume that these durations generalize to vowel quality.
-However, findings from sub-categorical phonetic shifts triggered by other factors (such as convergence towards an interlocutor) do evidence rapid shifts in vowel quality within comparable time-frames \citep[e.g.][]{pardo2010expressing,babel2010dialect,babel2012evidence}.
-Thus in the current study, we present switched and non-switched tokens randomly within the same test block to induce a bilingual mode.
-We expect the intervening words between two subsequent targets to undo any residual effects of transfer.
-
-\alert{discussion: found that like VOT, vowel quality shifts due to switching are also highly localized, and reset in an order of seconds.
-This behavior is similar to other context-driven shifts in the literature}
-
-\subsection{Asymmetries between languages in extent and direction of transfer} \label{asymmetries}
-A recurring pattern that emerges from existing research on short-term phonetic transfer is that the extent and patterns of phonetic shift are not equivalent across the two languages of the bilingual speaker.
-Studies vary greatly in the nature of differences that they report.
-The potential causes for such asymmetry are of interest because they point towards the factors that mediate cross-language phonetic interaction.
-In this section, we discuss two proposed sources of such asymmetry that are relevant to the present study:
-
-\subsubsection*{L1 vs. L2 status} 
-Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} posits that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another.
-Thus, the patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories.
-In non-switched production, this is evidenced through a pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf.
-\textsc{equivalence classification}-- \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.
-
-In mixed-language production, the role of language status is less clear-- studies have reported unidirectional influence of L1 on L2 \citep[][VOT]{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \citep[][VOT]{tsui2019impact, olson2013bilingual}, (\cite{elias2017effects}, vowel quality) bidirectional convergence \citep[][VOT]{bullock2009trying, olson2016role}, divergence \citep[][VOT]{bullock2009trying,vsimavckova2018patterns}, and no influence \citep[][vowel quality]{muldner2019phonetics}, \citep[][phonological process]{schwartz2015language}.
-Since most existing studies measure shifts in VOT, the observed asymmetries between languages have been variously explained either in terms of their L1 vs L2 status, or language-specific differences between long- and short-lag VOT languages.
-%(cf sec.
-\ref{sound systems}).
-
-\cite{olson2013bilingual} first reported a unidirectional VOT shift of L1 towards L2 in two different groups -- native speakers of Spanish (short-lag) and English (long-lag), matched for proficiency and age of L2 acquisition.
-This established that beyond language-specific differences, the L1 vs L2 status of the language does mediate transfer.
-They interpret this asymmetry in terms of the Inhibitory Control Model (ICM) \citep{green1998mental}: to select a phonetic realization from one language, the other must be \textsc{inhibited}.
-Since L1 is the `stronger' language, the amount of inhibition required on L1 while using an L2 is greater than the inhibition required on an L2 while using the L1.
-This greater initial inhibition on L1 means that switching into L1 incurs a greater \textsc{switch cost} (more cross-language influence) than switching into L2.
-Lower switch cost results in a lack of visible cross-language transfer effects on L2.
-\cite{tsui2019impact} report comparable results, but only in participants who were not equally dominant in both languages.
-Balanced bilinguals did not demonstrate any transfer effects in VOT.
-They propose that this is because balanced bilinguals have better inhibitory control (low switch cost in both languages) due to greater experience with language switching.
-
-Since differences between long-lag and short-lag languages independently affect the phonetic realization of VOT, focusing on other phonetic features as sites for transfer can avoid this conflation, and clarify the precise effect of language status.
-However, given the implications of the ICM, it is necessary to first establish that these L2 categories can indeed be affected by dynamic interference, particularly in bilinguals with extensive language-switching experience.
-The few existing studies on vowel quality \citep{simonet2014phonetic,muldner2019phonetics,elias2017effects} or phonological processes \citep{simonet2020increased,schwartz2015language} have examined transfer effects on L1 categories.
-As far as we are aware, there is no work yet on short-term transfer in L2 vowels.
-Thus, this study builds on the existing work by examining whether the L2 vowel quality of proficient bilinguals can be affected by mixed-language processing.
-
-\alert{discussion: more about ICM, assumptions about processing}
-
-\subsubsection*{Differences between sound systems of the languages} \label{sound systems} In addition to the cognitive factors discussed above, many researchers have attributed the observed asymmetries to language-internal factors.
-Specifically, a pattern that emerges from the body of work on VOT is that a shift in long-lag VOT towards the short-lag norms of the other language is much more consistent and systematic than the converse \citep{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid}.
-In general, short-lag VOTs seem to resist accommodative shift.
-\cite{bullock2009trying} suggest that because long-lag languages offer a greater range of acceptable VOTs, there is more `room' for movement.
-In contrast, shift in short-lag languages could risk the loss of a phonological contrast (VOT being the primary cue for voicing contrast in these languages).
-This is a language-specific phonological constraint on transfer.
-In a similar vein, \cite{antoniou2011inter} observe that while the VOT of English stops shifted towards Greek (a short-lag language), the reverse was seen only in a specific subset of sounds-- word-medial non-nasalized stops.
-They attribute this to the low frequency of these sounds in Greek, making them less ``stable", and thus more amenable to cross-language influence.
-This suggests that the distribution of sounds in a language could affect the kind of transfer patterns observed.
-
-These results highlight that beyond interaction during the cognitive processing of sounds, phonetic transfer is ultimately a linguistic phenomenon, and thus subject to language-specific phonological constraints.
-Therefore, to make meaningful generalizations, it is imperative to consider data from a variety of language pairs.
-Existing research on phonetic transfer largely revolves around a limited set of phonologically related languages.
-Thus, the present study extends the scope of this research to a new pair of languages--- Indian English and Bengali.
-
-Since Bengali has a four-way laryngeal contrast, VOT does not encode a binary voiced-voiceless distinction.
-Moreover, VOT is not the only relevant cue for the corresponding phonological categories in the language \citep{dmitrieva2020acoustic}.
-This means that any effect of L1 Bengali on L2 English VOT is likely to be affected by additional phonological factors.
-Thus, avoiding any potential confounds was another reason we chose to focus on vowel quality as a target for transfer in the present study.
-
-\subsection{Asymmetries between sounds}\label{asymmetry between sounds}
-
-Asymmetries in the extent and patterns of transfer have not only been observed between languages, but also between different sounds/features of a language, in both long-term and transient interactions.
-Studies that have examined multiple sound categories have found that interactions between individual sound pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep[e.g.][vowel quality]{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot treated as an atomic measure.
-In light of the discussion about VOT in section \ref{sound systems}, this is not surprising --- if a general linguistic principle of `room for movement' and contrast constrains transfer, then we should expect it to apply to individual sounds too.
-Once again, this emphasizes the importance of examining a wider range of sound contrasts.
-
-In the present study, we focus on two vowel categories in Indian English-- the mid-central vowel \nt{2} and the low front vowel \nt{\ae}. There are two plausible sources of asymmetry between these:
-\begin{enumerate}
-	\item Position in the IE vowel space: compared to \nt{\ae}, \nt{2} exists in a part of the vowel space that has a lower vowel density.
-This affords a greater latitude for movement without risking the loss of a contrast, particularly in the vowel height (F1) dimension.
-Thus, considering purely phonological constraints on IE, we expect that a greater degree of shift is possible in \nt{2} compared to \nt{\ae}.
-Any movement in \nt{\ae} is expected to be primarily in the backness (F2) dimension.
-	\item Target of transfer: findings from existing research on cross-language phonetic interaction suggest that changes in production during mixed-language use are not random, but rather targeted with respect to categories in the other language.
-Thus, another source of asymmetry is the fact that the category \nt{2} is absent in Bengali, whereas \nt{\ae} is a common category across both languages (cf.
-sec.\ref{vowel systems}).
-\end{enumerate}
-
-%table
-
-Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} posits that sound categories which are common across languages influence each other because they share a common acoustic-phonetic space.
-Thus, if Bengali and English differ in their canonical realizations of \nt{\ae}, then we should expect the English \nt{\ae} to shift towards the corresponding Bengali category in the mixed condition.
-
-%Other categories can in principle be produced with native-like accuracy, showing little effect of transfer.
-Given that our participants are highly proficient bilinguals, and assuming that cross-language correspondences remain the same in representation and processing, this would lead us to expect more transfer in the common category, \nt{\ae}.
-
-There is no obvious competing L1 category during the production of \nt{2}.
-However, it is unlikely that this should altogether preclude a shift in \nt{2}, since at least one existing study has reported transfer effects on a non-common vowel category in a comparable paradigm \citep{simonet2014phonetic}.
-Here, participants' Catalan \nt{O} shifted towards an acoustically close category \nt{o}, which is common to Catalan and Spanish, suggesting that movement might be towards a set of phonetic norms rather than a specific corresponding category.
-
-We expect the vowel \nt{2} to shift towards an acoustically/perceptually close category in IE -- a probable candidate being the low vowel \nt{a:}, as anecdotal observations of Bengali-accented English suggest that the sounds are perceived as close by Bengali speakers.
-
-In the present study, our research question concerns whether L2 vowels are produced differently in unilingual vs mixed conditions.
-We do not elicit unilingual Bengali productions.
-To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we used production data from the freely available Bengali corpus, SHRUTI \citep{shruticorpus}.
-Formant values were extracted from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, to get formant estimates from the steady state, which should most closely approximate the production target.
-Next, we compared these to our participants' productions in the unilingual English condition to understand the relative positions of the categories.
-\alert{fig.
-..
-plots-- from current plot- remove the dotted line.
-Filter by context, then remove group=context from figure code} F1-F2 plots show that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.
-
-\subsection{A note on paradigms: connected speech or not?} \label{paradigms}
-
-This section motivates a secondary aim of this study: to verify if the extent and patterns of phonetic interaction in bilingual participants are independently mediated by the choice of experimental paradigm.
-Studies of phonetic transfer during mixed-language use have largely employed spontaneous or scripted code-switching (CS) to elicit mixed productions.
-CS is defined as the use of multiple languages within a single utterance, i.e.
-in connected speech \citep{myers1993dueling}.
-Other paradigms which have been used include cued picture-naming, delayed repetition, interpreting across languages, reading word-lists etc.
-\cite{olson2013bilingual} argues that code-switching introduces discursive factors such as discourse planning and pragmatic considerations, which ultimately reflect patterns of language practice, rather than cognitive behavior.
-This finds support from studies which report transfer effects \textsc{before} the switch point \citep{bullock2009trying}, or other evidence for planning such as hyperarticulation \citep{muldner2019phonetics} and interlocutor-awareness in spontaneous CS \citep{khattab2013phonetic}.
-To avoid such confounds, \cite{olson2013bilingual} endorses cued picture-naming (where pictures are named in rapid succession in either language, based on the given language cue) as an alternative paradigm to isolate purely phonetic effects in controlled setups.
-This observation has important implications for how the existing literature on short-term phonetic interaction is interpreted: if the discursive properties of a CS paradigm independently affect the phonetic outcome of mixed-language use, this calls to question the comparability of results across studies using these different paradigms.
-Because the reported outcomes of short-term phonetic interaction vary so greatly across studies (cf.
-section \ref{asymmetries}), generalizations about the mechanisms that underlie these patterns often rely crucially on the assumption of such comparability.
-Therefore, identifying paradigm-specific differences is particularly relevant for this line of research.
-
-Note that while there is evidence for planning in CS, precisely how this affects phonetic interaction cannot be predicted from the existing literature--- while \cite{bullock2009trying} found tokens immediately preceding the switch point to be phonetically distinct from the rest, one participant group converged towards the switch language (explained in terms of proficiency) while the other showed divergence (explained in terms of extra-linguistic factors).
-The participants in Muldner et al.'s \citeyearpar{muldner2019phonetics} study hyperarticulated switched tokens (indicating some degree of planning), and did not evidence a significant shift during mixed-language use.
-Moreover, recall that the findings of \cite{olson2016role} suggest that there is a limit on the extent of transfer--- multiple factors do not necessarily result in additive effects.
-Thus, the presence of additional factors in CS does not necessitate that this should alter the extent of observable transfer compared to any other paradigm.
-It is possible that the outcomes in the studies discussed above are better attributed to some other aspect of the experimental setup, rather than to the act of code-switching in particular.
-
-The precise contribution of various (linguistic and extra-linguistic) factors and their interactions to the phonetics mixed-language use merits detailed investigation in its own right.
-In the present study, however, our focus is limited to verifying whether, in an experimental setting, the postulated discursive elements of a CS paradigm alter the extent of observable transfer compared to a paradigm that arguably lacks these--- cued picture-naming.
-We are less concerned with the exact nature of differences than with their comparability.
-While researchers have successfully employed each of these paradigms to induce phonetic interaction, comparison across studies is difficult since they also differ in other factors.
-We provide a starting point for such comparison by eliciting productions of the same tokens from the same set of participants, in both paradigms.
-A difference in results will suggest that the planning involved in connected speech and code-switching affects transfer above and beyond the phonetic outcome of switching itself, which has implications that future work should consider.
-
-\subsection{Research questions and hypotheses}\label{questions_and_hypotheses}
-\begin{itemize}
-	\item \textbf{L2 vowels}: Are L2 vowel categories of proficient bilingual speakers affected by dynamic phonetic transfer during mixed-language production, as evidenced by differences in the spectral properties of L2 vowels in switched and non-switched tokens?
-	\item \textbf{Asymmetry}: If so, we expect an asymmetry between the vowels in the extent and direction of shift.
-Based on comparisons with Bengali vowels from \cite{dutta2021} we expect \nt{2} to be lowered and fronted in the mixed condition.
-\nt{\ae} is expected to show lowering, but to a lesser extent than \nt{2}, as it is in a part of the IE vowel space that has greater vowel density.
-	\item \textbf{Paradigm}: Is there an observable difference between the extent of transfer in a cued picture-naming vs. a code-switching paradigm? 
-\end{itemize}
-
-\section{Methodology} \label{methodology}
-The experiment had the following factorial design: 10 participants* 20 target words* 2 conditions* 2 tasks* 4 iterations = 3200 target tokens.
-
-\subsection{Participants} \label{participants}
-
-10 Bengali-English bilingual speakers (male=5, age range 19 to 28), completed the tasks.
-All the participants speak other languages in addition to Bengali and English.
-At the time of recording, they were living in or around the EFL University campus in Hyderabad, which has a linguistically diverse population.
-Speakers were selected on the basis of responses to a Language Background Questionnaire to ensure comparable LSRW (listening, speaking, reading, writing) skills in L1 and L2 across participants, and were compensated for their time.
-
-\subsection{Stimuli}
-Twenty monosyllabic English words containing the vowels \nt{2} or \nt{\ae} were selected as target words.
-To minimize coarticulatory effects on the vowel, the pre-vocalic consonants were all voiced plosives (\nt{b} or \nt{d}).
-10 monosyllabic English words that do not contain the target vowel served as filler items.
-We avoided English words that are lexicalized loanwords in Bengali.
-%The list of words can be found in sec.\ref{appendix}.
-
-\subsection{Procedure}
-Participants were recorded individually in a sound attenuated room. %, using a ... microphone type/description?
-Productions were recorded in Praat \citep{boersma2016praat} at a 44.1K sampling rate.
-Every participant performed four iterations each of two tasks-- cued picture-naming and code-switching.
-Utterances in the two conditions (unilingual and mixed) were randomly interspersed within each iteration.
-In the unilingual condition, the target word was preceded by another English word in the utterance.
-In the mixed condition, the target was preceded by a Bengali word and thus produced as a switch from Bengali.
-Every target word appeared twice during a test block; once in a unilingual and once in a mixed condition.
-Participants alternated between the two tasks, and were allowed to take a short break after each test block.
-The tasks are described below:
-
-\begin{enumerate}[]
-	\item Cued picture-naming: Participants were presented with slides in the following sequence:
-	\begin{itemize}
-		\item Language cue- a word in either English or Bengali orthography
-		\item A picture displayed for 50ms; named by the participant in either English or Bengali, depending on the language cue 
-		\item Target -- an English word; read out by the participant
-		\item Distracter math problem
-	\end{itemize}
-	Since all the target words were in English, the language cue was used to manipulate the \textsc{preceding} word, giving target words in non-switched (unilingual) and switched (mixed) utterances.
-The orthographies of English and Bengali are visually distinct, and were therefore used to cue language.
-
-	\item Code-switching: Each slide contained one target word embedded in either an English (unilingual) or a Bengali (mixed; target produced as a switch from Bengali) carrier sentence which was read out by the participant.
-The carrier sentences are given below:
-	\begin{itemize}
-		\item Unilingual sentence: That is a yellow [\textit{target word}].
-		\item Mixed-language sentence: \textipa{o-\:ta \ae k-\:ta kalo} [\textit{target word}]. 
-	\end{itemize}
-
-Gloss and translation for Bengali sentence:
-
-o-\textipa{\:t}a \qquad \textipa{\ae}k-\textipa{\:t}a \quad kalo [\textit{target word}]\\
-that-\textsc{clf}  one-\textsc{clf}  black [\textit{target word}]\\
-``that is a black [\textit{target word}]"\\
-
-The sound preceding the target word was uniform across all sentences (the final sound in \textit{yellow} is produced as [o] in Indian English), to minimize any possible differences due to coarticulatory effects on the target vowel.
-A specifier-noun construction was used to ensure identical word order in both languages, giving a uniform prosodic context for the target word across condition.
-In presenting the mixed-language utterances, we used mixed orthography.
-\\
-
-\end{enumerate}
-\subsection{Acoustic analysis}\label{analysis}
-The first iteration of each task was treated as a trial, and excluded from further analyses.
-An additional 67 tokens (2.79\%) were excluded due to errors in presentation/production, giving 2333 target word tokens (\nt{2}: 1137, \nt{\ae}: 1196) for analysis.
-Recordings were manually segmented and annotated in Praat \citep{boersma2016praat}, and the first and second formant frequencies (F1 and F2) of the target vowels \nt{2, \ae} were measured at five points-- in 10\% increments starting at 5\% into the vowel.
-This gave measures of F1 and F2 at 5\%, 15\%, 25\%, 35\% and 45\% into the vowel, allowing us to observe any dynamic changes in the extent of transfer over the course of producing the vowel.
-Individual vocal tract differences affect formant frequencies across speakers.
-To avoid this confound, we used Lobanov normalization (using the PhonR package \citep{phonR} in R \citep{r}, representing each speaker's formant frequencies as the scaled distance from their own mean (\alert{citation for lobanov normalization?}).
-Plotting the vowels \alert{(figure)} shows a difference between male and female speakers, such that male speakers have overall lower formant frequencies.
-Speaker normalization using the Lobanov method does not get rid of this difference.
-Since gender-related differences in formant frequencies are not of interest in this study, we normalized individual values by gender (representing each value as a scaled distance from the group mean).
-Statistical analyses were performed on these normalized values.
-
-\section{Results} \label{results}
-
-All statistical analyses were performed in R \citep{r}.
-We used linear mixed effects models to test the effect of experimental conditions on vowel height (F1) and backness (F2) respectively.
-Since formants were extracted from five points over the course of articulating the vowel, time was coded numerically (1-5), and treated as an independent variable.
-This was scaled and centered prior to analysis and treated as a continuous predictor.
-We fitted models using the \emph{lme4} package \citep{lme4}, and used \emph{lmerTest} \citep{lmerTest} to obtain estimates of p values for individual variables.
-For all mixed models, the alpha criterion was set at $|t| > 2 $.
-The dependent variables were normalized F1 and F2, and we fitted separate models for these.
-The independent variables or fixed factors, along with their levels, are given in table \ref{table variables} \alert{add variable type with each: factor, numerical, ...}.
-Subject and Item were treated as random factors.\\
-
-< Insert Table \ref{table variables} about here > \\
-
-\subsection{Models}
-
-\subsubsection*{Fixed effects}
-
-Our study involved two IE vowels. Vowel category is expected to systematically affect formant frequencies (FFs).
-
-The experimental treatment of interest is Context, i.e.
-whether the target item was presented in a unilingual (English) or mixed (Bengali) context.
-Our main research question (\bf{L2 vowels}) asks whether L2 vowel quality (formant frequencies) of proficient bilinguals is affected by the Context of utterance.
-We hypothesize that the direction of shift in the mixed condition is towards the corresponding L1 category.
-Thus, we expect the effect of Context on formant frequencies, i.e.
-the direction of shift, to be affected by the vowel category (c.f.
-section \ref{asymmetry between sounds}).
-Thus, our main effect of interest is the interaction term Context*Vowel.
-This corresponds to the transfer pattern.
-
-Our third research question (\bf{Paradigm}) asks whether the language-mixing paradigm, or Task, affects the extent and direction of transfer.
-If this is the case, we expect Task to interact with the transfer term, giving a three way Task*Context*Vowel interaction.
-
-Since we measured FFs at five points in the vowel, Time is expected to affect the FF.
-However, the precise effect of formant dynamics is expected to depend on the vowel category.
-Thus, we expect the interaction of Vowel*Time to predict formant frequency.
-If transfer patterns change across time, then we should expect it to interact with the transfer term, giving a Context*Vowel*Time interaction.
-
-\subsubsection*{Random effects structure}
-
-Subject and Item were treated as random factors.
-Given our hypotheses, we included random intercepts for Subject and Item, by-subject random slopes for Vowel (formant targets for vowel categories differ across individuals) and Context (individuals differ in their response to the experimental condition), and a by-item random slope for Context (the effect of experimental conditions differ across words).
-
-We tested the significance of our hypothesized explanatory variables using t-values and p-values for individual variables in the models, and tested overall model fit by comparing the full model to a corresponding null model that lacks the variable of interest using ANOVAs.
-The following subsections summarize and discuss the model outputs for F1 and F2 respectively.
-
-\subsection{Vowel Height -- F1}
-
-\subsubsection*{Model output summary}
-
-First, to test hypothesis 1 (\bf{L2 transfer}), we fitted a model with vowel*time, and vowel*context as fixed effects.
-The transfer term (vowel*context) is significant in this model (model coefficient/ unstandardized effect size ($\beta$)= -2.330e-01, standard error (SE)=  5.244e-02, t= -4.44, p=0.0002).
-Comparing this to a null model that lacks the transfer term shows that the full model is significantly better ($\chi^2(2) = 14.42, p=  0.0007$).
-To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction.
-The full model was better ($\chi^2(1) = 12.51, p= 0.0004$).
-
-To test the hypothesis \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects.
-This three-way interaction term is significant ($\beta$= -1.955e-01, SE=  5.453e-02, t= -3.59, p= 0.0003).
-We also find a significant effect of task on FFs ($\beta$=3.300e-01, SE=2.668e-02, t=12.371, p=2e-16).
-Comparing this model to null models that lack the fixed effect Task ($\chi^2(4) = 572.61, p=  2.2e-16$) shows that the full model is better.
-Because Task has an independent effect on F1, and adding interactions increases model parameters, we fitted a second null model with Task as a separate fixed effect without the interaction term to test whether the interaction term is needed.
-However, the full model was still a better fit for the data ($\chi^2(3) = 23.82, p=  2.722e-05$).
-Next, we tested two other null models to verify that the full model with the three-way interaction term is indeed the optimal model for the data, by replacing the three-way interaction term with: (i) vowel*context + task*context (to test if task only interacts with context, rather than the transfer term vowel*context).
-The full model is better ($\chi^2(2) = 23.28, p= 8.78e-06$); (ii) vowel*context + task*vowel (to test if task only interacts with vowel, rather than the transfer term vowel*context).
-The full model is better ($\chi^2(2) = 13.36, p= 0.001$).
-This shows that the three-way interaction is the best fit for the data.
-Because an independent effect of task on FFs is unexpected (c.f.
-), we wanted to see if the effect might be local, i.e.
-decrease over time.
-We added a task*time interaction term to the full model, which is significant ($\beta$= -1.216e-01, SE=  1.347e-02, t= -9.028, p= 2e-16 ) and improves model fit ($\chi^2(1) = 81.22, p= 2.2e-16$).
-
-Next, to examine the dynamics of transfer, we added a three-way interaction between the transfer term (vowel*context) and time.
-This term has a marginally significant effect ($\beta$= -4.604e-02, SE=  2.692e-02, t= -1.71, p= 0.08), but does not significantly improve model fit compared to a null model lacking the vowel*context*time term ($\chi^2(2) = 3.41, p= 0.18$).
-
-Thus, the optimal model was: f1 ~ time*vowel + vowel*context*task + time*task + (1 + context|word) + (1 + context + vowel|subject).
-Table \ref{table_f1} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model.
-\alert{p-vales smaller than 0.05 are in bold.}
-
-\subsubsection*{Interpretation}
-We extracted formant estimates from five uniformly spaced points in the vowel, from the time at 5\% into the vowel (at the start of articulation) to 45\% (at the mid-part, or steady state, which is expected to correspond to the articulatory target for the category).
-As expected, we see that F1 values are significantly predicted by Time in the model, reflecting th dynamic nature of articulation which causes acoustic variation over the course of time.
-The positive slope of the estimate for Time indicates that F1 increases, i.e.
-the two categories in the study show lowering over the course of articulation.
-Moreover, the significant Time*Vowel interaction shows that the F1 trajectory is different for the two vowels.
-This is expected, as the categories are expected to present different articulatory targets.
-
- Vowel is not a significant predictor of F1.
-This appears surprising, given that the phonological category is expected to systematically affect vowel quality.
-However, a boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two categories \nt{2, \ae} and the extent of height difference between them.
-Note that these measures are only from the English (unilingual) condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching.
-Given that estimates for fixed effects are pooled across participants, this explains the lack of reliable differences between \nt{2} and \nt{\ae}.
-The negative slope of the estimate for Context shows that F1 is lower, i.e.
-vowels are higher, in the unilingual (e) condition compared to the mixed (b) condition.
-Put more intuitively, vowels are lowered in the mixed condition.
-However, the lack significance indicates that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels.
-Instead, the significant interaction between Vowel and Context signals an asymmetrical shift in the categories \nt{2} and \nt{\ae} in the mixed-language condition.
-Specifically, the Vowel*Context interaction has a negative slope, showing that the lowering is greater in \nt{2}.
-
-The results also reveal a significant effect of Task, such that the sentence task (code-switching) leads to higher F1 values in both vowels relative to the cued picture-naming task as: vowels are lowered in the code-switching task, independent of experimental condition.
-This is surprising, as we have no aproiri reason to expect a uniform difference in vowel quality across tasks.
-A significant interaction with Time with a positive slope of the estimate shows that this effect is local: it is strongest at the beginning of the vowel, and reduces over the course of the vowel.
-This is consistent with our assumption that the unexpected Task effect is an artifact of the paradigm, possibly resulting from coarticulatory effects with the preceding sound in the carrier phrase.
-Of greater interest to us is the significant Vowel*Context*Task interaction, which indicates a difference in transfer patterns across paradigms.
-Specifically, the negative slope of the estimate suggests that transfer in the F1 (height) dimension are greater in the code-switching (s) task compared to the picture-naming (p) task.\\
-
-< Insert Figure \ref{boxplot_F1} about here >\\
-
-\subsection{Vowel backness -- F2}
-
-\subsubsection*{Model output summary}
-
-First, to test hypothesis 1 (\bf{L2 transfer}), we fitted a model with vowel*time, and vowel*context as fixed effects.
-The transfer term (vowel*context) is significant in this model (standardized effect size ($\beta$)= 1.234e-01, standardized error (SE)=  2.055e-02, t= 6.00, p=5.46e-06).
-Comparing this to a null model that lacks the transfer term shows that the full model is significantly better ($\chi^2(2) = 16.20, p= 0.0003$).
-To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction.
-The full model was better ($\chi^2(1) = 16.02, p= 6.245e-05$).
-
-To test the hypothesis \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects.
-While task has an independent effect on F2 ($\beta$= -8.758e-02, SE= 1.756e-02, t= -4.98, p= 6.18e-07), the three-way interaction term is not significant ($\beta$= 3.349e-02, SE= 3.565e-02, t= 0.94, p= 0.34).
-However, there are significant two-way interactions between task*context ($\beta$= -1.099e-01, SE= 2.483e-02, t= -4.427, p= 9.66e-06) and task*vowel ($\beta$= -8.983e-02, SE= 2.535e-02, t= -3.544, p= 0.000396).
-Comparing this to a null model that lacks the fixed effect Task confirms that the full model is still a better fit for the data ($\chi^2(4) = 572.61, p=  2.2e-16$), indicating that an optimal model should contain the term Task.
-To confirm whether Task interacts with any of the other terms, we fitted a model with Task as a separate fixed effect without the interaction term.
-Comparing this to the full (interaction) model shows that the latter is still a better fit ($\chi^2(3) = 44.8, p= 1.02e-09$).
-Next, we tested two other null models to verify that the three-way interaction term is indeed optimal for the data, by replacing the three-way interaction term with: (i) vowel*context + task*context (to test if task only interacts with context, rather than the transfer term vowel*context).
-The full model is better ($\chi^2(2) = 17.28, p= 0.0001$); (ii) vowel*context + task*vowel (to test if task only interacts with vowel, rather than the transfer term vowel*context).
-The full model is better ($\chi^2(2) = 28.45, p= 6.633e-07$).
-This shows that the three-way interaction is the best fit for the data.
-
-Because an independent effect of task on FFs is unexpected \alert{(c.f.
-next section)}, we wanted to see if the effect might be local, i.e.
-decrease over time.
-We added a task*time interaction term to the full model, which is significant ($\beta$= 6.237e-02, SE=  8.876e-03, t= 7.027, p= 2.23e-12 ) and improves model fit ($\chi^2(1) = 49.271, p= 2.23e-12$).
-
-Next, to examine the dynamics of transfer, we added a three-way interaction between the transfer term (vowel*context) and time.
-This term is not significant ($\beta$= -6.274e-03, SE= 1.774e-02, t= -0.35, p= 0.72), and the addition does not significantly improve model fit compared to a null model lacking the vowel*context*time term ($\chi^2(2) = 2.2443, p= 0.32$).
-
-Thus, the optimal model is: vowel*time + vowel*context*task + task*time + (1+context|word) + (1+context+vowel|subject).
-Table \ref{table_f2} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model.
-\alert{p-vales smaller than 0.05 are in bold.}
-
-Statistical analyses reveal that the vowel \nt{2} is lowered and fronted, while \nt{\ae} is fronted, in the mixed condition.
-Previous literature on phonetic transfer suggests that such shifts are not random, but rather directional and targeted.
-We want to understand how the direction of shift compares to corresponding Bengali categories.
-Figure \ref{vowels_e_b} summarizes the findings of the experiment ($F1XF2$ plots for vowel productions in the unilingual and mixed conditions, in comparison to the related Bengali categories from \cite{dutta2021}).
-This confirms our hypotheses (c.f.
-section \ref{asymmetry between sounds}) that \nt{\ae} moves towards the corresponding \nt{\ae} category in Bengali, whereas the shift in \nt{2} is towards the related Bengali category \nt{a:}.
-The following sections discuss the results for vowel height and backness respectively.
-
-< Insert Figure \ref{vowels_e_b} about here >
-
-\subsection*{Interpretation}
-
-As with F1, we find that Time significantly influences F2 values in the model, reflecting the dynamic nature of articulation which causes acoustic variation over the course of time.
-Moreover, the Time*Vowel interaction shows that this trajectory differs across the two vowels: while the positive slope for Time indicates that F2 increases, i.e.
-vowels are fronted with time, the negative slope of the Time*Vowel interaction shows that this effect is smaller for \nt{2}.
-This is expected, as figure \alert{ref} confirms that the category \nt{\ae} is overall more fronted than \nt{2}.
-
-Unlike in the F1 dimension, Vowel is a significant predictor of F2, with \nt{2} having significantly lower F2, i.e. produced as more backed, than \nt{\ae}.
-
-As with vowel height, there is no significant independent effect of Context
-
-The estimates for Vowel show that excepting the beginning of the vowel, the category \nt{2} is consistently produced as more backed than \nt{\ae}.
-Unlike vowel height, Context has an independent effect on vowel backness, such that both the vowels in this study have higher F2 values in the Bengali (mixed) condition, i.e.
-they are fronted in the mixed condition.
-A significant interaction between Vowel and Context reveals an asymmetry in the extent of fronting between the two vowels -- the shift is greater in the case of \nt{\ae}.
-We see a significant effect of Task on F2, such that vowels in the Sentence task (code-switching) were more backed than those in the picture-naming task.
-As with F1, this is an unexpected result, as we have no principled reason to expect code-switching to induce an across-board backing of vowels.
-A lack of significant interaction with Context suggests that there is no difference in the extent of shift between the two paradigms.
-The results of ANOVAs comparing the full model to a null model that lacks the Vowel*Context interaction factor reveal that as with vowel height, the change in vowel backness between the two conditions is not significant at 5\%, and increases over the course of the vowel.\\
-
-< Insert Table \ref{table_f2_fixed_effects} about here >
-
-\subsection{Power analysis}
-We were able to recruit 10 participants (c.f.
-section \ref{participants}).
-To confirm whether the experiment design was powerful enough to detect the effects we are investigating, we carried out post-hoc power analyses using the mixedpower package \citep{kumle2021estimating} in R.
+The aim of the analysis was to explore the power of our experiment design to detect the effects we are investigating.
+The package simulates responses based on a given dataset and runs the specified analyses on these new data, estimating power by calculating how frequently statistical significance is obtained in the simulated data.
 The target of the analysis was to calculate the power for detecting an effect for our main variables of interest: vowel*context (to test the hypotheses L2-vowels and Asymmetry), and vowel*context*task (to test the hypothesis Paradigm).
@@
 Thus, a key issue in power analyses is to estimate the expected effect size for the treatment/factor of interest \citep{brysbaert2018power, kumle2021estimating}.
-This may be expressed as a standardized effect size or estimate ($\beta$), Cohen's d, or partial $\eta ^2$.
-
-Using observed effect sizes from the data for power analyses is a bad practice and has been shown to be meaningless \citep{hoenig2001abuse, gelman2019observed_effect_size}, as such "observed power" depends on the p-value.
+This may be expressed by a variety of measures, including Cohen's d, $\eta ^2$, and model estimates ($\beta$).
+Using observed effect sizes from the data for power analyses has been shown to be meaningless \citep{hoenig2001abuse, gelman2019observed_effect_size}, since such ``observed power" is a function of the p-value.
 Therefore, we need a reasonable estimate of effect size.
@@
 An alternative approach is to define a smallest effect size that is meaningful to us (smallest effect size of interest; SESOI \citep{kumle2021estimating}), and test whether the study has enough power to detect such an effect, if it were to exist.
-This needs to be theoretically motivated.
-To the best of our knowledge, there is no existing literature on typical effect sizes for spectral differences in cross-language vowel production, and not many studies that report model estimates.
-Only more discussion among researchers in this field will lead to a consensus on what is considered a typical or minimally meaningful cross-language effect.
+This value needs to be theoretically motivated.
+To the best of our knowledge, there is no existing literature on typical effect sizes for spectral differences in cross-language vowel production, and not many studies that report effect sizes.
+Only more discussion in this field will lead to a consensus on what is considered a typical or minimally meaningful cross-language effect.
+Similarly, while differences between language-switching paradigms have been informally discussed in previous literature, we are not aware of any study that has measured these differences in a controlled experimental setup, and thus no reported effect sizes for between-paradigm differences in phonetic transfer to act as a guideline.
 
-Similarly, while differences between language-switching paradigms have been informally discussed in previous literature, we are not aware of any study that has measured these differences in a controlled experimental setup, and thus no reported effect sizes to act as a guideline.
-
-Cohen (cite) arbitrarily classifies effect sizes (cohen's d values) as small, medium, and large.
-Although there ar arguments against the meaningfulness of these categories \alert{citations}, these are often used as ballpark values for power analyses in psychological studies (for a discussion, see \cite{brysbaert2018power}).
+Cohen \citeyearpar{cohen1988statistical} classifies effect sizes (Cohen's d values) as small ($\leq$0.2), medium (0.5), and large ($\geq$ 0.8).
+Although the usefulness of this classification has been debated in the literature, these are often used as ballpark values for power analyses in psychological studies (for a discussion, see \cite{brysbaert2018power}).
 However, this approach does not generalize well to mixed-effects analyses, particularly for complex models with multiple random effects and interactions, as is typical in psycholinguistic and bilingualism research.
-"Determining the SESOI for (G)LMMs is difficult in a simulation-based approach where effect sizes are indicated through the model’s unstandardized beta coefficients...relating effect sizes to beta coefficients in complex models is far from trivial and the authors therefore refrain from making specific recommendations.
+``Determining the SESOI for (G)LMMs is difficult in a simulation-based approach where effect sizes are indicated through the model’s unstandardized beta coefficients...relating effect sizes to beta coefficients in complex models is far from trivial and the authors therefore refrain from making specific recommendations.
 \citep{kumle2021estimating}".
@@
 The results of a power analysis are only as meaningful as the assumptions that go into it.
-To avoid making arbitrary assumptions, we followed the approach in \cite{brysbaert2018power} (and discussed in \cite{kumle2021estimating}) of taking effect sized of the study of interest and directly modifying all the beta coefficients by a fixed percentage.
-Note that we are not assuming that doing so gets us closer to a "true effect size".
-Instead, we run simulations with a range of effect sized and report the power of our study to plausibly detect an effect of that size.
-The results are shown in the plots in figure ...
-These show power over a range of sample sizes, given the assumed effect size.
+To avoid making arbitrary assumptions, we followed the approach in \cite{brysbaert2018power} (and discussed in \cite{kumle2021estimating}) of extracting the effect sizes from the study of interest and directly modifying all the beta coefficients by a fixed percentage.
+Note that we are not assuming that doing so gets us closer to a ``true effect size".
+Instead, we run simulations with a range of effect sizes and report the power of our study to plausibly detect an effect of that size.
+The results are shown in the plots in figures \ref{figure_power_analysis_f1} and \ref{figure_power_analysis_f2}.
+Each panel presents power analysis at a different effect size, and is based on 500 simulations.
+The plots show power (y-axis) over a range of sample sizes (x-axis), given the assumed effect size.
 
---> in cover letter: power analyses for complex mixed models are not trivial and is an active pursuit.
-While we have done our best to follow available literature, we would love to have more feedback on this.
-
-\section{Discussion}
-
-Taken as a whole, these results bear on our original research questions in the following ways: regarding our first question, whether L2 vowels of proficient bilinguals can be affected by short-term transfer from L1 (\textbf{L2 vowels}), our results confirm that this indeed happens during mixed-language use.
-Both the common category \nt{\ae} and the L2-exclusive category \nt{2} show spectral shifts during mixed-language productions compared to unilingual productions.
-As hypothesized (\textbf{Asymmetry}), the results reveal asymmetries in the extent and direction of shifts between the two vowels.
-Specifically, \nt{2} is lowered and fronted, while \nt{\ae} is fronted, with the extent of shift being greater in \nt{2}.
-Comparision with canonical Bengali vowel productions shows that the shift in \nt{\ae} is towards the corresponding Bengali category, while that in \nt{2} is towards the Bengali category \nt{a:}.
-Regarding our third research question, whether paradigm used to elicit language switching independently affects patterns of transfer (\textbf{Paradigm}), analyzing the role of Task in the results confirms this.
-We find a greater degree of transfer during code-switching compared to the cued picture-naming task.
-However, the difference is small, and needs to be replicated (?).
-These results and their implications are discussed in the following subsections.
-
-\subsection{Phonetic transfer in L2 vowels}
-
-One of the main aims of this study is to understand short-term phonetic transfer in L2, towards understanding language processing in proficient bilinguals.
-Two interrelated questions, discussed in section \ref{introduction} (....) bear on this: (i) what causes transfer in mixed-language speech; and (ii) what causes observed asymmetries between transfer in L1 vs L2.
-Together, these serve as guidelines for how we posit that the language systems of th bilingual exist in relation to one another, and consequently, what a cognitive model should look like-- any model should be able to account for these observations.
-
--- what causes transfer: language mode vs switching
-
-\alert{discussion: We presented data in the same session, such that the language mode (participants' overall language ``setting" remained the same).
-The results show differences between between the switched and non-switched tokens in spite of this.
-this confirms that the local act of switching does cause transfer, even in a bilingual language mode, paralleling results from \cite{olson2016role, tsui2019impact, piccinini2015voice}.
-All of these examined the local effect of switching on a temporal feature of speech, namely, VOT.
-The results of the present study indicate that the same happens in vowels.
-Moreover, the result of transfer is interesting-- in VOT, it can only increase or decrease, which makes it impossible to know if it is the particular consonant that becomes more like its L1 counterpart, or if there is some more system-wide shift.
---* check discussion section of these three studies and see if they have talked about this --* Because vowel quality measures show directionality as well as magnitude, we can see finer patterns.
-We find that it is not merely a shift towards a Bengali counterpart, but rather a shift towards Bengali \textit{norms}-- a category that doesn't exist in Bengali moves towards a related Bengali category.
-This result parallels \cite{simonet2014phonetic}, who found the Catalan ...
-of bilingual speakers shifts towards a related L1 Spanish category ..
-in a bilingual language mode (although, this study did not examine switched speech).
-(check discussion section once) What does this say about what happens during online processing and speech planning?
---- look at discussion section of the mode vs switching papers (those three + ones that found no local effect of switching): what do they say about this?
-Can this finding be accounted for with that?}
-
-\alert{It makes sense that a bilingual language mode makes the representations from both languages available.
-But what do these representations look like, and how are these processed during the act of using language?
-Examining speech during switching can shed light on this.
-We argue that what we see here is an example of \textit{creativity}-- beyond a simple L2-L1 mapping, there is a method to the movement such that it makes reference to entities beyond the level of the individual sound-- i.e., the system-level.
-If representations are combined in creative ways, then are we truly ``moving between" languages, in a real sense, during online processing within a single discourse?
-Can we think of intermediate representations, not of individual sounds, but rather of sound systems?
-While more detailed empirical data will eventually enable us to map out patterns of transfer during switched speech, this interpretation is in line with the kind of model proposed by \cite{khattab2013phonetic} based on observation from the use of phonological variables in bilingual children (check-- this the one where they talked about moving between systems-- spectrum of choices?).
-This visualizes/conceptualizes the bilingual language system as a \textit{continuum of sound systems}, from which the bilingual speaker is constantly choosing, guided by a variety of different pressures.
-We propose that vowels are a good feature to examine: in addition to continuous measures (as opposed to phonological variables), they also give us the 2-dimensional data (as opposed to VOT) that is suited to further testing and refining such a model.}
-
---- differences between L1 vs L2 transfer:
-
-Studies of short-term phonetic transfer consistently find differences between transfer patterns in L1 vs L2.
-The bulk of this literature is based on the feature VOT.
-While vowels have been show to induce L2-to-L1 influence, the present study demonstrates L1-to-L2 transfer.
-
-Previous studies with VOT have often found a lack of parallel transfer from L1 to L2.
-One postulated model that is consistent with this finding is the ICM, first proposed by \cite{green1986control} at the level of the lexicon, but extended to sound systems by \cite{olson2013bilingual,tsui2019impact}.
-This proposes that ...\alert{read Olson2013 discussion, more about assumptions of ICM and consequences about models of processing}  As discussed in section \ref{causes}, vowels provide a fertile testing ground for this theory in sound systems.
-
-Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.
-In the context of an ICM--based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, this is a significant finding: it demonstrates that the postulated lower switch cost on an L2 does not entirely mask transfer effects on L2 vowels.
-Moreover, our findings differ from \cite{tsui2019impact} who report on the basis of VOT measurements that balanced bilinguals do not evidence an effect of language-switching.
-\citeauthor{tsui2019impact} attribute the absence of transfer effects to greater experience with language switching in these participants, leading to lower switch-costs in both languages.
-We find that even proficient bilinguals with extensive experience of language switching demonstrate cross-language transfer effects in vowel production.
-Thus, following from the discussion in sec.
-\ref{asymmetries}, it might be fruitful for future studies to focus on vowels as a potential site for transfer.
-This would both expand the empirical scope of existing research (cf.
-sec.
-\ref{asymmetry between sounds}) and avoid confounding factors arising from independent differences between long- and short-lag VOT languages.
-More generally, these findings point to a need for expanding the range of features in which transfer effects are studied, in order to make robust generalizations about the mechanisms underlying observed patterns.
-
-\alert{this is an example where studying different features allows us to compare the predictions of different underlying process models.
-While most studies usually focus on a single feature due to ease of creating materials, there have been studies which have measured multiple features (e.g.
-..., or no need?).
-Measuring multiple features within a single design, with the same group of participants, would give a powerful tool to compare different models at little added cost to researchers.
-This would sidestep the unavoidable confounds that come with comparing results across studies, or even replication}
-
-\subsection{Asymmetries in Phonetic Transfer}
-The literature on short-term phonetic transfer has frequently noted differences in transfer patterns between different sound categories in a languages.
-This has been attributed to both general linguistic principles such as contrast preservation, and language-specific factors such as frequency and distribution of categories.
-Examining these asymmetries more closely, and understanding their sources, helps us figure out what factors affect/moderate transfer and thus, by extension, what level of representation and processing this interaction happens at, and the nature of short-term intra-speaker variation.
-The present study examined transfer in two L2 categories, \nt{2} and \nt{\ae}. We hypothesized an asymmetrical shift between the two vowels (\textbf{Asymmetry}), owing to differences in:
- \begin{itemize}
- 	\item their position in the Indian English vowel space: \nt{\ae} is in a more crowded part of the vowel space
- 	\item the presence/absence of a corresponding category in the L1 Bengali inventory
- \end{itemize}
-
-The significant interaction between Vowel and Context in both F1 and F2 in the present study confirm that the experimental condition affected both height and backness differently for each vowel.
-Examining the model coeffieicnets ($\beta$) values shows that the transfer effect is greater for ...
-compared to ...
-(\alert{check interaction effect and unpack}) a greater degree of shift in \nt{2}, possibly owing to its position in a less dense part of the IE vowel space.
-The category \nt{2} is lowered and fronted in the mixed language condition as expected (Figure \ref{vowels_e_b}).
-The direction of shift in \nt{\ae} is surprising-- it is fronted in the mixed-language condition.
-While phonological constraints on IE suggested that a shift in the F1 dimension could potentially risk the loss of contrast and is thus expected to be minimal, Figure \ref{vowels_e_b} shows that \nt{\ae} in Bengali is not fronted with respect to the English category.
-Thus, fronting in the mixed condition cannot be explained in relation the former, at least from the present data.
-\alert{remove-- Recall that the Bengali vowel productions used here are from a different set of speakers than our participants.
-Assuming some amount of individual differences in the precise phonetic realization of vowel categories in both languages, this could be a potential source of explanation for this result.} Data from a larger number of speakers for both languages, as well as within-speaker comparisons of English and Bengali vowels, are required to clarify this.
-
-\alert{The fact that nt shows more shift-- shows that this influence is not just at the level of individual sounds (that way would expect more shift in the common category because it's "confusing".
---check what \cite{simonet2014phonetic} says about this, about why shift happens.
-But this-- shows that the L1 influence is at a system-level.
-More inclined towards using a Bengali-like phonological system.
-Has implications for cognitive models of language and what happens during short-term interaction)}
-
-Switching paradigm
-Our third research question \textbf{Paradigm} concerned the effect of experimental paradigm on the phonetic outcome of mixed-language use.
-\cite{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
-If such discursive properties independently affect the phonetic outcome of language mixing in an experimental setup, this calls to question the comparability of results across studies using these different paradigms.
-We verified this by comparing productions of the same items from the same group of participants in a scripted CS paradigm and a cued picture-naming paradigm.
-We found no significant interaction of Task either with Context, or with the Vowel*Context interaction, suggesting that the extent of phonetic transfer was not affected by the paradigm.
-There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in observable transfer at all.
-This would mean that the pre-switch effects observed, for example, in \cite{muldner2019phonetics} and \cite{bullock2009trying} are better interpreted as resulting from some other aspect of the experimental setup, than the act of code-switching itself.
-While planning could additionally affect the words preceding the switch, the effects on the switched token itself do not change.
-(ii) Our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
-Thus, another possible explanation is that the discursive effects discussed by \cite{olson2013bilingual} are triggered by more naturalistic paradigms where attention to meaning and interlocutor are important.
-More within-study comparisons with less structured paradigms, and spontaneous code-switching data, can clarify this.
-The results of the present study indicate that for the purposes of observing dynamic phonetic transfer in an experimental setup, elicitation through cued-picture-naming vs.
-(controlled) code-switching does not significantly alter the behavior under study.
-Thus, the choice of paradigm can be based on other considerations.
-Moreover, this suggests that (all else being equal) the results from such studies can be compared directly.
-
-The paradigm did affect vowel quality, but this applied uniformly across experimental conditions: all vowels were more backed and lowered in the code-switching task.
-Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this lowering is an artifact of the particular stimuli used here---resulting from some corarticulatory effect that was absent in the picture-naming task.
-While this is intriguing, it is beyond the scope of the present study.
-\\
-
-\section{Conclusion}
-This study examined the phonetic outcome of mixed-language production in a group of highly proficient bilingual speakers of Bengali and Indian English.
-Specifically, we examined the spectral quality of two L2 vowel categories \nt{2} and \nt{\ae} produced in two experimental conditions: unilingual (English) and mixed (switching from Bengali to English).
-Comparisons of F1 and F2 reveal that mixed language processing temporarily increases phonetic interaction between the languages, causing \nt{2} to converge towards a related L1 category \nt{a:}.
-The extent and direction of shift is asymmetrical across the two vowel categories, which parallels findings in prior studies.
-This asymmetry appears to be mediated by the language-specific distribution of vowel categories.
-These findings evidence an effect of phonetic transfer on L2 vowels for the first time, and raise questions for proposed inhibition-based accounts of cross-language phonetic interaction.
-Mixed productions were elicited in two switching paradigms: code-switching and cued picture-naming, to verify whether the postulated discursive elements in code-switching independently influence the outcome of phonetic transfer.
-Results suggest that in an experimental setup, elicitation through cued-picture-naming vs.
-(controlled) code-switching does not significantly alter the behavior under study.
-
-This study has a number of limitations: (i) the `baseline' values for Bengali vowel categories were estimated from a different set of speakers than the participants of this study, potentially obscuring the role of individual differences in the realization of vowel categories; within-speaker comparisons of L1 and L2 vowels is likely to offer a more fine-grained view of the direction of shift in relation to L1 categories.
-(ii) we used a frame sentence to elicit code-switched productions such that not all the sentences were pragmatically well-formed.
-This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study.
-Future work should address this by considering data from more naturalistic code-switching paradigms.
-
-\section*{Abbreviations}
-\textsc{clf} = classifier
-
-\section*{Data availability}
-The materials, data files, and analysis code for this study are openly available in OSF at:  \url{https://doi.org/10.17605/OSF.IO/DSB2X}.
-
-\section*{Competing interests}
-The authors declare none.
-
-%\printbibliography %for use with biblatex; comment out if you use natbib
-\bibliography{references} %for use with natbib; comment out if you use biblatex, and change 'sample' by the name of your bib-file
+As expected, power increases as a function of sample size for all effects.
+However, of greater interest to us is the differences between the power curves in the four panels (a,b,c,d).
+These demonstrate how results of the power analysis vary as a function of assumed effect size.
+Note the the range of effect sizes used here is fairly narrow, varying between 15\% smaller and 15\% greater than the observed effect sizes.
+These differences demonstrate the importance of theoretically justified effect size estimates before attempting to make inferences based on power analyses.
+This applies equally to aposteriori power analyses done to make decisions about sample sizes prior to data collection (to reiterate, post-hoc power analyses are best avoided).
+This emphasizes the importance for bilingualism studies to report effect sizes as a part of their results, which can facilitate discussions and meta-analyses that eventually lead to such norms being established in the field.
+\footnote{Thanks to the anonymous reviewer at BLC for initiating a fruitful discussion around this.}
 
 \newpage
-\section*{Tables}
-\begin{table}[h]
-	\caption{Fixed effects}
-	%\centering
-	\resizebox{\textwidth}{!}{%
-		\begin{tabular}{@{}ll@{}}
-			\toprule
-			Variable name                                                 & Levels                                                            \\ \midrule
-			Gender                                                 & male, female                                                      \\
-			Vowel                                                  & \nt{2}, \nt{\ae} \\
-			Task (corresponding to the two paradigms)              & p (Picture: cued picture-naming, s (Sentence: code-switching)     \\
-			Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)                       \\
-			Time                                                   & continuous variable, range -1.6 to 1.6, scaled and centered\\
-			 \bottomrule
-		\end{tabular}%
-	}
-	\label{table variables}
-\end{table}
+\begin{figure} \label{figure_power_analysis_f1}
+	\centering
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f1_minus15} 
+		\caption{Effect size = observed effect size - 15\%} \label{f1_minus15}
+	\end{subfigure}
 
-\newpage
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f1_minus5} 
+		\caption{Effect size = observed effect size - 5\%} \label{f1_minus5}
+	\end{subfigure}
 
-% Please add the following required packages to your document preamble:
-% \usepackage{booktabs}
-\begin{table}[h]
+	\begin{subfigure}[t]{0.8\textwidth}
 		\centering
-	\caption{Optimal model: F1}
-	\label{table_f1}
-	\begin{tabular}{@{}lllll@{}}
-		\toprule
-		& Estimate  & Std. Error & t value & p value          \\ \midrule
-		(Intercept)           & -0.053474 & 0.178435   & -0.300  & 0.769            \\
-		time                  & 0.139140  & 0.011556   & 12.040  & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}                & -0.147480 & 0.223022   & -0.661  & 0.519            \\
-		contexte              & -0.013419 & 0.068357   & -0.196  & 0.846            \\
-		tasks                 & 0.330034  & 0.026586   & 12.414  & \textbf{< 0.001 ***} \\
-		time:vowel\textipa{2}           & -0.170892 & 0.013469   & -12.688 & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte       & -0.137062 & 0.060884   & -2.251  & \textbf{0.031 *}   \\
-		vowel\textipa{2}:tasks          & 0.009002  & 0.038498   & 0.234   & 0.815            \\
-		contexte:tasks        & 0.073341  & 0.037597   & 1.951   & 0.051 .          \\
-		time:tasks            & -0.121604 & 0.013469   & -9.028  & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte:tasks & -0.195613 & 0.054338   & -3.600  & \textbf{< 0.001 ***} \\ \bottomrule
-	\end{tabular}
-\end{table}
+		\includegraphics[width=\textwidth]{powerplot_f1_plus5} 
+		\caption{Effect size = observed effect size + 5\%} \label{f1_plus5}
+	\end{subfigure}
 
-\newpage
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f1_plus15} 
+		\caption{Effect size = observed effect size + 15\%} \label{f1_plus15}
+	\end{subfigure}
 
-% Please add the following required packages to your document preamble:
-% \usepackage{booktabs}
-\begin{table}[h]
+	\caption{Power analysis for fixed effects of interest at varying sample sizes: F1}
+
+\end{figure}
+
+\begin{figure}[!htb] \label{figure_power_analysis_f2}
 	\centering
-	\caption{Optimal model: F2}
-	\label{table_f2}
-	\begin{tabular}{@{}lllll@{}}
-		\toprule
-		& Estimate & Std. Error & t value & p value          \\ \midrule
-		(Intercept)           & 0.23202  & 0.22100    & 1.05    & 0.304            \\
-		vowel\textipa{2}                & -0.38596 & 0.14858    & -2.60   & \textbf{0.015 *}   \\
-		time                  & 0.04287  & 0.00761    & 5.63    & \textbf{< 0.001 ***} \\
-		contexte              & -0.01720 & 0.02823    & -0.61   & 0.548            \\
-		tasks                 & -0.08758 & 0.01752    & -5.00   & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:time           & -0.16289 & 0.00888    & -18.35  & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte       & 0.11043  & 0.02751    & 4.01    & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:tasks          & -0.08983 & 0.02530    & -3.55   & \textbf{< 0.001 ***} \\
-		contexte:tasks        & -0.10990 & 0.02477    & -4.44   & \textbf{< 0.001 ***} \\
-		time:tasks            & 0.06237  & 0.00888    & 7.03    & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte:tasks & 0.03349  & 0.03557    & 0.94    & 0.34652          \\ \bottomrule
-	\end{tabular}
-\end{table}
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f2_minus15} 
+		\caption{Effect size = observed effect size - 15\%} \label{f2_minus15}
+	\end{subfigure}
 
-f1: contexte had -ve slope, vt also has -ve slope, so vt is moving up in e ie down in b condition.
-vowel*context has -ve slope, which means that vt is moving more than ae.
-Now, vowel*context*task has -ve slope, which means in the s task, compared to p, there is  more downward movement.
-So more transfer in s task.
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f2_minus5} 
+		\caption{Effect size = observed effect size - 5\%} \label{f2_minus5}
+	\end{subfigure}
 
-f2: vt has -ve slope, contexte has -ve slope, which means e condition e things are more back, ie vowels are fronted in b condition.
-But vowel*context has a +ve slope-- vt is less backed in the e condition that ae, ie it is less fronted in the b condition, ie ae is fronted more, or shows more transfer effect, than vt.
-vowel*context*task also has +ve slope, which means thatthe -ve effect of contexte are further reduced when the task is s, ie fronting in mixed condition is less in s task
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f2_plus5} 
+		\caption{Effect size = observed effect size + 5\%} \label{f2_plus5}
+	\end{subfigure}
 
-\newpage
-\section*{Figures}
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f2_plus15} 
+		\caption{Effect size = observed effect size + 15\%} \label{f2_plus15}
+	\end{subfigure}
 
-\begin{figure}[h] %\centering
-	\includegraphics[scale=1]{vowels_e_b_final}
-	\caption{Vowel quality in unilinual English, mixed, and baseline Bengali productions}
-	\label{vowels_e_b}
+	\caption{Power analysis for fixed effects of interest at varying sample sizes: F2}
+
 \end{figure}
@@
 
-\begin{figure}[h] %\centering
-
-	\includegraphics[scale=1]{vowel_by_subject_ggplot}
-	\caption{Subject-wise mean F1 in unilingual English productions}
-	\label{boxplot_F1}
-\end{figure}
+\clearpage
+\bibliography{references}
 
 \end{document}
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: t
-%%% TeX-engine: luatex
-%%% End:

### 2022-10-01 RCode/transfer_analysis.Rmd
trial commit; 2nd round revisions

@@
 
-options(scipen=999) # turn off scientific e notation, show full numbers. To turn on: options(scipen=0)
+options(scipen=0) # turn off scientific e notation: options(scipen=999), show full numbers. To turn on: options(scipen=0)
 ```
@@
 ```{r}
-## final result plot:
+## final result plot (unilingual IE + mixed + baseline Bengali):
 
@@
 
-# raw FFs
 with(df_lobanov_gender_time, plotVowels(f1, f2, vowel, group = gender,
@@
 ```
+
+---
+
+### 29.7.2022 (second round revision):
+
+gender normalizing the raw (non-z-score normalized) data
+
+```{r}
+
+gender_normalized <- with(df_time, normLobanov(cbind(f1,f2), group=gender))
+
+df_gender_time <- df_time %>%
+  select(!c(f1,f2)) %>%
+  mutate(f1 = gender_normalized$f1,
+         f2 = gender_normalized$f2)
+
+rm(gender_normalized)
+
+```
+
+```{r}
+
+# Plots before gender normalization:
+
+xlim = c(2500, 700)
+ylim = c(1400, 100)
+
+with(df_time, plotVowels(f1, f2, vowel, group = gender,
+                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = vowel,
+                               var.col.by = vowel, var.sty.by = gender, 
+                               cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
+                               ellipse.line = T,  ellipse.fill = T,
+                               poly.line = F, poly.order = c("\\ae","\\vt"),
+                               xlim = xlim, ylim = ylim, 
+                               legend.kwd = "bottomright", legend.args = gender,
+                               fill.opacity = 0.05, pretty = TRUE))
+
+```
+
+```{r}
+
+# Plots after gender normalization:
+
+xlim = c(3, -3)
+ylim = c(4, -4)
+
+with(df_gender_time, plotVowels(f1, f2, vowel, group = gender,
+                               plot.tokens = F, plot.means = TRUE , pch.tokens = vowel, pch.means = vowel,
+                               var.col.by = vowel, var.sty.by = gender, 
+                               cex.tokens = 0.8, cex.means = 1, alpha.means = 0.5,alpha.tokens = 0.3,
+                               ellipse.line = T,  ellipse.fill = T,
+                               poly.line = F, poly.order = c("\\ae","\\vt"),
+                               xlim = xlim, ylim = ylim, 
+                               legend.kwd = "bottomright", legend.args = gender,
+                               fill.opacity = 0.05, pretty = TRUE))
+
+```
+
+---
 
@@
 phon_acco.model_f1 = lmer(f1 ~ 
-                         time*vowel + vowel*context*task + time*task +
+                         time*vowel + vowel*context +
                          (1+context|word) + (1+context+vowel|subject), 
@@
 null.model = lmer(f1~
-                   time*vowel + vowel*context*task + time*task +
+                   time*vowel +
                    (1+context|word) + (1+context+vowel|subject), 
@@
 #modelsummary(phon_acco.model)
+
+```
+
+---
+
+### 29.7.2022 (second round revision):
+
+Fitting models to non-z-score-normalized data (only gender normalized)
+
+```{r, eval=FALSE}
+
+temp_raw <- df_gender_time %>%
+  mutate(time = as.numeric(time)) %>%
+  mutate(time = scale(time))
+
+# F1
+
+# full model
+raw_phon_acco.model_f1 = lmer(f1 ~ 
+                         time*vowel + task*vowel*context +
+                         (1+context|word) + (1+context+vowel|subject), 
+                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_phon_acco.model_f1)
+
+# null model
+raw_null.model = lmer(f1~
+                   time*vowel + vowel*context +
+                   (1+context|word) + (1+context+vowel|subject), 
+                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_null.model)
+
+anova(raw_phon_acco.model_f1, raw_null.model)
 
@@
 ```{r, eval=FALSE, include=FALSE}
+
 # scratch pad: additional code for manuscript revision
@@
 ```
+
+## 29.7.2022 (second round revision):
+
+Content for separate .rmd file:
+
+Fitting models to non-z-score-normalized data (only gender normalized)
+
+```{r, eval=FALSE}
+
+temp_raw <- df_gender_time %>%
+  mutate(time = as.numeric(time)) %>%
+  mutate(time = scale(time))
+
+```
+
+### F1
+
+First, to test our first research question (L2 vowels), we fitted a model with vowel\*time, and vowel\*context as fixed effects.
+Comparing this to a null model that lacks the transfer term shows that the full model is significantly better.
+To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction.
+The full model was better.
+
+```{r}
+
+# full model
+raw_phon_acco.model_f1 = lmer(f1 ~ 
+                         time*vowel + vowel*context +
+                         (1+context|word) + (1+context+vowel|subject), 
+                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_phon_acco.model_f1)
+
+# null model 1
+raw_null.model = lmer(f1~
+                   time*vowel +
+                   (1+context|word) + (1+context+vowel|subject), 
+                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_null.model)
+
+anova(raw_phon_acco.model_f1, raw_null.model)
+
+# null model 2
+raw_null.model = lmer(f1~
+                   time*vowel + context +
+                   (1+context|word) + (1+context+vowel|subject), 
+                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_null.model)
+
+anova(raw_phon_acco.model_f1, raw_null.model)
+
+```
+
+To test the hypothesis Paradigm, we fitted a model with vowel*time and task*vowel*context as fixed effects.
+This three-way interaction term is significant.
+Comparing this model to a null model that lacks the fixed effect Task shows that the full model is better.
+Because Task has an independent effect on F1, and adding interactions increases model parameters, we fitted a second null model with Task as a separate fixed effect without the interaction term
+to test whether the interaction term is needed. However, the full model was still a better fit for the data:
+
+```{r}
+
+# full model
+raw_phon_acco.model_f1 = lmer(f1 ~ 
+                         time*vowel + task*vowel*context +
+                         (1+context|word) + (1+context+vowel|subject), 
+                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_phon_acco.model_f1)
+
+# null model 1
+raw_null.model = lmer(f1~
+                   time*vowel + vowel*context +
+                   (1+context|word) + (1+context+vowel|subject), 
+                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_null.model)
+
+anova(raw_phon_acco.model_f1, raw_null.model)
+
+# null model 2
+raw_null.model = lmer(f1~
+                   time*vowel + vowel*context + task +
+                   (1+context|word) + (1+context+vowel|subject), 
+                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_null.model)
+
+anova(raw_phon_acco.model_f1, raw_null.model)
+
+```
+
+Because an independent effect of task on FFs is unexpected (c.f.
+model interpretation below), we wanted to examine if the effect might be local, i.e.
+decrease over time.
+We added a task*time interaction term to the full model, which is significant and improves model fit.
+
+```{r}
+
+# full model
+raw_phon_acco.model_f1 = lmer(f1 ~ 
+                         time*vowel + task*vowel*context + task*time +
+                         (1+context|word) + (1+context+vowel|subject), 
+                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_phon_acco.model_f1)
+
+# null model 
+raw_null.model = lmer(f1~
+                   time*vowel + task*vowel*context +
+                   (1+context|word) + (1+context+vowel|subject), 
+                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_null.model)
+
+anova(raw_phon_acco.model_f1, raw_null.model)
+
+```
+
+Next, to examine the dynamics of transfer, we added a three-way interaction between the transfer term (vowel\*context) and time.
+This term has a significant effect, but does not significantly improve model fit compared to a null model lacking the vowel\*context\*time term.
+
+```{r}
+
+# full model
+raw_phon_acco.model_f1 = lmer(f1 ~ 
+                         time*vowel + task*vowel*context + task*time + vowel*context*time +
+                         (1+context|word) + (1+context+vowel|subject), 
+                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_phon_acco.model_f1)
+
+# null model 
+raw_null.model = lmer(f1~
+                   time*vowel + task*vowel*context + task*time +
+                   (1+context|word) + (1+context+vowel|subject), 
+                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_null.model)
+
+anova(raw_phon_acco.model_f1, raw_null.model)
+
+```
+
+Thus, the optimal model was: f1 ~ time\*vowel + vowel\*context\*task + time\*task + (1 + context|word) + (1 + context + vowel|subject).
+
+### F2
+
+First, to test our first research question (L2 vowels), we fitted a model with vowel\*time, and vowel\*context as fixed effects.
+Comparing this to a null model that lacks the transfer term shows that the full model is significantly better.
+To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction.
+The full model was better.
+
+```{r}
+
+# full model
+raw_phon_acco.model_f2 = lmer(f2 ~ 
+                         time*vowel + vowel*context +
+                         (1+context|word) + (1+context+vowel|subject), 
+                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_phon_acco.model_f2)
+
+# null model 1
+raw_null.model = lmer(f2~
+                   time*vowel +
+                   (1+context|word) + (1+context+vowel|subject), 
+                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_null.model)
+
+anova(raw_phon_acco.model_f2, raw_null.model)
+
+# null model 2
+raw_null.model = lmer(f2~
+                   time*vowel + context +
+                   (1+context|word) + (1+context+vowel|subject), 
+                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_null.model)
+
+anova(raw_phon_acco.model_f2, raw_null.model)
+
+```
+
+To test the hypothesis Paradigm, we fitted a model with vowel\*time and task\*vowel\*context as fixed effects.
+While task has an independent effect on F2, the three-way interaction term is not significant.
+However, there are significant two-way interactions between task\*context and task\*vowel.
+Comparing this to a null model
+that lacks the fixed effect Task confirms that the full model is still a better fit for the data, indicating that an optimal model should contain the term Task.
+To confirm whether Task interacts with any of the other terms, we fitted a model with Task as a separate fixed effect without the interaction term.
+Comparing this to the full (interaction) model shows that the latter is still a better fit.
+
+```{r}
+
+# full model
+raw_phon_acco.model_f2 = lmer(f2 ~ 
+                         time*vowel + task*vowel*context +
+                         (1+context|word) + (1+context+vowel|subject), 
+                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_phon_acco.model_f2)
+
+# null model 1
+raw_null.model = lmer(f2~
+                   time*vowel + vowel*context +
+                   (1+context|word) + (1+context+vowel|subject), 
+                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_null.model)
+
+anova(raw_phon_acco.model_f2, raw_null.model)
+
+# null model 2
+raw_null.model = lmer(f2~
+                   time*vowel + vowel*context + task +
+                   (1+context|word) + (1+context+vowel|subject), 
+                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_null.model)
+
+anova(raw_phon_acco.model_f2, raw_null.model)
+
+```
+
+Because an independent effect of task on FFs is unexpected (c.f.
+model interpretation below), we wanted to examine if the effect might be local, i.e.
+decrease over time.
+We added a task\*time interaction term to the full model, which is significant and improves model fit.
+
+```{r}
+
+# full model
+raw_phon_acco.model_f2 = lmer(f2 ~ 
+                         time*vowel + task*vowel*context + task*time +
+                         (1+context|word) + (1+context+vowel|subject), 
+                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_phon_acco.model_f2)
+
+# null model
+raw_null.model = lmer(f2~
+                   time*vowel + task*vowel*context +
+                   (1+context|word) + (1+context+vowel|subject), 
+                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_null.model)
+
+anova(raw_phon_acco.model_f2, raw_null.model)
+
+```
+
+Next, to examine the dynamics of transfer, we added a three-way interaction between the transfer term (vowel\*context) and time.
+This term is not significant, and does not significantly improve model fit compared to a null model lacking the vowel\*context\*time term.
+
+```{r}
+
+# full model
+raw_phon_acco.model_f2 = lmer(f2 ~ 
+                         time*vowel + task*vowel*context + task*time + vowel*context*time +
+                         (1+context|word) + (1+context+vowel|subject), 
+                         data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_phon_acco.model_f2)
+
+# null model
+raw_null.model = lmer(f2~
+                   time*vowel + task*vowel*context + task*time +
+                   (1+context|word) + (1+context+vowel|subject), 
+                   data= temp_raw, REML=FALSE, control = lmerControl(optimizer = "bobyqa"))
+
+summary(raw_null.model)
+
+anova(raw_phon_acco.model_f2, raw_null.model)
+
+```
+
+Thus, the optimal model is: f2 ~ vowel\*time + vowel\*context\*task + task\*time + (1+context|word) + (1+context+vowel|subject).

### 2022-10-01 draft_BLC/BLC_AM-ID_manuscript.tex
trial commit; 2nd round revisions

@@
 
-%Word count: 9263
+Word count (29.7.2022): 13388 -- remove 4388
 
@@
 Bi/multilingual speakers can distinguish between the phonetic norms of their languages and maintain separate sound categories for each language \citep{caramazza1973acquisition,macleod2010impact,bosch2003simultaneous}.
-However, these categories are not autonomous-- they influence each other across languages in both perception and production \citep[e.g.][]{flege1995second,fowler2008cross,flege2002assessing}, and the nature of such interaction provides crucial insights into how language `systems' are cognitively represented and processed.
+However, these categories are not autonomous--- they influence each other across languages in both perception and production \citep[e.g.][]{flege1995second,fowler2008cross,flege2002assessing}, and the nature of such interaction provides crucial insights into how language `systems' are cognitively represented and processed.
 Based on the productions of proficient bilingual speakers of Bengali and English, this study reports that cross-language phonetic influence temporarily increases during mixed-language use.%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
@@
 The present study is concerned with the latter.
-In production, this kind of influence has been variously termed \textsc{transfer, drift, accommodation,} and \textsc{interference}-- we use the term \textsc{``transfer"} here, to indicate any interaction between two sets of phonetic norms.
+In production, this kind of influence has been variously termed \textsc{transfer, drift, accommodation,} and \textsc{interference}.
+We use the term \textsc{``transfer"} here, to indicate any interaction between two sets of phonetic norms.
 
@@
 A majority of these studies use temporal properties of consonants (in particular, voice onset time or VOT) to measure transfer.
-However, the reported results vary greatly across studies, and appear to be contingent upon both language-specific features and language experience of the participants.
+However, reported results in the literature vary greatly across studies, as discussed in the following subsections, and appear to be contingent upon both language-specific features and language experience of the participants.
 This emphasizes the importance of considering data from a wider variety of
@@
 'Types of features'--separate line?- Brief mention - discursive factors in VOT.
-populations, language pairs,  and phonetic features in order to make meaningful generalizations. Widespread multilingualism in the Indian subcontinent suggests that phonetic
-%line--rephrase
-behavior in these populations can be particularly valuable towards understanding the nature of such cross-language interactions, as they are likely to reflect real-world experience with mixed-language processing.
+populations, language pairs,  and phonetic features in order to make meaningful generalizations.
+Widespread multilingualism in the Indian subcontinent suggests that phonetic  behavior in these populations can be particularly valuable towards understanding the nature of such cross-language interactions, as they are likely to reflect real-world experience with mixed-language processing.
 However, there is no work yet on short-term phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language.
@@
 The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India.
-We measure spectral properties: F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2.
+We measure spectral properties, F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2.
 Mixed-language data is elicited in two switching paradigms: cued picture-naming and code-switching.
@@
 The rest of this section is structured as follows: \ref{bengali_english_in_india} provides background information on Bengali and Indian English focusing on vowel systems.
-\ref{causes} discusses two potential sources of transfer during language processing-- global co-activation of multiple languages vs.
+\ref{causes} discusses two potential sources of transfer during language processing: global co-activation of multiple languages vs.
 the local act of switching between languages.
@@
 In \ref{paradigms}, we motivate a secondary aim of this study: to verify the postulated differences between two paradigms for eliciting short-term phonetic interaction in an experimental setup, namely cued picture-naming and code-switching.
-\ref{questions_and_hypotheses} summarizes the research questions and hypotheses.
+Finally, we summarize the research questions and hypotheses.
 
@@
 In India, more than 97 million people speak Bengali as a first language (Census of India, 2011), mostly in the state of West Bengal.
-A majority of this population also speaks other additional languages.
+A majority of this population also speaks additional languages.
 
@@
 
-The socio-linguistic facts about English in India also suggest that for the present population, mixed-language processing of English and an L1 is an ecologically valid paradigm that is likely to be a part of everyday language experience.
+The socio-linguistic facts about English in India also suggest that, for the present population, mixed-language processing of English and an L1 is an ecologically valid paradigm that is likely to be a part of everyday language experience.
 Thus, data from this population is valuable to the understanding of short-term phonetic interaction.
@@
 The vowel inventory of Western Bangla (the variety spoken by the participants in this study) consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \citep{garry2001facts}.
-The vowel chart in figure \ref{vowels_bengali} shows their distribution in the F1XF2 space.
+The vowel chart in figure \ref{vowels_bengali} shows their distribution in the F1$\times$F2 formant space.
 Note that there is no mid-central vowel category.
 
-The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2/3:, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \citep{wells1982accents, masica1972sound}.
-The vowel chart in figure \ref{vowels_ie} shows their distribution in the F1XF2 space.
+The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2/3:, a:, O, o, U, u} \citep{wells1982accents, masica1972sound}.
+The vowel chart in figure \ref{vowels_ie} shows their distribution in the F1$\times$F2 formant space.
 A single mid-central vowel corresponds to the categories \nt{2,@,3:}, which are treated as distinct in many native varieties of English \citep{nihalani1979indian,wells1982accents,hickey2005legacies,bansal1969intelligibility}.
@@
 	\item The local point of switch itself
-	\item Global co-activation of two languages -- \textsc{bilingual language mode} \citep{grosjean1998studying} 
+	\item Global co-activation of two languages: \textsc{bilingual language mode} \citep{grosjean1998studying} 
 \end{enumerate}
@@
 Overall, results suggest that: (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep[][vowel quality]{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use--- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \citep[][VOT]{olson2016role,tsui2019impact} or spontaneous conversation \citep[][VOT]{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
-(iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \cite[][VOT]{olson2016role} found no additive effects of language mode, \cite[][VOT]{olson2013bilingual} found a balanced language context to inhibit transfer compared to unbalanced contexts.
+(iii) How the two sources interact to influence the final outcome of transfer is not fully understood: \cite[][VOT]{olson2016role} found no additive effects of language mode, \cite[][VOT]{olson2013bilingual} found a balanced language context to inhibit transfer compared to unbalanced contexts.
 Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days (\cite{elias2017effects}; vowel quality), \citep[][VOT]{schwartz2015language, bullock2009trying,antoniou2011inter, vsimavckova2015immediate,vsimavckova2018patterns}.
@@
 In a bilingual mode, both language systems are expected to be nearly equally accessible throughout the test block.
-Thus, a consistent difference between switched and non-switched tokens in such a paradigm would be possible only if the effects were highly localized-- if not, we should expect a gradual convergence over the course of the experiment.
+Thus, a consistent difference between switched and non-switched tokens in such a paradigm would be possible only if the effects were highly localized--- if not, we should expect a gradual convergence over the course of the experiment.
 This is discussed in the next section.
@@
 In non-switched production, this is evidenced through a pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf.
-\textsc{equivalence classification}-- \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.
-
-In mixed-language production, the role of language status is less clear-- studies have reported unidirectional influence of L1 on L2 \citep[][VOT]{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \citep[][VOT]{tsui2019impact, olson2013bilingual}, (\cite{elias2017effects}, vowel quality) bidirectional convergence \citep[][VOT]{bullock2009trying, olson2016role}, divergence \citep[][VOT]{bullock2009trying,vsimavckova2018patterns}, and no influence \citep[][vowel quality]{muldner2019phonetics}, \citep[][phonological process]{schwartz2015language}.
+\textsc{equivalence classification}: \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.
+
+In mixed-language production, the role of language status is less clear--- studies have reported unidirectional influence of L1 on L2 \citep[][VOT]{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \citep[][VOT]{tsui2019impact, olson2013bilingual}, (\cite{elias2017effects}, vowel quality) bidirectional convergence \citep[][VOT]{bullock2009trying, olson2016role}, divergence \citep[][VOT]{bullock2009trying,vsimavckova2018patterns}, and no influence \citep[][vowel quality]{muldner2019phonetics}, \citep[][phonological process]{schwartz2015language}.
 Since most existing studies measure shifts in VOT, the observed asymmetries between languages have been variously explained either in terms of their L1 vs L2 status, or language-specific differences between long- and short-lag VOT languages.
@@
 
-\cite{olson2013bilingual} first reported a unidirectional VOT shift of L1 towards L2 in two different groups -- native speakers of Spanish (short-lag) and English (long-lag), matched for proficiency and age of L2 acquisition.
+\cite{olson2013bilingual} first reported a unidirectional VOT shift of L1 towards L2 in two different groups: native speakers of Spanish (short-lag) and English (long-lag), matched for proficiency and age of L2 acquisition.
 This established that beyond language-specific differences, the L1 vs L2 status of the language does mediate transfer.
@@
 This is a language-specific phonological constraint on transfer.
-In a similar vein, \cite{antoniou2011inter} observe that while the VOT of English stops shifted towards Greek (a short-lag language), the reverse was seen only in a specific subset of sounds-- word-medial non-nasalized stops.
+In a similar vein, \cite{antoniou2011inter} observe that while the VOT of English stops shifted towards Greek (a short-lag language), the reverse was seen only in a specific subset of sounds: word-medial non-nasalized stops.
 They attribute this to the low frequency of these sounds in Greek, making them less ``stable", and thus more amenable to cross-language influence.
@@
 
-In the present study, we focus on two vowel categories in Indian English-- the mid-central vowel \nt{2} and the low front vowel \nt{\ae}. There are two plausible sources of asymmetry between these:
+In the present study, we focus on two vowel categories in Indian English: the mid-central vowel \nt{2} and the low front vowel \nt{\ae}. There are two plausible sources of asymmetry between these:
 \begin{enumerate}
@@
 
-We expect the vowel \nt{2} to shift towards an acoustically/perceptually close category in IE -- a probable candidate being the low vowel \nt{a:}, as anecdotal observations of Bengali-accented English suggest that the sounds are perceived as close by Bengali speakers.
+We expect the vowel \nt{2} to shift towards an acoustically/perceptually close category in IE--- a probable candidate being the low vowel \nt{a:}, as anecdotal observations of Bengali-accented English suggest that the sounds are perceived as close by Bengali speakers.
 
@@
 
-The F1XF2 plot indicates that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.
+The F1$\times$F2 plot indicates that the Bengali category \nt{\ae} is more fronted, and \nt{a:} lower and more fronted, compared to the English categories \nt{\ae} and \nt{2} respectively.
 
@@
 
-Note that while there is evidence for planning in CS, precisely how this affects phonetic interaction cannot be predicted from the existing literature--- while \cite{bullock2009trying} found tokens immediately preceding the switch point to be phonetically distinct from the rest, one participant group converged towards the switch language (explained in terms of proficiency) while the other showed divergence (explained in terms of extra-linguistic factors).
+Note that while there is evidence for planning in CS, precisely how this affects phonetic interaction cannot be predicted from the existing literature: while \cite{bullock2009trying} found tokens immediately preceding the switch point to be phonetically distinct from the rest, one participant group converged towards the switch language (explained in terms of proficiency) while the other showed divergence (explained in terms of extra-linguistic factors).
 The participants in Muldner et al.'s \citeyearpar{muldner2019phonetics} study hyperarticulated switched tokens (indicating some degree of planning), and did not evidence a significant shift during mixed-language use.
@@
 The precise contribution of various (linguistic and extra-linguistic) factors and their interactions to the phonetics mixed-language use merits detailed investigation in its own right.
-In the present study, however, our focus is limited to verifying whether, in an experimental setting, the postulated discursive elements of a CS paradigm alter the extent of observable transfer compared to a paradigm that arguably lacks these--- cued picture-naming.
+In the present study, however, our focus is limited to verifying whether, in an experimental setting, the postulated discursive elements of a CS paradigm alter the extent of observable transfer compared to a paradigm that arguably lacks these: cued picture-naming.
 We are less concerned with the exact nature of differences than with their comparability.
@@
 
-\subsection{Research questions and hypotheses}\label{questions_and_hypotheses}
+\subsection*{Research questions and hypotheses}\label{questions_and_hypotheses}
 \begin{itemize}
-	\item \textbf{L2 vowels}: Are L2 vowel categories of proficient bilingual speakers affected by dynamic phonetic transfer during mixed-language production, as evidenced by differences in the spectral properties of L2 vowels in switched and non-switched tokens?
-	\item \textbf{Asymmetry}: If so, we expect an asymmetry between the vowels in the extent and direction of shift.
-Based on comparisons with Bengali vowels from \cite{dutta2021} we expect \nt{2} to be lowered and fronted in the mixed condition.
-\nt{\ae} is expected to show lowering, but to a lesser extent than \nt{2}, as it is in a part of the IE vowel space that has greater vowel density.
-	\item \textbf{Paradigm}: Is there an observable difference between the extent of transfer in a cued picture-naming vs. a code-switching paradigm? 
+	\item \textbf{RQ1---L2 vowels}: Are L2 vowel categories of proficient bilingual speakers affected by dynamic phonetic transfer during mixed-language production?
+	\item \textbf{H1---L2 vowels}: We expect to see differences in the spectral properties of L2 vowels in switched vs non-switched tokens
+	\item \textbf{RQ2---Asymmetry}: Do patterns of transfer differ between the two vowel categories?
+	\item \textbf{H2---Asymmetry}: Based on existing literature evidencing targeted and phonologically-constrained shifts, and comparisons with Bengali vowels from \cite{dutta2021}, we expect \nt{2} to be lowered and fronted in the mixed condition.
+\nt{\ae} is expected to be fronted.
+We expect the extent of shift in \nt{\ae} to be lesser than \nt{2}, as it is in a part of the IE vowel space that has greater vowel density.
+	\item \textbf{RQ3---Paradigm}: Is there an observable difference between the extent of transfer in a cued picture-naming vs. a code-switching paradigm? 
+	\item \textbf{H3---Paradigm}: While code-switching as a source of discursive factors that are likely to affect transfer behavior, this has not been systematically studied.
+Existing literature does not allow us to predict precisely what form this effect might take (c.f.
+discussion in Section \ref{paradigms}).
+Thus, this question is exploratory, and meant to serve as a starting point for more within-study comparisons
 \end{itemize}
@@
 ``plurality" as defined by \cite{khubchandani2021plural}).
-In such a scenario, it is not surprising that language-switching is common, expected, as well as used is creative ways with communicative purpose.
+In such a scenario, it is not surprising that language-switching is common, expected, as well as used in creative ways with communicative purpose.
 Responses to the section on language switching demonstrate this.
@@
 Productions were recorded in Praat \citep{boersma2016praat} at a 44.1K sampling rate.
-Each participant completed four iterations each of two tasks-- cued picture-naming and code-switching.
+Each participant completed four iterations each of two tasks--- cued picture-naming and code-switching.
 Utterances in the two conditions (unilingual and mixed) were randomly interspersed within each iteration.
@@
 	\begin{itemize}
-		\item Language cue- a word in either English or Bengali orthography
+		\item Language cue: a word in either English or Bengali orthography
 		\item A picture displayed for 50 ms; named by the participant in either English or Bengali, depending on the language cue.
 The picture of a fish was used in all trials, prompting the participant to say either \nt{fIS} or \nt{ma: \textteshlig \super{h}}, depending on the language cue
-		\item Target -- an English word; read out by the participant
+		\item Target: an English word; read out by the participant
 		\item Distracter math problem
@@
 An additional 67 tokens (2.79\%) were excluded due to errors in presentation/production, giving 2333 target word tokens (\nt{2}: 1137, \nt{\ae}: 1196) for analysis.
-Recordings were manually segmented and annotated in Praat \citep{boersma2016praat}, and the first and second formant frequencies (F1 and F2) of the target vowels \nt{2, \ae} were measured at five points-- in 10\% increments starting at 5\% into the vowel.
+Recordings were manually segmented and annotated in Praat \citep{boersma2016praat}, and the first and second formant frequencies (F1 and F2) of the target vowels \nt{2, \ae} were measured at five points: in 10\% increments starting at 5\% into the vowel.
 This gave measures of F1 and F2 at 5\%, 15\%, 25\%, 35\% and 45\% into the vowel, allowing us to observe any dynamic changes in the extent of transfer over the course of producing the vowel.
 Individual vocal tract differences affect formant frequencies across speakers.
-To avoid this confound, we used Lobanov normalization (using the PhonR package \citep{phonR} in R \citep{r}, representing each speaker's formant frequencies as the scaled distance from their own mean).
+To reduce between-speaker differences, we used z-score normalization (using the PhonR package \citep{phonR} in R \citep{r}, representing each speaker's formant frequencies as the scaled distance from their own mean).
 Plotting these vowels shows a difference between male and female speakers, such that male speakers have overall lower formant frequencies (figure \ref{gender_before}).
-Speaker normalization using the Lobanov method does not get rid of this difference.
+Speaker normalization using the z-score method does not get rid of this difference.
 Since gender-related differences in formant frequencies are not of interest in this study, we normalized individual values by gender (representing each value as a scaled distance from the group mean; figure \ref{gender_after}).
@@
 Figure \ref{vowels_unilingual_ie_bengali} shows the unilingual IE productions of the participants in relation to average positions of related vowel categories in Bengali.
-Figure \ref{vowels_e_b} overlays productions in the mixed condition over this, showing $F1XF2$ plots for IE vowel productions in the unilingual and mixed conditions, in comparison to the related Bengali categories from the SHRUTI corpus \cite{shruticorpus}.
+Figure \ref{vowels_e_b} overlays productions in the mixed condition over this, showing F1$\times$F2 vowel plots for IE vowel productions in the unilingual and mixed conditions, in comparison to the related Bengali categories from the SHRUTI corpus \cite{shruticorpus}.
 Given that the Bengali corpus data represent average vowel positions, and are not from the same set of speakers as the participants of this study, we refrained from including these in statistical analyses.
@@
 Both the common category \nt{\ae} and the L2-exclusive category \nt{2} show spectral shifts during mixed-language productions compared to unilingual productions.
-As hypothesized (c.f.
-our second research question \textbf{Asymmetry}), the results reveal asymmetries in the extent and direction of shifts between the two vowels.
+\alert{As hypothesized (c.f.
+our second research question \textbf{Asymmetry})}, the results reveal asymmetries in the extent and direction of shift between the two vowels.
 Specifically, \nt{2} is lowered and fronted, while \nt{\ae} is fronted, with the extent of shift being greater in \nt{2}.
 Comparison with canonical Bengali vowel productions shows that the shift in \nt{\ae} is towards the corresponding Bengali category \nt{\ae}, while that in \nt{2} is towards the Bengali category \nt{a:}.
-Regarding our third research question of whether the paradigm used to elicit language switching independently affects patterns of transfer (\textbf{Paradigm}), analyzing the role of Task in the results confirms this.
+Regarding \alert{our third research question of whether the paradigm} used to elicit language switching independently affects patterns of transfer (\textbf{Paradigm}), analyzing the role of Task in the results confirms this.
 We find a greater degree of transfer during code-switching compared to the cued picture-naming task.
@@
 The results of the present study indicate that the same happens in vowels.
-Moreover, the exact effect of transfer is interesting-- VOT is a one-dimensional measure.
+Moreover, the exact effect of transfer is interesting: VOT is a one-dimensional measure.
 Thus, observing an increase or decrease in VOT in a mixed language condition makes it difficult to infer whether a consonant has moved towards a specific L1/L2 counterpart, or reflects a more general system-wide shift in articulatory norms.
@@
 One point to note is that participants rated their reading proficiency in Bengali to be lower than that in English.
-Given that the stimuli in this task were presented visually, it is a potential source of explanation.
+Given that the stimuli in this task were presented visually, differences in reading proficiency are a potential source of explanation.
 However, anecdotal observation suggests that phonetic transfer in L2 productions is common in daily language use too, even in the absence of reading.
@@
 \subsection{Asymmetries in Phonetic Transfer}
-The literature on short-term phonetic transfer has frequently noted differences in transfer patterns between different sound categories in a languages.
+The literature on short-term phonetic transfer has frequently noted differences in transfer patterns between different sound categories in a language.
 This has been attributed to both general linguistic principles such as contrast preservation, and language-specific factors such as frequency and distribution of categories.
@@
 
-Our results also indicate asymmetry in the direction of shift between the two categories: the category \nt{2} is lowered and fronted in the mixed language condition, which \nt{\ae} is fronted.
+Our results also indicate asymmetry in the direction of shift between the two categories: the category \nt{2} is lowered and fronted in the mixed language condition, while \nt{\ae} is fronted.
 On the basis of previous literature indicating that both long-term and short-term shifts in phonological categories of multilingual speakers is targeted towards specific categories in the other language, we hypothesized (Section \ref{asymmetry between sounds}) that participants' IE \nt{\ae} would shift towards the Bengali \nt{\ae}, and \nt{2} towards the Bengali \nt{a:}.
@@
 In both tasks, Bengali and English words were presented in their respective orthographies.
-In the picture-naming task, this was used to cue language-- as described in  section \ref{procedure}, participants saw one slide at a time.
+In the picture-naming task, this was used to cue language--- as described in  section \ref{procedure}, participants saw one slide at a time.
 In the CS task, mixed utterances were presented in mixed orthography.
@@
 A closer look at the dynamics of this effect indicate that it is greatest at the beginning of articulation, and decreases steadily with time.
-Given that it is localized at the beginning of the target vowel, it seems likely that this lowering is an artifact of the particular stimuli used here-- resulting from some corarticulatory effect with the preceding word in the carrier phrase that was absent in the picture-naming task.
+Given that it is localized at the beginning of the target vowel, it seems likely that this lowering is an artifact of the particular stimuli used here--- resulting from some corarticulatory effect with the preceding word in the carrier phrase that was absent in the picture-naming task.
 While intriguing, a more detailed analysis of this effect is beyond the scope of the present study.

### 2022-10-01 draft_BLC/BLC_AM-ID_manuscript.tex
table for participant info

@@
 \usepackage{easyReview}
+
+%\usepackage[noae]{Sweave} %to get bold text?
 
@@
 This study investigates the phonetic outcome of mixed-language speech in Bengali and Indian English, towards understanding cross-language interaction in highly proficient bilinguals.
-We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in unilingual vs.
-mixed productions.
-Results reveal asymmetrical shifts in both vowels, towards related L1 categories.
-We interpret this as a temporary increase in cross-language phonetic interaction during mixed-language use, leading to a shift towards L1 norms, evidencing transfer effects on L2 vowels for the first time.
+We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in code-switched (mixed) vs.
+non-switched productions.
+Results reveal asymmetrical shifts in both vowels during mixed productions, towards related L1 categories.
+We interpret this as a temporary increase in cross-language phonetic interaction during mixed-language use, leading to a shift towards L1 norms, evidencing transfer effects on L2 vowels.
 We elicit mixed productions through two tasks: cued picture-naming and code-switching, to assess if experimental paradigm independently influences the behavior under study.
@@
 The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India.
-We measure spectral properties, F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2.
+We measure spectral properties, F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's baseline production of L2.
 Mixed-language data is elicited in two switching paradigms: cued picture-naming and code-switching.
@@
 Therefore we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode.
-To preserve ecological validity, we elicited both unilingual and switched utterances in a bilingual language mode.
+To preserve ecological validity, we elicited both switched and non-switched utterances in a bilingual language mode.
 Any observed differences in this paradigm would arguably result from interaction during online processing of sounds while switching between languages.
@@
 
-In the present study, our research question concerns whether L2 vowels are produced differently in unilingual vs mixed conditions.
+In the present study, our research question concerns whether L2 vowels are produced differently in nonswitch vs switch conditions.
 Previous literature on phonetic transfer suggests that such shifts are not random, but rather directional and targeted.
-Although our primary research question merely concerns whether productions in unilingual vs mixed conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories.
+Although our primary research question merely concerns whether productions in the two conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories.
 To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we processed and used existing speech data from a freely available Bengali corpus, SHRUTI \citep{shruticorpus}.
@@
 To estimate average vowel positions, we extracted formant values from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, to get estimates from the steady state, which should most closely approximate the production target.
-Next, we compared these to our participants' productions in the unilingual English condition to understand the relative positions of these categories.
+Next, we compared these to our participants' productions in the non-switched English condition to understand the relative positions of these categories.
 This is visualized in figure \ref{vowels_unilingual_ie_bengali}.
@@
 	\item \textbf{RQ2---Asymmetry}: Do patterns of transfer differ between the two vowel categories?
-	\item \textbf{H2---Asymmetry}: Based on existing literature evidencing targeted and phonologically-constrained shifts, and comparisons with Bengali vowels from \cite{dutta2021}, we expect \nt{2} to be lowered and fronted in the mixed condition.
+	\item \textbf{H2---Asymmetry}: Since existing literature demonstrates targeted and phonologically-constrained shifts, we expect transfer effects to be category-specific.
+Based on the position of IE vowels compared to those of Bengali vowels from \cite{shruticorpus}, we expect \nt{2} to be lowered and fronted in the mixed condition.
 \nt{\ae} is expected to be fronted.
-We expect the extent of shift in \nt{\ae} to be lesser than \nt{2}, as it is in a part of the IE vowel space that has greater vowel density.
+We hypothesize that the extent of shift in \nt{\ae} to be lesser than \nt{2}, as it is in a part of the IE vowel space that has greater vowel density.
 	\item \textbf{RQ3---Paradigm}: Is there an observable difference between the extent of transfer in a cued picture-naming vs. a code-switching paradigm? 
-	\item \textbf{H3---Paradigm}: While code-switching as a source of discursive factors that are likely to affect transfer behavior, this has not been systematically studied.
+	\item \textbf{H3---Paradigm}: While researchers have discussed code-switching as a source of discursive factors likely to affect transfer behavior, this has not been systematically studied.
 Existing literature does not allow us to predict precisely what form this effect might take (c.f.
 discussion in Section \ref{paradigms}).
-Thus, this question is exploratory, and meant to serve as a starting point for more within-study comparisons
+Thus, this question is exploratory, and meant to serve as a starting point for more within-study comparisons.
 \end{itemize}
@@
 \subsection{Participants} \label{participants}
-Participants were recruited through an open call circulated on social media at the EFL University campus in Hyderabad.
-Interested volunteers filled a brief language background questionnaire.
-These responses were used to screen participants, in the manner described below.
-As noted in section \ref{bengali_english_in_india}, India has an extremely multilingual an linguistically diverse population.
+Participants were recruited through an open call at the EFL University campus in Hyderabad.
+Interested volunteers completed a brief language background questionnaire which was used to screen participants as described below.
+
+As noted in section \ref{bengali_english_in_india}, the population is multilingual and linguistically diverse.
 The university campus comprises of students from different parts of the country and world, with varying L1s and different varieties of English.
-Our aim in screening participants was to minimize between-participant random variability in the study, while maintaining ecological validity considering the actual day-to-day settings in which transfer behaviors likely take place.
+Our aim in screening participants was to minimize between-participant random variability, while maintaining ecological validity considering the actual day-to-day settings in which transfer behaviors likely take place.
 We chose two metrics to do this: (i) where the participant grew up; and (ii) amount of formal education in Bengali.
@@
 Thus, we wanted to ensure that any between-participant asymmetries in the pattern of shift reflected individual differences in transfer, rather than different acoustic targets.
-Another reason to control for location was that it is a fairly reliable predictor for language(s) of education, owing to how the education system functions in the country (we elaborate on this below).
+Another reason to control for location was that it is a fairly reliable predictor for language(s) of education(see Supplementary materials for details).
 
 The second metric is Medium of Instruction (MoI) in school.
-Most schools in India follow the Three Language Policy, wherein formal education is provided in English, Hindi, and a regional language.
-(The idea and implementation of this policy, and its implications for linguistic minorities, has been a matter of much public debate as well as scholarly work; for discussions see, e.g.
-\cite{tollefson2014language, jhingran2009hundreds, khubchandani1997language, mohanty2009multilingual, ramanathan2005rethinking}).
 Sustained formal education in a language is a reliable predictor of LSRW (listening, speaking, reading, writing) skills.
-Since this study uses written stimuli, we wanted to ensure that all participants were sufficiently comfortable with reading and speaking both Bengali and English, to avoid confounds and uncontrolled variability due to reading/task effort.
-
-With these considerations in mind, we only recruited participants who had grown up in the state of West Bengal and lived there for the majority of their lives, had parents who spoke Bengali, and had received at least 5 years of formal education in Bengali (all respondents had received 5+ years of education in English and were pursuing university degrees that are taught in English), and self-reported proficiency in speaking, understanding, and reading Bengali.
+Since this study uses written stimuli, we wanted to ensure that all participants were sufficiently comfortable with reading and speaking both Bengali and English, to avoid variability due to reading/task effort.
+
+Given these considerations, we only recruited participants who had grown up in the state of West Bengal and lived there for the majority of their lives, had parents who both spoke Bengali, and had received at least 5 years of formal education in Bengali (all respondents had received 5+ years of education in English and were pursuing university degrees that are taught in English), and self-reported being proficient in speaking, understanding, and reading Bengali.
 
@@
 Using the inclusion criteria reported above, 10 volunteers (5 female, 5 male, age range 19 to 28) were invited to participate in a speech production study, and were compensated for their time.
-All participants reported normal hearing, and normal or corrected-to-normal vision.
 After completion of the study, we followed up with a detailed language background survey at a later date to learn more about the language profiles of the participants.
@@
 Together, these incorporate questions about language history, usage, attitudes, and switching experience.
-The complete questionnaire, along with responses, is available in the ``Data" component of the OSF repository for this project (c.f.
+\alert{The complete questionnaire, along with responses, is available in the ``Data" component of the OSF repository for this project (c.f.
 section \ref{data_availability}).
-
-In addition to Bengali and English, all participants reported knowing other languages, the most common being Hindi.
-Since this situation is representative of the present population and our study concerns only Bengali-English interaction, we only focused on these languages in the survey.
-The average age of acquisition was 2.6 for English (range 0 i.e.
-since birth to 6) and 0.25 (range 0 to 2) for Bengali.
-Thus, all participants can be described as simultaneous bilinguals, having acquired both English and Bengali at a young age.
-They report being comfortable using both languages at a young age, with the average being 7.5 for English and 0.3 for Bengali (on a likert scale rating where 0=for as long as I can remember, 5=since I was 5 years old, and so on).
-As mentioned earlier, all participants had attended schools where Bengali and English were both used as Medium of Instruction (MoI).
-On average, participants had had more than 16 years (range 7 to 20+) of formal education in English and 11.8 years (range 8 to 15) in Bengali at the time of the study.
-As described in the inclusion criteria, all participants reported having spent over 15 years in a Bengali-speaking region.
-At the time of recording, they were living in or around the EFL University campus in Hyderabad, which has a linguistically diverse population.
-In terms of usage, participants show a range of language usage patterns.
-Almost 100\% of the participants report exclusively using Bengali with little to no English while interacting with family, and English at school/the workplace.
-However, language use with friends varies greatly, likely reflecting the language composition of the friend circle.
-The questionnaire asked participants to rate their proficiency in each language.
-All participants rated themselves as highly proficient in speaking and understanding both Bengali and English.
-While self-ratings for reading and writing English were at ceiling, ratings for Bengali literacy (reading and writing) showed some variation, with an average of 4.87 (on a 0-6 likert scale, range 3 to 6) for reading and 4.25 (range 3 to 6) for writing.
-This is a common situation for many educated sections of society, since institutions of higher education beyond the school level largely use English (for discussions about the language politics of higher education in India, see, for example, \cite{mohanty2009multilingual}).
-The language attitude section of the questionnaire reveals interesting insights about the linguistic landscape of the population, and how people experience language in such a setting.
-When asked how much ``like themselves" participants feel while using each language, response were at ceiling for both languages.
-Similarly, most participants identified equally strongly with both an ``English-speaking" and a ``Bengali-speaking" culture, possibly because for most people, such linguistic-cultural identities are not clearly delineated and not exclusive of one another.
-These responses shed light on how multiple languages fit seamlessly into different facets of daily life to create a ``continuous" linguistic experience (c.f.
-``plurality" as defined by \cite{khubchandani2021plural}).
-In such a scenario, it is not surprising that language-switching is common, expected, as well as used in creative ways with communicative purpose.
-Responses to the section on language switching demonstrate this.
-On average, participants do not report being often unable to recall words in the target language while speaking Bengali (average rating 2.5 on a 1-6 likert scale where 1=never and 6=always) or English (average rating 2.3), indicating that switching is not strictly necessary for filling lexical gaps.
-Indeed, participants do not report switching between languages without awareness (1.8) or control (1.25).
-In spite of this, participants report a proclivity to code-switch (rating 3.5).
-Importantly, a large number of participants report volition while switching languages (4.12), and many report this to be situationally motivated (3.12).
-These responses indicate that participants are highly aware of their switching habits and move between languages for communicative purposes.
-This fact, combined with the linguistic situation described here and in section \ref{bengali_english_in_india}, are important to bear in mind while interpreting the results of a transfer study in the present population.
+--footnotes?}
 
 Given our inclusion criteria, we were able to recruit 10 participants for this study.
-Although existing studies in the field have used comparable participant numbers (e.g ), we acknowledge that this is a small sample size.
+Although existing studies in the field have used comparable participant numbers, we acknowledge that this is a small sample size.
 As detailed above, even after a recruitment call that already controlled for language background and education level (recruiting Bengali-English bilinguals in a university campus), we were able to include only 10 out of 28 volunteers (35.7\%) in the final study.
@@
 In settings where linguistic variability at the population level is much greater than these canonically studied groups, establishing comparable levels of experimental control requires significantly greater resources.
-Given that small sample sizes are likely to have low statistical power, it is fruitful to treat these findings as preliminary.
-Replication with larger sample sizes will help to confirm the observed effects, and provide much-needed data diversity (c.f.
+Given that small sample sizes are likely to have low statistical power, replication with larger sample sizes are important to confirm the observed effects, and provide much-needed data diversity (c.f.
 discussion in section \ref{introduction}).
@@
 \subsection{Procedure} \label{procedure}
-Participants were recorded individually in a sound attenuated room. %, using a ... microphone type/description?
+Participants were recorded individually in a sound attenuated room.
 Productions were recorded in Praat \citep{boersma2016praat} at a 44.1K sampling rate.
 Each participant completed four iterations each of two tasks--- cued picture-naming and code-switching.
-Utterances in the two conditions (unilingual and mixed) were randomly interspersed within each iteration.
-In the unilingual condition, the target word was preceded by another English word in the utterance.
+Utterances in the two conditions (nonswitch and switch/mixed) were randomly interspersed within each iteration.
+In the nonswitch condition, the target word was preceded by another English word in the utterance.
 In the mixed condition, the target was preceded by a Bengali word and thus produced as a switch from Bengali.
-Every target word appeared twice during a test block; once in a unilingual and once in a mixed condition.
+Every target word appeared twice during a test block; once in a nonswitch and once in a switch condition.
 Participants alternated between the two tasks, and were allowed to take a short break after each test block.
@@
 	\end{itemize}
-	Since all the target words were in English, the language cue was used to manipulate the preceding word, giving target words in non-switched (unilingual) and switched (mixed) utterances.
+	Since all the target words were in English, the language cue was used to manipulate the preceding word, giving target words in non-switched and switched (mixed) utterances.
 The orthographies of English and Bengali are visually distinct, and were therefore used to cue language.
 
-	\item Code-switching: Each slide contained one target word embedded in either an English (unilingual) or a Bengali (mixed; target produced as a switch from Bengali) carrier sentence which was read out by the participant.
+	\item Code-switching: Each slide contained one English target word embedded in either an English (nonswitch) or a Bengali (mixed; target produced as a switch from Bengali) carrier sentence which was read out by the participant.
 The carrier sentences are given below:
@@
 The experimental treatment of interest is Context, i.e.
-whether the target item was presented in a unilingual (English) or mixed (Bengali) context.
+whether the English target item was presented in an English (nonswitch) or Bengali (switch/mixed) context.
 Our main research question (\bf{L2 vowels}) asks whether L2 vowel quality (formant frequencies) of proficient bilinguals is affected by the Context of utterance.
@@
 the discussion in Section \ref{asymmetry between sounds}, leading to our second research question \textbf{Asymmetry}).
-Thus, our main effect of interest is the interaction term Context*Vowel.
+Thus, the main effect of interest is the interaction term Context*Vowel.
 This corresponds to the transfer pattern.
@@
 
-First, to test our first research question (\bf{L2 vowels}), we fitted a model with vowel*time, and vowel*context as fixed effects.
+First, to test our first and second research questions (\bf{L2 vowels and Asymmetry}), we fitted a model with vowel*time, and vowel*context as fixed effects.
 The transfer term (vowel*context) is significant in this model (model coefficient/ unstandardized effect size ($\beta$)= -2.330e-01, standard error (SE)=  5.244e-02, t= -4.44, p=0.0002).
@@
 
-To test the hypothesis \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects.
+To examine the third research question, \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects.
 This three-way interaction term is significant ($\beta$= -1.955e-01, SE=  5.453e-02, t= -3.59, p= 0.0003).
@@
 However, a boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two IE categories \nt{2, \ae} and the extent of height difference between them.
-Note that these measures are only from the English (unilingual) condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching.
+Note that these measures are only from the nonswitch condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching.
 Given that estimates for fixed effects are pooled across participants, this explains the lack of reliable differences between \nt{2} and \nt{\ae}.
@@
 The negative slope of the estimate for Context shows that F1 is lower, i.e.
-vowels are higher, in the unilingual (e) condition compared to the mixed (b) condition.
+vowels are higher, in the nonswitch (e) condition compared to the mixed (b) condition.
 Put more intuitively, vowels are lowered in the mixed condition.
@@
 The lack of a significant Vowel*Context*Time interaction shows that transfer patterns do not change across time.
-This indicates a uniform scaling of formant trajectories in the mixed condition, suggesting that the underlying acoustic targets for these vowels are different during unilingual vs mixed utterances.
+This indicates a uniform scaling of formant trajectories in the mixed condition, suggesting that the underlying acoustic targets for these vowels are different during non-switched vs mixed utterances.
 
@@
 
-First, to test hypothesis 1 (\bf{L2 transfer}), we fitted a model with vowel*time, and vowel*context as fixed effects.
+First, to test RQ1 (\textbf{L2 transfer}) and RQ2(\textbf{Asymmetry}), we fitted a model with vowel*time, and vowel*context as fixed effects.
 The transfer term (vowel*context) is significant in this model (standardized effect size ($\beta$)= 1.234e-01, standardized error (SE)=  2.055e-02, t= 6.00, p=5.46e-06).
@@
 
-To test the hypothesis \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects.
+To test RQ3 (\textbf{Paradigm}), we fitted a model with vowel*time and task*vowel*context as fixed effects.
 While task has an independent effect on F2 ($\beta$= -8.758e-02, SE= 1.756e-02, t= -4.98, p= 6.18e-07), the three-way interaction term is not significant ($\beta$= 3.349e-02, SE= 3.565e-02, t= 0.94, p= 0.34).
@@
 The negative slope of the estimate indicates that on average, vowels have lower F2 i.e.
-are backed, in the unilingual (e) condition.
+are backed, in the nonswitch (e) condition.
 In other words, vowels are fronted in the mixed condition.
@@
 shifts are targeted with respect to L1 categories.
-As discussed in section \ref{asymmetry between sounds}, although our primary research question is about whether productions in unilingual vs mixed conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories.
-Figure \ref{vowels_unilingual_ie_bengali} shows the unilingual IE productions of the participants in relation to average positions of related vowel categories in Bengali.
-Figure \ref{vowels_e_b} overlays productions in the mixed condition over this, showing F1$\times$F2 vowel plots for IE vowel productions in the unilingual and mixed conditions, in comparison to the related Bengali categories from the SHRUTI corpus \cite{shruticorpus}.
+As discussed in section \ref{asymmetry between sounds}, although our primary research question is about whether productions in switch vs nonswitch conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories.
+Figure \ref{vowels_unilingual_ie_bengali} shows the non-switched IE productions of the participants in relation to average positions of related vowel categories in Bengali.
+Figure \ref{vowels_e_b} overlays productions in the mixed condition over this, showing F1$\times$F2 vowel plots for IE vowel productions in the nonswitch and mixed conditions, in comparison to the related Bengali categories from the SHRUTI corpus \cite{shruticorpus}.
 Given that the Bengali corpus data represent average vowel positions, and are not from the same set of speakers as the participants of this study, we refrained from including these in statistical analyses.
@@
 Taken as a whole, these results bear on our original research questions in the following ways: regarding our first question of whether L2 vowels of proficient bilinguals can be affected by short-term transfer from L1 (\textbf{L2 vowels}), our results indicate that this indeed happens during mixed-language use.
-Both the common category \nt{\ae} and the L2-exclusive category \nt{2} show spectral shifts during mixed-language productions compared to unilingual productions.
-\alert{As hypothesized (c.f.
-our second research question \textbf{Asymmetry})}, the results reveal asymmetries in the extent and direction of shift between the two vowels.
+Both the common category \nt{\ae} and the L2-exclusive category \nt{2} show spectral shifts during mixed-language productions compared to non-switched productions.
+As hypothesized (c.f.
+RQ2 and H2, \textbf{Asymmetry}), the results reveal asymmetries in the extent and direction of shift between the two vowels.
 Specifically, \nt{2} is lowered and fronted, while \nt{\ae} is fronted, with the extent of shift being greater in \nt{2}.
 Comparison with canonical Bengali vowel productions shows that the shift in \nt{\ae} is towards the corresponding Bengali category \nt{\ae}, while that in \nt{2} is towards the Bengali category \nt{a:}.
-Regarding \alert{our third research question of whether the paradigm} used to elicit language switching independently affects patterns of transfer (\textbf{Paradigm}), analyzing the role of Task in the results confirms this.
+Regarding our third research question of whether the paradigm used to elicit language switching independently affects patterns of transfer (RQ3: \textbf{Paradigm}), analyzing the role of Task in the results confirms this.
 We find a greater degree of transfer during code-switching compared to the cued picture-naming task.
@@
 An often discussed issue in transfer research is whether cross-language interaction at the phonetic level during language mixing is triggered by the co-activation of languages that makes representations from both languages available, or by the local act of switching during online processing and articulation of those sounds.
-In this study, we presented unilingual and mixed stimuli in the same session, such that the language mode (participants' global language ``setting") remained the same throughout.
+In this study, we presented nonswitched English and code-switched Bengali-English stimuli in the same session, such that the language mode (participants' global language ``setting") remained the same throughout.
 In spite of this, results demonstrate consistent differences between switched and non-switched tokens.
@@
 The time course of such shifts provides an insight into this.
-Vowels in the mixed condition do not start out the same, eventually deviating in acoustics from their unilingual counterparts.
+Vowels in the mixed condition do not start out the same, eventually deviating in acoustics from their nonswitch counterparts.
 Instead, we see a difference between the conditions right from the start of articulation.
@@
 As discussed in section \ref{causes}, vowels provide a fertile testing ground for this theory in sound systems, as they avoid the confound of independent language-specific differences between long- and short-lag VOT languages.
-Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.
+Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline non-switched productions.
 In the context of an ICM-based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, this is a significant finding: it demonstrates that the postulated lower switch cost on an L2 does not entirely mask transfer effects on L2 vowels.
@@
 
-Our third research question \textbf{Paradigm} concerned the effect of experimental paradigm on the phonetic outcome of mixed-language use.
+Our third research question, \textbf{Paradigm}, concerned the effect of experimental paradigm on the phonetic outcome of mixed-language use.
 \cite{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
@@
 This study investigated how the phonetic systems of highly proficient bilingual speakers interact during online speech processing, by examining the phonetic outcome of mixed-language production in speakers of Bengali and Indian English.
-Specifically, we examined the spectral quality of two L2 vowel categories \nt{2} and \nt{\ae} produced in two experimental conditions: unilingual (English) and mixed (switching from Bengali to English).
+Specifically, we examined the spectral quality of two L2 vowel categories \nt{2} and \nt{\ae} produced in two experimental conditions: nonswitch (English) and switch/mixed (switching from Bengali to English).
 Comparisons of F1 and F2 across conditions reveal that mixed language processing temporarily increases phonetic interaction between the languages, causing the English \nt{\ae} of participants to converge towards the Bengali category \nt{\ae}, and \nt{2} to converge towards a related L1 category \nt{a:}.
@@
 \section*{Tables}
+
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+% \usepackage{multirow}
+% \usepackage{graphicx}
+\begin{table}[h]
+	\caption{Key biographical information summarized across participants. Scales: age 0(since birth)---5(since age 5) etc.; likert scale: 0(not proficient at all)---6(highly proficient)}
+	\label{table_participants}
+	\resizebox{\textwidth}{!}{%
+		\begin{tabular}{@{}lllllllll@{}}
+			\toprule
+			\multicolumn{2}{l}{\multirow{}{}{}}               & \multicolumn{3}{l}{Bengali} &  & \multicolumn{3}{l}{English} \\ \cmidrule(l){3-9} 
+			\multicolumn{2}{l}{}                                & Mean    & SD     & Range    &  & Mean   & SD     & Range     \\ \midrule
+			\multicolumn{2}{l}{Age of acquisition}              & 0.25    & 0.71   & 0 - 2    &  & 2.6    & 2.13   & 0 - 6     \\
+			\multicolumn{2}{l}{Years of formal education}       & 11.8    & 4.92   & 8 - 15   &  & 16+    & 4.15   & 7 - 20+   \\
+			\multirow{}{}{ Self-rated proficiency} & speaking  & 6       & 0      & 6 - 6    &  & 6      & 0      & 6 - 6     \\
+			& listening & 5.87    & 0.35   & 5 - 6    &  & 6      & 0      & 6 - 6     \\
+			& reading   & 4.87    & 1.24   & 3 - 6    &  & 6      & 0      & 6 - 6     \\ \bottomrule
+		\end{tabular}%
+	}
+\end{table}
+\newpage
+
 \begin{table}[h]
@@
 			Task (corresponding to the two paradigms)              & p (Picture: cued picture-naming), s (Sentence: code-switching)     \\
-			Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)                       \\
+			Context (corresponding to the experimental conditions) & e (English: nonswitch), b (Bengali: switch/mixed)                       \\
 			Time                                                   & continuous variable, range -1.6 to 1.6, scaled and centered\\
@@
 		\includestandalone[mode=image]{picture_task_eng}
-		\caption{English (non-switch) trial} \label{picture_task_english}
+		\caption{English (nonswitch) trial} \label{picture_task_english}
 	\end{subfigure}
@@
 		\includestandalone[mode=image]{picture_task_bengali} 
-		\caption{Bengali (switch) trial} \label{picture_trial_bengali}
+		\caption{Bengali-English (switch) trial} \label{picture_trial_bengali}
 	\end{subfigure}

### 2022-10-01 draft_BLC/supplementary_material.tex
table for participant info

--- 2022-05-09 draft_BLC/BLC_AM-ID_manuscript.tex
@@
 \captionsetup[figure]{skip=0pt,singlelinecheck=off}
+\usepackage{subcaption}
+
+\usepackage{standalone} %to include separate .tex files that contain mages
 
@@
 
+%\usepackage[noae]{Sweave} %to get bold text?
+
 \title{Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals}
@@
 
-%Word count: 9263
+\subsection{Participants} \label{participants}
+Participants were recruited through an open call at the EFL University campus in Hyderabad.
+Interested volunteers completed a brief language background questionnaire which was used to screen participants as described below.
 
-\begin{singlespace}
+As noted in section \ref{bengali_english_in_india}, India has an extremely multilingual an linguistically diverse population.
+The university campus comprises of students from different parts of the country and world, with varying L1s and different varieties of English.
+Our aim in screening participants was to minimize between-participant random variability, while maintaining ecological validity considering the actual day-to-day settings in which transfer behaviors likely take place.
+We chose two metrics to do this: (i) where the participant grew up; and (ii) amount of formal education in Bengali.
+The first ensures that the variety of Bengali that individuals are exposed to in their surroundings is similar across participants.
+This would mean that the target representations for Bengali vowels would be expected to be similar.
+This was especially important because we were not eliciting unilingual L1 Bengali data in this study, and were using an existing data source to estimate the position of Bengali vowels in the formant space.
+Based on previous literature (c.f.
+section \ref{asymmetry between sounds}), we expect that L2 vowels shift towards related L1 categories during mixed language utterances.
+Thus, we wanted to ensure that any between-participant asymmetries in the pattern of shift reflected individual differences in transfer, rather than different acoustic targets.
+Another reason to control for location was that it is a fairly reliable predictor for language(s) of education, owing to how the education system functions in the country (see Supplementary materials for details).
 
-\begin{myparindent}{0pt}
-Short title: Phonetic transfer\\
+The second metric is Medium of Instruction (MoI) in school.
+Most schools in India follow the Three Language Policy, wherein formal education is provided in English, Hindi, and a regional language.
+\alert{(The idea and implementation of this policy, and its implications for linguistic minorities, has been a matter of much public debate as well as scholarly work; for discussions see, e.g.
+\cite{tollefson2014language, jhingran2009hundreds, khubchandani1997language, mohanty2009multilingual, ramanathan2005rethinking}).} Sustained formal education in a language is a reliable predictor of LSRW (listening, speaking, reading, writing) skills.
+Since this study uses written stimuli, we wanted to ensure that all participants were sufficiently comfortable with reading and speaking both Bengali and English, to avoid variability due to reading/task effort.
 
-Full title: Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals\\
+Given these considerations, we only recruited participants who had grown up in the state of West Bengal and lived there for the majority of their lives, had parents who both spoke Bengali, and had received at least 5 years of formal education in Bengali (all respondents had received 5+ years of education in English and were pursuing university degrees that are taught in English), and self-reported being proficient in speaking, understanding, and reading Bengali.
 
-Authors:\\
+A total of 28 bilingual speakers of Bengali and English (18 female) responded to the recruitment call.
+Using the inclusion criteria reported above, 10 volunteers (5 female, 5 male, age range 19 to 28) were invited to participate in a speech production study, and were compensated for their time.
+All participants reported normal hearing, and normal or corrected-to-normal vision.
+After completion of the study, we followed up with a detailed language background survey at a later date to learn more about the language profiles of the participants.
+This survey was a modified and consolidated version of three popular bilingualism profiling tools: the Bilingual Language Profile (BLP) \citep{blp}, the Language Experience and Proficiency Questionnaire (LEAP-Q) \citep{leap-q} which incorporates language history and self-assessed proficiency, and the Bilingual Switching Questionnaire (BSWQ) \citep{language_switching_questionnaire}.
+Together, these incorporate questions about language history, usage, attitudes, and switching experience.
+\alert{table ...
+summarizes the key information.
+The complete questionnaire, along with responses, is available in the ``Data" component of the OSF repository for this project (c.f.
+section \ref{data_availability}).
+--footnotes?}
 
-Auromita Mitra
+In addition to Bengali and English, all participants reported knowing other languages, the most common being Hindi.
+Since this situation is representative of the present population and our study concerns only Bengali-English interaction, we only focused on these languages in the survey.
+The average age of acquisition was 2.6 for English (range 0 i.e.
+since birth to 6) and 0.25 (range 0 to 2) for Bengali.
+Thus, all participants can be described as simultaneous bilinguals, having acquired both English and Bengali at a young age.
+They report being comfortable using both languages at a young age, with the average being 7.5 for English and 0.3 for Bengali (on a likert scale rating where 0=for as long as I can remember, 5=since I was 5 years old, and so on).
+As mentioned earlier, all participants had attended schools where Bengali and English were both used as Medium of Instruction (MoI).
+On average, participants had had more than 16 years (range 7 to 20+) of formal education in English and 11.8 years (range 8 to 15) in Bengali at the time of the study.
+As described in the inclusion criteria, all participants reported having spent over 15 years in a Bengali-speaking region.
+At the time of recording, they were living in or around the EFL University campus in Hyderabad, which has a linguistically diverse population.
+In terms of usage, participants show a range of language usage patterns.
+Almost 100\% of the participants report exclusively using Bengali with little to no English while interacting with family, and English at school/the workplace.
+However, language use with friends varies greatly, likely reflecting the language composition of the friend circle.
+The questionnaire asked participants to rate their proficiency in each language.
+All participants rated themselves as highly proficient in speaking and understanding both Bengali and English.
+While self-ratings for reading and writing English were at ceiling, ratings for Bengali literacy (reading and writing) showed some variation, with an average of 4.87 (on a 0-6 likert scale, range 3 to 6) for reading and 4.25 (range 3 to 6) for writing.
+This is a common situation for many educated sections of society, since institutions of higher education beyond the school level largely use English (for discussions about the language politics of higher education in India, see, for example, \cite{mohanty2009multilingual}).
+The language attitude section of the questionnaire reveals interesting insights about the linguistic landscape of the population, and how people experience language in such a setting.
+When asked how much ``like themselves" participants feel while using each language, response were at ceiling for both languages.
+Similarly, most participants identified equally strongly with both an ``English-speaking" and a ``Bengali-speaking" culture, possibly because for most people, such linguistic-cultural identities are not clearly delineated and not exclusive of one another.
+These responses shed light on how multiple languages fit seamlessly into different facets of daily life to create a ``continuous" linguistic experience (c.f.
+``plurality" as defined by \cite{khubchandani2021plural}).
+In such a scenario, it is not surprising that language-switching is common, expected, as well as used in creative ways with communicative purpose.
+Responses to the section on language switching demonstrate this.
+On average, participants do not report being often unable to recall words in the target language while speaking Bengali (average rating 2.5 on a 1-6 likert scale where 1=never and 6=always) or English (average rating 2.3), indicating that switching is not strictly necessary for filling lexical gaps.
+Indeed, participants do not report switching between languages without awareness (1.8) or control (1.25).
+In spite of this, participants report a proclivity to code-switch (rating 3.5).
+Importantly, a large number of participants report volition while switching languages (4.12), and many report this to be situationally motivated (3.12).
+These responses indicate that participants are highly aware of their switching habits and move between languages for communicative purposes.
+This fact, combined with the linguistic situation described here and in section \ref{bengali_english_in_india}, are important to bear in mind while interpreting the results of a transfer study in the present population.
 
-EFL University,
+Given our inclusion criteria, we were able to recruit 10 participants for this study.
+Although existing studies in the field have used comparable participant numbers, we acknowledge that this is a small sample size.
+As detailed above, even after a recruitment call that already controlled for language background and education level (recruiting Bengali-English bilinguals in a university campus), we were able to include only 10 out of 28 volunteers (35.7\%) in the final study.
+Note that the criteria used here establish only a minimal level of between-participant similarity in exposure, dominance, habits, and proficiency, such as is extremely common in most comparable studies.
+This demonstrates the degree of linguistic variability in the present population.
+Norms for experimental research on bilingualism have largely been established in a Western context.
+In settings where linguistic variability at the population level is much greater than these canonically studied groups, establishing comparable levels of experimental control requires significantly greater resources.
+Given that small sample sizes are likely to have low statistical power, replication with larger sample sizes are important to confirm the observed effects, and provide much-needed data diversity (c.f.
+discussion in section \ref{introduction}).
+A more detailed discussion of power analyses in bilingualism research is included as supplementary material on OSF (c.f.
+section \ref{data_availability}).
+We hope that this can facilitate more conversation around establishing norms for power and sample size in our field.
 
-Hyderabad,
-
-India\\
-
-Indranil Dutta
-
-Jadavpur University,
-
-Kolkata,
-
-India\\
-
-Acknowledgments:\\
-We are grateful to Maumita Bhowmik and Anannya Mondal for their help with data collection and processing, and to all the participants of the study.\\
-
-Address for correspondence:\\
-Auromita Mitra,
-
-EFL University,
-
-Hyderabad- 500007,
-
-auromita.mitra@gmail.com
-
-\end{myparindent}
-
-\end{singlespace}
-
-\newpage
-\begin{myparindent}{0pt}
-Abstract:\\
-This study investigates the phonetic outcome of mixed-language speech in Bengali and Indian English, towards understanding cross-language interaction in highly proficient bilinguals.
-We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in unilingual vs.
-mixed productions.
-Results reveal dynamic and asymmetrical changes in both vowels;  \nt{2} shifting towards a related L1 category \nt{a:}.
-We interpret this as a temporary increase in cross-language phonetic interaction during mixed-language use, leading to a shift towards L1 norms, evidencing transfer effects on L2 vowels for the first time.
-We elicit mixed productions through two tasks: cued picture-naming and code-switching, to assess if experimental paradigm independently influences the behavior under study.
-Results reveal no difference across paradigms.
-We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction, postulated discursive factors in code-switching, and the issue of comparability between paradigms in transfer studies.\\
-
-Keywords: phonetic transfer; vowel quality; bilingual speech production; L1-L2 interaction; Indian English; Bengali
-
-\end{myparindent}
-
-\newpage
-
-\section{Introduction: Phonetic transfer}
-
-Bi/multilingual speakers can distinguish between the phonetic norms of their languages and maintain separate sound categories for each language \citep{caramazza1973acquisition,macleod2010impact,bosch2003simultaneous}.
-However, these categories are not autonomous-- they influence each other across languages in both perception and production \citep[e.g.][]{flege1995second,fowler2008cross,flege2002assessing}, and the nature of such interaction provides crucial insights into how language `systems' are cognitively represented and processed.
-Based on the productions of proficient bilingual speakers of Bengali and English, this study reports that cross-language phonetic influence temporarily increases during mixed-language use.%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
-%show cross-language-- rephrase
-
-%keep just the types here, and put the next 2 lines in the causes&effects section?
-Cross-language influence at the level of sounds can be studied in broadly two kinds of conditions:
-(i) While a bilingual speaker is operating in any one of their languages, by comparing bilingual speech to monolingual norms \citep[e.g.][]{guion2003vowel,caramazza1973acquisition,flege1987production}; %These studies often compare the speech of bilinguals to monolingual norms, and view the influence as changes to long-term memory representations as a result of acquiring an L2 \citep{guion2003vowel,caramazza1973acquisition,flege1987production};
-(ii) When both languages of a bilingual speaker are co-activated, by comparing speakers' mixed-language speech to their own norms while using a single language \citep[e.g.][]{grosjean1994going, bullock2009trying,elias2017effects, simonet2014phonetic}.
-%These compare productions during mixed-language use to participants' own productions while using a single language.
-The resulting influence is variously thought to involve short-term memory, online processing costs, language mode, and context-awareness.
-The present study is concerned with the latter.
-In production, this kind of influence has been variously termed \textsc{transfer, drift, accommodation,} and \textsc{interference}-- we use the term \textsc{transfer} here, to indicate any interaction between two sets of phonetic norms.
-
-Existing research on the phonetic effects of mixed-language production has largely focused on a limited set of phonologically related languages.
-A majority of these studies use temporal properties of consonants (in particular, voice onset time or VOT) to measure transfer.
-However, the reported results vary greatly across studies, and appear to be contingent upon both language-specific features and language experience of the participants.
-This emphasizes the importance of considering data from a wider variety of
-%elaborate on 'great variation'-- differences between language pairs, plus sociolinguistic situation (effects of language experience etc)-- clearer.
-'Types of features'--separate line?- Brief mention - discursive factors in VOT.
-populations, language pairs,  and phonetic features in order to make meaningful generalizations. Widespread multilingualism in the Indian subcontinent suggests that phonetic
-%line--rephrase
-behavior in these populations can be particularly valuable towards understanding the nature of such cross-language interactions, as they are likely to reflect real-world experience with mixed-language processing.
-However, there is no work yet on short-term phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language.
-
-The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India.
-We measure spectral properties: F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2.
-Mixed-language data is elicited in two switching paradigms: cued picture-naming and code-switching.
-We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction.
-The results demonstrate a dynamic effect of transfer on L2 vowels during mixed-language production, \alert{but no difference in the extent of transfer between the two paradigms --and slightly different patterns of shift among the paradigms?}.
-We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.
-
-The rest of this section is structured as follows: \ref{bengali_english_in_india} provides background information on Bengali and Indian English focusing on vowel systems.
-\ref{causes} discusses two potential sources of transfer during language processing-- global co-activation of multiple languages vs.
-the local act of switching between languages.
-Given the highly multilingual setting that characterizes this population, we note that the former is expected to be a constant feature and therefore any observed phonetic interaction is better understood as resulting from the latter.
-The design of the present study is motivated on the basis of this discussion.
-A logical consequence of this model is that such phonetic interaction must be highly localized.
-\ref{duration} elaborates on this by reviewing literature on the duration and `reversibility' of transfer effects.
-A recurring pattern that emerges from existing research on short-term phonetic transfer is that both languages of the bilingual speaker are not equally affected, although the nature of these reported differences varies greatly across studies.
-In \ref{asymmetries}, we review some of these findings, focusing on proposed sources of asymmetries, in order to highlight the factors that mediate such phonetic interaction (or fail to do so).
-We argue that these results point to an urgent need for expanding the range of language pairs and sound categories examined in order to make meaningful generalizations about the nature of short-term phonetic transfer.
-Following from this, \ref{asymmetry between sounds} discusses reported asymmetries between different sound categories of a single language, and develops the hypotheses of the present study on the basis of these.
-In \ref{paradigms}, we motivate a secondary aim of this study: to verify the postulated differences between two paradigms for eliciting short-term phonetic interaction in an experimental setup, namely cued picture-naming and code-switching.
-\ref{questions_and_hypotheses} summarizes the research questions and hypotheses.
-
-Existing studies have studied phonetic transfer along different acoustic features.
-While discussing the literature in subsequent sections, we will often indicate the acoustic feature in parentheses following a citation, for clarity.
-
-\subsection{Bengali and English in India} \label{bengali_english_in_india}
-
-\subsubsection*{Demography}
-
-Bengali (also, Bangla) is an Indo-Aryan language primarily spoken in India and Bangladesh.
-In India, more than 97 million people speak Bengali as a first language (Census of India, 2011), mostly in the state of West Bengal.
-A majority of this population also speaks other additional languages.
-
-Indian English (IE) refers to the variety of English that has developed in the Indian subcontinent.
-In India, it is spoken as an L2 by 129 million people (Census of India, 2011).
-English is one of the two official languages, used in education, law, media, as a lingua franca mainly for an educated elite in most metropolitan regions, and carries a high prestige value \citep{pandey201517, kachru1981english, tollefson2014language, kachru1983indianization}.
-
-\alert{note that there are large regional variations in Indian English, and the literature on IEphonology is thus not "standardized" or uncontested.
-For the present study, we focus on a variety of IE spoken by an educated elite, used in news channels, taught in classrooms (citations)}
-
-Recent research comparing the segmental and suprasegmental properties of IE spoken in different parts of the country report many commonalities (see \cite{sirsa2013effects} for a review).
-This suggests that in spite of the varied L1s of its speakers, IE has a target phonology that is distinct from any of these, as well as from other native varieties of English.
-Thus, it is fruitful to think of regional variations as resulting from L1-influence on a common underlying target.
-Excepting a small minority of specific communities \citep{pandey201517, wells1982accents, coelho1997anglo}, there are very few L1 speakers of IE.
-
-Considering the language usage patterns suggested by this demography, the present study focuses specifically on short-term phonetic interaction during mixed-language use, because:
-
-\begin{enumerate}[label=(\roman*)]
-	\item In multilingual populations, long-term representations of sound categories are expected to be affected by multiple languages. 
-	\item Given that examples of an IE phonology without L1 `influence' are very rare, it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a given baseline condition.
-\end{enumerate}
-
-The socio-linguistic facts about English in India also suggest that for the present population, mixed-language processing of English and an L1 is an ecologically valid paradigm that is likely to be a part of everyday language experience.
-Thus, data from this population is valuable to the understanding of short-term phonetic interaction.
-
-\subsubsection*{Vowel systems} \label{vowel systems}
-The vowel inventory of Western Bangla (the variety spoken by the participants in this study) consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \citep{garry2001facts}.
-Note that there is no mid-central vowel category.
-
-The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \citep{wells1982accents, masica1972sound}.
-%Their distribution in the F1XF2 space is shown in (insert table 1.7, Pandey).
-A single mid-central vowel corresponds to the categories \nt{2,@,3:}, which are treated as distinct in many native varieties of English \citep{nihalani1979indian,wells1982accents,hickey2005legacies,bansal1969intelligibility}.
-Since the English items used in this study are traditionally transcribed with \nt{2}, we use this symbol to indicate the mid-central vowel throughout.
-
-\subsection{What causes transfer and what does it affect?}\label{causes}
-
-Existing research has distinguished changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf.
-Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during accessing, processing, or articulation of these categories (cf.
-transfer vs.
-interference \citep{grosjean2012attempt}; competence vs.
-performance interference \citep{paradis1993linguistic}).
-Given their transient nature, dynamic changes in production during mixed-language use are generally attributed to the latter, e.g.
-online processing costs \citep[][VOT]{olson2013bilingual,tsui2019impact,vsimavckova2015immediate}, language mode \citep[][vowel quality]{simonet2014phonetic}, context-awareness \citep[][phonological variables]{khattab2013phonetic}.
-What triggers this interaction?
-\cite{olson2016role} argues that while cross-language phonetic effects are largely measured at the point of language switch, it could have two potential sources:
-\begin{enumerate}
-	\item The local point of switch itself
-	\item Global co-activation of two languages -- \textsc{bilingual language mode} \citep{grosjean1998studying} 
-\end{enumerate}
-
-A number of studies have specifically manipulated language mode, both in the presence and absence of switching.
-Overall, results suggest that: (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep[][vowel quality]{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use--- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \citep[][VOT]{olson2016role,tsui2019impact} or spontaneous conversation \citep[][VOT]{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
-(iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \cite[][VOT]{olson2016role} found no additive effects of language mode, \cite[][VOT]{olson2013bilingual} found a balanced language context to inhibit transfer compared to unbalanced contexts.
-Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days (\cite{elias2017effects}; vowel quality), \citep[][VOT]{schwartz2015language, bullock2009trying,antoniou2011inter, vsimavckova2015immediate,vsimavckova2018patterns}.
-
-\cite{grosjean1998studying} suggests that various aspects of the communicative setting, including exposure to (spoken or written) stimuli in multiple languages and awareness of the interlocutor's being bilingual, could trigger a bilingual language mode.
-The participants in the present study are in an environment which largely contains mixed-language input, multilingual interlocutors, and no stable ambient language.
-Therefore we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode.
-To preserve ecological validity, we elicited both unilingual and switched utterances in a bilingual language mode.
-Any observed differences in this paradigm would arguably result from interaction during online processing of sounds while switching between languages.
-
-In a bilingual mode, both language systems are expected to be nearly equally accessible throughout the test block.
-Thus, a consistent difference between switched and non-switched tokens in such a paradigm would be possible only if the effects were highly localized-- if not, we should expect a gradual convergence over the course of the experiment.
-This is discussed in the next section.
-
-\subsection{Duration of transfer effects} \label{duration}
-
-Research comparing bilingual speakers' phonological systems with monolinguals treats them as relatively stable over time.
-The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods \citep[e.g.][]{guion2003vowel, caramazza1973acquisition}.
-However, longitudinal studies %(citations)
-and between-subject comparisons of bilingual speakers with different durations of L2 exposure suggest that interaction between the sound systems is dynamic: over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts.
-For example, \cite[][vowel quality]{bohn1992production}
-observed that approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language influence.
-These have been interpreted as changes to long-term phonological representations.
-
-Moreover, changes due to transfer are not necessarily unidirectional or irreversible.
-\cite[][VOT]{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language.
-This not only evidenced that cross-language interaction can be triggered in an order of months, but also that the effects can be reversed within a similar time range.
-A more recent study by \cite[][VOT]{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks).
-Based on a short-term longitudinal study, \cite[][VOT]{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2.
-For some sounds and features (though not others, cf.
-sec.\ref{asymmetry between sounds}), transfer was additive over time.
-How long these effects last in the absence of regular L2 input was not tested.
-% if effects are cumulative, then the difference between switched and non-switched tokes should be lesser in later test blocks (previous two paragraphs needed?)
-
-The research discussed above concerns situations where a speaker is operating in any one of their languages.
-When it comes to the phonetics of mixed-language use, the majority of existing studies only analyze the switched (target) token.
-Thus, there are few direct measurements of the duration of short-term transfer effects.
-One study which measured this in a code-switching paradigm \citep[][VOT]{bullock2009trying} did not find any residual effects on the matrix language following a switch, suggesting that transfer during code-switching is localized.
-Indirect evidence in support of this comes from experiments which have elicited switched and non-switched tokens in the same test block and still reported differences between the two \citep[e.g.][VOT]{tsui2019impact,olson2013bilingual}, suggesting that changes due to transfer are quickly `reset'--- in an order of seconds.\\
-Note that all the studies discussed above measure VOT, which is a temporal feature.
-We have no apriori reason to assume that these durations generalize to vowel quality.
-However, findings from sub-categorical phonetic shifts triggered by other factors (such as convergence towards an interlocutor) do evidence rapid shifts in vowel quality within comparable time-frames \citep[e.g.][]{pardo2010expressing,babel2010dialect,babel2012evidence}.
-Thus in the current study, we present switched and non-switched tokens randomly within the same test block to induce a bilingual mode.
-We expect the intervening words between two subsequent targets to undo any residual effects of transfer.
-
-\alert{discussion: found that like VOT, vowel quality shifts due to switching are also highly localized, and reset in an order of seconds.
-This behavior is similar to other context-driven shifts in the literature}
-
-\subsection{Asymmetries between languages in extent and direction of transfer} \label{asymmetries}
-A recurring pattern that emerges from existing research on short-term phonetic transfer is that the extent and patterns of phonetic shift are not equivalent across the two languages of the bilingual speaker.
-Studies vary greatly in the nature of differences that they report.
-The potential causes for such asymmetry are of interest because they point towards the factors that mediate cross-language phonetic interaction.
-In this section, we discuss two proposed sources of such asymmetry that are relevant to the present study:
-
-\subsubsection*{L1 vs. L2 status} 
-Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} posits that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another.
-Thus, the patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories.
-In non-switched production, this is evidenced through a pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf.
-\textsc{equivalence classification}-- \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.
-
-In mixed-language production, the role of language status is less clear-- studies have reported unidirectional influence of L1 on L2 \citep[][VOT]{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \citep[][VOT]{tsui2019impact, olson2013bilingual}, (\cite{elias2017effects}, vowel quality) bidirectional convergence \citep[][VOT]{bullock2009trying, olson2016role}, divergence \citep[][VOT]{bullock2009trying,vsimavckova2018patterns}, and no influence \citep[][vowel quality]{muldner2019phonetics}, \citep[][phonological process]{schwartz2015language}.
-Since most existing studies measure shifts in VOT, the observed asymmetries between languages have been variously explained either in terms of their L1 vs L2 status, or language-specific differences between long- and short-lag VOT languages.
-%(cf sec.
-\ref{sound systems}).
-
-\cite{olson2013bilingual} first reported a unidirectional VOT shift of L1 towards L2 in two different groups -- native speakers of Spanish (short-lag) and English (long-lag), matched for proficiency and age of L2 acquisition.
-This established that beyond language-specific differences, the L1 vs L2 status of the language does mediate transfer.
-They interpret this asymmetry in terms of the Inhibitory Control Model (ICM) \citep{green1998mental}: to select a phonetic realization from one language, the other must be \textsc{inhibited}.
-Since L1 is the `stronger' language, the amount of inhibition required on L1 while using an L2 is greater than the inhibition required on an L2 while using the L1.
-This greater initial inhibition on L1 means that switching into L1 incurs a greater \textsc{switch cost} (more cross-language influence) than switching into L2.
-Lower switch cost results in a lack of visible cross-language transfer effects on L2.
-\cite{tsui2019impact} report comparable results, but only in participants who were not equally dominant in both languages.
-Balanced bilinguals did not demonstrate any transfer effects in VOT.
-They propose that this is because balanced bilinguals have better inhibitory control (low switch cost in both languages) due to greater experience with language switching.
-
-Since differences between long-lag and short-lag languages independently affect the phonetic realization of VOT, focusing on other phonetic features as sites for transfer can avoid this conflation, and clarify the precise effect of language status.
-However, given the implications of the ICM, it is necessary to first establish that these L2 categories can indeed be affected by dynamic interference, particularly in bilinguals with extensive language-switching experience.
-The few existing studies on vowel quality \citep{simonet2014phonetic,muldner2019phonetics,elias2017effects} or phonological processes \citep{simonet2020increased,schwartz2015language} have examined transfer effects on L1 categories.
-As far as we are aware, there is no work yet on short-term transfer in L2 vowels.
-Thus, this study builds on the existing work by examining whether the L2 vowel quality of proficient bilinguals can be affected by mixed-language processing.
-
-\alert{discussion: more about ICM, assumptions about processing}
-
-\subsubsection*{Differences between sound systems of the languages} \label{sound systems} In addition to the cognitive factors discussed above, many researchers have attributed the observed asymmetries to language-internal factors.
-Specifically, a pattern that emerges from the body of work on VOT is that a shift in long-lag VOT towards the short-lag norms of the other language is much more consistent and systematic than the converse \citep{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid}.
-In general, short-lag VOTs seem to resist accommodative shift.
-\cite{bullock2009trying} suggest that because long-lag languages offer a greater range of acceptable VOTs, there is more `room' for movement.
-In contrast, shift in short-lag languages could risk the loss of a phonological contrast (VOT being the primary cue for voicing contrast in these languages).
-This is a language-specific phonological constraint on transfer.
-In a similar vein, \cite{antoniou2011inter} observe that while the VOT of English stops shifted towards Greek (a short-lag language), the reverse was seen only in a specific subset of sounds-- word-medial non-nasalized stops.
-They attribute this to the low frequency of these sounds in Greek, making them less ``stable", and thus more amenable to cross-language influence.
-This suggests that the distribution of sounds in a language could affect the kind of transfer patterns observed.
-
-These results highlight that beyond interaction during the cognitive processing of sounds, phonetic transfer is ultimately a linguistic phenomenon, and thus subject to language-specific phonological constraints.
-Therefore, to make meaningful generalizations, it is imperative to consider data from a variety of language pairs.
-Existing research on phonetic transfer largely revolves around a limited set of phonologically related languages.
-Thus, the present study extends the scope of this research to a new pair of languages--- Indian English and Bengali.
-
-Since Bengali has a four-way laryngeal contrast, VOT does not encode a binary voiced-voiceless distinction.
-Moreover, VOT is not the only relevant cue for the corresponding phonological categories in the language \citep{dmitrieva2020acoustic}.
-This means that any effect of L1 Bengali on L2 English VOT is likely to be affected by additional phonological factors.
-Thus, avoiding any potential confounds was another reason we chose to focus on vowel quality as a target for transfer in the present study.
-
-\subsection{Asymmetries between sounds}\label{asymmetry between sounds}
-
-Asymmetries in the extent and patterns of transfer have not only been observed between languages, but also between different sounds/features of a language, in both long-term and transient interactions.
-Studies that have examined multiple sound categories have found that interactions between individual sound pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep[e.g.][vowel quality]{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot treated as an atomic measure.
-In light of the discussion about VOT in section \ref{sound systems}, this is not surprising --- if a general linguistic principle of `room for movement' and contrast constrains transfer, then we should expect it to apply to individual sounds too.
-Once again, this emphasizes the importance of examining a wider range of sound contrasts.
-
-In the present study, we focus on two vowel categories in Indian English-- the mid-central vowel \nt{2} and the low front vowel \nt{\ae}. There are two plausible sources of asymmetry between these:
-\begin{enumerate}
-	\item Position in the IE vowel space: compared to \nt{\ae}, \nt{2} exists in a part of the vowel space that has a lower vowel density.
-This affords a greater latitude for movement without risking the loss of a contrast, particularly in the vowel height (F1) dimension.
-Thus, considering purely phonological constraints on IE, we expect that a greater degree of shift is possible in \nt{2} compared to \nt{\ae}.
-Any movement in \nt{\ae} is expected to be primarily in the backness (F2) dimension.
-	\item Target of transfer: findings from existing research on cross-language phonetic interaction suggest that changes in production during mixed-language use are not random, but rather targeted with respect to categories in the other language.
-Thus, another source of asymmetry is the fact that the category \nt{2} is absent in Bengali, whereas \nt{\ae} is a common category across both languages (cf.
-sec.\ref{vowel systems}).
-\end{enumerate}
-
-%table
-
-Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} posits that sound categories which are common across languages influence each other because they share a common acoustic-phonetic space.
-Thus, if Bengali and English differ in their canonical realizations of \nt{\ae}, then we should expect the English \nt{\ae} to shift towards the corresponding Bengali category in the mixed condition.
-
-%Other categories can in principle be produced with native-like accuracy, showing little effect of transfer.
-Given that our participants are highly proficient bilinguals, and assuming that cross-language correspondences remain the same in representation and processing, this would lead us to expect more transfer in the common category, \nt{\ae}.
-
-There is no obvious competing L1 category during the production of \nt{2}.
-However, it is unlikely that this should altogether preclude a shift in \nt{2}, since at least one existing study has reported transfer effects on a non-common vowel category in a comparable paradigm \citep{simonet2014phonetic}.
-Here, participants' Catalan \nt{O} shifted towards an acoustically close category \nt{o}, which is common to Catalan and Spanish, suggesting that movement might be towards a set of phonetic norms rather than a specific corresponding category.
-
-We expect the vowel \nt{2} to shift towards an acoustically/perceptually close category in IE -- a probable candidate being the low vowel \nt{a:}, as anecdotal observations of Bengali-accented English suggest that the sounds are perceived as close by Bengali speakers.
-
-In the present study, our research question concerns whether L2 vowels are produced differently in unilingual vs mixed conditions.
-We do not elicit unilingual Bengali productions.
-To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we used production data from the freely available Bengali corpus, SHRUTI \citep{shruticorpus}.
-Formant values were extracted from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, to get formant estimates from the steady state, which should most closely approximate the production target.
-Next, we compared these to our participants' productions in the unilingual English condition to understand the relative positions of the categories.
-\alert{fig.
-..
-plots-- from current plot- remove the dotted line.
-Filter by context, then remove group=context from figure code} F1-F2 plots show that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.
-
-\subsection{A note on paradigms: connected speech or not?} \label{paradigms}
-
-This section motivates a secondary aim of this study: to verify if the extent and patterns of phonetic interaction in bilingual participants are independently mediated by the choice of experimental paradigm.
-Studies of phonetic transfer during mixed-language use have largely employed spontaneous or scripted code-switching (CS) to elicit mixed productions.
-CS is defined as the use of multiple languages within a single utterance, i.e.
-in connected speech \citep{myers1993dueling}.
-Other paradigms which have been used include cued picture-naming, delayed repetition, interpreting across languages, reading word-lists etc.
-\cite{olson2013bilingual} argues that code-switching introduces discursive factors such as discourse planning and pragmatic considerations, which ultimately reflect patterns of language practice, rather than cognitive behavior.
-This finds support from studies which report transfer effects \textsc{before} the switch point \citep{bullock2009trying}, or other evidence for planning such as hyperarticulation \citep{muldner2019phonetics} and interlocutor-awareness in spontaneous CS \citep{khattab2013phonetic}.
-To avoid such confounds, \cite{olson2013bilingual} endorses cued picture-naming (where pictures are named in rapid succession in either language, based on the given language cue) as an alternative paradigm to isolate purely phonetic effects in controlled setups.
-This observation has important implications for how the existing literature on short-term phonetic interaction is interpreted: if the discursive properties of a CS paradigm independently affect the phonetic outcome of mixed-language use, this calls to question the comparability of results across studies using these different paradigms.
-Because the reported outcomes of short-term phonetic interaction vary so greatly across studies (cf.
-section \ref{asymmetries}), generalizations about the mechanisms that underlie these patterns often rely crucially on the assumption of such comparability.
-Therefore, identifying paradigm-specific differences is particularly relevant for this line of research.
-
-Note that while there is evidence for planning in CS, precisely how this affects phonetic interaction cannot be predicted from the existing literature--- while \cite{bullock2009trying} found tokens immediately preceding the switch point to be phonetically distinct from the rest, one participant group converged towards the switch language (explained in terms of proficiency) while the other showed divergence (explained in terms of extra-linguistic factors).
-The participants in Muldner et al.'s \citeyearpar{muldner2019phonetics} study hyperarticulated switched tokens (indicating some degree of planning), and did not evidence a significant shift during mixed-language use.
-Moreover, recall that the findings of \cite{olson2016role} suggest that there is a limit on the extent of transfer--- multiple factors do not necessarily result in additive effects.
-Thus, the presence of additional factors in CS does not necessitate that this should alter the extent of observable transfer compared to any other paradigm.
-It is possible that the outcomes in the studies discussed above are better attributed to some other aspect of the experimental setup, rather than to the act of code-switching in particular.
-
-The precise contribution of various (linguistic and extra-linguistic) factors and their interactions to the phonetics mixed-language use merits detailed investigation in its own right.
-In the present study, however, our focus is limited to verifying whether, in an experimental setting, the postulated discursive elements of a CS paradigm alter the extent of observable transfer compared to a paradigm that arguably lacks these--- cued picture-naming.
-We are less concerned with the exact nature of differences than with their comparability.
-While researchers have successfully employed each of these paradigms to induce phonetic interaction, comparison across studies is difficult since they also differ in other factors.
-We provide a starting point for such comparison by eliciting productions of the same tokens from the same set of participants, in both paradigms.
-A difference in results will suggest that the planning involved in connected speech and code-switching affects transfer above and beyond the phonetic outcome of switching itself, which has implications that future work should consider.
-
-\subsection{Research questions and hypotheses}\label{questions_and_hypotheses}
-\begin{itemize}
-	\item \textbf{L2 vowels}: Are L2 vowel categories of proficient bilingual speakers affected by dynamic phonetic transfer during mixed-language production, as evidenced by differences in the spectral properties of L2 vowels in switched and non-switched tokens?
-	\item \textbf{Asymmetry}: If so, we expect an asymmetry between the vowels in the extent and direction of shift.
-Based on comparisons with Bengali vowels from \cite{dutta2021} we expect \nt{2} to be lowered and fronted in the mixed condition.
-\nt{\ae} is expected to show lowering, but to a lesser extent than \nt{2}, as it is in a part of the IE vowel space that has greater vowel density.
-	\item \textbf{Paradigm}: Is there an observable difference between the extent of transfer in a cued picture-naming vs. a code-switching paradigm? 
-\end{itemize}
-
-\section{Methodology} \label{methodology}
-The experiment had the following factorial design: 10 participants* 20 target words* 2 conditions* 2 tasks* 4 iterations = 3200 target tokens.
-
-\subsection{Participants} \label{participants}
-
-10 Bengali-English bilingual speakers (male=5, age range 19 to 28), completed the tasks.
-All the participants speak other languages in addition to Bengali and English.
-At the time of recording, they were living in or around the EFL University campus in Hyderabad, which has a linguistically diverse population.
-Speakers were selected on the basis of responses to a Language Background Questionnaire to ensure comparable LSRW (listening, speaking, reading, writing) skills in L1 and L2 across participants, and were compensated for their time.
-
-\subsection{Stimuli}
-Twenty monosyllabic English words containing the vowels \nt{2} or \nt{\ae} were selected as target words.
-To minimize coarticulatory effects on the vowel, the pre-vocalic consonants were all voiced plosives (\nt{b} or \nt{d}).
-10 monosyllabic English words that do not contain the target vowel served as filler items.
-We avoided English words that are lexicalized loanwords in Bengali.
-%The list of words can be found in sec.\ref{appendix}.
-
-\subsection{Procedure}
-Participants were recorded individually in a sound attenuated room. %, using a ... microphone type/description?
-Productions were recorded in Praat \citep{boersma2016praat} at a 44.1K sampling rate.
-Every participant performed four iterations each of two tasks-- cued picture-naming and code-switching.
-Utterances in the two conditions (unilingual and mixed) were randomly interspersed within each iteration.
-In the unilingual condition, the target word was preceded by another English word in the utterance.
-In the mixed condition, the target was preceded by a Bengali word and thus produced as a switch from Bengali.
-Every target word appeared twice during a test block; once in a unilingual and once in a mixed condition.
-Participants alternated between the two tasks, and were allowed to take a short break after each test block.
-The tasks are described below:
-
-\begin{enumerate}[]
-	\item Cued picture-naming: Participants were presented with slides in the following sequence:
-	\begin{itemize}
-		\item Language cue- a word in either English or Bengali orthography
-		\item A picture displayed for 50ms; named by the participant in either English or Bengali, depending on the language cue 
-		\item Target -- an English word; read out by the participant
-		\item Distracter math problem
-	\end{itemize}
-	Since all the target words were in English, the language cue was used to manipulate the \textsc{preceding} word, giving target words in non-switched (unilingual) and switched (mixed) utterances.
-The orthographies of English and Bengali are visually distinct, and were therefore used to cue language.
-
-	\item Code-switching: Each slide contained one target word embedded in either an English (unilingual) or a Bengali (mixed; target produced as a switch from Bengali) carrier sentence which was read out by the participant.
-The carrier sentences are given below:
-	\begin{itemize}
-		\item Unilingual sentence: That is a yellow [\textit{target word}].
-		\item Mixed-language sentence: \textipa{o-\:ta \ae k-\:ta kalo} [\textit{target word}]. 
-	\end{itemize}
-
-Gloss and translation for Bengali sentence:
-
-o-\textipa{\:t}a \qquad \textipa{\ae}k-\textipa{\:t}a \quad kalo [\textit{target word}]\\
-that-\textsc{clf}  one-\textsc{clf}  black [\textit{target word}]\\
-``that is a black [\textit{target word}]"\\
-
-The sound preceding the target word was uniform across all sentences (the final sound in \textit{yellow} is produced as [o] in Indian English), to minimize any possible differences due to coarticulatory effects on the target vowel.
-A specifier-noun construction was used to ensure identical word order in both languages, giving a uniform prosodic context for the target word across condition.
-In presenting the mixed-language utterances, we used mixed orthography.
-\\
-
-\end{enumerate}
-\subsection{Acoustic analysis}\label{analysis}
-The first iteration of each task was treated as a trial, and excluded from further analyses.
-An additional 67 tokens (2.79\%) were excluded due to errors in presentation/production, giving 2333 target word tokens (\nt{2}: 1137, \nt{\ae}: 1196) for analysis.
-Recordings were manually segmented and annotated in Praat \citep{boersma2016praat}, and the first and second formant frequencies (F1 and F2) of the target vowels \nt{2, \ae} were measured at five points-- in 10\% increments starting at 5\% into the vowel.
-This gave measures of F1 and F2 at 5\%, 15\%, 25\%, 35\% and 45\% into the vowel, allowing us to observe any dynamic changes in the extent of transfer over the course of producing the vowel.
-Individual vocal tract differences affect formant frequencies across speakers.
-To avoid this confound, we used Lobanov normalization (using the PhonR package \citep{phonR} in R \citep{r}, representing each speaker's formant frequencies as the scaled distance from their own mean (\alert{citation for lobanov normalization?}).
-Plotting the vowels \alert{(figure)} shows a difference between male and female speakers, such that male speakers have overall lower formant frequencies.
-Speaker normalization using the Lobanov method does not get rid of this difference.
-Since gender-related differences in formant frequencies are not of interest in this study, we normalized individual values by gender (representing each value as a scaled distance from the group mean).
-Statistical analyses were performed on these normalized values.
-
-\section{Results} \label{results}
-
-All statistical analyses were performed in R \citep{r}.
-We used linear mixed effects models to test the effect of experimental conditions on vowel height (F1) and backness (F2) respectively.
-Since formants were extracted from five points over the course of articulating the vowel, time was coded numerically (1-5), and treated as an independent variable.
-This was scaled and centered prior to analysis and treated as a continuous predictor.
-We fitted models using the \emph{lme4} package \citep{lme4}, and used \emph{lmerTest} \citep{lmerTest} to obtain estimates of p values for individual variables.
-For all mixed models, the alpha criterion was set at $|t| > 2 $.
-The dependent variables were normalized F1 and F2, and we fitted separate models for these.
-The independent variables or fixed factors, along with their levels, are given in table \ref{table variables} \alert{add variable type with each: factor, numerical, ...}.
-Subject and Item were treated as random factors.\\
-
-< Insert Table \ref{table variables} about here > \\
-
-\subsection{Models}
-
-\subsubsection*{Fixed effects}
-
-Our study involved two IE vowels. Vowel category is expected to systematically affect formant frequencies (FFs).
-
-The experimental treatment of interest is Context, i.e.
-whether the target item was presented in a unilingual (English) or mixed (Bengali) context.
-Our main research question (\bf{L2 vowels}) asks whether L2 vowel quality (formant frequencies) of proficient bilinguals is affected by the Context of utterance.
-We hypothesize that the direction of shift in the mixed condition is towards the corresponding L1 category.
-Thus, we expect the effect of Context on formant frequencies, i.e.
-the direction of shift, to be affected by the vowel category (c.f.
-section \ref{asymmetry between sounds}).
-Thus, our main effect of interest is the interaction term Context*Vowel.
-This corresponds to the transfer pattern.
-
-Our third research question (\bf{Paradigm}) asks whether the language-mixing paradigm, or Task, affects the extent and direction of transfer.
-If this is the case, we expect Task to interact with the transfer term, giving a three way Task*Context*Vowel interaction.
-
-Since we measured FFs at five points in the vowel, Time is expected to affect the FF.
-However, the precise effect of formant dynamics is expected to depend on the vowel category.
-Thus, we expect the interaction of Vowel*Time to predict formant frequency.
-If transfer patterns change across time, then we should expect it to interact with the transfer term, giving a Context*Vowel*Time interaction.
-
-\subsubsection*{Random effects structure}
-
-Subject and Item were treated as random factors.
-Given our hypotheses, we included random intercepts for Subject and Item, by-subject random slopes for Vowel (formant targets for vowel categories differ across individuals) and Context (individuals differ in their response to the experimental condition), and a by-item random slope for Context (the effect of experimental conditions differ across words).
-
-We tested the significance of our hypothesized explanatory variables using t-values and p-values for individual variables in the models, and tested overall model fit by comparing the full model to a corresponding null model that lacks the variable of interest using ANOVAs.
-The following subsections summarize and discuss the model outputs for F1 and F2 respectively.
-
-\subsection{Vowel Height -- F1}
-
-\subsubsection*{Model output summary}
-
-First, to test hypothesis 1 (\bf{L2 transfer}), we fitted a model with vowel*time, and vowel*context as fixed effects.
-The transfer term (vowel*context) is significant in this model (model coefficient/ unstandardized effect size ($\beta$)= -2.330e-01, standard error (SE)=  5.244e-02, t= -4.44, p=0.0002).
-Comparing this to a null model that lacks the transfer term shows that the full model is significantly better ($\chi^2(2) = 14.42, p=  0.0007$).
-To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction.
-The full model was better ($\chi^2(1) = 12.51, p= 0.0004$).
-
-To test the hypothesis \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects.
-This three-way interaction term is significant ($\beta$= -1.955e-01, SE=  5.453e-02, t= -3.59, p= 0.0003).
-We also find a significant effect of task on FFs ($\beta$=3.300e-01, SE=2.668e-02, t=12.371, p=2e-16).
-Comparing this model to null models that lack the fixed effect Task ($\chi^2(4) = 572.61, p=  2.2e-16$) shows that the full model is better.
-Because Task has an independent effect on F1, and adding interactions increases model parameters, we fitted a second null model with Task as a separate fixed effect without the interaction term to test whether the interaction term is needed.
-However, the full model was still a better fit for the data ($\chi^2(3) = 23.82, p=  2.722e-05$).
-Next, we tested two other null models to verify that the full model with the three-way interaction term is indeed the optimal model for the data, by replacing the three-way interaction term with: (i) vowel*context + task*context (to test if task only interacts with context, rather than the transfer term vowel*context).
-The full model is better ($\chi^2(2) = 23.28, p= 8.78e-06$); (ii) vowel*context + task*vowel (to test if task only interacts with vowel, rather than the transfer term vowel*context).
-The full model is better ($\chi^2(2) = 13.36, p= 0.001$).
-This shows that the three-way interaction is the best fit for the data.
-Because an independent effect of task on FFs is unexpected (c.f.
-), we wanted to see if the effect might be local, i.e.
-decrease over time.
-We added a task*time interaction term to the full model, which is significant ($\beta$= -1.216e-01, SE=  1.347e-02, t= -9.028, p= 2e-16 ) and improves model fit ($\chi^2(1) = 81.22, p= 2.2e-16$).
-
-Next, to examine the dynamics of transfer, we added a three-way interaction between the transfer term (vowel*context) and time.
-This term has a marginally significant effect ($\beta$= -4.604e-02, SE=  2.692e-02, t= -1.71, p= 0.08), but does not significantly improve model fit compared to a null model lacking the vowel*context*time term ($\chi^2(2) = 3.41, p= 0.18$).
-
-Thus, the optimal model was: f1 ~ time*vowel + vowel*context*task + time*task + (1 + context|word) + (1 + context + vowel|subject).
-Table \ref{table_f1} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model.
-\alert{p-vales smaller than 0.05 are in bold.}
-
-\subsubsection*{Interpretation}
-We extracted formant estimates from five uniformly spaced points in the vowel, from the time at 5\% into the vowel (at the start of articulation) to 45\% (at the mid-part, or steady state, which is expected to correspond to the articulatory target for the category).
-As expected, we see that F1 values are significantly predicted by Time in the model, reflecting th dynamic nature of articulation which causes acoustic variation over the course of time.
-The positive slope of the estimate for Time indicates that F1 increases, i.e.
-the two categories in the study show lowering over the course of articulation.
-Moreover, the significant Time*Vowel interaction shows that the F1 trajectory is different for the two vowels.
-This is expected, as the categories are expected to present different articulatory targets.
-
- Vowel is not a significant predictor of F1.
-This appears surprising, given that the phonological category is expected to systematically affect vowel quality.
-However, a boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two categories \nt{2, \ae} and the extent of height difference between them.
-Note that these measures are only from the English (unilingual) condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching.
-Given that estimates for fixed effects are pooled across participants, this explains the lack of reliable differences between \nt{2} and \nt{\ae}.
-The negative slope of the estimate for Context shows that F1 is lower, i.e.
-vowels are higher, in the unilingual (e) condition compared to the mixed (b) condition.
-Put more intuitively, vowels are lowered in the mixed condition.
-However, the lack significance indicates that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels.
-Instead, the significant interaction between Vowel and Context signals an asymmetrical shift in the categories \nt{2} and \nt{\ae} in the mixed-language condition.
-Specifically, the Vowel*Context interaction has a negative slope, showing that the lowering is greater in \nt{2}.
-
-The results also reveal a significant effect of Task, such that the sentence task (code-switching) leads to higher F1 values in both vowels relative to the cued picture-naming task as: vowels are lowered in the code-switching task, independent of experimental condition.
-This is surprising, as we have no aproiri reason to expect a uniform difference in vowel quality across tasks.
-A significant interaction with Time with a positive slope of the estimate shows that this effect is local: it is strongest at the beginning of the vowel, and reduces over the course of the vowel.
-This is consistent with our assumption that the unexpected Task effect is an artifact of the paradigm, possibly resulting from coarticulatory effects with the preceding sound in the carrier phrase.
-Of greater interest to us is the significant Vowel*Context*Task interaction, which indicates a difference in transfer patterns across paradigms.
-Specifically, the negative slope of the estimate suggests that transfer in the F1 (height) dimension are greater in the code-switching (s) task compared to the picture-naming (p) task.\\
-
-< Insert Figure \ref{boxplot_F1} about here >\\
-
-\subsection{Vowel backness -- F2}
-
-\subsubsection*{Model output summary}
-
-First, to test hypothesis 1 (\bf{L2 transfer}), we fitted a model with vowel*time, and vowel*context as fixed effects.
-The transfer term (vowel*context) is significant in this model (standardized effect size ($\beta$)= 1.234e-01, standardized error (SE)=  2.055e-02, t= 6.00, p=5.46e-06).
-Comparing this to a null model that lacks the transfer term shows that the full model is significantly better ($\chi^2(2) = 16.20, p= 0.0003$).
-To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction.
-The full model was better ($\chi^2(1) = 16.02, p= 6.245e-05$).
-
-To test the hypothesis \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects.
-While task has an independent effect on F2 ($\beta$= -8.758e-02, SE= 1.756e-02, t= -4.98, p= 6.18e-07), the three-way interaction term is not significant ($\beta$= 3.349e-02, SE= 3.565e-02, t= 0.94, p= 0.34).
-However, there are significant two-way interactions between task*context ($\beta$= -1.099e-01, SE= 2.483e-02, t= -4.427, p= 9.66e-06) and task*vowel ($\beta$= -8.983e-02, SE= 2.535e-02, t= -3.544, p= 0.000396).
-Comparing this to a null model that lacks the fixed effect Task confirms that the full model is still a better fit for the data ($\chi^2(4) = 572.61, p=  2.2e-16$), indicating that an optimal model should contain the term Task.
-To confirm whether Task interacts with any of the other terms, we fitted a model with Task as a separate fixed effect without the interaction term.
-Comparing this to the full (interaction) model shows that the latter is still a better fit ($\chi^2(3) = 44.8, p= 1.02e-09$).
-Next, we tested two other null models to verify that the three-way interaction term is indeed optimal for the data, by replacing the three-way interaction term with: (i) vowel*context + task*context (to test if task only interacts with context, rather than the transfer term vowel*context).
-The full model is better ($\chi^2(2) = 17.28, p= 0.0001$); (ii) vowel*context + task*vowel (to test if task only interacts with vowel, rather than the transfer term vowel*context).
-The full model is better ($\chi^2(2) = 28.45, p= 6.633e-07$).
-This shows that the three-way interaction is the best fit for the data.
-
-Because an independent effect of task on FFs is unexpected \alert{(c.f.
-next section)}, we wanted to see if the effect might be local, i.e.
-decrease over time.
-We added a task*time interaction term to the full model, which is significant ($\beta$= 6.237e-02, SE=  8.876e-03, t= 7.027, p= 2.23e-12 ) and improves model fit ($\chi^2(1) = 49.271, p= 2.23e-12$).
-
-Next, to examine the dynamics of transfer, we added a three-way interaction between the transfer term (vowel*context) and time.
-This term is not significant ($\beta$= -6.274e-03, SE= 1.774e-02, t= -0.35, p= 0.72), and the addition does not significantly improve model fit compared to a null model lacking the vowel*context*time term ($\chi^2(2) = 2.2443, p= 0.32$).
-
-Thus, the optimal model is: vowel*time + vowel*context*task + task*time + (1+context|word) + (1+context+vowel|subject).
-Table \ref{table_f2} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model.
-\alert{p-vales smaller than 0.05 are in bold.}
-
-Statistical analyses reveal that the vowel \nt{2} is lowered and fronted, while \nt{\ae} is fronted, in the mixed condition.
-Previous literature on phonetic transfer suggests that such shifts are not random, but rather directional and targeted.
-We want to understand how the direction of shift compares to corresponding Bengali categories.
-Figure \ref{vowels_e_b} summarizes the findings of the experiment ($F1XF2$ plots for vowel productions in the unilingual and mixed conditions, in comparison to the related Bengali categories from \cite{dutta2021}).
-This confirms our hypotheses (c.f.
-section \ref{asymmetry between sounds}) that \nt{\ae} moves towards the corresponding \nt{\ae} category in Bengali, whereas the shift in \nt{2} is towards the related Bengali category \nt{a:}.
-The following sections discuss the results for vowel height and backness respectively.
-
-< Insert Figure \ref{vowels_e_b} about here >
-
-\subsection*{Interpretation}
-
-As with F1, we find that Time significantly influences F2 values in the model, reflecting the dynamic nature of articulation which causes acoustic variation over the course of time.
-Moreover, the Time*Vowel interaction shows that this trajectory differs across the two vowels: while the positive slope for Time indicates that F2 increases, i.e.
-vowels are fronted with time, the negative slope of the Time*Vowel interaction shows that this effect is smaller for \nt{2}.
-This is expected, as figure \alert{ref} confirms that the category \nt{\ae} is overall more fronted than \nt{2}.
-
-Unlike in the F1 dimension, Vowel is a significant predictor of F2, with \nt{2} having significantly lower F2, i.e. produced as more backed, than \nt{\ae}.
-
-As with vowel height, there is no significant independent effect of Context
-
-The estimates for Vowel show that excepting the beginning of the vowel, the category \nt{2} is consistently produced as more backed than \nt{\ae}.
-Unlike vowel height, Context has an independent effect on vowel backness, such that both the vowels in this study have higher F2 values in the Bengali (mixed) condition, i.e.
-they are fronted in the mixed condition.
-A significant interaction between Vowel and Context reveals an asymmetry in the extent of fronting between the two vowels -- the shift is greater in the case of \nt{\ae}.
-We see a significant effect of Task on F2, such that vowels in the Sentence task (code-switching) were more backed than those in the picture-naming task.
-As with F1, this is an unexpected result, as we have no principled reason to expect code-switching to induce an across-board backing of vowels.
-A lack of significant interaction with Context suggests that there is no difference in the extent of shift between the two paradigms.
-The results of ANOVAs comparing the full model to a null model that lacks the Vowel*Context interaction factor reveal that as with vowel height, the change in vowel backness between the two conditions is not significant at 5\%, and increases over the course of the vowel.\\
-
-< Insert Table \ref{table_f2_fixed_effects} about here >
-
-\subsection{Power analysis}
-We were able to recruit 10 participants (c.f.
-section \ref{participants}).
-To confirm whether the experiment design was powerful enough to detect the effects we are investigating, we carried out post-hoc power analyses using the mixedpower package \citep{kumle2021estimating} in R.
-The target of the analysis was to calculate the power for detecting an effect for our main variables of interest: vowel*context (to test the hypotheses L2-vowels and Asymmetry), and vowel*context*task (to test the hypothesis Paradigm).
-
-Power is a function of sample size and effect size.
-Thus, a key issue in power analyses is to estimate the expected effect size for the treatment/factor of interest \citep{brysbaert2018power, kumle2021estimating}.
-This may be expressed as a standardized effect size or estimate ($\beta$), Cohen's d, or partial $\eta ^2$.
-
-Using observed effect sizes from the data for power analyses is a bad practice and has been shown to be meaningless \citep{hoenig2001abuse, gelman2019observed_effect_size}, as such "observed power" depends on the p-value.
-Therefore, we need a reasonable estimate of effect size.
-One way to do this is to use effect sizes reported in other existing literature (however, see \cite{brysbaert2018power} who recommends against using effect sizes from single published studies, as these tend to be inflated), or meta analyses.
-An alternative approach is to define a smallest effect size that is meaningful to us (smallest effect size of interest; SESOI \citep{kumle2021estimating}), and test whether the study has enough power to detect such an effect, if it were to exist.
-This needs to be theoretically motivated.
-To the best of our knowledge, there is no existing literature on typical effect sizes for spectral differences in cross-language vowel production, and not many studies that report model estimates.
-Only more discussion among researchers in this field will lead to a consensus on what is considered a typical or minimally meaningful cross-language effect.
-
-Similarly, while differences between language-switching paradigms have been informally discussed in previous literature, we are not aware of any study that has measured these differences in a controlled experimental setup, and thus no reported effect sizes to act as a guideline.
-
-Cohen (cite) arbitrarily classifies effect sizes (cohen's d values) as small, medium, and large.
-Although there ar arguments against the meaningfulness of these categories \alert{citations}, these are often used as ballpark values for power analyses in psychological studies (for a discussion, see \cite{brysbaert2018power}).
-However, this approach does not generalize well to mixed-effects analyses, particularly for complex models with multiple random effects and interactions, as is typical in psycholinguistic and bilingualism research.
-"Determining the SESOI for (G)LMMs is difficult in a simulation-based approach where effect sizes are indicated through the model’s unstandardized beta coefficients...relating effect sizes to beta coefficients in complex models is far from trivial and the authors therefore refrain from making specific recommendations.
-\citep{kumle2021estimating}".
-
-The results of a power analysis are only as meaningful as the assumptions that go into it.
-To avoid making arbitrary assumptions, we followed the approach in \cite{brysbaert2018power} (and discussed in \cite{kumle2021estimating}) of taking effect sized of the study of interest and directly modifying all the beta coefficients by a fixed percentage.
-Note that we are not assuming that doing so gets us closer to a "true effect size".
-Instead, we run simulations with a range of effect sized and report the power of our study to plausibly detect an effect of that size.
-The results are shown in the plots in figure ...
-These show power over a range of sample sizes, given the assumed effect size.
-
---> in cover letter: power analyses for complex mixed models are not trivial and is an active pursuit.
-While we have done our best to follow available literature, we would love to have more feedback on this.
-
-\section{Discussion}
-
-Taken as a whole, these results bear on our original research questions in the following ways: regarding our first question, whether L2 vowels of proficient bilinguals can be affected by short-term transfer from L1 (\textbf{L2 vowels}), our results confirm that this indeed happens during mixed-language use.
-Both the common category \nt{\ae} and the L2-exclusive category \nt{2} show spectral shifts during mixed-language productions compared to unilingual productions.
-As hypothesized (\textbf{Asymmetry}), the results reveal asymmetries in the extent and direction of shifts between the two vowels.
-Specifically, \nt{2} is lowered and fronted, while \nt{\ae} is fronted, with the extent of shift being greater in \nt{2}.
-Comparision with canonical Bengali vowel productions shows that the shift in \nt{\ae} is towards the corresponding Bengali category, while that in \nt{2} is towards the Bengali category \nt{a:}.
-Regarding our third research question, whether paradigm used to elicit language switching independently affects patterns of transfer (\textbf{Paradigm}), analyzing the role of Task in the results confirms this.
-We find a greater degree of transfer during code-switching compared to the cued picture-naming task.
-However, the difference is small, and needs to be replicated (?).
-These results and their implications are discussed in the following subsections.
-
-\subsection{Phonetic transfer in L2 vowels}
-
-One of the main aims of this study is to understand short-term phonetic transfer in L2, towards understanding language processing in proficient bilinguals.
-Two interrelated questions, discussed in section \ref{introduction} (....) bear on this: (i) what causes transfer in mixed-language speech; and (ii) what causes observed asymmetries between transfer in L1 vs L2.
-Together, these serve as guidelines for how we posit that the language systems of th bilingual exist in relation to one another, and consequently, what a cognitive model should look like-- any model should be able to account for these observations.
-
--- what causes transfer: language mode vs switching
-
-\alert{discussion: We presented data in the same session, such that the language mode (participants' overall language ``setting" remained the same).
-The results show differences between between the switched and non-switched tokens in spite of this.
-this confirms that the local act of switching does cause transfer, even in a bilingual language mode, paralleling results from \cite{olson2016role, tsui2019impact, piccinini2015voice}.
-All of these examined the local effect of switching on a temporal feature of speech, namely, VOT.
-The results of the present study indicate that the same happens in vowels.
-Moreover, the result of transfer is interesting-- in VOT, it can only increase or decrease, which makes it impossible to know if it is the particular consonant that becomes more like its L1 counterpart, or if there is some more system-wide shift.
---* check discussion section of these three studies and see if they have talked about this --* Because vowel quality measures show directionality as well as magnitude, we can see finer patterns.
-We find that it is not merely a shift towards a Bengali counterpart, but rather a shift towards Bengali \textit{norms}-- a category that doesn't exist in Bengali moves towards a related Bengali category.
-This result parallels \cite{simonet2014phonetic}, who found the Catalan ...
-of bilingual speakers shifts towards a related L1 Spanish category ..
-in a bilingual language mode (although, this study did not examine switched speech).
-(check discussion section once) What does this say about what happens during online processing and speech planning?
---- look at discussion section of the mode vs switching papers (those three + ones that found no local effect of switching): what do they say about this?
-Can this finding be accounted for with that?}
-
-\alert{It makes sense that a bilingual language mode makes the representations from both languages available.
-But what do these representations look like, and how are these processed during the act of using language?
-Examining speech during switching can shed light on this.
-We argue that what we see here is an example of \textit{creativity}-- beyond a simple L2-L1 mapping, there is a method to the movement such that it makes reference to entities beyond the level of the individual sound-- i.e., the system-level.
-If representations are combined in creative ways, then are we truly ``moving between" languages, in a real sense, during online processing within a single discourse?
-Can we think of intermediate representations, not of individual sounds, but rather of sound systems?
-While more detailed empirical data will eventually enable us to map out patterns of transfer during switched speech, this interpretation is in line with the kind of model proposed by \cite{khattab2013phonetic} based on observation from the use of phonological variables in bilingual children (check-- this the one where they talked about moving between systems-- spectrum of choices?).
-This visualizes/conceptualizes the bilingual language system as a \textit{continuum of sound systems}, from which the bilingual speaker is constantly choosing, guided by a variety of different pressures.
-We propose that vowels are a good feature to examine: in addition to continuous measures (as opposed to phonological variables), they also give us the 2-dimensional data (as opposed to VOT) that is suited to further testing and refining such a model.}
-
---- differences between L1 vs L2 transfer:
-
-Studies of short-term phonetic transfer consistently find differences between transfer patterns in L1 vs L2.
-The bulk of this literature is based on the feature VOT.
-While vowels have been show to induce L2-to-L1 influence, the present study demonstrates L1-to-L2 transfer.
-
-Previous studies with VOT have often found a lack of parallel transfer from L1 to L2.
-One postulated model that is consistent with this finding is the ICM, first proposed by \cite{green1986control} at the level of the lexicon, but extended to sound systems by \cite{olson2013bilingual,tsui2019impact}.
-This proposes that ...\alert{read Olson2013 discussion, more about assumptions of ICM and consequences about models of processing}  As discussed in section \ref{causes}, vowels provide a fertile testing ground for this theory in sound systems.
-
-Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.
-In the context of an ICM--based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, this is a significant finding: it demonstrates that the postulated lower switch cost on an L2 does not entirely mask transfer effects on L2 vowels.
-Moreover, our findings differ from \cite{tsui2019impact} who report on the basis of VOT measurements that balanced bilinguals do not evidence an effect of language-switching.
-\citeauthor{tsui2019impact} attribute the absence of transfer effects to greater experience with language switching in these participants, leading to lower switch-costs in both languages.
-We find that even proficient bilinguals with extensive experience of language switching demonstrate cross-language transfer effects in vowel production.
-Thus, following from the discussion in sec.
-\ref{asymmetries}, it might be fruitful for future studies to focus on vowels as a potential site for transfer.
-This would both expand the empirical scope of existing research (cf.
-sec.
-\ref{asymmetry between sounds}) and avoid confounding factors arising from independent differences between long- and short-lag VOT languages.
-More generally, these findings point to a need for expanding the range of features in which transfer effects are studied, in order to make robust generalizations about the mechanisms underlying observed patterns.
-
-\alert{this is an example where studying different features allows us to compare the predictions of different underlying process models.
-While most studies usually focus on a single feature due to ease of creating materials, there have been studies which have measured multiple features (e.g.
-..., or no need?).
-Measuring multiple features within a single design, with the same group of participants, would give a powerful tool to compare different models at little added cost to researchers.
-This would sidestep the unavoidable confounds that come with comparing results across studies, or even replication}
-
-\subsection{Asymmetries in Phonetic Transfer}
-The literature on short-term phonetic transfer has frequently noted differences in transfer patterns between different sound categories in a languages.
-This has been attributed to both general linguistic principles such as contrast preservation, and language-specific factors such as frequency and distribution of categories.
-Examining these asymmetries more closely, and understanding their sources, helps us figure out what factors affect/moderate transfer and thus, by extension, what level of representation and processing this interaction happens at, and the nature of short-term intra-speaker variation.
-The present study examined transfer in two L2 categories, \nt{2} and \nt{\ae}. We hypothesized an asymmetrical shift between the two vowels (\textbf{Asymmetry}), owing to differences in:
- \begin{itemize}
- 	\item their position in the Indian English vowel space: \nt{\ae} is in a more crowded part of the vowel space
- 	\item the presence/absence of a corresponding category in the L1 Bengali inventory
- \end{itemize}
-
-The significant interaction between Vowel and Context in both F1 and F2 in the present study confirm that the experimental condition affected both height and backness differently for each vowel.
-Examining the model coeffieicnets ($\beta$) values shows that the transfer effect is greater for ...
-compared to ...
-(\alert{check interaction effect and unpack}) a greater degree of shift in \nt{2}, possibly owing to its position in a less dense part of the IE vowel space.
-The category \nt{2} is lowered and fronted in the mixed language condition as expected (Figure \ref{vowels_e_b}).
-The direction of shift in \nt{\ae} is surprising-- it is fronted in the mixed-language condition.
-While phonological constraints on IE suggested that a shift in the F1 dimension could potentially risk the loss of contrast and is thus expected to be minimal, Figure \ref{vowels_e_b} shows that \nt{\ae} in Bengali is not fronted with respect to the English category.
-Thus, fronting in the mixed condition cannot be explained in relation the former, at least from the present data.
-\alert{remove-- Recall that the Bengali vowel productions used here are from a different set of speakers than our participants.
-Assuming some amount of individual differences in the precise phonetic realization of vowel categories in both languages, this could be a potential source of explanation for this result.} Data from a larger number of speakers for both languages, as well as within-speaker comparisons of English and Bengali vowels, are required to clarify this.
-
-\alert{The fact that nt shows more shift-- shows that this influence is not just at the level of individual sounds (that way would expect more shift in the common category because it's "confusing".
---check what \cite{simonet2014phonetic} says about this, about why shift happens.
-But this-- shows that the L1 influence is at a system-level.
-More inclined towards using a Bengali-like phonological system.
-Has implications for cognitive models of language and what happens during short-term interaction)}
-
-Switching paradigm
-Our third research question \textbf{Paradigm} concerned the effect of experimental paradigm on the phonetic outcome of mixed-language use.
-\cite{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
-If such discursive properties independently affect the phonetic outcome of language mixing in an experimental setup, this calls to question the comparability of results across studies using these different paradigms.
-We verified this by comparing productions of the same items from the same group of participants in a scripted CS paradigm and a cued picture-naming paradigm.
-We found no significant interaction of Task either with Context, or with the Vowel*Context interaction, suggesting that the extent of phonetic transfer was not affected by the paradigm.
-There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in observable transfer at all.
-This would mean that the pre-switch effects observed, for example, in \cite{muldner2019phonetics} and \cite{bullock2009trying} are better interpreted as resulting from some other aspect of the experimental setup, than the act of code-switching itself.
-While planning could additionally affect the words preceding the switch, the effects on the switched token itself do not change.
-(ii) Our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
-Thus, another possible explanation is that the discursive effects discussed by \cite{olson2013bilingual} are triggered by more naturalistic paradigms where attention to meaning and interlocutor are important.
-More within-study comparisons with less structured paradigms, and spontaneous code-switching data, can clarify this.
-The results of the present study indicate that for the purposes of observing dynamic phonetic transfer in an experimental setup, elicitation through cued-picture-naming vs.
-(controlled) code-switching does not significantly alter the behavior under study.
-Thus, the choice of paradigm can be based on other considerations.
-Moreover, this suggests that (all else being equal) the results from such studies can be compared directly.
-
-The paradigm did affect vowel quality, but this applied uniformly across experimental conditions: all vowels were more backed and lowered in the code-switching task.
-Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this lowering is an artifact of the particular stimuli used here---resulting from some corarticulatory effect that was absent in the picture-naming task.
-While this is intriguing, it is beyond the scope of the present study.
-\\
-
-\section{Conclusion}
-This study examined the phonetic outcome of mixed-language production in a group of highly proficient bilingual speakers of Bengali and Indian English.
-Specifically, we examined the spectral quality of two L2 vowel categories \nt{2} and \nt{\ae} produced in two experimental conditions: unilingual (English) and mixed (switching from Bengali to English).
-Comparisons of F1 and F2 reveal that mixed language processing temporarily increases phonetic interaction between the languages, causing \nt{2} to converge towards a related L1 category \nt{a:}.
-The extent and direction of shift is asymmetrical across the two vowel categories, which parallels findings in prior studies.
-This asymmetry appears to be mediated by the language-specific distribution of vowel categories.
-These findings evidence an effect of phonetic transfer on L2 vowels for the first time, and raise questions for proposed inhibition-based accounts of cross-language phonetic interaction.
-Mixed productions were elicited in two switching paradigms: code-switching and cued picture-naming, to verify whether the postulated discursive elements in code-switching independently influence the outcome of phonetic transfer.
-Results suggest that in an experimental setup, elicitation through cued-picture-naming vs.
-(controlled) code-switching does not significantly alter the behavior under study.
-
-This study has a number of limitations: (i) the `baseline' values for Bengali vowel categories were estimated from a different set of speakers than the participants of this study, potentially obscuring the role of individual differences in the realization of vowel categories; within-speaker comparisons of L1 and L2 vowels is likely to offer a more fine-grained view of the direction of shift in relation to L1 categories.
-(ii) we used a frame sentence to elicit code-switched productions such that not all the sentences were pragmatically well-formed.
-This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study.
-Future work should address this by considering data from more naturalistic code-switching paradigms.
-
-\section*{Abbreviations}
-\textsc{clf} = classifier
-
-\section*{Data availability}
-The materials, data files, and analysis code for this study are openly available in OSF at:  \url{https://doi.org/10.17605/OSF.IO/DSB2X}.
-
-\section*{Competing interests}
-The authors declare none.
-
-%\printbibliography %for use with biblatex; comment out if you use natbib
-\bibliography{references} %for use with natbib; comment out if you use biblatex, and change 'sample' by the name of your bib-file
-
-\newpage
-\section*{Tables}
-\begin{table}[h]
-	\caption{Fixed effects}
-	%\centering
-	\resizebox{\textwidth}{!}{%
-		\begin{tabular}{@{}ll@{}}
-			\toprule
-			Variable name                                                 & Levels                                                            \\ \midrule
-			Gender                                                 & male, female                                                      \\
-			Vowel                                                  & \nt{2}, \nt{\ae} \\
-			Task (corresponding to the two paradigms)              & p (Picture: cued picture-naming, s (Sentence: code-switching)     \\
-			Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)                       \\
-			Time                                                   & continuous variable, range -1.6 to 1.6, scaled and centered\\
-			 \bottomrule
-		\end{tabular}%
-	}
-	\label{table variables}
-\end{table}
-
-\newpage
-
-% Please add the following required packages to your document preamble:
-% \usepackage{booktabs}
-\begin{table}[h]
-	\centering
-	\caption{Optimal model: F1}
-	\label{table_f1}
-	\begin{tabular}{@{}lllll@{}}
-		\toprule
-		& Estimate  & Std. Error & t value & p value          \\ \midrule
-		(Intercept)           & -0.053474 & 0.178435   & -0.300  & 0.769            \\
-		time                  & 0.139140  & 0.011556   & 12.040  & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}                & -0.147480 & 0.223022   & -0.661  & 0.519            \\
-		contexte              & -0.013419 & 0.068357   & -0.196  & 0.846            \\
-		tasks                 & 0.330034  & 0.026586   & 12.414  & \textbf{< 0.001 ***} \\
-		time:vowel\textipa{2}           & -0.170892 & 0.013469   & -12.688 & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte       & -0.137062 & 0.060884   & -2.251  & \textbf{0.031 *}   \\
-		vowel\textipa{2}:tasks          & 0.009002  & 0.038498   & 0.234   & 0.815            \\
-		contexte:tasks        & 0.073341  & 0.037597   & 1.951   & 0.051 .          \\
-		time:tasks            & -0.121604 & 0.013469   & -9.028  & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte:tasks & -0.195613 & 0.054338   & -3.600  & \textbf{< 0.001 ***} \\ \bottomrule
-	\end{tabular}
-\end{table}
-
-\newpage
-
-% Please add the following required packages to your document preamble:
-% \usepackage{booktabs}
-\begin{table}[h]
-	\centering
-	\caption{Optimal model: F2}
-	\label{table_f2}
-	\begin{tabular}{@{}lllll@{}}
-		\toprule
-		& Estimate & Std. Error & t value & p value          \\ \midrule
-		(Intercept)           & 0.23202  & 0.22100    & 1.05    & 0.304            \\
-		vowel\textipa{2}                & -0.38596 & 0.14858    & -2.60   & \textbf{0.015 *}   \\
-		time                  & 0.04287  & 0.00761    & 5.63    & \textbf{< 0.001 ***} \\
-		contexte              & -0.01720 & 0.02823    & -0.61   & 0.548            \\
-		tasks                 & -0.08758 & 0.01752    & -5.00   & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:time           & -0.16289 & 0.00888    & -18.35  & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte       & 0.11043  & 0.02751    & 4.01    & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:tasks          & -0.08983 & 0.02530    & -3.55   & \textbf{< 0.001 ***} \\
-		contexte:tasks        & -0.10990 & 0.02477    & -4.44   & \textbf{< 0.001 ***} \\
-		time:tasks            & 0.06237  & 0.00888    & 7.03    & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte:tasks & 0.03349  & 0.03557    & 0.94    & 0.34652          \\ \bottomrule
-	\end{tabular}
-\end{table}
-
-f1: contexte had -ve slope, vt also has -ve slope, so vt is moving up in e ie down in b condition.
-vowel*context has -ve slope, which means that vt is moving more than ae.
-Now, vowel*context*task has -ve slope, which means in the s task, compared to p, there is  more downward movement.
-So more transfer in s task.
-
-f2: vt has -ve slope, contexte has -ve slope, which means e condition e things are more back, ie vowels are fronted in b condition.
-But vowel*context has a +ve slope-- vt is less backed in the e condition that ae, ie it is less fronted in the b condition, ie ae is fronted more, or shows more transfer effect, than vt.
-vowel*context*task also has +ve slope, which means thatthe -ve effect of contexte are further reduced when the task is s, ie fronting in mixed condition is less in s task
-
-\newpage
-\section*{Figures}
-
-\begin{figure}[h] %\centering
-	\includegraphics[scale=1]{vowels_e_b_final}
-	\caption{Vowel quality in unilinual English, mixed, and baseline Bengali productions}
-	\label{vowels_e_b}
-\end{figure}
-
-\newpage
-
-\begin{figure}[h] %\centering
-
-	\includegraphics[scale=1]{vowel_by_subject_ggplot}
-	\caption{Subject-wise mean F1 in unilingual English productions}
-	\label{boxplot_F1}
-\end{figure}
+\bibliography{references}
 
 \end{document}
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: t
-%%% TeX-engine: luatex
-%%% End:

### 2022-10-01 draft_BLC/supplementary_material.tex
added supplementary material file

@@
 \usepackage{subcaption}
-
-\usepackage{standalone} %to include separate .tex files that contain mages
 
@@
 
-%\usepackage[noae]{Sweave} %to get bold text?
+%-----------------------------
 
-\title{Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals}
+%-----------------------
 
-% \author[Auromita \& Indranil]% short form of the author names for the running header. If no short author is given, no authors print in the headers.
-% {%as many authors as you like, each separated by \AND.
-	%   \spauthor{Auromita Mitra\\ 
-		%   \institute{The EFL University}\\
-		%   \small{%105, Bd. Raspail, 75005 Paris\\
-			%   auromita.mitra@gmail.com}
-		%   }
-	%   \AND
-	%   \spauthor{Indranil Dutta \\
-		%   \institute{Jadavpur University}\\
-		%   \small{%Warmoesberg 26, 1000 Brussel\\
-			%   indranildutta.lnl@jadavpuruniversity.in}
-		%   }%
-	% }
-%\usepackage{gb4e}
+\title{Supplementary material}
+\date{}
+\maketitle
 
@@
 	\bibliographystyle{apacite}
-	%\sffamily
-	%\maketitle
 
-\subsection{Participants} \label{participants}
+\section*{Participant and demographic information}
+
 Participants were recruited through an open call at the EFL University campus in Hyderabad.
@@
 
+\section*{Supplementary material: power analysis}
+
+Owing to participant availability given our inclusion criteria, we were able to recruit 10 participants for this study (c.f.
+section Participants in main article).
+As part of the peer review process, we carried out post-hoc power analyses using the \textit{mixedpower} package \citep{kumle2021estimating} in R.
+Although there is a body of literature that debates the usefulness of post-hoc power analyses in interpreting the findings of a completed study where statistical analyses have already been performed \citep{lakens_2021,dziak2020interpretation,lenth2007post,gelman2019observed_effect_size}, we have included this section as a discussion around considerations for power calculation in bilingualism research.
+
+The aim of the analysis was to explore the power of our experiment design to detect the effects we are investigating.
+The package simulates responses based on a given dataset and runs the specified analyses on these new data, estimating power by calculating how frequently statistical significance is obtained in the simulated data.
+The target of the analysis was to calculate the power for detecting an effect for our main variables of interest: vowel*context (to test the hypotheses L2-vowels and Asymmetry), and vowel*context*task (to test the hypothesis Paradigm).
+
+Power is a function of sample size and effect size.
+Thus, a key issue in power analyses is to estimate the expected effect size for the treatment/factor of interest \citep{brysbaert2018power, kumle2021estimating}.
+This may be expressed by a variety of measures, including Cohen's d, $\eta ^2$, and model estimates ($\beta$).
+Using observed effect sizes from the data for power analyses has been shown to be meaningless \citep{hoenig2001abuse, gelman2019observed_effect_size}, since such ``observed power" is a function of the p-value.
+Therefore, we need a reasonable estimate of effect size.
+One way to do this is to use effect sizes reported in other existing literature (however, see \cite{brysbaert2018power} who recommends against using effect sizes from single published studies, as these tend to be inflated), or meta analyses.
+An alternative approach is to define a smallest effect size that is meaningful to us (smallest effect size of interest; SESOI \citep{kumle2021estimating}), and test whether the study has enough power to detect such an effect, if it were to exist.
+This value needs to be theoretically motivated.
+To the best of our knowledge, there is no existing literature on typical effect sizes for spectral differences in cross-language vowel production, and not many studies that report effect sizes.
+Only more discussion in this field will lead to a consensus on what is considered a typical or minimally meaningful cross-language effect.
+Similarly, while differences between language-switching paradigms have been informally discussed in previous literature, we are not aware of any study that has measured these differences in a controlled experimental setup, and thus no reported effect sizes for between-paradigm differences in phonetic transfer to act as a guideline.
+
+Cohen \citeyearpar{cohen1988statistical} classifies effect sizes (Cohen's d values) as small ($\leq$0.2), medium (0.5), and large ($\geq$ 0.8).
+Although the usefulness of this classification has been debated in the literature, these are often used as ballpark values for power analyses in psychological studies (for a discussion, see \cite{brysbaert2018power}).
+However, this approach does not generalize well to mixed-effects analyses, particularly for complex models with multiple random effects and interactions, as is typical in psycholinguistic and bilingualism research.
+``Determining the SESOI for (G)LMMs is difficult in a simulation-based approach where effect sizes are indicated through the model’s unstandardized beta coefficients...relating effect sizes to beta coefficients in complex models is far from trivial and the authors therefore refrain from making specific recommendations.
+\citep{kumle2021estimating}".
+
+The results of a power analysis are only as meaningful as the assumptions that go into it.
+To avoid making arbitrary assumptions, we followed the approach in \cite{brysbaert2018power} (and discussed in \cite{kumle2021estimating}) of extracting the effect sizes from the study of interest and directly modifying all the beta coefficients by a fixed percentage.
+Note that we are not assuming that doing so gets us closer to a ``true effect size".
+Instead, we run simulations with a range of effect sizes and report the power of our study to plausibly detect an effect of that size.
+The results are shown in the plots in figures \ref{figure_power_analysis_f1} and \ref{figure_power_analysis_f2}.
+Each panel presents power analysis at a different effect size, and is based on 500 simulations.
+The plots show power (y-axis) over a range of sample sizes (x-axis), given the assumed effect size.
+
+As expected, power increases as a function of sample size for all effects.
+However, of greater interest to us is the differences between the power curves in the four panels (a,b,c,d).
+These demonstrate how results of the power analysis vary as a function of assumed effect size.
+Note the the range of effect sizes used here is fairly narrow, varying between 15\% smaller and 15\% greater than the observed effect sizes.
+These differences demonstrate the importance of theoretically justified effect size estimates before attempting to make inferences based on power analyses.
+This applies equally to aposteriori power analyses done to make decisions about sample sizes prior to data collection (to reiterate, post-hoc power analyses are best avoided).
+This emphasizes the importance for bilingualism studies to report effect sizes as a part of their results, which can facilitate discussions and meta-analyses that eventually lead to such norms being established in the field.
+\footnote{Thanks to the anonymous reviewer at BLC for initiating a fruitful discussion around this.}
+
+\newpage
+\begin{figure} \label{figure_power_analysis_f1}
+	\centering
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f1_minus15} 
+		\caption{Effect size = observed effect size - 15\%} \label{f1_minus15}
+	\end{subfigure}
+
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f1_minus5} 
+		\caption{Effect size = observed effect size - 5\%} \label{f1_minus5}
+	\end{subfigure}
+
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f1_plus5} 
+		\caption{Effect size = observed effect size + 5\%} \label{f1_plus5}
+	\end{subfigure}
+
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f1_plus15} 
+		\caption{Effect size = observed effect size + 15\%} \label{f1_plus15}
+	\end{subfigure}
+
+	\caption{Power analysis for fixed effects of interest at varying sample sizes: F1}
+
+\end{figure}
+
+\begin{figure}[!htb] \label{figure_power_analysis_f2}
+	\centering
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f2_minus15} 
+		\caption{Effect size = observed effect size - 15\%} \label{f2_minus15}
+	\end{subfigure}
+
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f2_minus5} 
+		\caption{Effect size = observed effect size - 5\%} \label{f2_minus5}
+	\end{subfigure}
+
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f2_plus5} 
+		\caption{Effect size = observed effect size + 5\%} \label{f2_plus5}
+	\end{subfigure}
+
+	\begin{subfigure}[t]{0.8\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{powerplot_f2_plus15} 
+		\caption{Effect size = observed effect size + 15\%} \label{f2_plus15}
+	\end{subfigure}
+
+	\caption{Power analysis for fixed effects of interest at varying sample sizes: F2}
+
+\end{figure}
+
+\newpage
+
+\clearpage
 \bibliography{references}

### 2022-10-01 draft_BLC/BLC_AM-ID_manuscript.tex
added supplementary material file
shortened discussion, results, introduction

@@
 
-The rest of this section is structured as follows: \ref{bengali_english_in_india} provides background information on Bengali and Indian English focusing on vowel systems.
-\ref{causes} discusses two potential sources of transfer during language processing: global co-activation of multiple languages vs.
-the local act of switching between languages.
-Given the highly multilingual setting that characterizes this population, we note that the former is expected to be a constant feature and therefore any observed phonetic interaction is better understood as resulting from the latter.
-The design of the present study is motivated on the basis of this discussion.
-A logical consequence of this model is that such phonetic interaction must be highly localized.
-\ref{duration} elaborates on this by reviewing literature on the duration and `reversibility' of transfer effects.
-A recurring pattern that emerges from existing research on short-term phonetic transfer is that both languages of the bilingual speaker are not equally affected, although the nature of these reported differences varies greatly across studies.
-In \ref{asymmetries}, we review some of these findings, focusing on proposed sources of asymmetries, in order to highlight the factors that mediate such phonetic interaction (or fail to do so).
-We argue that these results point to an urgent need for expanding the range of language pairs and sound categories examined in order to make meaningful generalizations about the nature of short-term phonetic transfer.
-Following from this, \ref{asymmetry between sounds} discusses reported asymmetries between different sound categories of a single language, and develops the hypotheses of the present study on the basis of these.
-In \ref{paradigms}, we motivate a secondary aim of this study: to verify the postulated differences between two paradigms for eliciting short-term phonetic interaction in an experimental setup, namely cued picture-naming and code-switching.
-Finally, we summarize the research questions and hypotheses.
-
 Existing studies have studied phonetic transfer along different acoustic features.
@@
 Note that ``Indian English" is not a monolithic entity; there are large regional variations in the English speech of Indian speakers, and the literature on IE phonology is thus neither ``standardized" nor uncontested.
-For the present study, we focus on a variety of IE spoken by an educated elite, used in news channels, often taught as a target in classrooms, and generally considered to be free of any particular regional affiliation \citep[c.f.][who reports this variety to be ``a \textit{de-facto} norm", and thus calls it General(ized) Indian English]{masica1972sound}.
+Recent research suggests that (see \cite{sirsa2013effects} for a review).
+IE has a target phonology that is distinct from the varied L1s of its speakers, as well as from other native varieties of English.
+For the present study, we focus on ``General(ized) Indian English" \citep{masica1972sound}.
 This sound system is described in section \ref{vowel systems}.
-However, recent research comparing the segmental and suprasegmental properties of IE spoken in different parts of the country also report many commonalities (see \cite{sirsa2013effects} for a review).
-It has thus been suggested that in spite of the varied L1s of its speakers, IE has a target phonology that is distinct from any of these, as well as from other native varieties of English.
-Thus, it is fruitful to think of regional variations as resulting from L1-influence on a common underlying target.
-Excepting a small minority of specific communities \citep{pandey201517, wells1982accents, coelho1997anglo}, there are very few L1 speakers of IE.
-Thus, English is largely learned and used as an L2.
-
-Considering the language usage patterns suggested by this demography, the present study focuses specifically on short-term phonetic interaction during mixed-language use, because:
+
+We focus specifically on short-term phonetic interaction during mixed-language use, because:
 
@@
 	\item In multilingual populations, long-term representations of sound categories are expected to be affected by multiple languages. 
-	\item Given that examples of an IE phonology without L1 `influence' are very rare, it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a given baseline condition.
+	\item Given the rarity of IE phonology without L1 `influence' (it is largely spoken as an L2) it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a baseline condition.
 \end{enumerate}
 
-The socio-linguistic facts about English in India also suggest that, for the present population, mixed-language processing of English and an L1 is an ecologically valid paradigm that is likely to be a part of everyday language experience.
-Thus, data from this population is valuable to the understanding of short-term phonetic interaction.
+The socio-linguistic facts about English in India also suggest that, make mixed-language processing of English and an L1 an ecologically valid paradigm to understand short-term phonetic interaction.
 
@@
 What triggers this interaction?
-\cite{olson2016role} argues that while cross-language phonetic effects are largely measured at the point of language switch, it could have two potential sources:
-\begin{enumerate}
-	\item The local point of switch itself
-	\item Global co-activation of two languages: \textsc{bilingual language mode} \citep{grosjean1998studying} 
-\end{enumerate}
+\cite{olson2016role} argues that cross-language phonetic effects could have two potential sources: (i) the local point of switch, and (ii) global co-activation of two languages: \textsc{bilingual language mode} \citep{grosjean1998studying}.
 
@@
 
-\cite{grosjean1998studying} suggests that various aspects of the communicative setting, including exposure to (spoken or written) stimuli in multiple languages and awareness of the interlocutor's being bilingual, could trigger a bilingual language mode.
-The participants in the present study are in an environment which largely contains mixed-language input, multilingual interlocutors, and no stable ambient language.
-Therefore we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode.
-To preserve ecological validity, we elicited both switched and non-switched utterances in a bilingual language mode.
+Since participants of this study are in an environment that largely contains mixed-language input and multilingual interlocutors, we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode \citep[c.f.][]{grosjean1998studying}.
+Thus, we elicited all utterances in a bilingual language mode.
 Any observed differences in this paradigm would arguably result from interaction during online processing of sounds while switching between languages.
@@
 Note that all the studies of mixed-language production discussed above measure VOT, which is a temporal feature.
-We have no apriori reason to assume that these durations generalize to vowel quality.
+We have no \textit{a priori} reason to assume that these durations generalize to vowel quality.
 However, findings from sub-categorical phonetic shifts triggered by other factors (such as convergence towards an interlocutor) do evidence rapid shifts in vowel quality within comparable time-frames \citep[e.g.][]{pardo2010expressing,babel2010dialect,babel2012evidence}.
-Thus in the current study, we present switched and non-switched tokens randomly within the same test block to induce a bilingual mode.
+In this study, we present switched and non-switched tokens randomly within the same test block to induce a bilingual mode.
 We expect the intervening words between two subsequent targets to undo any residual effects of transfer.
@@
 
-\subsubsection*{Differences between sound systems of the languages} \label{sound systems} In addition to the cognitive factors discussed above, many researchers have attributed the observed asymmetries to language-internal factors.
-Specifically, a pattern that emerges from the body of work on VOT is that a shift in long-lag VOT towards the short-lag norms of the other language is much more consistent and systematic than the converse \citep{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid}.
-In general, short-lag VOTs seem to resist accommodative shift.
-\cite{bullock2009trying} suggest that because long-lag languages offer a greater range of acceptable VOTs, there is more `room' for movement.
-In contrast, shift in short-lag languages could risk the loss of a phonological contrast (VOT being the primary cue for voicing contrast in these languages).
-This is a language-specific phonological constraint on transfer.
-In a similar vein, \cite{antoniou2011inter} observe that while the VOT of English stops shifted towards Greek (a short-lag language), the reverse was seen only in a specific subset of sounds: word-medial non-nasalized stops.
-They attribute this to the low frequency of these sounds in Greek, making them less ``stable", and thus more amenable to cross-language influence.
-This suggests that the distribution of sounds in a language could affect the kind of transfer patterns observed.
-
-These results highlight that beyond interaction during the cognitive processing of sounds, phonetic transfer is ultimately a linguistic phenomenon, and thus subject to language-specific phonological constraints.
-Therefore, to make meaningful generalizations, it is imperative to consider data from a variety of language pairs.
-Existing research on phonetic transfer largely revolves around a limited set of phonologically related languages.
+\subsubsection*{Differences between sound systems of the languages} \label{sound systems} In addition to the cognitive factors discussed above, many researchers have attributed asymmetries to language-internal factors, such as the shape, size, and composition of the sound inventory.
+
+For example, the frequently reported asymmetries between long- and short-lag VOT languages \citep[see, e.g.][]{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid} and postulated frequency effects \citep{antoniou2011inter} highlight that beyond interaction during the cognitive processing of sounds, phonetic transfer is ultimately a linguistic phenomenon, subject to language-specific phonological constraints.
+Therefore, it is imperative to consider data from a variety of language pairs.
+Existing research on phonetic transfer largely focuses on a limited set of phonologically related languages.
 Thus, the present study extends the scope of this research to a new pair of languages--- Indian English and Bengali.
-
-Since Bengali has a four-way laryngeal contrast, VOT does not encode a binary voiced-voiceless distinction.
-Moreover, VOT is not the only relevant cue for the corresponding phonological categories in the language \citep{dmitrieva2020acoustic}.
-This means that any effect of L1 Bengali on L2 English VOT is likely to be affected by additional phonological factors.
-Thus, avoiding any potential confounds was another reason we chose to focus on vowel quality as a target for transfer in the present study.
 
@@
 Together, these incorporate questions about language history, usage, attitudes, and switching experience.
+Table \ref{table_participants} summarizes some key biographical data from the survey.
 \alert{The complete questionnaire, along with responses, is available in the ``Data" component of the OSF repository for this project (c.f.
 section \ref{data_availability}).
---footnotes?}
+A more detailed discussion of the linguistic situation and language profiles is included as supplementary material.--footnote?}
 
@@
 
-\subsubsection*{Random effects structure}
-
-Subject and Item were treated as random factors.
-Given our hypotheses, we included random intercepts for Subject and Item, by-subject random slopes for Vowel (formant targets for vowel categories differ across individuals) and Context (individuals differ in their response to the experimental condition), and a by-item random slope for Context (the effect of experimental conditions differ across words).
-
-We tested the significance of our hypothesized explanatory variables using t-values and p-values for individual variables in the models, and tested overall model fit by comparing the full model to a corresponding null model that lacks the variable of interest using ANOVAs.
-The following subsections summarize and discuss the model outputs for F1 and F2 respectively.
-
 \subsection{Vowel Height -- F1}
-
-\subsubsection*{Model output summary}
-
-First, to test our first and second research questions (\bf{L2 vowels and Asymmetry}), we fitted a model with vowel*time, and vowel*context as fixed effects.
-The transfer term (vowel*context) is significant in this model (model coefficient/ unstandardized effect size ($\beta$)= -2.330e-01, standard error (SE)=  5.244e-02, t= -4.44, p=0.0002).
-Comparing this to a null model that lacks the transfer term shows that the full model is significantly better ($\chi^2(2) = 14.42, p=  0.0007$).
-To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction.
-The full model was better ($\chi^2(1) = 12.51, p= 0.0004$).
-
-To examine the third research question, \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects.
-This three-way interaction term is significant ($\beta$= -1.955e-01, SE=  5.453e-02, t= -3.59, p= 0.0003).
-We also find a significant effect of task on FFs ($\beta$=3.300e-01, SE=2.668e-02, t=12.371, p=2e-16).
-Comparing this model to a null model that lacks the fixed effect Task ($\chi^2(4) = 572.61, p=  2.2e-16$) shows that the full model is better.
-Because Task has an independent effect on F1, and adding interactions increases model parameters, we fitted a second null model with Task as a separate fixed effect without the interaction term to test whether the interaction term is needed.
-However, the full model was still a better fit for the data ($\chi^2(3) = 23.82, p=  2.722e-05$).
-Next, we tested two other null models to verify that the full model with the three-way interaction term is indeed the optimal model for the data, by replacing the three-way interaction term with: (i) vowel*context + task*context (to test if task only interacts with context, rather than the transfer term vowel*context).
-The full model is better ($\chi^2(2) = 23.28, p= 8.78e-06$); (ii) vowel*context + task*vowel (to test if task only interacts with vowel, rather than the transfer term vowel*context).
-The full model is better ($\chi^2(2) = 13.36, p= 0.001$).
-This shows that the three-way interaction is the best fit for the data.
-Because an independent effect of task on FFs is unexpected (c.f.
-model interpretation below), we wanted to examine if the effect might be local, i.e.
-decrease over time.
-We added a task*time interaction term to the full model, which is significant ($\beta$= -1.216e-01, SE=  1.347e-02, t= -9.028, p= 2e-16 ) and improves model fit ($\chi^2(1) = 81.22, p= 2.2e-16$).
-
-Next, to examine the dynamics of transfer, we added a three-way interaction between the transfer term (vowel*context) and time.
-This term has a marginally significant effect ($\beta$= -4.604e-02, SE=  2.692e-02, t= -1.71, p= 0.08), but does not significantly improve model fit compared to a null model lacking the vowel*context*time term ($\chi^2(2) = 3.41, p= 0.18$).
 
@@
 \subsubsection*{Model interpretation}
-We extracted formant estimates from five uniformly spaced time points in the vowel: from the time at 5\% into the vowel (at the start of articulation) to 45\% (at the mid-part, or steady state, which is expected to correspond most closely to the articulatory target for the category).
-As expected, we see that F1 values are significantly predicted by Time in the model, reflecting the dynamic nature of articulation which causes acoustic variation over the course of time.
+We see that F1 values are significantly predicted by Time in the model, reflecting the dynamic nature of articulation which causes acoustic variation over the course of time.
 The positive slope of the estimate for Time indicates that F1 increases, i.e.
@@
 
-Vowel is not a significant predictor of F1.
-This is initially surprising, given that the phonological category is expected to systematically affect vowel quality.
-However, a boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two IE categories \nt{2, \ae} and the extent of height difference between them.
-Note that these measures are only from the nonswitch condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching.
-Given that estimates for fixed effects are pooled across participants, this explains the lack of reliable differences between \nt{2} and \nt{\ae}.
+A boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two IE categories \nt{2, \ae} and the extent of height difference between them.
+These measures are only from the nonswitch condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching.
 
@@
 Put more intuitively, vowels are lowered in the mixed condition.
-However, the lack significance indicates that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels.
-Instead, the significant interaction between Vowel and Context signals that the extent and direction of shift is moderated by vowel category.
-There is an asymmetrical shift in the categories \nt{2} and \nt{\ae} in the mixed-language condition.
+However, the lack of significance indicates that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels.
+There is an asymmetrical shift in the categories \nt{2} and \nt{\ae} in the mixed-language condition, as shown by the significant interaction between Vowel and Context.
 Specifically, the Vowel*Context interaction has a negative slope, showing that the lowering is greater in \nt{2}.
-The lack of a significant Vowel*Context*Time interaction shows that transfer patterns do not change across time.
-This indicates a uniform scaling of formant trajectories in the mixed condition, suggesting that the underlying acoustic targets for these vowels are different during non-switched vs mixed utterances.
-
-The results also reveal a significant effect of Task, such that the sentence task (code-switching) leads to higher F1 values in both vowels relative to the cued picture-naming task as: vowels are lowered in the code-switching task, independent of experimental condition.
-This is surprising, as we have no aproiri reason to expect a uniform difference in vowel quality across tasks.
-A significant interaction with Time with a positive slope of the estimate shows that this effect is local: it is strongest at the beginning of the vowel, and reduces over the course of the vowel.
-This is consistent with the interpretation that the unexpected Task effect is an artifact of the paradigm, possibly resulting from coarticulatory effects owing to the specific stimuli used in this study.
-Of greater interest to us is the significant Vowel*Context*Task interaction, which indicates a difference in transfer patterns across paradigms.
-Specifically, the negative slope of the estimate suggests that transfer in the F1 (height) dimension are greater in the code-switching (s) task compared to the picture-naming (p) task.\\
+The Vowel*Context*Time interaction is not significant: transfer patterns do not change across time.
+
+Independent of experimental condition, there is a significant effect of Task: the sentence task (code-switching) leads to higher F1 values in both vowels relative to the cued picture-naming task.
+This is surprising, as we have no \textit{a priori} reason to expect a uniform difference in vowel quality across tasks.
+A significant interaction with Time is strongest at the beginning of the vowel, indicatin a local effect.
+This is consistent with the interpretation that the unexpected Task effect is an artifact of the paradigm.
+A significant Vowel*Context*Task interaction indicates a difference in transfer patterns such that transfer in the F1 (height) dimension is greater in the code-switching (s) task compared to the picture-naming (p) task.\\
 
@@
 \subsection{Vowel backness -- F2}
-
-\subsubsection*{Model output summary}
-
-First, to test RQ1 (\textbf{L2 transfer}) and RQ2(\textbf{Asymmetry}), we fitted a model with vowel*time, and vowel*context as fixed effects.
-The transfer term (vowel*context) is significant in this model (standardized effect size ($\beta$)= 1.234e-01, standardized error (SE)=  2.055e-02, t= 6.00, p=5.46e-06).
-Comparing this to a null model that lacks the transfer term shows that the full model is significantly better ($\chi^2(2) = 16.20, p= 0.0003$).
-To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction.
-The full model was better ($\chi^2(1) = 16.02, p= 6.245e-05$).
-
-To test RQ3 (\textbf{Paradigm}), we fitted a model with vowel*time and task*vowel*context as fixed effects.
-While task has an independent effect on F2 ($\beta$= -8.758e-02, SE= 1.756e-02, t= -4.98, p= 6.18e-07), the three-way interaction term is not significant ($\beta$= 3.349e-02, SE= 3.565e-02, t= 0.94, p= 0.34).
-However, there are significant two-way interactions between task*context ($\beta$= -1.099e-01, SE= 2.483e-02, t= -4.427, p= 9.66e-06) and task*vowel ($\beta$= -8.983e-02, SE= 2.535e-02, t= -3.544, p= 0.000396).
-Comparing this to a null model that lacks the fixed effect Task confirms that the full model is still a better fit for the data ($\chi^2(4) = 572.61, p=  2.2e-16$), indicating that an optimal model should contain the term Task.
-To confirm whether Task interacts with any of the other terms, we fitted a model with Task as a separate fixed effect without the interaction term.
-Comparing this to the full (interaction) model shows that the latter is still a better fit ($\chi^2(3) = 44.8, p= 1.02e-09$).
-Next, we tested two other null models to verify that the three-way interaction term is indeed optimal for the data, by replacing the three-way interaction term with: (i) vowel*context + task*context (to test if task only interacts with context, rather than the transfer term vowel*context).
-The full model is better ($\chi^2(2) = 17.28, p= 0.0001$); (ii) vowel*context + task*vowel (to test if task only interacts with vowel, rather than the transfer term vowel*context).
-The full model is better ($\chi^2(2) = 28.45, p= 6.633e-07$).
-This shows that the three-way interaction is the best fit for the data.
-
-Because an independent effect of task on FFs is unexpected (c.f.
-model interpretation below), we wanted to see if the effect might be local, i.e.
-decrease over time.
-We added a task*time interaction term to the full model, which is significant ($\beta$= 6.237e-02, SE=  8.876e-03, t= 7.027, p= 2.23e-12 ) and improves model fit ($\chi^2(1) = 49.271, p= 2.23e-12$).
-
-Next, to examine the dynamics of transfer, we added a three-way interaction between the transfer term (vowel*context) and time.
-This term is not significant ($\beta$= -6.274e-03, SE= 1.774e-02, t= -0.35, p= 0.72), and the addition does not significantly improve model fit compared to a null model lacking the vowel*context*time term ($\chi^2(2) = 2.2443, p= 0.32$).
 
@@
 
-Taken as a whole, these results bear on our original research questions in the following ways: regarding our first question of whether L2 vowels of proficient bilinguals can be affected by short-term transfer from L1 (\textbf{L2 vowels}), our results indicate that this indeed happens during mixed-language use.
-Both the common category \nt{\ae} and the L2-exclusive category \nt{2} show spectral shifts during mixed-language productions compared to non-switched productions.
-As hypothesized (c.f.
-RQ2 and H2, \textbf{Asymmetry}), the results reveal asymmetries in the extent and direction of shift between the two vowels.
-Specifically, \nt{2} is lowered and fronted, while \nt{\ae} is fronted, with the extent of shift being greater in \nt{2}.
-Comparison with canonical Bengali vowel productions shows that the shift in \nt{\ae} is towards the corresponding Bengali category \nt{\ae}, while that in \nt{2} is towards the Bengali category \nt{a:}.
-Regarding our third research question of whether the paradigm used to elicit language switching independently affects patterns of transfer (RQ3: \textbf{Paradigm}), analyzing the role of Task in the results confirms this.
-We find a greater degree of transfer during code-switching compared to the cued picture-naming task.
-However, this difference is limited to the F1 dimension, and needs to be replicated.
-These results and their implications are discussed in the following subsections.
-
 \subsection{Phonetic transfer in L2 vowels}
@@
 \paragraph{\textbf{What causes transfer: language mode vs switching}}
-An often discussed issue in transfer research is whether cross-language interaction at the phonetic level during language mixing is triggered by the co-activation of languages that makes representations from both languages available, or by the local act of switching during online processing and articulation of those sounds.
 In this study, we presented nonswitched English and code-switched Bengali-English stimuli in the same session, such that the language mode (participants' global language ``setting") remained the same throughout.
@@
 \paragraph{Differences between L1 vs L2 transfer:}
-Studies of short-term phonetic transfer consistently find differences between transfer patterns in L1 vs L2.
-The bulk of this literature is based on the feature VOT.
-While vowels have been shown to induce L2-to-L1 influence, the present study demonstrates L1-to-L2 transfer in vowels.
-
 Previous studies of language-switching have often reported phonetic transfer from L2 to L1, but a lack of parallel transfer from L1 to L2.
-One postulated model that is consistent with this finding is the Inhibitory Control Model (ICM), first proposed by \cite{green1986control} at the lexical level, but extended to sound systems by \cite{olson2013bilingual,tsui2019impact}.
+One postulated model that is consistent with this finding is the Inhibitory Control Model (ICM) \cite{green1986control,olson2013bilingual,tsui2019impact}.
 This proposes that the lack of transfer effects on L2 reflects the lower inhibition required on L2 representations in the first place, which makes switching into L2 easier, compared to the converse situation, when the high initial inhibition on L1 induces phonetic transfer while switching back into L1.

### 2022-10-08 draft_BLC/BLC_AM-ID_manuscript.tex
shortening

@@
 
-Note that ``Indian English" is not a monolithic entity; there are large regional variations in the English speech of Indian speakers, and the literature on IE phonology is thus neither ``standardized" nor uncontested.
-Recent research suggests that (see \cite{sirsa2013effects} for a review).
-IE has a target phonology that is distinct from the varied L1s of its speakers, as well as from other native varieties of English.
-For the present study, we focus on ``General(ized) Indian English" \citep{masica1972sound}.
-This sound system is described in section \ref{vowel systems}.
+``Indian English" is not a monolithic entity; there are large regional variations in the English speech of Indian speakers, and the literature on IE phonology is thus neither ``standardized" nor uncontested.
+Recent research suggests that IE has a target phonology that is distinct from the varied L1s of its speakers, as well as from other native varieties of English (see \cite{sirsa2013effects} for a review).
+For the present study, we focus on ``General(ized) Indian English" \citep{masica1972sound} (\ref{vowel systems}).
 
@@
 	\item In multilingual populations, long-term representations of sound categories are expected to be affected by multiple languages. 
-	\item Given the rarity of IE phonology without L1 `influence' (it is largely spoken as an L2) it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a baseline condition.
+	\item Given the rarity of IE phonology without L1 `influence' (it is almost exclusively spoken as an L2) it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a baseline condition.
 \end{enumerate}
 
-The socio-linguistic facts about English in India also suggest that, make mixed-language processing of English and an L1 an ecologically valid paradigm to understand short-term phonetic interaction.
+The socio-linguistic facts about English in India also suggest that mixed-language processing of English and an L1 an ecologically valid paradigm to understand short-term phonetic interaction.
 
 \subsubsection*{Vowel systems} \label{vowel systems}
-The vowel inventory of Western Bangla (the variety spoken by the participants in this study) consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \citep{garry2001facts}.
+The vowel inventory of Western Bangla consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \citep{garry2001facts}.
 The vowel chart in figure \ref{vowels_bengali} shows their distribution in the F1$\times$F2 formant space.
@@
 Since participants of this study are in an environment that largely contains mixed-language input and multilingual interlocutors, we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode \citep[c.f.][]{grosjean1998studying}.
-Thus, we elicited all utterances in a bilingual language mode.
+Thus, we elicited all utterances in the same bilingual test block.
 Any observed differences in this paradigm would arguably result from interaction during online processing of sounds while switching between languages.
@@
 
-Research comparing bilingual speakers' phonological systems with monolinguals treats them as relatively stable over time.
-The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods \citep[e.g.][]{guion2003vowel, caramazza1973acquisition}.
-However, longitudinal studies %(citations)
-and between-subject comparisons of bilingual speakers with different durations of L2 exposure suggest that interaction between the sound systems is dynamic: over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts.
-For example, \cite[][vowel quality]{bohn1992production}
-observed that approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language influence.
-These have been interpreted as changes to long-term phonological representations.
-
+Longitudinal studies and between-subject comparisons of speakers with different durations of L2 exposure suggest that interaction between bilingual sound systems is dynamic \citep[e.g.][]{bohn1992production}.
 Moreover, changes due to transfer are not necessarily unidirectional or irreversible.
-\cite[][VOT]{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language.
-This not only evidenced that cross-language interaction can be triggered in an order of months, but also that the effects can be reversed within a similar time range.
+\cite[][VOT]{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language, showing that cross-language interaction can be triggered in an order of months, and also reversed within a similar time range.
 A more recent study by \cite[][VOT]{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks).
@@
 How long these effects last in the absence of regular L2 input was not tested.
-% if effects are cumulative, then the difference between switched and non-switched tokes should be lesser in later test blocks (previous two paragraphs needed?)
 
@@
 \subsubsection*{L1 vs. L2 status} 
-Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} posits that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another.
+Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} proposes that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another.
 Thus, the patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories.
-In non-switched production, this is evidenced through a pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf.
+In non-switched production, this is evidenced through a L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf.
 \textsc{equivalence classification}: \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.
@@
 They interpret this asymmetry in terms of the Inhibitory Control Model (ICM) \citep{green1998mental}: to select a phonetic realization from one language, the other must be \textsc{inhibited}.
-Since L1 is the `stronger' language, the amount of inhibition required on L1 while using an L2 is greater than the inhibition required on an L2 while using the L1.
+Greater inhibition is required on L1 while using an L2 than the reverse.
 This greater initial inhibition on L1 means that switching into L1 incurs a greater \textsc{switch cost} (more cross-language influence) than switching into L2.
@@
 To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we processed and used existing speech data from a freely available Bengali corpus, SHRUTI \citep{shruticorpus}.
-The procedure is described below:
+We elaborate on the corpus details and procedure in the Supplementary materials.
 
-\subsubsection*{Baseline Bengali data from the SHRUTI corpus}
-
-The SHRUTI corpus is the largest free corpus of Western Bengali speech that we are aware of.
-It comprises of 21.64 hours of read speech, containing 7383 unique sentences (22012 unique words) from Bengali newspapers in different domains, covering the most frequently used words of the language.
-The speakers are 34 adults (age range 20-40, 75\% male, 25\% female) who have grown up in the Indian state of West Bengal, and speak standard colloquial Bengali.
-All of the speakers have university-level education or higher.
-Given these parallels in linguistic, regional, and socio-economic background, we believe that Bengali speech from this corpus is representative of the average target Bengali vowel categories that are of interest in this study.
-To estimate average vowel positions, we extracted formant values from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, to get estimates from the steady state, which should most closely approximate the production target.
-Next, we compared these to our participants' productions in the non-switched English condition to understand the relative positions of these categories.
+We extracted formant values from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, and compared these to our participants' productions in the non-switched English condition.
 This is visualized in figure \ref{vowels_unilingual_ie_bengali}.
-
-Note that we refrained from using corpus data for average Indian English vowel positions, opting instead to use the participant data from this study, because IE has high regional variability (c.f.
-discussion in section \ref{bengali_english_in_india}).
-To the best of our knowledge, there is no freely available corpus of IE speech from Bengali speakers in India.
-
 The F1$\times$F2 plot indicates that the Bengali category \nt{\ae} is more fronted, and \nt{a:} lower and more fronted, compared to the English categories \nt{\ae} and \nt{2} respectively.
@@
 Although existing studies in the field have used comparable participant numbers, we acknowledge that this is a small sample size.
-As detailed above, even after a recruitment call that already controlled for language background and education level (recruiting Bengali-English bilinguals in a university campus), we were able to include only 10 out of 28 volunteers (35.7\%) in the final study.
-Note that the criteria used here establish only a minimal level of between-participant similarity in exposure, dominance, habits, and proficiency, such as is extremely common in most comparable studies.
+As detailed above, even after a recruitment call that already controlled for language background and education level, we were able to include only 10 out of 28 volunteers (35.7\%) in the final study.
+The criteria used here establish only a minimal level of between-participant similarity in exposure, dominance, habits, and proficiency, such as is extremely common in most comparable studies.
 This demonstrates the degree of linguistic variability in the present population.
@@
 In settings where linguistic variability at the population level is much greater than these canonically studied groups, establishing comparable levels of experimental control requires significantly greater resources.
-Given that small sample sizes are likely to have low statistical power, replication with larger sample sizes are important to confirm the observed effects, and provide much-needed data diversity (c.f.
-discussion in section \ref{introduction}).
+Given that small sample sizes are likely to have low statistical power, replication with larger sample sizes are important to confirm the observed effects, and provide much-needed data diversity.
 A more detailed discussion of power analyses in bilingualism research is included as supplementary material on OSF (c.f.
@@
 Our study involved two IE vowels.
-Vowel category is expected to systematically affect formant frequencies (FFs).
-Thus, Vowel is a fixed effect in the models.
-The experimental treatment of interest is Context, i.e.
+Vowel is a fixed effect, since it is expected to systematically affect formant frequencies (FFs).
+The experimental treatment is Context, i.e.
 whether the English target item was presented in an English (nonswitch) or Bengali (switch/mixed) context.
-Our main research question (\bf{L2 vowels}) asks whether L2 vowel quality (formant frequencies) of proficient bilinguals is affected by the Context of utterance.
+Our first research question (\bf{RQ1---L2 vowels}) asks whether L2 vowel quality of proficient bilinguals is affected by the Context of utterance.
 We hypothesize that the direction of such shift in the mixed condition, if any, is towards a corresponding L1 category.
 Therefore, we expect the effect of Context on formant frequencies, i.e.
-the direction of shift, to be moderated by the vowel category (c.f.
-the discussion in Section \ref{asymmetry between sounds}, leading to our second research question \textbf{Asymmetry}).
+the direction of shift, to be moderated by the vowel category (see Section \ref{asymmetry between sounds}, leading to our second research question \textbf{RQ2---Asymmetry}).
 Thus, the main effect of interest is the interaction term Context*Vowel.
 This corresponds to the transfer pattern.
-Our third research question (\bf{Paradigm}) asks whether the language-mixing paradigm, or Task, affects the extent and direction of transfer.
+Our third research question (\bf{RQ3---Paradigm}) asks whether the language-mixing paradigm, or Task, affects the extent and direction of transfer.
 If this is the case, we expect Task to interact with the transfer term, giving a three way Task*Context*Vowel interaction.
@@
 
-Thus, the optimal model was: f1 $\sim$ time*vowel + vowel*context*task + time*task + (1 + context|word) + (1 + context + vowel|subject).
+The sequence of models fitted can be found in the Supplementary materials.
+The optimal model was: f1 $\sim$ time*vowel + vowel*context*task + time*task + (1 + context|word) + (1 + context + vowel|subject).
 Table \ref{table_f1} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model.
@@
 This is surprising, as we have no \textit{a priori} reason to expect a uniform difference in vowel quality across tasks.
-A significant interaction with Time is strongest at the beginning of the vowel, indicatin a local effect.
-This is consistent with the interpretation that the unexpected Task effect is an artifact of the paradigm.
+A significant interaction with Time is strongest at the beginning of the vowel, indicating a local effect.
 A significant Vowel*Context*Task interaction indicates a difference in transfer patterns such that transfer in the F1 (height) dimension is greater in the code-switching (s) task compared to the picture-naming (p) task.\\
@@
 \subsection{Vowel backness -- F2}
-
-Thus, the optimal model is: f2 $\sim$ vowel*time + vowel*context*task + task*time + (1+context|word) + (1+context+vowel|subject).
+The optimal model was: f2 $\sim$ vowel*time + vowel*context*task + task*time + (1+context|word) + (1+context+vowel|subject).
 Table \ref{table_f2} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model.
@@
 
-As with F1, we find that Time significantly influences F2 values in the model, reflecting the dynamic nature of speech which causes acoustic variation over the course of articulation.
-Moreover, the Time*Vowel interaction shows that this trajectory differs across the two vowels: while the positive slope for Time indicates that F2 increases i.e.
-vowels are fronted with time, the negative slope of the Time*Vowel interaction shows that this effect is smaller for \nt{2}.
-This is expected, as figure \ref{vowels_unilingual_ie_bengali} confirms that the category \nt{\ae} is overall more fronted than \nt{2}.
-
+As with F1, we find that Time significantly influences F2 values as expected.
 Unlike in the F1 dimension, Vowel is a significant predictor of F2, with \nt{2} having significantly lower F2, i.e.
 being produced as more backed, than \nt{\ae}.
-As with vowel height, there is no significant independent effect of Context, showing that mixed language use does not lead to uniform fronting or backing across vowels.
+There is no significant independent effect of Context, showing that mixed language use does not lead to uniform fronting or backing across vowels.
 This is not surprising, as we expect movement to be towards the corresponding L1 category, and thus differ across vowels.
-The negative slope of the estimate indicates that on average, vowels have lower F2 i.e.
-are backed, in the nonswitch (e) condition.
-In other words, vowels are fronted in the mixed condition.
-This effect is greater in \nt{\ae}, as indicated by the negative slope of the Vowel*Context interaction.
-In the F2 dimension, too, the lack of a significant Vowel*Context*Time interaction shows that transfer patterns do not change across time, suggesting that the underlying acoustic targets are different in the two conditions.
+The negative slope of the estimate indicates that on average, vowels are fronted in the mixed condition.
+This effect is greater in \nt{\ae}.
+The lack of a significant Vowel*Context*Time interaction shows that transfer patterns do not change across time, suggesting that the underlying acoustic targets are different in the two conditions.
 
-We see a significant effect of Task on F2, such that vowels in the code-switching (s) task have lower F2, i.e.
+We see a significant effect of Task on F2: vowels in the code-switching (s) task have lower F2, i.e.
 are more backed than those in the picture-naming (p) task.
-As with F1, this is an unexpected result; we have no principled reason to expect code-switching to induce an across-board backing of vowels.
-Vowels are backed in the code-switching task, independent of experimental condition.
-A significant interaction with Time, with a positive slope for the estimate, shows that this effect is local: it is strongest at the beginning of the vowel, and reduces over the course of the vowel.
-This aligns with the possibility that the unexpected Task effect is an artifact of the paradigm and the particular stimuli used.
+As with F1, this is an unexpected result.
+The significant interaction with Time shows that this effect is local and decreases over time.
 The Vowel*Context*Task interaction is of interest, as it relates to our research question of how the test paradigm affects transfer.
@@
 Statistical analyses reveal that the vowel \nt{2} is lowered and fronted, while \nt{\ae} is fronted, in the mixed condition.
-Based on previous literature, we expect that such shifts in vowel categories result from increased cross-language influence during mixed-language use, i.e.
-shifts are targeted with respect to L1 categories.
 As discussed in section \ref{asymmetry between sounds}, although our primary research question is about whether productions in switch vs nonswitch conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories.
@@
 
-The main aim of this study was to examine short-term phonetic transfer in L2, towards understanding language processing in proficient bilinguals.
+The aim of this study was to examine short-term phonetic transfer in L2, towards understanding language processing in proficient bilinguals.
 Two interrelated questions, discussed in section \ref{introduction}, bear on this: (i) what causes transfer in mixed-language speech; and (ii) what causes observed asymmetries between transfer in L1 vs L2.
-Together, these serve as the basis for our theories about the language systems of multilingual individuals exist in relation to one another, and consequently, what cognitive models of language should look like.
 
 \paragraph{\textbf{What causes transfer: language mode vs switching}}
-In this study, we presented nonswitched English and code-switched Bengali-English stimuli in the same session, such that the language mode (participants' global language ``setting") remained the same throughout.
-In spite of this, results demonstrate consistent differences between switched and non-switched tokens.
+In this study, we presented nonswitched English and code-switched Bengali-English stimuli in the same session and found consistent differences between switched and non-switched tokens.
 This confirms that the local act of switching does trigger transfer, even in a bilingual language mode, paralleling results from \cite{olson2016role, tsui2019impact, piccinini2015voice}.
-These studies all examined the effect of switching on a temporal feature of speech, namely voice onset time or VOT.
-The results of the present study indicate that the same happens in vowels.
-Moreover, the exact effect of transfer is interesting: VOT is a one-dimensional measure.
-Thus, observing an increase or decrease in VOT in a mixed language condition makes it difficult to infer whether a consonant has moved towards a specific L1/L2 counterpart, or reflects a more general system-wide shift in articulatory norms.
-Since vowel quality measures show directionality as well as magnitude, we can observe finer details in the patterns of shift.
-In the present study, we find that productions of English vowels do not merely shift towards a Bengali counterpart, but rather towards a set of Bengali-like \textit{norms}: a vowel category that doesn't exist in Bengali nonetheless moves towards a related (perceptually close) Bengali category.
-This result parallels \cite{simonet2014phonetic}, who found the Catalan \nt{O} of bilingual speakers shifts towards a related Spanish category, \nt{o}, in a bilingual language mode (note that they did not examine code-switched speech, and focused on vowel height as a target of transfer).
-The study showed that co-activating multiple languages within a discourse causes a transient increase in cross-language phonetic interaction, leading to changes in both L1 and L2 (moderated by language dominance).
-This is interpreted as a follows: a bilingual mode temporarily co-activates both language systems, making both L1 and L2 representations available during processing.
-These representations interact while producing speech in such a language mode, leading to cross-language phonetic transfer.
-But what does this interaction look like, and how are these abstract representations processed during the act of producing speech?
-Examining speech during switching can shed light on this.
+These studies focused on a temporal feature, VOT.
+Our results indicate that the same happens in vowels.
+The exact effect of transfer is interesting: since VOT is a one-dimensional measure, observing a change in VOT in a mixed language condition makes it difficult to infer whether a consonant has moved towards a specific L1/L2 counterpart, or reflects a more general system-wide shift in articulatory norms.
+Vowel quality measures, showing both directionality and magnitude, facilitate this.
+We find that productions of English vowels do not merely shift towards a Bengali counterpart, but rather towards a set of Bengali-like \textit{norms}: a vowel category that doesn't exist in Bengali nonetheless moves towards a related (perceptually close) Bengali category.
+This parallels results from \cite{simonet2014phonetic} for Catalan and Spanish (they examined language mode, without code-switching).
+The study showed that co-activating multiple languages within a discourse causes a transient increase in cross-language phonetic interaction, leading to changes in both L1 and L2.
+This is interpreted as a follows: a bilingual mode temporarily co-activates both language systems, making both L1 and L2 representations available during processing, whose interactions lead to cross-language transfer.
+But what does this interaction look like, and how are these abstract representations processed during speech?
+Examining transfer during switching can provide insights.
 
-The time course of such shifts provides an insight into this.
-Vowels in the mixed condition do not start out the same, eventually deviating in acoustics from their nonswitch counterparts.
-Instead, we see a difference between the conditions right from the start of articulation.
+The time course of such shifts is telling: we see a difference between the conditions right from the start of articulation, suggesting different acoustic targets.
 This suggests that such interaction involves the speech planning stage.
-The pattern of transfer observed in this study, particularly for \nt{2}, also demonstrates that beyond a simple L2-L1 mapping, transfer during online processing there makes reference to entities beyond the level of the individual sound.
-Specifically, such transfer demonstrates a system-level interaction: a sound that does not exist in the L1 repertoire nonetheless acquires a different target, and the target is related to the L1 system in a meaningful way.
+The pattern of transfer observed in this study, particularly for \nt{2}, also demonstrates that beyond a simple L2-L1 mapping, transfer during online processing makes reference to entities beyond the level of the individual sound, demonstrating system-level interaction: a sound that does not exist in the L1 repertoire nonetheless acquires a different target, and the target is related to the L1 system in a meaningful way.
 This cannot be reasonably explained as articulatory inertia carried over from uttering L1 sounds.
@@
 If representations are combined in creative ways during online speech processing within an utterance, then can we think of intermediate representations not just for individual sounds, but rather for sound systems?
-While more detailed empirical data will eventually enable us to clarify such questions, this approach aligns with sociophonetic understandings of bilingual phonetic variation, for example as discussed by \cite{khattab2009phonetic,khattab2013phonetic} based on the use of phonological variables by Arabic-English bilingual children.
-This visualizes the bilingual phonetic system as ``a rich store of phonological representations", comprising of both language-internal and cross-language variation, from which the bilingual speaker is constantly choosing, guided by a variety of pressures.
+While more detailed empirical data is needed to answer such questions, this approach aligns with sociophonetic understandings of bilingual phonetic variation, for example as discussed by \cite{khattab2009phonetic,khattab2013phonetic} based on the use of phonological variables by Arabic-English bilingual children.
+She visualizes the bilingual phonetic system as ``a rich store of phonological representations", comprising of both language-internal and cross-language variation, from which the bilingual speaker is constantly choosing, guided by a variety of pressures.
 We propose that vowels are a good feature to examine the plausibility of such systems as cognitive models for language processing: in addition to continuous measures (unlike phonological variables), they also provide two-dimensional data (unlike VOT) that is suited to further test and refine predictions from such a model.
@@
 \subsection{Asymmetries in Phonetic Transfer}
-The literature on short-term phonetic transfer has frequently noted differences in transfer patterns between different sound categories in a language.
-This has been attributed to both general linguistic principles such as contrast preservation, and language-specific factors such as frequency and distribution of categories.
-Examining these asymmetries more closely, and understanding their sources, helps us figure out what factors affect/moderate transfer and thus, by extension, what level of representation and processing this interaction happens at, and the nature of short-term intra-speaker variation.
-The present study examined transfer in two L2 categories, \nt{2} and \nt{\ae}. We hypothesized an asymmetrical shift between the two vowels (\textbf{Asymmetry}), owing to differences in:
- \begin{itemize}
- 	\item their position in the Indian English vowel space: \nt{\ae} is in a more crowded part of the vowel space
- 	\item the presence/absence of a corresponding category in the L1 Bengali inventory: \nt{\ae} is a common phonological category in Bengali and English, while \nt{2} does not exist as a category in Bengali
- \end{itemize}
 
-The results of the study indicate that the effect of the experimental condition differed across the two vowels, showing differences in both the magnitude and direction of shift.
-Overall, the magnitude of shift is greater in \nt{2}.
+Our results indicate differing effects of experimental conditions across the two vowels in both magnitude and direction of shift.
+Overall, shift is greater in \nt{2}.
 Recall (figure \ref{vowels_ie}) that the category \nt{\ae} is in a denser part of the IE vowel space than \nt{2}.
@@
 In contrast, \nt{2} (indicated as \nt{@} in the chart) is in the central part of the vowel space and is not flanked by any immediate neighboring categories.
-Thus, the difference in magnitude of transfer aligns with our hypothesis \textbf{Asymmetry} that the extent of shift in individual phonological categories is constrained by the linguistic principle of contrast preservation.
+Thus, the difference in magnitude of transfer aligns with our hypothesis \textbf{H2---Asymmetry} that the extent of shift in individual categories is constrained by the linguistic principle of contrast preservation.
 An alternative way to interpret this result is that the initial distance between IE \nt{2} and its target L1 category \nt{a:} was greater than that between the IE and Bengali \nt{\ae} categories, leading to a greater visible shift in the former.
-However, our results also parallel previous studies with VOT that have consistently observed greater shifts in the long-lag VOT language (which has a greater acceptable range of VOT, i.e.
-more room for movement) compared to the short-lag VOT language, and attribute these differences to the language-specific distribution of categories \citep{antoniou2011inter, bullock2009trying}.
+However, our results find parallels in previous studies with VOT that have consistently observed greater shifts in the long-lag VOT language (which has a greater acceptable range of VOT) compared to the short-lag language \citep[e.g.][]{antoniou2011inter, bullock2009trying}.
 This suggests that a general principle of ``room for movement" does play a role in moderating transfer.
@@
 This has implications both for our understanding of phonetic transfer as a phenomenon (e.g.
-on which mental representations and at what level of processing and is this occurring?) and language as a system (e.g.
-how have transfer phenomena shaped phonological systems in multilingual settings where languages have co-existed for long periods of time?
+which mental representations and at what level of processing are involved?) and language as a system (e.g.
+how have transfer phenomena shaped phonological systems in multilingual settings, where languages have co-existed for long periods of time?
 What are the linguistic implications of increasingly cosmopolitan urban spaces and identities?).
 
-Our results also indicate asymmetry in the direction of shift between the two categories: the category \nt{2} is lowered and fronted in the mixed language condition, while \nt{\ae} is fronted.
-On the basis of previous literature indicating that both long-term and short-term shifts in phonological categories of multilingual speakers is targeted towards specific categories in the other language, we hypothesized (Section \ref{asymmetry between sounds}) that participants' IE \nt{\ae} would shift towards the Bengali \nt{\ae}, and \nt{2} towards the Bengali \nt{a:}.
+Our results also indicate asymmetry in the direction of shift: the category \nt{2} is lowered and fronted in the mixed language condition, while \nt{\ae} is fronted.
+On the basis of previous literature indicating that both long-term and short-term shifts in phonological categories of multilingual speakers are targeted with respect to the other language, we hypothesized that participants' IE \nt{\ae} would shift towards the Bengali \nt{\ae}, and \nt{2} towards the Bengali \nt{a:}.
 As figure \ref{vowels_e_b} indicates, comparing transfer patterns with Bengali categories from \cite{shruticorpus} suggests that this is borne out.
-The results of \nt{2} align with \citeauthor[][]{simonet2014phonetic}'s (2014) finding that a category which is not common to L1 and L2 nonetheless shifts towards an acoustically/perceptually close category.
-Once again, we see that transfer phenomena make reference to system-level representations.
 
@@
 
-Our third research question, \textbf{Paradigm}, concerned the effect of experimental paradigm on the phonetic outcome of mixed-language use.
-\cite{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
-If such discursive properties independently affect the phonetic outcome of language mixing in an experimental setup, this calls to question the comparability of results across studies using these different paradigms.
-We verified this by comparing productions of the same items from the same group of participants in a scripted CS paradigm and a cued picture-naming paradigm.
-We found that the direction of transfer did not change across the two paradigms, with \nt{2} getting lowered and fronted, and \nt{\ae} getting fronted, in the mixed condition in both cases.
+\cite{olson2013bilingual} posits that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", proposing cued picture-naming as an alternative.
+We tested whether this might independently affect the outcome of transfer, by comparing productions from the same participants in these two tasks.
+We found that the direction of transfer did not change across paradigms.
 However, there was a difference in the magnitude of transfer.
@@
 This asymmetry is not seen in the F2 dimension.
-One way to think about this is that during code-switching, there is a greater degree of language co-activation because switches are being made within an utterance, where planning is required: Since participants see the entire utterance and then read it aloud, there is room for utterance planning.
+One way to account for this is that during code-switching, there is a greater degree of language co-activation because switches are being made within an utterance, where planning is required: Since participants see the entire utterance and then read it aloud, there is room for utterance planning.
 Compared to this, in the cued picture-naming paradigm participants see the language cue immediately prior the target word, inducing extremely ``local" switches and thus a smaller degree of language co-activation.
-While both of these are local switches and take place in a bilingual language mode where both languages are already co-activated, it is possible that this minute difference in time-frame induces a difference of degrees in the extent of cross-language phonetic interaction.
+While both of these are local switches and take place in a bilingual language mode, it is possible that this minute difference in time-frame induces slightly different degrees of cross-language phonetic interaction.
 This aligns with the findings of \cite{bullock2009trying}, who found evidence of transfer in pre-switch tokens, suggesting planning effects.
-In our study specifically, another factor that could have contributed to this is orthography.
+In our study specifically, orthography was another factor that could have potentially affected this.
 In both tasks, Bengali and English words were presented in their respective orthographies.
-In the picture-naming task, this was used to cue language--- as described in  section \ref{procedure}, participants saw one slide at a time.
-In the CS task, mixed utterances were presented in mixed orthography.
-It is possible that seeing mixed orthography, particularly for participants who are used to processing mixed-language visual input in their surroundings, triggered increased language co-activation, leading to greater cross-language interaction.
+In the picture-naming task, participants saw one word at a time.
+The CS task in contrast presented mixed utterances in mixed orthography.
+It is possible that seeing mixed orthography, particularly for participants accustomed to processing mixed-language visual input in their surroundings, triggered increased language co-activation, leading to greater cross-language interaction.
 
@@
 (controlled) code-switching does not significantly alter the behavior under study.
-However, elicitation through code-switching offers greater planning, potentially increasing the magnitude of observable effects.
+However, elicitation through code-switching allows for more planning, potentially increasing the magnitude of observable effects.
 
-%There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in observable transfer at all.
-This would mean that the pre-switch effects observed, for example, in \cite{muldner2019phonetics} and \cite{bullock2009trying} are better interpreted as resulting from some other aspect of the experimental setup, than the act of code-switching itself.
-While planning could additionally affect the words preceding the switch, the effects on the switched token itself do not change.
-(ii)
-
-An unexpected result in the present study is that the paradigm directly affected vowel quality, across experimental conditions: all vowels were more backed and lowered in the code-switching task.
+An unexpected result in this study is that the paradigm directly affected vowel quality: all vowels were more backed and lowered in the code-switching task.
 There is no principled reason to expect this.

### 2022-10-08 draft_BLC/supplementary_material.tex
shortening

@@
 
+\subsubsection*{Baseline Bengali data from the SHRUTI corpus}
+
+The SHRUTI corpus is the largest free corpus of Western Bengali speech that we are aware of.
+It comprises of 21.64 hours of read speech, containing 7383 unique sentences (22012 unique words) from Bengali newspapers in different domains, covering the most frequently used words of the language.
+The speakers are 34 adults (age range 20-40, 75\% male, 25\% female) who have grown up in the Indian state of West Bengal, and speak standard colloquial Bengali.
+All of the speakers have university-level education or higher.
+Given these parallels in linguistic, regional, and socio-economic background, we believe that Bengali speech from this corpus is representative of the average target Bengali vowel categories that are of interest in this study.
+To estimate average vowel positions, we extracted formant values from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, to get estimates from the steady state, which should most closely approximate the production target.
+Next, we compared these to our participants' productions in the non-switched English condition to understand the relative positions of these categories.
+This is visualized in figure \ref{vowels_unilingual_ie_bengali}.
+
+Note that we refrained from using corpus data for average Indian English vowel positions, opting instead to use the participant data from this study, because IE has high regional variability (c.f.
+discussion in section \ref{bengali_english_in_india}).
+To the best of our knowledge, there is no freely available corpus of IE speech from Bengali speakers in India.
+
 \clearpage

### 2022-10-10 draft_BLC/blc_revision1_reviewer_response.txt
trial commit; 2nd round revisions
added reviewer response

@@
 Thank you for the opportunity to submit a revised draft of our manuscript “Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals” for publication in Bilingualism: Language and Cognition.
-We appreciate the time and effort from your end and that of the reviewers in providing feedback on our manuscript.
+We appreciate the time and effort from your end and that of the reviewer in providing feedback on our manuscript.
 We have incorporated the suggestions made by the reviewer.
-Please see below, for a point-by-point response to the reviewers’ comments and concerns.
+Please see below, for a point-by-point response to the reviewer's comments and concerns.
 All page numbers refer to the revised manuscript file uploaded in the portal.
@@
 
-Authors' response (AR):
+Authors' response (AR): Thank you for facilitating a productive discussion on this issue, and for pointing us to these references.
+We have removed the power analysis from the main manuscript.
+As per the editor's suggestion, we have included it as Supplementary Material.
+As suggested, we have acknowledged the small sample size and need for replication in the main text, on page ...
 
@@
 
-AR: In this task the picture, named in either Bengali or English, serves to prime the participant for that particular language.
-The following target word then either complies with the prime (unilingual condition) or deviates from it (inducing switching; mixed condition).
+AR: We have now added a schematic figure showing examples of a nonswitched and a switched trial (figure 4, page...). The trials proceeded as follows:
+(i) see a word in Bengali/English orthography. This is the cue. It is NOT read aloud.
+(ii)see a picture of a fish. Based on the language cue in the previous slide, name it aloud in either English (say "fish") or Bengali (say "matS").  
+(iii) see an English word, read it aloud. This is the target. If the previous slide was named in English this counts as a nonswitch trial; if it was named in Bengali this counts as a switched trial. 
+(iv) solve a distractor match problem before proceeding to the next trial
 
@@
 
-AR:
+AR: Thanks for pointing this out; we agree that "Lobanov normalization" is not the appropriate term here, given the lack of all corner vowels.
+We have replaced this with the term "z-score normalization".
+The target was to represent all formant values on a uniform scale across speakers, removing speaker-specific value ranges, so that the relative positions of the two vowels and shift between conditions could be compared directly.
+Since the data had productions of the same two categories from all speakers, we felt comfortable that this would not affect the interpretability of the values.
+This was useful for first-pass visualizations to understand speakers' vowel systems (e.g.
+fig.
+2 in the text, which shows that speakers vary greatly in their default realizations of the two IE vowels), and visualize gender differences without confounds from individual differences (e.g.
+figure 6).
+Since we were expecting high individual variability, we did this before running any further analyses.
+We agree that given the specific statistical tools used here, this was not essential to the statistical analyses.
+We re-ran the analysis with non-normalized values to verify this, and found identical results.
+We have included the code and output from both analyses in the OSF repository for the project, and added a footnote in the main text addressing this (pg.
+...).
 
@@
 
-AR:
+AR: We have replaced the terminology throughout the text, and used "nonswitch" and "switch/mixed" to refer to the two conditions.
 
@@
 
-AR:
+AR: We have revised these.
 
 The following are minor comments:
+
+9.      Page 14, line 43. “cannot be treated as an atomic measure.” I don’t know what this means. Could you explain or revise wording?
+
+AR: By "atomic", we mean a single measure that characterizes a property.
+Since previous literature has reported that different categories within a language can differ in their patterns of transfer, and that shifts in individual categories may not be identical to the patterns of system-wide shift, we cannot have a single indivisibe ("atomic") measure such as "extent of transfer in L1"-- we must view the overall transfer as resulting from shifts constrained by different factors (cognitive, linguistic).
+To do this, we need to examine transfer in a wider range of sound contrasts.
+
+10.
+Page 19, section 1.7.
+Not sure this list merits its own section.
+I believe this should be incorporated in the previous section.
+I also believe the question might need to be revised.
+You do not only check if vowel categories are modified in one setting relative to the other setting.
+You also check the direction and size (or extent) of the modification.
+Should you have a main research question and some follow-up, secondary questions focused on directionality and extent-of-change?
+18.
+Page 46, line 9.
+“Our third research question Paradigm concerned…” I am a bit confused.
+I understood “Paradigm” to be a hypothesis, not a research question.
+Please, check section 1.7 and decide if Asymmetry and Paradigm and research questions or hypotheses.
+If they are questions, what are the hypotheses in this study?
+Also, when referring to Paradigm and Asymmetry in the Discussion section, please notice that, in most cases, surrounding commas are needed: “Our third research question, Paradigm, concerned…”
+
+AR: We have modified our Research Questions and Hypotheses section on the basis of these comments (pg. ...).
+
+16.
+Page 43, full last paragraph in page.
+This paragraph is quite intriguing.
+It seems it contains the gist of a great idea or observation.
+However, the idea is so vague that the paragraph becomes empty.
+Do the authors have something to say?
+Please, say it clearly (explaining it in some detail), even if speculative (you have a right to speculation after having explained the data).
+Are they not ready to explain their idea in some detail?
+Please, delete hints to ideas of they are not to be explained.
+
+AR: In this paragraph, we suggest that (i) examining multiple acoustic features within a single study would allow us to compare predictions made by different models for phonetic transfer; (ii) given its communicative relevance (as demonstrated by sociophonetic literature), communication-centered approaches might be more appropriate than inhibition-based ones.
+For the same reason, data from native multilingual populations is valuable, since these settings provide obvious communicative motivations for transfer.
+We have revised the wording in the paragraph, and hope it comes across more clearly now.
 
@@
 
-9.      Page 14, line 43. “cannot be treated as an atomic measure.” I don’t know what this means. Could you explain or revise wording?
-
-10.
-Page 19, section 1.7.
-Not sure this list merits its own section.
-I believe this should be incorporated in the previous section.
-I also believe the question might need to be revised.
-You do not only check if vowel categories are modified in one setting relative to the other setting.
-You also check the direction and size (or extent) of the modification.
-Should you have a main research question and some follow-up, secondary questions focused on directionality and extent-of-change?
-
 11.     Page 22, lines 10-13. “all participants … (n = 10)”. I don’t think including n = 10 is necessary here. You have said before you had 10 participants and this sentence already says “all.”
@@
 
-16.
-Page 43, full last paragraph in page.
-This paragraph is quite intriguing.
-It seems it contains the gist of a great idea or observation.
-However, the idea is so vague that the paragraph becomes empty.
-Do the authors have something to say?
-Please, say it clearly (explaining it in some detail), even if speculative (you have a right to speculation after having explained the data).
-Are they not ready to explain their idea in some detail?
-Please, delete hints to ideas of they are not to be explained.
-
 17.     Page 44, line 11: “in a languages” > “in a language” or “in languages”
 
-18.
-Page 46, line 9.
-“Our third research question Paradigm concerned…” I am a bit confused.
-I understood “Paradigm” to be a hypothesis, not a research question.
-Please, check section 1.7 and decide if Asymmetry and Paradigm and research questions or hypotheses.
-If they are questions, what are the hypotheses in this study?
-Also, when referring to Paradigm and Asymmetry in the Discussion section, please notice that, in most cases, surrounding commas are needed: “Our third research question, Paradigm, concerned…”
+AR: Thanks for pointing these out! We have revised the relevant parts of the manuscript.
 
 Congratulations on a fine study and thank you for the opportunity to learn about this study before it was published. I hope some of my comments are found to be useful.
+
+AR: Thank you for the detailed and constructive feedback! We hope we have been able to address all the concerns.

### 2022-10-10 draft_BLC/BLC_AM-ID_manuscript.tex
added reviewer response

@@
 
-Word count (29.7.2022): 13388 -- remove 4388
+Word count (7.10.2022): 9650
 
@@
 
-Bengali (also, Bangla) is an Indo-Aryan language primarily spoken in India and Bangladesh.
+Bengali is an Indo-Aryan language primarily spoken in India and Bangladesh.
 In India, more than 97 million people speak Bengali as a first language (Census of India, 2011), mostly in the state of West Bengal.
@@
 
-Indian English (IE) refers to the variety of English that has developed in the Indian subcontinent.
-In India, it is spoken as an L2 by 129 million people (Census of India, 2011).
-English is one of the two official languages, used in education, law, media, as a lingua franca mainly for an educated elite in most metropolitan regions, and carries a high prestige value \citep{pandey201517, kachru1981english, tollefson2014language, kachru1983indianization}.
+Indian English (IE) refers to the variety of English spoken in the Indian subcontinent.
+It is spoken as an L2 by 129 million people (Census of India, 2011).
+English is one of two official languages, used in education, law, media, as a lingua franca mainly by an educated elite in most metropolitan regions, and carries high prestige value \citep{pandey201517, kachru1981english, tollefson2014language, kachru1983indianization}.
 
@@
 
-The socio-linguistic facts about English in India also suggest that mixed-language processing of English and an L1 an ecologically valid paradigm to understand short-term phonetic interaction.
+The socio-linguistic facts around English in India also suggest that mixed-language processing of English and an L1 is an ecologically valid paradigm to understand short-term phonetic interaction.
 
@@
 
-The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2/3:, a:, O, o, U, u} \citep{wells1982accents, masica1972sound}.
+The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2/3:, a:, O, o, U, u} \citep{wells1982accents, masica1972sound}\footnote{\nt{@, 2, and 3:} are treated as a single phoneme in many descriptions of IE; the slashes between the symbols indicate possible notations for the same category}.
 The vowel chart in figure \ref{vowels_ie} shows their distribution in the F1$\times$F2 formant space.
@@
 
-Existing research has distinguished changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf.
-Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during accessing, processing, or articulation of these categories (cf.
-transfer vs.
+Existing research has distinguished changes in category \textsc{representations} due to the acquisition of multiple sound systems (Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during accessing, processing, or articulation of these categories (transfer vs.
 interference \citep{grosjean2012attempt}; competence vs.
@@
 
-Since participants of this study are in an environment that largely contains mixed-language input and multilingual interlocutors, we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode \citep[c.f.][]{grosjean1998studying}.
+Since participants of this study are in an environment that largely contains mixed-language input and multilingual interlocutors, we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode \citep[see][]{grosjean1998studying}.
 Thus, we elicited all utterances in the same bilingual test block.
@@
 Based on a short-term longitudinal study, \cite[][VOT]{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2.
-For some sounds and features (though not others, cf.
-sec.\ref{asymmetry between sounds}), transfer was additive over time.
+For some sounds and features (though not others, see sec.\ref{asymmetry between sounds}), transfer was additive over time.
 How long these effects last in the absence of regular L2 input was not tested.
@@
 Thus, the patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories.
-In non-switched production, this is evidenced through a L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf.
+In non-switched production, this is evidenced through pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf.
 \textsc{equivalence classification}: \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.
@@
 	\item Target of transfer: findings from existing research on cross-language phonetic interaction suggest that changes in production during mixed-language use are not random, but rather targeted with respect to categories in the other language.
-Thus, another source of asymmetry is the fact that the category \nt{2} is absent in Bengali, whereas \nt{\ae} is a common category across both languages (cf.
-sec.\ref{vowel systems}).
+Thus, another source of asymmetry is the fact that the category \nt{2} is absent in Bengali, whereas \nt{\ae} is a common category across both languages (see sec.\ref{vowel systems}).
 \end{enumerate}
@@
 This observation has important implications for how the existing literature on short-term phonetic interaction is interpreted: if the discursive properties of a CS paradigm independently affect the phonetic outcome of mixed-language use, this calls to question the comparability of results across studies using these different paradigms.
-Because the reported outcomes of short-term phonetic interaction vary so greatly across studies (cf.
-section \ref{asymmetries}), generalizations about the mechanisms that underlie these patterns often rely crucially on the assumption of such comparability.
+Because the reported outcomes of short-term phonetic interaction vary so greatly across studies (see section \ref{asymmetries}), generalizations about the mechanisms that underlie these patterns often rely crucially on the assumption of such comparability.
 Therefore, identifying paradigm-specific differences is particularly relevant for this line of research.
@@
 	\item \textbf{H3---Paradigm}: While researchers have discussed code-switching as a source of discursive factors likely to affect transfer behavior, this has not been systematically studied.
-Existing literature does not allow us to predict precisely what form this effect might take (c.f.
-discussion in Section \ref{paradigms}).
+Existing literature does not allow us to predict precisely what form this effect might take (see discussion in Section \ref{paradigms}).
 Thus, this question is exploratory, and meant to serve as a starting point for more within-study comparisons.
@@
 This was especially important because we were not eliciting unilingual L1 Bengali data in this study, and were using an existing data source to estimate the position of Bengali vowels in the formant space.
-Based on previous literature (c.f.
-section \ref{asymmetry between sounds}), we expect that L2 vowels shift towards related L1 categories during mixed language utterances.
+Based on previous literature (see section \ref{asymmetry between sounds}), we expect that L2 vowels shift towards related L1 categories during mixed language utterances.
 Thus, we wanted to ensure that any between-participant asymmetries in the pattern of shift reflected individual differences in transfer, rather than different acoustic targets.
@@
 Table \ref{table_participants} summarizes some key biographical data from the survey.
-\alert{The complete questionnaire, along with responses, is available in the ``Data" component of the OSF repository for this project (c.f.
-section \ref{data_availability}).
+\alert{The complete questionnaire, along with responses, is available in the ``Data" component of the OSF repository for this project (see section \ref{data_availability}).
 A more detailed discussion of the linguistic situation and language profiles is included as supplementary material.--footnote?}
@@
 Given that small sample sizes are likely to have low statistical power, replication with larger sample sizes are important to confirm the observed effects, and provide much-needed data diversity.
-A more detailed discussion of power analyses in bilingualism research is included as supplementary material on OSF (c.f.
-section \ref{data_availability}).
+A more detailed discussion of power analyses in bilingualism research is included as supplementary material on OSF (see section \ref{data_availability}).
 We hope that this can facilitate more conversation around establishing norms for power and sample size in our field.
@@
 We avoided English words that are already lexicalized loanwords in Bengali.
-The complete list of target words can be found in the ``Materials" component of the OSF repository for this project (c.f.
-section \ref{data_availability}).
+The complete list of target words can be found in the ``Materials" component of the OSF repository for this project (see section \ref{data_availability}).
 
@@
 \begin{enumerate}[]
-	\item Cued picture-naming: Participants were presented with slides in the following sequence:
+	\item Cued picture-naming: Participants were presented with slides in the following sequence (see fig. \ref{figure_picture_task}):
 	\begin{itemize}
-		\item Language cue: a word in either English or Bengali orthography
-		\item A picture displayed for 50 ms; named by the participant in either English or Bengali, depending on the language cue.
-The picture of a fish was used in all trials, prompting the participant to say either \nt{fIS} or \nt{ma: \textteshlig \super{h}}, depending on the language cue
-		\item Target: an English word; read out by the participant
+		\item Language cue: a word in either English or Bengali orthography, not read aloud
+		\item A picture displayed for 50 ms; named aloud by the participant in either English or Bengali, depending on the language cue.
+The picture of a fish was used in all trials, prompting the participant to say either \nt{fIS} (English) or \nt{ma: \textteshlig \super{h}} (Bengali)
+		\item Target: an English word; read aloud by the participant
 		\item Distracter math problem
 	\end{itemize}
-	Since all the target words were in English, the language cue was used to manipulate the preceding word, giving target words in non-switched and switched (mixed) utterances.
-The orthographies of English and Bengali are visually distinct, and were therefore used to cue language.
+	The language cue was used to manipulate the word preceding the target, giving target words in non-switched and switched (mixed) utterances.
 
@@
 
-Carrier sentences were chosen such that the sound preceding the target word was uniform across all sentences (the final sound in \textit{yellow} is produced as [o] in Indian English), to minimize any possible differences due to coarticulatory effects on the target vowel.
+Carrier sentences were chosen such that the sound preceding the target word was uniform across all sentences (the final sound in \textit{yellow} is produced as [o] in Indian English), to minimize differences due to coarticulatory effects on the target vowel.
 To ensure identical word order in both languages, we chose a specifier-noun construction, as this has a parallel structure in Bengali and English.
@@
 \\
+
+<insert fig. \ref{figure_picture_task} about here>
 
@@
 To reduce between-speaker differences, we used z-score normalization (using the PhonR package \citep{phonR} in R \citep{r}, representing each speaker's formant frequencies as the scaled distance from their own mean).
-Plotting these vowels shows a difference between male and female speakers, such that male speakers have overall lower formant frequencies (figure \ref{gender_before}).
-Speaker normalization using the z-score method does not get rid of this difference.
+\footnote{Since our experimental treatments were between-subject and statistical analyses use mixed-effects regression, this step would not affect the results, as pointed out by a reviewer.
+This was done to allow for more intuitive values and visualizations, rater than statistical necessity.
+The code for parallel statistical analysis using non-z-score-normalized values can be found in the OSF repository.} Plotting these vowels shows a difference between male and female speakers, such that male speakers have overall lower formant frequencies (figure \ref{gender_before}).
+%Speaker normalization using the z-score method does not get rid of this difference.
 Since gender-related differences in formant frequencies are not of interest in this study, we normalized individual values by gender (representing each value as a scaled distance from the group mean; figure \ref{gender_after}).
@@
 Figure \ref{vowels_unilingual_ie_bengali} shows the non-switched IE productions of the participants in relation to average positions of related vowel categories in Bengali.
-Figure \ref{vowels_e_b} overlays productions in the mixed condition over this, showing F1$\times$F2 vowel plots for IE vowel productions in the nonswitch and mixed conditions, in comparison to the related Bengali categories from the SHRUTI corpus \cite{shruticorpus}.
+Figure \ref{vowels_e_b} overlays productions in the mixed condition over figure \ref{vowels_unilingual_ie_bengali}, showing F1$\times$F2 vowel plots for IE vowel productions in the nonswitch and mixed conditions, in comparison to the related Bengali categories from the SHRUTI corpus \citep{shruticorpus}.
 Given that the Bengali corpus data represent average vowel positions, and are not from the same set of speakers as the participants of this study, we refrained from including these in statistical analyses.
@@
 \ref{asymmetries}, it might be fruitful for future studies to focus on vowels as a potential site for transfer.
-This would both expand the empirical scope of existing research (cf.
-sec.
+This would both expand the empirical scope of existing research (see sec.
 \ref{asymmetry between sounds}) and avoid confounding factors arising from independent differences between long- and short-lag VOT languages.
@@
 
-Beyond methodological considerations, however, we argue that these differences ultimately raise a broader question about a purely inhibition-based understanding of transfer.
-In light of a compelling body of literature on the sociophonetic and communicative implications of phonetic variation along the L1-L2 continuum \citep[e.g.][]{khattab2013phonetic,agnihotri1979processes,stuart2011hybridity}, is it wise to think of transfer as resulting from an inability to suppress and separate language systems?
-Given that young children are able to use minute phonetic variation in sophisticated ways, might it not be more intuitive to think of transfer as a communication-enabling feature of language, even at a purely cognitive level?
-Research on the cognitive mechanisms underlying cross-language interaction and variation in bilingual speech more generally is still an emerging and exciting line of inquiry.
+Beyond methodological considerations, we argue that these differences ultimately raise a broader question about a purely inhibition-based understanding of transfer.
+In spite of a compelling body of literature on the sociophonetic and communicative implication of phonetic variation along the L1-L2 continuum \citep[e.g.][]{khattab2013phonetic,agnihotri1979processes,stuart2011hybridity}, is it wise to think of transfer as resulting from an inability to suppress and separate language systems?
+Given that young children are able to use minute phonetic variation in sophisticated ways, it might be more intuitive to think of transfer as a communication-enabling feature of language, even at a purely cognitive level.
+Research on the cognitive mechanisms underlying cross-language interaction and variation in bilingual speech is still an emerging and exciting line of inquiry.
 We believe that a communication-centered approach to bilingual variation is essential to understand how bilingual language processing functions in the real world.
-To this end, native multilingual populations serve as an important source of data on language behavior that has been largely ignored, but should be harnessed.
+To this end, highly proficient native multilingual populations serve as an important source of data on language behavior.
 
@@
 
+\begin{figure}[!htb] \label{figure_picture_task}
+	\centering
+	\begin{subfigure}[t]{0.49\textwidth}
+		\centering
+		\includestandalone[mode=image]{picture_task_eng}
+		\caption{English (nonswitch) trial} \label{picture_task_english}
+	\end{subfigure}
+	\begin{subfigure}[t]{0.49\textwidth}
+		\centering
+		\includestandalone[mode=image]{picture_task_bengali} 
+		\caption{Bengali-English (switch) trial} \label{picture_trial_bengali}
+	\end{subfigure}
+	\caption{Picture-naming task sample sequence: language cue (not read out), picture (named as per lang cue), target (read out), distractor}
+\end{figure}
+
+\newpage
+
 \begin{figure}[h] %\centering
@@
 
-\newpage
-
-\begin{figure}[!htb] \label{figure_picture_task}
-	\centering
-	\begin{subfigure}[t]{0.49\textwidth}
-		\centering
-		\includestandalone[mode=image]{picture_task_eng}
-		\caption{English (nonswitch) trial} \label{picture_task_english}
-	\end{subfigure}
-	\begin{subfigure}[t]{0.49\textwidth}
-		\centering
-		\includestandalone[mode=image]{picture_task_bengali} 
-		\caption{Bengali-English (switch) trial} \label{picture_trial_bengali}
-	\end{subfigure}
-\caption{Picture-naming task sample sequence: language cue (not read out), picture (named as per lang cue), target (read out), distractor}
-\end{figure}
-
 \end{document}

### 2022-10-10 draft_BLC/extra_text_from_manuscript/extra.tex
added reviewer response

--- 2022-05-09 draft_BLC/BLC_AM-ID_manuscript.tex
@@
-% !TeX document-id = {6d544fe2-4255-4108-8022-d809ff1d2c53}
-% TeX program = xelatexmk %doesn't work, no xelatexmk option under 'build'. Changed default engine to xelatex (options-->configure-->build-->default build-->xelatex)
-% see http://info.semprag.org/basics for a full description of this template
-%\documentclass[times,linguex]{glossa}
-\documentclass[12 pt]{article}
-\usepackage[letterpaper]{geometry}
-\usepackage{times}
-\geometry{top=1.5in, bottom=1.0in, left=1.0in, right=1.0in}
-
-\usepackage[hidelinks]{hyperref}
-%\renewcommand{\sectionautorefname}{\S}
-\usepackage{xcolor}
-\hypersetup{
-	colorlinks,
-	linkcolor={blue!50!black},
-	citecolor={blue!50!black},
-	urlcolor={blue!80!black}
-}
-
-\usepackage{graphicx} %for images
-\graphicspath{{images/}}
-\usepackage{caption}
-\captionsetup[table]{skip=0pt,singlelinecheck=off}
-\captionsetup[figure]{skip=0pt,singlelinecheck=off}
-
-\usepackage{booktabs}
-\usepackage{enumitem} %customize list numberings
-
-\usepackage[natbibapa]{apacite}
-\usepackage{natbib}
-
-\usepackage{tipa}
-\newcommand{\nt}[1]{\textipa{[#1]}} % narrow transcription
-\newcommand{\wt}[1]{\textipa{/#1/}} % wide transcription
-
-\usepackage{fancyhdr}
-\pagestyle{fancy}
-\fancyhf{}
-\renewcommand{\headrulewidth}{0pt}
-\fancyhead[R]{\thepage}
-
-\usepackage{setspace}
-\doublespacing
-
-\usepackage{comment}
-
-\newlength\mystoreparindent
-\newenvironment{myparindent}[1]{%
-	\setlength{\mystoreparindent}{\the\parindent}
-	\setlength{\parindent}{#1}
-}{%
-	\setlength{\parindent}{\mystoreparindent}
-}
-
-\usepackage{easyReview}
-
-\title{Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals}
-
-% \author[Auromita \& Indranil]% short form of the author names for the running header. If no short author is given, no authors print in the headers.
-% {%as many authors as you like, each separated by \AND.
-%   \spauthor{Auromita Mitra\\ 
-%   \institute{The EFL University}\\
-%   \small{%105, Bd. Raspail, 75005 Paris\\
-%   auromita.mitra@gmail.com}
-%   }
-%   \AND
-%   \spauthor{Indranil Dutta \\
-%   \institute{Jadavpur University}\\
-%   \small{%Warmoesberg 26, 1000 Brussel\\
-%   indranildutta.lnl@jadavpuruniversity.in}
-%   }%
-% }
-%\usepackage{gb4e}
-
-\begin{document}
-\bibliographystyle{apacite}
-%\sffamily
-%\maketitle
-
-%Word count: 9263
-
-\begin{singlespace}
-
-\begin{myparindent}{0pt}
-Short title: Phonetic transfer\\
-
-Full title: Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals\\
-
-Authors:\\
-
-Auromita Mitra
-
-EFL University,
-
-Hyderabad,
-
-India\\
-
-Indranil Dutta
-
-Jadavpur University,
-
-Kolkata,
-
-India\\
-
-Acknowledgments:\\
-We are grateful to Maumita Bhowmik and Anannya Mondal for their help with data collection and processing, and to all the participants of the study.\\
-
-Address for correspondence:\\
-Auromita Mitra,
-
-EFL University,
-
-Hyderabad- 500007,
-
-auromita.mitra@gmail.com
-
-\end{myparindent}
-
-\end{singlespace}
-
-\newpage
-\begin{myparindent}{0pt}
-Abstract:\\
-This study investigates the phonetic outcome of mixed-language speech in Bengali and Indian English, towards understanding cross-language interaction in highly proficient bilinguals.
-We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in unilingual vs.
-mixed productions.
-Results reveal dynamic and asymmetrical changes in both vowels;  \nt{2} shifting towards a related L1 category \nt{a:}.
-We interpret this as a temporary increase in cross-language phonetic interaction during mixed-language use, leading to a shift towards L1 norms, evidencing transfer effects on L2 vowels for the first time.
-We elicit mixed productions through two tasks: cued picture-naming and code-switching, to assess if experimental paradigm independently influences the behavior under study.
-Results reveal no difference across paradigms.
-We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction, postulated discursive factors in code-switching, and the issue of comparability between paradigms in transfer studies.\\
-
-Keywords: phonetic transfer; vowel quality; bilingual speech production; L1-L2 interaction; Indian English; Bengali
-
-\end{myparindent}
-
-\newpage
-
-\section{Introduction: Phonetic transfer}
-
-Bi/multilingual speakers can distinguish between the phonetic norms of their languages and maintain separate sound categories for each language \citep{caramazza1973acquisition,macleod2010impact,bosch2003simultaneous}.
-However, these categories are not autonomous-- they influence each other across languages in both perception and production \citep[e.g.][]{flege1995second,fowler2008cross,flege2002assessing}, and the nature of such interaction provides crucial insights into how language `systems' are cognitively represented and processed.
-Based on the productions of proficient bilingual speakers of Bengali and English, this study reports that cross-language phonetic influence temporarily increases during mixed-language use.%Given that a majority of the world's population is multilingual, this is a central concern for any realistic theory of language.
-%show cross-language-- rephrase
-
-%keep just the types here, and put the next 2 lines in the causes&effects section?
-Cross-language influence at the level of sounds can be studied in broadly two kinds of conditions:
-(i) While a bilingual speaker is operating in any one of their languages, by comparing bilingual speech to monolingual norms \citep[e.g.][]{guion2003vowel,caramazza1973acquisition,flege1987production}; %These studies often compare the speech of bilinguals to monolingual norms, and view the influence as changes to long-term memory representations as a result of acquiring an L2 \citep{guion2003vowel,caramazza1973acquisition,flege1987production};
-(ii) When both languages of a bilingual speaker are co-activated, by comparing speakers' mixed-language speech to their own norms while using a single language \citep[e.g.][]{grosjean1994going, bullock2009trying,elias2017effects, simonet2014phonetic}.
-%These compare productions during mixed-language use to participants' own productions while using a single language.
-The resulting influence is variously thought to involve short-term memory, online processing costs, language mode, and context-awareness.
-The present study is concerned with the latter.
-In production, this kind of influence has been variously termed \textsc{transfer, drift, accommodation,} and \textsc{interference}-- we use the term \textsc{transfer} here, to indicate any interaction between two sets of phonetic norms.
-
-Existing research on the phonetic effects of mixed-language production has largely focused on a limited set of phonologically related languages.
-A majority of these studies use temporal properties of consonants (in particular, voice onset time or VOT) to measure transfer.
-However, the reported results vary greatly across studies, and appear to be contingent upon both language-specific features and language experience of the participants.
-This emphasizes the importance of considering data from a wider variety of
-%elaborate on 'great variation'-- differences between language pairs, plus sociolinguistic situation (effects of language experience etc)-- clearer.
-'Types of features'--separate line?- Brief mention - discursive factors in VOT.
-populations, language pairs,  and phonetic features in order to make meaningful generalizations. Widespread multilingualism in the Indian subcontinent suggests that phonetic
-%line--rephrase
-behavior in these populations can be particularly valuable towards understanding the nature of such cross-language interactions, as they are likely to reflect real-world experience with mixed-language processing.
-However, there is no work yet on short-term phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language.
-
-The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India.
-We measure spectral properties: F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2.
-Mixed-language data is elicited in two switching paradigms: cued picture-naming and code-switching.
-We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction.
-The results demonstrate a dynamic effect of transfer on L2 vowels during mixed-language production, \alert{but no difference in the extent of transfer between the two paradigms --and slightly different patterns of shift among the paradigms?}.
-We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.
-
 The rest of this section is structured as follows: \ref{bengali_english_in_india} provides background information on Bengali and Indian English focusing on vowel systems.
-\ref{causes} discusses two potential sources of transfer during language processing-- global co-activation of multiple languages vs.
+\ref{causes} discusses two potential sources of transfer during language processing: global co-activation of multiple languages vs.
 the local act of switching between languages.
@@
 In \ref{paradigms}, we motivate a secondary aim of this study: to verify the postulated differences between two paradigms for eliciting short-term phonetic interaction in an experimental setup, namely cued picture-naming and code-switching.
-\ref{questions_and_hypotheses} summarizes the research questions and hypotheses.
+Finally, we summarize the research questions and hypotheses.
 
-Existing studies have studied phonetic transfer along different acoustic features.
-While discussing the literature in subsequent sections, we will often indicate the acoustic feature in parentheses following a citation, for clarity.
+Taken as a whole, these results bear on our original research questions in the following ways: regarding our first question of whether L2 vowels of proficient bilinguals can be affected by short-term transfer from L1 (\textbf{L2 vowels}), our results indicate that this indeed happens during mixed-language use.
+Both the common category \nt{\ae} and the L2-exclusive category \nt{2} show spectral shifts during mixed-language productions compared to non-switched productions.
+As hypothesized (c.f.
+RQ2 and H2, \textbf{Asymmetry}), the results reveal asymmetries in the extent and direction of shift between the two vowels.
+Specifically, \nt{2} is lowered and fronted, while \nt{\ae} is fronted, with the extent of shift being greater in \nt{2}.
+Comparison with canonical Bengali vowel productions shows that the shift in \nt{\ae} is towards the corresponding Bengali category \nt{\ae}, while that in \nt{2} is towards the Bengali category \nt{a:}.
+Regarding our third research question of whether the paradigm used to elicit language switching independently affects patterns of transfer (RQ3: \textbf{Paradigm}), analyzing the role of Task in the results confirms this.
+We find a greater degree of transfer during code-switching compared to the cued picture-naming task.
+However, this difference is limited to the F1 dimension, and needs to be replicated.
+These results and their implications are discussed in the following subsections.
 
-\subsection{Bengali and English in India} \label{bengali_english_in_india}
+An often discussed issue in transfer research is whether cross-language interaction at the phonetic level during language mixing is triggered by the co-activation of languages that makes representations from both languages available, or by the local act of switching during online processing and articulation of those sounds.
 
-\subsubsection*{Demography}
-
-Bengali (also, Bangla) is an Indo-Aryan language primarily spoken in India and Bangladesh.
-In India, more than 97 million people speak Bengali as a first language (Census of India, 2011), mostly in the state of West Bengal.
-A majority of this population also speaks other additional languages.
-
-Indian English (IE) refers to the variety of English that has developed in the Indian subcontinent.
-In India, it is spoken as an L2 by 129 million people (Census of India, 2011).
-English is one of the two official languages, used in education, law, media, as a lingua franca mainly for an educated elite in most metropolitan regions, and carries a high prestige value \citep{pandey201517, kachru1981english, tollefson2014language, kachru1983indianization}.
-
-\alert{note that there are large regional variations in Indian English, and the literature on IEphonology is thus not "standardized" or uncontested.
-For the present study, we focus on a variety of IE spoken by an educated elite, used in news channels, taught in classrooms (citations)}
-
-Recent research comparing the segmental and suprasegmental properties of IE spoken in different parts of the country report many commonalities (see \cite{sirsa2013effects} for a review).
-This suggests that in spite of the varied L1s of its speakers, IE has a target phonology that is distinct from any of these, as well as from other native varieties of English.
-Thus, it is fruitful to think of regional variations as resulting from L1-influence on a common underlying target.
-Excepting a small minority of specific communities \citep{pandey201517, wells1982accents, coelho1997anglo}, there are very few L1 speakers of IE.
-
-Considering the language usage patterns suggested by this demography, the present study focuses specifically on short-term phonetic interaction during mixed-language use, because:
-
-\begin{enumerate}[label=(\roman*)]
-	\item In multilingual populations, long-term representations of sound categories are expected to be affected by multiple languages. 
-	\item Given that examples of an IE phonology without L1 `influence' are very rare, it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a given baseline condition.
-\end{enumerate}
-
-The socio-linguistic facts about English in India also suggest that for the present population, mixed-language processing of English and an L1 is an ecologically valid paradigm that is likely to be a part of everyday language experience.
-Thus, data from this population is valuable to the understanding of short-term phonetic interaction.
-
-\subsubsection*{Vowel systems} \label{vowel systems}
-The vowel inventory of Western Bangla (the variety spoken by the participants in this study) consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \citep{garry2001facts}.
-Note that there is no mid-central vowel category.
-
-The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2, a:, O, o, U, u}, represented by the lexical set KIT, FLEECE, DRESS, FACE, and TRAP, STRUT, PALM, LOT/CLOTH, GOAT, FOOT, GOOSE \citep{wells1982accents, masica1972sound}.
-%Their distribution in the F1XF2 space is shown in (insert table 1.7, Pandey).
-A single mid-central vowel corresponds to the categories \nt{2,@,3:}, which are treated as distinct in many native varieties of English \citep{nihalani1979indian,wells1982accents,hickey2005legacies,bansal1969intelligibility}.
-Since the English items used in this study are traditionally transcribed with \nt{2}, we use this symbol to indicate the mid-central vowel throughout.
-
-\subsection{What causes transfer and what does it affect?}\label{causes}
-
-Existing research has distinguished changes in category \textsc{representations} due to the acquisition of multiple sound systems (cf.
-Speech Learning Model \citep{flege1995second,flege2007language}; Perceptual Assimilation Model-L2 \citep{best2007nonnative}), from interaction during accessing, processing, or articulation of these categories (cf.
-transfer vs.
-interference \citep{grosjean2012attempt}; competence vs.
-performance interference \citep{paradis1993linguistic}).
-Given their transient nature, dynamic changes in production during mixed-language use are generally attributed to the latter, e.g.
-online processing costs \citep[][VOT]{olson2013bilingual,tsui2019impact,vsimavckova2015immediate}, language mode \citep[][vowel quality]{simonet2014phonetic}, context-awareness \citep[][phonological variables]{khattab2013phonetic}.
-What triggers this interaction?
-\cite{olson2016role} argues that while cross-language phonetic effects are largely measured at the point of language switch, it could have two potential sources:
-\begin{enumerate}
-	\item The local point of switch itself
-	\item Global co-activation of two languages -- \textsc{bilingual language mode} \citep{grosjean1998studying} 
-\end{enumerate}
-
-A number of studies have specifically manipulated language mode, both in the presence and absence of switching.
-Overall, results suggest that: (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep[][vowel quality]{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use--- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \citep[][VOT]{olson2016role,tsui2019impact} or spontaneous conversation \citep[][VOT]{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
-(iii) How the two sources interact to influence the final outcome of transfer is not fully understood-- \cite[][VOT]{olson2016role} found no additive effects of language mode, \cite[][VOT]{olson2013bilingual} found a balanced language context to inhibit transfer compared to unbalanced contexts.
-Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days (\cite{elias2017effects}; vowel quality), \citep[][VOT]{schwartz2015language, bullock2009trying,antoniou2011inter, vsimavckova2015immediate,vsimavckova2018patterns}.
-
-\cite{grosjean1998studying} suggests that various aspects of the communicative setting, including exposure to (spoken or written) stimuli in multiple languages and awareness of the interlocutor's being bilingual, could trigger a bilingual language mode.
-The participants in the present study are in an environment which largely contains mixed-language input, multilingual interlocutors, and no stable ambient language.
-Therefore we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode.
-To preserve ecological validity, we elicited both unilingual and switched utterances in a bilingual language mode.
-Any observed differences in this paradigm would arguably result from interaction during online processing of sounds while switching between languages.
-
-In a bilingual mode, both language systems are expected to be nearly equally accessible throughout the test block.
-Thus, a consistent difference between switched and non-switched tokens in such a paradigm would be possible only if the effects were highly localized-- if not, we should expect a gradual convergence over the course of the experiment.
-This is discussed in the next section.
-
-\subsection{Duration of transfer effects} \label{duration}
-
-Research comparing bilingual speakers' phonological systems with monolinguals treats them as relatively stable over time.
-The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods \citep[e.g.][]{guion2003vowel, caramazza1973acquisition}.
-However, longitudinal studies %(citations)
-and between-subject comparisons of bilingual speakers with different durations of L2 exposure suggest that interaction between the sound systems is dynamic: over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts.
-For example, \cite[][vowel quality]{bohn1992production}
-observed that approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language influence.
-These have been interpreted as changes to long-term phonological representations.
-
-Moreover, changes due to transfer are not necessarily unidirectional or irreversible.
-\cite[][VOT]{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language.
-This not only evidenced that cross-language interaction can be triggered in an order of months, but also that the effects can be reversed within a similar time range.
-A more recent study by \cite[][VOT]{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks).
-Based on a short-term longitudinal study, \cite[][VOT]{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2.
-For some sounds and features (though not others, cf.
-sec.\ref{asymmetry between sounds}), transfer was additive over time.
-How long these effects last in the absence of regular L2 input was not tested.
-% if effects are cumulative, then the difference between switched and non-switched tokes should be lesser in later test blocks (previous two paragraphs needed?)
-
-The research discussed above concerns situations where a speaker is operating in any one of their languages.
-When it comes to the phonetics of mixed-language use, the majority of existing studies only analyze the switched (target) token.
-Thus, there are few direct measurements of the duration of short-term transfer effects.
-One study which measured this in a code-switching paradigm \citep[][VOT]{bullock2009trying} did not find any residual effects on the matrix language following a switch, suggesting that transfer during code-switching is localized.
-Indirect evidence in support of this comes from experiments which have elicited switched and non-switched tokens in the same test block and still reported differences between the two \citep[e.g.][VOT]{tsui2019impact,olson2013bilingual}, suggesting that changes due to transfer are quickly `reset'--- in an order of seconds.\\
-Note that all the studies discussed above measure VOT, which is a temporal feature.
-We have no apriori reason to assume that these durations generalize to vowel quality.
-However, findings from sub-categorical phonetic shifts triggered by other factors (such as convergence towards an interlocutor) do evidence rapid shifts in vowel quality within comparable time-frames \citep[e.g.][]{pardo2010expressing,babel2010dialect,babel2012evidence}.
-Thus in the current study, we present switched and non-switched tokens randomly within the same test block to induce a bilingual mode.
-We expect the intervening words between two subsequent targets to undo any residual effects of transfer.
-
-\alert{discussion: found that like VOT, vowel quality shifts due to switching are also highly localized, and reset in an order of seconds.
-This behavior is similar to other context-driven shifts in the literature}
-
-\subsection{Asymmetries between languages in extent and direction of transfer} \label{asymmetries}
-A recurring pattern that emerges from existing research on short-term phonetic transfer is that the extent and patterns of phonetic shift are not equivalent across the two languages of the bilingual speaker.
-Studies vary greatly in the nature of differences that they report.
-The potential causes for such asymmetry are of interest because they point towards the factors that mediate cross-language phonetic interaction.
-In this section, we discuss two proposed sources of such asymmetry that are relevant to the present study:
-
-\subsubsection*{L1 vs. L2 status} 
-Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} posits that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another.
-Thus, the patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories.
-In non-switched production, this is evidenced through a pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf.
-\textsc{equivalence classification}-- \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.
-
-In mixed-language production, the role of language status is less clear-- studies have reported unidirectional influence of L1 on L2 \citep[][VOT]{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \citep[][VOT]{tsui2019impact, olson2013bilingual}, (\cite{elias2017effects}, vowel quality) bidirectional convergence \citep[][VOT]{bullock2009trying, olson2016role}, divergence \citep[][VOT]{bullock2009trying,vsimavckova2018patterns}, and no influence \citep[][vowel quality]{muldner2019phonetics}, \citep[][phonological process]{schwartz2015language}.
-Since most existing studies measure shifts in VOT, the observed asymmetries between languages have been variously explained either in terms of their L1 vs L2 status, or language-specific differences between long- and short-lag VOT languages.
-%(cf sec.
-\ref{sound systems}).
-
-\cite{olson2013bilingual} first reported a unidirectional VOT shift of L1 towards L2 in two different groups -- native speakers of Spanish (short-lag) and English (long-lag), matched for proficiency and age of L2 acquisition.
-This established that beyond language-specific differences, the L1 vs L2 status of the language does mediate transfer.
-They interpret this asymmetry in terms of the Inhibitory Control Model (ICM) \citep{green1998mental}: to select a phonetic realization from one language, the other must be \textsc{inhibited}.
-Since L1 is the `stronger' language, the amount of inhibition required on L1 while using an L2 is greater than the inhibition required on an L2 while using the L1.
-This greater initial inhibition on L1 means that switching into L1 incurs a greater \textsc{switch cost} (more cross-language influence) than switching into L2.
-Lower switch cost results in a lack of visible cross-language transfer effects on L2.
-\cite{tsui2019impact} report comparable results, but only in participants who were not equally dominant in both languages.
-Balanced bilinguals did not demonstrate any transfer effects in VOT.
-They propose that this is because balanced bilinguals have better inhibitory control (low switch cost in both languages) due to greater experience with language switching.
-
-Since differences between long-lag and short-lag languages independently affect the phonetic realization of VOT, focusing on other phonetic features as sites for transfer can avoid this conflation, and clarify the precise effect of language status.
-However, given the implications of the ICM, it is necessary to first establish that these L2 categories can indeed be affected by dynamic interference, particularly in bilinguals with extensive language-switching experience.
-The few existing studies on vowel quality \citep{simonet2014phonetic,muldner2019phonetics,elias2017effects} or phonological processes \citep{simonet2020increased,schwartz2015language} have examined transfer effects on L1 categories.
-As far as we are aware, there is no work yet on short-term transfer in L2 vowels.
-Thus, this study builds on the existing work by examining whether the L2 vowel quality of proficient bilinguals can be affected by mixed-language processing.
-
-\alert{discussion: more about ICM, assumptions about processing}
-
-\subsubsection*{Differences between sound systems of the languages} \label{sound systems} In addition to the cognitive factors discussed above, many researchers have attributed the observed asymmetries to language-internal factors.
-Specifically, a pattern that emerges from the body of work on VOT is that a shift in long-lag VOT towards the short-lag norms of the other language is much more consistent and systematic than the converse \citep{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid}.
-In general, short-lag VOTs seem to resist accommodative shift.
-\cite{bullock2009trying} suggest that because long-lag languages offer a greater range of acceptable VOTs, there is more `room' for movement.
-In contrast, shift in short-lag languages could risk the loss of a phonological contrast (VOT being the primary cue for voicing contrast in these languages).
-This is a language-specific phonological constraint on transfer.
-In a similar vein, \cite{antoniou2011inter} observe that while the VOT of English stops shifted towards Greek (a short-lag language), the reverse was seen only in a specific subset of sounds-- word-medial non-nasalized stops.
-They attribute this to the low frequency of these sounds in Greek, making them less ``stable", and thus more amenable to cross-language influence.
-This suggests that the distribution of sounds in a language could affect the kind of transfer patterns observed.
-
-These results highlight that beyond interaction during the cognitive processing of sounds, phonetic transfer is ultimately a linguistic phenomenon, and thus subject to language-specific phonological constraints.
-Therefore, to make meaningful generalizations, it is imperative to consider data from a variety of language pairs.
-Existing research on phonetic transfer largely revolves around a limited set of phonologically related languages.
-Thus, the present study extends the scope of this research to a new pair of languages--- Indian English and Bengali.
-
-Since Bengali has a four-way laryngeal contrast, VOT does not encode a binary voiced-voiceless distinction.
-Moreover, VOT is not the only relevant cue for the corresponding phonological categories in the language \citep{dmitrieva2020acoustic}.
-This means that any effect of L1 Bengali on L2 English VOT is likely to be affected by additional phonological factors.
-Thus, avoiding any potential confounds was another reason we chose to focus on vowel quality as a target for transfer in the present study.
-
-\subsection{Asymmetries between sounds}\label{asymmetry between sounds}
-
-Asymmetries in the extent and patterns of transfer have not only been observed between languages, but also between different sounds/features of a language, in both long-term and transient interactions.
-Studies that have examined multiple sound categories have found that interactions between individual sound pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep[e.g.][vowel quality]{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot treated as an atomic measure.
-In light of the discussion about VOT in section \ref{sound systems}, this is not surprising --- if a general linguistic principle of `room for movement' and contrast constrains transfer, then we should expect it to apply to individual sounds too.
-Once again, this emphasizes the importance of examining a wider range of sound contrasts.
-
-In the present study, we focus on two vowel categories in Indian English-- the mid-central vowel \nt{2} and the low front vowel \nt{\ae}. There are two plausible sources of asymmetry between these:
-\begin{enumerate}
-	\item Position in the IE vowel space: compared to \nt{\ae}, \nt{2} exists in a part of the vowel space that has a lower vowel density.
-This affords a greater latitude for movement without risking the loss of a contrast, particularly in the vowel height (F1) dimension.
-Thus, considering purely phonological constraints on IE, we expect that a greater degree of shift is possible in \nt{2} compared to \nt{\ae}.
-Any movement in \nt{\ae} is expected to be primarily in the backness (F2) dimension.
-	\item Target of transfer: findings from existing research on cross-language phonetic interaction suggest that changes in production during mixed-language use are not random, but rather targeted with respect to categories in the other language.
-Thus, another source of asymmetry is the fact that the category \nt{2} is absent in Bengali, whereas \nt{\ae} is a common category across both languages (cf.
-sec.\ref{vowel systems}).
-\end{enumerate}
-
-%table
-
-Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} posits that sound categories which are common across languages influence each other because they share a common acoustic-phonetic space.
-Thus, if Bengali and English differ in their canonical realizations of \nt{\ae}, then we should expect the English \nt{\ae} to shift towards the corresponding Bengali category in the mixed condition.
-
-%Other categories can in principle be produced with native-like accuracy, showing little effect of transfer.
-Given that our participants are highly proficient bilinguals, and assuming that cross-language correspondences remain the same in representation and processing, this would lead us to expect more transfer in the common category, \nt{\ae}.
-
-There is no obvious competing L1 category during the production of \nt{2}.
-However, it is unlikely that this should altogether preclude a shift in \nt{2}, since at least one existing study has reported transfer effects on a non-common vowel category in a comparable paradigm \citep{simonet2014phonetic}.
-Here, participants' Catalan \nt{O} shifted towards an acoustically close category \nt{o}, which is common to Catalan and Spanish, suggesting that movement might be towards a set of phonetic norms rather than a specific corresponding category.
-
-We expect the vowel \nt{2} to shift towards an acoustically/perceptually close category in IE -- a probable candidate being the low vowel \nt{a:}, as anecdotal observations of Bengali-accented English suggest that the sounds are perceived as close by Bengali speakers.
-
-In the present study, our research question concerns whether L2 vowels are produced differently in unilingual vs mixed conditions.
-We do not elicit unilingual Bengali productions.
-To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we used production data from the freely available Bengali corpus, SHRUTI \citep{shruticorpus}.
-Formant values were extracted from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, to get formant estimates from the steady state, which should most closely approximate the production target.
-Next, we compared these to our participants' productions in the unilingual English condition to understand the relative positions of the categories.
-\alert{fig.
-..
-plots-- from current plot- remove the dotted line.
-Filter by context, then remove group=context from figure code} F1-F2 plots show that the Bengali category \nt{\ae} is lower, and \nt{a:} lower and more fronted, than the English categories \nt{\ae} and \nt{2} respectively.
-
-\subsection{A note on paradigms: connected speech or not?} \label{paradigms}
-
-This section motivates a secondary aim of this study: to verify if the extent and patterns of phonetic interaction in bilingual participants are independently mediated by the choice of experimental paradigm.
-Studies of phonetic transfer during mixed-language use have largely employed spontaneous or scripted code-switching (CS) to elicit mixed productions.
-CS is defined as the use of multiple languages within a single utterance, i.e.
-in connected speech \citep{myers1993dueling}.
-Other paradigms which have been used include cued picture-naming, delayed repetition, interpreting across languages, reading word-lists etc.
-\cite{olson2013bilingual} argues that code-switching introduces discursive factors such as discourse planning and pragmatic considerations, which ultimately reflect patterns of language practice, rather than cognitive behavior.
-This finds support from studies which report transfer effects \textsc{before} the switch point \citep{bullock2009trying}, or other evidence for planning such as hyperarticulation \citep{muldner2019phonetics} and interlocutor-awareness in spontaneous CS \citep{khattab2013phonetic}.
-To avoid such confounds, \cite{olson2013bilingual} endorses cued picture-naming (where pictures are named in rapid succession in either language, based on the given language cue) as an alternative paradigm to isolate purely phonetic effects in controlled setups.
-This observation has important implications for how the existing literature on short-term phonetic interaction is interpreted: if the discursive properties of a CS paradigm independently affect the phonetic outcome of mixed-language use, this calls to question the comparability of results across studies using these different paradigms.
-Because the reported outcomes of short-term phonetic interaction vary so greatly across studies (cf.
-section \ref{asymmetries}), generalizations about the mechanisms that underlie these patterns often rely crucially on the assumption of such comparability.
-Therefore, identifying paradigm-specific differences is particularly relevant for this line of research.
-
-Note that while there is evidence for planning in CS, precisely how this affects phonetic interaction cannot be predicted from the existing literature--- while \cite{bullock2009trying} found tokens immediately preceding the switch point to be phonetically distinct from the rest, one participant group converged towards the switch language (explained in terms of proficiency) while the other showed divergence (explained in terms of extra-linguistic factors).
-The participants in Muldner et al.'s \citeyearpar{muldner2019phonetics} study hyperarticulated switched tokens (indicating some degree of planning), and did not evidence a significant shift during mixed-language use.
-Moreover, recall that the findings of \cite{olson2016role} suggest that there is a limit on the extent of transfer--- multiple factors do not necessarily result in additive effects.
-Thus, the presence of additional factors in CS does not necessitate that this should alter the extent of observable transfer compared to any other paradigm.
-It is possible that the outcomes in the studies discussed above are better attributed to some other aspect of the experimental setup, rather than to the act of code-switching in particular.
-
-The precise contribution of various (linguistic and extra-linguistic) factors and their interactions to the phonetics mixed-language use merits detailed investigation in its own right.
-In the present study, however, our focus is limited to verifying whether, in an experimental setting, the postulated discursive elements of a CS paradigm alter the extent of observable transfer compared to a paradigm that arguably lacks these--- cued picture-naming.
-We are less concerned with the exact nature of differences than with their comparability.
-While researchers have successfully employed each of these paradigms to induce phonetic interaction, comparison across studies is difficult since they also differ in other factors.
-We provide a starting point for such comparison by eliciting productions of the same tokens from the same set of participants, in both paradigms.
-A difference in results will suggest that the planning involved in connected speech and code-switching affects transfer above and beyond the phonetic outcome of switching itself, which has implications that future work should consider.
-
-\subsection{Research questions and hypotheses}\label{questions_and_hypotheses}
-\begin{itemize}
-	\item \textbf{L2 vowels}: Are L2 vowel categories of proficient bilingual speakers affected by dynamic phonetic transfer during mixed-language production, as evidenced by differences in the spectral properties of L2 vowels in switched and non-switched tokens?
-	\item \textbf{Asymmetry}: If so, we expect an asymmetry between the vowels in the extent and direction of shift.
-Based on comparisons with Bengali vowels from \cite{dutta2021} we expect \nt{2} to be lowered and fronted in the mixed condition.
-\nt{\ae} is expected to show lowering, but to a lesser extent than \nt{2}, as it is in a part of the IE vowel space that has greater vowel density.
-	\item \textbf{Paradigm}: Is there an observable difference between the extent of transfer in a cued picture-naming vs. a code-switching paradigm? 
-\end{itemize}
-
-\section{Methodology} \label{methodology}
-The experiment had the following factorial design: 10 participants* 20 target words* 2 conditions* 2 tasks* 4 iterations = 3200 target tokens.
-
-\subsection{Participants} \label{participants}
-
-10 Bengali-English bilingual speakers (male=5, age range 19 to 28), completed the tasks.
-All the participants speak other languages in addition to Bengali and English.
-At the time of recording, they were living in or around the EFL University campus in Hyderabad, which has a linguistically diverse population.
-Speakers were selected on the basis of responses to a Language Background Questionnaire to ensure comparable LSRW (listening, speaking, reading, writing) skills in L1 and L2 across participants, and were compensated for their time.
-
-\subsection{Stimuli}
-Twenty monosyllabic English words containing the vowels \nt{2} or \nt{\ae} were selected as target words.
-To minimize coarticulatory effects on the vowel, the pre-vocalic consonants were all voiced plosives (\nt{b} or \nt{d}).
-10 monosyllabic English words that do not contain the target vowel served as filler items.
-We avoided English words that are lexicalized loanwords in Bengali.
-%The list of words can be found in sec.\ref{appendix}.
-
-\subsection{Procedure}
-Participants were recorded individually in a sound attenuated room. %, using a ... microphone type/description?
-Productions were recorded in Praat \citep{boersma2016praat} at a 44.1K sampling rate.
-Every participant performed four iterations each of two tasks-- cued picture-naming and code-switching.
-Utterances in the two conditions (unilingual and mixed) were randomly interspersed within each iteration.
-In the unilingual condition, the target word was preceded by another English word in the utterance.
-In the mixed condition, the target was preceded by a Bengali word and thus produced as a switch from Bengali.
-Every target word appeared twice during a test block; once in a unilingual and once in a mixed condition.
-Participants alternated between the two tasks, and were allowed to take a short break after each test block.
-The tasks are described below:
-
-\begin{enumerate}[]
-	\item Cued picture-naming: Participants were presented with slides in the following sequence:
-	\begin{itemize}
-		\item Language cue- a word in either English or Bengali orthography
-		\item A picture displayed for 50ms; named by the participant in either English or Bengali, depending on the language cue 
-		\item Target -- an English word; read out by the participant
-		\item Distracter math problem
-	\end{itemize}
-	Since all the target words were in English, the language cue was used to manipulate the \textsc{preceding} word, giving target words in non-switched (unilingual) and switched (mixed) utterances.
-The orthographies of English and Bengali are visually distinct, and were therefore used to cue language.
-
-	\item Code-switching: Each slide contained one target word embedded in either an English (unilingual) or a Bengali (mixed; target produced as a switch from Bengali) carrier sentence which was read out by the participant.
-The carrier sentences are given below:
-	\begin{itemize}
-		\item Unilingual sentence: That is a yellow [\textit{target word}].
-		\item Mixed-language sentence: \textipa{o-\:ta \ae k-\:ta kalo} [\textit{target word}]. 
-	\end{itemize}
-
-Gloss and translation for Bengali sentence:
-
-o-\textipa{\:t}a \qquad \textipa{\ae}k-\textipa{\:t}a \quad kalo [\textit{target word}]\\
-that-\textsc{clf}  one-\textsc{clf}  black [\textit{target word}]\\
-``that is a black [\textit{target word}]"\\
-
-The sound preceding the target word was uniform across all sentences (the final sound in \textit{yellow} is produced as [o] in Indian English), to minimize any possible differences due to coarticulatory effects on the target vowel.
-A specifier-noun construction was used to ensure identical word order in both languages, giving a uniform prosodic context for the target word across condition.
-In presenting the mixed-language utterances, we used mixed orthography.
-\\
-
-\end{enumerate}
-\subsection{Acoustic analysis}\label{analysis}
-The first iteration of each task was treated as a trial, and excluded from further analyses.
-An additional 67 tokens (2.79\%) were excluded due to errors in presentation/production, giving 2333 target word tokens (\nt{2}: 1137, \nt{\ae}: 1196) for analysis.
-Recordings were manually segmented and annotated in Praat \citep{boersma2016praat}, and the first and second formant frequencies (F1 and F2) of the target vowels \nt{2, \ae} were measured at five points-- in 10\% increments starting at 5\% into the vowel.
-This gave measures of F1 and F2 at 5\%, 15\%, 25\%, 35\% and 45\% into the vowel, allowing us to observe any dynamic changes in the extent of transfer over the course of producing the vowel.
-Individual vocal tract differences affect formant frequencies across speakers.
-To avoid this confound, we used Lobanov normalization (using the PhonR package \citep{phonR} in R \citep{r}, representing each speaker's formant frequencies as the scaled distance from their own mean (\alert{citation for lobanov normalization?}).
-Plotting the vowels \alert{(figure)} shows a difference between male and female speakers, such that male speakers have overall lower formant frequencies.
-Speaker normalization using the Lobanov method does not get rid of this difference.
-Since gender-related differences in formant frequencies are not of interest in this study, we normalized individual values by gender (representing each value as a scaled distance from the group mean).
-Statistical analyses were performed on these normalized values.
-
-\section{Results} \label{results}
-
-All statistical analyses were performed in R \citep{r}.
-We used linear mixed effects models to test the effect of experimental conditions on vowel height (F1) and backness (F2) respectively.
-Since formants were extracted from five points over the course of articulating the vowel, time was coded numerically (1-5), and treated as an independent variable.
-This was scaled and centered prior to analysis and treated as a continuous predictor.
-We fitted models using the \emph{lme4} package \citep{lme4}, and used \emph{lmerTest} \citep{lmerTest} to obtain estimates of p values for individual variables.
-For all mixed models, the alpha criterion was set at $|t| > 2 $.
-The dependent variables were normalized F1 and F2, and we fitted separate models for these.
-The independent variables or fixed factors, along with their levels, are given in table \ref{table variables} \alert{add variable type with each: factor, numerical, ...}.
-Subject and Item were treated as random factors.\\
-
-< Insert Table \ref{table variables} about here > \\
-
-\subsection{Models}
-
-\subsubsection*{Fixed effects}
-
-Our study involved two IE vowels. Vowel category is expected to systematically affect formant frequencies (FFs).
-
-The experimental treatment of interest is Context, i.e.
-whether the target item was presented in a unilingual (English) or mixed (Bengali) context.
-Our main research question (\bf{L2 vowels}) asks whether L2 vowel quality (formant frequencies) of proficient bilinguals is affected by the Context of utterance.
-We hypothesize that the direction of shift in the mixed condition is towards the corresponding L1 category.
-Thus, we expect the effect of Context on formant frequencies, i.e.
-the direction of shift, to be affected by the vowel category (c.f.
-section \ref{asymmetry between sounds}).
-Thus, our main effect of interest is the interaction term Context*Vowel.
-This corresponds to the transfer pattern.
-
-Our third research question (\bf{Paradigm}) asks whether the language-mixing paradigm, or Task, affects the extent and direction of transfer.
-If this is the case, we expect Task to interact with the transfer term, giving a three way Task*Context*Vowel interaction.
-
-Since we measured FFs at five points in the vowel, Time is expected to affect the FF.
-However, the precise effect of formant dynamics is expected to depend on the vowel category.
-Thus, we expect the interaction of Vowel*Time to predict formant frequency.
-If transfer patterns change across time, then we should expect it to interact with the transfer term, giving a Context*Vowel*Time interaction.
+Studies of short-term phonetic transfer consistently find differences between transfer patterns in L1 vs L2.
+The bulk of this literature is based on the feature VOT.
+While vowels have been shown to induce L2-to-L1 influence, the present study demonstrates L1-to-L2 transfer in vowels.
 
@@
 
-\subsection{Vowel Height -- F1}
+We extracted formant estimates from five uniformly spaced time points in the vowel: from the time at 5\% into the vowel (at the start of articulation) to 45\% (at the mid-part, or steady state, which is expected to correspond most closely to the articulatory target for the category).
+As expected, w
+
+Vowel is not a significant predictor of F1. This is initially surprising, given that the phonological category is expected to systematically affect vowel quality. However, a
+
+Given that estimates for fixed effects are pooled across participants, this explains the lack of reliable differences between \nt{2} and \nt{\ae}.
+
+This indicates a uniform scaling of formant trajectories in the mixed condition, suggesting that the underlying acoustic targets for these vowels are different during non-switched vs mixed utterances.
 
@@
 
-First, to test hypothesis 1 (\bf{L2 transfer}), we fitted a model with vowel*time, and vowel*context as fixed effects.
+First, to test our first and second research questions (\bf{L2 vowels and Asymmetry}), we fitted a model with vowel*time, and vowel*context as fixed effects.
 The transfer term (vowel*context) is significant in this model (model coefficient/ unstandardized effect size ($\beta$)= -2.330e-01, standard error (SE)=  5.244e-02, t= -4.44, p=0.0002).
@@
 
-To test the hypothesis \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects.
+To examine the third research question, \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects.
 This three-way interaction term is significant ($\beta$= -1.955e-01, SE=  5.453e-02, t= -3.59, p= 0.0003).
 We also find a significant effect of task on FFs ($\beta$=3.300e-01, SE=2.668e-02, t=12.371, p=2e-16).
-Comparing this model to null models that lack the fixed effect Task ($\chi^2(4) = 572.61, p=  2.2e-16$) shows that the full model is better.
+Comparing this model to a null model that lacks the fixed effect Task ($\chi^2(4) = 572.61, p=  2.2e-16$) shows that the full model is better.
 Because Task has an independent effect on F1, and adding interactions increases model parameters, we fitted a second null model with Task as a separate fixed effect without the interaction term to test whether the interaction term is needed.
@@
 
-Thus, the optimal model was: f1 ~ time*vowel + vowel*context*task + time*task + (1 + context|word) + (1 + context + vowel|subject).
+Thus, the optimal model was: f1 $\sim$ time*vowel + vowel*context*task + time*task + (1 + context|word) + (1 + context + vowel|subject).
 Table \ref{table_f1} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model.
-\alert{p-vales smaller than 0.05 are in bold.}
-
-\subsubsection*{Interpretation}
-We extracted formant estimates from five uniformly spaced points in the vowel, from the time at 5\% into the vowel (at the start of articulation) to 45\% (at the mid-part, or steady state, which is expected to correspond to the articulatory target for the category).
-As expected, we see that F1 values are significantly predicted by Time in the model, reflecting th dynamic nature of articulation which causes acoustic variation over the course of time.
-The positive slope of the estimate for Time indicates that F1 increases, i.e.
-the two categories in the study show lowering over the course of articulation.
-Moreover, the significant Time*Vowel interaction shows that the F1 trajectory is different for the two vowels.
-This is expected, as the categories are expected to present different articulatory targets.
-
- Vowel is not a significant predictor of F1.
-This appears surprising, given that the phonological category is expected to systematically affect vowel quality.
-However, a boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two categories \nt{2, \ae} and the extent of height difference between them.
-Note that these measures are only from the English (unilingual) condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching.
-Given that estimates for fixed effects are pooled across participants, this explains the lack of reliable differences between \nt{2} and \nt{\ae}.
-The negative slope of the estimate for Context shows that F1 is lower, i.e.
-vowels are higher, in the unilingual (e) condition compared to the mixed (b) condition.
-Put more intuitively, vowels are lowered in the mixed condition.
-However, the lack significance indicates that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels.
-Instead, the significant interaction between Vowel and Context signals an asymmetrical shift in the categories \nt{2} and \nt{\ae} in the mixed-language condition.
-Specifically, the Vowel*Context interaction has a negative slope, showing that the lowering is greater in \nt{2}.
-
-The results also reveal a significant effect of Task, such that the sentence task (code-switching) leads to higher F1 values in both vowels relative to the cued picture-naming task as: vowels are lowered in the code-switching task, independent of experimental condition.
-This is surprising, as we have no aproiri reason to expect a uniform difference in vowel quality across tasks.
-A significant interaction with Time with a positive slope of the estimate shows that this effect is local: it is strongest at the beginning of the vowel, and reduces over the course of the vowel.
-This is consistent with our assumption that the unexpected Task effect is an artifact of the paradigm, possibly resulting from coarticulatory effects with the preceding sound in the carrier phrase.
-Of greater interest to us is the significant Vowel*Context*Task interaction, which indicates a difference in transfer patterns across paradigms.
-Specifically, the negative slope of the estimate suggests that transfer in the F1 (height) dimension are greater in the code-switching (s) task compared to the picture-naming (p) task.\\
-
-< Insert Figure \ref{boxplot_F1} about here >\\
-
-\subsection{Vowel backness -- F2}
+p-vales smaller than 0.05 are starred.
 
@@
 
-Thus, the optimal model is: vowel*time + vowel*context*task + task*time + (1+context|word) + (1+context+vowel|subject).
+Thus, the optimal model is: f2 $\sim$ vowel*time + vowel*context*task + task*time + (1+context|word) + (1+context+vowel|subject).
 Table \ref{table_f2} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model.
-\alert{p-vales smaller than 0.05 are in bold.}
+p-values smaller than 0.05 are in starred.
 
-Statistical analyses reveal that the vowel \nt{2} is lowered and fronted, while \nt{\ae} is fronted, in the mixed condition.
-Previous literature on phonetic transfer suggests that such shifts are not random, but rather directional and targeted.
-We want to understand how the direction of shift compares to corresponding Bengali categories.
-Figure \ref{vowels_e_b} summarizes the findings of the experiment ($F1XF2$ plots for vowel productions in the unilingual and mixed conditions, in comparison to the related Bengali categories from \cite{dutta2021}).
-This confirms our hypotheses (c.f.
-section \ref{asymmetry between sounds}) that \nt{\ae} moves towards the corresponding \nt{\ae} category in Bengali, whereas the shift in \nt{2} is towards the related Bengali category \nt{a:}.
-The following sections discuss the results for vowel height and backness respectively.
+For the present study, we focus on a variety of IE spoken by an educated elite, used in news channels, often taught as a target in classrooms, and generally considered to be free of any particular regional affiliation \citep[c.f.][who reports this variety to be ``a \textit{de-facto} norm", and thus calls it General(ized) Indian English]{masica1972sound}.
+This sound system is described in section \ref{vowel systems}.
+However, recent research comparing the segmental and suprasegmental properties of IE spoken in different parts of the country also report many commonalities (see \cite{sirsa2013effects} for a review).
+It has thus been suggested that in spite of the varied L1s of its speakers, IE has a target phonology that is distinct from any of these, as well as from other native varieties of English.
+Thus, it is fruitful to think of regional variations as resulting from L1-influence on a common underlying target.
+Excepting a small minority of specific communities \citep{pandey201517, wells1982accents, coelho1997anglo}, there are very few L1 speakers of IE.
+Thus, English is largely learned and used as an L2.
 
-< Insert Figure \ref{vowels_e_b} about here >
+Specifically, a pattern that emerges from the body of work on VOT is that a shift in long-lag VOT towards the short-lag norms of the other language is much more consistent and systematic than the converse \citep{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid}.
+In general, short-lag VOTs seem to resist accommodative shift.
+\cite{bullock2009trying} suggest that because long-lag languages offer a greater range of acceptable VOTs, there is more `room' for movement.
+In contrast, shift in short-lag languages could risk the loss of a phonological contrast (VOT being the primary cue for voicing contrast in these languages).
+This is a language-specific phonological constraint on transfer.
+In a similar vein, \cite{antoniou2011inter} observe that while the VOT of English stops shifted towards Greek (a short-lag language), the reverse was seen only in a specific subset of sounds: word-medial non-nasalized stops.
+They attribute this to the low frequency of these sounds in Greek, making them less ``stable", and thus more amenable to cross-language influence.
+This suggests that the distribution of sounds in a language could affect the kind of transfer patterns observed.
 
-\subsection*{Interpretation}
+Since Bengali has a four-way laryngeal contrast, VOT does not encode a binary voiced-voiceless distinction.
+Moreover, VOT is not the only relevant cue for the corresponding phonological categories in the language \citep{dmitrieva2020acoustic}.
+This means that any effect of L1 Bengali on L2 English VOT is likely to be affected by additional phonological factors.
+Thus, avoiding any potential confounds was another reason we chose to focus on vowel quality as a target for transfer in the present study.
 
-As with F1, we find that Time significantly influences F2 values in the model, reflecting the dynamic nature of articulation which causes acoustic variation over the course of time.
-Moreover, the Time*Vowel interaction shows that this trajectory differs across the two vowels: while the positive slope for Time indicates that F2 increases, i.e.
+Research comparing bilingual speakers' phonological systems to monolinguals treats them as relatively stable over time.
+The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods \citep[e.g.][]{guion2003vowel, caramazza1973acquisition}.
+
+over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts.
+For example, \cite[][vowel quality]{bohn1992production}
+observed that approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language influence.
+
+F2 model interpretation:
+As with F1, we find that Time significantly influences F2 values in the model, reflecting the dynamic nature of speech which causes acoustic variation over the course of articulation.
+Moreover, the Time*Vowel interaction shows that this trajectory differs across the two vowels: while the positive slope for Time indicates that F2 increases i.e.
 vowels are fronted with time, the negative slope of the Time*Vowel interaction shows that this effect is smaller for \nt{2}.
-This is expected, as figure \alert{ref} confirms that the category \nt{\ae} is overall more fronted than \nt{2}.
+This is expected, as figure \ref{vowels_unilingual_ie_bengali} confirms that the category \nt{\ae} is overall more fronted than \nt{2}.
 
-Unlike in the F1 dimension, Vowel is a significant predictor of F2, with \nt{2} having significantly lower F2, i.e. produced as more backed, than \nt{\ae}.
+Unlike in the F1 dimension, Vowel is a significant predictor of F2, with \nt{2} having significantly lower F2, i.e.
+being produced as more backed, than \nt{\ae}.
+As with vowel height, there is no significant independent effect of Context, showing that mixed language use does not lead to uniform fronting or backing across vowels.
+This is not surprising, as we expect movement to be towards the corresponding L1 category, and thus differ across vowels.
+The negative slope of the estimate indicates that on average, vowels have lower F2 i.e.
+are backed, in the nonswitch (e) condition.
+In other words, vowels are fronted in the mixed condition.
+This effect is greater in \nt{\ae}, as indicated by the negative slope of the Vowel*Context interaction.
+In the F2 dimension, too, the lack of a significant Vowel*Context*Time interaction shows that transfer patterns do not change across time, suggesting that the underlying acoustic targets are different in the two conditions.
 
-As with vowel height, there is no significant independent effect of Context
+We see a significant effect of Task on F2, such that vowels in the code-switching (s) task have lower F2, i.e.
+are more backed than those in the picture-naming (p) task.
+As with F1, this is an unexpected result; we have no principled reason to expect code-switching to induce an across-board backing of vowels.
+Vowels are backed in the code-switching task, independent of experimental condition.
+A significant interaction with Time, with a positive slope for the estimate, shows that this effect is local: it is strongest at the beginning of the vowel, and reduces over the course of the vowel.
+This aligns with the possibility that the unexpected Task effect is an artifact of the paradigm and the particular stimuli used.
+The Vowel*Context*Task interaction is of interest, as it relates to our research question of how the test paradigm affects transfer.
+Unlike vowel height, we do not find patterns of shift in vowel backness (F2) to differ across paradigms, as indicated by the lack of a significant effect for the fixed effect.
 
-The estimates for Vowel show that excepting the beginning of the vowel, the category \nt{2} is consistently produced as more backed than \nt{\ae}.
-Unlike vowel height, Context has an independent effect on vowel backness, such that both the vowels in this study have higher F2 values in the Bengali (mixed) condition, i.e.
-they are fronted in the mixed condition.
-A significant interaction between Vowel and Context reveals an asymmetry in the extent of fronting between the two vowels -- the shift is greater in the case of \nt{\ae}.
-We see a significant effect of Task on F2, such that vowels in the Sentence task (code-switching) were more backed than those in the picture-naming task.
-As with F1, this is an unexpected result, as we have no principled reason to expect code-switching to induce an across-board backing of vowels.
-A lack of significant interaction with Context suggests that there is no difference in the extent of shift between the two paradigms.
-The results of ANOVAs comparing the full model to a null model that lacks the Vowel*Context interaction factor reveal that as with vowel height, the change in vowel backness between the two conditions is not significant at 5\%, and increases over the course of the vowel.\\
-
-< Insert Table \ref{table_f2_fixed_effects} about here >
-
-\subsection{Power analysis}
-We were able to recruit 10 participants (c.f.
-section \ref{participants}).
-To confirm whether the experiment design was powerful enough to detect the effects we are investigating, we carried out post-hoc power analyses using the mixedpower package \citep{kumle2021estimating} in R.
-The target of the analysis was to calculate the power for detecting an effect for our main variables of interest: vowel*context (to test the hypotheses L2-vowels and Asymmetry), and vowel*context*task (to test the hypothesis Paradigm).
-
-Power is a function of sample size and effect size.
-Thus, a key issue in power analyses is to estimate the expected effect size for the treatment/factor of interest \citep{brysbaert2018power, kumle2021estimating}.
-This may be expressed as a standardized effect size or estimate ($\beta$), Cohen's d, or partial $\eta ^2$.
-
-Using observed effect sizes from the data for power analyses is a bad practice and has been shown to be meaningless \citep{hoenig2001abuse, gelman2019observed_effect_size}, as such "observed power" depends on the p-value.
-Therefore, we need a reasonable estimate of effect size.
-One way to do this is to use effect sizes reported in other existing literature (however, see \cite{brysbaert2018power} who recommends against using effect sizes from single published studies, as these tend to be inflated), or meta analyses.
-An alternative approach is to define a smallest effect size that is meaningful to us (smallest effect size of interest; SESOI \citep{kumle2021estimating}), and test whether the study has enough power to detect such an effect, if it were to exist.
-This needs to be theoretically motivated.
-To the best of our knowledge, there is no existing literature on typical effect sizes for spectral differences in cross-language vowel production, and not many studies that report model estimates.
-Only more discussion among researchers in this field will lead to a consensus on what is considered a typical or minimally meaningful cross-language effect.
-
-Similarly, while differences between language-switching paradigms have been informally discussed in previous literature, we are not aware of any study that has measured these differences in a controlled experimental setup, and thus no reported effect sizes to act as a guideline.
-
-Cohen (cite) arbitrarily classifies effect sizes (cohen's d values) as small, medium, and large.
-Although there ar arguments against the meaningfulness of these categories \alert{citations}, these are often used as ballpark values for power analyses in psychological studies (for a discussion, see \cite{brysbaert2018power}).
-However, this approach does not generalize well to mixed-effects analyses, particularly for complex models with multiple random effects and interactions, as is typical in psycholinguistic and bilingualism research.
-"Determining the SESOI for (G)LMMs is difficult in a simulation-based approach where effect sizes are indicated through the model’s unstandardized beta coefficients...relating effect sizes to beta coefficients in complex models is far from trivial and the authors therefore refrain from making specific recommendations.
-\citep{kumle2021estimating}".
-
-The results of a power analysis are only as meaningful as the assumptions that go into it.
-To avoid making arbitrary assumptions, we followed the approach in \cite{brysbaert2018power} (and discussed in \cite{kumle2021estimating}) of taking effect sized of the study of interest and directly modifying all the beta coefficients by a fixed percentage.
-Note that we are not assuming that doing so gets us closer to a "true effect size".
-Instead, we run simulations with a range of effect sized and report the power of our study to plausibly detect an effect of that size.
-The results are shown in the plots in figure ...
-These show power over a range of sample sizes, given the assumed effect size.
-
---> in cover letter: power analyses for complex mixed models are not trivial and is an active pursuit.
-While we have done our best to follow available literature, we would love to have more feedback on this.
-
-\section{Discussion}
-
-Taken as a whole, these results bear on our original research questions in the following ways: regarding our first question, whether L2 vowels of proficient bilinguals can be affected by short-term transfer from L1 (\textbf{L2 vowels}), our results confirm that this indeed happens during mixed-language use.
-Both the common category \nt{\ae} and the L2-exclusive category \nt{2} show spectral shifts during mixed-language productions compared to unilingual productions.
-As hypothesized (\textbf{Asymmetry}), the results reveal asymmetries in the extent and direction of shifts between the two vowels.
-Specifically, \nt{2} is lowered and fronted, while \nt{\ae} is fronted, with the extent of shift being greater in \nt{2}.
-Comparision with canonical Bengali vowel productions shows that the shift in \nt{\ae} is towards the corresponding Bengali category, while that in \nt{2} is towards the Bengali category \nt{a:}.
-Regarding our third research question, whether paradigm used to elicit language switching independently affects patterns of transfer (\textbf{Paradigm}), analyzing the role of Task in the results confirms this.
-We find a greater degree of transfer during code-switching compared to the cued picture-naming task.
-However, the difference is small, and needs to be replicated (?).
-These results and their implications are discussed in the following subsections.
-
-\subsection{Phonetic transfer in L2 vowels}
-
-One of the main aims of this study is to understand short-term phonetic transfer in L2, towards understanding language processing in proficient bilinguals.
-Two interrelated questions, discussed in section \ref{introduction} (....) bear on this: (i) what causes transfer in mixed-language speech; and (ii) what causes observed asymmetries between transfer in L1 vs L2.
-Together, these serve as guidelines for how we posit that the language systems of th bilingual exist in relation to one another, and consequently, what a cognitive model should look like-- any model should be able to account for these observations.
-
--- what causes transfer: language mode vs switching
-
-\alert{discussion: We presented data in the same session, such that the language mode (participants' overall language ``setting" remained the same).
-The results show differences between between the switched and non-switched tokens in spite of this.
-this confirms that the local act of switching does cause transfer, even in a bilingual language mode, paralleling results from \cite{olson2016role, tsui2019impact, piccinini2015voice}.
-All of these examined the local effect of switching on a temporal feature of speech, namely, VOT.
-The results of the present study indicate that the same happens in vowels.
-Moreover, the result of transfer is interesting-- in VOT, it can only increase or decrease, which makes it impossible to know if it is the particular consonant that becomes more like its L1 counterpart, or if there is some more system-wide shift.
---* check discussion section of these three studies and see if they have talked about this --* Because vowel quality measures show directionality as well as magnitude, we can see finer patterns.
-We find that it is not merely a shift towards a Bengali counterpart, but rather a shift towards Bengali \textit{norms}-- a category that doesn't exist in Bengali moves towards a related Bengali category.
-This result parallels \cite{simonet2014phonetic}, who found the Catalan ...
-of bilingual speakers shifts towards a related L1 Spanish category ..
-in a bilingual language mode (although, this study did not examine switched speech).
-(check discussion section once) What does this say about what happens during online processing and speech planning?
---- look at discussion section of the mode vs switching papers (those three + ones that found no local effect of switching): what do they say about this?
-Can this finding be accounted for with that?}
-
-\alert{It makes sense that a bilingual language mode makes the representations from both languages available.
-But what do these representations look like, and how are these processed during the act of using language?
-Examining speech during switching can shed light on this.
-We argue that what we see here is an example of \textit{creativity}-- beyond a simple L2-L1 mapping, there is a method to the movement such that it makes reference to entities beyond the level of the individual sound-- i.e., the system-level.
-If representations are combined in creative ways, then are we truly ``moving between" languages, in a real sense, during online processing within a single discourse?
-Can we think of intermediate representations, not of individual sounds, but rather of sound systems?
-While more detailed empirical data will eventually enable us to map out patterns of transfer during switched speech, this interpretation is in line with the kind of model proposed by \cite{khattab2013phonetic} based on observation from the use of phonological variables in bilingual children (check-- this the one where they talked about moving between systems-- spectrum of choices?).
-This visualizes/conceptualizes the bilingual language system as a \textit{continuum of sound systems}, from which the bilingual speaker is constantly choosing, guided by a variety of different pressures.
-We propose that vowels are a good feature to examine: in addition to continuous measures (as opposed to phonological variables), they also give us the 2-dimensional data (as opposed to VOT) that is suited to further testing and refining such a model.}
-
---- differences between L1 vs L2 transfer:
-
-Studies of short-term phonetic transfer consistently find differences between transfer patterns in L1 vs L2.
-The bulk of this literature is based on the feature VOT.
-While vowels have been show to induce L2-to-L1 influence, the present study demonstrates L1-to-L2 transfer.
-
-Previous studies with VOT have often found a lack of parallel transfer from L1 to L2.
-One postulated model that is consistent with this finding is the ICM, first proposed by \cite{green1986control} at the level of the lexicon, but extended to sound systems by \cite{olson2013bilingual,tsui2019impact}.
-This proposes that ...\alert{read Olson2013 discussion, more about assumptions of ICM and consequences about models of processing}  As discussed in section \ref{causes}, vowels provide a fertile testing ground for this theory in sound systems.
-
-Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline unilingual productions.
-In the context of an ICM--based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, this is a significant finding: it demonstrates that the postulated lower switch cost on an L2 does not entirely mask transfer effects on L2 vowels.
-Moreover, our findings differ from \cite{tsui2019impact} who report on the basis of VOT measurements that balanced bilinguals do not evidence an effect of language-switching.
-\citeauthor{tsui2019impact} attribute the absence of transfer effects to greater experience with language switching in these participants, leading to lower switch-costs in both languages.
-We find that even proficient bilinguals with extensive experience of language switching demonstrate cross-language transfer effects in vowel production.
-Thus, following from the discussion in sec.
-\ref{asymmetries}, it might be fruitful for future studies to focus on vowels as a potential site for transfer.
-This would both expand the empirical scope of existing research (cf.
-sec.
-\ref{asymmetry between sounds}) and avoid confounding factors arising from independent differences between long- and short-lag VOT languages.
-More generally, these findings point to a need for expanding the range of features in which transfer effects are studied, in order to make robust generalizations about the mechanisms underlying observed patterns.
-
-\alert{this is an example where studying different features allows us to compare the predictions of different underlying process models.
-While most studies usually focus on a single feature due to ease of creating materials, there have been studies which have measured multiple features (e.g.
-..., or no need?).
-Measuring multiple features within a single design, with the same group of participants, would give a powerful tool to compare different models at little added cost to researchers.
-This would sidestep the unavoidable confounds that come with comparing results across studies, or even replication}
-
-\subsection{Asymmetries in Phonetic Transfer}
-The literature on short-term phonetic transfer has frequently noted differences in transfer patterns between different sound categories in a languages.
+The literature on short-term phonetic transfer has frequently noted differences in transfer patterns between different sound categories in a language.
 This has been attributed to both general linguistic principles such as contrast preservation, and language-specific factors such as frequency and distribution of categories.
 Examining these asymmetries more closely, and understanding their sources, helps us figure out what factors affect/moderate transfer and thus, by extension, what level of representation and processing this interaction happens at, and the nature of short-term intra-speaker variation.
-The present study examined transfer in two L2 categories, \nt{2} and \nt{\ae}. We hypothesized an asymmetrical shift between the two vowels (\textbf{Asymmetry}), owing to differences in:
+We examined transfer in two L2 categories, \nt{2} and \nt{\ae}, and hypothesized an asymmetrical shift between the two vowels owing to  differences in their position in the
 \begin{itemize}
 	\item their position in the Indian English vowel space: \nt{\ae} is in a more crowded part of the vowel space
- 	\item the presence/absence of a corresponding category in the L1 Bengali inventory
+	\item the presence/absence of a corresponding category in the L1 Bengali inventory: \nt{\ae} is a common phonological category in Bengali and English, while \nt{2} does not exist as a category in Bengali
 \end{itemize}
 
-The significant interaction between Vowel and Context in both F1 and F2 in the present study confirm that the experimental condition affected both height and backness differently for each vowel.
-Examining the model coeffieicnets ($\beta$) values shows that the transfer effect is greater for ...
-compared to ...
-(\alert{check interaction effect and unpack}) a greater degree of shift in \nt{2}, possibly owing to its position in a less dense part of the IE vowel space.
-The category \nt{2} is lowered and fronted in the mixed language condition as expected (Figure \ref{vowels_e_b}).
-The direction of shift in \nt{\ae} is surprising-- it is fronted in the mixed-language condition.
-While phonological constraints on IE suggested that a shift in the F1 dimension could potentially risk the loss of contrast and is thus expected to be minimal, Figure \ref{vowels_e_b} shows that \nt{\ae} in Bengali is not fronted with respect to the English category.
-Thus, fronting in the mixed condition cannot be explained in relation the former, at least from the present data.
-\alert{remove-- Recall that the Bengali vowel productions used here are from a different set of speakers than our participants.
-Assuming some amount of individual differences in the precise phonetic realization of vowel categories in both languages, this could be a potential source of explanation for this result.} Data from a larger number of speakers for both languages, as well as within-speaker comparisons of English and Bengali vowels, are required to clarify this.
-
-\alert{The fact that nt shows more shift-- shows that this influence is not just at the level of individual sounds (that way would expect more shift in the common category because it's "confusing".
---check what \cite{simonet2014phonetic} says about this, about why shift happens.
-But this-- shows that the L1 influence is at a system-level.
-More inclined towards using a Bengali-like phonological system.
-Has implications for cognitive models of language and what happens during short-term interaction)}
-
-Switching paradigm
-Our third research question \textbf{Paradigm} concerned the effect of experimental paradigm on the phonetic outcome of mixed-language use.
 \cite{olson2013bilingual} proposes that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", and thus proposes cued picture-naming as a preferable alternative.
 If such discursive properties independently affect the phonetic outcome of language mixing in an experimental setup, this calls to question the comparability of results across studies using these different paradigms.
-We verified this by comparing productions of the same items from the same group of participants in a scripted CS paradigm and a cued picture-naming paradigm.
-We found no significant interaction of Task either with Context, or with the Vowel*Context interaction, suggesting that the extent of phonetic transfer was not affected by the paradigm.
-There are at least two plausible reasons for this-- (i) it is possible that the planning effects of CS do not translate to an increase/decrease in observable transfer at all.
-This would mean that the pre-switch effects observed, for example, in \cite{muldner2019phonetics} and \cite{bullock2009trying} are better interpreted as resulting from some other aspect of the experimental setup, than the act of code-switching itself.
-While planning could additionally affect the words preceding the switch, the effects on the switched token itself do not change.
-(ii) Our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
-Thus, another possible explanation is that the discursive effects discussed by \cite{olson2013bilingual} are triggered by more naturalistic paradigms where attention to meaning and interlocutor are important.
-More within-study comparisons with less structured paradigms, and spontaneous code-switching data, can clarify this.
-The results of the present study indicate that for the purposes of observing dynamic phonetic transfer in an experimental setup, elicitation through cued-picture-naming vs.
-(controlled) code-switching does not significantly alter the behavior under study.
-Thus, the choice of paradigm can be based on other considerations.
-Moreover, this suggests that (all else being equal) the results from such studies can be compared directly.
-
-The paradigm did affect vowel quality, but this applied uniformly across experimental conditions: all vowels were more backed and lowered in the code-switching task.
-Given that at least for F1, the effect is limited to the early stages of articulating the vowel, it is plausible that this lowering is an artifact of the particular stimuli used here---resulting from some corarticulatory effect that was absent in the picture-naming task.
-While this is intriguing, it is beyond the scope of the present study.
-\\
-
-\section{Conclusion}
-This study examined the phonetic outcome of mixed-language production in a group of highly proficient bilingual speakers of Bengali and Indian English.
-Specifically, we examined the spectral quality of two L2 vowel categories \nt{2} and \nt{\ae} produced in two experimental conditions: unilingual (English) and mixed (switching from Bengali to English).
-Comparisons of F1 and F2 reveal that mixed language processing temporarily increases phonetic interaction between the languages, causing \nt{2} to converge towards a related L1 category \nt{a:}.
-The extent and direction of shift is asymmetrical across the two vowel categories, which parallels findings in prior studies.
-This asymmetry appears to be mediated by the language-specific distribution of vowel categories.
-These findings evidence an effect of phonetic transfer on L2 vowels for the first time, and raise questions for proposed inhibition-based accounts of cross-language phonetic interaction.
-Mixed productions were elicited in two switching paradigms: code-switching and cued picture-naming, to verify whether the postulated discursive elements in code-switching independently influence the outcome of phonetic transfer.
-Results suggest that in an experimental setup, elicitation through cued-picture-naming vs.
-(controlled) code-switching does not significantly alter the behavior under study.
-
-This study has a number of limitations: (i) the `baseline' values for Bengali vowel categories were estimated from a different set of speakers than the participants of this study, potentially obscuring the role of individual differences in the realization of vowel categories; within-speaker comparisons of L1 and L2 vowels is likely to offer a more fine-grained view of the direction of shift in relation to L1 categories.
-(ii) we used a frame sentence to elicit code-switched productions such that not all the sentences were pragmatically well-formed.
-This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study.
-Future work should address this by considering data from more naturalistic code-switching paradigms.
-
-\section*{Abbreviations}
-\textsc{clf} = classifier
-
-\section*{Data availability}
-The materials, data files, and analysis code for this study are openly available in OSF at:  \url{https://doi.org/10.17605/OSF.IO/DSB2X}.
-
-\section*{Competing interests}
-The authors declare none.
-
-%\printbibliography %for use with biblatex; comment out if you use natbib
-\bibliography{references} %for use with natbib; comment out if you use biblatex, and change 'sample' by the name of your bib-file
-
-\newpage
-\section*{Tables}
-\begin{table}[h]
-	\caption{Fixed effects}
-	%\centering
-	\resizebox{\textwidth}{!}{%
-		\begin{tabular}{@{}ll@{}}
-			\toprule
-			Variable name                                                 & Levels                                                            \\ \midrule
-			Gender                                                 & male, female                                                      \\
-			Vowel                                                  & \nt{2}, \nt{\ae} \\
-			Task (corresponding to the two paradigms)              & p (Picture: cued picture-naming, s (Sentence: code-switching)     \\
-			Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)                       \\
-			Time                                                   & continuous variable, range -1.6 to 1.6, scaled and centered\\
-			 \bottomrule
-		\end{tabular}%
-	}
-	\label{table variables}
-\end{table}
-
-\newpage
-
-% Please add the following required packages to your document preamble:
-% \usepackage{booktabs}
-\begin{table}[h]
-	\centering
-	\caption{Optimal model: F1}
-	\label{table_f1}
-	\begin{tabular}{@{}lllll@{}}
-		\toprule
-		& Estimate  & Std. Error & t value & p value          \\ \midrule
-		(Intercept)           & -0.053474 & 0.178435   & -0.300  & 0.769            \\
-		time                  & 0.139140  & 0.011556   & 12.040  & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}                & -0.147480 & 0.223022   & -0.661  & 0.519            \\
-		contexte              & -0.013419 & 0.068357   & -0.196  & 0.846            \\
-		tasks                 & 0.330034  & 0.026586   & 12.414  & \textbf{< 0.001 ***} \\
-		time:vowel\textipa{2}           & -0.170892 & 0.013469   & -12.688 & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte       & -0.137062 & 0.060884   & -2.251  & \textbf{0.031 *}   \\
-		vowel\textipa{2}:tasks          & 0.009002  & 0.038498   & 0.234   & 0.815            \\
-		contexte:tasks        & 0.073341  & 0.037597   & 1.951   & 0.051 .          \\
-		time:tasks            & -0.121604 & 0.013469   & -9.028  & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte:tasks & -0.195613 & 0.054338   & -3.600  & \textbf{< 0.001 ***} \\ \bottomrule
-	\end{tabular}
-\end{table}
-
-\newpage
-
-% Please add the following required packages to your document preamble:
-% \usepackage{booktabs}
-\begin{table}[h]
-	\centering
-	\caption{Optimal model: F2}
-	\label{table_f2}
-	\begin{tabular}{@{}lllll@{}}
-		\toprule
-		& Estimate & Std. Error & t value & p value          \\ \midrule
-		(Intercept)           & 0.23202  & 0.22100    & 1.05    & 0.304            \\
-		vowel\textipa{2}                & -0.38596 & 0.14858    & -2.60   & \textbf{0.015 *}   \\
-		time                  & 0.04287  & 0.00761    & 5.63    & \textbf{< 0.001 ***} \\
-		contexte              & -0.01720 & 0.02823    & -0.61   & 0.548            \\
-		tasks                 & -0.08758 & 0.01752    & -5.00   & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:time           & -0.16289 & 0.00888    & -18.35  & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte       & 0.11043  & 0.02751    & 4.01    & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:tasks          & -0.08983 & 0.02530    & -3.55   & \textbf{< 0.001 ***} \\
-		contexte:tasks        & -0.10990 & 0.02477    & -4.44   & \textbf{< 0.001 ***} \\
-		time:tasks            & 0.06237  & 0.00888    & 7.03    & \textbf{< 0.001 ***} \\
-		vowel\textipa{2}:contexte:tasks & 0.03349  & 0.03557    & 0.94    & 0.34652          \\ \bottomrule
-	\end{tabular}
-\end{table}
-
-f1: contexte had -ve slope, vt also has -ve slope, so vt is moving up in e ie down in b condition.
-vowel*context has -ve slope, which means that vt is moving more than ae.
-Now, vowel*context*task has -ve slope, which means in the s task, compared to p, there is  more downward movement.
-So more transfer in s task.
-
-f2: vt has -ve slope, contexte has -ve slope, which means e condition e things are more back, ie vowels are fronted in b condition.
-But vowel*context has a +ve slope-- vt is less backed in the e condition that ae, ie it is less fronted in the b condition, ie ae is fronted more, or shows more transfer effect, than vt.
-vowel*context*task also has +ve slope, which means thatthe -ve effect of contexte are further reduced when the task is s, ie fronting in mixed condition is less in s task
-
-\newpage
-\section*{Figures}
-
-\begin{figure}[h] %\centering
-	\includegraphics[scale=1]{vowels_e_b_final}
-	\caption{Vowel quality in unilinual English, mixed, and baseline Bengali productions}
-	\label{vowels_e_b}
-\end{figure}
-
-\newpage
-
-\begin{figure}[h] %\centering
-
-	\includegraphics[scale=1]{vowel_by_subject_ggplot}
-	\caption{Subject-wise mean F1 in unilingual English productions}
-	\label{boxplot_F1}
-\end{figure}
-
-\end{document}
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: t
-%%% TeX-engine: luatex
-%%% End:

### 2022-10-11 draft_BLC/blc_revision1_reviewer_response.txt
bengali picture task figure

@@
 fig.
-2 in the text, which shows that speakers vary greatly in their default realizations of the two IE vowels), and visualize gender differences without confounds from individual differences (e.g.
+2 in the text, which shows that speakers vary greatly in their default realizations of the two IE vowels), and to visualize gender differences without confounds from individual differences (e.g.
 figure 6).

### 2022-10-11 draft_BLC/BLC_AM-ID_manuscript.tex
bengali picture task figure

@@
 \footnote{Since our experimental treatments were between-subject and statistical analyses use mixed-effects regression, this step would not affect the results, as pointed out by a reviewer.
-This was done to allow for more intuitive values and visualizations, rater than statistical necessity.
-The code for parallel statistical analysis using non-z-score-normalized values can be found in the OSF repository.} Plotting these vowels shows a difference between male and female speakers, such that male speakers have overall lower formant frequencies (figure \ref{gender_before}).
+This was done to allow for more intuitive values and visualizations, rather than statistical necessity.
+The code and analyses using non-normalized values can be found in the OSF repository.} Plotting these vowels shows a difference between male and female speakers, such that male speakers have overall lower formant frequencies (figure \ref{gender_before}).
 %Speaker normalization using the z-score method does not get rid of this difference.
@@
 		\centering
+		\includegraphics[scale=1]{images/picture_task_eng}
+		\caption{English (nonswitch) trial} \label{picture_task_english}
+	\end{subfigure}
+	\begin{subfigure}[t]{0.49\textwidth}
+		\centering
+		\includegraphics[scale=1]{images/picture_task_bengali} 
+		\caption{Bengali-English (switch/mixed) trial} \label{picture_trial_bengali}
+	\end{subfigure}
+	\caption{Picture-naming task sample sequence: language cue (not read out), picture (named as per lang cue), target (read out), distractor}
+\end{figure}
+
+\begin{comment}
+	\begin{figure}[!htb] \label{figure_picture_task}
+	\centering
+	\begin{subfigure}[t]{0.49\textwidth}
+		\centering
 		\includestandalone[mode=image]{picture_task_eng}
@@
 \end{figure}
-
+\end{comment}
 \newpage

### 2022-10-19 draft_BLC/blc_revision1_reviewer_response.txt
2nd round revisions final

--- 2022-10-10 draft_BLC/blc_revision1_reviewer_response.txt
@@
-Manuscript BLC-21-0133.R1
+Manuscript BLC-21-0133.R2
 Response to Reviewers
@@
 We have removed the power analysis from the main manuscript.
-As per the editor's suggestion, we have included it as Supplementary Material.
-As suggested, we have acknowledged the small sample size and need for replication in the main text, on page ...
+As per the Editor's suggestion, we have included it as Supplementary Material.
+As suggested, we have acknowledged the small sample size and need for replication in the main text, on page 19.
 
@@
 
-AR: We have now added a schematic figure showing examples of a nonswitched and a switched trial (figure 4, page...). The trials proceeded as follows:
+AR: We have now added a schematic figure showing examples of a nonswitched and a switched trial (figure 4, page 51). The trials proceeded as follows:
 (i) see a word in Bengali/English orthography. This is the cue. It is NOT read aloud.
-(ii)see a picture of a fish. Based on the language cue in the previous slide, name it aloud in either English (say "fish") or Bengali (say "matS").  
+(ii) see a picture of a fish. Based on the language cue in the previous slide, name it aloud in either English (say "fish") or Bengali (say "matS").  
 (iii) see an English word, read it aloud. This is the target. If the previous slide was named in English this counts as a nonswitch trial; if it was named in Bengali this counts as a switched trial. 
@@
 We have replaced this with the term "z-score normalization".
-The target was to represent all formant values on a uniform scale across speakers, removing speaker-specific value ranges, so that the relative positions of the two vowels and shift between conditions could be compared directly.
+The target was to remove speaker-specific ranges in formant values, so that the relative positions of the two vowels and shift between conditions could be compared directly.
 Since the data had productions of the same two categories from all speakers, we felt comfortable that this would not affect the interpretability of the values.
@@
 fig.
-2 in the text, which shows that speakers vary greatly in their default realizations of the two IE vowels), and visualize gender differences without confounds from individual differences (e.g.
-figure 6).
+7 (pg.
+54) in the text, which shows that speakers vary greatly in their default realizations of the two IE vowels), and to visualize gender differences without confounds from individual differences (e.g.
+figure 6, pg.
+53).
 Since we were expecting high individual variability, we did this before running any further analyses.
-We agree that given the specific statistical tools used here, this was not essential to the statistical analyses.
+We agree that given the specific statistical methods used here, this was not essential for the statistical analyses.
 We re-ran the analysis with non-normalized values to verify this, and found identical results.
 We have included the code and output from both analyses in the OSF repository for the project, and added a footnote in the main text addressing this (pg.
-...).
+22).
 
@@
 
-AR: We have modified our Research Questions and Hypotheses section on the basis of these comments (pg. ...).
+AR: We have modified our Research Questions and Hypotheses section on the basis of these comments (pg. 16).
 
@@
 For the same reason, data from native multilingual populations is valuable, since these settings provide obvious communicative motivations for transfer.
-We have revised the wording in the paragraph, and hope it comes across more clearly now.
+We have revised the wording in the paragraph (pg.
+31), and hope it comes across more clearly now.
 

### 2022-10-19 draft_BLC/BLC_AM-ID_manuscript.tex
2nd round revisions final

@@
 
-Word count (7.10.2022): 9650
-
 \begin{singlespace}
@@
 \begin{myparindent}{0pt}
+
+Word count (17.10.2022): 9289 \\
+
 Short title: Phonetic transfer\\
@@
 
-EFL University,
+New York University,
 
-Hyderabad,
-
-India\\
+New York\\
 
@@
 
-Kolkata,
-
-India\\
+Kolkata\\
 
 Acknowledgments:\\
-We are grateful to Maumita Bhowmik and Anannya Mondal for their help with data collection and processing, and to all the participants of the study.\\
+We are grateful to Maumita Bhowmik and Anannya Mondal for their help with data collection and processing, and to all the participants of the study.
+We also thank the anonymous reviewers at BLC for their detailed feedback and engagement with the contents of this paper.
+This study was conducted at the English and Foreign Languages University (EFLU), Hyderabad.
+\\
 
@@
 
-EFL University,
+Department of linguistics,
 
-Hyderabad- 500007,
+New York University,
 
-auromita.mitra@gmail.com
+10 Washington Pl,
+
+New York, NY --- 10003
+
+auromita.mitra@nyu.edu
 
@@
 The present study is concerned with the latter.
-In production, this kind of influence has been variously termed \textsc{transfer, drift, accommodation,} and \textsc{interference}.
+In production, this kind of influence has been variously termed \textsc{``transfer'', ``drift'', ``accommodation'',} and \textsc{``interference''}.
 We use the term \textsc{``transfer"} here, to indicate any interaction between two sets of phonetic norms.
@@
 A majority of these studies use temporal properties of consonants (in particular, voice onset time or VOT) to measure transfer.
-However, reported results in the literature vary greatly across studies, as discussed in the following subsections, and appear to be contingent upon both language-specific features and language experience of the participants.
+However, reported results vary greatly across studies, as discussed in the following subsections, and appear to be contingent upon both language-specific features and language experience of the participants.
 This emphasizes the importance of considering data from a wider variety of
@@
 populations, language pairs,  and phonetic features in order to make meaningful generalizations.
-Widespread multilingualism in the Indian subcontinent suggests that phonetic  behavior in these populations can be particularly valuable towards understanding the nature of such cross-language interactions, as they are likely to reflect real-world experience with mixed-language processing.
+Widespread multilingualism in the Indian subcontinent suggests that phonetic  behavior in these populations can be particularly valuable towards understanding the nature of such cross-language interactions, as it is likely to reflect real-world experience with mixed-language processing.
 However, there is no work yet on short-term phonetic transfer in the Indian subcontinent, or in any Indo-Aryan language.
@@
 We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction.
-The results demonstrate a time-invariant effect of transfer on L2 vowels during mixed-language production, and parallel pattern but different magnitude of transfer between the two paradigms.
+The results demonstrate a shift in L2 vowel quality during mixed-language production, with parallel patterns but different magnitude of transfer between the two paradigms.
 We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.
@@
 
-``Indian English" is not a monolithic entity; there are large regional variations in the English speech of Indian speakers, and the literature on IE phonology is thus neither ``standardized" nor uncontested.
+Indian English is not a monolithic entity; there are large regional variations in the population, and the literature on IE phonology is thus neither ``standardized" nor uncontested.
 Recent research suggests that IE has a target phonology that is distinct from the varied L1s of its speakers, as well as from other native varieties of English (see \cite{sirsa2013effects} for a review).
 For the present study, we focus on ``General(ized) Indian English" \citep{masica1972sound} (\ref{vowel systems}).
-
 We focus specifically on short-term phonetic interaction during mixed-language use, because:
@@
 \subsubsection*{Vowel systems} \label{vowel systems}
-The vowel inventory of Western Bangla consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \citep{garry2001facts}.
+The vowel inventory of Western Bangla consists of \nt{i, e, \ae, a:, O, o, u}, and their nasalized counterparts \citep{garry2001facts}.
 The vowel chart in figure \ref{vowels_bengali} shows their distribution in the F1$\times$F2 formant space.
@@
 
-The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2/3:, a:, O, o, U, u} \citep{wells1982accents, masica1972sound}\footnote{\nt{@, 2, and 3:} are treated as a single phoneme in many descriptions of IE; the slashes between the symbols indicate possible notations for the same category}.
+The vowel system of IE contains the monophthongs \nt{I, i, E, e, \ae, @/2/3, a, O, o, U, u} \citep{wells1982accents, masica1972sound}\footnote{\nt{@, 2, and 3} are treated as a single phoneme in many descriptions of IE; the slashes between the symbols indicate possible notations for the same category.}.
 The vowel chart in figure \ref{vowels_ie} shows their distribution in the F1$\times$F2 formant space.
@@
 Overall, results suggest that: (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep[][vowel quality]{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use--- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \citep[][VOT]{olson2016role,tsui2019impact} or spontaneous conversation \citep[][VOT]{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
-(iii) How the two sources interact to influence the final outcome of transfer is not fully understood: \cite[][VOT]{olson2016role} found no additive effects of language mode, \cite[][VOT]{olson2013bilingual} found a balanced language context to inhibit transfer compared to unbalanced contexts.
-Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a separate monolingual test block, separated by a few hours to days (\cite{elias2017effects}; vowel quality), \citep[][VOT]{schwartz2015language, bullock2009trying,antoniou2011inter, vsimavckova2015immediate,vsimavckova2018patterns}.
+(iii) How the two sources interact to influence the final outcome of transfer is not fully understood: \citet[][VOT]{olson2016role} found no additive effects of language mode, \citet[][VOT]{olson2013bilingual} found a balanced language context to inhibit transfer compared to unbalanced contexts.
+Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a monolingual test block, separated by a few hours to days \citep[][vowel quality]{elias2017effects}, \citep[][VOT]{schwartz2015language, bullock2009trying,antoniou2011inter, vsimavckova2015immediate,vsimavckova2018patterns}.
 
@@
 Moreover, changes due to transfer are not necessarily unidirectional or irreversible.
-\cite[][VOT]{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language, showing that cross-language interaction can be triggered in an order of months, and also reversed within a similar time range.
-A more recent study by \cite[][VOT]{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks).
-Based on a short-term longitudinal study, \cite[][VOT]{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2.
-For some sounds and features (though not others, see sec.\ref{asymmetry between sounds}), transfer was additive over time.
+\citet[][VOT]{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language, showing that cross-language interaction can be triggered in an order of months, and also reversed within a similar time range.
+A more recent study by \citet[][VOT]{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks).
+Based on a short-term longitudinal study, \citet[][VOT]{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2.
+For some sounds and features (but not others, see sec.\ref{asymmetry between sounds}), transfer was additive over time.
 How long these effects last in the absence of regular L2 input was not tested.
@@
 \subsection{Asymmetries between languages in extent and direction of transfer} \label{asymmetries}
-A recurring pattern that emerges from existing research on short-term phonetic transfer is that the extent and patterns of phonetic shift are not equivalent across the two languages of the bilingual speaker.
-Studies vary greatly in the nature of differences that they report.
-The potential causes for such asymmetry are of interest because they point towards the factors that mediate cross-language phonetic interaction.
+A recurring pattern that emerges from the literature on short-term phonetic transfer is that the extent and patterns of phonetic shift are not equivalent across the two languages of the bilingual speaker.
+Studies vary greatly in the exact differences they report.
+The sources of such asymmetry are of interest because they point towards the factors that mediate cross-language phonetic interaction.
 In this section, we discuss two proposed sources of such asymmetry that are relevant to the present study:
@@
 \subsubsection*{L1 vs. L2 status} 
-Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} proposes that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another.
-Thus, the patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories.
-In non-switched production, this is evidenced through pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf.
-\textsc{equivalence classification}: \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.
+Flege's Speech Learning Model (SLM) \citeyearpar[][]{flege1995second,flege2007language} proposes that the sound categories of a bilingual speaker exist in a common phonological space--- in principle, both L1 and L2 categories can influence one another.
+Thus, patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories.
+In non-switched production, this is evidenced through pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast \citep[see equivalence classification:][]{flege1984limits,flege1987production}, and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.
 
-In mixed-language production, the role of language status is less clear--- studies have reported unidirectional influence of L1 on L2 \citep[][VOT]{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \citep[][VOT]{tsui2019impact, olson2013bilingual}, (\cite{elias2017effects}, vowel quality) bidirectional convergence \citep[][VOT]{bullock2009trying, olson2016role}, divergence \citep[][VOT]{bullock2009trying,vsimavckova2018patterns}, and no influence \citep[][vowel quality]{muldner2019phonetics}, \citep[][phonological process]{schwartz2015language}.
+In mixed-language production, the role of language status is less clear--- studies have reported unidirectional influence of L1 on L2 \citep[][VOT]{balukas2015spanish,antoniou2011inter,vsimavckova2015immediate,goldrick2014language}, L2 on L1 \citep[][VOT]{tsui2019impact, olson2013bilingual}, (\cite{elias2017effects}, vowel quality), bidirectional convergence \citep[][VOT]{bullock2009trying, olson2016role}, divergence \citep[][VOT]{bullock2009trying,vsimavckova2018patterns}, and no influence \citep[][vowel quality]{muldner2019phonetics}, \citep[][phonological process]{schwartz2015language}.
 Since most existing studies measure shifts in VOT, the observed asymmetries between languages have been variously explained either in terms of their L1 vs L2 status, or language-specific differences between long- and short-lag VOT languages.
@@
 They interpret this asymmetry in terms of the Inhibitory Control Model (ICM) \citep{green1998mental}: to select a phonetic realization from one language, the other must be \textsc{inhibited}.
-Greater inhibition is required on L1 while using an L2 than the reverse.
-This greater initial inhibition on L1 means that switching into L1 incurs a greater \textsc{switch cost} (more cross-language influence) than switching into L2.
-Lower switch cost results in a lack of visible cross-language transfer effects on L2.
+Greater inhibition is required on L1 while using an L2.
+This greater initial inhibition on L1 means that switching into L1 incurs a greater \textsc{``switch cost''} (more cross-language influence) than switching into L2.
+Lower switch costs translate to a lack of visible cross-language transfer effects on L2.
 \cite{tsui2019impact} report comparable results, but only in participants who were not equally dominant in both languages.
@@
 However, given the implications of the ICM, it is necessary to first establish that these L2 categories can indeed be affected by dynamic interference, particularly in bilinguals with extensive language-switching experience.
-The few existing studies on vowel quality \citep{simonet2014phonetic,muldner2019phonetics,elias2017effects} or phonological processes \citep{simonet2020increased,schwartz2015language} have examined transfer effects on L1 categories.
-As far as we are aware, there is no work yet on short-term transfer in L2 vowels.
-Thus, this study builds on the existing work by examining whether the L2 vowel quality of proficient bilinguals can be affected by mixed-language processing.
+There are relatively few studies on other phonetic features such as vowel quality \citep[see][]{simonet2014phonetic,muldner2019phonetics,elias2017effects} or phonological processes \citep[see][]{simonet2020increased,schwartz2015language}, and many of these focus on L1 categories as the target for transfer.
+%As far as we are aware, there is no work yet on short-term transfer in L2 vowels.
+This study builds on existing work by examining whether the L2 vowel quality of proficient bilinguals can be affected by mixed-language processing.
 
 \subsubsection*{Differences between sound systems of the languages} \label{sound systems} In addition to the cognitive factors discussed above, many researchers have attributed asymmetries to language-internal factors, such as the shape, size, and composition of the sound inventory.
-
-For example, the frequently reported asymmetries between long- and short-lag VOT languages \citep[see, e.g.][]{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid} and postulated frequency effects \citep{antoniou2011inter} highlight that beyond interaction during the cognitive processing of sounds, phonetic transfer is ultimately a linguistic phenomenon, subject to language-specific phonological constraints.
+For example, frequently reported asymmetries between long- and short-lag VOT languages \citep[see, e.g.][]{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid} and postulated frequency effects \citep{antoniou2011inter} highlight that beyond interaction during the cognitive processing of sounds, phonetic transfer is ultimately a linguistic phenomenon, subject to language-specific phonological constraints.
 Therefore, it is imperative to consider data from a variety of language pairs.
 Existing research on phonetic transfer largely focuses on a limited set of phonologically related languages.
-Thus, the present study extends the scope of this research to a new pair of languages--- Indian English and Bengali.
+The present study extends the scope of this research to a new pair of languages--- Indian English and Bengali.
 
@@
 Asymmetries in the extent and patterns of transfer have not only been observed between languages, but also between different sounds/features of a language, in both long-term and transient interactions.
-Studies that have examined multiple sound categories have found that interactions between individual sound pairs do not necessarily reflect the overall pattern of global (system-wide) shift \citep[e.g.][vowel quality]{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot treated as an atomic measure.
+Studies that have examined multiple sound categories have found that interactions between individual sound pairs do not necessarily parallel the overall pattern of global (system-wide) shift \citep[e.g.][vowel quality]{chang2012rapid,elias2017effects}, suggesting that `extent of transfer' cannot treated as an atomic measure.
 In light of the discussion about VOT in section \ref{sound systems}, this is not surprising --- if a general linguistic principle of `room for movement' and contrast constrains transfer, then we should expect it to apply to individual sounds too.
@@
 Any movement in \nt{\ae} is expected to be primarily in the backness (F2) dimension.
-	\item Target of transfer: findings from existing research on cross-language phonetic interaction suggest that changes in production during mixed-language use are not random, but rather targeted with respect to categories in the other language.
-Thus, another source of asymmetry is the fact that the category \nt{2} is absent in Bengali, whereas \nt{\ae} is a common category across both languages (see sec.\ref{vowel systems}).
+	\item Target of transfer: existing research on cross-language phonetic interaction suggests that changes in production during mixed-language use are not random, but rather targeted with respect to categories in the other language.
+Thus, another source of asymmetry is the fact that the category \nt{2} is absent in Bengali, whereas \nt{\ae} is a common category across both languages (see sec.
+\ref{vowel systems}).
 \end{enumerate}
@@
 
-In the present study, our research question concerns whether L2 vowels are produced differently in nonswitch vs switch conditions.
-Previous literature on phonetic transfer suggests that such shifts are not random, but rather directional and targeted.
-Although our primary research question merely concerns whether productions in the two conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories.
+Our research questions in this study concern whether L2 vowels are produced differently in nonswitch vs switch conditions.
+Existing literature suggests that such shifts are directional and targeted.
+Although our primary concern is whether productions in the two conditions differ, we wanted to understand how the direction of shift, if any, relates to corresponding Bengali vowel categories.
 To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we processed and used existing speech data from a freely available Bengali corpus, SHRUTI \citep{shruticorpus}.
-We elaborate on the corpus details and procedure in the Supplementary materials.
 
-We extracted formant values from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, and compared these to our participants' productions in the non-switched English condition.
+We extracted formant values from syllable-initial productions of \nt{\ae} and \nt{a:} at 55\% into the vowel, and compared these to our participants' productions in the non-switched English condition.
 This is visualized in figure \ref{vowels_unilingual_ie_bengali}.
 The F1$\times$F2 plot indicates that the Bengali category \nt{\ae} is more fronted, and \nt{a:} lower and more fronted, compared to the English categories \nt{\ae} and \nt{2} respectively.
+We elaborate on the corpus details and procedure in the Supplementary materials.
 
@@
 in connected speech \citep{myers1993dueling}.
-Other paradigms which have been used include cued picture-naming, delayed repetition, interpreting across languages, reading word-lists etc.
+Other paradigms which have included cued picture-naming, delayed repetition, interpreting across languages, reading word-lists etc.
 \cite{olson2013bilingual} argues that code-switching introduces discursive factors such as discourse planning and pragmatic considerations, which ultimately reflect patterns of language practice, rather than cognitive behavior.
-This finds support from studies which report transfer effects \textsc{before} the switch point \citep{bullock2009trying}, or other evidence for planning such as hyperarticulation \citep{muldner2019phonetics} and interlocutor-awareness in spontaneous CS \citep{khattab2013phonetic}.
+This finds support from studies that report transfer effects \textsc{before} the switch point \citep{bullock2009trying}, or other evidence for planning such as hyperarticulation \citep{muldner2019phonetics} and interlocutor-awareness in spontaneous CS \citep{khattab2013phonetic}.
 To avoid such confounds, \cite{olson2013bilingual} endorses cued picture-naming (where pictures are named in rapid succession in either language, based on the given language cue) as an alternative paradigm to isolate purely phonetic effects in controlled setups.
@@
 Thus, the presence of additional factors in CS does not necessitate that this should alter the extent of observable transfer compared to any other paradigm.
-It is possible that the outcomes in the studies discussed above are better attributed to some other aspect of the experimental setup, rather than to the act of code-switching in particular.
+It is possible that the outcomes in the studies discussed above are better attributed to some other aspect of the experimental setup, rather than the act of code-switching in particular.
 
@@
 This was especially important because we were not eliciting unilingual L1 Bengali data in this study, and were using an existing data source to estimate the position of Bengali vowels in the formant space.
-Based on previous literature (see section \ref{asymmetry between sounds}), we expect that L2 vowels shift towards related L1 categories during mixed language utterances.
+We expect that L2 vowels shift towards related L1 categories during mixed language utterances.
 Thus, we wanted to ensure that any between-participant asymmetries in the pattern of shift reflected individual differences in transfer, rather than different acoustic targets.
@@
 Table \ref{table_participants} summarizes some key biographical data from the survey.
-\alert{The complete questionnaire, along with responses, is available in the ``Data" component of the OSF repository for this project (see section \ref{data_availability}).
-A more detailed discussion of the linguistic situation and language profiles is included as supplementary material.--footnote?}
+The complete questionnaire and responses are available in the Data component of the OSF repository for this project (see section \ref{data_availability}).
+A more detailed discussion of the linguistic situation in India and language profiles of participants is included as Supplementary material.
 
@@
 Given that small sample sizes are likely to have low statistical power, replication with larger sample sizes are important to confirm the observed effects, and provide much-needed data diversity.
-A more detailed discussion of power analyses in bilingualism research is included as supplementary material on OSF (see section \ref{data_availability}).
-We hope that this can facilitate more conversation around establishing norms for power and sample size in our field.
+A more detailed discussion of power analyses in bilingualism research is included as Supplementary material.
+We hope that this can facilitate more conversation around establishing norms for statistical power and sample size in our field.
 
 \subsection{Stimuli}
-Twenty monosyllabic English words: ten containing the vowel \nt{2} and ten containing \nt{\ae}, were selected as target words.
+Twenty monosyllabic English words, ten containing the vowel \nt{2} and ten containing \nt{\ae}, were selected as target words.
 To minimize coarticulatory effects on the vowel, the pre-vocalic consonants were all voiced plosives (\nt{b} or \nt{d}).
@@
 \begin{enumerate}[]
-	\item Cued picture-naming: Participants were presented with slides in the following sequence (see fig. \ref{figure_picture_task}):
+	\item Cued picture-naming: Participants were presented with slides in the following sequence (see fig. \ref{picture_task_english},\ref{picture_trial_bengali}):
 	\begin{itemize}
-		\item Language cue: a word in either English or Bengali orthography, not read aloud
+		\item Language cue: a word in either English or Bengali orthography, read silently
 		\item A picture displayed for 50 ms; named aloud by the participant in either English or Bengali, depending on the language cue.
@@
 
-<insert fig. \ref{figure_picture_task} about here>
+<insert fig. \ref{picture_task_english},\ref{picture_trial_bengali} about here>
 
@@
 
+< Insert Table \ref{table_f1} about here >
+
 \subsection{Vowel backness -- F2}
@@
 
-.
+\vspace{3cm}
 
 Statistical analyses reveal that the vowel \nt{2} is lowered and fronted, while \nt{\ae} is fronted, in the mixed condition.
-As discussed in section \ref{asymmetry between sounds}, although our primary research question is about whether productions in switch vs nonswitch conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories.
+As discussed in section \ref{asymmetry between sounds}, we wanted to understand how the direction of this shift compares to corresponding Bengali categories.
 Figure \ref{vowels_unilingual_ie_bengali} shows the non-switched IE productions of the participants in relation to average positions of related vowel categories in Bengali.
@@
 Given that the Bengali corpus data represent average vowel positions, and are not from the same set of speakers as the participants of this study, we refrained from including these in statistical analyses.
-However, figure \ref{vowels_e_b} provides a heuristic confirmation for our informal hypotheses in section \ref{asymmetry between sounds}) that the IE \nt{\ae} moves towards the corresponding \nt{\ae} category in Bengali, whereas the shift in \nt{2} is towards the related Bengali category \nt{a:}.
+However, figure \ref{vowels_e_b} provides a heuristic confirmation for our informal hypotheses in section \ref{asymmetry between sounds} that the IE \nt{\ae} moves towards the corresponding \nt{\ae} category in Bengali, whereas the shift in \nt{2} is towards the related Bengali category \nt{a:}.\\
 The following sections discuss the implications of these results.
@@
 
-\paragraph{\textbf{What causes transfer: language mode vs switching}}
+\paragraph{\textbf{What causes transfer---language mode vs switching:}}
+
 In this study, we presented nonswitched English and code-switched Bengali-English stimuli in the same session and found consistent differences between switched and non-switched tokens.
@@
 This parallels results from \cite{simonet2014phonetic} for Catalan and Spanish (they examined language mode, without code-switching).
-The study showed that co-activating multiple languages within a discourse causes a transient increase in cross-language phonetic interaction, leading to changes in both L1 and L2.
+The \cite{simonet2014phonetic} study showed that co-activating multiple languages within a discourse causes a transient increase in cross-language phonetic interaction, leading to changes in both L1 and L2.
 This is interpreted as a follows: a bilingual mode temporarily co-activates both language systems, making both L1 and L2 representations available during processing, whose interactions lead to cross-language transfer.
 But what does this interaction look like, and how are these abstract representations processed during speech?
-Examining transfer during switching can provide insights.
+Examining transfer during switching can provide insights into this.
 
-The time course of such shifts is telling: we see a difference between the conditions right from the start of articulation, suggesting different acoustic targets.
+The time course of such shifts is telling: we see a difference between conditions right from the start of articulation, suggesting different acoustic targets.
 This suggests that such interaction involves the speech planning stage.
@@
 While more detailed empirical data is needed to answer such questions, this approach aligns with sociophonetic understandings of bilingual phonetic variation, for example as discussed by \cite{khattab2009phonetic,khattab2013phonetic} based on the use of phonological variables by Arabic-English bilingual children.
-She visualizes the bilingual phonetic system as ``a rich store of phonological representations", comprising of both language-internal and cross-language variation, from which the bilingual speaker is constantly choosing, guided by a variety of pressures.
+They visualize the bilingual phonetic system as ``a rich store of phonological representations", comprising of both language-internal and cross-language variation, from which the bilingual speaker is constantly choosing, guided by a variety of pressures.
 We propose that vowels are a good feature to examine the plausibility of such systems as cognitive models for language processing: in addition to continuous measures (unlike phonological variables), they also provide two-dimensional data (unlike VOT) that is suited to further test and refine predictions from such a model.
@@
 This proposes that the lack of transfer effects on L2 reflects the lower inhibition required on L2 representations in the first place, which makes switching into L2 easier, compared to the converse situation, when the high initial inhibition on L1 induces phonetic transfer while switching back into L1.
-As discussed in section \ref{causes}, vowels provide a fertile testing ground for this theory in sound systems, as they avoid the confound of independent language-specific differences between long- and short-lag VOT languages.
+As discussed in section \ref{causes}, vowels provide a fertile testing ground for this theory at the level of sounds, as they avoid the confound of independent language-specific differences between long- and short-lag VOT languages.
 Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline non-switched productions.
 In the context of an ICM-based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, this is a significant finding: it demonstrates that the postulated lower switch cost on an L2 does not entirely mask transfer effects on L2 vowels.
-There are a couple of possible explanations for this.
-To begin with, the present study focuses solely on L1 (Bengali) to L2 (English) transfer, and not the converse.
-It is theoretically possible that these participant are L2 English dominant.
+There are a few possible explanations for this.
+The present study focuses solely on L1 (Bengali) to L2 (English) transfer, and not the converse.
+It is theoretically possible that these participants are L2-dominant.
 That would mean that switching from L1 into L2 induces a greater switch cost, leading to transfer effects on switched L2 productions.
@@
 While more empirical data could clarify this, the language background of the present participant pool makes this scenario unlikely.
-All participants report regularly using both English and Bengali in a variety of settings, and feeling equally comfortable using bot languages.
-Listening and speaking skills are expected to be most relevant to phonetic transfer in daily language use scenarios, and participants' self-ratings for these skills are at ceiling for both languages.
-One point to note is that participants rated their reading proficiency in Bengali to be lower than that in English.
+All participants report regularly using both English and Bengali in a variety of settings, and feeling equally comfortable using both languages.
+Listening and speaking skills are expected to be most relevant to phonetic transfer in daily language use, and participants' self-ratings for these skills are at ceiling for both languages.
+One point to note is that participants on average rated their reading proficiency in Bengali to be lower than that in English.
 Given that the stimuli in this task were presented visually, differences in reading proficiency are a potential source of explanation.
@@
 \citeauthor{tsui2019impact} attribute the absence of transfer effects to greater experience with language switching in these participants, leading to lower switch-costs in both languages.
-As noted in \ref{participants}, the participants in this study are highly proficient bilinguals with extensive experience of language switching.
-Moreover, they live in an environment where they are constantly exposed to mixed-language stimuli, and are thus expected to be habituated to drawing on multiple language systems during online language processing.
+As noted in sec.
+\ref{participants}, the participants in this study are highly proficient bilinguals with extensive experience of language switching.
+Moreover, they live in an environment where they are constantly exposed to mixed-language stimuli: we expected that they are habituated to drawing on multiple language systems during online language processing.
 They nonetheless demonstrate cross-language transfer effects in L2 vowel production.
@@
 \ref{asymmetry between sounds}) and avoid confounding factors arising from independent differences between long- and short-lag VOT languages.
-More generally, these findings point to a need for expanding the range of features in which transfer effects are studied, in order to make robust generalizations about the mechanisms underlying observed patterns.
+More generally, these findings point to a need for expanding the range of features in which transfer is studied, in order to make robust generalizations about the mechanisms underlying observed patterns.
 This is an example where studying different features of speech (e.g.
 VOT vs vowel quality) allows us to compare the predictions of different underlying process models.
-Transfer studies usually focus on a single feature due to ease of creating materials.
+Transfer studies usually focus on a single feature.
 Measuring multiple features within a single design, with the same group of participants, promises to be a powerful tool to compare different models at little added cost to researchers.
@@
 Beyond methodological considerations, we argue that these differences ultimately raise a broader question about a purely inhibition-based understanding of transfer.
-In spite of a compelling body of literature on the sociophonetic and communicative implication of phonetic variation along the L1-L2 continuum \citep[e.g.][]{khattab2013phonetic,agnihotri1979processes,stuart2011hybridity}, is it wise to think of transfer as resulting from an inability to suppress and separate language systems?
+In spite of a compelling body of literature on the sociophonetic and communicative implications of phonetic variation along the L1-L2 continuum \citep[e.g.][]{khattab2013phonetic,agnihotri1979processes,stuart2011hybridity}, is it wise to think of transfer as resulting from an inability to suppress and separate language systems?
 Given that young children are able to use minute phonetic variation in sophisticated ways, it might be more intuitive to think of transfer as a communication-enabling feature of language, even at a purely cognitive level.
@@
 In contrast, \nt{2} (indicated as \nt{@} in the chart) is in the central part of the vowel space and is not flanked by any immediate neighboring categories.
-Thus, the difference in magnitude of transfer aligns with our hypothesis \textbf{H2---Asymmetry} that the extent of shift in individual categories is constrained by the linguistic principle of contrast preservation.
+Thus, the difference in magnitude of transfer aligns with our hypothesis \textbf{H2---Asymmetry}, that the extent of shift in individual categories is constrained by the linguistic principle of contrast preservation.
 An alternative way to interpret this result is that the initial distance between IE \nt{2} and its target L1 category \nt{a:} was greater than that between the IE and Bengali \nt{\ae} categories, leading to a greater visible shift in the former.
@@
 This suggests that a general principle of ``room for movement" does play a role in moderating transfer.
-More broadly, these results collectively suggest that phonetic transfer (i) makes reference to system-level representations that go beyond individual sounds; and (ii) respects linguistic principles.
+More broadly, these results suggest that phonetic transfer (i) makes reference to system-level representations that go beyond individual sounds; and (ii) respects linguistic principles.
 This has implications both for our understanding of phonetic transfer as a phenomenon (e.g.
@@
 We found that the direction of transfer did not change across paradigms.
-However, there was a difference in the magnitude of transfer.
-Specifically, in the F1 dimension, the extent of phonetic transfer is greater in the code-switching task compared to the cued picture-naming task.
+However, there was a difference in the magnitude of transfer: in the F1 dimension, the extent of phonetic transfer is greater in the code-switching task compared to the cued picture-naming task.
 This asymmetry is not seen in the F2 dimension.
@@
 
-We do not see any difference in the pattern of transfer, or the relative behaviors of the two categories.
+We do not find any differences in the pattern of transfer, or the relative behaviors of the two categories.
 Our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
@@
 Given that it is localized at the beginning of the target vowel, it seems likely that this lowering is an artifact of the particular stimuli used here--- resulting from some corarticulatory effect with the preceding word in the carrier phrase that was absent in the picture-naming task.
-While intriguing, a more detailed analysis of this effect is beyond the scope of the present study.
+While intriguing, a more detailed investigation of this effect is beyond the scope of the present study.
 
-Although the findings of this study are interesting, we acknowledge that it has a number of limitations, including (i) a small sample size; (ii) a frame sentence to elicit code-switched productions such that not all the sentences were pragmatically well-formed.
+Although the findings of this study are interesting, we acknowledge that it has a number of limitations, including (i) a small sample size; (ii) a frame sentence for eliciting code-switching such that not all the sentences were pragmatically well-formed.
 This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study; (iii) focus on a single acoustic feature; (iv) the absence of unilingual Bengali data from the same set of speakers as the participants of the study, which would have allowed us to compare each individual's transfer patterns to their own Bengali productions, rather than to average Bengali vowel formants.
@@
 Comparisons of F1 and F2 across conditions reveal that mixed language processing temporarily increases phonetic interaction between the languages, causing the English \nt{\ae} of participants to converge towards the Bengali category \nt{\ae}, and \nt{2} to converge towards a related L1 category \nt{a:}.
-The extent and direction of shift is asymmetrical across the two vowel categories, which parallels findings in prior studies.
+The extent and direction of shift is asymmetrical across the two categories, which parallels findings in prior studies.
 This asymmetry appears to be mediated by the language-specific distribution of vowel categories.
-These findings evidence an effect of phonetic transfer on L2 vowels for the first time, and raise questions for proposed inhibition-based accounts of cross-language phonetic interaction.
+These findings evidence an effect of phonetic transfer on L2 vowels, and raise questions for proposed inhibition-based accounts of cross-language phonetic interaction.
 Mixed productions were elicited in two switching paradigms: code-switching and cued picture-naming, to verify whether the postulated discursive elements in code-switching independently influence the outcome of phonetic transfer.

### 2022-11-05 draft_BLC/BLC_AM-ID_manuscript.tex
reduced word count

@@
 
-Word count (17.10.2022): 9289 \\
+Word count (5.11.2022): 9089 \\
 
@@
 We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in code-switched (mixed) vs.
-non-switched productions.
+nonswitched productions.
 Results reveal asymmetrical shifts in both vowels during mixed productions, towards related L1 categories.
@@
 We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.
-
-Existing studies have studied phonetic transfer along different acoustic features.
-While discussing the literature in subsequent sections, we will often indicate the acoustic feature in parentheses following a citation, for clarity.
+In subsequent sections, we will indicate the acoustic feature in parentheses following a citation, for clarity.
 
@@
 A number of studies have specifically manipulated language mode, both in the presence and absence of switching.
-Overall, results suggest that: (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep[][vowel quality]{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use--- studies comparing switched and non-switched tokens produced in the same test block (identical language mode) \citep[][VOT]{olson2016role,tsui2019impact} or spontaneous conversation \citep[][VOT]{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
+Overall, results suggest that: (i) In the absence of other manipulations, productions in a bilingual language mode show increased cross-language influence compared to a monolingual mode \citep[][vowel quality]{simonet2020increased,simonet2014phonetic}; (ii) However, language mode is not the sole source of influence during mixed language use--- studies comparing switched and nonswitched tokens produced in the same test block (identical language mode) \citep[][VOT]{olson2016role,tsui2019impact} or spontaneous conversation \citep[][VOT]{piccinini2015voice}, have still reported a difference, suggesting that independently of mode, switching between languages triggers a local increase in cross-language transfer.
 (iii) How the two sources interact to influence the final outcome of transfer is not fully understood: \citet[][VOT]{olson2016role} found no additive effects of language mode, \citet[][VOT]{olson2013bilingual} found a balanced language context to inhibit transfer compared to unbalanced contexts.
-Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and non-switched tokens in a monolingual test block, separated by a few hours to days \citep[][vowel quality]{elias2017effects}, \citep[][VOT]{schwartz2015language, bullock2009trying,antoniou2011inter, vsimavckova2015immediate,vsimavckova2018patterns}.
+Other studies have not analyzed the two separately, eliciting switched tokens in a bilingual test block and nonswitched tokens in a monolingual test block, separated by a few hours to days \citep[][vowel quality]{elias2017effects}, \citep[][VOT]{schwartz2015language, bullock2009trying,antoniou2011inter, vsimavckova2015immediate,vsimavckova2018patterns}.
 
-Since participants of this study are in an environment that largely contains mixed-language input and multilingual interlocutors, we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode \citep[see][]{grosjean1998studying}.
-Thus, we elicited all utterances in the same bilingual test block.
-Any observed differences in this paradigm would arguably result from interaction during online processing of sounds while switching between languages.
+Since participants of this study are in an environment containing mixed-language input and multilingual interlocutors, we expect that phonetic transfer during everyday language use takes place in a bilingual mode \citep[see][]{grosjean1998studying}.
+We elicited all utterances in the same bilingual test block.
+Observed differences in this paradigm would arguably result from interaction during online processing of sounds while switching between languages.
 
-In a bilingual mode, both language systems are expected to be nearly equally accessible throughout the test block.
-Thus, a consistent difference between switched and non-switched tokens in such a paradigm would be possible only if the effects were highly localized--- if not, we should expect a gradual convergence over the course of the experiment.
+In bilingual mode, both language systems are expected to be nearly equally accessible throughout the test block.
+Thus, a consistent difference between switched and nonswitched tokens in such a paradigm would be possible only if the effects were highly localized--- if not, we should expect a gradual convergence over the course of the experiment.
 This is discussed in the next section.
@@
 Longitudinal studies and between-subject comparisons of speakers with different durations of L2 exposure suggest that interaction between bilingual sound systems is dynamic \citep[e.g.][]{bohn1992production}.
-Moreover, changes due to transfer are not necessarily unidirectional or irreversible.
-\citet[][VOT]{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language, showing that cross-language interaction can be triggered in an order of months, and also reversed within a similar time range.
-A more recent study by \citet[][VOT]{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks).
+Changes due to transfer are not necessarily unidirectional or irreversible.
+\citet[][VOT]{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language, showing that cross-language interaction can be triggered in an order of months, and reversed within a similar time range.
+A recent study by \citet[][VOT]{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks).
 Based on a short-term longitudinal study, \citet[][VOT]{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2.
@@
 
-The research discussed above concerns situations where a speaker is operating in any one of their languages.
-When it comes to the phonetics of mixed-language use, the majority of existing studies only analyze the switched (target) token.
-Thus, there are few direct measurements of the duration of short-term transfer effects.
+Studies discussed above concern situations where a speaker is operating in any one of their languages.
+For mixed-language use, the majority of existing studies only analyze the switched (target) token--- there are few direct measurements of the duration of short-term transfer effects.
 One study which measured this in a code-switching paradigm \citep[][VOT]{bullock2009trying} did not find any residual effects on the matrix language following a switch, suggesting that transfer during code-switching is localized.
-Indirect evidence in support of this comes from experiments which have elicited switched and non-switched tokens in the same test block and still reported differences between the two \citep[e.g.][VOT]{tsui2019impact,olson2013bilingual}, suggesting that changes due to transfer are quickly `reset'--- in an order of seconds.\\
+Indirect evidence for this comes from experiments that have elicited switched and nonswitched tokens in the same test block and still reported differences between the two \citep[e.g.][VOT]{tsui2019impact,olson2013bilingual}, suggesting that changes due to transfer are quickly `reset'--- in an order of seconds.\\
 Note that all the studies of mixed-language production discussed above measure VOT, which is a temporal feature.
@@
 However, findings from sub-categorical phonetic shifts triggered by other factors (such as convergence towards an interlocutor) do evidence rapid shifts in vowel quality within comparable time-frames \citep[e.g.][]{pardo2010expressing,babel2010dialect,babel2012evidence}.
-In this study, we present switched and non-switched tokens randomly within the same test block to induce a bilingual mode.
+In this study, we present switched and nonswitched tokens randomly within the same test block to induce a bilingual mode.
 We expect the intervening words between two subsequent targets to undo any residual effects of transfer.
@@
 \subsection{Asymmetries between languages in extent and direction of transfer} \label{asymmetries}
-A recurring pattern that emerges from the literature on short-term phonetic transfer is that the extent and patterns of phonetic shift are not equivalent across the two languages of the bilingual speaker.
+A recurring pattern in the literature on short-term phonetic transfer is that the extent and patterns of phonetic shift are not equivalent across the two languages of the bilingual speaker.
 Studies vary greatly in the exact differences they report.
@@
 Flege's Speech Learning Model (SLM) \citeyearpar[][]{flege1995second,flege2007language} proposes that the sound categories of a bilingual speaker exist in a common phonological space--- in principle, both L1 and L2 categories can influence one another.
-Thus, patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories.
-In non-switched production, this is evidenced through pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast \citep[see equivalence classification:][]{flege1984limits,flege1987production}, and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.
+Patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories.
+In nonswitched production, this is evidenced through pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast \citep[see equivalence classification:][]{flege1984limits,flege1987production}, and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.
 
@@
 \subsubsection*{Differences between sound systems of the languages} \label{sound systems} In addition to the cognitive factors discussed above, many researchers have attributed asymmetries to language-internal factors, such as the shape, size, and composition of the sound inventory.
-For example, frequently reported asymmetries between long- and short-lag VOT languages \citep[see, e.g.][]{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid} and postulated frequency effects \citep{antoniou2011inter} highlight that beyond interaction during the cognitive processing of sounds, phonetic transfer is ultimately a linguistic phenomenon, subject to language-specific phonological constraints.
+For example, frequently reported asymmetries between long- and short-lag VOT languages \citep[see, e.g.,][]{tobin2017phonetic, olson2016role,bullock2009trying,antoniou2011inter,chang2012rapid} and postulated frequency effects \citep{antoniou2011inter} highlight that beyond interaction during the cognitive processing of sounds, phonetic transfer is ultimately a linguistic phenomenon, subject to language-specific phonological constraints.
 Therefore, it is imperative to consider data from a variety of language pairs.
@@
 
-We extracted formant values from syllable-initial productions of \nt{\ae} and \nt{a:} at 55\% into the vowel, and compared these to our participants' productions in the non-switched English condition.
+We extracted formant values from syllable-initial productions of \nt{\ae} and \nt{a:} at 55\% into the vowel, and compared these to our participants' productions in the nonswitched English condition.
 This is visualized in figure \ref{vowels_unilingual_ie_bengali}.
@@
 	\item \textbf{RQ1---L2 vowels}: Are L2 vowel categories of proficient bilingual speakers affected by dynamic phonetic transfer during mixed-language production?
-	\item \textbf{H1---L2 vowels}: We expect to see differences in the spectral properties of L2 vowels in switched vs non-switched tokens
+	\item \textbf{H1---L2 vowels}: We expect to see differences in the spectral properties of L2 vowels in switched vs nonswitched tokens
 	\item \textbf{RQ2---Asymmetry}: Do patterns of transfer differ between the two vowel categories?
@@
 \subsection{Participants} \label{participants}
-Participants were recruited through an open call at the EFL University campus in Hyderabad.
-Interested volunteers completed a brief language background questionnaire which was used to screen participants as described below.
+Participants were recruited through an open call at the EFL University campus in Hyderabad. Volunteers completed a brief language background questionnaire, which was used to screen participants.
 
-As noted in section \ref{bengali_english_in_india}, the population is multilingual and linguistically diverse.
-The university campus comprises of students from different parts of the country and world, with varying L1s and different varieties of English.
-Our aim in screening participants was to minimize between-participant random variability, while maintaining ecological validity considering the actual day-to-day settings in which transfer behaviors likely take place.
-We chose two metrics to do this: (i) where the participant grew up; and (ii) amount of formal education in Bengali.
-The first ensures that the variety of Bengali that individuals are exposed to in their surroundings is similar across participants.
-This would mean that the target representations for Bengali vowels would be expected to be similar.
-This was especially important because we were not eliciting unilingual L1 Bengali data in this study, and were using an existing data source to estimate the position of Bengali vowels in the formant space.
-We expect that L2 vowels shift towards related L1 categories during mixed language utterances.
-Thus, we wanted to ensure that any between-participant asymmetries in the pattern of shift reflected individual differences in transfer, rather than different acoustic targets.
-Another reason to control for location was that it is a fairly reliable predictor for language(s) of education(see Supplementary materials for details).
+The university campus has a linguistically diverse population with varying L1s and varieties of English.
+Our aim was to minimize between-participant random variability, while maintaining ecological validity considering the actual day-to-day settings in which transfer behaviors likely take place.
+We chose two metrics for this: (i) where the participant grew up; and (ii) Medium of Instruction (MoI) in school.
+(i) ensures that the variety of Bengali that individuals were exposed to is similar across participants, resulting in similar target representations for Bengali vowel categories.
+This was especially important because our study does not elicit unilingual Bengali vowels, instead using existing corpus data to estimate the position of Bengali vowels in the formant space.
+Since we expect L2 vowels to shift towards related L1 categories during mixed language utterances, we wanted to ensure that any between-participant asymmetries reflected individual differences in transfer, rather than different acoustic targets.
+Another reason to control for location was that it is a fairly reliable predictor for language(s) of education (see supplementary materials for details).
+(ii) reflects sustained formal education in a language, which predicts LSRW (listening, speaking, reading, writing) skills.
+Since this study uses written stimuli, we wanted to ensure that all participants were sufficiently comfortable with reading and speaking both Bengali and English, to avoid variability due to task effort.
 
-The second metric is Medium of Instruction (MoI) in school.
-Sustained formal education in a language is a reliable predictor of LSRW (listening, speaking, reading, writing) skills.
-Since this study uses written stimuli, we wanted to ensure that all participants were sufficiently comfortable with reading and speaking both Bengali and English, to avoid variability due to reading/task effort.
+Given these considerations, we recruited participants who (i) grew up and lived in the state of West Bengal for the majority of their lives; (ii) had Bengali-speaking parents; (iii) received at least 5 years of formal education in Bengali (all respondents had 5+ years of education in English and were pursuing degrees taught in English); and (iv) self-reported LSRW proficiency in Bengali.
 
-Given these considerations, we only recruited participants who had grown up in the state of West Bengal and lived there for the majority of their lives, had parents who both spoke Bengali, and had received at least 5 years of formal education in Bengali (all respondents had received 5+ years of education in English and were pursuing university degrees that are taught in English), and self-reported being proficient in speaking, understanding, and reading Bengali.
-
-A total of 28 bilingual speakers of Bengali and English (18 female) responded to the recruitment call.
-Using the inclusion criteria reported above, 10 volunteers (5 female, 5 male, age range 19 to 28) were invited to participate in a speech production study, and were compensated for their time.
-After completion of the study, we followed up with a detailed language background survey at a later date to learn more about the language profiles of the participants.
-This survey was a modified and consolidated version of three popular bilingualism profiling tools: the Bilingual Language Profile (BLP) \citep{blp}, the Language Experience and Proficiency Questionnaire (LEAP-Q) \citep{leap-q} which incorporates language history and self-assessed proficiency, and the Bilingual Switching Questionnaire (BSWQ) \citep{language_switching_questionnaire}.
-Together, these incorporate questions about language history, usage, attitudes, and switching experience.
+28 bilingual speakers of Bengali and English (18 female) responded to the recruitment call.
+Using the inclusion criteria reported above, 10 volunteers (5 female, 5 male, age range 19--28) participated in the study, and were compensated for their time.
+We followed up with a detailed language background survey at a later date to learn more about the language profiles of the participants.
+This survey was a modified and consolidated version of three popular bilingualism profiling tools: the Bilingual Language Profile (BLP) \citep{blp}, the Language Experience and Proficiency Questionnaire (LEAP-Q) \citep{leap-q}, and the Bilingual Switching Questionnaire (BSWQ) \citep{language_switching_questionnaire}.
+Together, these incorporate questions about language history, usage, self-assessed proficiency, attitudes, and switching experience.
 Table \ref{table_participants} summarizes some key biographical data from the survey.
 The complete questionnaire and responses are available in the Data component of the OSF repository for this project (see section \ref{data_availability}).
-A more detailed discussion of the linguistic situation in India and language profiles of participants is included as Supplementary material.
+A detailed discussion of the linguistic situation in India and language profiles of participants is included as supplementary material.
 
-Given our inclusion criteria, we were able to recruit 10 participants for this study.
+< insert table \ref{table_participants} about here >\\
+
 Although existing studies in the field have used comparable participant numbers, we acknowledge that this is a small sample size.
@@
 Given that small sample sizes are likely to have low statistical power, replication with larger sample sizes are important to confirm the observed effects, and provide much-needed data diversity.
-A more detailed discussion of power analyses in bilingualism research is included as Supplementary material.
-We hope that this can facilitate more conversation around establishing norms for statistical power and sample size in our field.
+A discussion of power analyses in bilingualism research is included as supplementary material.
+%We hope that this can facilitate more conversation around establishing norms for statistical power and sample size in our field.
 
 \subsection{Stimuli}
-Twenty monosyllabic English words, ten containing the vowel \nt{2} and ten containing \nt{\ae}, were selected as target words.
-To minimize coarticulatory effects on the vowel, the pre-vocalic consonants were all voiced plosives (\nt{b} or \nt{d}).
-10 monosyllabic English words that do not contain the target vowel served as filler items.
+Twenty monosyllabic English words, ten each containing the vowels \nt{2} and \nt{\ae}, were selected as target words.
+To minimize coarticulatory effects on the vowel, pre-vocalic consonants were all voiced plosives (\nt{b} or \nt{d}).
 We avoided English words that are already lexicalized loanwords in Bengali.
+10 monosyllabic English words that do not contain the target vowels served as filler items.
 The complete list of target words can be found in the ``Materials" component of the OSF repository for this project (see section \ref{data_availability}).
@@
 Participants were recorded individually in a sound attenuated room.
-Productions were recorded in Praat \citep{boersma2016praat} at a 44.1K sampling rate.
-Each participant completed four iterations each of two tasks--- cued picture-naming and code-switching.
+Productions were recorded in Praat \citep{boersma2016praat} at a 44.1k sampling rate.
+Participants completed four iterations each of two tasks--- cued picture-naming and code-switching.
 Utterances in the two conditions (nonswitch and switch/mixed) were randomly interspersed within each iteration.
@@
 	\end{itemize}
-	The language cue was used to manipulate the word preceding the target, giving target words in non-switched and switched (mixed) utterances.
+	The language cue was used to manipulate the word preceding the target, giving target words in nonswitched and switched (mixed) utterances.
 
@@
 
-Carrier sentences were chosen such that the sound preceding the target word was uniform across all sentences (the final sound in \textit{yellow} is produced as [o] in Indian English), to minimize differences due to coarticulatory effects on the target vowel.
-To ensure identical word order in both languages, we chose a specifier-noun construction, as this has a parallel structure in Bengali and English.
+Carrier sentences were such that the sound preceding the target word was uniform across all sentences (the final sound in ``yellow'' is [o] in Indian English), to minimize differences in coarticulatory effects on the target vowel.
+We chose a specifier-noun construction, which has identical word-order in Bengali and English.
 This gave a uniform prosodic context for the target word across conditions.
@@
 Thus, we expect the interaction of Vowel*Time to predict formant frequency.
-If transfer patterns change across time, then we should expect it to interact with the transfer term, giving a Context*Vowel*Time interaction.
+If transfer patterns change across time, then we should expect Time to interact with the transfer term, giving a Context*Vowel*Time interaction.
 
@@
 
+< Insert Table \ref{table_f1} about here >\\
+
 \subsubsection*{Model interpretation}
-We see that F1 values are significantly predicted by Time in the model, reflecting the dynamic nature of articulation which causes acoustic variation over the course of time.
-The positive slope of the estimate for Time indicates that F1 increases, i.e.
-the two categories in the study show lowering, over the course of articulation.
-Moreover, the significant Time*Vowel interaction shows that the F1 trajectory is different for the two vowels.
-This is expected, as the categories are expected to present different articulatory targets.
-
-A boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two IE categories \nt{2, \ae} and the extent of height difference between them.
+F1 values are significantly predicted by Time in the model, reflecting the dynamic nature of articulation which causes acoustic variation over the course of time.
+%The significant Time*Vowel interaction shows that the F1 trajectory differs between the vowels as expected, since the two categories have different acoustic targets.
+Vowel is not a significant predictor of F1.
+This is initially surprising, given that the phonological category is expected to systematically affect vowel quality.
+However, a boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two IE categories \nt{2, \ae} and the extent of height difference between them.
 These measures are only from the nonswitch condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching.
@@
 vowels are higher, in the nonswitch (e) condition compared to the mixed (b) condition.
-Put more intuitively, vowels are lowered in the mixed condition.
+More intuitively, vowels are lowered in the mixed condition.
 However, the lack of significance indicates that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels.
-There is an asymmetrical shift in the categories \nt{2} and \nt{\ae} in the mixed-language condition, as shown by the significant interaction between Vowel and Context.
+Instead, there is an asymmetrical shift in the categories \nt{2} and \nt{\ae}, as shown by the significant interaction between Vowel and Context.
 Specifically, the Vowel*Context interaction has a negative slope, showing that the lowering is greater in \nt{2}.
@@
 This is surprising, as we have no \textit{a priori} reason to expect a uniform difference in vowel quality across tasks.
-A significant interaction with Time is strongest at the beginning of the vowel, indicating a local effect.
-A significant Vowel*Context*Task interaction indicates a difference in transfer patterns such that transfer in the F1 (height) dimension is greater in the code-switching (s) task compared to the picture-naming (p) task.\\
+A significant interaction with Time shows that this effect is strongest at the beginning of the vowel and decreases with time, indicating a local effect.
+The Vowel*Context*Task interaction is of interest, as it relates to our research question of how the test paradigm affects transfer.
+The significant effect of this interaction indicates a difference in transfer patterns such that transfer in the F1 (height) dimension is greater in the code-switching (s) task compared to the picture-naming (p) task.\\
 
 < Insert Figure \ref{boxplot_F1} about here >\\
-
-< Insert Table \ref{table_f1} about here >
 
@@
 
+< Insert Table \ref{table_f2} about here >\\
+
 \subsubsection*{Model interpretation}
@@
 As with F1, we find that Time significantly influences F2 values as expected.
-Unlike in the F1 dimension, Vowel is a significant predictor of F2, with \nt{2} having significantly lower F2, i.e.
-being produced as more backed, than \nt{\ae}.
-There is no significant independent effect of Context, showing that mixed language use does not lead to uniform fronting or backing across vowels.
-This is not surprising, as we expect movement to be towards the corresponding L1 category, and thus differ across vowels.
-The negative slope of the estimate indicates that on average, vowels are fronted in the mixed condition.
-This effect is greater in \nt{\ae}.
+Vowel is a significant predictor of F2, with \nt{2} having lower F2, i.e.
+being more backed, than \nt{\ae}.
+The negative slope of the estimate for Context indicates that on average, vowels are fronted in the mixed condition.
+The lack of significance for this effect shows that mixed productions do not lead to uniform fronting or backing across vowels.
+This aligns with our expectation for movement to be towards the corresponding L1 category, and thus differ across vowels.
+Instead, the significant Vowel*Context interaction indicates that backing is greater in \nt{\ae}.
 The lack of a significant Vowel*Context*Time interaction shows that transfer patterns do not change across time, suggesting that the underlying acoustic targets are different in the two conditions.
@@
 The significant interaction with Time shows that this effect is local and decreases over time.
-The Vowel*Context*Task interaction is of interest, as it relates to our research question of how the test paradigm affects transfer.
-Unlike vowel height, we do not find patterns of shift in vowel backness (F2) to differ across paradigms, as indicated by the lack of a significant effect for the fixed effect.
-
-< Insert Table \ref{table_f2} about here >
+Unlike vowel height, we do not find patterns of shift in vowel backness (F2) to differ across paradigms, as indicated by the lack of a significant effect for the Vowel*Context*Task interaction.
 
@@
 
-Statistical analyses reveal that the vowel \nt{2} is lowered and fronted, while \nt{\ae} is fronted, in the mixed condition.
+Statistical analyses reveal that \nt{2} is lowered and fronted, while \nt{\ae} is fronted, in the mixed condition.
 As discussed in section \ref{asymmetry between sounds}, we wanted to understand how the direction of this shift compares to corresponding Bengali categories.
 Figure \ref{vowels_unilingual_ie_bengali} shows the non-switched IE productions of the participants in relation to average positions of related vowel categories in Bengali.
-Figure \ref{vowels_e_b} overlays productions in the mixed condition over figure \ref{vowels_unilingual_ie_bengali}, showing F1$\times$F2 vowel plots for IE vowel productions in the nonswitch and mixed conditions, in comparison to the related Bengali categories from the SHRUTI corpus \citep{shruticorpus}.
+Figure \ref{vowels_e_b} overlays productions in the mixed condition over figure \ref{vowels_unilingual_ie_bengali}, showing F1$\times$F2 plots for IE vowel productions in the nonswitch and mixed conditions, in comparison to the related Bengali categories from the SHRUTI corpus \citep{shruticorpus}.
 Given that the Bengali corpus data represent average vowel positions, and are not from the same set of speakers as the participants of this study, we refrained from including these in statistical analyses.
@@
 
-In this study, we presented nonswitched English and code-switched Bengali-English stimuli in the same session and found consistent differences between switched and non-switched tokens.
+In this study, we presented nonswitched English and code-switched Bengali-English stimuli in the same session and found consistent differences between switched and nonswitched tokens.
 This confirms that the local act of switching does trigger transfer, even in a bilingual language mode, paralleling results from \cite{olson2016role, tsui2019impact, piccinini2015voice}.
@@
 As discussed in section \ref{causes}, vowels provide a fertile testing ground for this theory at the level of sounds, as they avoid the confound of independent language-specific differences between long- and short-lag VOT languages.
-Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline non-switched productions.
+Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline nonswitched productions.
 In the context of an ICM-based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, this is a significant finding: it demonstrates that the postulated lower switch cost on an L2 does not entirely mask transfer effects on L2 vowels.
@@
 The materials, data files, and analysis code for this study are openly available in OSF at:  \url{https://doi.org/10.17605/OSF.IO/DSB2X}.
+
+\section*{Supplementary material}
+All supplementary material accompanying this paper can be found at \url{https://osf.io/2q3ga/?view_only=d0f3cd8f38a1461fab738fd0e698f6dc}.
 

### 2022-11-07 draft_BLC/BLC_AM-ID_manuscript.tex
reduced word count: discussion

@@
 
-Word count (5.11.2022): 9089 \\
+Word count: \\
 
@@
 
-The aim of this study was to examine short-term phonetic transfer in L2, towards understanding language processing in proficient bilinguals.
-Two interrelated questions, discussed in section \ref{introduction}, bear on this: (i) what causes transfer in mixed-language speech; and (ii) what causes observed asymmetries between transfer in L1 vs L2.
+This study examined short-term phonetic transfer in L2, towards understanding language processing in proficient bilinguals.
+Two interrelated questions (section \ref{introduction}), bear on this: (i) what causes transfer in mixed-language speech; (ii) what causes observed asymmetries between transfer in L1 vs L2.
 
@@
 
-In this study, we presented nonswitched English and code-switched Bengali-English stimuli in the same session and found consistent differences between switched and nonswitched tokens.
-This confirms that the local act of switching does trigger transfer, even in a bilingual language mode, paralleling results from \cite{olson2016role, tsui2019impact, piccinini2015voice}.
+We presented nonswitched English and code-switched Bengali-English stimuli in the same test block and found consistent differences in vowel quality.
+This confirms that the local act of switching triggers transfer, even in a bilingual language mode, paralleling results from \citet{olson2016role, tsui2019impact, piccinini2015voice}.
 These studies focused on a temporal feature, VOT.
-Our results indicate that the same happens in vowels.
-The exact effect of transfer is interesting: since VOT is a one-dimensional measure, observing a change in VOT in a mixed language condition makes it difficult to infer whether a consonant has moved towards a specific L1/L2 counterpart, or reflects a more general system-wide shift in articulatory norms.
-Vowel quality measures, showing both directionality and magnitude, facilitate this.
+Our results demonstrate a similar effect in vowels.
+Since VOT is a one-dimensional measure, a difference across conditions is insufficient to infer whether the category has shifted towards a specific L1/L2 counterpart, or reflects a more general system-wide shift in articulatory norms.
+Vowel quality measures, having both directionality and magnitude, enable this.
 We find that productions of English vowels do not merely shift towards a Bengali counterpart, but rather towards a set of Bengali-like \textit{norms}: a vowel category that doesn't exist in Bengali nonetheless moves towards a related (perceptually close) Bengali category.
-This parallels results from \cite{simonet2014phonetic} for Catalan and Spanish (they examined language mode, without code-switching).
+This parallels findings from \cite{simonet2014phonetic} for Catalan and Spanish (they manipulated language mode, without code-switching).
 The \cite{simonet2014phonetic} study showed that co-activating multiple languages within a discourse causes a transient increase in cross-language phonetic interaction, leading to changes in both L1 and L2.
@@
 The time course of such shifts is telling: we see a difference between conditions right from the start of articulation, suggesting different acoustic targets.
-This suggests that such interaction involves the speech planning stage.
-The pattern of transfer observed in this study, particularly for \nt{2}, also demonstrates that beyond a simple L2-L1 mapping, transfer during online processing makes reference to entities beyond the level of the individual sound, demonstrating system-level interaction: a sound that does not exist in the L1 repertoire nonetheless acquires a different target, and the target is related to the L1 system in a meaningful way.
+This indicates that such interaction involves the speech planning stage.
+The pattern of transfer observed in this study, particularly for \nt{2}, also demonstrates that beyond a simple L2-L1 mapping, transfer during online processing makes reference to entities above the level of individual sounds, demonstrating system-level interaction: a sound that does not exist in the L1 repertoire nonetheless acquires a different target, and the target is related to the L1 system in a meaningful way.
 This cannot be reasonably explained as articulatory inertia carried over from uttering L1 sounds.
 It is also difficult to account for in a purely exemplar framework, as the discourse does not contain any instances of ``L1 \nt{2}" that could affect the L2 category.
-If representations are combined in creative ways during online speech processing within an utterance, then can we think of intermediate representations not just for individual sounds, but rather for sound systems?
-While more detailed empirical data is needed to answer such questions, this approach aligns with sociophonetic understandings of bilingual phonetic variation, for example as discussed by \cite{khattab2009phonetic,khattab2013phonetic} based on the use of phonological variables by Arabic-English bilingual children.
-They visualize the bilingual phonetic system as ``a rich store of phonological representations", comprising of both language-internal and cross-language variation, from which the bilingual speaker is constantly choosing, guided by a variety of pressures.
+If representations are combined in creative ways during online speech processing, then can we think of intermediate representations not just for individual sounds, but rather for sound systems?
+While more detailed empirical data is needed to answer such questions, this approach aligns with sociophonetic understandings of bilingual phonetic variation, e.g.
+as discussed by \cite{khattab2009phonetic,khattab2013phonetic} based on the use of phonological variables by Arabic-English bilingual children.
+These visualize the bilingual phonetic system as ``a rich store of phonological representations", comprising of both language-internal and cross-language variation, from which the bilingual speaker is constantly choosing, guided by a variety of pressures.
 We propose that vowels are a good feature to examine the plausibility of such systems as cognitive models for language processing: in addition to continuous measures (unlike phonological variables), they also provide two-dimensional data (unlike VOT) that is suited to further test and refine predictions from such a model.
@@
 \paragraph{Differences between L1 vs L2 transfer:}
-Previous studies of language-switching have often reported phonetic transfer from L2 to L1, but a lack of parallel transfer from L1 to L2.
-One postulated model that is consistent with this finding is the Inhibitory Control Model (ICM) \cite{green1986control,olson2013bilingual,tsui2019impact}.
-This proposes that the lack of transfer effects on L2 reflects the lower inhibition required on L2 representations in the first place, which makes switching into L2 easier, compared to the converse situation, when the high initial inhibition on L1 induces phonetic transfer while switching back into L1.
-As discussed in section \ref{causes}, vowels provide a fertile testing ground for this theory at the level of sounds, as they avoid the confound of independent language-specific differences between long- and short-lag VOT languages.
+Existing studies on language-switching have often reported phonetic transfer from L2 to L1, no parallel transfer from L1 to L2.
+One postulated model consistent with this finding is the Inhibitory Control Model (ICM) \cite{green1986control,olson2013bilingual,tsui2019impact}.
+%This proposes that the lack of transfer effects on L2 reflects the lower initial inhibition required on L2 representations, which makes switching into L2 easier, compared to the converse situation, where the high initial inhibition on L1 induces phonetic transfer while switching back into L1.
+As discussed in section \ref{causes}, vowels provide a fertile testing ground for this theory at the level of sounds.
 Both vowels in this study showed systematic shifts in the mixed-language condition compared to baseline nonswitched productions.
-In the context of an ICM-based analysis of between-language asymmetries in phonetic transfer as proposed by \cite{olson2013bilingual}, this is a significant finding: it demonstrates that the postulated lower switch cost on an L2 does not entirely mask transfer effects on L2 vowels.
+In the context of an ICM-based analysis of between-language asymmetries in phonetic transfer, as proposed by \cite{olson2013bilingual}, this demonstrates that the postulated lower switch cost on L2 does not entirely mask transfer effects on L2 vowels.
 There are a few possible explanations for this.
@@
 That would mean that switching from L1 into L2 induces a greater switch cost, leading to transfer effects on switched L2 productions.
-This would lead us to expect that if the same participants are asked to switch from English to Bengali, we would not see any evidence of transfer.
-While more empirical data could clarify this, the language background of the present participant pool makes this scenario unlikely.
-All participants report regularly using both English and Bengali in a variety of settings, and feeling equally comfortable using both languages.
+We should then expect a lack of transfer while switching from English to Bengali.
+While more empirical data could clarify this, the language background of these participants makes this scenario unlikely.
+All individuals report regularly using both English and Bengali in a variety of settings, and feeling equally comfortable using both languages.
 Listening and speaking skills are expected to be most relevant to phonetic transfer in daily language use, and participants' self-ratings for these skills are at ceiling for both languages.
-One point to note is that participants on average rated their reading proficiency in Bengali to be lower than that in English.
+One point to note is that participants on average rated their reading proficiency in Bengali to be lower than in English.
 Given that the stimuli in this task were presented visually, differences in reading proficiency are a potential source of explanation.
@@
 Moreover, our findings differ from \cite{tsui2019impact}, who report on the basis of VOT measurements that balanced bilinguals do not evidence an effect of language-switching.
-\citeauthor{tsui2019impact} attribute the absence of transfer effects to greater experience with language switching in these participants, leading to lower switch-costs in both languages.
+\citeauthor{tsui2019impact} attribute this to greater language-switching experience in these participants, leading to lower switch-costs in both languages.
 As noted in sec.
-\ref{participants}, the participants in this study are highly proficient bilinguals with extensive experience of language switching.
-Moreover, they live in an environment where they are constantly exposed to mixed-language stimuli: we expected that they are habituated to drawing on multiple language systems during online language processing.
+\ref{participants}, participants in this study are highly proficient bilinguals with extensive experience of language switching.
+Moreover, their environment constantly exposes them to mixed-language stimuli: we expect that they are habituated to drawing on multiple language systems during online language processing.
 They nonetheless demonstrate cross-language transfer effects in L2 vowel production.
-One way to think about these differences is that vowels are generally more prone phonetic transfer than temporal features such as VOT.
-Thus, following from the discussion in sec.
-\ref{asymmetries}, it might be fruitful for future studies to focus on vowels as a potential site for transfer.
-This would both expand the empirical scope of existing research (see sec.
-\ref{asymmetry between sounds}) and avoid confounding factors arising from independent differences between long- and short-lag VOT languages.
-More generally, these findings point to a need for expanding the range of features in which transfer is studied, in order to make robust generalizations about the mechanisms underlying observed patterns.
-This is an example where studying different features of speech (e.g.
-VOT vs vowel quality) allows us to compare the predictions of different underlying process models.
-Transfer studies usually focus on a single feature.
-Measuring multiple features within a single design, with the same group of participants, promises to be a powerful tool to compare different models at little added cost to researchers.
-This would sidestep the unavoidable confounds that come with comparing results across studies, or even replication.
+There are a few different ways to approach this:
 
-Beyond methodological considerations, we argue that these differences ultimately raise a broader question about a purely inhibition-based understanding of transfer.
-In spite of a compelling body of literature on the sociophonetic and communicative implications of phonetic variation along the L1-L2 continuum \citep[e.g.][]{khattab2013phonetic,agnihotri1979processes,stuart2011hybridity}, is it wise to think of transfer as resulting from an inability to suppress and separate language systems?
-Given that young children are able to use minute phonetic variation in sophisticated ways, it might be more intuitive to think of transfer as a communication-enabling feature of language, even at a purely cognitive level.
+One possible explanation is that vowels are generally more susceptible to phonetic transfer than temporal features such as VOT.
+Following from the discussion in sec.
+\ref{asymmetries}, it might therefore be fruitful for future studies to focus on vowels as the site for transfer.
+More generally, these findings point to a need for expanding the range of features studied, in order to make robust generalizations about the mechanisms underlying observed patterns of transfer.
+This is an example where examining different features of speech (e.g.
+VOT vs.
+vowel quality) allows us to compare predictions from different underlying process models.
+Measuring multiple features within a single study promises to be a powerful tool for interpretation and theorizing, at little added cost to researchers.
+This would sidestep the unavoidable confounds that come with comparing results across studies.
+
+Beyond methodological considerations, however, we argue that these differences ultimately raise broader questions about a purely inhibition-based understanding of transfer.
+Given the compelling body of literature on the socio-communicative significance of phonetic variation along the L1-L2 continuum \citep[e.g.][]{khattab2013phonetic,agnihotri1979processes,stuart2011hybridity}, is it wise to think of transfer as resulting from an inability to suppress and separate language systems?
+Seeing as even young children are capable of using minute phonetic variation in sophisticated ways, it might be more intuitive to think of transfer as a communication-enabling feature of language, even at a purely cognitive level.
 Research on the cognitive mechanisms underlying cross-language interaction and variation in bilingual speech is still an emerging and exciting line of inquiry.
 We believe that a communication-centered approach to bilingual variation is essential to understand how bilingual language processing functions in the real world.
-To this end, highly proficient native multilingual populations serve as an important source of data on language behavior.
+To this end, highly proficient native multilingual populations serve as an important source of data.
 
@@
 
-Our results indicate differing effects of experimental conditions across the two vowels in both magnitude and direction of shift.
+Our results indicate differences across the two vowels in both magnitude and direction of shift.
 Overall, shift is greater in \nt{2}.
@@
 Specifically: \nt{\ae} is already at the lowermost part of the vowel space.
-Moreover, large movements upwards in the F1 (vowel height) dimension would risk a conflation of the category with the neighboring \nt{E}.
-In contrast, \nt{2} (indicated as \nt{@} in the chart) is in the central part of the vowel space and is not flanked by any immediate neighboring categories.
+Large movements upwards in the F1 (vowel height) dimension would risk a conflation of the category with the neighboring \nt{E}.
+In contrast, \nt{2} (indicated as \nt{@} in the chart) is in the central part of the vowel space and not flanked by any immediate neighboring categories.
 Thus, the difference in magnitude of transfer aligns with our hypothesis \textbf{H2---Asymmetry}, that the extent of shift in individual categories is constrained by the linguistic principle of contrast preservation.
 An alternative way to interpret this result is that the initial distance between IE \nt{2} and its target L1 category \nt{a:} was greater than that between the IE and Bengali \nt{\ae} categories, leading to a greater visible shift in the former.
-However, our results find parallels in previous studies with VOT that have consistently observed greater shifts in the long-lag VOT language (which has a greater acceptable range of VOT) compared to the short-lag language \citep[e.g.][]{antoniou2011inter, bullock2009trying}.
+However, our results find parallels in previous studies with VOT that have consistently observed greater shifts in the long-lag VOT language (greater acceptable range of VOT) compared to the short-lag language \citep[e.g.][]{antoniou2011inter, bullock2009trying}.
 This suggests that a general principle of ``room for movement" does play a role in moderating transfer.
-More broadly, these results suggest that phonetic transfer (i) makes reference to system-level representations that go beyond individual sounds; and (ii) respects linguistic principles.
+These results suggest that phonetic transfer (i) makes reference to system-level representations that go beyond individual sounds; and (ii) respects linguistic principles.
 This has implications both for our understanding of phonetic transfer as a phenomenon (e.g.
@@
 Our results also indicate asymmetry in the direction of shift: the category \nt{2} is lowered and fronted in the mixed language condition, while \nt{\ae} is fronted.
-On the basis of previous literature indicating that both long-term and short-term shifts in phonological categories of multilingual speakers are targeted with respect to the other language, we hypothesized that participants' IE \nt{\ae} would shift towards the Bengali \nt{\ae}, and \nt{2} towards the Bengali \nt{a:}.
+On the basis of previous literature indicating that both long-term and short-term shifts in categories of multilingual speakers are targeted with respect to the other language, we hypothesized that participants' IE \nt{\ae} would shift towards the Bengali \nt{\ae}, and \nt{2} towards the Bengali \nt{a:}.
 As figure \ref{vowels_e_b} indicates, comparing transfer patterns with Bengali categories from \cite{shruticorpus} suggests that this is borne out.
@@
 
-\cite{olson2013bilingual} posits that in both scripted and spontaneous code-switching paradigms, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", proposing cued picture-naming as an alternative.
+\cite{olson2013bilingual} posits that in both scripted and spontaneous code-switching, ``discursive properties may partially drive the production of code-switches, masking underlying effects of interaction", proposing cued picture-naming as an alternative.
 We tested whether this might independently affect the outcome of transfer, by comparing productions from the same participants in these two tasks.
-We found that the direction of transfer did not change across paradigms.
+The direction of transfer did not change across paradigms.
 However, there was a difference in the magnitude of transfer: in the F1 dimension, the extent of phonetic transfer is greater in the code-switching task compared to the cued picture-naming task.
 This asymmetry is not seen in the F2 dimension.
-One way to account for this is that during code-switching, there is a greater degree of language co-activation because switches are being made within an utterance, where planning is required: Since participants see the entire utterance and then read it aloud, there is room for utterance planning.
-Compared to this, in the cued picture-naming paradigm participants see the language cue immediately prior the target word, inducing extremely ``local" switches and thus a smaller degree of language co-activation.
+It is possible that code-switching induces a greater degree of language co-activation because switches are being made within an utterance, which requires planning: participants see the entire sentence and then read it aloud.
+Compared to this, the cued picture-naming paradigm presents the language cue immediately prior to the target word, inducing extremely ``local" switches and thus a smaller degree of language co-activation.
 While both of these are local switches and take place in a bilingual language mode, it is possible that this minute difference in time-frame induces slightly different degrees of cross-language phonetic interaction.
 This aligns with the findings of \cite{bullock2009trying}, who found evidence of transfer in pre-switch tokens, suggesting planning effects.
-In our study specifically, orthography was another factor that could have potentially affected this.
-In both tasks, Bengali and English words were presented in their respective orthographies.
+In our study specifically, orthography was another factor that might have affected this.
 In the picture-naming task, participants saw one word at a time.
 The CS task in contrast presented mixed utterances in mixed orthography.
-It is possible that seeing mixed orthography, particularly for participants accustomed to processing mixed-language visual input in their surroundings, triggered increased language co-activation, leading to greater cross-language interaction.
+It is possible that seeing mixed orthography, particularly for individuals accustomed to processing mixed-language visual input in their surroundings, triggered increased language co-activation, leading to greater cross-language interaction.
 
-We do not find any differences in the pattern of transfer, or the relative behaviors of the two categories.
+We do not find any differences in the pattern of transfer across paradigms.
 Our study used a frame sentence to induce code switching, and not all the items led to pragmatically meaningful sentences.
@@
 More within-study comparisons with less structured paradigms, and spontaneous code-switching data, can clarify this.
-The results of the present study indicate that for the purposes of observing dynamic phonetic transfer in an experimental setup, elicitation through cued-picture-naming vs.
+The results of this study indicate that for the purposes of observing dynamic phonetic transfer in an experimental setup, elicitation through cued-picture-naming vs.
 (controlled) code-switching does not significantly alter the behavior under study.
@@
 There is no principled reason to expect this.
-A closer look at the dynamics of this effect indicate that it is greatest at the beginning of articulation, and decreases steadily with time.
-Given that it is localized at the beginning of the target vowel, it seems likely that this lowering is an artifact of the particular stimuli used here--- resulting from some corarticulatory effect with the preceding word in the carrier phrase that was absent in the picture-naming task.
+A closer look at the dynamics of this effect indicate that it is greatest at the beginning of articulation, and decreases with time.
+Given that it is localized at the beginning of the target vowel, it likely an artifact of the particular stimuli used here.
 While intriguing, a more detailed investigation of this effect is beyond the scope of the present study.
@@
 Although the findings of this study are interesting, we acknowledge that it has a number of limitations, including (i) a small sample size; (ii) a frame sentence for eliciting code-switching such that not all the sentences were pragmatically well-formed.
-This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study; (iii) focus on a single acoustic feature; (iv) the absence of unilingual Bengali data from the same set of speakers as the participants of the study, which would have allowed us to compare each individual's transfer patterns to their own Bengali productions, rather than to average Bengali vowel formants.
+This could have blocked some of the discursive effects that might result from speech planning in ordinary connected speech, thus obscuring differences between the two paradigms under study; (iii) focus on a single acoustic feature; (iv) the absence of unilingual Bengali data from the same set of speakers as the participants of the study, which would have allowed us to compare each individual's transfer patterns to their own Bengali productions.
 Future work should address these.
-Our results also raise a number of open questions for future research, particularly concerning the paradigm differences and between-sound asymmetries in phonetic transfer.
+Our results also raise a number of open questions for future research, particularly concerning paradigm differences and between-sound asymmetries in phonetic transfer.
 
 \section{Conclusion}
-This study investigated how the phonetic systems of highly proficient bilingual speakers interact during online speech processing, by examining the phonetic outcome of mixed-language production in speakers of Bengali and Indian English.
-Specifically, we examined the spectral quality of two L2 vowel categories \nt{2} and \nt{\ae} produced in two experimental conditions: nonswitch (English) and switch/mixed (switching from Bengali to English).
+This study investigated how the sound systems of highly proficient bilingual speakers interact during online speech processing, by investigating the phonetic outcome of mixed-language production in speakers of Bengali and Indian English.
+We examined the spectral quality of L2 vowel categories \nt{2} and \nt{\ae} produced in two experimental conditions: nonswitch (English) and switch/mixed (switching from Bengali to English).
 Comparisons of F1 and F2 across conditions reveal that mixed language processing temporarily increases phonetic interaction between the languages, causing the English \nt{\ae} of participants to converge towards the Bengali category \nt{\ae}, and \nt{2} to converge towards a related L1 category \nt{a:}.
@@
 Mixed productions were elicited in two switching paradigms: code-switching and cued picture-naming, to verify whether the postulated discursive elements in code-switching independently influence the outcome of phonetic transfer.
-Results suggest that in an experimental setup, elicitation through cued-picture-naming vs.
-(controlled) code-switching does not significantly alter the behavior under study.
+Results suggest that in an experimental setup, paradigm does not significantly alter the behavior under study.
 However, the magnitude of observed transfer is greater in the code-switching paradigm.
-These findings add to empirical data, highlight the need for diversifying the set of language pairs and populations in transfer studies, and encourage more within-study comparisons of paradigms and features, towards establishing better methodological norms.
+These findings add to empirical data, highlight the need for diversifying the set of language pairs and populations in transfer studies, and encourage more within-study comparisons of paradigms and acoustic features, towards establishing better methodological norms.
 Ultimately, these results contribute to ongoing discussion about the implications of short-term phonetic interaction as a window to understanding language as a mental and social system.

### 2022-11-07 draft_BLC/extra_text_from_manuscript/extra.tex
reduced word count
reduced word count: discussion

@@
 As expected, w
+
+The significant Time*Vowel interaction shows that the F1 trajectory differs between the vowels as expected, since the two categories have different acoustic targets.
 
@@
 If such discursive properties independently affect the phonetic outcome of language mixing in an experimental setup, this calls to question the comparability of results across studies using these different paradigms.
+
+The exact nature of transfer is interesting: since VOT is a one-dimensional measure, observing a change in VOT in a mixed language condition makes it difficult to infer whether a consonant has moved towards a specific L1/L2 counterpart, or reflects a more general system-wide shift in articulatory norms.
+Vowel quality measures, showing both directionality and magnitude, facilitate this.
+We find that productions of English vowels do not merely shift towards a Bengali counterpart, but rather towards a set of Bengali-like \textit{norms}: a vowel category that doesn't exist in Bengali nonetheless moves towards a related (perceptually close) Bengali category.
+This parallels results from \cite{simonet2014phonetic} for Catalan and Spanish (they examined language mode, without code-switching).
+The \cite{simonet2014phonetic} study showed that co-activating multiple languages within a discourse causes a transient increase in cross-language phonetic interaction, leading to changes in both L1 and L2.
+This is interpreted as a follows: a bilingual mode temporarily co-activates both language systems, making both L1 and L2 representations available during processing, whose interactions lead to cross-language transfer.
+But what does this interaction look like, and how are these abstract representations processed during speech?
+Examining transfer during switching can provide insights into this.
+
+This proposes that the lack of transfer effects on L2 reflects the lower initial inhibition required on L2 representations, which makes switching into L2 easier, compared to the converse situation, where the high initial inhibition on L1 induces phonetic transfer while switching back into L1.
